### 2023-03-29 ARB/standards/work-in-progress/document/tex/boilerplate/about.tex
add tex subdir

\def\tightlist{}
\chapter*{Introduction}
\section*{History of the International \rexx{} Language Symposium}
In 1990, Cathie Dager of SLAC\footnote{Stanford Linear Accelerator Center, since 2008 SLAC National Accelerator Laboratory} convened the organizing committee for the first independent \rexx{}\footnote{Cowlishaw, M. F., \textbf{The REXX Language} (second edition), ISBN 0-13-780651-5, Prentice-Hall, 1990.} Symposium for Developers and Users. SLAC continued to organize this annual event until the middle of the 1990's when the \rexx{}LA took over that responsibility. Symposia have been held annually since 1990.
\section*{About RexxLA}
During the 1993 Symposium in La Jolla, California, plans for a \rexx{} User Group materialized. The \rexx{} Language Association (\rexx{}LA), as it was called, is an independent, non-profit organization dedicated to promoting the use and understanding of the \rexx{} programming language. \rexx{}LA manages several open source implementations of \rexx{}.
\section*{The selection procedure}Presentation proposals are sollicited yearly using a CFP\footnote{Call For Papers.} procedure, after which the RexxLA symposium comittee reviews them and votes which presentations are selected for the symposium. The presentations are peer reviewed before being presented. Presenters are not compensated for their presentations.
\input{symposium}

### 2023-03-29 ARB/standards/work-in-progress/document/tex/boilerplate/acknowledgements.tex
add tex subdir

\small
\chapter*{Acknowledgments}
\begin{multicols*}{2}
RexxLA thanks everybody who has been involved with the \rexx{} symposia over more than 30 years. Mike Cowlishaw for being the Father of \rexx{}, Rick McGuire for being the driving force behind all IBM implementations of \rexx{}, Chip Davis who is Past-President for Life, Mark Hessling for maintaining Regina and building our wonderful symposium application and web site (which enable this publication). Simon Nash for being the architect of Oryx, which became Object Rexx. Jon Wolfers for all his work in converting the website to the database-driven system, and being our treasurer. Gil Barmwater for being our vice-president and treasurer for many years. Rony Flatscher for his teaching and enthousiasm, and for connecting ooRexx to the Java Virtual Machine and its libraries. All the presenters for investing their time to deliver the message. Terry Fuller for being the current vice-president and his help during the symposia. Lee Peedin for his work as President and Vice President for many years. Cathie Dager for being a \rexx{} ``Spark Plug'', Pam Taylor for all her work for RexxLA. Virgil Hein and Matthew Emmons for being our representatives within IBM and assisting us through the open sourcing of the IBM products. We thank IBM for graciously donating two of their \rexx{} implementations, Object Rexx and NetRexx, to RexxLA and the open source community. We thank Per-Olov Jonsson for financing and running our automated test facility - and for all his work on that very important environment. David Ashley his contributions in the early days of ``Open'' Object Rexx. Marc Remes for saving NetRexx from the Java Platform Modules System. Erich Steinbock and Jean-Louis Faucher for their contributions (and future contributions) to ooRexx. 
\\[0.5cm]
RexxLA remembers our departed, Kurt Maerker, Brian Marks, Les Koehler, Mark Miesfeld, Oliver Sims and Kermit Kiser, for all their work and pleasant company.
\end{multicols*}

### 2023-03-29 ARB/standards/work-in-progress/document/tex/boilerplate/bookmeta.tex
add tex subdir

\section*{Publication Data}
\textcopyright  Copyright The Rexx Language Association, \splice{nrc -noverbose -exec ../boilerplate/TexYear}
%\\

All original material in this publication is published under the
Creative Commons - Share Alike 3.0 License as stated at
\url{https://creativecommons.org/licenses/by-nc-sa/3.0/us/legalcode}.\\[0.5cm]

A publication of \textbf{RexxLA Press}\\[0.5cm]
The responsible publisher of this edition is identified as \emph{IBizz
IT Services and Consultancy}, Amsteldijk 14, 1074 HR Amsterdam, a
registered company governed by the laws of the Kingdom of The
Netherlands.\\[1cm]
The RexxLA Symposium Series is registered under ISSN 1534-8954\\
The \edition{} edition is registered under ISBN \isbn \\[1cm]
\psset{unit=1in}
\begin{pspicture}(3.5,1in)
  \psbarcode{1534-8954}{includetext guardwhitespace}{issn}
\end{pspicture}
\begin{pspicture}(3.5,1in)
  \psbarcode{\isbn}{includetext guardwhitespace}{isbn}
\end{pspicture}
%\section*{Publication History}
\\[0.5cm]
\begin{description}
\item[2023-03-14] First printing
\end{description}
\newpage
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2023-03-29 ARB/standards/work-in-progress/document/tex/standard/foreword.tex
tex skeleton

\hypertarget{foreword}{%
\chapter{Foreword}\label{foreword}}

\hypertarget{purpose}{%
\section{Purpose}\label{purpose}}

This standard provides an unambiguous definition of the programming
language Rexx. Its purpose is to facilitate portability of Rexx programs
for use on a wide variety of computer systems. History The computer
programming language Rexx was designed by Mike Cowlishaw to satisfy the
following principal aims: - to provide a highly readable command
programming language for the benefit of programmers and program readers,
users and maintainers; - to incorporate within this language program
design features such as natural data typing and control structures which
would contribute to rapid, efficient and accurate program development; -
to define a language whose implementations could be both reliable and
efficient on a wide variety of computing platforms.

In November, 1990, X3 announced the formation of a new technical
committee, X3J18, to develop an American National Standard for Rexx.
This standard was published as ANSI X3.274-1996.

The popularity of ``Object Oriented'' programming, and the need for Rexx
to work with objects created in various ways, led to Rexx extensions and
to a second X3J18 project which produced this standard. (Ed - hopefully)

\hypertarget{committee-lists}{%
\section{Committee lists}\label{committee-lists}}

(Here)

This standard was prepared by the Technical Development Committee for
Rexx, X3J18. There are annexes in this standard; they are informative
and are not considered part of this standard.

Suggestions for improvement of this standard will be welcome. They
should be sent to the

Information Technology Industry Council, 1250 Eye Street, NW, Washington
DC 20005-3922.

This standard was processed and approved for submittal to ANSI by the
Accredited Standards Committee on Information Processing Systems, NCITS.
Committee approval of this standard does not necessarily imply that all
committee members voted for its approval. At the time it approved this
standard, the NCITS Committee had the following members:

To be inserted The people who contributed to Technical Committee J18 on
Rexx, which developed this standard, include:

### 2023-05-01 ARB/standards/work-in-progress/document/tex/standard/foreword.tex
some standard formatting work

@@
 programming language Rexx was designed by Mike Cowlishaw to satisfy the
-following principal aims: - to provide a highly readable command
-programming language for the benefit of programmers and program readers,
-users and maintainers; - to incorporate within this language program
-design features such as natural data typing and control structures which
-would contribute to rapid, efficient and accurate program development; -
+following principal aims:
+
+\begin{itemize}
+\item
+  to provide a highly readable command programming language for the
+  benefit of programmers and program readers, users and maintainers;
+\item
+  to incorporate within this language program design features such as
+  natural data typing and control structures which would contribute to
+  rapid, efficient and accurate program development;
+\item
   to define a language whose implementations could be both reliable and
   efficient on a wide variety of computing platforms.
+\end{itemize}
 

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/annexb.tex
complete, but very deficient, extended standard

\hypertarget{annex-b}{%
\chapter{Annex B}\label{annex-b}}

(informative)

Method of definition

This annex describes the methods chosen to describe Rexx for this
standard.

Definitions

Definitions are given for some terms which are both used in this
standard and also may be used elsewhere. This does not include names of
syntax constructions; for example, group, which are distinguished in
this standard by the use of italic font.

Conformance

Note that irrespective of how this standard is written, the obligation
on a conforming processor is only to achieve the defined results, not to
follow the algorithms in this standard.

Notation

The notation used to describe functions provided by the configuration is
like a Rexx function call but it is not defined as a Rexx function call
since a Rexx function call is described in terms of one of these
configuration functions.

Note that the mechanism of a returned string with a distinguishing first
character is part of the notation used in this standard to explain the
functions; implementations may use a different mechanism. Notation for
completion response and conditions

The testing of `X' and `S' indicators is made implicit, for brevity.
Even when written as a subroutine call, each use of a configuration
routine implies the testing. Thus:

call Config Time

implies

\#Response = Config Time()

if left (\#Response,1) == `X' then call \#Raise `SYNTAX', 5.1, substr
(\#Response, 2) if left (\#Response,1) == `S' then call \#Raise
`SYNTAX', 48.1, substr(\#Response, 2)

Source programs and character sets

The characters required by Rexx are identified by name, with a glyph
associated so that they can be printed in this standard. Alternative
names are shown as a convenience for the reader.

Notation

Note that nnn is not specifying the syntax of a program; it is
specifying the notation used in this standard for describing syntax.

Lexical level

Productions nnn and nnn contain a recursion of comment. Apart from this
recursion, the lexical level is a finite state automaton.

Syntax level

This syntax shows a null\_clause list, which is minimally a semicolon,
being required in places where programmers do not normally write
semicolons, for example after `THEN'. This is because the `THEN' implies
a semicolon. This approach to the syntax was taken to allow the rule
`semicolons separate clauses' to define `clauses'.

The precedence rules for the operators are built into this grammar

Data Model

The following explanation of data in terms of Classic Rexx may be
helpful. References to clauses of the existing standard have 274 as a
prefix.

We start with the data model from the first Standard - a number of
variable pools. Two mechanisms, the external access of section 274.5.13
(API\_Drop etc) and the internal of 274.7.1 (Var\_Drop etc). Pools are
numbered, with pool 0 reserved for reserved names (.MN etc) and pool N-1
being related to pool N as the caller's pool. The symbols which index
the pools are distinquished as tailed or non-tailed. The items in the
pool have attributes `exposed'', `dropped', and `implicit'. The values
in the pools are string values.

An extra scope is used for `state variables' used in the definition of
the standard. These follow the same lookup rules in a conceptual and
separate pool.

The first change necessary is to define the values in the pools as
references. For string values this is just a change in definition style,
since a reference always followed to a string value is semantically
identical with the notion of having the value in the pool. However,
references open the possibility of referencing

non-strings, which can behave in a changed way while still being refered
to by the same reference. (Mutable objects)

It is reasonable that the definition should have the pools reference one
another rather than use numbered pools. It is difficult to have a notion
of numbering the pools when any object can have a set of variables
associated with it.

Assignment is defined as assignment of references. The language could
have been designed differently, for example to make assignment behave
like the COPY method, but assignment of references is the natural,
powerful, choice.

If pools are not numbered, the notation of the first standard, where
some state variables use the \#Level number as part of their names, will
not suffice. An appropriate solution is to say that each variable pool
can have state variables and user program variables in it. Placing the
state variables that are per-procedure-level in the variable pool for
their level avoids the need to specify \#Level in their tails. There are
pre-existing objects such as all possible values that can be written as
literals and the objects accessed by .SYSTEM etc. Further objects are
created by the NEW method.

Editorial note: It looks nice to unify: an object \emph{is} a variable
pool and a variable pool \emph{is} an object. There is some awkwardness
describing the classic API\_ function as applying to an object. There
don't seem to be difficulties in defining any object behaviour we want
in terms of state variables that refer from one object to another.

Evaluation (Definitions written as code) There is no single definitional
mechanism for describing semantics that is predominantly used in
standards describing programming languages, except for the use of prose.
The committee has chosen to define some parts of this standard using
algorithms written in Rexx. This has the advantages of being rigorous
and familiar to many of the intended readers of this standard. It has
the potential disadvantage of circularity - a definition based on an
assumption that the reader already understands what is being defined.
Circularity has been avoided by: - specifying the language
incrementally, so that the algorithms for more complex details are
specified in code that uses only more simple Rexx. For example, the
notion that an expression evaluates to a result can be understood by the
reader even without a complete specification of all operators and
built-in functions that might be used in the expression; - specifying
the valid syntax of Rexx programs without using Rexx coding. The method
used, Backus Normal Form, can adequately be introduced by prose.
Ultimately, some understanding of programming languages is assumed in
the reader (just as the ability to read prose is assumed) but any
remaining circularity in this standard is harmless. The comparison of
two single characters is an example of such a circularity;
Config\_Compare can compare two characters but the outcome can only be
tested by comparing characters. It has to be assumed that the reader
understands such a comparison. Some of the definition using code is
repeating earlier definition in prose. This duplication is to make the
document easier to understand when read from front to back. Note that
the layout of the code, in the choices of instructions-per-line,
indentations etc., is not significant. (The layout style used follows
the examples in the base reference and it is deliberate that the DO and
END of a group are not at the same alignment.) The code is not intended
as an example of good programming practice or style. The variables in
this code cannot be directly referenced by any program, even if the
spelling of some VAR\_SYMBOL coincides. These variables, referred to as
state variables, are referenced throughout this document; they are not
affected by any execution activity involving scopes. Some of more
significant variables and routines are written with \# as their first
character. The following list of them is intended as an aid to
understanding the code. The index of this standard shows the main usage,
but not all usage, of these names. The following are constants set by
the configuration, by Config\_Constants: \#Configuration is used for
PARSE SOURCE. \#Version is used for PARSE VERSION. \#Bif\_Digits.
represents numeric digits settings, tails are built-in function names.
\#Limit\_Digits is the maximum significant digits.
\#Limit\_EnvironmentName is a maximum length.

\#Limit\_ExponentDigits is the maximum digits in an exponent.
\#Limit\_Literal is a maximum length.

\#Limit\_Messagelnsert is a maximum length.

\#Limit\_Name is a maximum length.

\#Limit\_String is a maximum length.

\#Limit\_TraceData is a maximum length.

These are named outputs of configuration routines:

\#Response is used to hold the result from a configuration routine.
\#Indicator is used to hold the leftmost character of Response.
\#Outcome is the main outcome of a configuration routine.

\#RC is set by Contig\_Command.

\#NoSource is set by Config\_NoSource.

\#Time is set by Config\_Time

\#Adjust\textless Index ``\#Adjust'' \#``\,'' \textgreater{} is set by
Config\_Time

These variables are set up with output from configuration routines:

\#Howlnvoked records from API\_Start, for use by PARSE SOURCE.

\#Source records from API\_ Start for use by PARSE SOURCE.

\#AIIBlanks\textless Index ``\#AllBlanks'' \# ``\,'' \textgreater{} is a
string including Blank and equivalents. \#ErrorText.MsgNumber is the
text as altered by limits.

\#SourceLine. is a record of the source, retained unless NoSource is
set. \#SourceLine.0 is a count of lines.

\#Pool is a reference to the current variable pool.

These are variables not initialized from the configuration:

\#Level is a count of invocation depth, starting at one.

\#NewLevel equals \#Level plus one.

\#Pool1 is a reference to the variable pool current when the first
instruction was executed.

\#Upper is a reference to the variable pool which will be current when
the current PROCEDURE ends. \#Loop is a count of loop nesting.

\#LineNumber is the line number of the current clause.

\#Symbol is a symbol after tails replacement.

\#API\_Enabled determines when the application programming interface for
variable pools is available. \#Test is the Greater/Lesser/Equal result.

\#InhibitPauses is a numeric trace control.

\#InhibitTrace is a numeric trace control.

\#AtPause is on when executing interactive input.

\#AllowProcedure provides a check for the label needed before a
procedure.

\#DatatypeResult is a by-product of DATATYPE().

\#Condition is a condition, eg `SYNTAX'.

\#Trace\_QueryPrior detects an external request for tracing.

\#Tracelnstruction detects TRACE as interactive input.

These are variables that are per-Level, that is, have \#Level as a tail
component:

\#lsFunction. indicates a function call.

\#lsProcedure. indicates indicates the routine is a procedure.

\#Condition. indicates whether the routine is handling a condition.

\#ArgExists.\#Level.ArgNumber indicates whether an argument exists.
(Initialized from API\_Start for Level=1)

\#Arg.\#Level.ArgNumber provides the value of an argument. (Initialized
from API\_Start for Level=1) When ArgNumber=0 this gives a count of the
arguments.

\#Tracing. is the trace setting letter.

\#Interactive. indicates when tracing is interactive. (`?' trace
setting) \#ClauseLocal. ensures that DATE/TIME are consistent across a
clause. \#ClauseTime. is the TIME/DATE frozen for the clause.

\#StartTime. is for `Elapsed' time calculations.

\#Digits. is the current numeric digits.

\#Form. is the current numeric form.

\#Fuzz. is the current numeric fuzz.

These are qualified by \#Condition as well as \#Level:

\#Enabling. is `ON', `OFF' or 'DELAYED''.

\#Instruction. is `CALL' or `SIGNAL'

\#TrapName. is the label.

\#ConditionDescription. is for CONDITION(`D')

\#ConditionExtra. is for CONDITION(`E')

\#ConditionInstruction. is for CONDITION(`T')

\#PendingNow. indicates a DELAYED condition. \#PendingDescription. is
the description of a DELAYED condition. \#PendingExtra. is the extra
description fora DELAYED condition. \#EventLevel. is the \#Level at
which an event was DELAYED.

These are qualified by ACTIVE, ALTERNATE, or TRANSIENT as well as
\#Level:

\#Env\_Name. is the environment name.

\#Env\_Type. is the type of a resource, and is additionally qualified by
input/output/error distinction. \#Env\_Resource. is the name of a
resource, and is additionally qualified by input/output/error
distinction. \#Env\_Position. is INPUT or APPEND or REPLACE, and is
additionally qualified by input/output/error distinction.

These are variables that are per-loop:

\#ldentity. is the control variable.

\#Repeat. is the repetition count.

\#By. is the increment.

\#To. is the limit.

\#For. is that count.

\#lterate. holds a position in code describing DO instruction semantics.
\#Once. holds a position in code describing DO instruction semantics.

\#Leave. holds a position in code describing DO instruction semantics.

These are variables that are per-stream:

\#Charin\_Position.

\#Charout\_Position.

\#Linein\_Position.

\#Lineout\_Position.

\#StreamState. records ERROR state for return by STREAM built-in
function.

These are commonly used prefixes: Config\_ is used for a function
provided by the configuration. API\_is used for an application
programming interface.

Trap\_ is used for a routine called from the processor, not provided by
it. Var\_ is used for the routines operating on the variable pools.

These are notation routines, only available to code in this standard:

\#Contains checks whether some construct is in the source. \#Instance
returns the content of some construct in the source. \#Evaluate returns
the value of some construct in the source. \#Execute causes execution of
some construct in the source. \#Parses checks whether a string matches
some construct. \#Clause notes some position in the code.

\#Goto continues execution at some noted position.

\#Retry causes execution to continue at a previous clause.

These are frequently used routines: \#Raise is a routine for condition
raising. \#Trace is a routine for trace output.

\#TraceSource is a routine to trace the source program. \#CheckArgs
processes the arguments to a built-in function.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/built-in_classes.tex
complete, but very deficient, extended standard

\hypertarget{built-in-classes}{%
\chapter{Built-in classes}\label{built-in-classes}}

\hypertarget{notation}{%
\section{Notation}\label{notation}}

The built-in classes are defined mainly through code. The code refers to
state variables. This is solely a notation used in this standard.

\hypertarget{object-class-and-method}{%
\section{Object, class and method}\label{object-class-and-method}}

These objects provide the basis for class structure.

\hypertarget{the-object-class}{%
\subsection{The object class}\label{the-object-class}}

::class object

::method new class Returns a new instance of the receiver class.

call Config ObjectNew return \#Outcome

::method `==' `==' with no argument gives a hash value in OOI.

call Config ObjectCompare \#Receiver, \#Arg.1

if \#Outcome == `equal' then return `1'

else return `0'

::method `\textless\textgreater!'

use arg a

return \self==a

::method `\textgreater\textless{}' forward message
`\textless\textgreater{}'

::method `=' forward message `=='

::method `=! forward message'\textless\textgreater'

::method `==' forward message `\textless\textgreater{}'

::method copy Returns a copy of the receiver object. The copied object
has the same methods as the receiver object and an equivalent set of
object variables, with the same values.

call Config ObjectCopy \#Receiver

return \#Outcome

Since we have var\_empty we could save a primitive by rendering `new' as
`copy' plus `empty'. ::method defaultname

Returns a short human-readable string representation for the object.

call var\_value \#Receiver, `\#Human', `0'

return \#Outcome

This field would have been filled in at `NEW' time.

::method `OBJECTNAMES' /* rvSTRING */

Sets the receiver object's name to the specified string.

call var\_set \#Receiver, \#ObjectName, `0', \#Arg.1 return

Initialized to \#Human? Or ObjectName does forwarding until assigned to?

::method objectname

Returns the receiver object's name (which the OBJECTNAME= method sets).
call var\_value \#Receiver, \#ObjectName, `0' return \#Outcome

::method string

Returns a human-readable string representation for the object. return
\#Receiver\textasciitilde ObjectName

::method class Returns the class object that received the message that
created the object.

call var value \#Receiver, \#IsA, `0' return \#Outcome

::method setmethod /* rSTRING oSTRING/METHOD/ARRAY */ Adds a method to
the receiver object's collection of object methods.

Is `object methods' what is intended; you add to a class without adding
to its instance methods? Yes. if \#ArgExists.2 then m = Arg.2

else m = .NIL call set\_var \#Receiver, `METHODS.'\#Arg.1, `1', m return
::method hasmethod /* rSTRING */

Returns 1 (true) if the receiver object has a method with the specified
name (translated to uppercase); otherwise, returns 0 (false).

This presumably means inherited as well as SETMETHOD ones. What about
ones set to .NIL?

Need to use the same search as for sending.

::method unsetmethod private

Removes a method from the receiver object's collection of object
methods. Use var\_drop

Private means Receiver = Self check.

::method request /* rSTRING */ Returns an object of the specified class,
or the NIL object if the request cannot be satisfied. t = `MAKE'\#Arg.1

if \#Receiver\textasciitilde hasmethod(t) then return .NIL forward
message(t) array ()

::method run private /* rMETHOD Ugh keyoptions */ Runs the specified
method. The method has access to the object variables of the receiver
object, just as if the receiver object had defined the method by using
SETMETHOD.

::smethod startat Undocumented?

::method start /* rMESSAGE oArglist */

Returns a message object and sends it a START message to start
concurrent processing. ::method init

Performs any required object initialization.

\hypertarget{the-class-class}{%
\subsection{The class class}\label{the-class-class}}

::class class

Lots of these methods are both class and instance. \textbar{} don't know
whether to list them twice. ::method new class /* OARGLIST */

Returns a new instance of the receiver class, whose object methods are
the instance methods of the class. This method initializes a new
instance by running its INIT methods.

::method subclass class

Returns a new subclass of the receiver class.

::method subclasses class

Returns the immediate subclasses of the receiver class in the form of a
single-index array of the required size.

::method define class /* rSTRING oMETHOD */

Incorporates the method object in the receiver class's collection of
instance methods. The method name is translated to upper case.

::method delete

Removes the receiver class's definition for the method name specified.

Builtin classes cannot be altered.

::method method class /* rSTRING */

Returns the method object for the receiver class's definition for the
method name given.

Do we have to keep saying ``method object'' as opposed to ``method''
because ``method name'' exists? ::method querymixinclass

Returns 1 (true) if the class is a mixin class or 0 (false) otherwise.

::method mixinclass class /* 3 of em */

Returns a new mixin subclass of the receiver class.

::method inherit class /* rCLASS oCLASS */

Causes the receiver class to inherit the instance and class methods of
the class object specified. The

optional class is a class object that specifies the position of the new
superclass in the list of superclasses. ::method uninherit class /*
rCLASSOBJ */

Nullifies the effect of any previous INHERIT message sent to the
receiver for the class specified. ::smethod enhanced class /*
rCOLLECTION oArgs */

Returns an enhanced new instance of the receiver class, with object
methods that are the instance

methods of the class enhanced by the methods in the specified collection
of methods. ::method baseclass class

Returns the base class associated with the class. If the class is a
mixin class, the base class is the first superclass that is not also a
mixin class. If the class is not a mixin class, then the base class is
the class receiving the BASECLASS message.

::method superclasses class Returns the immediate superclasses of the
receiver class in the form of a single-index array of the

required size. ::method id class

Returns a string that is the class identity (instance SUBCLASS and
MIXINCLASS methods.)

::method metaclass class

Returns the receiver class's default metaclass. ::method methods class
/* oCLASSOBJECT */

Returns a supplier object for all the instance methods of the receiver
class and its superclasses, if no argument is specified. \#\#\# The
method class

::class method

::method new class /* rSTRING rSOURCE */ Returns a new instance of
method class, which is an executable representation of the code
contained in

the source. ::method setprivate

Specifies that a method is a private method. ::method setprotected

::method setsecuritymanager

::method setguarded Reverses any previous SETUNGUARDED messages,
restoring the receiver to the default guarded

status. ::method setunguarded

Lets an object run a method even when another method is active on the
same object. If a method object

does not receive a SETUNGUARDED message, it requires exclusive use of
its object variable pool. ::method source

Returns the method source code as a single index array of source lines.
::method interface

::method setinterface

11.3 The string class The string class provides conventional strings and
numbers.

Some differences from REXX class of NetRexx. ::class string

::method new class

::method `' We can do all the operators by appeal to classic section 7.
::method'-' ::method `-'

use arg a return \a General problem of making the error message come
right.

:method :method :method :method :method :method :method :method :method
:method :method :method :method :method :method :method :method :method
:method :method :method :method :method :method :method :method :method
:method :method

:method :method :method :method :method :method :method :method :method
:method :method :method :method :method :method :method :method :method
:method :method :method :method :method :method :method :method :method
:method :method :method :method :method :method :method :method :method
:method :method :method :method :method

All

centre center changestr compare copies counstr datatype delstr delword
insert lastpos left length overlay pos reverse right space strip substr
subword translate verify word wordindex wordlength wordpos words abs
format max

min

sign trunc B2x bitand bitor bitxor C2D

C2x

D2xX ::smethod D2C ::smethod X2B ::smethod X2C ::smethod X2D ::method
string

::method makestring \#\#\# The array class

The main features of a single dimension array are provided by the
configuration. This section defines further methods and
multi-dimensional arrays.

To be done. Dimensionality set at first use. Count commas, not classic
arg().

::class array

::method new class /* 0 or more WHOLE\textgreater=0 */

Returns a new empty array.

::method of class /* 0 or more ANY */

Returns a newly created single-index array containing the specified
value objects. ::method put /* rANY one or more WHOLE\textgreater0 */

Makes the object value a member item of the array and associates it with
the specified index or indexes. ::method ' {[}{]}=' /* 1 or more
WHOLE\textgreater0O */

This method is the same as the PUT method.

::method at /* 1 or more WHOLE\textgreater0O */

Returns the item associated with the specified index or indexes.

::method `{[}{]}' /* 1 or more WHOLE\textgreater0O */

Returns the same value as the AT method.

::method remove /* 1 or more WHOLE\textgreater0O */

Returns and removes the member item with the specified index or indexes
from the array. ::method hasindex /* 1 or more WHOLE\textgreater0O */

Returns 1 (true) if the array contains an item associated with the
specified index or indexes. Returns 0 (false) otherwise.

::method items /* (None) \emph{/ Returns the number of items in the
collection. ::method dimension /} OWHOLE\textgreater0O */

Returns the current size (upper bound) of dimension specified (a
positive whole number). If you omit the argument this method returns the
dimensionality (number of dimensions) of the array.

::method size /* (None) */

Returns the number of items that can be placed in the array before it
needs to be extended.

::method first /* (None) */

Returns the index of the first item in the array, or the NIL object if
the array is empty.

::method last /* (None) */

Returns the index of the last item in the array, or the NIL object if
the array is empty.

::method next /* rcWHOLE\textgreater O */

Returns the index of the item that follows the array item having the
specified index or returns the NIL object if the item having that index
is last in the array.

::method previous /* rcWHOLE\textgreater O */

Returns the index of the item that precedes the array item having index
index or the NIL object if the item having that index is first in the
array.

::method makearray /* (None) */

Returns a single-index array with the same number of items as the
receiver object. Any index with no associated item is omitted from the
new array.

Returns a new array (of the same class as the receiver) containing
selected items from the receiver array. The first item in the new array
is the item corresponding to index start (the first argument) in the
receiver

array. ::method supplier /* (None) \emph{/ Returns a supplier object for
the collection. ::method section /} rcWHOLE\textgreater0O
oOWHOLE\textgreater=0 */

\hypertarget{the-supplier-class}{%
\section{The supplier class}\label{the-supplier-class}}

A supplier object enumerates the items a collection contained at the
time of the supplier's creation. ::class supplier

::method new class /* rANYARRAY rINDEXARRAY */ Returns a new supplier
object.

::method index

Returns the index of the current item in the collection. ::method next

Moves to the next item in the collection.

::method item

Returns the current item in the collection.

::method available

Returns 1 (true) if an item is available from the supplier (that is, if
the ITEM method would return a value). Returns 0 (false) otherwise.

11.5 The message class ::class message

::method init class /* Ugh */

Initializes the message object for sending\ldots\ldots{}

::method completed

Returns 1 if the message object has completed its message; returns 0
otherwise. ::method notify /* rMESSAGE */

Requests notification about the completion of processing for the message
SEND or START sends. ::method start /* oANY */

Sends the message for processing concurrently with continued processing
of the sender. ::method send /* oANY */

Returns the result (if any) of sending the message.

::method result

Returns the result of the message SEND or START sends.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/built-in_functions.tex
complete, but very deficient, extended standard

\hypertarget{built-in-functions}{%
\chapter{Built-in functions}\label{built-in-functions}}

\hypertarget{notation}{%
\section{Notation}\label{notation}}

The built-in functions are defined mainly through code. The code refers
to state variables. This is solely a notation used in this standard.

The code refers to functions with names that start with `Config\_';
these are the functions described in section nnn.

The code is specified as an external routine that produces a result from
the values \#Bif (which is the name of the built-in function),
\#Bif\_Arg.0 (the number of arguments), \#Bif\_Arg.i and
\#Bif\_ArgExists.i (which are the argument data.)

The value of \#Level is the value for the clause which invoked the
built-in function.

The code either returns the result of the built-in or exits with an
indication of a condition that the invocation of the built-in raises.

The code below uses built-in functions. Such a use invokes another use
of this code with a new value of \#Level. On these invocations, the
CheckArgs function is not relevant.

Numeric settings as follows are used in the code. When an argument is
being checked as a number by `NUM' or `WHOLENUM' the settings are those
current in the caller. When an argument is being checked as an integer
by an item containing `WHOLE' the settings are those for the particular
built-in function. Elsewhere the settings have sufficient numeric digits
to avoid values which would require exponential notation.

\hypertarget{routines-used-by-built-in-functions}{%
\section{Routines used by built-in
functions}\label{routines-used-by-built-in-functions}}

The routine CheckArgs is concerned with checking the arguments to the
built-in. The routines Time2Date and Leap are for date calculations.
ReRadix is used for radix conversion. The routine Raise raises a
condition and does not return.

\hypertarget{argument-checking}{%
\subsection{Argument checking}\label{argument-checking}}

/* Check arguments. Some further checks will be made in particular
built-ins.\emph{/ /} The argument to CheckArgs is a checklist for the
allowable arguments. */

/* NUM, WHOLENUM and WHOLE have a side-effect, `normalizing' the number.
*/

/* Calls to raise syntax conditions will not return. */

CheckArgs: CheckList = arg(1) /* This refers to the argument of
CheckArgs. */

/* Move the checklist information from a string to individual variables
\emph{/ ArgType. = '\,'\,' ArgPos = 0 /} To count arguments \emph{/
MinArgs = 0 do j = 1 to length (CheckList) ArgPos = ArgPos+1 /} Count
the required arguments. \emph{/ if substr(CheckList,j,1) == `r' then
MinArgs = MinArgs + 1 /} Collect type information. \emph{/ do while j
\textless{} length(CheckList) j=eajae1 t = substr(CheckList,j,1) if t=='
' then leave ArgType.ArgPos = ArgType.ArgPos \textbar\textbar{} t end /}
A single space delimits parts. */ end j MaxArgs = ArgPos

/* Check the number of arguments to the built-in, in this instance. */
NumArgs = \#Bif Arg.0

if NumArgs \textless{} MinArgs then call Raise 40.3, MinArgs

if NumArgs \textgreater{} MaxArgs then call Raise 40.4, MaxArgs

/* Check the type(s) of the arguments to the built-in. */ do ArgPos = 1
to NumArgs if \#Bif ArgExists.ArgPos then call CheckType else if ArgPos
\textless= MinArgs then call Raise 40.5, ArgPos end ArgPos

/* No errors found by CheckArgs. */

return CheckType: Value = \#Bif Arg.ArgPos Type = ArgType.ArgPos select
when Type == `ANY' then nop /* Any string \emph{/ when Type == `NUM'
then do /} Any number */

/* This check is made with the caller's digits setting. \emph{/ if
\Cdatatype(Value, `N') then if \#DatatypeResult==`E' then call Raise
40.9, ArgPos, Value else call Raise 40.11, ArgPos, Value \#Bif
Arg.ArgPos=\#DatatypeResult /} Update argument copy. */ end

when Type == `WHOLE' then do /* Whole number \emph{/ /} This check is
made with digits setting for the built-in. */ if \Edatatype(Value,`W')
then call Raise 40.12, ArgPos, Value \#Bif\_ Arg.ArgPos=\#DatatypeResult
end

when Type == `WHOLE\textgreater=0' then do /* Non-negative whole number
*/ if \Edatatype(Value,`W') then call Raise 40.12, ArgPos, Value if
\#DatatypeResult \textless{} 0 then call Raise 40.13, ArgPos, Value
\#Bif\_ Arg.ArgPos=\#DatatypeResult end

when Type == `WHOLE\textgreater0O' then do /* Positive whole number */
if \Edatatype(Value,`W') then call Raise 40.12, ArgPos, Value if
\#DatatypeResult \textless= 0 then call Raise 40.14, ArgPos, Value
\#Bif\_ Arg.ArgPos=\#DatatypeResult end

when Type == `WHOLENUM' then do /* D2X type whole number \emph{/ /} This
check is made with digits setting of the caller. */ if
\Cdatatype(Value,`W') then call Raise 40.12, ArgPos, Value \#Bif\_
Arg.ArgPos=\#DatatypeResult end

when Type == `WHOLENUM\textgreater=0' then do /* D2X Non-negative whole
number */ if \Cdatatype(Value,`W') then call Raise 40.12, ArgPos, Value
if \#DatatypeResult \textless{} 0 then call Raise 40.13, ArgPos, Value
\#Bif\_ Arg.ArgPos=\#DatatypeResult end

when Type == `0 90' then do /* Errortext */

if \Edatatype(Value,`N') then call Raise 40.11, ArgPos, Value

Value=\#DatatypeResult

\#Bif\_ Arg.ArgPos=Value

Major=Value \% 1

Minor=Value - Major

if Major \textless{} 0 \textbar{} Major \textgreater{} 90 \textbar{}
Minor \textgreater{} .9 \textbar{} pos(`E',Value)\textgreater0 then call
Raise 40.17, Value /* ArgPos will be 1 */

end

when Type == `PAD' then do /* Single character, usually a pad. */

if length(Value) = 1 then call Raise 40.23, ArgPos, Value end

when Type == `HEX' then /* Hexadecimal string \emph{/ if
\datatype (Value, `X') then call Raise 40.25, Value /} ArgPos will be 1
\emph{/ when Type == `BIN' then /} Binary string \emph{/ if
\datatype(Value,`B') then call Raise 40.24, Value /} ArgPos will be 1
\emph{/ when Type == `SYM' then /} Symbol */

if \datatype (Value, `S') then call Raise 40.26, Value /* ArgPos will be
1 */

when Type == `STREAM' then do call Config Stream Qualify Value if left
(\#Response, 1) == `B' then call Raise 40.27, Value /* ArgPos will be 1
*/ end

when Type = `ACEFILNOR' then do /* Trace \emph{/ Val = Value /} Allow
`?' alone */

if val == `?' then do /* Allow leading `?' */ if left(Val,1) == `?' then
Val = substr(Val,2)

if pos(translate(left(Val, 1)), `ACEFILNOR') = 0 then call Raise 40.28,
ArgPos, Type, Val

end end otherwise do /* Options \emph{/ /} The checklist item is a list
of allowed characters */ if Value == '\,' then

call Raise 40.21, ArgPos \#Bif Arg.ArgPos = translate(left(Value, 1)) if
pos(\#Bif Arg.ArgPos, Type) = 0 then call Raise 40.28, ArgPos, Type,
Value end

end /* Select */ return

Cdatatype: /* This check is made with the digits setting of the caller.
\emph{/ /} \#DatatypeResult will be set by use of datatype() */

numeric digits \#Digits.\#Level

numeric form value \#Form.\#Level

return datatype(arg(1), arg(2))

Edatatype: /* This check is made with digits setting for the particular
built-in. \emph{/ /} \#DatatypeResult will be set by use of datatype()
*/

numeric digits \#Bif Digits.\#Bif

numeric form scientific

return datatype(arg(1),arg(2))

10.2.2 Date calculations Time2Date: if arg(1) \textless{} 0 then call
Raise 40.18 if arg(1) \textgreater= 315537897600000000 then call Raise
40.18 return Time2Date2 (arg(1))

Time: procedure /* This routine is essentially the code from the
standard, put in stand-alone form. The only `tricky bit' is that there
is no Rexx way for it to fail with the same error codes as a ``real''
implementation would. It can however give a SYNTAX error, albeit not the
desirable one. This causing of an error is done by returning with no
value. Since the routine will have been called as a function, this
produces an error. */

/* Backslash is avoided as some systems don't handle that negation sign.
\emph{/ if argQ)\textgreater3 then returm numeric digits 18 if
arg(1,`E') then if pos(translate(left(arg(1),1)),``CEHLMNRS'')=0 then
return /} (The standard would also allow `O' but what this code is
running on would not.) \emph{/ if arg(3,`E') then if
pos(translate(left(arg(3),1)),``CHLMNS'')=0 then return /} Tf the third
argument is given then the second is mandatory. \emph{/ if arg(3,`E') \&
arg(2,`E')=0 then returm /} Default the first argument. \emph{/ if
arg(1,`E') then Option = translate(left(arg(1),1)) else Option = `N'' /}
Tf there is no second argument, the current time is returned. \emph{/ if
arg(2,`E') = 0 then if arg(1,`E') then return ``TIME'(arg(1)) else
return `TIME'() /} One cannot convert to elapsed times. */ if
pos(Option, `ERO') \textgreater{} 0 then returm InValue = arg(2) if
arg(3,`E') then InOption = arg(3) else InOption = `N' HH =0 MM = 0 SS =0
HourAdjust = 0 select

when InOption == `C' then do parse var InValue HH `:'. +1 MM +2 XX if HH
= 12 then HH =0 if XX == `pm' then HourAdjust = 12

end when InOption == `H' then HH = InValue

when InOption == `L'' \textbar{} InOption == `N' then parse var InValue
HH `:' MM `:' SS when InOption == `M' then MM = InValue otherwise SS =
InValue end if datatype(HH,`W')=0 \textbar{} datatype(MM,`W')=0
\textbar{} datatype(SS,`N')=0 then returm HH = HH + HourAdjust /*
Convert to microseconds \emph{/ Micro = trunc((((HH } 60) + MM) * 60 +
SS) * 1000000) /* There is no special message for time-out-of-range; the
bad-format message is used. \emph{/ if Micro\textless0O \textbar{} Micro
\textgreater{} 24}3600* 1000000 then returm /* Reconvert to further
check the original. */ if TimeFormat(Micro,InOption) == InValue then
return TimeFormat(Micro, Option) return

TimeFormat: procedure /* Convert from microseconds to given format.
\emph{/ /} The day will be irrelevant; actually it will be the first day
possible. */ x = Time2Date2(arg(1)) parse value x with Year Month Day
Hour Minute Second Microsecond Base Days select when arg(2) == `C' then
select when Hour\textgreater12 then return
Hour-12':`right(Minute,2,'0'')'pm' when Hour=12 then return
`12:'right(Minute,2,'0')`pm' when Hour\textgreater0 then return
Hour':`right(Minute,2,'0')`am' when Hour=0 then return
`12:'right(Minute,2,'0')`am'

when arg(2) == `H' then return Hour when arg(2) == `L' then return
right(Hour,?2,`0')`:'right(Minute,2,'0')`:'right(Second,2,'0'),
\textbar\textbar{} `.'right(Microsecond,6,'0') when arg(2) == `M' then
return 60\emph{Hour+Minute when arg(2) == `N' then return
right(Hour,?2,`0')`:'right(Minute,2,'0')`:'right(Second,2,'0') otherwise
/} arg(2) ==`S' \emph{/ return 3600}Hour+60* Minute+Second end

Time2Date2: Procedure /* Convert a timestamp to a date. Argument is a
timestamp (the number of microseconds relative to 0001 01 01
00:00:00.000000) Returns a date in the form: year month day hour minute
second microsecond base days */

/* Argument is relative to the virtual date 0001 01 01 00:00:00.000000
*/ Time = arg(1)

Second = Time \% 1000000 \} Microsecond = Time // 1000000 Minute =
Second \% 60 ; Second = Second // 60 Hour = Minute \% 60 ; Minute =
Minute // 60 Day = Hour \% 24 ;\} Hour = Hour // 24

/* At this point, the days are the days since the 0001 base date. */
BaseDays = Day Day = Day + 1

/* Compute either the fitting year, or some year not too far earlier.
Compute the number of days left on the first of January of this year. */

Year = Day \% 366 Day = Day - (Year*365 + Year\%4 - Year\%100 +
Year\%400) Year = Year +1

/* Now if the number of days left is larger than the number of days in
the year we computed, increment the year, and decrement the number of
days accordingly. */ do while Day \textgreater{} (365 + Leap (Year) )

Day = Day - (365 + Leap(Year) )

Year = Year + 1 end

/* At this point, the days left pertain to this year. */ YearDays = Day

/* Now step through the months, increment the number of the month, and
decrement the number of days accordingly (taking into consideration that
in a leap year February has 29 days), until further reducing the number
of days and incrementing the month would lead to a negative number of
days */ Days = `31 28 31 30 31 30 31 31 30 31 30 31' do Month = 1 to
words (Days)

ThisMonth = Word(Days, Month) + (Month = 2) * Leap (Year)

if Day \textless= ThisMonth then leave

Day = Day - ThisMonth end

return Year Month Day Hour Minute Second Microsecond BaseDays YearDays

Leap: procedure /* Return 1 if the year given as argument is a leap
year, or 0 otherwise. */ return (arg(1)//4 = 0) \& ((arg(1)//100
\textless\textgreater{} 0) \textbar{} (arg(1)//400 = 0))

10.2.1. Radix conversion

ReRadix: /* Converts Arg(1) from radix Arg(2) to radix Arg(3) */

procedure

Subj ect=arg(1)

FromRadix=arg (2)

ToRadix=arg (3)

/* Radix range is 2-16. Conversion is via decimal */

Integer=0

do j=1 to length (Subject) /* Individual digits have already been
checked for range. \emph{/ Integer=Integer}FromRadix+pos (substr
(Subject,j,1),`0123456789ABCDEF')-1 end

rete

do while Integer\textgreater0

r= substr(`0123456789ABCDEF',1 + Integer // ToRadix, 1)
\textbar\textbar{} r Integer = Integer \% ToRadix end

/* When between 2 and 16, there is no zero suppression. \emph{/ if
FromRadix = 2 \& ToRadix = 16 then r=eright(r, (length(Subject)+3) \% 4,
`0') else if FromRadix = 16 \& ToRadix = 2 then reright(r,
length(Subject) } 4, `0') return r

\hypertarget{raising-the-syntax-condition}{%
\subsection{Raising the SYNTAX
condition}\label{raising-the-syntax-condition}}

Raise:

/* These 40.nn messages always include the built-in name as an
insert.\emph{/ call \#Raise `SYNTAX', arg(1), \#Bif, arg(2), arg(3),
arg(4) /} \#Raise does not return. */

\hypertarget{character-built-in-functions}{%
\section{Character built-in
functions}\label{character-built-in-functions}}

These functions process characters or words in strings. Character
positions are numbered from one at the left. Words are delimited by
blanks and their equivalents, word positions are counted from one at the
left.

\hypertarget{abbrev}{%
\subsection{ABBREV}\label{abbrev}}

ABBREV returns `1' if the second argument is equal to the leading
characters of the first and the length of the second argument is not
less than the third argument.

call CheckArgs `rANY rANY oWHOLE\textgreater=0'

Subject \#Bif Arg.1

Subj \#Bif Arg.2

if \#Bif\_ArgExists.3 then Length = \#Bif Arg.3 else Length = length
(Subj)

Condl = length(Subject) \textgreater= length(Subj) Cond2 = length(Subj)
\textgreater= Length Cond3 = substr(Subject, 1, length(Subj)) == Subj

return Condl \& Cond2 \& Cond3

\hypertarget{center}{%
\subsection{CENTER}\label{center}}

CENTER returns a string with the first argument centered in it. The
length of the result is the second argument and the third argument
specifies the character to be used for padding.

call CheckArgs `rANY rWHOLE\textgreater=0 oPAD'

String = \#Bif Arg.1

Length = \#Bif Arg.2

if \#Bif\_ArgExists.3 then Pad = \#Bif Arg.3 else Pad = ' '!

Trim = length(String) - Length

if Trim \textgreater{} 0 then return substr(String, Trim \% 2 + 1,
Length)

return overlay(String, copies(Pad, Length), -Trim \% 2 + 1)

\hypertarget{centre}{%
\subsection{CENTRE}\label{centre}}

This is an alternative spelling for the CENTER built-in function.

\hypertarget{changestr}{%
\subsection{CHANGESTR}\label{changestr}}

CHANGESTR replaces all occurrences of the first argument within the
second argument, replacing them with the third argument.

call CheckArgs ```rANY rANY rANY'

Output = '\,' Position = 1 do forever FoundPos = pos(\#Bif Arg.1, \#Bif
Arg.2, Position) if FoundPos = 0 then leave Output = Output
\textbar\textbar{} substr(\#Bif\_ Arg.2, Position, FoundPos - Position),
\textbar\textbar{} \#Bif Arg.3 Position = FoundPos + length(\#Bif Arg.1)
end return Output \textbar\textbar{} substr(\#Bif Arg.2, Position)

\hypertarget{compare}{%
\subsection{COMPARE}\label{compare}}

COMPARE returns `0' if the first and second arguments have the same
value. Otherwise, the result is the position of the first character that
is not the same in both strings.

call CheckArgs `rANY rANY oPAD'

Strl = \#Bif\_Arg.1

Str2 = \#Bif\_Arg.2

if \#Bif\_ArgExists.3 then Pad else Pad

\#Bif Arg.3

/* Compare the strings from left to right one character at a time */ if
length(Str1) \textgreater{} length(Str2) then do

Length = length(Str1)

Str2=left (Str2, Length, Pad)

end else do

Length = length(Str2)

Strl=left (Str1, Length, Pad)

end

do i= 1 to Length if substr(Stri, i, 1) == substr(Str2, i, 1) then
return i end

return 0

\hypertarget{copies}{%
\subsection{COPIES}\label{copies}}

COPIES returns concatenated copies of the first argument. The second
argument is the number of copies.

call CheckArgs ```rANY rWHOLE\textgreater=0' Output = '\,' do \#Bif
Arg.2 Output = Output \textbar\textbar{} \#Bif\_Arg.1 end

return Output

\hypertarget{countstr}{%
\subsection{COUNTSTR}\label{countstr}}

COUNTSTR counts the appearances of the first argument in the second
argument.

call CheckArgs ```rANY rANY'

Output = 0

Position = pos (\#Bif Arg.1,\#Bif Arg.2)

do while Position \textgreater{} 0 Output = Output + 1 Position =
pos(\#Bif Arg.1, \#Bif Arg.2, Position + length(\#Bif Arg.1)) end

return Output

\hypertarget{datatype}{%
\subsection{DATATYPE}\label{datatype}}

DATATYPE tests for characteristics of the first argument. The second
argument specifies the particular test.

call CheckArgs `rANY oABLMNSUWX'

/* As well as returning the type, the value for a `NUM' is set in
\#DatatypeResult. This is a convenience when DATATYPE is used

by CHECKARGS. */

String = \#Bif Arg.1

/* If no second argument, DATATYPE checks whether the first is a number.
*/ if \#Bif\_ArgExists.2 then return DtypeOne()

Type = \#Bif\_ Arg.2 /* Null strings are a special case. */

if String == '\,' then do if Type == ``X'' then return 1 if Type ==
``B'' then return 1 return 0 end

/* Several of the options are shorthands for VERIFY \emph{/
azl=``abcdefghijklmnopqrstuvwxyz'' AZU= ``ABCDEFGHIJKLMNOPQRSTUVWXYZ''
DO09=``0123456789'' if Type == ``A'' then return
verify(String,az1\textbar\textbar AzU\textbar\textbar D09)=0 if Type ==
``B'' then do /} Check blanks in allowed places. \emph{/ if pos (left
(String,1),\#Al1Blanks)\textgreater0 then return 0 if pos (right
(String,1),\#Al1Blanks)\textgreater0 then return 0 BinaryDigits=0 do j =
length(String) by -1 tol ce = substr(String,j,1) if
pos(c,\#Al11Blanks)\textgreater0 then do /} Blanks need four
BinaryDigits to the right of them. */ if BinaryDigits//4 = 0 then return
0 end else do if verify(c,``01'') = 0 then return 0 BinaryDigits =
BinaryDigits + 1

end end j

return 1

end /* B */ if Type == ``L'' then return (verify (String,azl1)=0) if
Type == ``M'' then return (verify (String, azl1\textbar\textbar AZU)=0)
if Type == ``N'' then return (datatype (String) ==``NUM'') if Type ==
``S'' then return(symbol (String) ==`BAD') if Type == ``U'' then return
(verify (String, AZU)=0) if Type == ``W'' then do

/* It may not be a number. \emph{/ if DtypeOne(String) == `CHAR' then
return `0' /} It can be ``Whole'' even if originally in exponential
notation, provided it can be written as non-exponential. \emph{/ if
pos(`E',\#DatatypeResult)\textgreater0 then return `0' /} It won't be
``Whole'' if there is a non-zero after the decimal point. */
InFraction=`0'

do j = 1 to length (String) ce = substr(String,j,1) if
pos(c,`Ee')\textgreater0 then leave j if InFraction \&
pos(c,`+-')\textgreater0 then leave j if c == `.' then InFraction=`1'
else if InFraction \& a==`0' then return 0 end j /* All tests for Whole
passed. \emph{/ \#DatatypeResult = \#DatatypeResult \% 1 return 1 end /}
W \emph{/ /} Type will be ``x'' \emph{/ if pos (left
(String,1),\#Al1Blanks)\textgreater0 then return 0 if pos (right
(String,1),\#Al1Blanks)\textgreater0 then return 0 HexDigits=0 do
j=length(String) by -1 to 1 c=substr (String,j,1) if
pos(c,\#Al11Blanks)\textgreater0 then do /} Blanks need a pair of
HexDigits to the right of them. \emph{/ if HexDigits//2 = 0 then return
0 end else do if verify(c,``abcdefABCDEF''D09) = 0 then return 0
HexDigits=HexDigits+1 end end return 1 /} end X */

DtypeOne: /* See section nnn for the syntax of a number. \emph{/
\#DatatypeResult = `S' /} If not syntactically a number \emph{/ Residue
= strip(String) /} Blanks are allowed at both ends. */ if Residue ==
'\,' then return ``CHAR'' Sign = '\,' if left(Residue,1) == `+'
\textbar{} left(Residue,1) == `-' then do

Sign = left(Residue, 1) Residue = strip(substr(Residue,2),`L') /* Blanks
after sign */

end if Residue == '\,' then return ``CHAR'' /* Now testing Number,
section nnn */ if left(Residue,1) == `.' then do Residue = substr
(Residue, 2) Before = '\,' After = DigitRun() if After == '\,' then
return ``CHAR'' end else do Before = DigitRun() if Before == '\,' then
return ``CHAR'' if left(Residue,1) == `.' then do Residue = substr
(Residue, 2) After = DigitRun() end end Exponent = 0 if Residue == '\,'
then do if left(Residue, 1) == `e' \& left(Residue, 1) == `E' then

return ``CHAR'' Residue = substr (Residue, 2)

if Residue == '\,' then return ``CHAR''

Esign = '\,'

if left(Residue, 1) == `+' \textbar{} left(Residue, 1) == `-' then do
Esign = left(Residue, 1) Residue = substr (Residue, 2) if Residue ==
'\,' then return ``CHAR'' end

Exponent = DigitRun()

if Exponent == '\,' then return ``CHAR''

Exponent = Esign \textbar\textbar{} Exponent

end

if Residue == '\,' then return ``CHAR''

102 /\emph{DATATYPE tests for exponent out of range. }/

\#DatatypeResult = `E' /* If exponent out of range */ Before =
strip(Before,`L',`0') if Before == '\,' then Before = `0'

Exponent = Exponent + length(Before) -1 /* For SCIENTIFIC */

/* ``Engineering notation causes powers of ten to expressed as a
multiple of 3 - the integer part may therefore range from 1 through
9910.'' \emph{/ g=1 if \#Form.\#Level == `E' then do /} Adjustment to
make exponent a multiple of 3 */ g = Exponent//3 if g \textless{} 0 then
g = g + Exponent = Exponent - end

/* Check on the exponent. */ if Exponent \textgreater{} \#Limit
ExponentDigits then return ``CHAR'' if -\#Limit ExponentDigits
\textgreater{} Exponent then return ``CHAR''

/* Format to the numeric setting of the caller of DATATYPE */ numeric
digits \#Digits.\#Level

numeric form value \#Form.\#Level

\#DatatypeResult = 0 + \#Bif\_Arg.1

return ``NUM''

DigitRun: Outcome = '\,' do while Residue == '\,' if pos(left (Residue,
1), `0123456789') = 0 then leave

Outcome = Outcome \textbar\textbar{} left(Residue, 1) Residue =
substr(Residue, 2) end

return Outcome

\hypertarget{delstr}{%
\subsection{DELSTR}\label{delstr}}

DELSTR deletes the sub-string of the first argument which begins at the
position given by the second argument. The third argument is the length
of the deletion.

call CheckArgs `rANY rWHOLE\textgreater0 oWHOLE\textgreater=0' String
\#Bif Arg.1

Num \#Bif Arg.2 if \#Bif\_ArgExists.3 then Len = \#Bif\_Arg.3

if Num \textgreater{} length(String) then return String

Output = substr(String, 1, Num - 1) if \#Bif\_ArgExists.3 then if Num +
Len \textless= length(String) then Output = Output \textbar\textbar{}
substr(String, Num + Len) return Output

\hypertarget{delword}{%
\subsection{DELWORD}\label{delword}}

DELWORD deletes words from the first argument. The second argument
specifies position of the first word to be deleted and the third
argument specifies the number of words.

call CheckArgs `rANY rWHOLE\textgreater0 oWHOLE\textgreater=0' String
\#Bif Arg.1

Num \#Bif Arg.2 if \#Bif\_ArgExists.3 then Len = \#Bif\_Arg.3

if Num \textgreater{} words(String) then return String

EndLeft = wordindex(String, Num) - 1 Output = left(String, EndLeft) if
\#Bif\_ArgExists.3 then do BeginRight = wordindex(String, Num + Len)

if BeginRight\textgreater0 then Output =

end return Output

\hypertarget{insert}{%
\subsection{INSERT}\label{insert}}

Output \textbar\textbar{} substr(String, BeginRight)

INSERT insets the first argument into the second. The third argument
gives the position of the character before the insert and the fourth
gives the length of the insert. The fifth is the padding character.

call CheckArgs

New \#Bif\_ Arg.1 Target \#Bif Arg.2 if \#Bif\_ArgExists.3

then else then else then else

Num Num Length = \#Bif Arg.4 Length = length (New) Pad \#Bif Arg.5 Pad
ro

\#Bif Arg.3 0

if \#Bif\_ArgExists.4

if \#Bif\_ArgExists.5

return left(Target, Num, Pad) left (New, Length, Pad), substr(Target,
Num + 1)

``

\hypertarget{lastpos}{%
\subsection{LASTPOS}\label{lastpos}}

/* To left of insert /* New string inserted \emph{/ /} To right of
insert

`rANY rANY oWHOLE\textgreater=0 oWHOLE\textgreater=0 oPAD'

\emph{/ }/

LASTPOS returns the position of the last occurrence of the first
argument within the second. The third

argument is a starting position for the search.

call CheckArgs `rANY rANY oWHOLE\textgreater0'

Needle = \#Bif\_Arg.1 Haystack = \#Bif Arg.2 if \#Bif\_ArgExists.3 then
Start = \#Bif\_Arg.3 else Start = length(Haystack)

NeedleLength = length (Needle) if NeedleLength = 0 then return 0 Start =
Start - NeedleLength + 1 do i= Start by -1 while i \textgreater{} 0 if
substr(Haystack, i, NeedleLength) end i return 0

\hypertarget{left}{%
\subsection{LEFT}\label{left}}

LEFT returns characters that are on the left of the first argument.

length of the result and the third is the padding character. call
CheckArgs `rANY rWHOLE\textgreater=0 oPAD'

if \#Bif\_ArgExists.3 then Pad else Pad

\#Bif Arg.3

return substr(\#Bif Arg.1, 1, \#Bif\_ Arg.2, Pad)

10.1.14 LENGTH

Needle then return i

The second argument specifies the

Length returns a count of the number of characters in the argument.

call CheckArgs `rANY'

String = \#Bif Arg.1

\#Response = Config Length(String)

Length = \#Outcome

call Config Substr \#Response, 1

if \#Outcome == `E' then return Length

/* Here if argument was not a character string. call Config C2B String
call \#Raise `SYNTAX',

23.1, b2x(\#Outcome)

\emph{/ /} No return to here */

\hypertarget{overlay}{%
\subsection{OVERLAY}\label{overlay}}

OVERLAY overlays the first argument onto the second. The third argument
is the starting position of the

overlay. The fourth argument is the length of the overlay and the fifth
is the padding character.

call CheckArgs `rANY rANY oWHOLE\textgreater0 oOWHOLE\textgreater=0
oPAD'

New = \#Bif\_Arg.1

Target = \#Bif Arg.2

if \#Bif\_ArgExists.3 then Num = \#Bif\_Arg.3 else Num = 1

if \#Bif\_ArgExists.4 then Length = \#Bif\_ Arg.4 else Length = length
(New)

if \#Bif\_ArgExists.5 then Pad = \#Bif\_ Arg.5 else Pad = ' ' return
left(Target, Num - 1, Pad), /* To left of overlay \emph{/
\textbar\textbar{} left (New, Length, Pad), /} New string overlaid
\emph{/ \textbar\textbar{} substr (Target, Num + Length) /} To right of
overlay */

10.1.16 POS POS returns the position of the first argument within the
second.

call CheckArgs `rANY rANY oWHOLE\textgreater0'

Needle \#Bif Arg.1

Haystack = \#Bif Arg.2

if \#Bif\_ ArgExists.3 then Start else Start

\#Bif Arg.3 1

if length(Needle) = 0 then return 0

do i = Start to length (Haystack) +1-length (Needle) if substr(Haystack,
i, length(Needle)) == Needle then return i end i

return 0

\hypertarget{reverse}{%
\subsection{REVERSE}\label{reverse}}

REVERSE returns its argument, swapped end for end.

call CheckArgs `rANY'

String

\#Bif Arg.1 Output = '\,' do i= 1 to length (String) Output =
substr(String,i,1) \textbar\textbar{} Output end return Output

\hypertarget{right}{%
\subsection{RIGHT}\label{right}}

RIGHT returns characters that are on the right of the first argument.
The second argument specifies the

length of the result and the third is the padding character. call
CheckArgs `rANY rWHOLE\textgreater=0 oPAD'

String = \#Bif Arg.1 Length = \#Bif Arg.2 if \#Bif\_ArgExists.3 then Pad

\#Bif Arg.3 else Pad '

Trim = length(String) - Length if Trim \textgreater= 0 then return
substr(String,Trim + 1) return copies(Pad, -Trim) \textbar\textbar{}
String /* Pad string on the left */

\hypertarget{space}{%
\subsection{SPACE}\label{space}}

SPACE formats the blank-delimited words in the first argument with pad
characters between each word. The second argument is the number of pad
characters between each word and the third is the pad character.

call CheckArgs `rANY oOWHOLE\textgreater=0 oPAD'

String = \#Bif Arg.1

if \#Bif\_ArgExists.2 then Num = \#Bif\_ Arg.2 else Num = 1 if \#Bif
ArgExists.3 then Pad = \#Bif Arg.3 else Pad = ' '! Padding = copies(Pad,
Num) Output = subword(String, 1, 1) do i = 2 to words (String) Output =
Output \textbar\textbar{} Padding \textbar\textbar{} subword(String, i,
1)

end return Output

\hypertarget{strip}{%
\subsection{STRIP}\label{strip}}

STRIP removes characters from its first argument. The second argument
specifies whether the deletions are leading characters, trailing
characters or both. Each character deleted is equal to the third
argument, or equivalent to a blank if the third argument is omitted.

call CheckArgs `rANY oLTB oPAD'

String = \#Bif Arg.1 if \#Bif\_ArgExists.2 then Option = \#Bif Arg.2
else Option = `B' if \#Bif\_ ArgExists.3 then Unwanted = \#Bif\_ Arg.3
else Unwanted = \#A11Blanks\textless Index ``\#Al11Blanks'' \# ``\,''
\textgreater{}

if Option == `L' \textbar{} Option == `B' then do /* Strip leading
characters */ do while String == '\,' \& pos(left(String, 1), Unwanted)
\textgreater{} 0 String = substr(String, 2) end end

if Option == `T' \textbar{} Option == `B' then do /* Strip trailing
characters \emph{/ do while String == '\,' \& pos(right(String, 1),
Unwanted) \textgreater{} 0 String = left(String, length(String) -1) end
/} of while */ end return String

\hypertarget{substr}{%
\subsection{SUBSTR}\label{substr}}

SUBSTR returns a sub-string of the first argument. The second argument
specifies the position of the first character and the third specifies
the length of the sub-string. The fourth argument is the padding

character.

call CheckArgs `rANY rWHOLE\textgreater0 oOWHOLE\textgreater=0 oPAD'

String = \#Bif Arg.1 Num = \#Bif\_ Arg.2 if \#Bif\_ArgExists.3 then
Length = \#Bif Arg.3 else Length = max(length (String) +1-Num, 0) if
\#Bif\_ArgExists.4 then Pad = \#Bif\_ Arg.4 else Pad = ' `! Output =''
do Length \#Response Config Substr(String,Num) /* Attempt to fetch
character.*/ Character \#Outcome

Num = Num + 1 call Config Substr \#Response,1 /* Was there such a
character? */ if \#Outcome == `E' then do

/* Here if argument was not a character string. */

call Config C2B String call \#Raise `SYNTAX', 23.1, b2x(\#Outcome) /* No
return to here */

end if \#Outcome == `M' then Character = Pad Output=Output \textbar{}
\textbar{} Character end

return Output

\hypertarget{subword}{%
\subsection{SUBWORD}\label{subword}}

SUBWORD returns a sub-string of the first argument, comprised of words.
The second argument is the position in the first argument of the first
word of the sub-string. The third argument is the number of words in the
sub-string.

call CheckArgs `rANY rWHOLE\textgreater0 oWHOLE\textgreater=0'

String \#Bif Arg.1

Num \#Bif Arg.2

if \#Bif\_ ArgExists.3 then Length else Length

\#Bif\_Arg.3 length(String) /* Avoids call \emph{/ /} to WORDS() */

if Length = 0 then return '\,'

/* Find position of first included word */

Start = wordindex (String, Num)

if Start = 0 then return '\,' /* Start is beyond end */

/* Find position of first excluded word */ End = wordindex (String,
Num+Length) if End = 0 then End = length(String)+1

Output=substr (String, Start, End-Start)

/* Drop trailing blanks */

do while Output == '\,' if pos (right (Output,1),\#A11Blanks) = 0 then
leave Output = left (Output, length (Output) -1) end

return Output

\hypertarget{translate}{%
\subsection{TRANSLATE}\label{translate}}

TRANSLATE returns the characters of its first argument with each
character either unchanged or translated to another character.

call CheckArgs `rANY oANY oANY oPAD' String = \#Bif Arg.1 /* If neither
input nor output tables, uppercase. */ if \#Bif ArgExists.2 \&
\#Bif\_ArgExists.3 then do Output = '\,' do j=1 to length (String)
\#Response = Config Upper (substr (String,j,1))

Output = Output \textbar\textbar{} \#Outcome end j

return Output

end

/* The input table defaults to all characters. \emph{/ if \#Bif
ArgExists.3 then do \#Response = Config Xrange() Tablei = \#Outcome end
else Tablei = \#Bif\_Arg.3 /} The output table defaults to null \emph{/
if \#Bif\_ArgExists.2 then Tableo = \#Bif Arg.2 else Tableo = '\,' /}
The tables are made the same length */ if \#Bif\_ArgExists.4 then Pad =
\#Bif\_ Arg.4 else Pad = ' ' Tableo=left (Tableo, length (Tablei) , Pad)

107 Output='\,' do j=1 to length (String) c=substr (String,j,1)
k=pos(c,Tablei) if k=0 then Output=Output \textbar\textbar c else
Output=Output \textbar{} \textbar{} substr (Tableo,k,1) end j return
Output

\hypertarget{verify}{%
\subsection{VERIFY}\label{verify}}

VERIFY checks that its first argument contains only characters that are
in the second argument, or that it contains no characters from the
second argument; the third argument specifies which check is made. The
result is `0', or the position of the character that failed
verification. The fourth argument is a starting position for the check.

call CheckArgs `rANY rANY oMN oWHOLE\textgreater0'

String = \#Bif Arg.1

Reference = \#Bif\_Arg.2

if \#Bif\_ ArgExists.3 then Option \#Bif Arg.3 else Option 'N!

if \#Bif\_ ArgExists.4 then Start = \#Bif\_Arg.4 else Start = 1

Last = length(String) if Start \textgreater{} Last then return 0 if
Reference == '\,' then if Option == `N' then return Start else return 0

do i = Start to Last t = pos(substr(String, i, 1), Reference) if Option
== `N' then do if t = 0 then return i /* Return position of NoMatch
character. \emph{/ end else if t \textgreater{} 0 then return i /}
Return position of Matched character. */ end i return 0

\hypertarget{word}{%
\subsection{WORD}\label{word}}

WORD returns the word from the first argument at the position given by
the second argument.

call CheckArgs `rANY rwWHOLE\textgreater0'

return subword(\#Bif Arg.1, \#Bif\_Arg.2, 1)

\hypertarget{wordindex}{%
\subsection{WORDINDEX}\label{wordindex}}

WORDINDEX returns the character position in the first argument of a word
in the first argument. The second argument is the word position of that
word.

call CheckArgs `rANY rwWHOLE\textgreater0'

String Num

\#Bif Arg.1 \#Bif Arg.2

/* Find starting position */

Start = 1 Count = 0 do forever Start = verify(String,
\#Al1Blanks\textless Index ``\#A11Blanks'' \# ``\,'' \textgreater, `N',
Start) /\emph{ Find non-blank }/ if Start = 0 then return 0 /* Start is
beyond end \emph{/ Count = Count + 1 /} Words found \emph{/ if Count =
Num then leave Start = verify(String, \#Al1Blanks\textless Index
``\#Al11Blanks'' \# ``\,'' \textgreater, `M', Start + 1) /}

Find blank */

if Start = 0 then return 0 /* Start is beyond end */ end return Start

\hypertarget{wordlength}{%
\subsection{WORDLENGTH}\label{wordlength}}

WORDLENGTH returns the number of characters in a word from the first
argument. The second argument is the word position of that word.

call CheckArgs `rANY rwWHOLE\textgreater0'

return length(subword(\#Bif Arg.1, \#Bif\_Arg.2, 1))

\hypertarget{wordpos}{%
\subsection{WORDPOS}\label{wordpos}}

WORDPOS finds the leftmost occurrence in the second argument of the
sequence of words in the first argument. The result is `0' or the word
position in the second argument of the first word of the matched
sequence. Third argument is a word position for the start of the search.

call CheckArgs `rANY rANY oWHOLE\textgreater0'

Phrase = \#Bif\_Arg.1

String = \#Bif Arg.2

if \#Bif\_ArgExists.3 then Start = \#Bif\_Arg.3 else Start = 1

Phrase = space (Phrase) PhraseWords = words (Phrase) if PhraseWords = 0
then return 0 String = space (String) StringWords = words (String) do
WordNumber = Start to StringWords - PhraseWords + 1 if Phrase ==
subword(String, WordNumber, PhraseWords) then return WordNumber end
WordNumber return 0

\hypertarget{words}{%
\subsection{WORDS}\label{words}}

WORDS counts the number of words in its argument.

call CheckArgs `rANY'

do Count = 0 by 1 if subword(\#Bif Arg.1, Count + 1) == '\,' then return
Count end Count

\hypertarget{xrange}{%
\subsection{XRANGE}\label{xrange}}

XRANGE returns an ordered string of all valid character encodings in the
specified range.

call CheckArgs `oPAD oPAD'

if \#Bif\_ArgExists.1 then \#Bif\_Arg.1 mr if \#Bif\_ArgExists.2 then
\#Bif Arg.2 \#Response = Config Xrange(\#Bif Arg.1, \#Bif Arg.2) return
\#Outcome

\hypertarget{arithmetic-built-in-functions}{%
\section{Arithmetic built-in
functions}\label{arithmetic-built-in-functions}}

These functions perform arithmetic at the numeric settings current at
the invocation of the built-in function. Note that CheckArgs formats any
`NUM' (numeric) argument.

\hypertarget{abs}{%
\subsection{ABS}\label{abs}}

ABS returns the absolute value of its argument.

call CheckArgs `rNUM' Number=\#Bif Arg.1

if left (Number,1) = `-' then Number = substr (Number, 2) return Number

\hypertarget{format}{%
\subsection{FORMAT}\label{format}}

FORMAT formats its first argument. The second argument specifies the
number of characters to be used for the integer part and the third
specifies the number of characters for the decimal part. The fourth
argument specifies the number of characters for the exponent and the
fifth determines when exponeniial notation is used.

call CheckArgs, `rNUM OWHOLE\textgreater=0 OWHOLE\textgreater=0
OWHOLE\textgreater=0 OWHOLE\textgreater=0'

if \#Bif\_ArgExists.2 then Before \#Bif Arg.2 if \#Bif\_ArgExists.3 then
After \#Bif Arg.3 if \#Bif\_ArgExists.4 then Expp \#Bif Arg.4

if \#Bif\_ArgExists.5 then Expt = \#Bif Arg.5

/* In the simplest case the first is the only argument. */ Number=\#Bif
Arg.1

if \#Bif\_Arg.0 \textless{} 2 then return Number

/* Dissect the Number. It is in the normal Rexx format. */ parse var
Number Mantissa `E' Exponent

if Exponent == '\,' then Exponent = 0 Sign = 0 if left (Mantissa,1) ==
`-' then do Sign = 1 Mantissa = substr(Mantissa, 2) end parse var
Mantissa Befo `.' Afte

/* Count from the left for the decimal point. */

Point = length (Befo)

/* Sign Mantissa and Exponent now reflect the Number. Befo Afte and
Point reflect Mantissa. */

/* The fourth and fifth arguments allow for exponential notation.
\emph{/ /} Decide whether exponential form to be used, setting ShowExp.
\emph{/ ShowExp = 0 if \#Bif\_ArgExists.4 \#Bif ArgExists.5 then do if
\#Bif\_ArgExists.5 then Expt = \#Digits.\#Level /} Decide whether
exponential form to be used. \emph{/ if (Point + Exponent)
\textgreater{} Expt then ShowExp = 1 /} Digits before rule. \emph{/
LeftOfPoint = 0 if length(Befo) \textgreater{} 0 then LeftOfPoint = Befo
/} Value left of the point */

/* Digits after point rule for exponentiation: */

/* Count zeros to right of point. */

ze=O0

do while substr(Afte,z+1,1) == `0' Zeze+41tl end

if LeftOfPoint = 0 \& (z - Exponent) \textgreater{} 5 then ShowExp = 1

/* An extra rule for exponential form: */ if \#Bif\_ArgExists.4 then if
Expp = 0 then ShowExp = 0

/* Construct the exponential part of the result. */ if ShowExp then do

Exponent = Exponent + ( Point - 1 ) Point = 1 /* As required for
'SCIENTIFIC! */ if \#Form.\#Level == `ENGINEERING' then do while
Exponent//3 = 0

Point = Point+1l

Exponent = Exponent-1

end end

if \ShowExp then Point = Point + Exponent end /* Expp or Expt given
\emph{/ else do /} Even if Expp and Expt are not given, exponential
notation will be used if the original Number+0 done by CheckArgs led to
it. */ if Exponent = 0 then do ShowExp = 1

110 111

end end

/* ShowExp now indicates whether to show an exponent, Exponent is its
value. \emph{/ /} Make this a Number without a point.

Integer = Befo\textbar\textbar Afte

*/

/* Make sure Point position isn't disjoint from Integer. \emph{/ if
Point\textless1 then do /} Extra zeros on the left. */

Integer = Point = 1 end

copies(`0',1 - Point)

if Point \textgreater{} length(Integer) then Integer =
left(Integer,Point,`0') /* And maybe on the right.

/* Deal with right of decimal point first since that can affect the

\textbar\textbar{} Integer

left. Ensure the requested number of digits there. Afters =
length(Integer) -Point

if \#Bif\_ArgExists.3 = 0 then After = /* Make Afters match the
requested After */

do while Afters \textless{} After Afters = Afters+1 Integer = Integer'0'
end

if Afters \textgreater{} After then do

/* Round by adding 5 at the right place.

Afters

r=substr (Integer, Point + After + 1,

Integer = left (Integer,

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
\end{enumerate}

Point + After)

if r \textgreater= `5' then Integer = Integer + 1 /* This can leave the
result zero. If Integer = 0 then Sign = 0

/* The case when rounding makes the integer longer is an awkward

*/

*/

one. The exponent will have to be adjusted. */ if length(Integer)
\textgreater{} Point + After then do

Point = Point+l end

if ShowExp = 1 then do

Exponent=Exponent + (Point - 1)

Point = 1 /* As required for 'SCIENTIFIC! */

if form() = `ENGINEERING' then do while Exponent//3 = 0

Point = Point+

1

Exponent = Exponent-1

end end

t = Point-length (Integer) if t \textgreater{} 0 then Integer =
Integer\textbar\textbar copies(`0',t)

end /* Rounded */

/* Right part is final if After \textgreater{} 0 then Afte else Afte

now. */ `.'\textbar{} \textbar substr (Integer, Point+1,After)

/* Now deal with the integer part of the result. Integer = left
(Integer, Point)

if \#Bif ArgExists.2 =

0 then Before

/* Make Point match Before */ if Point \textgreater{} Before - Sign then
call Raise

do while Point\textless Before Point = Point+1l Integer = '0'Integer end

40.38,

*/

Point + Sign /* Note default.

2,

/* Find the Sign position and blank leading zeroes.

re ter Triggered = 0

do j = 1 to length (Integer) Digit = substr(Integer,j,1) /* Triggered is
set when sign inserted or blanking finished. if Triggered = 1 then do

r= r\textbar\textbar Daigit iterate end

*/

/* Note default.

\#Bif Arg.1

*/

*/

*/

*/

\emph{/ /} If before sign insertion point then blank out zero. */

if Digit = `0' then if ats cael = `0' \& j+l\textless length(Integer)
then do re r ' ' iterate end /* j is the sign insertion point. */ if
Digit = `0' \& j = length(Integer) then Digit = ' '

if Sign = 1 then Digit = `-' r= xr\textbar\textbar Digit Triggered = 1
end j Number = r\textbar\textbar Afte

if ShowExp = 1 then do /* Format the exponent. \emph{/ Expart = '\,'
SignExp = 0 if Exponent\textless0 then do SignExp = 1 Exponent =
-Exponent end /} Make the exponent to the requested width. */ if \#Bif\_
ArgExists.4 = 0 then Expp = length (Exponent) if length(Exponent)
\textgreater{} Expp then call Raise 40.38, 4, \#Bif\_ Arg.1
Exponent=right (Exponent,Expp,`0') if Exponent = 0 then do if \#Bif\_
ArgExists.4 then Expart = copies(' ',expp+2) end else if SignExp = 0
then Expart else Expart Number = Number \textbar{} \textbar Expart end
return Number

`E+' Exponent `E-' Exponent

\hypertarget{max}{%
\subsection{MAX}\label{max}}

MAX returns the largest of its arguments.

if \#Bif\_Arg.0 \textless1 then call Raise 40.3, 1 call CheckArgs
`rNUM'\textbar\textbar copies(' rNUM', \#Bif Arg.0 - 1)

Max = \#Bif Arg.1

do i = 2 to \#Bif Arg.0 by 1 Next = \#Bif Arg.i if Max \textless{} Next
then Max = Next end i

return Max

\hypertarget{min}{%
\subsection{MIN}\label{min}}

MIN returns the smallest of its arguments.

if \#Bif\_Arg.0 \textless1 then call Raise 40.3, 1 call CheckArgs
`rNUM'\textbar\textbar copies(' rNUM', \#Bif Arg.0 - 1)

Min = \#Bif Arg.1

do i = 2 to \#Bif Arg.0 by 1 Next = \#Bif Arg.i if Min \textgreater{}
Next then Min = Next end i

return Min

\hypertarget{sign}{%
\subsection{SIGN}\label{sign}}

SIGN returns `1', `0' or `-1' according to whether its argument is
greater than, equal to, or less than zero.

call CheckArgs `rNUM'

Number = \#Bif Arg.1

select

when Number \textless{} 0 then Output = -1 when Number = 0 then Output =
0 when Number \textgreater{} 0 then Output = 1

end return Output

\hypertarget{trunc}{%
\subsection{TRUNC}\label{trunc}}

TRUNC returns the integer part of its argument, or the integer part plus
a number of digits after the decimal point, specified by the second
argument.

call CheckArgs `rNUM oWHOLE\textgreater=0'

Number = \#Bif Arg.1 if \#Bif\_ArgExists.2 then Num

\#Bif Arg.2 else Num 0

Integer =(10**Num * Number) \%1 if Num=0 then return Integer

t=length (Integer) -Num if t\textless=0 then return
`0.'right(Integer,Num,'0') else return insert(`.',Integer,t)

\hypertarget{state-built-in-functions}{%
\section{State built-in functions}\label{state-built-in-functions}}

These functions return values from the state of the execution.

\hypertarget{address}{%
\subsection{ADDRESS}\label{address}}

ADDRESS returns the name of the environment to which commands are
currently being submitted. Optionally, under control by the argument, it
also returns information on the targets of command output and the source
of command input.

call CheckArgs `oEINO'

if \#Bif\_ArgExists.1 then Optionl = \#Bif\_Arg.1 else Optionl=`N'

if Optionl == `N' then return \#Env\_Name.ACTIVE. \#Level

Tail = Optionl' .ACTIVE. '\#Level return \#Env\_Position.Tail
\#Env\_Type.Tail \#Env\_Resource.Tail

\hypertarget{arg}{%
\subsection{ARG}\label{arg}}

ARG returns information about the argument strings to a program or
routine, or the value of one of those strings.

ArgData = `OWHOLE\textgreater0 oENO' if \#Bif\_ ArgExists.2 then ArgData
= `rWHOLE\textgreater0 rENO' call CheckArgs ArgData

if \#Bif\_ArgExists.1 then return \#Arg.\#Level.0

ArgNum=\#Bif Arg.1

if \#Bif\_ArgExists.2 then return \#Arg.\#Level.ArgNum

if \#Bif\_Arg.2 ==`0O' then return \#ArgExists.\#Level.ArgNum else
return \#ArgExists.\#Level .ArgNum

\hypertarget{condition}{%
\subsection{CONDITION}\label{condition}}

CONDITION returns information associated with the current condition.

call CheckArgs `oCDEIS'

/* Values are null if this is not following a condition. */ if
\#Condition.\#Level == '\,' then do \#ConditionDescription.\#Level =
'\,' \#ConditionExtra.\#Level = '\,' \#ConditionInstruction.\#Level end

Option=\#Bif Arg.1

if Option==`C' then return \#Condition.\#Level

if Option==`D' then return \#ConditionDescription. \#Level if
Option==`E' then return \#ConditionExtra.\#Level

if Option==`I' then return \#ConditionInstruction. \#Level /* State is
the current state. */

if \#Condition.\#Level = '\,' then return ``\,'' return
\#Enabling.\#Condition.\#Level \#\#\# DIGITS

DIGITS returns the current setting of NUMERIC DIGITS. call CheckArgs
'\,'

return \#Digits.\#Level

\hypertarget{errortext}{%
\subsection{ERRORTEXT}\label{errortext}}

ERRORTEXT returns the unexpanded text of the message which is identified
by the first argument. A second argument of `S' selects the standard
English text, otherwise the text may be translated to another national
language. This translation is not shown in the code below.

call CheckArgs `r0\_90 oSN'

msgcode = \#Bif Arg.1

if \#Bif\_ ArgExists.2 then Option else Option

return \#ErrorText .msgcode

\hypertarget{form}{%
\subsection{FORM}\label{form}}

FORM returns the current setting of NUMERIC FORM.

\#Bif\_Arg.2 tint

call CheckArgs '\,'

return \#Form.\#Level

\hypertarget{fuzz}{%
\subsection{FUZZ}\label{fuzz}}

FUZZ returns the current setting of NUMERIC FUZZ.

call CheckArgs '\,'

return \#Fuzz.\#Level

\hypertarget{sourceline}{%
\subsection{SOURCELINE}\label{sourceline}}

If there is no argument, SOURCELINE returns the number of lines in the
program, or `0' if the source program is not being shown on this
execution. If there is an argument it specifies the number of the line
of the source program to be returned.

call CheckArgs `oWHOLE\textgreater0'

if \#Bif\_ArgExists.1 then return \#SourceLine.0 Num = \#Bif\_Arg.1 if
Num \textgreater{} \#SourceLine.0O then call Raise 40.34, Num,
\#SourceLine.0 return \#SourceLine.Num

\hypertarget{trace}{%
\subsection{TRACE}\label{trace}}

TRACE returns the trace setting currently in effect, and optionally
alters the setting.

call CheckArgs `oACEFILNOR' /* Also checks for `?!' \emph{/ /} With no
argument, this a simple query. */ Output=\#Tracing.\#Level

if \#Interactive.\#Level then Output = `?'\textbar\textbar Output if
\#Bif\_ArgExists.1 then return Output

Value=\#Bif Arg.1

\#Interactive.\#Level=0

/* A question mark sets the interactive flag. */ if left(Value,1)==`?'
then do

\#Interactive.\#Level = 1

Value=substr (Value, 2)

end /* Absence of a letter leaves the setting unchanged. */ if
Value=='\,' then do

Value=translate (left (Value,1)) if Value==`0O' then
\#Interactive.\#Level=`0' \#Tracing.\#Level = Value end return Output

10.4 Conversion built-in functions

Conversions between Binary form, Decimal form, and heXadecimal form do
not depend on the encoding (see nnn) of the character data.

Conversion to Coded form gives a result which depends on the encoding.
Depending on the encoding, the result may be a string that does not
represent any sequence of characters.

\hypertarget{b2xx}{%
\subsection{B2xX}\label{b2xx}}

B2X performs binary to hexadecimal conversion.

call CheckArgs `rBIN'

String = space (\#Bif Arg.1,0) return ReRadix(String,2,16)

\hypertarget{bitand}{%
\subsection{BITAND}\label{bitand}}

The functions BITAND, BITOR and BITXOR operate on encoded character
data. Each binary digit from the encoding of the first argument is
processed in conjunction with the corresponding bit from the second
argument.

call CheckArgs `rANY oANY oPAD'

Stringl = \#Bif\_ Arg.1 if \#Bif ArgExists.2 then String2

\#Bif Arg.2 else String2 rr

/* Presence of a pad implies character strings. */ if \#Bif\_ArgExists.3
then if length(Stringl) \textgreater{} length(String2) then String2=left
(String2,length(String1l),\#Bif\_Arg.3) else
Stringl=left(Stringl,length(String2),\#Bif\_Arg.3)

/* Change to manifest bit representation. \emph{/ \#Response=Config
C2B(String1) String1l=\#Outcome \#Response=Config C2B(String2)
String2=\#0utcome /} Exchange if necessary to make shorter second. */ if
length(Stringl)\textless length(String2) then do

t=Stringl

Stringl=String2

String2=t

end

/* Operate on common length of those bit strings. */ r='\,' do j=1 to
length (String2)

bl=substr (Stringl,j,1)

b2=substr (String2,j,1)

select when \#Bif=`BITAND' then b1=b1\&b2 when \#Bif=`BITOR' then
bl=b1\textbar b2

115 when \#Bif=`BITXOR' then b1=b1\&\&b2 end r=r\textbar\textbar bl end
j rer \textbar\textbar{} right (Stringl, length (String1)
-length(String2) )

/* Convert back to encoded characters. */ return x2c (b2x(r))

\hypertarget{bitor}{%
\subsection{BITOR}\label{bitor}}

See nnn

\hypertarget{bitxor}{%
\subsection{BITXOR}\label{bitxor}}

See nnn

\hypertarget{c2d}{%
\subsection{C2D}\label{c2d}}

C2D performs coded to decimal conversion.

call CheckArgs `rANY oWHOLE\textgreater=0' if length(\#Bif Arg.1)=0 then
return 0

if \#Bif\_ArgExists.2 then do /* Size specified \emph{/ Size =
\#Bif\_Arg.2 if Size = 0 then return 0 /} Pad will normally be zeros
\emph{/ t=right (\#Bif Arg.1,Size,left (xrange(),1)) /} Convert to
manifest bit \emph{/ call Config C2B t /} And then to signed decimal.
\emph{/ Sign = Left (\#Outcome,1) \#Outcome = substr(\#Outcome, 2)
t=ReRadix (\#Outcome, 2,10) /} Sign indicates 2s-complement. \emph{/ if
Sign then t=t-2\textbf{length(\#Outcome) if abs(t) \textgreater{} 10 }
\#Digits.\#Level - 1 then call Raise 40.35, t return t end /} Size not
specified. */ call Config C2B \#Bif\_Arg.1 t = ReRadix(\#Outcome, 2,10)
if t \textgreater{} 10 ** \#Digits.\#Level - 1 then call Raise 40.35, t
return t

10.4.6 C2Xx C2X performs coded to hexadecimal conversion.

call CheckArgs `rANY' if length(\#Bif Arg.1) = 0 then return '\,'

call Config C2B \#Bif\_Arg.1 return ReRadix (\#Outcome,2,16)

\hypertarget{d2c}{%
\subsection{D2C}\label{d2c}}

D2C performs decimal to coded conversion.

`rWHOLENUM\textgreater=0'! `rWHOLENUM rWHOLE\textgreater=0'

if \#Bif\_ArgExists.2 then ArgData else ArgData call CheckArgs ArgData

/* Convert to manifest binary \emph{/ Subject = abs(\#Bif Arg.1) r =
ReRadix(Subject,10,2) /} Make length a multiple of 8, as required for
Config B2C */ Length = length(r) do while Length//8 = 0 Length =
Length+1 end r= right(r,Length,`0')

/* 2s-complement for negatives. */ if \#Bif\_Arg.1\textless0 then do
Subject = 2**length(r)-Subject r = ReRadix(Subject,10,2) end /* Convert
to characters */ \#Response = Config B2C(r) Output = \#Outcome if
\#Bif\_ArgExists.2 then return Output

/* Adjust the length with appropriate characters. */ if
\#Bif\_Arg.1\textgreater=0 then return right (Output,
\#Bif\_Arg.2,left(xrange(),1)) else return right (Output, \#Bif
Arg.2,right (xrange(),1))

\hypertarget{d2x}{%
\subsection{D2X}\label{d2x}}

D2X performs decimal to hexadecimal conversion. if \#Bif\_ArgExists.2
then ArgData = `rWHOLENUM\textgreater=0' else ArgData = `rWHOLENUM
rWHOLE\textgreater=0'

call CheckArgs ArgData

/* Convert to manifest hexadecimal \emph{/ Subject = abs(\#Bif Arg.1 ) r
= ReRadix(Subject,10,16) /} Twos-complement for negatives */ if
\#Bif\_Arg.1\textless0 then do Subject = 16**length(r) -Subject r =
ReRadix(Subject,10,16) end if \#Bif\_ArgExists.2 then return r /* Adjust
the length with appropriate characters. */ if \#Bif\_
Arg.1\textgreater=0 then return right(r,\#Bif Arg.2,`0') else return
right(r,\#Bif Arg.2,`F')

\hypertarget{x2b}{%
\subsection{X2B}\label{x2b}}

X2B performs hexadecimal to binary conversion.

call CheckArgs `rHEX'

Subject = \#Bif Arg.1

if Subject == '\,' then return '\,'

/* Blanks were checked by CheckArgs, here they are ignored. */ Subject =
space (Subject, 0)

return ReRadix(translate (Subject) ,16,2)

\hypertarget{x2c}{%
\subsection{X2C}\label{x2c}}

X2C performs hexadecimal to coded character conversion.

call CheckArgs `rHEX'

Subject = \#Bif Arg.1

if Subject == '\,' then return '\,'

Subject = space (Subject, 0)

/* Convert to manifest binary */

r = ReRadix(translate (Subject) ,16,2)

/* Convert to character */

Length = 8*( (length (Subject) +1) \%2) \#Response = Config
B2C(right(r,Length,`0')) return \#Outcome

\hypertarget{x2d}{%
\subsection{X2D}\label{x2d}}

X2D performs hexadecimal to decimal conversion.

call CheckArgs `rHEX OWHOLE\textgreater=0'

Subject = \#Bif Arg.1 if Subject == '\,' then return `0'

Subject = translate (space (Subject,0))

if \#Bif ArgExists.2 then

Subject = right (Subject, \#Bif Arg.2,`0') if Subject =='\,' then return
`0' /* Note the sign */ if \#Bif ArgExists.2 then SignBit

else SignBit

/* Convert to decimal \emph{/ r = ReRadix(Subject,16,10) /}
Twos-complement */ if SignBit then r = 2**(4*\#Bif Arg.2) - r if
abs(r)\textgreater10 ** \#Digits.\#Level - 1 then call Raise 40.35, t
return r

left (x2b (Subject) ,1) ror

\hypertarget{inputoutput-built-in-functions}{%
\section{Input/Output built-in
functions}\label{inputoutput-built-in-functions}}

The configuration shall provide the ability to access streams. Streams
are identified by character string identifiers and provide for the
reading and writing of data. They shall support the concepts of
characters, lines, and positioning. The input/output built-in functions
interact with one another, and they make use of Config\_ functions, see
nnn. When the operations are successful the following characteristics
shall be exhibited: - The CHARIN/CHAROUT functions are insensitive to
the lengths of the arguments. The data written to a stream by CHAROUT
can be read by a different number of CHARINs. - The CHARIN/CHAROUT
functions are reflective, that is, the concatenation of the data read
from a persistent stream by CHARIN (after positioning to 1, while
CHARS(Stream)\textgreater0), will be the same as the concatenation of
the data put by CHAROUT. - All characters can be used as CHARIN/CHAROUT
data. - The CHARS(Stream, `N') function will return zero only when a
subsequent read (without positioning) is guaranteed to raise the
NOTREADY condition. - The LINEIN/LINEOUT functions are sensitive to the
length of the arguments, that is, the length of a line written by
LINEOUT is the same as the length of the string returned by successful
LINEIN of the line. - Some characters, call them line-banned characters,
cannot reliably be used as data for LINEIN/LINEOUT. If these are not
used, LINEIN/LINEOUT is reflective. If they are used, the result is not
defined. The set of characters which are line-barred is a property of
the configuration. - The LINES(Stream, `N') function will return zero
only when a subsequent LINEIN (without positioning) is guaranteed to
raise the NOTREADY condition. - When a persistent stream is repositioned
and written to with CHAROUT, the previously written data is not lost,
except for the data overwritten by this latest CHAROUT. - When a
persistent stream is repositioned and written to with LINEOUT, the
previously written data is not lost, except for the data overwritten by
this latest LINEOUT, which may leave lines partially overwritten. \#\#\#
CHARIN CHARIN returns a string read from the stream named by the first
argument.

call CheckArgs `oSTREAM oOWHOLE\textgreater0 oOWHOLE\textgreater=0'

if \#Bif\_ArgExists.1 then Stream else Stream \#StreamState.Stream =
'\,' /* Argument 2 is positioning. */ if \#Bif\_ArgExists.2 then do
\#Response = Config Stream Position(Stream, `CHARIN', \#Bif Arg.2)

\#Bif Arg.1

if left(\#Response, 1) == `R' then call Raise 40.41, 2, \#Bif\_ Arg.2 if
left(\#Response, 1) == `T' then call Raise 40.42,Stream end

/* Argument 3 is how many. \emph{/ if \#Bif\_ArgExists.3 then Count else
Count if Count = 0 then do call Config Stream Charin Stream, `NULL' /}
``Touch'' the stream \emph{/ return '! end /} The unit may be eight bits
(as characters) or one character. */ call Config Stream Query Stream

\#Bif Arg.3 1

Mode = \#Outcome

do until Count = 0

\#Response = Config Stream Charin(Stream, `CHARIN')

if left(\#Response, 1) == `N' then do if left (\#Response, 1) == `E'
then \#StreamState.Stream = `ERROR' /* This call will return. */ call
\#Raise `NOTREADY', Stream, substr(\#Response, 2) leave end

r = r\textbar{} \textbar\#Outcome

Count = Count-1

end if Mode == `B' then do call Config B2C r r = \#Outcome end return r

\hypertarget{charout}{%
\subsection{CHAROUT}\label{charout}}

CHAROUT returns the count of characters remaining after attempting to
write the second argument to the stream named by the first argument.

call CheckArgs `oSTREAM oANY oWHOLE\textgreater0'

if \#Bif\_ArgExists.1 then Stream else Stream

\#Bif Arg.1

\#StreamState.Stream = '\,' if \#Bif ArgExists.2 \& \#Bif\_ArgExists.3
then do /* Position to end of stream. */ \#Response = Config Stream
Close (Stream) if left (\#Response,1) == `T' then call Raise
40.42,Stream return 0 end

if \#Bif\_ArgExists.3 then do /* Explicit positioning. */ \#Response =
Config Stream Position(Stream, `CHAROUT', \#Bif Arg.3)

if left (\#Response,1) == `T' then call Raise 40.42,Stream if
left(\#Response, 1) == `R' then call Raise 40.41, 3, \#Bif\_ Arg.3 end

if \#Bif\_ArgExists.2 \textbar{} \#Bif\_Arg.2 == '\,' then do call
Config Stream \_Charout Stream, `NULL' /* ``Touch'' the stream */ return
0 end

String = \#Bif Arg.2 call Config Stream Query Stream Mode = \#Outcome if
Mode == `B' then do call Config C2B String String = \#Outcome Stride = 8
Residue = length(String)/8 end else do Stride = 1 Residue = length
(String) end

Cursor = 1 do while Residue\textgreater0 Piece = substr (String, Cursor,
Stride) Cursor = Cursor+Stride call Config Stream Charout Stream, Piece
if left (\#Response, 1) == `N' then do if left (\#Response, 1) == `E'
then \#StreamState.Stream = `ERROR' call \#Raise `NOTREADY', Stream,
substr(\#Response, 2)

return Residue end Residue = Residue - 1 end return 0

\hypertarget{chars}{%
\subsection{CHARS}\label{chars}}

CHARS indicates whether there are characters remaining in the named
stream. Optionally, it returns a

count of the characters remaining and immediately available.

call CheckArgs `oSTREAM oCN'

if \#Bif\_ArgExists.1 then Stream = \#Bif\_Arg.1 else Stream = '\,'

if \#Bif\_ArgExists.2 then Option = \#Bif Arg.2 else Option = `N'

call Config Stream Count Stream, `CHARS', Option return \#Outcome

\hypertarget{linein}{%
\subsection{LINEIN}\label{linein}}

LINEIN reads a line from the stream named by the first argument, unless
the third argument is Zero.

call CheckArgs `oSTREAM oOWHOLE\textgreater0 oOWHOLE\textgreater=0'

if \#Bif\_ArgExists.1 then Stream else Stream \#StreamState.Stream =
'\,' if \#Bif\_ArgExists.2 then do \#Response = Config Stream
Position(Stream, `LINEIN', \#Bif Arg2) if left (\#Response, 1) `T' then
call Raise 40.42,Stream if left (\#Response, 1) `R' then call Raise
40.41, 2, \#Bif\_ Arg.2 end if \#Bif\_ArgExists.3 then Count \#Bif Arg.3
else Count 1 if Count\textgreater1 then call Raise 40.39, Count if Count
= 0 then do call Config Stream Charin Stream, `NULL' /* ``Touch'' the
stream \emph{/ return `! end /* A configuration may recognise lines even
in 'binary' mode. }/ call Config Stream Query Stream Mode = \#Outcome re
ter t = \#Linein Position.Stream /* Config Stream Charin will alter
\#Linein Position. */ do until t = \#Linein Position.Stream \#Response =
Config Stream Charin(Stream, `LINEIN') if left(\#Response, 1) == `N'
then do if left (\#Response, 1) == `E' then \#StreamState.Stream =
`ERROR' call \#Raise `NOTREADY', Stream, substr(\#Response, 2) leave end
r = r\textbar\textbar\#Outcome end if Mode == `B' then do call Config
B2C r r= \#Outcome end return r

\#Bif Arg.1

\hypertarget{lineout}{%
\subsection{LINEOUT}\label{lineout}}

LINEOUT returns `1' or `0', indicating whether the second argument has
been successfully written to the

stream named by the first argument. A result of `1' means an
unsuccessful write. call CheckArgs `oSTREAM oANY oWHOLE\textgreater0'

if \#Bif\_ArgExists.1 then Stream else Stream

\#Bif Arg.1

\#StreamState.Stream = '\,'

if \#Bif ArgExists.2 \& \#Bif\_ArgExists.3 then do

/* Position to end of stream.

*/

\#Response = Config Stream Close (Stream) if left (\#Response,1) == `T'
then call Raise 40.42,Stream

return 0 end

if \#Bif\_ArgExists.3 then do

\#Response = Config Stream Position(Stream, `LINEOUT', \#Bif Arg.3)

if left(\#Response, 1) == `T' then call Raise 40.42,Stream if left
(\#Response, 1) == `R' then call Raise 40.41, 3, \#Bif\_Arg.3 end

if \#Bif ArgExists.2 then do

call Config Stream \_Charout Stream, '\,' /* ``Touch'' the stream */

return 0 end

String \#Bif Arg.2 Stride 1 call Config Stream Query Stream Mode =
\#Outcome if Mode == `B' then do call Config C2B String String =
\#Outcome Stride = 8 Residue = length(String)/8 end else do Stride = 1
Residue = length(String) end Cursor = 1 do while Residue \textgreater{}
0

Piece = substr (String, Cursor, Stride)

Cursor = Cursor+Stride

call Config Stream Charout Stream, Piece then do

then \#StreamState.Stream = `ERROR'

call \#Raise `NOTREADY', Stream, substr(\#Response, 2)

if left(\#Response, 1) == `N' if left (\#Response, 1) == `E' return 1
end Residue = Residue-1 end call Config Stream Charout Stream, return 0
\#\#\# LINES

`EOL'

LINES returns the number of lines remaining in the named stream.

call CheckArgs `oSTREAM oCN'

if \#Bif\_ArgExists.1 then Stream else Stream if \#Bif\_ ArgExists.2
then Option else Option

Call Config Stream Count Stream, return \#Outcome

\hypertarget{qualify}{%
\subsection{QUALIFY}\label{qualify}}

\#Bif Arg.1 ter \#Bif\_Arg.2 tint

LINES', Option

QUALIFY returns a name for the stream named by the argument. The two
names are currently associated with the same resource and the result of
QUALIFY may be more persistently associated with

that resource. call CheckArgs `oSTREAM'

if \#Bif\_ArgExists.1 then Stream else Stream

\#Bif Arg.1 \#Response = Config Stream Qualified (Stream) return
\#Outcome

\hypertarget{stream}{%
\subsection{STREAM}\label{stream}}

STREAM returns a description of the state of, or the result of an
operation upon, the stream named by the first argument.

/* Third argument is only correct with `C' */

if \#Bif ArgExists.2 \& translate(left(\#Bif Arg.2, 1)) == `C' then
ArgData = `rSTREAM rCDS rANY'

else ArgData = `rSTREAM oCDS'

call CheckArgs ArgData

Stream = \#Bif\_Arg.1

if \#Bif\_ ArgExists.2 then Operation = \#Bif\_Arg.2 else Operation =
`S' Select when Operation == `C' then do

call Config Stream Command Stream, \#Bif Arg.3 return \#Outcome

end when Operation == `D' then do \#Response = Config Stream State
(Stream) return substr(\#Response, 2) end when Operation == `S' then do
if StreamState.Stream == `ERROR' then return `ERROR' \#Response = Config
Stream State (Stream)

`N' then return `READY' `U' then return `UNKNOWN'

if left (\#Response, if left (\#Response, return `NOTREADY' end

end

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  ==
\item
  ==
\end{enumerate}

\hypertarget{other-built-in-functions}{%
\section{Other built-in functions}\label{other-built-in-functions}}

\hypertarget{date}{%
\subsection{DATE}\label{date}}

DATE with fewer than two arguments returns the local date. Otherwise it
converts the second argument (which has a format given by the third
argument) to the format specified by the first argument. If there are
fourth or fifth arguments, they describe the treatment of separators
between fields of the date.

call CheckArgs `oBDEMNOSUW oANY oOBDENOSU oSEP oSEP' /* If the third
argument is given then the second is mandatory. */ if \#Bif\_ArgExists.3
\& \#Bif\_ArgExists.2 then

call Raise 40.19, '\,', \#Bif Arg.3

if \#Bif\_ ArgExists.1 then Option else Option

\#Bif Arg.1 tint

/* The date/time is `frozen' throughout a clause. \emph{/ if
\#ClauseTime.\#Level == '\,' then do \#Response = Config Time ()
\#ClauseTime.\#Level = \#Time \#ClauseLocal.\#Level = \#Time +
\#Adjust\textless Index ``\#Adjust'' \# ``\,'' \textgreater{} end /}
English spellings are used, even if messages not in English are used. */
Months = `January February March April May June July', `August September
October November December' WeekDays = `Monday Tuesday Wednesday Thursday
Friday Saturday Sunday'

/* If there is no second argument, the current date is returned. */ if
\#Bif ArgExists.2 then return DateFormat (\#ClauseLocal.\#Level, Option)

/* If there is a second argument it provides the date to be converted.
*/

Value = \#Bif\_Arg.2 if \#Bif\_ ArgExists.3 then InOption else InOption
if Option == `S' then OutSeparator else OutSeparator if
\#Bif\_ArgExists.4 then do

if OutSeparator == `x' then call OutSeparator = \#Bif.Arg.4 end

if InOption == `S' then InSeparator

else InSeparator if \#Bif\_ArgExists.5 then do

\#Bif\_Arg.3 'nt

translate (Option, ``xx/x //x'',``BDEMNOUW'')

Raise 40.46, Option, 4

translate(InOption,``xx/ //'',``BDENOU'')

if InSeparator == `x' then call Raise 40.46, InOption, 5

InSeparator = \#Bif.Arg.5 end /* First try for Year Month Day */ Logic =
`NS' select when InOption == `N' then do if InSeparator == '\,' then do

if length(Value)\textless9 then return Year = right (Value, 4)

MonthIs = substr (right (Value,7),1,3) Day = left (Value, length (Value)
-7)

end else

parse var Value Day (InSeparator) MonthIs (InSeparator) Year

do Month = 1 to 12 if left(word(Months, Month), 3)

== MonthIs then leave

parse var Value Year (InSeparator) Month (InSeparator) Day

end Month

end

when InOption == `S' then if InSeparator == '\,' then

parse var Value Year +4 Month +2 Day

else

otherwise Logic = `EOU' /* or BD */

end

/* Next try for year without century */

parse var Value Day (InSeparator) Month (InSeparator) YY

parse var Value YY (InSeparator) Month (InSeparator) Day

parse var Value Month (InSeparator) Day (InSeparator) YY

if logic = `EOU' then Select when InOption == `E' then if InSeparator ==
'\,' then parse var Value Day +2 Month +2 YY else when InOption == `O'
then if InSeparator == '\,' then parse var Value YY +2 Month +2 Day else
when InOption == `U' then if InSeparator == '\,' then parse var Value
Month +2 Day +2 YY else otherwise Logic = `BD' end if Logic = `EOU' then
do

/* The century is assumed, on the basis of the current year. */

if datatype(YY,`W')=0 then return

YearNow = left(`DATE'(`S'),4)

Year = YY

do while Year \textless{} YearNow-50 Year = Year + 100

end

end /* Century assumption */

if Logic \textless\textgreater{} `BD' then do /* Convert Month \& Day to
Days of year. \emph{/ if datatype(Month,`W')=0 \textbar{}
datatype(Day,`W')=0 \textbar{} datatype(Year,`W')=0 then return Days =
word(`0 31 59 90 120 151 181 212 243 273 304 334',Month), +
(Month\textgreater2)}Leap(Year) + Day-1

end else

if datatype(Value,`W')=0 then

return

if InOption == `D' then do

Year = left(`DATE' (`S'),4)

Days = Value - 1 /* `D' includes current day */ end

/* Convert to BaseDays */ if InOption \textless\textgreater{} `B' then

BaseDays = (Year-1)*365 + (Year-1)\%4 - (Year-1)\%100 + (Year-1)\%400 +
Days else

Basedays = Value

/* Convert to microseconds from 0001 \emph{/ Micro = BaseDays } 86400 *
1000000

/* Reconvert to check the original. (eg for Month = 99) */

if DateFormat (Micro,InOption,InSeparator) == Value then call Raise
40.19, Value, InOption

return DateFormat (Micro, Option, OutSeparator)

DateFormat: /* Convert from microseconds to given format. */ parse value
Time2Date(arg(1)) with, Year Month Day Hour Minute Second Microsecond
Base Days

select

when arg(2) == `B' then

return Base when arg(2) == `D' then

return Days when arg(2) == `E' then

return right (Day,2,`0') (arg(3)) right (Month,2,`0') (arg(3)) right
(Year,2,`0') when arg(2) == `M' then

return word (Months ,Month) when arg(2) == `N' then

return (Day) (arg(3)) left (word(Months,Month),3) (arg(3))right
(Year,4,`0') when arg(2) == `O' then

return right (Year,2,`0') (arg(3))right (Month,2,`0') (arg(3))right
(Day,2,`0') when arg(2) == `S' then

return right (Year,4,`0') (arg(3))right (Month,2,`0') (arg(3))right
(Day,2,`0') when arg(2) == `U' then

return right (Month,2,`0') (arg(3)) right (Day,2,`0') (arg(3)) right
(Year,2,`0') otherwise /* arg(2) == `W' */

return word (Weekdays, 1+Base//7)

end

\hypertarget{queued}{%
\subsection{QUEUED}\label{queued}}

QUEUED returns the number of lines remaining in the external data queue.

call CheckArgs '\,'

\#Response = Config Queued() return \#Outcome

\hypertarget{random}{%
\subsection{RANDOM}\label{random}}

RANDOM returns a quasi-random number.

call CheckArgs `oWHOLE\textgreater=0 oWHOLE\textgreater=0
oWHOLE\textgreater=0'

if \#Bif\_Arg.0 = 1 then do Minimum = 0

Maximum = \#Bif Arg.1 if Maximum\textgreater100000 then call Raise
40.31, Maximum

end else do if \#Bif\_ArgExists.1 then Minimum = \#Bif Arg.1 else
Minimum = 0 if \#Bif\_ArgExists.2 then Maximum = \#Bif\_Arg.2 else
Maximum = 999

end

if Maximum-Minimum\textgreater100000 then call Raise 40.32, Minimum,
Maximum

if Maximum-Minimum\textless0 then call Raise 40.33, Minimum, Maximum

if \#Bif\_ArgExists.3 then call Config Random Seed \#Bif\_Arg.3 call
Config Random Next Minimum, Maximum return \#Outcome

\hypertarget{symbol}{%
\subsection{SYMBOL}\label{symbol}}

The function SYMBOL takes one argument, which is evaluated. Let String
be the value of that argument. If Config\_Length(String) returns an
indicator `E' then the SYNTAX condition 23.1 shall be raised. Otherwise,
if the syntactic recognition described in section nnn would not
recognize String as a symbol then the result of the function SYMBOL is
'BAD''.

If String would be recognized as a symbol the result of the function
SYMBOL depends on the outcome of accessing the value of that symbol, see
nnn. If the final use of Var\_Value leaves the indicator with value `D'
then the result of the function SYMBOL is `LIT', otherwise `VAR'.

\hypertarget{time}{%
\subsection{TIME}\label{time}}

TIME with less than two arguments returns the local time within the day,
or an elapsed time. Otherwise it converts the second argument (which has
a format given by the third argument) to the format specified by the
first argument.

call CheckArgs `oCEHLMNORS oANY oCHLMNS' /* If the third argument is
given then the second is mandatory. */ if \#Bif\_ArgExists.3 \&
\#Bif\_ArgExists.2 then

call Raise 40.19, '\,', \#Bif Arg.3

if \#Bif\_ ArgExists.1 then Option else Option

\#Bif Arg.1 tint

/* The date/time is `frozen' throughout a clause. */

if \#ClauseTime.\#Level == '\,' then do \#Response = Config Time ()
\#ClauseTime.\#Level = \#Time \#ClauseLocal.\#Level = \#Time +
\#Adjust\textless Index ``\#Adjust'' \# ``\,'' \textgreater{} end

/* If there is no second argument, the current time is returned. */ if
\#Bif ArgExists.2 then return TimeFormat (\#ClauseLocal.\#Level, Option)

/* If there is a second argument it provides the time to be converted.
*/ if pos(Option, `ERO') \textgreater{} 0 then call Raise 40.29, Option
InValue = \#Bif Arg.2

if \#Bif\_ArgExists.3 then InOption = \#Bif\_Arg.3 else InOption =

HH = 0

MM = 0

ss = 0

HourAdjust = 0

select

when InOption == `C' then do parse var InValue HH `:' . +1 MM +2 XX

if HH = 12 then

125 HH = 0

if XX == `pm' then HourAdjust = 12 end when InOption == `H' then HH =
InValue when InOption == `L' \textbar{} InOption == `N' then parse var
InValue HH `:' MM `:' SS when InOption == `M' then MM = InValue
otherwise SS = InValue

end

if datatype(HH,`W')=0 \textbar{} datatype(MM,`W')=0 \textbar{}
datatype(SS,`N')=0 then call Raise 40.19, InValue, InOption

HH = HH + HourAdjust

/* Convert to microseconds */

Micro = trunc((((HH * 60) + MM) * 60 + SS) * 1000000)

/* There is no special message for time-out-of-range; the bad-format

message is used. */

if Micro\textless0 \textbar{} Micro \textgreater{} 24\emph{3600}1000000
then call Raise 40.19, InValue, InOption

/* Reconvert to check the original. (eg for hour = 99) */

if TimeFormat (Micro,InOption) == InValue then call Raise 40.19,
InValue, InOption

return TimeFormat (Micro, Option)

end /* Conversion */

TimeFormat: procedure /* Convert from microseconds to given format.
\emph{/ /} The day will be irrelevant; actually it will be the first day
possible. \emph{/ x = Time2Date2(arg(1)) parse value x with Year Month
Day Hour Minute Second Microsecond Base Days select when arg(2) == `C'
then select when Hour\textgreater12 then return Hour-12':`right
(Minute,2,'0')`pm' when Hour=12 then return `12:'right(Minute,2,'0')`pm'
when Hour\textgreater0 then return Hour':`right (Minute,2,'0')`am' when
Hour=0 then return `12:'right(Minute,2,'0')`am' end when arg(2) == `H'
then return Hour when arg(2) == `L' then return right
(Hour,2,`0')`:'right (Minute,2,'0')`:'right(Second,2,'0'),
\textbar\textbar{} `.'right (Microsecond,6,'0') when arg(2) == `M' then
return 60}Hour+Minute when arg(2) == `N' then return right
(Hour,2,`0')`:'right (Minute,2,'0')`:'right(Second,2,'0') otherwise /*
arg(2) == `S' \emph{/ return 3600}Hour+60*Minute+Second end

Time2Date: /* These are checks on the range of the date. */ if arg(1)
\textless{} 0 then call Raise 40.19, InValue, InOption if arg(1)
\textgreater= 315537897600000000 then call Raise 40.19, InValue,
InOption return Time2Date2 (arg(1))

10.6.1 VALUE VALUE returns the value of the symbol named by the first
argument, and optionally assigns it a new

value.

`rANY oOANY oANY'

if \#Bif ArgExists.3 then ArgData `rSYM oANY oOANY'

else ArgData call CheckArgs ArgData

Subject = \#Bif Arg.1

if \#Bif\_ArgExists.3 then do /* An external pool, or the reserved pool.
\emph{/ /} The reserved pool uses a null string as its pool identifier.
*/ Pool = \#Bif\_Arg.3

if Pool == '\,' then do Subject = `.' \textbar\textbar{} translate
(Subject) /* The dot on the name is implied. \emph{/ Value =
.environment {[}Subject{]} /} Was the translate redundant? */

if \#Bif ArgExists.2 then .environment {[}Subject{]} = \#Bif\_ Arg.2
return Value end

/* Fetch the original value */

\#Response = Config Get (Pool, Subject)

\#Indicator = left (\#Response,1)

if \#Indicator == `F' then call Raise 40.36, Subject if \#Indicator ==
`P' then

call Raise 40.37, Pool Value = \#Outcome if \#Bif\_ArgExists.2 then do
/* Set the new value. \emph{/ \#Response = Config Set(Pool,Subject,\#Bif
Arg.2) if \#Indicator == `P' then call Raise 40.37, Pool if \#Indicator
== `F' then call Raise 40.36, Subject end /} Return the original value.
\emph{/ return Value end /} Not external \emph{/ Subject =
translate(Subject) /} See nnn \emph{/ Pp = pos(Subject, `.') if p = 0
\textbar{} p = length(Subject) then do /} Not compound \emph{/
\#Response = Var Value(\#Pool, Subject, `0') /} The caller, in the code
of the standard, may need to test whether the Subject was dropped.
\emph{/ \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
ArgExists.2 then \#Response = Var Set(\#Pool, Subject, `0', \#Bif\_
Arg.2) return Value end /} Compound \emph{/ Expanded = left(Subject,p-1)
/} The stem \emph{/ do forever Start = p+l1 Pp = pos(Subject,`.',Start)
if p = 0 then p = length(Subject) Item = substr(Subject,Start,p-Start)
/} Tail component symbol \emph{/ if Item=='\,' then if
pos(left(Item,1),`0123456789') = 0 then do \#Response = Var
Value(\#Pool, Item, `0') Item = \#Outcome end /} Add tail component. */
Expanded = Expanded'.'Item end \#Response = Var Value(\#Pool, Expanded,
`1') \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
ArgExists.2 then \#Response = Var Set(\#Pool, Expanded, `1', \#Bif\_
Arg.2)

return Value

\hypertarget{queued-1}{%
\subsection{QUEUED}\label{queued-1}}

QUEUED returns the number of lines remaining in the external data queue.

call CheckArgs '\,' \#Response = Config Queued() return \#Outcome

\hypertarget{random-1}{%
\subsection{RANDOM}\label{random-1}}

RANDOM returns a quasi-random number.

call CheckArgs `oWHOLE\textgreater=0 oWHOLE\textgreater=0
oWHOLE\textgreater=0'

if \#Bif\_Arg.0 = 1 then do Minimum = 0 Maximum = \#Bif\_Arg.1 if
Maximum\textgreater100000 then call Raise 40.31, Maximum

end else do if \#Bif\_ArgExists.1 then Minimum = \#Bif Arg.1 else
Minimum = 0 if \#Bif\_ArgExists.2 then Maximum = \#Bif\_Arg.2 else
Maximum = 999

end

if Maximum-Minimum\textgreater100000 then call Raise 40.32, Minimum,
Maximum

if Maximum-Minimum\textless0 then call Raise 40.33, Minimum, Maximum

if \#Bif\_ArgExists.3 then call Config Random Seed \#Bif\_Arg.3 call
Config Random Next Minimum, Maximum return \#Outcome

\hypertarget{symbol-1}{%
\subsection{SYMBOL}\label{symbol-1}}

The function SYMBOL takes one argument, which is evaluated. Let String
be the value of that argument. If Config\_Length(String) returns an
indicator `E' then the SYNTAX condition 23.1 shall be raised. Otherwise,
if the syntactic recognition described in section nnn would not
recognize String as a symbol then the result of the function SYMBOL is
'BAD''.

If String would be recognized as a symbol the result of the function
SYMBOL depends on the outcome of accessing the value of that symbol, see
nnn. If the final use of Var\_Value leaves the indicator with value `D'
then the result of the function SYMBOL is `LIT', otherwise `VAR'.

\hypertarget{time-1}{%
\subsection{TIME}\label{time-1}}

TIME with less than two arguments returns the local time within the day,
or an elapsed time. Otherwise it converts the second argument (which has
a format given by the third argument) to the format specified by the
first argument.

call CheckArgs `oCEHLMNORS oANY oCHLMNS' /* If the third argument is
given then the second is mandatory. */ if \#Bif\_ArgExists.3 \&
\#Bif\_ArgExists.2 then

call Raise 40.19, '\,', \#Bif Arg.3

if \#Bif\_ ArgExists.1 then Option else Option

\#Bif Arg.1 tint

/* The date/time is `frozen' throughout a clause. */

if \#ClauseTime.\#Level == '\,' then do \#Response = Config Time ()
\#ClauseTime.\#Level = \#Time \#ClauseLocal.\#Level = \#Time +
\#Adjust\textless Index ``\#Adjust'' \# ``\,'' \textgreater{} end

/* If there is no second argument, the current time is returned. */ if
\#Bif ArgExists.2 then return TimeFormat (\#ClauseLocal.\#Level, Option)

/* If there is a second argument it provides the time to be

converted. */ if pos(Option, `ERO') \textgreater{} 0 then

128 call Raise 40.29, Option InValue = \#Bif Arg.2

if \#Bif\_ArgExists.3 then InOption = \#Bif\_Arg.3 else InOption = `N'
HH = 0 MM = 0 ss = 0 HourAdjust = 0 select when InOption == `C' then do
parse var InValue HH `:' . +1 MM +2 XX if XX == `pm' then HourAdjust =
12 end when InOption == `H' then HH = InValue when InOption == `L'
\textbar{} InOption == `N' then parse var InValue HH `:' MM `:' SS when
InOption == `M' then MM = InValue otherwise SS = InValue end

if \datatype(HH,`W') \textbar{} \datatype(MM,`W') \textbar{}
\datatype(SS,`N') then call Raise 40.19, InValue, InOption

HH = HH + HourAdjust

/* Convert to microseconds */

Micro = trunc((((HH * 60) + MM) * 60 + SS) * 1000000)

/* Reconvert to check the original. (eg for hour = 99) */

if TimeFormat (Micro,InOption) == InValue then call Raise 40.19,
InValue, InOption

return TimeFormat (Micro, Option)

end /* Conversion */

TimeFormat: /* Convert from microseconds to given format. */ parse value
Time2Date(arg(1)) with, Year Month Day Hour Minute Second Microsecond
Base Days select when arg(2) == `C' then if Hour\textgreater12 then
return Hour-12':`right (Minute,2,'0')`pm'

else return Hour':`right (Minute,2,'0')`am' when arg(2) == `E'
\textbar{} arg(2) == `R' then do

/* Special case first time */

if \#StartTime.\#Level == '\,' then do \#StartTime.\#Level
\#ClauseTime.\#Level return `0O'

end Output = \#ClauseTime.\#Level-\#StartTime. \#Level if arg(2) == `R'
then

\#StartTime.\#Level = \#ClauseTime.\#Level return Output * 1E-6 end /* E
or R */ when arg(2) == `H' then return Hour when arg(2) == `L' then
return right (Hour,2,`0')`:'right (Minute,2,'0')`:'right(Second,2,'0'),
\textbar\textbar{} `.'right (Microsecond,6,'0')

when arg(2) == `M' then return 60\emph{Hour+Minute when arg(2) == `N'
then return right (Hour,2,`0')`:'right
(Minute,2,'0')`:'right(Second,2,'0') when arg(2) == `O' then return
trunc(\#ClauseLocal.\#Level - \#ClauseTime.\#Level) otherwise /} arg(2)
== `S' \emph{/ return 3600}Hour+60*Minute+Second end 10.6.5 VALUE

VALUE returns the value of the symbol named by the first argument, and
optionally assigns it a new value.

`rANY oOANY oANY' `rSYM OANY oANY'

if \#Bif ArgExists.3 then ArgData else ArgData call CheckArgs ArgData

129 130

Subject = \#Bif Arg.1

if \#Bif\_ArgExists.3 then do /* An external pool, or the reserved pool.
\emph{/ /} The reserved pool uses a null string as its pool identifier.
*/ Pool = \#Bif\_Arg.3

if Pool == '\,' then do Subject = `.' \textbar\textbar{} translate
(Subject) /* The dot on the name is implied. \emph{/ Value =
.environment {[}Subject{]} /} Was the translate redundant? */

if \#Bif ArgExists.2 then .environment {[}Subject{]} = \#Bif\_ Arg.2
return Value end

/* Fetch the original value */

\#Response = Config Get (Pool, Subject)

\#Indicator = left (\#Response,1)

if \#Indicator == `F' then call Raise 40.36, Subject if \#Indicator ==
`P' then

call Raise 40.37, Pool Value = \#Outcome if \#Bif\_ArgExists.2 then do
/* Set the new value. \emph{/ \#Response = Config Set(Pool,Subject,\#Bif
Arg.2) if \#Indicator == `P' then call Raise 40.37, Pool if \#Indicator
== `F' then call Raise 40.36, Subject end /} Return the original value.
\emph{/ return Value end /} Not external \emph{/ Subject =
translate(Subject) /} See nnn \emph{/ Pp = pos(Subject, `.') if p = 0
\textbar{} p = length(Subject) then do /} Not compound \emph{/
\#Response = Var Value(\#Pool, Subject, `0') /} The caller, in the code
of the standard, may need to test whether the Subject was dropped.
\emph{/ \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
ArgExists.2 then \#Response = Var Set(\#Pool, Subject, `0', \#Bif\_
Arg.2) return Value end /} Compound \emph{/ Expanded = left(Subject,p-1)
/} The stem \emph{/ do forever Start = p+l1 Pp = pos(Subject,`.',Start)
if p = 0 then p = length(Subject) Item = substr(Subject,Start,p-Start)
/} Tail component symbol \emph{/ if Item=='\,' then if
pos(left(Item,1),`0123456789') = 0 then do \#Response = Var
Value(\#Pool, Item, `0') Item = \#Outcome end /} Add tail component. */
Expanded = Expanded'.'Item end \#Response = Var Value(\#Pool, Expanded,
`1') \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
ArgExists.2 then \#Response = Var Set(\#Pool, Expanded, `1', \#Bif\_
Arg.2)

return Value

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/configuration.tex
complete, but very deficient, extended standard

\hypertarget{configuration}{%
\chapter{Configuration}\label{configuration}}

Any implementation of this standard will be functioning within a
configuration. In practice, the boundary between what is implemented
especially to support Rexx and what is provided by the system will vary
from system to system. This clause describes what they shall together do
to provide the configuration for the Rexx language processing which is
described in this standard.

We don't want to add undue ``magic'' to this section. It seems we will
need the concept of a ``reference'' (equivalent to a machine address) so
that this section can at least have composite objects as arguments. (As
it already does but these are not Rexx objects)

Possibly we could unify ``reference'' with ``variable pool number''
since object one-to-one with its variable pool is a fair model. That way
we don't need a new primitive for comparison of two references.

JAVA is only a ``reference'' for NetRexx so some generalized JAVA-like
support is needed for that. It would provide the answers to what classes
were in the context, what their method signatures were etc.

\hypertarget{notation}{%
\section{Notation}\label{notation}}

The interface to the configuration is described in terms of functions.
The notation for describing the interface functionally uses the name
given to the function, followed by any arguments. This does not
constrain how a specific implementation provides the function, nor does
it imply that the order of arguments is significant for a specific
implementation.

The names of the functions are used throughout this standard; the names
used for the arguments are used only in this clause and nnn.

The name of a function refers to its usage. A function whose name starts
with

\begin{itemize}
\item
  Config\_ is used only from the language processor when processing
  programs;
\item
  API\_is part of the application programming interface and is
  accessible from programs which are not written in the Rexx language;
\item
  Trap\_ is not provided by the language processor but may be invoked by
  the language processor. As its result, each function shall return a
  completion Response. This is a string indicating how the function
  behaved. The completion response may be the character `N' indicating
  the normal behavior occurred; otherwise the first character is an
  indicator of a different behavior and the remainder shall be suitable
  as a human-readable description of the function's behavior.
\end{itemize}

This standard defines any additional results from Config\_ functions as
made available to the language processor in variables. This does not
constrain how a particular implementation should return these results.

\hypertarget{notation-for-completion-response-and-conditions}{%
\subsection{Notation for completion response and
conditions}\label{notation-for-completion-response-and-conditions}}

As alternatives to the normal indicator `N', each function may return a
completion response with indicator 'X' or `S'; other possible indicators
are described for each function explicitly. The indicator `X' means that
the function failed because resources were exhausted. The indicator `S'
shows that the configuration was unable to perform the function. Certain
indicators cause conditions to be raised. The possible raising of these
conditions is implicit in the use of the function; it is not shown
explicitly when the functions are used in this standard. The implicit
action is call \#Raise `SYNTAX', Message, Description where: \#Raise
raises the condition, see nnn. Message is determined by the indicator in
the completion response. If the indicator is `X' then Message is 5.1. If
the indicator is `S' then Message is 48.1. Description is the
description in the completion response. The `SYNTAX' condition 5.1 can
also be raised by any other activity of the language processor.

\hypertarget{processing-initiation}{%
\section{Processing initiation}\label{processing-initiation}}

The processing initiation interface consists of a function which the
configuration shall provide to invoke the language processor. We could
do REQUIRES in a macro-expansion way by adding an argument to
Contig\_SourceChar to specify the source file. However, I'm assuming we
will prefer to recursively ``run'' each required file. One of the
results of that will be the classes and methods made public by that
REQUIRES subject.

\hypertarget{api-start}{%
\subsection{API Start}\label{api-start}}

Syntax:

API Start(How, Source, Environment, Arguments, Streams, Traps, Provides)

where: How is one of `COMMAND', `FUNCTION', or `SUBROUTINE' and
indicates how the program is invoked.

What does OOI say for How when running REQUIREd files?

Source is an identification of the source of the program to be
processed.

Environment is the initial value of the environment to be used in
processing commands. This has components for the name of the environment
and how the input and output of commands is to be directed.

Arguments is the initial argument list to be used in processing. This
has components to specify the number of arguments, which arguments are
omitted, and the values of arguments that are not omitted.

Streams has components for the default input stream to be used and the
default output streams to be used.

Traps is the list of traps to be used in processing (see nnn). This has
components to specify whether each trap is omitted or not.

Semantics:

This function starts the execution of a Rexx program.

If the program was terminated due to a RETURN or EXIT instruction
without an expression the completion response is `N'.

If the program was terminated due to a RETURN or EXIT instruction with
an expression the indicator in the completion response is `R' and the
description of the completion response is the value of the expression.

If the program was terminated due to an error the indicator in the
completion response is `E' and the description in the completion
response comprises information about the error that terminated
processing.

If How was `REQUIRED' and the completion response was not `E', the
Provides argument is set to reference classes made available. See nnn
for the semantics of these classes.

\hypertarget{source-programs-and-character-sets}{%
\section{Source programs and character
sets}\label{source-programs-and-character-sets}}

The configuration shall provide the ability to access source programs
(see nnn). Source programs consist of characters belonging to the
following categories:

\begin{itemize}
\item
  syntactic\_characters;
\item
  extra\_letters;
\item
  other\_blank\_characters;
\item
  other\_negators;
\item
  other\_characters.
\end{itemize}

A character shall belong to only one category.

\hypertarget{syntactic_characters}{%
\subsection{Syntactic\_characters}\label{syntactic_characters}}

The following characters represent the category of characters called
syntactic\_characters, identified by their names. The glyphs used to
represent them in this document are also shown. Syntactic\_characters
shall be available in every configuration:

\begin{itemize}
\item
  \& ampersand;
\item
  apostrophe, single quotation mark, single quote;
\item
  asterisk, star;
\item
  blank, space;
\item
  A-Z capital letters A through Z;
\item
  colon;
\item
  , comma;
\item
  0-9 digits zero through nine;
\item
  = equal sign;
\item
  exclamation point, exclamation mark;
\item
  \begin{quote}
  greater-than sign;
  \end{quote}
\end{itemize}

hyphen, minus sign;

\textless{} less-than sign;

\begin{itemize}
\item
  {[} left bracket, left square bracket; (\_ left parenthesis;
\item
  \% percent sign;
\item
  . period, decimal point, full stop, dot;
\item
  plus sign;
\item
  ? question mark;
\item
  '' quotation mark, double quote; ~reverse slant, reverse solidus,
  backslash; {]} right bracket, right square bracket;
\item
  ) right parenthesis; ; semicolon; /\_ slant, solidus, slash; a-z small
  letters a through z;
\item
  \textasciitilde{} tilde, twiddle;
\item
  \_ underline, low line, underscore;
\item
  vertical line, bar, vertical bar.
\end{itemize}

\hypertarget{extra_letters}{%
\subsection{Extra\_letters}\label{extra_letters}}

A configuration may have a category of characters in source programs
called extra\_letters. Extra\_letters are determined by the
configuration.

\hypertarget{other_blank_characters}{%
\subsection{Other\_blank\_characters}\label{other_blank_characters}}

A configuration may have a category of characters in source programs
called other\_blank\_characters. Other\_blank\_characters are determined
by the configuration. Only the following characters represent possible
characters of this category:

\begin{itemize}
\item
  carriage return;
\item
  form feed;
\item
  horizontal tabulation;
\item
  new line;
\item
  vertical tabulation.
\end{itemize}

\hypertarget{other_negators}{%
\subsection{Other\_negators}\label{other_negators}}

A configuration may have a category of characters in source programs
called other\_negators. Other\_negators are determined by the
configuration. Only the following characters represent possible
characters of this category. The glyphs used to represent them in this
document are also shown:

\begin{itemize}
\item
  \begin{itemize}
  \tightlist
  \item
    circumflex accent, caret;
  \end{itemize}
\item
  --- not sign.
\end{itemize}

\hypertarget{other_characters}{%
\subsection{Other\_characters}\label{other_characters}}

A configuration may have a category of characters in source programs
called other\_characters. Other\_characters are determined by the
configuration.

\hypertarget{configuration-characters-and-encoding}{%
\section{Configuration characters and
encoding}\label{configuration-characters-and-encoding}}

The configuration characters and encoding interface consists of
functions which the configuration shall provide which are concerned with
the encoding of characters. The following functions shall be provided:

\begin{itemize}
\item
  Config\_SourceChar;
\item
  Config\_OtherBlankCharacters;
\item
  Config\_Upper;
\item
  Config\_Compare;
\item
  Config\_B2C;
\item
  Config\_C2B;
\item
  Config\_Substr;
\item
  Config\_Length;
\item
  Config\_Xrange.
\end{itemize}

\hypertarget{config_sourcechar}{%
\subsection{Config\_SourceChar}\label{config_sourcechar}}

Syntax:

Config SourceChar ()

Semantics: Supply the characters of the source program in sequence,
together with the EOL and EOS events. The EOL event represents the end
of a line. The EOS event represents the end of the source program. The
EOS event must only occur immediately after an EOL event. Either a
character or an event is supplied on each invocation, by setting
\#Outcome.

If this function is unable to supply a character because the source
program encoding is incorrect the indicator of the completion response
is `E' and the description of the completion information is the encoding
which is incorrect, in hexadecimal notation.

\hypertarget{config_otherblankcharacters}{%
\subsection{Config\_OtherBlankCharacters}\label{config_otherblankcharacters}}

Syntax: Config OtherBlankCharacters ()

Semantics: Get other\_blank\_characters (see nnn). Set \#Outcome to a
string of zero or more distinct characters in arbitrary order. Each
character is one that the configuration considers equivalent to the
character Blank for the purposes of parsing.

\hypertarget{config_upper}{%
\subsection{Config\_Upper}\label{config_upper}}

Syntax: Config Upper (Character)

where: Character is the character to be translated to uppercase.
Semantics: Translate Character to uppercase. Set \#Outcome to the
translated character. Characters which have been subject to this
translation are referred to as being in uppercase. Config\_Upper applied
to a character in uppercase must not change the character.

\hypertarget{config_lower}{%
\subsection{Config\_Lower}\label{config_lower}}

Syntax:

Config Lower (Character) where: Character is the character to be
translated to lowercase. Semantics: Translate Character to lowercase.
Set \#Outcome to the translated character. Characters which have been
subject to this translation are referred to as being in lowercase.
Config\_Lower applied to a character in lowercase must not change the
character. Config\_Upper of the outcome of Config\_Lower(Character)
shall be the original character.

\hypertarget{config_compare}{%
\subsection{Config\_Compare}\label{config_compare}}

Syntax: Config Compare(Characterl, Character2)

where:

Character1 is the character to be compared with Character2. Character2
is the character to be compared with Character1.

Semantics:

Compare two characters. Set \#Outcome to

\begin{itemize}
\item
  `equal' if Character1 is equal to Character2;
\item
  `greater' if Character1 is greater than Character2;
\item
  `lesser' if Character' is less than Character2. The function shall
  exhibit the following characteristics. If Config \_Compare(a,b)
  produces
\item
  `equal' then Config\_Compare(b,a) produces `equal';
\item
  `greater' then Config\_Compare(b,a) produces `lesser';
\item
  `lesser' then Config\_Compare(b,a) produces `greater';
\item
  `equal' and Config\_Compare(b,c) produces `equal' then
  Config\_Compare(a,c) produces `equal';
\item
  `greater' and Config\_Compare(b,c) produces `greater' then Config
  \_Compare(a,c) produces `greater';
\item
  `lesser' and Config \_Compare(b,c) produces `lesser' then
  Config\_Compare(a,c) produces `lesser';
\item
  `equal' then Config\_Compare(a,c) and Config\_Compare(b,c) produce the
  same value. Syntactic characters which are different characters shall
  not compare equal by Config\_Compare, see nnn.
\end{itemize}

\hypertarget{config-_b2c}{%
\subsection{Config \_B2C}\label{config-_b2c}}

Syntax: Config B2C (Binary) where: Binary is a sequence of digits, each
`0' or `1'. The number of digits shall be a multiple of eight.
Semantics:

Translate Binary to a coded string. Set \#Outcome to the resulting
string. The string may, or may not, correspond to a sequence of
characters.

\hypertarget{config_c2b}{%
\subsection{Config\_C2B}\label{config_c2b}}

Syntax: Config C2B (String)

where: String is a string. Semantics: Translate String to a sequence of
digits, each `0' or `1'. Set \#Outcome to the result. This function is
the inverse of Config\_B2C.

\hypertarget{config_substr}{%
\subsection{Config\_Substr}\label{config_substr}}

Syntax: Config Substr(String, n)

where: String is a string. nis an integer identifying a position within
String. Semantics: Copy the n-th character from String. The leftmost
character is the first character. Set Outcome to the resulting
character. If this function is unable to supply a character because
there is no n-th character in String the indicator of the completion
response is `M'. If this function is unable to supply a character
because the encoding of String is incorrect the indicator of the
completion response is `E' and the description of the completion
information is the encoding which is incorrect, in hexadecimal notation.

\hypertarget{config-length}{%
\subsection{Config Length}\label{config-length}}

Syntax: Config Length (String)

where: String is a string. Semantics: Set \#Outcome to the length of the
string, that is, the number of characters in the string. If this
function is unable to determine a length because the encoding of String
is incorrect, the indicator of the completion response is `E' and the
description of the completion information is the encoding which is
incorrect, in hexadecimal notation.

\hypertarget{config_xrange}{%
\subsection{Config\_Xrange}\label{config_xrange}}

Syntax:

Config Xrange(Characterl, Character2) where: Character1 is the null
string, or a single character. Character2 is the null string, or a
single character. Semantics: If Character1 is the null string then let
LowBound be a lowest ranked character in the character set according to
the ranking order provided by Config\_Compare; otherwise let LowBound be
Character1. If Character2 is the null string then let HighBound be a
highest ranked character in the character set according to the ranking
order provided by Config\_Compare; otherwise let HighBound be
Character2lf \#Outcome after Config\_Compare(LowBound,HighBound) has a
value of

\begin{itemize}
\item
  `equal' then \#Outcome is set to LowBound;
\item
  `lesser' then \#Outcome is set to the sequence of characters between
  LowBound and HighBound inclusively, in ranking order;
\item
  `greater' then \#Outcome is set to the sequence of characters
  HighBound and larger, in ranking order, followed by the sequence of
  characters LowBound and smaller, in ranking order.
\end{itemize}

\hypertarget{objects}{%
\section{Objects}\label{objects}}

The objects interface consists of functions which the configuration
shall provide for creating objects.

\hypertarget{config_objectnew}{%
\subsection{Config\_ObjectNew}\label{config_objectnew}}

Syntax: Config ObjectNew

Semantics:

Set \#Outcome to be a reference to an object. The object shall be
suitable for use as a variable pool, see nnn. This function shall never
return a value in \#Outcome which compares equal with the value returned
on another invokation of the function.

\hypertarget{config_array_size}{%
\subsection{Config\_Array\_Size}\label{config_array_size}}

Syntax: Config Array Size(Object, size) where: Object is an object. Size
is an integer greater or equal to 0. Semantics: The configuration should
prepare to deal efficiently with the object as an array with indexes
having values up to the value of size.

\hypertarget{config_array_put}{%
\subsection{Config\_Array\_Put}\label{config_array_put}}

Syntax: Config Array Put(Array, Item, Index)

where: Array is an array. Item is an object Index is an integer greater
or equal to 1. Semantics: The configuration shall record that the array
has Item associated with Index.

\hypertarget{config_array_at}{%
\subsection{Config\_Array\_At}\label{config_array_at}}

Syntax: Config Array At(Array, Index)

where: Array is an array. Index is an integer greater or equal to 1.
Semantics: The configuration shall return the item that the array has
associated with Index.

\hypertarget{config_array_hasindex}{%
\subsection{Config\_Array\_Hasindex}\label{config_array_hasindex}}

Syntax: Config Array At(Array, Index) where: Array is an array. Index is
an integer greater or equal to 1. Semantics: Return `1' if there is an
item in Array associated with Index, `0' otherwise.

\hypertarget{config_array_remove}{%
\subsection{Config\_Array\_Remove}\label{config_array_remove}}

Syntax: Config Array At(Array, Index)

where: Array is an array. Index is an integer greater or equal to 1.
Semantics: After this operation, no item is associated with the Index in
the Array.

\hypertarget{commands}{%
\section{Commands}\label{commands}}

The commands interface consists of a function which the configuration
shall provide for strings to be passed as commands to an environment.

See nnn and nnn for a description of language features that use
commands.

\hypertarget{config-_command}{%
\subsection{Config \_Command}\label{config-_command}}

Syntax:

Config Command(Environment, Command) where: Environment is the
environment to be addressed. It has components for:

\begin{itemize}
\item
  the name of the environment;
\item
  the name of a stream from which the command will read its input. The
  null string indicates use of the default input stream;
\item
  the name of a stream onto which the command will write its output. The
  null string indicates use of the default output stream. There is an
  indication of whether writing is to APPEND or REPLACE;
\item
  the name of a stream onto which the command will write its error
  output. The null string indicates use of the default error output
  stream. There is an indication of whether writing is to APPEND or
  REPLACE. Command is the command to be executed. Semantics: Perform a
  command.
\item
  set the indicator to `E' or `F' if the command ended with an ERROR
  condition, or a FAILURE condition, respectively;
\item
  set \#RC to the return code string of the command.
\end{itemize}

\hypertarget{external-routines}{%
\section{External routines}\label{external-routines}}

The external routines interface consists of a function which the
configuration shall provide to invoke external routines. See nnn and nnn
for a description of the language features that use external routines.

\hypertarget{config_externalroutine}{%
\subsection{Config\_ExternalRoutine}\label{config_externalroutine}}

Syntax: Config ExternalRoutine(How, NameType, Name, Environment,
Arguments, Streams, Traps) where: How is one of `FUNCTION' or
`SUBROUTINE' and indicates how the external routine is to be invoked.

NameType is a specification of whether the name was provided as a symbol
or as a string literal. Name is the name of the routine to be invoked.

Environment is an environment value with the same components as on API\_
Start.

Arguments is a specification of the arguments to the routine, with the
same components as on API Start.

Streams is a specification of the default streams, with the same
components as on API\_ Start.

Traps is the list of traps to be used in processing, with the same
components as on API\_ Start.

Semantics:

Invoke an external routine. Set \{Outcome to the result of the external
routine, or set the indicator of the completion response to `D' if the
external routine did not provide a result.

If this function is unable to locate the routine the indicator of the
completion response is `U'. As a result SYNTAX condition 43.1 is raised
implicitly.

If How indicated that a result from the routine was required but the
routine did not provide one the indicator of the completion response is
`H'. As a result SYNTAX condition 44.1 is raised implicitly.

If How indicated that a result from the routine was required but the
routine provided one that was too long (see \#Limit\_String in nnn) the
indicator of the completion response is `L'. As a result SYNTAX
condition 52 is raised implicitly.

If the routine failed in a way not indicated by some other indicator the
indicator of the completion response is `F'. As a result SYNTAX
condition 40.1 is raised implicitly.

\hypertarget{config_externalmethod}{%
\subsection{Config\_ExternalMethod}\label{config_externalmethod}}

OOI has external classes explicitly via the ::CLASS abc EXTERNAL
mechanism. Analogy with classic would also allow the subject of
::REQUIRES to be coded in non-Rexx. However ::REQUIRES subject is coded,
we need to gather in knowledge of its method names because of the search
algorithm that determines which method is called. Hence reasonable that
the ultimate external call is to a method. Perhaps combine
Config\_ExternalRoutine with Config\_ExternalMethod. There is a
terminology clash on ``environment''. Perhaps easiest to change the
classic to ``address\_environment''. (And make it part of new
``environment''?) There are terminology decisions to make about
``files'', ``programs'', and ``packages''. Possibly ``program'' is the
thing you run (and we don't say what it means physically), ``file'' is a
unit of scope (ROUTINEs in current file before those in REQUIREd), and
``package'' we don't use (since a software package from a shop would
probably have several files but not everything to run a program.) Using
``file'' this way may not be too bad since we used ``stream'' rather
than ``tile'' in the classic definition. The How parameter will need
`METHOD' as a value. Should API\_Start also allow 'METHOD''. If we pass
the new Environment we don't have to pass Streams separately.

Text of Config\_ExternalMethod waiting on such decisions. Syntax:

Config ExternalMethod (How, NameType, Name, Environment, Arguments,
Streams, Traps) where: How is one of `FUNCTION' or `SUBROUTINE' and
indicates how the external routine is to be invoked. NameType is a
specification of whether the name was provided as a symbol or as a
string literal. Name is the name of the routine to be invoked.
Environment is an environment value with the same components as on API\_
Start. Arguments is a specification of the arguments to the routine,
with the same components as on API\_Start. Streams is a specification of
the default streams, with the same components as on API\_ Start. Traps
is the list of traps to be used in processing, with the same components
as on API\_ Start. Semantics: Invoke an external routine. Set \{Outcome
to the result of the external routine, or set the indicator of the
completion response to `D' if the external routine did not provide a
result. If this function is unable to locate the routine the indicator
of the completion response is `U'. As a result SYNTAX condition 43.1 is
raised implicitly. If How indicated that a result from the routine was
required but the routine did not provide one the indicator of the
completion response is `H'. As a result SYNTAX condition 44.1 is raised
implicitly. If How indicated that a result from the routine was required
but the routine provided one that was too long (see \#Limit\_String in
nnn) the indicator of the completion response is `L'. As a result SYNTAX
condition 52 is raised implicitly. If the routine failed in a way not
indicated by some other indicator the indicator of the completion
response is `F'. As a result SYNTAX condition 40.1 is raised implicitly.
5.8 External data queue The external data queue interface consists of
functions which the configuration shall provide to manipulate an
external data queue mechanism. See nnn, nnn, nnn, nnn, and nnn for a
description of language features that use the external data queue. The
configuration shall provide an external data queue mechanism. The
following functions shall be provided:

\begin{itemize}
\item
  Config\_Push;
\item
  Config \_Queue;
\item
  Config\_ Pull;
\item
  Config\_Queued.
\end{itemize}

The configuration may permit the external data queue to be altered in
other ways. In the absence of such alterations the external data queue
shall be an ordered list. Config\_Push adds the specified string to one
end of the list, Config \_Queue to the other. Config\_Pull removes a
string from the end that Config\_Push adds to unless the list is empty.

\hypertarget{config-push}{%
\subsection{Config Push}\label{config-push}}

Syntax: Config Push(String)

where: String is the value to be retained in the external data queue.
Semantics: Add String as an item to the end of the external data queue
from which Config\_Pull will remove an item.

\hypertarget{contig_queue}{%
\subsection{Contig\_Queue}\label{contig_queue}}

Syntax: Config Queue (String)

where: String is the value to be retained in the external data queue.
Semantics: Add String as an item to the opposite end of the external
data queue from which Config\_Pull will remove an item.

\hypertarget{config_pull}{%
\subsection{Config\_Pull}\label{config_pull}}

Syntax: Config Pull()

Semantics: Retrieve an item from the end of the external data queue to
which Config\_Push adds an element to the list. Set \#Outcome to the
value of the retrieved item. If no item could be retrieved the indicator
of the completion response is `F'.

\hypertarget{contig_queued}{%
\subsection{Contig\_Queued}\label{contig_queued}}

Syntax: Config Queued ()

Semantics:

Get the count of items in the external data queue. Set \#Outcome to that
number.

\hypertarget{streams}{%
\section{Streams}\label{streams}}

The streams interface consists of functions which the configuration
shall provide to manipulate streams. See nnn, nnn, and nnn for a
description of language features which use streams. Streams are
identified by names and provide for the reading and writing of data.
They shall support the concepts of characters, lines, positioning,
default input stream and default output stream. The concept of a
persistent stream shall be supported and the concept of a transient
stream may be supported. A persistent stream is one where the content is
not expected to change except when the stream is explicitly acted on. A
transient stream is one where the data available is expected to vary
with time.

The concepts of binary and character streams shall be supported. The
content of a character stream is expected to be characters. The null
string is used as a name for both the default input stream and the
default output stream. The null string names the default output stream
only when it is an argument to the Config\_Stream\_Charout operation.

The following functions shall be provided:

\begin{itemize}
\item
  Config\_Stream\_Charin;
\item
  Config\_Stream\_Position;
\item
  Config\_Stream\_Command;
\item
  Config\_Stream\_State;
\item
  Config\_Stream\_Charout;
\item
  Config\_Stream\_Qualified;
\item
  Config\_Stream\_Unique;
\item
  Config\_Stream\_Query;
\item
  Config\_Stream\_Close;
\item
  Config\_Stream\_Count. The results of these functions are described in
  terms of the following stems with tails which are stream names:
\item
  \#Charin\_Position.Stream;
\item
  \#Charout\_Position.Stream;
\item
  \#Linein\_Position.Stream;
\item
  \#Lineout\_Position.Stream.
\end{itemize}

\hypertarget{config-_stream_charin}{%
\subsection{Config \_Stream\_Charin}\label{config-_stream_charin}}

Syntax:

Config Stream Charin(Stream, OperationType) where: Stream is the name of
the stream to be processed. OperationType is one of `CHARIN', `LINEIN',
or `NULL. Semantics: Read from a stream. Increase
\#Linein\_Position.Stream by one when the end-of-line indication is
encountered. Increase \#Charin\_Position.Stream when the indicator will
be 'N'. If OperationType is `CHARIN' the state variables describing the
stream will be affected as follows: - when the configuration is able to
provide data from a transient stream or the character at position
\#Charin\_Position.Stream of a persistent stream then \#Outcome shall be
set to contain the data. The indicator of the response shall be `N';

\begin{itemize}
\tightlist
\item
  when the configuration is unable to return data because the read
  position is at the end of a persistent stream then the indicator of
  the response shall be `O';
\item
  when the configuration is unable to return data from a transient
  stream because no data is available and no data is expected to become
  available then the indicator of the response shall be `O':
\item
  otherwise the configuration is unable to return data and does not
  expect to be able to return data by waiting; the indicator of the
  response shall be `E'. The data set in \#Outcome will either be a
  single character or will be a sequence of eight characters, each `0'
  or `1'. The choice is decided by the configuration. The eight
  character sequence indicates a binary stream, see nnn. If
  OperationType is `LINEIN' then the action is the same as if Operation
  had been `CHARIN' with the following additional possibility. If
  end-of-line is detected any character (or character sequence) which is
  an embedded indication of the end-of-line is skipped. The characters
  skipped contribute to the change of \#Charin\_Position.Stream.
  \#Outcome is the null string. If OperationType is `NULL' then the
  stream is accessed but no data is read.
\end{itemize}

\hypertarget{config_stream_position}{%
\subsection{Config\_Stream\_Position}\label{config_stream_position}}

Syntax:

Config Stream Position(Stream, OperationType, Position) where: Stream is
the name of the stream to be processed. Operation is `CHARIN', `LINEIN',
`CHAROUT', or `LINEOUT'. Position indicates where to position the
stream. Semantics: If the operation is `CHARIN' or `CHAROUT' then
Position is a character position, otherwise Position is a line position.
If Operation is `CHARIN' or `LINEIN' and the Position is beyond the
limit of the existing data then the indicator of the completion response
shall be `R'. Otherwise if Operation is `CHARIN' or `LINEIN' set
\#Charin\_Position.Stream to the position from which the next
Config\_Stream\_Charin on the stream shall read, as indicated by
Position. Set \#Linein\_Position.Stream to correspond with this
position. If Operation is `CHAROUT' or `LINEOQUT' and the Position is
more than one beyond the limit of existing data then the indicator of
the response shall be `R'. Otherwise if Operation is `CHAROUT' or
`LINEOUT' then \#Charout\_Position.Stream is set to the position at
which the next Config\_Stream\_Charout on the stream shall write, as
indicated by Position. Set \#Lineout\_Position.Stream to correspond with
this position. If this function is unable to position the stream because
the stream is transient then the indicator of the completion response
shall be `T'.

\hypertarget{config-_stream_command}{%
\subsection{Config \_Stream\_Command}\label{config-_stream_command}}

Syntax:

Config Stream Command (Stream, Command) where: Stream is the name of the
stream to be processed. Command is a configuration-specific command to
be performed against the stream. Semantics: Issue a
configuration-specific command against a stream. This may affect all
state variables describing Stream which hold position information. It
may alter the effect of any subsequent operation on the specified
stream. If the indicator is set to `N', \#Outcome shall be set to
information from the command.

\hypertarget{config_stream_state}{%
\subsection{Config\_Stream\_State}\label{config_stream_state}}

Syntax:

Config Stream State (Stream) where: Stream is the name of the stream to
be queried. Semantics: Set the indicator to reflect the state of the
stream. Return an indicator equal to the indicator that an immediately
subsequent Config\_Stream\_Charin(Stream, `CHARIN') would return.
Alternatively, return an indicator of `U'.

The remainder of the response shall be a configuration-dependent
description of the state of the stream.

\hypertarget{config_stream_charout}{%
\subsection{Config\_Stream\_Charout}\label{config_stream_charout}}

Syntax:

Config Stream Charout (Stream, Data) where: Stream is the name of the
stream to be processed. Data is the data to be written, or `EOL' to
indicate that an end-of-line indication is to be written, or a null
string. In the first case, if the stream is a binary stream then Data
will be eight characters, each `0' or `1', otherwise Data will be a
single character. Semantics: When Data is the null string, no data is
written. Otherwise write to the stream. The state variables describing
the stream will be affected as follows:

\begin{itemize}
\item
  when the configuration is able to write Data to a transient stream or
  at position \#Charout\_Position.Stream of a persistent stream then the
  indicator in the response shall be `N'. When Data is not `EOL' then
  \#Charout\_Position.Stream is increased by one. When Data is `EOL',
  then \#Lineout\_Position.Stream is increased by one and
  \#Charout\_Position.Stream is increased as necessary to account for
  any end-of-line indication embedded in the stream;
\item
  when the configuration is unable to write Data the indicator is set to
  `E'.
\end{itemize}

\hypertarget{config_stream_qualified}{%
\subsection{Config\_Stream\_Qualified}\label{config_stream_qualified}}

Syntax:

Config Stream Qualified (Stream) where: Stream is the name of the stream
to be processed. Semantics: Set \#Outcome to some name which identifies
Stream. Return a completion response with indicator `B' if the argument
is not acceptable to the configuration as identifying a stream.

\hypertarget{config_stream_unique}{%
\subsection{Config\_Stream\_Unique}\label{config_stream_unique}}

Syntax:

Config Stream Unique () Semantics: Set \#Outcome to a name that the
configuration recognizes as a stream name. The name shall not be a name
that the configuration associates with any existing data.

\hypertarget{config_stream_query}{%
\subsection{Config\_Stream\_Query}\label{config_stream_query}}

Syntax: Config Stream Query (Stream) where: Stream is the name of the
stream to be queried. Semantics: Set \#Outcome to `B' if the stream is a
binary stream, or to `C' if it is a character stream.

\hypertarget{config_stream_close}{%
\subsection{Config\_Stream\_Close}\label{config_stream_close}}

Syntax:

Config Stream Close (Stream) where: Stream is the name of the stream to
be closed. Semantics: \#Charout\_Position.Stream and
\#Lineout\_Position.Stream are set to 1 unless the stream has existing
data, in which case they are set ready to write immediately after the
existing data. If this function is unable to position the stream because
the stream is transient then the indicator of the completion response
shall be `T'.

\hypertarget{config-_stream_count}{%
\subsection{Config \_Stream\_Count}\label{config-_stream_count}}

Syntax: Config Stream Count (Stream, Operation, Option) where: Stream is
the name of the stream to be counted. Operation is `CHARS', or `LINES'.

Option is `N' or `C'. Semantics: If the option is `N', \#Outcome is set
to zero if:

\begin{itemize}
\item
  the file is transient and no more characters (or no more lines if the
  Operation is `LINES') are expected to be available, even after
  waiting;
\item
  the file is persistent and no more characters (or no more lines if the
  Operation is `LINES') can be obtained from this stream by
  Config\_Stream\_Charin before use of some function which resets
  \#Charin\_Position.Stream and \#Linein\_Position.Stream.
\end{itemize}

If the option is `N' and \#Outcome is set nonzero, \#Outcome shall be 1,
or be the number of characters (or the number of lines if Operation is
`LINES') which could be read from the stream before resetting.

If the option is `C', \#Outcome is set to zero if:

\begin{itemize}
\item
  the file is transient and no characters (or no lines if the Operation
  is `LINES') are available without waiting;
\item
  the file is persistent and no more characters (or no more lines if the
  Operation is `LINES') can be obtained from this stream by
  Config\_Stream\_Charin before use of some function which resets
  \#Charin\_Position.Stream and \#Linein\_Position.Stream. If the option
  is `C' and \#Outcome is set nonzero, \#Outcome shall be the number of
  characters (or the number of lines if the Operation is `LINES') which
  can be read from the stream without delay and before resetting.
\end{itemize}

\hypertarget{external-variable-pools}{%
\section{External variable pools}\label{external-variable-pools}}

The external variable pools interface consists of functions which the
configuration shall provide to manipulate variables in external variable
pools. See nnn for the VALUE built-in function which uses external
variable pools. The configuration shall provide an external variable
pools mechanism. The following functions shall be provided:

\begin{itemize}
\item
  Config\_Get;
\item
  Config\_Set.
\end{itemize}

The configuration may permit the external variable pools to be altered
in other ways.

\hypertarget{config-get}{%
\subsection{Config Get}\label{config-get}}

Syntax: Config Get (Poolid, Name) where: Poolid is an identification of
the external variable pool. Name is the name of a variable. Semantics:
Get the value of a variable with name Name in the external variable pool
Poolid. Set Outcome to this value. If Poolid does not identify an
external pool provided by this configuration, the indicator of the
completion response is `P'. If Name is not a valid name of a variable in
the external pool, the indicator of the completion response is `F'.

\hypertarget{config-set}{%
\subsection{Config Set}\label{config-set}}

Syntax: Config Set (Poolid, Name, Value)

where: Poolid is an identification of the external variable pool. Name
is the name of a variable. Value is the value to be assigned to the
variable. Semantics: Set a variable with name Name in the external
variable pool Poolid to Value. If Poolid does not identify an external
pool provided by this configuration, the indicator of the completion
response is `P'. If Name is not a valid name of a variable in the
external pool, the indicator of the completion response is `F'.

\hypertarget{configuration-characteristics}{%
\section{Configuration
characteristics}\label{configuration-characteristics}}

The configuration characteristics interface consists of a function which
the configuration shall provide which indicates choices decided by the
configuration.

\hypertarget{config_constants}{%
\subsection{Config\_Constants}\label{config_constants}}

Syntax:

Config Constants () Semantics: Set the values of the following state
variables:

\begin{itemize}
\item
  if there are any built-in functions which do not operate at NUMERIC
  DIGITS 9, then set variables \#Bif\_Digits. (with various tails which
  are the names of those built-in functions) to the values to be used;
\item
  set variables \#Limit\_Digits, \#Limit\_EnvironmentName,
  \#Limit\_ExponentDigits, \#Limit\_Literal, \#Limit\_Messagelnsert,
  \#Limit\_Name, \#Limit\_String, \#Limit\_TraceData to the relevant
  limits. A configuration shall allow a \#Limit\_Messagelnsert value of
  50 to be specified. A configuration shall allow a \#Limit\_TraceData
  value of 250 to be specified;
\item
  set \#Configuration to a string identifying the configuration;
\item
  set \#Version to a string identifying the language processor. It shall
  have five words. Successive words shall be separated by a blank
  character. The first four letters of the first word shall be `REXX'.
  The second word shall be the four characters `5.00'. The last three
  words comprise a date. This shall be in the format which is the
  default for the DATE() built-in function.
\item
  set .nil to a value which compares unequal with any other value that
  can occur in execution.
\item
  set .local .kernel .system?
\end{itemize}

\hypertarget{configuration-routines}{%
\section{Configuration routines}\label{configuration-routines}}

The configuration routines interface consists of functions which the
configuration shall provide which provide functions for a language
processor. The following functions shall be provided:

\begin{itemize}
\item
  Config\_Trace\_Query;
\item
  Config\_Trace\_Input;
\item
  Config\_Trace\_Output;
\item
  Config\_Default\_Input;
\item
  Config\_Default\_Output;
\item
  Config\_Initialization;
\item
  Config\_Termination;
\item
  Config\_Halt\_Query;
\item
  Config\_Halt\_Reset;
\item
  Config\_NoSource;
\item
  Config\_Time;
\item
  Config\_Random\_Seed;
\item
  Config\_Random\_Next.
\end{itemize}

\hypertarget{config_trace_query}{%
\subsection{Config\_Trace\_Query}\label{config_trace_query}}

Syntax: Config Trace Query ()

Semantics: Indicate whether external activity is requesting interactive
tracing. Set \#Outcome to `Yes' if interactive tracing is currently
requested. Otherwise set \#Outcome to `No'.

\hypertarget{config_trace_input}{%
\subsection{Config\_Trace\_Input}\label{config_trace_input}}

Syntax: Config Trace Input ()

Semantics: Set \#Outcome to a value from the source of trace input. The
source of trace input is determined by the configuration.

\hypertarget{config_trace_output}{%
\subsection{Config\_Trace\_Output}\label{config_trace_output}}

Syntax: Config Trace Output (Line)

where: Line is a string. Semantics:

Write String as a line to the destination of trace output. The
destination of trace output is defined by the configuration.

\hypertarget{config-_default_input}{%
\subsection{Config \_Default\_Input}\label{config-_default_input}}

Syntax: Config Default Input ()

Semantics: Set \#Outcome to the value that LINEIN( ) would return.

\hypertarget{config_default_output}{%
\subsection{Config\_Default\_Output}\label{config_default_output}}

Syntax: Config Default Output (Line)

where: Line is a string. Semantics: Write the string as a line in the
manner of LINEOUT( ,Line).

\hypertarget{config_initialization}{%
\subsection{Config\_Initialization}\label{config_initialization}}

Syntax:

Config Initialization ()

Semantics: This function is provided only as a counterpart to
Trap\_Initialization; in itself it does nothing except return the
response. An indicator of `F' gives rise to Msg3.1.

\hypertarget{config_termination}{%
\subsection{Config\_Termination}\label{config_termination}}

Syntax:

Config Termination ()

Semantics: This function is provided only as a counterpart to
Trap\_Termination; in itself it does nothing except return the response.
An indicator of `F' gives rise to Msg2.1.

\hypertarget{config_halt_query}{%
\subsection{Config\_Halt\_Query}\label{config_halt_query}}

Syntax: Config Halt Query ()

Semantics: Indicate whether external activity has requested a HALT
condition to be raised. Set \#Outcome to `Yes if HALT is requested.
Otherwise set \#Outcome to 'No'.

5.12.9 Config \_Halt\_Reset

Syntax: Config Halt Reset ()

Semantics:

Reset the configuration so that further attempts to cause a HALT
condition will be recognized.

\hypertarget{config-_nosource}{%
\subsection{Config \_NoSource}\label{config-_nosource}}

Syntax:

Config NoSource ()

Semantics: Indicate whether the source of the program may or may not be
output by the language processor. Set \#NoSource to `1' to indicate that
the source of the program may not be output by the language processor,
at various points in processing where it would otherwise be output.
Otherwise, set \#NoSource to `0'. A configuration shall allow any
program to be processed in such a way that Config\_NoSource() sets
\#NoSource to `0'. A configuration may allow any program to be processed
in such a way that Config\_NoSource() sets \#NoSource to `1'.

\hypertarget{config_time}{%
\subsection{Config\_Time}\label{config_time}}

Syntax:

Config Time ()

Semantics: Get a time stamp. Set \#Time to a string whose value is the
integer number of microseconds that have elapsed between 00:00:00 on
January first 0001 and the time that Config\_Time is called, at
longitude zero. Values sufficient to allow for any date in the year 9999
shall be supported. The value returned may be an approximation but shall
not be smaller than the value returned by a previous use of the
function.

Set \#Adjust\textless Index ``\#Adjust'' \#``\,'' \textgreater{} to an
integer number of microseconds. \#Adjust\textless Index ``\#Adjust'' \#
™ \textgreater{} reflects the difference between the local date/time and
the date/time corresponding to \#Time. \#Time + \#Adjust\textless Index
``\#Adjust'' \# ``\,'' \textgreater{} is the local date/time.

\hypertarget{config_random_seed}{%
\subsection{Config\_Random\_Seed}\label{config_random_seed}}

Syntax: Config Random Seed (Seed)

where: Seed is a sequence of up to \#Bif\_Digits. RANDOM digits.
Semantics: Set a seed, so that subsequent uses of Config\_Random\_Next
will reproducibly return quasi-random numbers.

\hypertarget{config_random_next}{%
\subsection{Config\_Random\_Next}\label{config_random_next}}

Syntax:

Config Random Next (Min, Max) where: Min is the lower bound, inclusive,
on the number returned in \#Outcome. Max is the upper bound, inclusive,
on the number returned in \#Outcome. Semantics: Set \#Outcome to a
quasi-random nonnegative integer in the range Min to Max.

\hypertarget{config_options}{%
\subsection{Config\_Options}\label{config_options}}

Syntax: Config Options (String) where: String is a string. Semantics: No
effect beyond the effects common to all Config\_ invocations. The value
of the string will have come from an OPTIONS instruction, see nnn.

\hypertarget{traps}{%
\section{Traps}\label{traps}}

The trapping interface consists of functions which may be provided by
the caller of API\_Start (see nnn) as a list of traps. Each trap may be
specified or omitted. The language processor shall invoke a specified
trap before, or instead of, using the corresponding feature of the
language processor itself. This correspondence is implied by the choice
of names; that is, a name beginning Trap\_ will correspond to a name
beginning Config\_ when the remainder of the name is the same.
Corresponding functions are called with the same interface, with one
exception. The exception is that a trap may return a null string. When a
trap returns a null string, the corresponding Config\_ function is
invoked; otherwise the invocation of the trap replaces the potential
invocation of the Config\_ function. In the rest of this standard, the
trapping mechanism is not shown explicitly. It is implied by the use of
a Config\_ function. The names of the traps are

\begin{itemize}
\item
  Trap\_Command;
\item
  Trap\_ExternalRoutine;
\item
  Trap\_Push;
\item
  Trap\_Queue;
\item
  Trap\_Pull;
\item
  Trap\_Queued;
\item
  Trap\_Trace\_Query;
\item
  Trap\_Trace\_Input;
\item
  Trap\_Trace\_Output;
\item
  Trap\_Default\_Input;
\item
  Trap\_Default\_Output;
\item
  Trap\_Initialization;
\item
  Trap\_Termination;
\item
  Trap\_Halt\_Query;
\item
  Trap\_Halt\_Reset.
\end{itemize}

\hypertarget{variable-pool}{%
\section{Variable pool}\label{variable-pool}}

How does this fit with variables as properties?

The variable pool interface consists of functions which the
configuration shall provide to manipulate the variables and to obtain
some characteristics of a Rexx program.

These functions can be called from programs not written in Rexx \_
commands and external routines invoked from a Rexx program, or traps
invoked from the language processor.

All the functions comprising the variable pool interface shall return
with an indication of whether an error occurred. They shall return
indicating an error and have no other effect, if \#API\_Enabled has a
value of `0' or if the arguments to them fail to meet the defined
syntactic constraints.

These functions interact with the processing of clauses. To define this
interaction, the functions are described here in terms of the processing
of variables, see nnn.

Some of these functions have an argument which is a symbol. A symbol is
a string. The content of the string shall meet the syntactic constraints
of the left hand side of an assignment. Conversion to uppercase and
substitution in compound symbols occurs as it does for the left hand
side of an assignment. The symbol identifies the variable to be operated
upon.

Some of the functions have an argument which is a direct symbol. A
direct symbol is a string. The content of this string shall meet the
syntactic constraints of a VAR\_SYMBOL in uppercase with no periods or
it shall be the concatenation of a part meeting the syntactic
constraints of a stem in uppercase, and a part that is any string. In
the former case the symbol identifies the variable to be operated upon.
In the latter case the variable to be operated on is one with the
specified stem and a tail which is the remainder of the direct symbol.

Functions that have an argument which is symbol or direct symbol shall
return an indication of whether the identified variable existed before
the function was executed. Clause nnn defines functions which manipulate
Rexx variable pools. Where possible the functions comprising the
variable pool interface are described in terms of the appropriate
invocations of the functions defined in nnn. The first parameter on
these calls is the state variable \#Pool. If these Var\_ functions do
not return an indicator `N', `R', or `D' then the API function shall
return an error indication.

\hypertarget{api-set}{%
\subsection{API Set}\label{api-set}}

Syntax: API Set(Symbol, Value) where: Symbol is a symbol. Value is the
string whose value is to be assigned to the variable. Semantics: Assign
the value of Value to the variable identified by Symbol. If Symbol
contains no periods or contains one period as its last character: Var\_
Set(\#Pool, Symbol, `0', Value) Otherwise: Var \_Set(\#Pool, \#Symbol,
`1', Value) where: \#Symbol is Symbol after any replacements in the tail
as described by nnn.

\hypertarget{api-value}{%
\subsection{API Value}\label{api-value}}

Syntax: API Value (Symbol)

where: Symbol is a symbol. Semantics: Return the value of the variable
identified by Symbol. If Symbol contains no periods or contains one

period as its last character this is the value of \#Outcome after: Var
\_Value(\#Pool, Symbol, `0')

Otherwise the value of \#Outcome after: Var Value(\#Pool, \#Symbol, `1')
where: \#Symbol is Symbol after any replacements in the tail as
described by nnn.

\hypertarget{api_drop}{%
\subsection{API\_Drop}\label{api_drop}}

Syntax: API Drop (Symbol)

where: Symbol is a symbol. Semantics: Drop the variable identified by
Symbol. If Symbol contains no periods or contains one period as its last
character: Var Drop(\#Pool, Symbol, `0') Otherwise: Var Drop(\#Pool,
\#Symbol, `1') where: \#Symbol is Symbol after any replacements in the
tail as described by nnn.

\hypertarget{api-setdirect}{%
\subsection{API SetDirect}\label{api-setdirect}}

Syntax: API SetDirect (Symbol, Value)

where: Symbol is a direct symbol. Value is the string whose value is to
be assigned to the variable. Semantics:

Assign the value of Value to the variable identified by Symbol. If the
Symbol contains no period: Var\_ Set(\#Pool, Symbol, `0', Value)

Otherwise: Var\_ Set(\#Pool, Symbol, `1', Value)

\hypertarget{api_valuedirect}{%
\subsection{API\_ValueDirect}\label{api_valuedirect}}

Syntax: API ValueDirect (Symbol)

where: Symbol is a direct symbol. Semantics: Return the value of the
variable identified by Symbol. If the Symbol contains no period: Var
\_Value(\#Pool, Symbol, `0') Otherwise: Var \_Value(\#Pool, Symbol, `1')

\hypertarget{api-dropdirect}{%
\subsection{API DropDirect}\label{api-dropdirect}}

Syntax: API DropDirect (Symbol)

where: Symbol is a direct symbol. Semantics:

Drop the variable identified by Symbol. If the Symbol contains no
period: Var Drop(\#Pool, Symbol, `0')

Otherwise: Var Drop(\#Pool, Symbol, `1')

\hypertarget{api-valueother}{%
\subsection{API ValueOther}\label{api-valueother}}

Syntax: API ValueOther (Qualifier) where: Qualifier is an indication
distinguishing the result to be returned including any necessary further
qualification. Semantics: Return characteristics of the program,
depending on the value of Qualifier. The possibilities for the value to
be returned are:

\begin{itemize}
\item
  the value of \#Source;
\item
  the value of \#Version;
\item
  the largest value of n such that \#ArgExists.1.n is `1', see nnn;
\item
  the value of \#Arg.1.n where n is an integer value provided as input.
\end{itemize}

\hypertarget{api-next}{%
\subsection{API Next}\label{api-next}}

Syntax: API Next ()

Semantics:

Returns both the name and the value of some variable in the variable
pool that does not have the attribute `dropped' or the attribute
`implicit' and is not a stem; alternatively return an indication that
there is no suitable name to return. When API\_Next is called it will
return a name that has not previously been returned; the order is
undefined. This process of returning different names will restart
whenever the Rexx processor executes Var\_Reset.

\hypertarget{api-nextvariable}{%
\subsection{API NextVariable}\label{api-nextvariable}}

Syntax: API NextVariable()

Semantics:

Returns both the name and the value of some variable in the variable
pool that does not have the attribute `dropped' or the attribute
`implicit'; alternatively, return an indication that there is no
suitable name to return. When API NextVariable is called it will return
data about a variable that has not previously been returned; the order
is undefined. This process of returning different names will restart
whenever the Rexx processor executes Var\_Reset. In addition to the name
and value, an indication of whether the variable was `tailed' will be
returned.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/conformance.tex
complete, but very deficient, extended standard

\hypertarget{conformance}{%
\chapter{Conformance}\label{conformance}}

\hypertarget{conformance-1}{%
\section{Conformance}\label{conformance-1}}

A conforming language processor shall not implement any variation of
this standard except where this standard permits. Such permitted
variations shall be implemented in the manner prescribed by this
standard and noted in the documentation accompanying the processor. A
conforming processor shall include in its accompanying documentation

\begin{itemize}
\item
  alist of all definitions or values for the features in this standard
  which are specified to be dependent on the configuration.
\item
  a statement of conformity, giving the complete reference of this
  standard (ANSI X3.274-1996) with which conformity is claimed.
\end{itemize}

\hypertarget{limits}{%
\section{Limits}\label{limits}}

Aside from the items listed here (and the assumed limitation in
resources of the configuration), a conforming language processor shall
not put numerical limits on the content of a program. Where a limit
expresses the limit on a number of digits, it shall be a multiple of
three. Other limits shall be one of the numbers one, five or twenty
five, or any of these multiplied by some power of ten. Limitations that
conforming language processors may impose are:

\begin{itemize}
\item
  NUMERIC DIGITS values shall be supported up to a value of at least
  nine hundred and ninety nine.
\item
  Exponents shall be supported. The limit of the absolute value of an
  exponent shall be at least as large as the largest number that can be
  expressed without an exponent in nine digits.
\item
  String lengths shall be supported. The limit on the length shall be at
  least as large as the largest number that can be expressed without an
  exponent in nine digits.
\item
  String literal length shall be supported up to at least two hundred
  and fifty.
\item
  Symbol length shall be supported up to at least two hundred and fifty.
\end{itemize}

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/definitions_and_document_notation.tex
complete, but very deficient, extended standard

\hypertarget{definitions-and-document-notation}{%
\chapter{Definitions and document
notation}\label{definitions-and-document-notation}}

Lots more for NetRexx

\hypertarget{definitions}{%
\section{Definitions}\label{definitions}}

\hypertarget{application-programming-interface}{%
\subsection{application programming
interface:}\label{application-programming-interface}}

A set of functions which allow access to some Rexx facilities from
non-Rexx programs.

\hypertarget{arguments}{%
\subsection{arguments:}\label{arguments}}

The expressions (separated by commas) between the parentheses of a
function call or following the name on a CALL instruction. Also the
corresponding values which may be accessed by a function or routine,
however invoked.

\hypertarget{built-in-function}{%
\subsection{built-in function:}\label{built-in-function}}

A function (which may be called as a subroutine) that is defined in
section nnn of this standard and can be used directly from a program.

\hypertarget{character-string}{%
\subsection{character string:}\label{character-string}}

A sequence of zero or more characters.

\hypertarget{clause}{%
\subsection{clause:}\label{clause}}

A section of the program, ended by a semicolon. The semicolon may be
implied by the end of a line or by some other constructs.

\hypertarget{coded}{%
\subsection{coded:}\label{coded}}

A coded string is a string which is not necessarily comprised of
characters. Coded strings can occur as arguments to a program, results
of external routines and commands, and the results of some built-in
functions, such as D2C.

\hypertarget{command}{%
\subsection{command:}\label{command}}

A clause consisting of just an expression is an instruction known as a
command. The expression is evaluated and the result is passed as a
command string to some external environment. \#\#\# condition:

A specific event, or state, which can be trapped by CALL ON or SIGNAL
ON.

\hypertarget{configuration}{%
\subsection{configuration:}\label{configuration}}

Any data-processing system, operating system and software used to
operate a language processor.

\hypertarget{conforming-language-processor}{%
\subsection{conforming language
processor:}\label{conforming-language-processor}}

A language processor which obeys all the provisions of this standard.

\hypertarget{construct}{%
\subsection{construct:}\label{construct}}

A named syntax grouping, for example ``expression'', ``do\_
specification''.

\hypertarget{default-error-stream}{%
\subsection{default error stream:}\label{default-error-stream}}

An output stream, determined by the configuration, on which error
messages are written.

\hypertarget{default-input-stream}{%
\subsection{default input stream:}\label{default-input-stream}}

An input stream having a name which is the null string. The use of this
stream may be implied.

\hypertarget{default-output-stream}{%
\subsection{default output stream:}\label{default-output-stream}}

An output stream having a name which is the null string. The use of this
stream may be implied.

\hypertarget{direct-symbol}{%
\subsection{direct symbol:}\label{direct-symbol}}

A symbol which, without any modification, names a variable in a variable
pool.

\hypertarget{directive}{%
\subsection{directive:}\label{directive}}

Clauses which begin with two colons are directives. Directives are not
executable, they indicate the structure of the program. Directives may
also be written with the two colons implied.

\hypertarget{dropped}{%
\subsection{dropped:}\label{dropped}}

A symbol which is in an unitialized state, as opposed to having had a
value assigned to it, is described as dropped. The names in a variable
pool have an attribute of `dropped' or `not-dropped'.

\hypertarget{encoding}{%
\subsection{encoding:}\label{encoding}}

The relation between a character string anda corresponding number. The
encoding of character strings is determined by the configuration.

\hypertarget{end-of-line}{%
\subsection{end-of-line:}\label{end-of-line}}

An event that occurs during the scanning of a source program. Normally
the end-of-lines will relate to the lines shown if the configuration
lists the program. They may, or may not, correspond to characters in the
source program.

\hypertarget{environment}{%
\subsection{environment:}\label{environment}}

The context in which a command may be executed. This is comprised of the
environment name, details of the resource that will provide input to the
command, and details of the resources that will receive output of the
command.

\hypertarget{environment-name}{%
\subsection{environment name:}\label{environment-name}}

The name of an external procedure or process that can execute commands.
Commands are sent to the current named environment, initially selected
externally but then alterable by using the ADDRESS instruction.

\hypertarget{error-number}{%
\subsection{error number:}\label{error-number}}

A number which identifies a particular situation which has occurred
during processing. The message prose associated with such a number is
defined by this standard.

\hypertarget{exposed}{%
\subsection{exposed:}\label{exposed}}

Normally, a symbol refers to a variable in the most recently established
variable pool. When this is not the case the variable is referred to as
an exposed variable.

\hypertarget{expression}{%
\subsection{expression:}\label{expression}}

The most general of the constructs which can be evaluated to produce a
single string value.

\hypertarget{external-data-queue}{%
\subsection{external data queue:}\label{external-data-queue}}

A queue of strings that is external to REXX programs in that other
programs may have access

to the queue whenever REXX relinquishes control to some other program.

\hypertarget{external-routine}{%
\subsection{external routine:}\label{external-routine}}

A function or subroutine that is neither built-in nor in the same
program as the CALL instruction or function call that invokes it.

\hypertarget{external-variable-pool}{%
\subsection{external variable pool:}\label{external-variable-pool}}

A named variable pool supplied by the configuration which can be
accessed by the VALUE built-in function.

\hypertarget{function}{%
\subsection{function:}\label{function}}

Some processing which can be invoked by name and will produce a result.
This term is used for both Rexx functions (See nnn) and functions
provided by the configuration (see n).

\hypertarget{identifier}{%
\subsection{identifier:}\label{identifier}}

The name of a construct.

\hypertarget{implicit-variable}{%
\subsection{implicit variable:}\label{implicit-variable}}

A tailed variable which is in a variable pool solely as a result of an
operation on its stem. The names in a variable pool have an attribute of
`implicit' or `not-implicit'.

\hypertarget{instruction}{%
\subsection{instruction:}\label{instruction}}

One or more clauses that describe some course of action to be taken by
the language processor.

\hypertarget{internal-routine}{%
\subsection{internal routine:}\label{internal-routine}}

A function or subroutine that is in the same program as the CALL
instruction or function call that invokes it.

\hypertarget{keyword}{%
\subsection{keyword:}\label{keyword}}

This standard specifies special meaning for some tokens which consist of
letters and have particular spellings, when used in particular contexts.
Such tokens, in these contexts, are keywords.

\hypertarget{label}{%
\subsection{label:}\label{label}}

A clause that consists of a single symbol or a literal followed by a
colon.

\hypertarget{language-processor}{%
\subsection{language processor:}\label{language-processor}}

Compiler, translator or interpreter working in combination with a
configuration.

\hypertarget{notation-function}{%
\subsection{notation function:}\label{notation-function}}

A function with the sole purpose of providing a notation for describing
semantics, within this standard. No Rexx program can invoke a notation
function.

\hypertarget{null-clause}{%
\subsection{null clause:}\label{null-clause}}

A clause which has no tokens.

\hypertarget{null-string}{%
\subsection{null string:}\label{null-string}}

A character string with no characters, that is, a string of length zero.

\hypertarget{production}{%
\subsection{production:}\label{production}}

The definition of a construct, in Backus-Naur form.

\hypertarget{return-code}{%
\subsection{return code:}\label{return-code}}

A string that conveys some information about the command that has been
executed. Return codes usually indicate the success or failure of the
command but can also be used to represent other information.

\hypertarget{routine}{%
\subsection{routine:}\label{routine}}

Some processing which can be invoked by name.

\hypertarget{state-variable}{%
\subsection{state variable:}\label{state-variable}}

A component of the state of progress in processing a program, described
in this standard by a named variable. No Rexx program can directly
access a state variable.

\hypertarget{stem}{%
\subsection{stem:}\label{stem}}

If a symbol naming a variable contains a period which is not the first
character, the part of the symbol up to and including the first period
is the stem.

\hypertarget{stream}{%
\subsection{stream:}\label{stream}}

Named streams are used as the sources of input and the targets of
output. The total semantics of such a stream are not defined in this
standard and will depend on the configuration. A stream may be a
permanent file in the configuration or may be something else, for
example the input from a keyboard.

\hypertarget{string}{%
\subsection{string:}\label{string}}

For many operations the unit of data is a string. It may, or may not, be
comprised of a sequence of characters which can be accessed
individually.

\hypertarget{subcode}{%
\subsection{subcode:}\label{subcode}}

The decimal part of an error number.

\hypertarget{subroutine}{%
\subsection{subroutine:}\label{subroutine}}

An internal, built-in, or external routine that may or may not return a
result string and is invoked by the CALL instruction. If it returns a
result string the subroutine can also be invoked by a function call, in
which case it is being called as a function. \#\#\# symbol:

A sequence of characters used as a name, see nnn. Symbols are used to
name variables, functions, etc.

\hypertarget{tailed-name}{%
\subsection{tailed name:}\label{tailed-name}}

The names in a variable pool have an attribute of `tailed' or
`non-tailed'. Otherwise identical names are distinct if their attributes
differ. Tailed names are normally the result of replacements in the tail
of a symbol, the part that follows a stem.

\hypertarget{token}{%
\subsection{token:}\label{token}}

The unit of low-level syntax from which high-level constructs are built.
Tokens are literal strings, symbols, operators, or special characters.

\hypertarget{trace}{%
\subsection{trace:}\label{trace}}

A description of some or all of the clauses of a program, produced as
each is executed.

\hypertarget{trap}{%
\subsection{trap:}\label{trap}}

A function provided by the user which replaces or augments some normal
function of the language processor.

\hypertarget{variable-pool}{%
\subsection{variable pool:}\label{variable-pool}}

A collection of the names of variables and their associated values.

\hypertarget{document-notation}{%
\section{Document notation}\label{document-notation}}

\hypertarget{rexx-code}{%
\subsection{Rexx Code}\label{rexx-code}}

Some Rexx code is used in this standard. This code shall be assumed to
have its private set of variables. Variables used in this code are not
directly accessible by the program to be processed. Comments in the code
are not part of the provisions of this standard.

\hypertarget{italics}{%
\subsection{Italics}\label{italics}}

Throughout this standard, except in Rexx code, references to the
constructs defined in section nnn are italicized.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/directives.tex
complete, but very deficient, extended standard

\hypertarget{directives}{%
\chapter{Directives}\label{directives}}

The syntax constructs which are introduced by the optional `::' token
are known as directives.

\hypertarget{notation}{%
\section{Notation}\label{notation}}

Notation functions are functions which are not directly accessible as
functions in a program but are used in this standard as a notation for
defining semantics.

Some notation functions allow reference to syntax constructs defined in
nnn. Which instance of the syntax construct in the program is being
referred to is implied; it is the one for which the semantics are being
specified.

The BNF\_primary referenced may be directly in the production or in some
component referenced in the

production, recursively. The components are considered in left to right
order. \#Contains (Identifier, BNF primary)

where:

Identifier is an identifier in a production (see nnn) defined in nnn.

BNF\_primary is a bnf\_primary (see nnn) in a production defined in nnn.
Return `1' if the production identitied by identifier contained a
bnf\_primary identified by BNF\_primary, otherwise return `0'.

\#Instance (Identifier, BNF primary) where: Identifier is an identifier
in a production defined in nnn. BNF\_primary is a Onf\_primary in a
production defined in nnn. Returns the content of the particular
instance of the BNF\_primary. If the BNF\_primary is a VAR\_SYMBOL this
is referred to as the symbol ``taken as a constant.''

\#Evaluate (Identifier, BNF primary) where: Identifier is an identifier
in a production defined in nnn. BNF\_primary is a Onf\_primary in a
production defined in nnn. Return the value of the BNF\_primary in the
production identified by Identifier.

\#Execute (Identifier, BNF primary) where: Identifier is an identifier
in a production defined in nnn. BNF\_primary is a Onf\_primary in a
production defined in nnn. Perform the instructions identified by the
BNF\_primary in the production identified by Identifier.

\#Parses (Value, BNF primary) where: Value is a string BNF\_primary is a
Onf\_primary in a production defined in nnn. Return `1' if Value matches
the definition of the BNF\_primary, by the rules of clause 6, `0'
otherwise.

\#Clause (Label) where: Label is a label in code used by this standard
to describe processing. Return an identification of that label. The
value of this identification is used only by the \#Goto notation
function.

\#Goto (Value) where:

Value identifies a label in code used by this standard to describe
processing. The description of processing continues at the identified
label.

\#Retry ()

This notation is used in the description of interactive tracing to
specify re-execution of the clause just previously executed. It has the
effect of transferring execution to the beginning of that clause, with
state variable \#Loop set to the value it had when that clause was
previously executed.

\hypertarget{initializing}{%
\section{Initializing}\label{initializing}}

Some of the initializing, now grouped in classic section 8.2.1 will have
to come here so that we have picked up anything from the START\_API that
needs to be passed on to the execution of REQUIRES subject.

We will be using some operations that are forward reference to what was
section nnn.

\hypertarget{program-initialization-and-message-texts}{%
\subsection{Program initialization and message
texts}\label{program-initialization-and-message-texts}}

Processing of a program begins when API\_Start is executed. A pool
becomes current for the reserved

variables. call Config ObjectNew \#ReservedPool = \#Outcome \#Pool =
\#ReservedPool Is it correct to make the reserved variables and the
builtin objects in the same pool?

Some of the values which affect processing of the program are parameters
of API\_ Start:

\#Howlnvoked is set to `COMMAND', `FUNCTION' or `SUBROUTINE' according
to the first parameter of APL Start.

\#Source is set to the value of the second parameter of API\_ Start.

The third parameter of API\_Start is used to determine the initial
active environment.

The fourth parameter of API Start is used to determine the arguments.
For each argument position \#ArgExists.1.ArgNumber is set `1' if there
is an argument present, `0' if not. ArgNumber is the number of the
argument position, counting from 1. If \#ArgExists.1.ArgNumber is `1'
then \#Arg.1.ArgNumber is set to the value of the corresponding
argument. If \#ArgExists.1.ArgNumber is `0' then \#Arg.1.Arg is set to
the null string. \#ArgExists.1.0 is set to the largest n for which
\#ArgExists.1.n is `1', or to zero if there is no such value of n.

Some of the values which affect processing of the program are provided
by the configuration: call Config OtherBlankCharacters

\#A11Blanks\textless Index ``\#Al1Blanks'' \# ``\,'' \textgreater{} = '
'\#Outcome /* ``Real'' blank concatenated with others */

\#Bif Digits. = 9

call Config Constants

-true = `1'

-false = `0O'

Objects in our model are only distinquished by the values within their
pool so we can construct the builtin classes incomplete and then
complete them with directives.

Can we initialize the methods of .nil by directives?

call Config ObjectNew

-List = \#Outcome

call var\_set .List, \#IsClass, `0', `1'

call var\_set .List, \#ID, `0', `List'

Some of the state variables set by this call are limits, and appear in
the text of error messages. The relation between message numbers and
message text is defined by the following list, where the message

number appears immediately before an `=' and the message text follows in
quotes.

\#ErrorText. stl

\#ErrorText.0.1 = `Error running , line :'

\#ErrorText.0.2 = `Error in interactive trace:'

\#ErrorText.0.3 = `Interactive trace. ``Trace Off'' to end debug.',
``ENTER to continue.'

\#ErrorText.2 = `Failure during finalization'

\#ErrorText.2.1 = `Failure during finalization: '

\#ErrorText.3 \#ErrorText.3.1

`Failure during initialization' `Failure during initialization: '

\#ErrorText.4 \#ErrorText.4.1

`Program interrupted' `Program interrupted with HALT condition: '

\#ErrorText.5 \#ErrorText.5.1

`System resources exhausted' `System resources exhausted: '

\#ErrorText.6 \#ErrorText.6.1

'Unmatched ``/\emph{'' or quote! 'Unmatched comment delimiter (''/}'')!

\#ErrorText.6.2 \#ErrorText.6.3

\#ErrorText.7 = \#ErrorText.7.1 =

\#ErrorText.7.2 =

\#ErrorText.7.3 =

\#ErrorText.8 = \#ErrorText.8.1 \#ErrorText.8.2

\#ErrorText.9 = \#ErrorText.9.1 \#ErrorText.9.2

\#ErrorText.10 \#ErrorText.10.1= \#ErrorText.10.2=

\#ErrorText.10.3=

\#ErrorText.10.4= \#ErrorText.10.5= \#ErrorText.10.6=

\#ErrorText.13 = \#ErrorText.13.1=

\#ErrorText.14 = \#ErrorText.14.1= \#ErrorText.14.2= \#ErrorText.14.3=
\#ErrorText.14.4=

\#ErrorText.15 = \#ErrorText.15.1=

\#ErrorText.15.2= \#ErrorText.15.3= \#ErrorText.15.4= \#ErrorText.16 =
\#ErrorText.16.1= \#ErrorText.16.2= \#ErrorText.16.3=

\#ErrorText.17 = \#ErrorText.17.1=

\#ErrorText.17.2=

\#ErrorText.18 =

\#ErrorText.18.1=

\#ErrorText.18.2=

``Unmatched single quote (`)'' `Unmatched double quote (``)'

`WHEN or OTHERWISE expected' ``SELECT on line `found ``''\,'

``SELECT on line `or END; found ``''\,'!

requires WHEN;',

`All WHEN expressions of SELECT on line are',

`false; OTHERWISE expected'

``Unexpected THEN or ELSE'

`THEN has no corresponding IF or WHEN clause' `ELSE has no corresponding
THEN clause'

``Unexpected WHEN or OTHERWISE' `WHEN has no corresponding SELECT'
`OTHERWISE has no corresponding SELECT'

`Unexpected or unmatched END'

`END has no corresponding DO or SELECT'

`END corresponding to DO on line ', `must have a symbol following that
matches', `the control variable (or no symbol);',

`found ``''\,'!

`END corresponding to DO on line ', `must not have a symbol following it
because', `there is no control variable;',

`found ``''\,'!

`END corresponding to SELECT on line ', `must not have a symbol
following;',

`found ``''\,'!

`END must not immediately follow THEN'

`END must not immediately follow ELSE'

`Invalid character in program' `Invalid character ``('`X)'' in program'

``Incomplete DO/SELECT/IF'

`DO instruction requires a matching END' `SELECT instruction requires a
matching END' `THEN requires a following instruction' `ELSE requires a
following instruction'

`Invalid hexadecimal or binary string'

`Invalid location of blank in position', ` in hexadecimal string'

`Invalid location of blank in position', ` in binary string'

`Only 0-9, a-f, A-F, and blank are valid in a', ``hexadecimal string;
found''``\,'

`Only 0, 1, and blank are valid in a',

`binary string; found ``''\,'

`Label not found'

`Label ``'' not found'

`Cannot SIGNAL to label ``'' because it is', `inside an IF, SELECT or DO
group'

`Cannot invoke label ``'' because it is', `inside an IF, SELECT or DO
group'

`Unexpected PROCEDURE'

``PROCEDURE is valid only when it is the first', `instruction executed
after an internal CALL', `or function invocation'

`The EXPOSE `instruction executed after a method invocation!' `THEN
expected'

`IF keyword on line requires', `matching THEN clause; found ``''\,'

`WHEN keyword on line requires',

requires WHEN, OTHERWISE,',

instruction is valid only when it is the first', \#ErrorText.19 =
\#ErrorText.19.1=

\#ErrorText.19.2= \#ErrorText.19.3= \#ErrorText.19.4= \#ErrorText.19.6=
\#ErrorText.19.7= \#ErrorText.19.8=

\#ErrorText.19.9=

\#ErrorText.19.11='String or symbol

\#ErrorText.19.12='String or symbol

\#ErrorText.19.13=

\#ErrorText.19.15='String or symbol

\#ErrorText.19.16='String or symbol

\#ErrorText.19.17=

Unsound now we are using ``term'?

65

\#ErrorText.20 = \#ErrorText.20.1= \#ErrorText.20.2= \#ErrorText.20.3=

\#ErrorText.21 \#ErrorText.21.1

\#ErrorText.22 = \#ErrorText.22.1=

\#ErrorText.23 = \#ErrorText.23.1=

\#ErrorText.24 = \#ErrorText.24.1= \#ErrorText.25 = \#ErrorText.25.1=
\#ErrorText.25.2= \#ErrorText.25.3= \#ErrorText.25.4= \#ErrorText.25.5=
\#ErrorText.25.6= \#ErrorText.25.7=

\#ErrorText.25.8=

`matching THEN clause; found ``''\,' `String or symbol expected' `String
or symbol expected after ADDRESS keyword;', ``found''``! `String or
symbol expected after CALL keyword;',''found ``''! `String or symbol
expected after NAME keyword;', ``found''``! `String or symbol expected
after SIGNAL keyword;',''found ``''! `String or symbol expected after
TRACE keyword;', ``found''``! `Symbol expected in parsing
pattern;',''found ``''! `String or symbol expected after REQUIRES;',
``found''``! `String or symbol expected after METHOD;',''found ``''!

expected after ROUTINE;', ``found''``!

expected after CLASS;', ``found''``! `String or symbol expected after
INHERIT;',''found ``''!

expected after METACLASS;', ``found''``!

expected after MIXINCLASS;`, ``found''``! `String or symbol expected
after SUBCLASS;',''found ``''! 'Name expected' `Name required; found
``''\,'

`Found ``'' where only a name is valid'

'Found ``'' where only a name or

`Invalid data on end of clause' `The clause ended at an unexpected
token;',

`found ``''\,'

'Invalid ``Invalid

'Invalid ``Invalid

'Invalid

character string' character string

data string' data string

TRACE request'

is valid'

''xX''

`'! xX''

`TRACE request letter must be one of',

' ``ACEFILNOR'';

found

``evalue\textgreater{}'''

`Invalid sub-keyword found' `CALL ON must be followed by one of the',

'keywords ;

found

``etoken\textgreater{}''!

`CALL OFF must be followed by one of the',

'keywords ;

found

``etoken\textgreater{}''!

``SIGNAL ON must be followed by one of the',

'keywords ;

found

``etoken\textgreater{}''!

`SIGNAL OFF must be followed by one of the',

'keywords ;

found

``etoken\textgreater{}''!

``ADDRESS WITH must be followed by one of the',

'keywords ;

found

``etoken\textgreater{}''!

``INPUT must be followed by one of the',

'keywords ; 'OUTPUT must be followed by 'keywords ; ``APPEND must be
followed by 'keywords ;

found

found

found

``etoken\textgreater{}''! one of the', ``etoken\textgreater{}''! one of
the', ``etoken\textgreater{}''!

\#ErrorText.25.9=

\#ErrorText.25.11= \#ErrorText.25.12= \#ErrorText.25.13=
\#ErrorText.25.14= \#ErrorText.25.15= \#ErrorText.25.16=
\#ErrorText.25.17=

\#ErrorText.25.18=

\#ErrorText.26 \#ErrorText.26.1=

\#ErrorText.26.2=

\#ErrorText.26.3=

\#ErrorText.26.4= \#ErrorText.26.5=

\#ErrorText.26.6=

\#ErrorText.26.7= \#ErrorText.26.8=

\#ErrorText.26.11

\#ErrorText.26.12

\#ErrorText.27 \#ErrorText.27.1=

\#ErrorText.28 \#ErrorText.28.1= \#ErrorText.28.2= \#ErrorText.28.3=

\#ErrorText.28.4=

\#ErrorText.29 \#ErrorText.29.1=

\#ErrorText.30 \#ErrorText.30.1= \#ErrorText.30.2=

\#ErrorText.31 \#ErrorText.31.1=

\#ErrorText.31.2=

\#ErrorText.31.3=

``REPLACE must be followed by one of the', `keywords ; found
``''\,'\,''NUMERIC FORM must be followed by one of the', `keywords ;
found ``''\,'

``PARSE must be followed by one of the', `keywords ; found ``''\,'

``UPPER must be followed by one of the', `keywords ; found ``''\,'

``ERROR must be followed by one of the', `keywords ; found
``''\,'\,''NUMERIC must be followed by one of the', `keywords ; found
``''\,' `FOREVER must be followed by one of the', ``keywords or nothing;
found''``'''PROCEDURE must be followed by the keyword', ``EXPOSE or
nothing; found''``\,'

``FORWARD must be followed by one of the the keywords',

`; found ``''\,'

`Invalid whole number'

`Whole numbers must fit within current DIGITS', `setting(); found
``''\,'

`Value of repetition count expression in DO instruction',

`must be zero or a positive whole number;', `found ``''\,'

`Value of FOR expression in DO instruction', `must be zero or a positive
whole number;', `found ``''\,'

``Positional pattern of parsing template', `must be a whole number;
found ``''\,'\,''NUMERIC DIGITS value',

`must be a positive whole number; ``NUMERIC FUZZ value',

`must be zero or a positive whole number;',

`found ``''\,'

`Number used in TRACE setting',

`must be a whole number; found ``''\,'

'Operand to right of the power operator (``**``)`, 'must be a whole
number; found ``''\,'

``Result of \% operation would need',

found ``'''

`exponential notation at current NUMERIC DIGITS '

``Result of \% operation used for // `, `operation would need',

`exponential notation at current NUMERIC DIGITS '

`Invalid DO syntax' `Invalid use of keyword ``'' in DO clause'

`Invalid LEAVE or ITERATE'

`LEAVE is valid only within a repetitive DO loop' `ITERATE is valid only
within a repetitive DO loop' `Symbol following LEAVE (``'') must',

`either match control variable of a current',

`DO loop or be omitted'

`Symbol following ITERATE (``'') must', `either match control variable
of a current',

`DO loop or be omitted'

`Environment name too long' `Environment name exceeds',

\#Limit EnvironmentName `characters; found ``''\,' `Name or string too
long'

`Name exceeds' \#Limit Name `characters' ``Literal string exceeds'
\#Limit Literal `characters' `Name starts with number or ``.''\,'!

`A value cannot be assigned to a number;',

`found ``''\,'

`Variable symbol must not start with a number;', `found ``''\,'

`Variable symbol must not start with a ``.'';',

`found ``''\,' \#ErrorText.33 = \#ErrorText.33.1=

\#ErrorText.33.2=

\#ErrorText.33.3=

\#ErrorText.34 = \#ErrorText.34.1=

`Invalid expression result' `Value of NUMERIC DIGITS (``'')',

'must exceed value of NUMERIC FUZZ

``(\textless¢value\textgreater)'''!

`Value of NUMERIC DIGITS (``'')',

`must not exceed'

\#Limit Digits

``Result of expression following NUMERIC FORM',

'must start with ``E''

or ``S''; found ``'''

``Logical value not''0'' or ``1''!

'Value of expression

following IF keyword',

`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.2= `Value
of expression following WHEN keyword',

`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.3= `Value
of expression following WHILE keyword',

`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.4= `Value
of expression following UNTIL keyword',

`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.5= `Value
of expression to left',

`of logical operator ``''\,',

`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.6= `Value
of expression to right',

`of logical operator ``''\,',

`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.35 = `Invalid
expression' \#ErrorText.35.1= `Invalid expression detected at ``''\,'

\#ErrorText.36

`Unmatched ``('' in expression'

\#ErrorText.37 = \#ErrorText.37.1= \#ErrorText.37.2=

'Unexpected n a n ny nt

`Unexpected ``,''! `Unmatched ``)'' in expression'

or

\#ErrorText.38 = \#ErrorText.38.1= \#ErrorText.38.2= \#ErrorText.38.3=

`Invalid template or pattern'

`Invalid parsing template detected at ``''\,' `Invalid parsing position
detected at ``''\,' ``PARSE VALUE instruction requires WITH keyword'

``Incorrect call to routine'

`External routine ``'' failed'

`Not enough arguments in invocation of ;', `minimum expected is '

`Too many arguments in invocation of ;', `maximum expected is '

`Missing argument in invocation of ;', `argument is required'
`ebif\textgreater{} argument ',

`exponent exceeds' \#Limit ExponentDigits ``found''``\,'

` argument ',

`must be a number; found ``''\,'! ` argument ',

\#ErrorText.40 = \#ErrorText.40.1= \#ErrorText.40.3= \#ErrorText.40.4=
\#ErrorText.40.5=

\#ErrorText.40.9= `digits;',

\#ErrorText.40.11=

\#ErrorText.40.12=

`must be a whole number; found ``''\,' \#ErrorText.40.13=` argument ',

`must be zero or positive; found ``''\,' \#ErrorText.40.14=` argument ',

`must be positive; found ``''\,'

\#ErrorText.40.17=` argument 1', `must have an integer part in the range
0:90 and a', `decimal part no larger than .9; found ``''\,'
\#ErrorText.40.18=` conversion must',

``have a year in the range 0001 to 9999! \#ErrorText.40.19=` argument 2,
``'', is not in the format',

`described by argument 3, ``''\,'

\#ErrorText.40.21=` argument must not be null' \#ErrorText.40.23=`
argument ',

`must be a single character; found ``''\,' \#ErrorText.40.24=` argument
1',

`must be a binary string; found ``''\,' \#ErrorText.40.25=` argument 1',

`must be a hexadecimal string; found ``''\,' \#ErrorText.40.26=`
argument 1',

`must be a valid symbol; found ``''\,' \#ErrorText.40.27=` argument 1',

`must be a valid stream name; found ``''\,' \#ErrorText.40.28=` argument
,',

`option must start with one of ``'';',

`found ``''\,'

\#ErrorText.40.29=` conversion to format ``'' is not allowed'
\#ErrorText.40.31=` argument 1 (``'') must not exceed 100000'
\#ErrorText.40.32=` the difference between argument 1 (``'') and',

`argument 2 (``'') must not exceed 100000' \#ErrorText.40.33=` argument
1 (``'') must be less than',

`or equal to argument 2 (``'')' \#ErrorText.40.34=` argument 1 (``'')
must be less than',

`or equal to the number of lines',

`in the program (\textless sourceline()\textgreater)'
\#ErrorText.40.35=` argument 1 cannot be expressed as a whole number;',

`found ``''\,'

\#ErrorText.40.36=` argument 1', `must be the name of a variable in the
pool;', `found ``''\,'

\#ErrorText.40.37=` argument 3',

`must be the name of a pool; found ``''\,' \#ErrorText.40.38=` argument
',

`is not large enough to format ``''\,' \#ErrorText.40.39=` argument 3 is
not zero or one; found ``''\,' \#ErrorText.40.41=` argument ',

`must be within the bounds of the stream;',

`found ``''\,'

\#ErrorText.40.42=` argument 1; cannot position on this stream;', `found
``''\,' \#ErrorText.40.45=` argument must be a single',

`non-alphanumeric character or the null string;',

' ``found '''

\#ErrorText.40.46=` argument 3, ``'', is a format incompatible',

`with separator specified in argument '

\#ErrorText.41 = `Bad arithmetic conversion' \#ErrorText.41.1=
`Non-numeric value (``'')', `to left of arithmetic operation ``''\,'

\#ErrorText.41.2= `Non-numeric value (``'')', `to right of arithmetic
operation ``''\,' \#ErrorText.41.3= `Non-numeric value (``'')',

`used with prefix operator ``''\,'

\#ErrorText.41.4= `Value of TO expression in DO instruction', `must be
numeric; found ``''\,'!

\#ErrorText.41.5= `Value of BY expression in DO instruction', `must be
numeric; found ``''\,'!

\#ErrorText.41.6= `Value of control variable expression of DO
instruction', `must be numeric; found ``''\,'!

\#ErrorText.41.7= `Exponent exceeds' \#Limit ExponentDigits `digits;',
`found ``''\,'

\#ErrorText.42 = `Arithmetic overflow/underflow' \#ErrorText.42.1l=
`Arithmetic overflow detected at', `Nevalue\textgreater{} ``;',
`exponent of result requires more than', \#Limit ExponentDigits `digits'
\#ErrorText.42.2= `Arithmetic underflow detected at',
`Nevalue\textgreater{} ``;`, `exponent of result requires more than',
\#Limit ExponentDigits `digits'\,''Arithmetic overflow; divisor must not
be zero'

\#ErrorText.42.3

``Routine not found' `Could not find routine ``''\,'

\#ErrorText.43 = \#ErrorText.43.1= \#ErrorText.44 = `Function did not
return data'

\#ErrorText.44.1= `No data returned from function ``''\,'

\#ErrorText.45 = `No data specified on function RETURN'
\#ErrorText.45.1= `Data expected on RETURN instruction because',
`routine ``'' was called as a function'

\#ErrorText.46 = `Invalid variable reference' \#ErrorText.46.1= `Extra
token (``'') found in variable', `reference; ``)'' expected'

\#ErrorText.47 = `Unexpected label' \#ErrorText.47.1l= `INTERPRET data
must not contain labels;', 'found ``''!

\#ErrorText.48 = `Failure in system service' \#ErrorText.48.1= `Failure
in system service: '

\#ErrorText.49 = `Interpretation Error' \#ErrorText.49.1=
`Interpretation Error: ' \#ErrorText.50 = `Unrecognized reserved symbol'

\#ErrorText.50.1= `Unrecognized reserved symbol ``''\,'

\#ErrorText.51 = `Invalid function name'

\#ErrorText.51.1= `Unquoted function names must not end with a period;',
`found ``''\,'

\#ErrorText.52

``Result returned by'''' is longer than', \#Limit String `characters'

\#ErrorText.53 = `Invalid option' \#ErrorText.53.1l= `Variable reference
expected', `after STREAM keyword; found ``''\,' \#ErrorText.53.2=
`Variable reference expected', `after STEM keyword; found ``''\,'
\#ErrorText.53.3= `Argument to STEM must have one period,', `as its last
character; found ``''\,' \#ErrorText.54 = `Invalid STEM value'
\#ErrorText.54.1= `For this use of STEM, the value of ``'' must be a',
`count of lines; found: ``''\,'

If the activity defined by clause 6 does not produce any error message,
execution of the program continues.

call Config NoSource

If Config\_NoSource has set \#NoSource to `0' the lines of source
processed by clause 6 are copied to \#SourceLine. , with \#SourceLine.O
being a count of the lines and \#SourceLine.n for n=1 to \#SourceLine.0
being the source lines in order.

If Config\_NoSource has set \#NoSource to `1' then \#SourceLine.0 is set
to 0. The following state variables affect tracing:

\#InhibitPauses = 0

\#InhibitTrace = 0

\#AtPause = 0 /* Off until interactive input being received. */

\#Trace QueryPrior = `No' An initial variable pool is established:

call Config ObjectNew

\#Pool = \#Outcome

\#P0011 = \#Pool

call Var\_Empty \#Pool

call Var\_Reset \#Pool

\#Level = 1 /* Level of invocation */ \#NewLevel = 2
\#IsFunction.\#Level = (\#HowInvoked == `FUNCTION')

For this first level, there is no previous level from which values are
inherited. The relevant fields are initialized.

\#Digits.\#Level = 9 /* Numeric Digits \emph{/ \#Form.\#Level =
`SCIENTIFIC' /} Numeric Form \emph{/ \#Fuzz.\#Level = 0 /} Numeric Fuzz
\emph{/ \#StartTime.\#Level = '\,' /} Elapsed time boundary */
\#LineNumber = '\,'

\#Tracing.\#Level = `N'

\#Interactive.\#Level = `0'

69 An environment is provided by the API\_ Start to become the initial
active environment to which commands will be addressed. The alternate
environment is made the same:

/* Call the environments ACTIVE, ALTERNATE, TRANSIENT where these are
never-initialized state variables.

Similarly call the redirections I O and E */

call EnvAssign ALTERNATE, \#Level, ACTIVE, \#Level

Conditions are initially disabled:

\#Enabling.SYNTAX.\#Level = `OFF' \#Enabling.HALT.\#Level = `OFF'
\#Enabling.ERROR.\#Level = `OFF' \#Enabling.FAILURE.\#Level = `OFF'
\#Enabling.NOTREADY.\#Level = `OFF' \#Enabling.NOVALUE.\#Level = `OFF'
\#Enabling.LOSTDIGITS.\#Level = `OFF'

\#PendingNow.HALT.\#Level = 0 \#PendingNow.ERROR.\#Level = 0
\#PendingNow.FAILURE.\#Level = 0 \#PendingNow.NOTREADY.\#Level = 0 /*
The following field corresponds to the results from the CONDITION
built-in function. */ \#Condition.\#Level = '\,' The opportunity is
provided for a trap to initialize the pool. \#API Enabled = `1' call
Var\_Reset \#Pool call Config Initialization \#API Enabled = `0' \#\#
REQUIRES For each requires in order of appearence: A use of Start\_API
with \#instance(requires, taken\_constant). Msg40.1 or a new if
completion `E'. Add Provides to an ordered collection. Not cyclic
because .LIST can be defined without defining REQUIRES but a fairly
profound forward reference. \#\# CLASS For each class in order of
appearence: \#ClassName = \#Instance(class, taken constant) call
var\_value \#ReservedPool, `\#CLASSES.'ClassName, '1' if \#Indicator ==
`D' then do call Config ObjectNew \#Class = \#Outcome call var\_set
\#ReservedPool, `\#CLASSES.'ClassName, '1', \#Class end else call
\#Raise `SYNTAX', nn.nn, \#ClassName

New instance of CLASS class added to list. Msg ``Duplicate ::CLASS
directive instruction''(?) \#\# METHOD

For each method in order of appearence: call Config ObjectNew \#Po00ol =
\#Outcome call Config ObjectSource (\#Pool) \#MethodName =
\#Instance(method, taken constant) call var\_value \#Class,
`\#METHODS.'\#MethodName, `1' if \#Indicator == `D' then call var set
\#Class, `\#METHODS.'\#MethodName, `1', \#Pool else call \#Raise
`SYNTAX', nn.nn, \#MethodName, \#ClassName

GUARDED \& public is default. if \#contains(method, `PRIVATE') then
m\textasciitilde setprivate; if \#contains(method, 'UNGUARDED)) then
m\textasciitilde setunguarded

Why is there a keyword for GUARDED but not for PUBLIC here?

Does CLASS option mean ENHANCE with Class class methods?

\#CurrentClass \textasciitilde class(\#instance(method,
taken\_constant), m)

For ATTRIBUTE, should we actually construct source for two methods?
ATTRIBUTE case needs test of null body. OO! doesn't have source (because
it actually traps UNKNOWN?).

For EXTERNAL test for null body. Simon Nash doc says ``Accessibility to
external methods \ldots{} is implementation-defined''. Left like that it
doesn't even tell us about search order. We will need a
Config\_ExternalClass to import the public names of the class.

\hypertarget{routine}{%
\section{ROUTINE}\label{routine}}

For each routine in order of appearence:

Add name (with duplicate check) to list for this file.

Extra step needed in the invocation search order. Although this is
nominally EXTERNAL we presumably wont use the external call mechanism.
(Except perhaps when the routine was made available by a REQUIRES; in
that case the PARSE SOURCE answer has to change.)

have the builtins-defined-by-directives elsewhere; it would make sense
if they wound up about here.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/evaluation.tex
complete, but very deficient, extended standard

\hypertarget{evaluation}{%
\chapter{Evaluation}\label{evaluation}}

The syntax section describes how expressions and the components of
expressions are written in a program. It also describes how operators
can be associated with the strings, symbols and function results which
are their operands.

This evaluation section describes what values these components have in
execution, or how they have no value because a condition is raised.

This section refers to the DATATYPE built-in function when checking
operands, see nnn. Except for considerations of limits on the values of
exponents, the test:

datatype (Subject) == `NUM' is equivalent to testing whether the subject
matches the syntax: num := {[}blank+{]} {[}`+' \textbar{} `-'{]}
{[}blank+{]} number {[}blank+{]}

For the syntax of number see nnn.

When the matching subject does not include a `-' the value is the value
of the number in the match, otherwise the value is the value of the
expression (0 - number).

The test:

datatype (Subject , `W')

is a test that the Subject matches that syntax and also has a value that
is ``whole'', that is has no non-zero fractional part.

When these two tests are made and the Subject matches the constraints
but has an exponent that is not

in the correct range of values then a condition is raised: call \#Raise
`SYNTAX', 41.7, Subject

This possibility is implied by the uses of DATATYPE and not shown
explicitly in the rest of this section nnn.

\hypertarget{variables}{%
\section{Variables}\label{variables}}

The values of variables are held in variable pools. The capabilities of
variable pools are listed here, together with the way each function will
be referenced in this definition.

The notation used here is the same as that defined in sections nnn and
nnn, including the fact that the Var\_ routines may return an indicator
of `N', `S' or `X'.

Each possible name in a variable pool is qualified as tailed or
non-tailed name; names with different qualification and the same
spelling are different items in the pool. For those Var\_ functions with
a third argument this argument indicates the qualification; it is `1'
when addressing tailed names or `0' when addressing non-tailed names.

Each item in a variable pool is associated with three attributes and a
value. The attributes are `dropped' or `not-dropped', `exposed' or
`not-exposed' and `implicit' or `not-implicit'.

A variable pool is associated with a reference denoted by the first
argument, with name Pool. The value of Pool may alter during execution.
The same name, in conjunction with different values of Pool, can
correspond to different values.

\hypertarget{var_empty}{%
\subsection{Var\_Empty}\label{var_empty}}

Var\_Empty (Pool)

The function sets the variable pool associated with the specified
reference to the state where every name is associated with attributes
`dropped', `implicit' and `not-exposed'.

\hypertarget{var-set}{%
\subsection{Var Set}\label{var-set}}

Var \_Set(Pool, Name, `0', Value)

The function operates on the variable pool with the specified reference.
The name is a non-tailed name. If the specified name has the `exposed'
attribute then Var\_Set operates on the variable pool referenced by
\#Upper in this pool and this rule is applied to that pool. When the
pool with attribute `not-exposed' for this name is determined the
specified value is associated with the specified name. It also
associates the attributes `not-dropped' and `not-implicit'. If that
attribute was previously `not-dropped' then the indicator returned is
`R'. The name is a stem if it contains just one period, as its rightmost
character. When the name is a stem Var\_Set(Pool, TailedName, `1',Value)
is executed for all possible valid tailed names which

have Name as their stem, and then those tailed-names are given the
attribute `implicit'. Var \_Set(Pool, Name, `1', Value)

The function operates on the variable pool with the specified reference.
The name is a tailed name. The left part of the name, up to and
including the first period, is the stem. The stem is a non-tailed name.
If the specified stem has the `exposed' attribute then Var\_Set operates
on the variable pool referenced by \#Upper in this pool and this rule is
applied to that pool. When the pool with attribute `not-exposed' for the
stem is determined the name is considered in that pool. If the name has
the `exposed' attribute then the

variable pool referenced by \#Upper in the pool is considered and this
rule applied to that pool. When the pool with attribute `not-exposed' is
determined the specified value is associated with the specified name. It
also associates the attributes `not-dropped' and `not-implicit' . If
that attribute was previously `not-dropped' then the indicator returned
is `R'.

\hypertarget{var_value}{%
\subsection{Var\_Value}\label{var_value}}

Var \_Value(Pool, Name, `0')

The function operates on the variable pool with the specified reference.
The name is a non-tailed name. If the specified name has the `exposed'
attribute then Var\_Value operates on the variable pool referenced by
\#Upper in this pool and this rule is applied to that pool. When the
pool with attribute `not-exposed' for this name is determined the
indicator returned is `D' if the name has `dropped' associated, `N'
otherwise. In the former case \#Outcome is set equal to Name, in the
latter case \#Outcome is set to the value most

recently associated with the name by Var\_Set. Var \_Value(Pool, Name,
`1')

The function operates on the variable pool with the specified reference.
The name is a tailed name. The left part of the name, up to and
including the first period, is the stem. The stem is a non-tailed name.
If the specified stem has the `exposed' attribute then Var\_Value
operates on the variable pool referenced by \#Upper in this pool and
this rule is applied to that pool. When the pool with attribute
`not-exposed' for the stem is determined the name is considered in that
pool. If the name has the `exposed' attribute then the variable pool
referenced by \#Upper in the pool is considered and this rule applied to
that pool. When the pool with attribute `not-exposed' is determined the
indicator returned is `D' if the name has `dropped' associated, `N'
otherwise. In the former case \#Outcome is set equal to Name, in the
latter case \#Outcome is set to the value most recently associated with
the name by Var\_Set.

\hypertarget{var_drop}{%
\subsection{Var\_Drop}\label{var_drop}}

Var \_Drop(Pool, Name, `0')

The function operates on the variable pool with the specified reference.
The name is a non-tailed name. If the specified name has the `exposed'
attribute then Var\_Drop operates on the variable pool referenced by
\#Upper in this pool and this rule is applied to that pool. When the
pool with attribute `not-exposed' for this name is determined the
attribute `dropped' is associated with the specified name. Also, when
the name is a stem, Var\_Drop(Pool,TailedName,`1') is executed for all
possible valid tailed names which have Name as astem.

Var \_Drop(Pool, Name, `1')

The function operates on the variable pool with the specified reference.
The name is a tailed name. The left part of the name, up to and
including the first period, is the stem. The stem is a non-tailed name.
If the specified stem has the `exposed' attribute then Var\_Drop
operates on the variable pool referenced by \#Upper in this pool and
this rule is applied to that pool. When the pool with attribute
`not-exposed' for the stem is determined the name is considered in that
pool. If the name has the `exposed' attribute then the variable pool
referenced by \#Upper in the pool is considered and this rule applied to
that pool. When the pool with attribute `not-exposed' is determined the
attribute `dropped' is associated with the specified name.

\hypertarget{var_expose}{%
\subsection{Var\_Expose}\label{var_expose}}

Var\_Expose (Pool, Name, `0')

The function operates on the variable pool with the specified reference.
The name is a non-tailed name. The attribute `exposed' is associated
with the specified name. Also, when the name is a stem,
Var\_Expose(Pool, TailedName,`1') is executed for all possible valid
tailed names which have Name as a stem.

Var\_Expose (Pool, Name, `1')

The function operates on the variable pool with the specified reference.
The name is a tailed name. The attribute `exposed' is associated with
the specified name.

\hypertarget{var_reset}{%
\subsection{Var\_Reset}\label{var_reset}}

Var\_ Reset (Pool)

The function operates on the variable pool with the specified reference.
It establishes the effect of subsequent API\_Next and API\_NextVariable
functions (see sections nnn and nnn). A Var\_Reset is implied by any
API\_ operation other than API\_Next and API\_NextVariable.

\hypertarget{symbols}{%
\section{Symbols}\label{symbols}}

For the syntax of a symbol see nnn.

The value of a symbol which is a NUMBER or a CONST\_SYMBOL which is not
a reserved symbol is the content of the appropriate token.

The value of a VAR\_SYMBOL which is ``taken as a constant'' is the
VAR\_SYMBOL itself, otherwise the VAR\_SYMBOL identifies a variable and
its value may vary during execution.

Accessing the value of a symbol which is not ``taken as a constant''
shall result in trace output, see nnn: if \#Tracing.\#Level == `I' then
call \#Trace Tag

where Tag is `\textgreater L\textgreater{}' unless the symbol is a
VAR\_SYMBOL which, when used as an argument to Var\_Value, does not
yield an indicator `D'. In that case, the Tag is
'\textgreater V\textgreater``.

\hypertarget{value-of-a-variable}{%
\section{Value of a variable}\label{value-of-a-variable}}

If VAR\_SYMBOL does not contain a period, or contains only one period as
its last character, the value of

the variable is the value associated with VAR\_SYMBOL in the variable
pool, that is \#Outcome after Var\_ Value (Pool, VAR SYMBOL, `0')

If the indicator is `D', indicating the variable has the `dropped'
attribute, the NOVALUE condition is raised; see nnn and nnn for
exceptions to this. \#Response = Var Value(Pool, VAR SYMBOL, `0') if
left(\#Response,1) == `D' then call \#Raise `NOVALUE', VAR\_SYMBOL, '\,'
If VAR\_SYMBOL contains a period which is not its last character, the
value of the variable is the value associated with the derived name.
7.3.1 Derived names A derived name is derived from a VAR\_SYMBOL as
follows:

VAR SYMBOL := Stem Tail

Stem := PlainSymbol `.'!

Tail i= (PlainSymbol \textbar{} `.' {[}PlainSymbol{]}) {[}`.'
{[}PlainSymbol1{]}{]}+ PlainSymbol := (general letter \textbar{} digit)+

The derived name is the concatenation of: - the Stem, without further
evaluation; - the Tail, with the PlainSymbols replaced by the values of
the symbols. The value of a PlainSymbol which does not start with a
digit is \#Outcome after Var\_ Value (Pool, PlainSymbol,`0') These
values are obtained without raising the NOVALUE condition.

If the indicator from the Var\_Value was not `D' then: if
\#Tracing.\#Level == `I' then call \#Trace
`\textgreater C\textgreater{}'

The value associated with a derived name is obtained from the variable
pool, that is \#Outcome after: Var\_Value(Pool,Derived Name,`1')

If the indicator is `D', indicating the variable has the `dropped'
attribute, the NOVALUE condition is raised; see nnn for an exception.

\hypertarget{value-of-a-reserved-symbol}{%
\subsection{Value of a reserved
symbol}\label{value-of-a-reserved-symbol}}

The value of a reserved symbol is the value of a variable with the
corresponding name in the reserved pool, see nnn.

\hypertarget{expressions-and-operators}{%
\section{Expressions and operators}\label{expressions-and-operators}}

Add a load of string coercions. Equality can operate on non-strings.
What if one operand non-string?

\hypertarget{the-value-of-a-term}{%
\subsection{The value of a term}\label{the-value-of-a-term}}

See nnn for the syntax of a ferm.

The value of a STRING is the content of the token; see nnn.

The value of a function is the value it returns, see nnn.

If a termis a symbol or STRING then the value of the term is the value
of that symbol or STRING.

If a term contains an expr\_alias the value of the term is the value of
the expr\_alias, see nnn.

\hypertarget{the-value-of-a-prefix_expression}{%
\subsection{The value of a
prefix\_expression}\label{the-value-of-a-prefix_expression}}

If the prefix\_expression is a term then the value of the
prefix\_expression is the value of the ferm, otherwise let rhs be the
value of the prefix\_expression within it\_\_ see nnn

If the prefix\_expression has the form `+' prefix\_expression then a
check is made: if datatype(rhs)==`NUM' then call \#Raise `SYNTAX',41.3,
rhs, `+'

and the value is the value of (0 + rhs).

If the prefix\_expression has the form `-' prefix\_expression then a
check is made: if datatype(rhs)==`NUM' then call \#Raise
`SYNTAX',41.3,rhs, `-'

and the value is the value of (0 - rhs).

If a prefix\_expression has the form not prefix\_expression then if rhs
== `0' then if rhs ==`1' then call \#Raise `SYNTAX', 34.6, not, rhs

See nnn for the value of the third argument to that \#Raise. If the
value of rhs is `0' then the value of the prefix\_expression value is
`1', otherwise it is `0'.

If the prefix\_expression is not a term then: if \#Tracing.\#Level ==
`I' then call \#Trace `\textgreater P\textgreater{}'

\hypertarget{the-value-of-a-power_expression}{%
\subsection{The value of a
power\_expression}\label{the-value-of-a-power_expression}}

See nnn for the syntax of a power\_expression.

If the power\_expression is a prefix\_expression then the value of the
power\_expression is the value of the prefix\_expression.

Otherwise, let Ihs be the value of power\_expression within it, and rhs
be the value of prefix\_expression within it.

\begin{verbatim}
if datatype(lhs)\=='NUM' then call #Raise 'SYNTAX',41.1,lhs,'**'

if \datatype(rhs,'W') then call #Raise 'SYNTAX',26.1,rhs,'**'
\end{verbatim}

The value of the power\_expression is

ArithOp(lhs,'**',rhs)

If the power\_expression is not a prefix\_expression then: if
\#Tracing.\#Level == `I' then call \#Trace
`\textgreater0O\textgreater{}'

\hypertarget{the-value-of-a-multiplication}{%
\subsection{The value of a
multiplication}\label{the-value-of-a-multiplication}}

See nnn for the syntax of a multiplication. If the multiplication is a
power\_expression then the value of the multiplication is the value of
the power\_expression. Otherwise, let Ihs be the value of multiplication
within it, and rns be the value of power\_expression within it. if
datatype(lhs)==`NUM' then call \#Raise `SYNTAX',41.1,lhs,multiplicative
operation if datatype(rhs)==`NUM' then call \#Raise
`SYNTAX',41.2,rhs,multiplicative operation

The value of the multiplication is ArithOp(lhs,multiplicative operation,
rhs)

If the multiplication is not a power\_expression then:

if \#Tracing.\#Level == `I' then call \#Trace
`\textgreater0O\textgreater{}'

\hypertarget{the-value-of-an-addition}{%
\subsection{The value of an addition}\label{the-value-of-an-addition}}

See nnn for the syntax of addition.

If the addition is a multiplication then the value of the addition is
the value of the multiplication. Otherwise, let Ihs be the value of
ad¢difion within it, and rhs be the value of the multiplication within
it. Let

operation be the adaltive\_operator. if datatype(lhs)==`NUM' then

call \#Raise `SYNTAX', 41.1, lhs, operation if datatype(rhs)==`NUM' then

call \#Raise `SYNTAX', 41.2, rhs, operation

If either of rhs or Ihs is not an integer then the value of the addition
is ArithOp(lhs, operation, rhs) Otherwise if the operation is `+' and
the length of the integer Ihs+rhs is not greater than \#Digits.\#Level

then the value of addition is lhs+rhs

Otherwise if the operation is `-' and the length of the integer Ihs-rhs
is not greater than \#Digits.\#Level then

the value of addition is lhs-rhs

Otherwise the value of the addition is ArithOp(lhs, operation, rhs)

If the addition is not a multiplication then: if \#Tracing.\#Level ==
`I' then call \#Trace `\textgreater0O\textgreater{}'

\hypertarget{the-value-of-a-concatenation}{%
\subsection{The value of a
concatenation}\label{the-value-of-a-concatenation}}

See nnn for the syntax of a concatenation. If the concatenation is an
addition then the value of the concatenation is the value of the
addition. Otherwise, let Ihs be the value of concatenation within it,
and rhs be the value of the additive\_expression within it. If the
concatenation contains `\textbar\textbar{}' then the value of the
concatenation will have the following characteristics:

\begin{itemize}
\item
  Config\_Length(Value) will be equal to
  Config\_Length(Ihs)+Config\_Length(rhs).
\item
  \#Outcome will be `equal' after each of:
\item
  Config\_Compare(Config\_Subsir(Ihs,n)\},Config\_Subsitr(Value,n)) for
  values of n not less than 1 and not more than Config\_Length(Ihs);
\item
  Config\_Compare(Config\_Subsir(rhs,n),Config\_Substr(Value,Config\_Length(Ihs)+n))
  for values of n not less than 1 and not more than Config\_Length(rhs).
  Otherwise the value of the concatenation will have the following
  characteristics:
\item
  Config\_Length(Value) will be equal to
  Config\_Length(Ihs)+1+Config\_Length(rhs).
\item
  \#Outcome will be `equal' after each of:
\item
  Config\_Compare(Config\_Subsir(Ihs,n)\},Config\_Subsitr(Value,n)) for
  values of n not less than 1 and not more than Config\_Length(Ihs);
\item
  Config\_Compare(' ',Config\_Substr(Value,Config\_Length(Ihs)\}+1));
\item
  Config\_Compare(Config\_Subsitr(rhs,n),Config\_Substr(Value,Config\_Length(Ins)+1+n))
  for values of n not less than 1 and not more than Config\_Length(rhs).
\end{itemize}

If the concatenation is not an addition then: if \#Tracing.\#Level ==
`I' then call \#Trace `\textgreater0O\textgreater{}'

\hypertarget{the-value-of-a-comparison}{%
\subsection{The value of a comparison}\label{the-value-of-a-comparison}}

See nnn for the syntax of a comparison.

If the comparison is a concatenation then the value of the comparison is
the value of the concatenation. Otherwise, let Ihs be the value of the
comparison within it, and rns be the value of the concatenation within
it.

If the comparison has a comparison\_operator that is a strict\_compare
then the variable \#Test is set as follows:

\#Test is set to `E'. Let Length be the smaller of Config\_Length(Ihs)
and Config\_Length(rhs). For values of n greater than O and not greater
than Length, if any, in ascending order, \#Test is set to the uppercased
first character of \#Outcome after:

Config\_Compare(Config\_Subsir(Ihs),Contfig\_Subsir(rhs)).

If at any stage this sets \#Test to a value other than `E' then the
setting of \#Test is complete. Otherwise, if Config\_Length(Ihs) is
greater than Config\_Length(rhs) then \#Test is set to `G' or if
Config\_Length(Ihs) is less than Config\_Length(rhs) then \#Test is set
to `L'.

If the comparison has a comparison\_operator that is a normal\_compare
then the variable \#Test is set as follows:

\begin{verbatim}
if datatype(lhs)\== 'NUM' | datatype(rhs)\== 'NUM' then do
/* Non-numeric non-strict comparison */
lhs=strip(lhs, 'B', ' ') /* ExtraBlanks not stripped */
rhs=strip(rhs, 'B', ' ')

if length(lhs)>length(rhs) then rhs=left (rhs, length (lhs) )
else lhs=left (lhs, length (rhs) )
if lhs>>rhs then #Test='G'
else if lhs<<rhs then #Test='L'
else #Test='E'

end
else do /* Numeric comparison */
if left(-lhs,1) == '-' & left(+rhs,1) \== '-' then #Test='G!
else if left(-rhs,1) == '-' & left(+lhs,1) \== '-' then #Test='L'
else do
Difference=lhs - rhs /* Will never raise an arithmetic condition. */
if Difference > 0 then #Test='G'
else if Difference < 0 then #Test='L'
else #Test='E'
end
end
The value of #Test, in conjunction with the operator in the comparison, determines the value of the
comparison.
The value of the comparison is '1' if
- #Test is 'E' and the operator is one of '="", '=="", '>=', <=", ‘\>', '\<', 'p>=', '<<=', \>>', or <<)

- #Test is 'G' and the operator is one of '>', '>=", ‘\<', ‘\=', '<>', '><', Nes", ‘>>! ‘p>’, or <<")
- #Test is 'L' and the operator is one of '<', <=", \>', \=', '<>', '><', \==', '<<', *<<=', or \>>'.
In all other cases the value of the comparison is '0'.

If the comparison is not a concatenation then:
if #Tracing.#Level == 'I' then call #Trace '>0O>'
\end{verbatim}

\hypertarget{the-value-of-an-and_expression}{%
\subsection{The value of an
and\_expression}\label{the-value-of-an-and_expression}}

See nnn for the syntax of an and\_expression.

If the and\_expression is a comparison then the value of the
and\_expression is the value of the comparison.

Otherwise, let Ihs be the value of the and\_expression within it, and
rhs be the value of the comparison within it.

if lhs == `0' then if lhs == `1' then call \#Raise
`SYNTAX',34.5,lhs,`\&'

if rhs == `0' then if rhs == `1' then call \#Raise
`SYNTAX',34.6,rhs,`\&'

Value=`0'

if lhs == `1' then if rhs == `1' then Value=`1'

If the and\_expression is not a comparison then:

if \#Tracing.\#Level == `I' then call \#Trace
`\textgreater0O\textgreater{}'

\hypertarget{the-value-of-an-expression}{%
\subsection{The value of an
expression}\label{the-value-of-an-expression}}

See nnn for the syntax of an expression.

The value of an expression, or an expr, is the value of the expr\_alias
within it.

If the expr\_alias is an and\_expression then the value of the
expr\_alias is the value of the and\_expression. Otherwise, let Ihs be
the value of the expr\_alias within it, and rhs be the value of the
and\_expression

within it. if lhs == `0' then if lhs == `1' then

call \#Raise `SYNTAX',34.5,lhs,or operator if rhs == `0' then if rhs ==
`1' then

call \#Raise `SYNTAX',34.6,rhs,or operator Value=`1' if lhs == `0' then
if rhs == `0' then Value=`0' If the or\_operator is `\&\&' then if lhs
== `1' then if rhs == `1' then Value=`0' If the expr\_alias is not an
and\_expression then: if \#Tracing.\#Level == `I' then call \#Trace
`\textgreater0O\textgreater{}'

The value of an expression or expr shall be traced when
\#Tracing.\#Level is `R'. The tag is `\textgreater=\textgreater{}' when

the value is used by an assignment and
`\textgreater\textgreater\textgreater{}' when it is not. if
\#Tracing.\#Level == `R' then call \#Trace Tag

\hypertarget{arithmetic-operations}{%
\subsection{Arithmetic operations}\label{arithmetic-operations}}

The user of this standard is assumed to know the results of the binary
operators `+' and `-' applied to signed or unsigned integers.

The code of ArithOp itself is assumed to operate under a sufficiently
high setting of numeric digits to avoid exponential notation.

\begin{verbatim}
ArithoOp:

arg Numberl, Operator, Number2
/* The Operator will be applied to Numberl and Number2 under the numeric
settings #Digits.#Level, #Form.#Level, #Fuzz.#Level */

/* The result is the result of the operation, or the raising of a 'SYNTAX' or
'LOSTDIGITS' condition. */

/* Variables with digit 1 in their names refer to the first argument of the
operation. Variables with digit 2 refer to the second argument. Variables
with digit 3 refer to the result. */

/* The quotations and page numbers are from the first reference in
Annex C of this standard. */

/* The operands are prepared first. (Page 130) Function Prepare does this,
separating sign, mantissa and exponent. */

v = Prepare (Numberl, #Digits.#Level)
parse var v Signl Mantissal Exponentl

v = Prepare (Number2, #Digits.#Level)
parse var v Sign2 Mantissa2 Exponent2

/* The calculation depends on the operator. The routines set Sign3
Mantissa3 and Exponent3. */

Comparator = ''

select
when Operator == '*' then call Multiply

when Operator
when Operator

' then call DivType
*' then call Power
when Operator %' then call DivType
when Operator '//" then call DivType
otherwise call AddSubComp

end

call PostOp /* Assembles Number3 */

if Comparator \== '' then do
/* Comparison requires the result of subtraction made into a logical */
/* value. */
t = '0'
select
when left (Number3,1) == '-' then
if wordpos(Comparator,'< <= <> >< \= \>') > 0 then t = '1'
when Number3 \== '0' then
if wordpos(Comparator,'> >= <> >< \= \<') > 0 then t = '1'
otherwise
if wordpos(Comparator,'>= = =< \< \>') > 0 then t = '1!
end
Number3 = t
end
return Number3 /* From ArithOp */
/* Activity before every operation: */
Prepare: /* Returns Sign Mantissa and Exponent */
/* Preparation of operands, Page 130 */
/* ",...terms being operated upon have leading zeros removed (noting the

position of any decimal point, and leaving just one zero if all the digits in

the number are zeros) and are then truncated to DIGITS+1 significant digits
(if necessary)..." */

arg Number, Digits

/* Blanks are not significant. */

/* The exponent is separated */

parse upper value space(Number,0) with Mantissa 'E' Exponent
if Exponent == '' then Exponent = '0'

/* The sign is separated and made explicit. */

Sign = '+' /* By default */
if left(Mantissa,1) == '-' then Sign = '-'
if verify (left (Mantissa,1),'+-') = 0 then Mantissa = substr(Mantissa,2)

/* Make the decimal point implicit; remove any actual Point from the
Mantissa. */
Pp = pos('.',Mantissa)
if p > 0 then Mantissa = delstr(Mantissa,p,1)
else p = 1+length (Mantissa)

/* Drop the leading zeros */
do q = 1 to length(Mantissa) - 1

if substr(Mantissa,g,1) \== '0' then leave
p=p-l
end q

Mantissa = substr(Mantissa,q)

/* Detect if Mantissa suggests more significant digits than DIGITS
caters for. */
do j = Digits+1 to length (Mantissa)
if substr(Mantissa,j,1) \== '0' then call #Raise 'LOSTDIGITS', Number
end j

/* Combine exponent with decimal point position, Page 127 */
/* "Exponential notation means that the number includes a power of ten

following an 'E' that indicates how the decimal point will be shifted. Thus

4E9 is just a shorthand way of writing 4000000000 " */

/* Adjust the exponent so that decimal point would be at right of
the Mantissa. */
Exponent = Exponent - (length(Mantissa) - p + 1)

/* Truncate if necessary */

t = length(Mantissa) - (Digits+1)

if t > 0 then do
Exponent = Exponent + t
Mantissa = left (Mantissa,Digits+1)
end

if Mantissa == '0' then Exponent = 0

return Sign Mantissa Exponent

/* Activity after every operation. */
/* The parts of the value are composed into a single string, Number3. */
PostOp:

/* Page 130 */
/* 'traditional' rounding */
t = length(Mantissa3) - #Digits.#Level
if t > 0 then do
/* 'traditional' rounding */
Mantissa3 = left (Mantissa3,#Digits.#Level+1) + 5
if length(Mantissa3) > #Digits.#Level+1 then
/* There was 'carry' */

Exponent3 = Exponent3 + 1

Mantissa3 = left (Mantissa3,#Digits.#Level)

Exponent3 = Exponent3 + t

end
/* "A result of zero is always expressed as a single character '0' "*/
if verify (Mantissa3,'0') = 0 then Number3 = '0'
else do

if Operator == '/' | Operator == '**' then do

/* Page 130 "For division, insignificant trailing zeros are removed
after rounding." */

/* Page 133 "... insignificant trailing zeros are removed." */
do q = length(Mantissa3) by -1 to 2
if substr(Mantissa3,q,1) \== '0' then leave
Exponent3 = Exponent3 + 1
end q
Mantissa3 = substr(Mantissa3,1,q)
end
if Floating() == 'E' then do /* Exponential format */

Exponent3 = Exponent3 + (length(Mantissa3)-1)

/* Page 136 "Engineering notation causes powers of ten to be expressed as a

multiple of 3 - the integer part may therefore range from 1 through
999." */

g=l

if #Form.#Level == 'E' then do

/* Adjustment to make exponent a multiple of 3 */
g = Exponent3//3 /* Recursively using ArithOp as
an external routine. */
if g < 0 then g =g+ 3
Exponent3 = Exponent3 - g
geaqgitl
if length(Mantissa3) < g then
Mantissa3 = left (Mantissa3,g,'0')
end /* Engineering */

/* Exact check on the exponent. */
if Exponent3 > #Limit ExponentDigits then

call #Raise 'SYNTAX', 42.1, Numberl, Operator, Number2
if -#Limit ExponentDigits > Exponent3 then

call #Raise 'SYNTAX', 42.2, Numberl, Operator, Number2

/* Insert any decimal [point. */

if length(Mantissa3) \= g then Mantissa3 = insert('.',Mantissa3,g)
/* Insert the E */
if Exponent3 >= 0 then Number3
else Number3
end /* Exponent format */
else do /* 'pure number' notation */
p = length(Mantissa3) + Exponent3 /* Position of the point within
Mantissa */
/* Add extra zeros needed on the left of the point. */
if p < 1 then do
Mantissa3 = copies('0',1 - p)| |Mantissa3
p=il
end
/* Add needed zeros on the right. */
if p > length(Mantissa3) then
Mantissa3 = Mantissa3||copies('0',p-length (Mantissa3) )
/* Format with decimal point. */
Number3 = Mantissa3

Mantissa3'E+'Exponent3
Mantissa3'E'Exponent3

if p < length(Number3) then Number3 = insert('.',Mantissa3,p)
else Number3 = Mantissa3
end /* pure */
if Sign3 == '-' then Number3 = '-'Number3
end /* Non-Zero */
return
/* This tests whether exponential notation is needed. */
Floating:
/* The rule in the reference has been improved upon. */
Ct = ter
if Exponent3+length(Mantissa3) > #Digits.#Level then t = 'E'
if length(Mantissa3) + Exponent3 < -5 then t = 'E'
return t
/* Add, Subtract and Compare. */

AddSubComp: /* Page 130 */
/* This routine is used for comparisons since comparison is
defined in terms of subtraction. Page 134 */
/* "Numeric comparison is affected by subtracting the two numbers (calculating

the difference) and then comparing the result with '0'." */
NowDigits = #Digits.#Level
if Operator \=='+' & Operator \== '-' then do

Comparator = Operator

/* Page 135 "The effect of NUMERIC FUZZ is to temporarily reduce the value
of NUMERIC DIGITS by the NUMERIC FUZZ value for each numeric comparison" */
NowDigits = NowDigits - #Fuzz.#Level

end
/* Page 130 "If either number is zero then the other number ... is used as
the result (with sign adjustment as appropriate). */
if Mantissa2 == '0' then do /* Result is the lst operand */
Sign3=Signl; Mantissa3 = Mantissal; Exponent3 = Exponentl
return ''
end

if Mantissal
Sign3 = Sign

== '0' then do /* Result is the 2nd operand */
2
if Operator \

; Mantissa3 = Mantissa2; Exponent3 = Exponent2
== '+' then if Sign3 = '+' then Sign3 rat
else Sign3

bat
return ''
end

/* The numbers may need to be shifted into alignment. */

/* Change to make the exponent to reflect a decimal point on the left,

so that right truncation/extension of mantissa doesn't alter exponent. */
Exponentl = Exponentl + length (Mantissal1)
Exponent2 = Exponent2 + length (Mantissa2)

/* Deduce the implied zeros on the left to provide alignment. */

Aligni = 0

Align2 = Exponentl - Exponent2

if Align2 > 0 then do /* Arg 1 provides a more significant digit */
Align2 = min(Align2,NowDigits+1) /* No point in shifting further. */
/* Shift to give Arg2 the same exponent as Argl */

Mantissa2 = copies('0',Align2) || Mantissa2
Exponent2 = Exponentl
end

if Align2 < 0 then do /* Arg 2 provides a more significant digit */
/* Shift to give Argl the same exponent as Arg2 */

Aligni = -Align2

Alignl = min(Align1l,NowDigits+1) /* No point in shifting further. */
Align2 = 0

Mantissal = copies('0',Alignl) || Mantissal

Exponentl = Exponent2

end

/* Maximum working digits is NowDigits+1. Footnote 41. */

SigDigits
SigDigits

max (length (Mantissal) , length (Mantissaz2) )
min (SigDigits,NowDigits+1)

/* Extend a mantissa with right zeros, if necessary. */
Mantissal = left (Mantissal,SigDigits,'0')

Mantissa2 = left (Mantissa2,SigDigits,'0')

/* The exponents are adjusted so that

the working numbers are integers, ie decimal point on the right. */

Exponent3 = Exponentl-SigDigits
Exponentl = Exponent3
Exponent2 = Exponent3
if Operator = '+' then
Mantissa3 = (Signl || Mantissal) + (Sign2 || Mantissa2)

else Mantissa3 (Signl || Mantigsal) - (Sign2 || Mantissa2)

/* Separate the sign */
if Mantissa3 < 0 then do

Sign3 = '-'
Mantissa3 = substr (Mantissa3,2)
end

else Sign3 = '+'

/* "The result is then rounded to NUMERIC DIGITS digits if necessary,
taking into account any extra (carry) digit on the left after addition,
but otherwise counting from the position corresponding to the most
significant digit of the terms being added or subtracted." */

if length(Mantissa3) > SigDigits then SigDigits = SigDigits+l
d = SigDigits - NowDigits /* Digits to drop. */
if d <= 0 then return
t = length(Mantissa3) - d /* Digits to keep. */
/* Page 130. "values of 5 through 9 are rounded up, values of 0 through 4 are
rounded down." */
if t > 0 then do
/* 'traditional' rounding */
Mantissa3 = left(Mantissa3, t +1) +5
if length(Mantissa3) > t+1 then
/* There was 'carry' */
/* Keep the extra digit unless it takes us over the limit. */
if t < NowDigits then t = t+l
else Exponent3 = Exponent3+1
Mantissa3 = left (Mantissa3,t)
Exponent3 = Exponent3 + d
end /* Rounding */
else Mantissa3 = '0'
return /* From AddSubComp */

/* Multiply operation: */

Multiply: /* p 131 */

/* Note the sign of the result */

if Signl == Sign2 then Sign3 = '+'
else Sign3 = '-'
/* Note the exponent */
Exponent3 = Exponentl + Exponent2
if Mantissal == '0' then do
Mantissa3 = '0'
return
end
/* Multiply the Mantissas */
Mantissa3 = ''

do q=1 to length (Mantissa2)
Mantissa3 = Mantissa3'0'
do substr(Mantissa2,q,1)
Mantissa3 = Mantissa3 + Mantissal
end
end q
return /* From Multiply */

/* Types of Division: */

DivType: /* p 131 */
/* Check for divide-by-zero */

if Mantissa2 == '0' then call #Raise 'SYNTAX',

/* Note the exponent of the result */
Exponent3 = Exponentl - Exponent2

/* Compute (one less than) how many digits will be in the integer

part of the result. */

IntDigits = length(Mantissal) - Length(Mantissa2) + Exponent3
/* In some cases, the result is known to be zero.
if Mantigsal = 0 | (IntDigits < 0 & Operator

Mantissa3 = 0
Sign3 = '+'
Exponent3 = 0
return
end
/* In some cases, the result is known to be to be the first argument.
if IntDigits < 0 & Operator == '//' then do
Mantissa3 = Mantissal
Sign3 = Signl
Exponent3 = Exponentl
return
end
/* Note the sign of the result. */
if Signl == Sign2 then Sign3 = '+'
else Sign3 = '-'

/* Make Mantissal at least as large as Mantissa2 so Mantissa2 can be
subtracted without causing leading zero to result. Page 131 */

az 0
do while Mantissa2 > Mantissal

Mantissal = Mantissal'0'
Exponent3 = Exponent3 - 1
aztadtl
end
/* Traditional divide */
Mantissa3 = ''

/* Subtract from part of Mantissal that has length of Mantissa2 */

left (Mantissal,length(Mantissa2) )
substr (Mantissal, length (Mantissa2)+1)
o forever

x
Y
d

/* Develop a single digit in z by repeated subtraction.

ze=O0
do forever
xX = kK - Mantissa2
if left(x,1) == '-' then leave
Zeze¢t+tl
end

x = x + Mantissa2 /* Recover from over-subtraction */
/* The digit becomes part of the result */

Mantissa3 = Mantissa3 || z

if Mantissa3 == '0' then Mantissa3 = '' /* A single leading

##

*f
'%') then do

*/
zero can happen. */
/* x||y is the current residue */
if y == '' then if x = 0 then leave /* Remainder is zero */

if length(Mantissa3) > #Digits.#Level then leave /* Enough digits
in the result */

/* Check type of division */
if Operator \== '/' then do
if IntDigits = 0 then leave
IntDigits = IntDigits - 1
end
/* Prepare for next digit */
/* Digits come from y, until that is exhausted. */
/* When y is exhausted an extra zero is added to Mantissal */
if y == '' then do
y = ror
Exponent3 = Exponent3 - 1
aztadtl
end
xX = xX | | left (y,1)
y = substr(y,2)
end /* Iterate for next digit. */

Remainder = x || y
Exponent3 = Exponent3 + length(y) /* The loop may have been left early.
/* Leading zeros are taken off the Remainder. */
do while length(Remainder) > 1 & Left (Remainder,1) == '0'
Remainder = substr (Remainder, 2)
end
if Operator \== '/' then do
/* Check whether % would fail, even if operation is // */

if Floating() 'E' then do
if Operator '%' then MsgNum
else MsgNum

/* Page 133. % could fail by needing exponential notation */

26.11
26.12

call #Raise 'SYNTAX', MsgNum, Numberl , Number2, #Digits.#Level

end
end
if Operator == '//' then do
/* We need the remainder */
Sign3 = Signl

Mantissa3 = Remainder
Exponent3 = Exponentl - a
end

return /* From DivType */

/* The Power operation: */

Power: /* page 132 */
/* The second argument should be an integer */
if \WholeNumber2() then call #Raise 'SYNTAX', 26.8, Number2
/* Lhs to power zero is always 1 */
if Mantissa2 == '0' then do
Sign3 = '+'
Mantissa3
Exponent3
return
end

i
ror

/* Pages 132-133 The Power algorithm */
Rhs = left (Mantissa2,length(Mantissa2)+Exponent2,'0')/* Explicit
integer form */
L length (Rhs)
b X2B(D2X(Rhs)) /* Makes Rhs in binary notation */
/* Ignore initial zeros */
do q=l1byl
if substr(b,q,1) \== '0' then leave
end q
ael
do forever
/* Page 133 "Using a precision of DIGITS+L+1" */
if substr(b,q,1) == '1' then do
a = Recursion('*',Signl || Mantissal'E'Exponent1)

*/
if left(a,2) == 'MN' then signal PowerFailed
end

/* Check for finished */

if q = length(b) then leave

/* Square a */

a = Recursion('*',a)
if left(a,2) == 'MN' then signal PowerFailed
q=qrtil
end
/* Divide into one for negative power */
if Sign2 == '-' then do
Sign2 = '+'
a = Recursion('/')
if left(a,2) == 'MN' then signal PowerFailed
end

/* Split the value up so that PostOp can put it together with rounding */
Parse value Prepare(a,#Digits.#Level+L+1) with Sign3 Mantissa3 Exponent3
return

PowerFailed:
/* Distinquish overflow and underflow */
ReWas = substr(a,4)
if Sign2 = '-' then if ReWas == '42.1' then RcWas
else RcWas
call #Raise 'SYNTAX', RcWas, Numberl, '**', Number2
/* No return */

"42.2!
"42.1!

WholeNumber2:
numeric digits Digits
if #Form.#Level == 'S' then numeric form scientific

else numeric form engineering
return datatype (Number2, 'W')

Recursion: /* Called only from '**! */
numeric digits #Digits.#Level + L + 1
signal on syntax name Overflowed
/* Uses ArithOp again under new numeric settings. */
if arg(1) == '/' then return 1l/a
else return a * arg(2)
Over flowed:
return 'MN '.MN
\end{verbatim}

\hypertarget{functions}{%
\section{Functions}\label{functions}}

\hypertarget{invocation}{%
\subsection{Invocation}\label{invocation}}

Invocation occurs when a function or a message\_term or a callis
evaluated. Invocation of a function may result in a value, in which
case:

if \#Tracing.\#Level == `I' then call \#Trace
`\textgreater F\textgreater{}' Invocation of a message\_term may result
in a value, in which case: if \#Tracing.\#Level == `I' then call \#Trace
`\textgreater M\textgreater{}'

\hypertarget{evaluation-of-arguments}{%
\subsection{Evaluation of arguments}\label{evaluation-of-arguments}}

The argument positions are the positions in the exoression\_list where
syntactically an expression occurs or could have occurred. Let ArgNumber
be the number of an argument position, counting from 1 at the left; the
range of ArgNumber is all whole numbers greater than zero.

For each value of ArgNumber, \#ArgExists.\#NewLevel.ArgNumber is set `1'
if there is an expression present, `O' if not.

From the left, if \#ArgExists.\#NewLevel.ArgNumber is `1' then
\#Arg.\#NewLevel.ArgNumber is set to the value of the corresponding
expression. If \#ArgExists.\#NewLevel.ArgNumber is `0' then
\#Arg.\#NewLevel.ArgNumber is set to the null string.

\#ArgExists.\#NewLevel.0 is set to the largest ArgNumber for which
\#ArgExists.\#NewLevel.ArgNumber is `1', or to zero if there is no such
value of ArgNumber.

\hypertarget{the-value-of-a-label}{%
\subsection{The value of a label}\label{the-value-of-a-label}}

The value of a LABEL, or of the taken\_constant in the function or
call\_instruction, is taken as a constant, see nnn. If the
taken\_constant is not a string\_literal it is a reference to the first
LABEL in the program which has the same value. The comparison is made
with the `==' operator.

If there is such a matching label and the label is trace-only (see nnn)
then a condition is raised: call \#Raise `SYNTAX', 16.3, taken constant

If there is such a matching label, and the label is not trace-only,
execution continues at the label with routine initialization (see nnn).
This is execution of an internal routine.

If there is no such matching label, or if the taken\_constant is a
string\_literal, further comparisons are made.

If the value of the taken\_constant matches the name of some built-in
function then that built-in function is invoked. The names of the
built-in functions are defined in section nnn and are in uppercase.

If the value does not match any built-in function name,
Config\_ExternalRoutine is used to invoke an external routine.

Whenever a matching label is found, the variables SIGL and .SIGL are
assigned the value of the line number of the clause which caused the
search for the label. In the case of an invocation resulting from a

condition occurring that shall be the clause in which the condition
occurred. Var \_ Set(\#Pool, `SIGL', `0', \#LineNumber) var Set(0 ,
`.SIGL', `0', \#LineNumber)

The name used in the invocation is held in \#Name.\#Level for possible
use in an error message from the RETURN clause, see nnn

\hypertarget{the-value-of-a-function}{%
\subsection{The value of a function}\label{the-value-of-a-function}}

A built-in function completes when it returns from the activity defined
in section nnn. The value of a built-in function is defined in section
nnn.

An internal routine completes when \#Level returns to the value it had
when the routine was invoked. The value of the internal function is the
value of the expression on the return which completed the routine. The
value of an external function is determined by Config\_ExternalRoutine.

\hypertarget{the-value-of-a-method}{%
\subsection{The value of a method}\label{the-value-of-a-method}}

A built-in method completes when it returns from the activity defined in
section n.~The value of a built-in method is defined in section n.

An internal method completes when \#Level returns to the value it had
when the routine was invoked. The value of the internal method is the
value of the expression on the return which completed the method. The
value of an external method is determined by Config\_ExternalMethod.

\hypertarget{the-value-of-a-message-term}{%
\subsection{The value of a message
term}\label{the-value-of-a-message-term}}

See nnn for the syntax of a message\_term. The value of the ferm within
a message\_term is called the receiver.

The receiver and any arguments of the term are evaluated, in left to
right order. r= \#evaluate(message term, term) If the message term
contains `\textasciitilde\textasciitilde{}' the value of the message
term is the receiver. Any effect on .Result? Otherwise the value of a
message\_term is the value of the method it invokes. The method invoked
is determined by the receiver and the taken\_constant and symbol. t =
\#Instance(message term, taken constant) If there is a symbol, it is
subject to a constraints. if \#contains (message term, symbol) then do
if r \textless\textgreater{} \#Self then

call \#Raise `SYNTAX', nn.n

/* OOI: ``Message search overrides can only be used from methods of the
target object.'' */

The search will progress from the object to its class and superclasses.
/* This is going to be circular because it describes the message lookup
algorithm and also uses messages. However for the messages in this code
the message names are chosen to be unique to a method so there is no
need to use this algorithm in deciding which method is intended. */

/* message term ::= receiver `\textasciitilde{}' taken constant `:'
VAR\_SYMBOL arguments */

/* This code reflects OOI - the arguments on the message don't affect
the method choice. */

/* This code selects a method based on its arguments, receiver,
taken\_constant, and symbol. */

/* This code is used in a context where \#Self is the receiver of the
method invocation which the subject message term is running under. */

\begin{verbatim}
SelectMethod:

/* If symbol given, receiver must be self. */
if arg(3,'E') then if arg(1)\==#Self then signal error /* syntax number? */

t arg(2) /* Will have been uppercased, unless a literal. */

x arg(1) /* Cursor through places to look for the method. */

Mixing 1 /* Off for potential mixins ignored because symbol given. */
Mixins -array~new /* to note any Mixins involved. */

/* Look in the method table of the object, if no 'symbol' given. */
if arg(3,'E') then do
Mixing = 0

end
else do
m = x~#MethodTable[t]
if m \== .nil then return m
end

do until x==.object
/* Follow the class hierarchy. */
x = x-class
/* Note any mixins for later reference. */
Mix = x~Inherited /* An array, ordered as the directive left-to-right. */

if Mix \== .nil then /* Append to the record. */
do j=1 to Mix~dimension (1)
Mixins [Mixins~dimension(1)+1] = Mix[j]
end

if Mixing do
/* Consider mixins only for superclasses of 'symbol'. */
do j=1 to Mixins~dimension (1)
/* Look at the baseclass of each. */
/* That is closest superclass not a mixin. */
s = Mixins[j]~class
do while s~Mixin /* Assert stop at .object if not before. */
s=s~class
end
if s==x then do
m=Mixins [j]~#MethodTable[t]
if m \== .nil then return m
end
end j
end /* Mixing */
if arg(3,'E') then if arg(3)==x then do
Mixing=1
end
if Mixing do
/* Consider non-Mixins */
m= x-#InstanceMethodTable[t]

if m \== .nil then return m
end

x=x~superclass

end

/* Try for UNKNOWN instead */
if t == 'UNKNOWN' then return .nil
if \arg(3,'E') then return SelectMethod arg(1),'UNKNOWN'
else return SelectMethod arg(1),'UNKNOWN',arg(3)
\end{verbatim}

\hypertarget{use-of-config_externalroutine}{%
\subsection{Use of
Config\_ExternalRoutine}\label{use-of-config_externalroutine}}

The values of the arguments to the use of Config\_ExternalRoutine, in
order, are:

The argument How is `SUBROUTINE' if the invocation is from a call,
```FUNCTION' if the invocation is from a function.

The argument NameType is `1' if the taken\_constant is a
string\_literal, `0' otherwise.

The argument Name is the value of the faken\_constant.

The argument Environment is the value of this argument on the API\_
Start which started this execution. The argument Arguments is the \#Arg.
and \#ArgExists. data.

The argument Streams is the value of this argument on the API\_Start
which started this execution.

The argument Traps is the value of this argument on the API\_Start which
started this execution. Var\_Reset is invoked and \#API\_Enabled set to
`1' before use of Config\_ExternalRoutine. \#API\_Enabled is set to `O'
after.

The response from Config\_ExternalRoutine is processed. If no conditions
are (implicitly) raised, \#Outcome is the value of the function.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/incompatibilities.tex
complete, but very deficient, extended standard

\hypertarget{incompatibilities}{%
\chapter{Incompatibilities}\label{incompatibilities}}

\hypertarget{the-incompatibilities-from-classic-rexx-are}{%
\chapter{The incompatibilities from Classic Rexx
are:}\label{the-incompatibilities-from-classic-rexx-are}}

Assignment of compound variables, as in ABC. = PQR., is an assignment of
references so that ABC. subsequently refers to the same object as PQR.,
as opposed to making the default value of ABC. that of PQR.. This change
was necessary to fit compound variables into the object framework, in
particular allowing USE ARG to handle compound variables as by-reference
parameters. The first reference of Annex B discouraged use of this
construct in Classic Rexx programs. ``Breakage'' of programs due to this
incompatibility is rare.

Also something in condition handling that I don't know the reason for.

\hypertarget{call}{%
\section{Call}\label{call}}

The call instruction has been extended to allow for a computed name of
the callee. Syntax considerations prevent a similar thing being done for
functions.

\hypertarget{concurrency}{%
\section{Concurrency}\label{concurrency}}

Meet 17 minutes

\hypertarget{guard}{%
\section{Guard}\label{guard}}

Meet 17 minutes

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/instructions.tex
complete, but very deficient, extended standard

\hypertarget{instructions}{%
\chapter{Instructions}\label{instructions}}

This clause describes the execution of instructions, and how the
sequence of execution can vary from the normal execution in order of
appearance in the program.

Execution of the program begins with its first clause.

If we left Routine initialization to here we can leave method
initialization. \#\# Method initialization

There is a pool for local variables.

call Config ObjectNew

\#Po00ol = \#Outcome

Set self and super

\hypertarget{routine-initialization}{%
\section{Routine initialization}\label{routine-initialization}}

If the routine is invoked as a function, \#lsFunction.\#NewLevel shall
be set to `1', otherwise to `0O'; this affects the processing of a
subsequent RETURN instruction.

\#AllowProcedure.\#NewLevel = `1'

Many of the initial values for a new invocation are inherited from the
caller's values. \#Digits.\#NewLevel = \#Digits.\#Level

\#Form.\#NewLevel = \#Form.\#Level

\#Fuzz.\#NewLevel = \#Fuzz.\#Level

\#StartTime.\#NewLevel = \#StartTime.\#Level

\#Tracing.\#NewLevel = \#Tracing.\#Level \#Interactive.\#NewLevel =
\#Interactive.\#Level

call EnvAssign ACTIVE, \#NewLevel, ACTIVE, \#Level call EnvAssign
ALTERNATE, \#NewLevel, ALTERNATE, \#Level

do t=1 to 7 Condition = word(`SYNTAX HALT ERROR FAILURE NOTREADY NOVALUE
LOSTDIGITS',t) \#Enabling.Condition.\#NewLevel =
\#Enabling.Condition.\#Level

\#Instruction.Condition.\#NewLevel = \#Instruction.Condition.\#Level
\#TrapName.Condition.\#NewLevel = \#TrapName.Condition.\#Level
\#EventLevel.Condition.\#NewLevel = \#EventLevel.Condition.\#Level end t

If this invocation is not caused by a condition occurring, see nnn, the
state variables for the CONDITION

built-in function are copied. \#Condition.\#NewLevel =
\#Condition.\#Level \#ConditionDescription.\#NewLevel =
\#ConditionDescription.\#Level \#ConditionExtra.\#NewLevel =
\#ConditionExtra.\#Level \#ConditionInstruction.\#NewLevel =
\#ConditionInstruction.\#Level Execution of the initialized routine
continues at the new level of invocation. \#Level = \#NewLevel
\#NewLevel = \#Level + 1

\hypertarget{clause-initialization}{%
\section{Clause initialization}\label{clause-initialization}}

The clause is traced before execution: if pos(\#Tracing.\#Level, `AIR')
\textgreater{} 0 then call \#TraceSource

The time of the first use of DATE or TIME will be retained throughout
the clause.

\#ClauseTime.\#Level = '\,' The state variable \#LineNumber is set to
the line number of the clause, see nnn. A clause other than a null
clause or label or procedure instruction sets:

\#AllowProcedure.\#Level = `0' /* See message 17.1 */

\hypertarget{clause-termination}{%
\section{Clause termination}\label{clause-termination}}

if \#InhibitTrace \textgreater{} 0 then \#InhibitTrace = \#InhibitTrace
- 1 Polling for a HALT condition occurs:

\#Response = Config Halt Query ()

if \#0utcome == `Yes' then do call Config Halt Reset

call \#Raise `HALT', substr(\#Response,2) /* May return */ end

At the end of each clause there is a check for conditions which occurred
and were delayed. It is acted on

if this is the clause in which the condition arose. do t=1 to 4
\#Condition=WORD(`HALT FAILURE ERROR NOTREADY',t) /* HALT can be
established during HALT handling. */ do while
\#PendingNow.\#Condition.\#Level \#PendingNow.\#Condition.\#Level = `0'
call \#Raise end end

Interactive tracing may be turned on via the configuration. Only a
change in the setting is significant. call Config Trace Query

if \#AtPause = 0 \& \#Outcome == `Yes' \& \#Trace QueryPrior == `No'
then do /* External request for Trace `?R!' */ \#Interactive.\#Level =
`1' \#Tracing.\#Level = `R' end

\#TraceQueryPrior = \#Outcome

Tracing just not the same with NetRexx.

When tracing interactively, pauses occur after the execution of each
clause except for CALL, DO the second or subsequent time through the
loop, END, ELSE, EXIT, ITERATE, LEAVE, OTHERWISE, RETURN, SIGNAL, THEN
and null clauses.

If the character `=' is entered in response to a pause, the prior clause
is re-executed.

Anything else entered will be treated as a string of one or more clauses
and executed by the language processor. The same rules apply to the
contents of the string executed by interactive trace as do for strings
executed by the INTERPRET instruction. If the execution of the string
generates a syntax error, the standard message is displayed but no
condition is raised. All condition traps are disabled during execution
of the string. During execution of the string, no tracing takes place
other than error or failure return codes from commands. The special
variable RC is not set by commands executed within the string, nor is
.RC.

If a TRACE instruction is executed within the string, the language
processor immediately alters the trace setting according to the TRACE
instruction encountered and leaves this pause point. If no TRACE
instruction is executed within the string, the language processor simply
pauses again at the same point in the program.

At a pause point: if \#AtPause = 0 \& \#Interactive.\#Level \&
\#InhibitTrace = 0 then do if \#InhibitPauses \textgreater{} 0 then
\#InhibitPauses = \#InhibitPauses-1 else do \#TraceInstruction = `0' do
forever

call Config Trace Query

if \#Outcome == `No' \& \#Trace QueryPrior == `Yes' then do /* External
request to stop tracing. */ \#Trace\_QueryPrior=\#Outcome
\#Interactive.\#Level = `0' \#Tracing.\#Level = `N' leave end

if \#Outcome == `Yes' \& \#Trace QueryPrior == `No' then do /* External
request for Trace `?R!' */ \#Trace QueryPrior = \#Outcome
\#Interactive.\#Level = `1' \#Tracing.\#Level = `R' leave end

if \#Interactive.\#Level \textbar{} \#TraceInstruction then leave

/* Accept input for immediate execution. */

call Config Trace Input

if length(\#0utcome) = 0 \textbar{} \#0Outcome == `=' then leave
\#AtPause = \#Level

interpret \#Outcome

\#AtPause = 0 end /* forever loop \emph{/ if \#Outcome == `=' then call
\#Retry /} With no return */ end end

\hypertarget{instruction}{%
\section{Instruction}\label{instruction}}

\hypertarget{address}{%
\subsection{ADDRESS}\label{address}}

For a definition of the syntax of this instruction, see nnn.

An external environment to which commands can be submitted is identified
by an environment name. Environment names are specified in the ADDRESS
instruction to identify the environment to which a command should be
sent.

I/O can be redirected when submitting commands to an external
environment. The submitted command's input stream can be taken from an
existing stream or from a set of compound variables with a common stem.
In the latter case (that is, when a stem is specified as the source for
the commands input stream) whole number tails are used to order input
for presentation to the submitted command. Stem.0 must contain a whole
number indicating the number of compound variables to be presented, and
stem. 1 through stem.n (where n=stem.0) are the compound variables to be
presented to the submitted command.

Similarly, the submitted command's output stream can be directed to a
stream, or to a set of compound variables with a given stem. In the
latter case (i.e., when a stem is specified as the destination) compound
variables will be created to hold the standard output, using whole
number tails as described above. Output redirection can specify a
REPLACE or APPEND option, which controls positioning prior to the
command's execution. REPLACE is the default.

I/O redirection can be persistently associated with an environment name.
The term ``environment'' is used to refer to an environment name
together with the I/O redirections.

At any given time, there will be two environments, the active
environment and the alternate environment. When an ADDRESS instruction
specifies a command to the environment, any specified I/O redirection
applies to that command's execution only, providing a third environment
for the duration of the instruction. When an ADDRESS command does not
contain a command, that ADDRESS command creates a new active
environment, which includes the specified I/O redirection.

The redirections specified on the ADDRESS instruction may not be
possible. If the configuration is aware that the command processor named
does not perform I/O in a manner compatible with the request, the value
of \#Env\_Type. may be set to `UNUSED' as an alternative to `STEM' and
`STREAM' where those values are assigned in the following code.

In the following code the particular use of \#Contains(address,
expression) refers to an expression immediately contained in the
address.

Addrinstr: /* If ADDRESS keyword alone, environments are swapped.
\emph{/ if \#Contains (address, taken constant), \& \#Contains
(address,valueexp), \& \#Contains (address, `WITH') then do call
EnvAssign TRANSIENT, \#Level, ACTIVE, \#Level call EnvAssign ACTIVE,
\#Level, ALTERNATE, \#Level call EnvAssign ALTERNATE, \#Level,
TRANSIENT, \#Level return end /} The environment name will be explicitly
specified. */ if \#Contains(address,taken constant) then Name =
\#Instance(address, taken \_ constant) else Name = \#Evaluate(valueexp,
expression) if length(Name) \textgreater{} \#LimitEnvironmentName then
call \#Raise `SYNTAX', 29.1, Name

if \#Contains(address,expression) then do /* The command is evaluated
(but not issued) at this point. */ Command = \#Evaluate (address,
expression)

if \#Tracing.\#Level == `C' \textbar{} \#Tracing.\#Level == `A' then do
call \#Trace '\textgreater\textgreater\textgreater! end

end

call AddressSetup /* Note what is specified on the ADDRESS instruction.
\emph{/ /} If there is no command, the persistent environment is being
set. */ if \#Contains(address,expression) then do

call EnvAssign ACTIVE, \#Level, TRANSIENT, \#Level

return

end

call CommandIssue Command /* See nnn \emph{/ return /} From Addrinstr */

AddressSetup: /* Note what is specified on the ADDRESS instruction, into
the TRANSIENT environment. \emph{/ EnvTail = `TRANSIENT.'\#Level /}
Initialize with defaults. */ \#Env\_Name.EnvTail = '\,'

\#Env\_ Type.I.EnvTail = `NORMAL' \#Env\_ Type.O.EnvTail = `NORMAL'
\#Env\_ Type.E.EnvTail = `NORMAL'

\#Env\_Resource.I.EnvTail = '\,' \#Env\_Resource.O.EnvTail = `!
\#Env\_Resource.E.EnvTail ='' /* APPEND / REPLACE does not apply to
input. */

\#Env\_Position.I.EnvTail = `INPUT' \#Env\_Position.O.EnvTail =
`REPLACE' \#Env\_Position.E.EnvTail = `REPLACE'

/* If anything follows ADDRESS, it will include the command processor
name.*/ \#Env\_Name.EnvTail = Name

/* Connections may be explicitly specified. */ if \#Contains (address,
connection) then do

if \#Contains(connection,input) then do /* input redirection */ if
\#Contains (resourcei, `STREAM') then do \#Env\_Type.I.EnvTail =
`STREAM' \#Env\_Resource.1.EnvTail=\#Evaluate(resourcei, VAR\_SYMBOL)
end if \#Contains (resourcei, `STEM') then do \#Env\_Type.I.EnvTail =
`STEM'

Temp=\#Instance (resourcei,VAR\_SYMBOL) if \#Parses(Temp, stem /* See
nnn \emph{/) then call \#Raise `SYNTAX', 53.3, Temp
\#Env\_Resource.I.EnvTail=Temp end end /} Input */

if \#Contains(connection,output) then /* output redirection */ call
NoteTarget O

if \#Contains(connection,error) then /* error redirection \emph{/ /} The
prose on the description of \#Contains specifies that the relevant
resourceo is used in NoteTarget. */ call NoteTarget E

end /* Connection */

return /* from AddressSetup */

NoteTarget:

/* Note the characteristics of an output resource. */

arg Which /* O or E */

if \#Contains (resourceo, `STREAM') then do
\#Env\_Type.Which.EnvTail=`STREAM'
\#Env\_Resource.Which.EnvTail=\#Evaluate(resourceo, VAR\_SYMBOL) end

if \#Contains(resourceo,`STEM') then do \#Env\_Type.Which.EnvTail=`STEM'
Temp=\#Instance (resourceo, VAR\_SYMBOL) if \#Parses(Temp, stem /* See
nnn */) then

call \#Raise `SYNTAX', 53.3, Temp

\#Env\_Resource.Which.EnvTail=Temp end

if \#Contains (resourceo,append) then
\#Env\_Position.Which.EnvTail=`APPEND' return /* From NoteTarget */

EnvAssign: /* Copy the values that name an environment and describe its
redirections. */ arg Lhs, LhsLevel, Rhs, RhsLevel
\#Env\_Name.Lhs.LhsLevel = \#Env\_Name.Rhs.RhsLevel \#Env\_
Type.I.Lhs.LhsLevel = \#Env\_Type.I.Rhs.RhsLevel \#Env\_
Resource.I.Lhs.LhsLevel = \#Env\_Resource.I.Rhs.RhsLevel
\#Env\_Position.I.Lhs.LhsLevel = \#Env\_Position.I.Rhs.RhsLevel \#Env\_
Type.O.Lhs.LhsLevel = \#Env\_Type.O.Rhs.RhsLevel
\#Env\_Resource.O.Lhs.LhsLevel = \#Env\_Resource.O.Rhs.RhsLevel
\#Env\_Position.O.Lhs.LhsLevel = \#Env\_Position.O.Rhs.RhsLevel \#Env\_
Type.E.Lhs.LhsLevel = \#Env\_Type.E.Rhs.RhsLevel
\#Env\_Resource.E.Lhs.LhsLevel \#Env\_Resource.E.Rhs.RhsLevel
\#Env\_Position.E.Lhs.LhsLevel \#Env\_Position.E.Rhs.RhsLevel return

\hypertarget{arg}{%
\subsection{ARG}\label{arg}}

For a definition of the syntax of this instruction, see nnn.

The ARG instruction is a shorter form of the equivalent instruction:
PARSE UPPER ARG template list

\hypertarget{assignment}{%
\subsection{Assignment}\label{assignment}}

Assignment can occur as the result of executing a clause containing an
assignment (see nnn and nnn), or as a result of executing the VALUE
built-in function, or as part of the execution of a PARSE instruction.
Assignment involves an expression and a VAR\_SYMBOL. The value of the
expression is determined; see nnn.

If the VAR\_SYMBOL does not contain a period, or contains only one
period as its last character, the

value is associated with the VAR\_SYMBOL: call Var Set \#Pool,VAR
SYMBOL, `0',Value

Otherwise, a name is derived, see nnn. The value is associated with the
derived name: call Var Set \#Pool,Derived Name,`1',Value

\hypertarget{call}{%
\subsection{CALL}\label{call}}

For a definition of the syntax of this instruction, see nnn.

The CALL instruction is used to invoke a routine, or is used to control
trapping of conditions. lf a vrefis specified that value is the name of
the routine to invoke:

if \#Contains (call, vref) then Name = \#Evaluate(vref, var\_symbol)

If a taken\_constant is specified, that name is used. if \#Contains
(call, taken constant) then Name = \#Instance(call, taken constant)

The name is used to invoke a routine, see nnn. If that routine does not
return a result the RESULT and

-RESULT variables become uninitialized: call Var Drop \#Pool, `RESULT',
`0! call Var Drop \#ReservedPool,'.RESULT', `0'

If the routine does return a result that value is assigned to RESULT and
.RESULT. See nnn for an exception to assigning results.

If the routine returns a result and the trace setting is `R' or `I' then
a trace with that result and a tag
'\textgreater\textgreater\textgreater'' shall be produced, associated
with the call instruction.

If a callon\_spec is specified: If \#Contains(call,callon spec) then do
Condition = \#Instance(callon\_spec,callable condition)

\#Instruction.Condition.\#Level = `CALL' If \#Contains(callon spec,
`OFF') then \#Enabling.Condition.\#Level = `OFF' else
\#Enabling.Condition.\#Level = `ON'

/* Note whether NAME supplied. */ If Contains (callon spec,taken
constant) then Name = \#Instance (callable condition, taken\_constant)

else

Name = Condition \#TrapName.Condition.\#Level = Name end

\hypertarget{command-to-the-configuration}{%
\subsection{Command to the
configuration}\label{command-to-the-configuration}}

For a definition of the syntax of a command, see nnn. A command that is
not part of an ADDRESS instruction is processed in the ACTIVE
environment.

Command = \#Evaluate(command, expression)

if \#Tracing.\#Level == `C' \textbar{} \#Tracing.\#Level == `A' then
call \#Trace `\textgreater\textgreater\textgreater!'

call EnvAssign TRANSIENT, \#Level, ACTIVE, \#Level

call CommandiIssue Command

Commandlssue is also used to describe the ADDRESS instruction:

CommandIssue: parse arg Cmd /* Issues the command, requested environment
is TRANSIENT \emph{/ /} This description does not require the command
processor to understand stems, so it uses an altered environment. */
call EnvAssign PASSED, \#Level, TRANSIENT, \#Level EnvTail =
`TRANSIENT.'\#Level

/* Note the command input. \emph{/ if \#Env\_Type.I.EnvTail = `STEM'
then do /} Check reasonableness of the stem. \emph{/ Stem =
\#Env\_Resource.I.EnvTail Lines = value(Stem'0') if \datatype(Lines,`W')
then call \#Raise `SYNTAX',54.1,Stem'0', Lines if Lines\textless0 then
call \#Raise `SYNTAX',54.1,Stem'0', Lines /} Use a stream for the stem
*/ \#Env\_ Type.I.PASSED.\#Level = `STREAM' call Config Stream Unique
InputStream = \#Outcome \#Env\_Resource.1I.PASSED.\#Level = InputStream
call charout InputStream , vl do j = 1 to Lines call lineout
InputStream, value(Stem \textbar\textbar{} j) end j call lineout
InputStream end

/* Note the command output. */

if \#Env\_Type.O.EnvTail = `STEM' then do Stem =
\#Env\_Resource.O.EnvTail if \#Env\_Position.O.EnvTail == `APPEND' then
do

/* Check that Stem.0 will accept incrementing. \emph{/ Lines=value
(Stem'0'); if \datatype(Lines,`W') then call \#Raise
`SYNTAX',54.1,Stem'0', Lines if Lines\textless0 then call \#Raise
`SYNTAX',54.1,Stem'0', Lines end else call value Stem'0',O /} Use a
stream for the stem */ \#Env\_Type.O.PASSED.\#Level = `STREAM' call
Config Stream Unique \#Env\_Resource.O.PASSED.\#Level = \#Outcome end

/* Note the command error stream. */

if \#Env\_Type.E.EnvTail = `STEM' then do Stem =
\#Env\_Resource.E.EnvTail if \#Env\_Position.E.EnvTail == `APPEND' then
do

/* Check that Stem.0 will accept incrementing. */ Lines=value (Stem'0');
if \datatype(Lines,`W') then call \#Raise `SYNTAX',54.1,Stem'0', Lines
if Lines\textless0 then call \#Raise `SYNTAX',54.1,Stem'0', Lines

end else call value Stem'0',0O /* Use a stream for the stem */ \#Env\_
Type.E.PASSED.\#Level = `STREAM' call Config Stream Unique
\#Env\_Resource.E.PASSED.\#Level = \#Outcome end

\#API Enabled = `1'

call Var\_Reset \#Pool

/* Specifying PASSED here implies all the

components of that environment. */

\#Response = Config Command (PASSED, Cmd) \#Indicator = left
(\#Response,1) Description = substr (\#Response, 2)

\#API Enabled = `0'

/* Recognize success and failure. */ if \#AtPause = 0 then do

call value `RC', \#RC

call var Set 0, `.RC', 0, \#RC end

select when \#Indicator==`N' then Temp=0

when \#Indicator==`F' then Temp=-1 /* Failure \emph{/ when
\#Indicator==`E' then Temp=1 /} Error \emph{/ end call Var Set 0, `.RS',
0, Temp /} Process the output \emph{/ if \#Env\_Type.O.EnvTail=`STEM'
then do /} get output into stem. \emph{/ Stem =
\#Env\_Resource.0O.EnvTail OutputStream =
\#Env\_Resource.0O.PASSED.\#Level do while lines (OutputStream)
\textgreater{} 0 call value Stem'0O',value(Stem'0')4+1 call value
Stem\textbar{} \textbar value(Stem'0'),linein (OutputStream) end end /}
Stemmed Output \emph{/ if \#Env\_Type.E.EnvTail=`STEM' then do /} get
error output into stem. */ Stem = \#Env\_Resource.E.EnvTail OutputStream
= \#Env\_Resource.E.PASSED.\#Level do while lines (OutputStream)
\textgreater{} 0 call value Stem'0O',value(Stem'0')4+1 call value
Stem\textbar{} \textbar value(Stem'0'),linein (OutputStream)

end end /* Stemmed Error output */ if \#Indicator == `N' \&
pos(\#Tracing.\#Level, `CAIR') \textgreater{} 0 then call \#Trace `+++'
if (\#Indicator == `N' \& \#Tracing.\#Level==`E'), \textbar{}
(\#Indicator==`F' \& (\#Tracing.\#Level==`F' \textbar{}
\#Tracing.\#Level==`N')) then do

call \#Trace `\textgreater\textgreater\textgreater!' call \#Trace `+++!'

end \#Condition=`FAILURE' if \#Indicator=`F' \&
\#Enabling.\#Condition.\#Level == `OFF' then call \#Raise `FAILURE' ,
Cmd else if \#Indicator=`E' \textbar{} \#Indicator=`F' then call \#Raise
`ERROR', Cmd

return /* From CommandIssue */

The configuration may choose to perform the test for message 54.1 before
or after issuing the command.

\hypertarget{do}{%
\subsection{DO}\label{do}}

For a definition of the syntax of this instruction, see nnn.

The DO instructions is used to group instructions together and
optionally to execute them repeatedly. Executing a do\_simple has the
same effect as executing a nop, except in its trace output. Executing
the do\_ending associated with a do\_simple has the same effect as
executing a nop, except in its trace output.

A do\_instruction that does not contain a do\_simple is equivalent,
except for trace output, to a sequence of instructions in the following
order.

\#Loop = \#Loop+1 \#Iterate.\#Loop = \#Clause (IterateLabel)
\#Once.\#Loop = \#Clause (OnceLabel) \#Leave.\#Loop = \#Clause
(LeaveLabel) if \#Contains (do specification,assignment) then
\#Identity.\#Loop = \#Instance(assignment, VAR SYMBOL) if \#Contains (do
specification, repexpr) then if \datatype(repexpr,`W') then call \#Raise
`SYNTAX', 26.2,repexpr else do \#Repeat.\#Loop = repexpr+0 if
\#Repeat.\#Loop\textless0 then call \#Raise
`SYNTAX',26.2,\#Repeat.\#Loop end if \#Contains (do
specification,assignment) then do \#StartValue.\#Loop = \#Evaluate
(assignment, expression) if datatype (\#StartValue.\#Loop) == `NUM' then
call \#Raise `SYNTAX', 41.6, \#StartValue.\#Loop \#StartValue.\#Loop =
\#StartValue.\#Loop + 0 if \#Contains (do specification,byexpr) then
\#By.\#Loop = 1 end

The following three assignments are made in the order in which `TO'``,
`BY' and `FOR' appear in docount; see nnn.

if \#Contains (do specification, toexpr) then do if datatype(toexpr) ==
`NUM' then call \#Raise `SYNTAX', 41.4, toexpr \#To.\#LOop = toexpr+0 if
\#Contains (do specification, byexpr) then do if datatype (byexpr)
==`NUM' then call \#Raise `SYNTAX', 41.5, byexpr \#By.\#Loop = byexpr+0
if \#Contains (do specification, forexpr) then do if \datatype(forexpr,
`W') then call \#Raise `SYNTAX', 26.3, forexpr \#For.\#Loop = forexpr+0
if \#For.\#Loop \textless0 then call \#Raise `SYNTAX', 26.3,
\#For.\#Loop end if \#Contains (do specification,assignment) then do
call value \#Identity.\#Loop, \#StartValue.\#Loop end if \#Contains (do
specification, `OVER') then do Value = \#Evaluate(dorep, expression)
\#OverArray.\#Loop = Value \textasciitilde{} makearray

\#Repeat.\#Loop = \#OverArray\textasciitilde items /* Count this
downwards as if repexpr. */ \#Iidentity.\#Loop = \#Instance(dorep, VAR
SYMBOL) end

call \#Goto \#Once.\#Loop /* to OnceLabel */

IterateLabel:

if \#Contains (do specification, untilexpr) then do Value =
\#Evaluate(untilexp, expression)

if Value == `1' then leave if Value == `0' then call \#Raise `SYNTAX',
34.4, Value end

if \#Contains (do specification, assignment) then do t = value
(\#Identity. \#Loop)

if \#Indicator == `D' then call \#Raise `NOVALUE', \#Identity.\#Loop
call value \#Identity.\#Loop, t + \#By.\#Loop end

OnceLabel:

if \#Contains (do specification, toexpr) then do if
\#By.\#Loop\textgreater=0 then do if value(\#Identity.\#Loop)
\textgreater{} \#To.\#Loop then leave end else do if
value(\#Identity.\#Loop) \textless{} \#To.\#Loop then leave end end

if \#Contains(dorep, repexpr) \textbar{} \#Contains(dorep, `OVER') then
do if \#Repeat.\#Loop = 0 then leave \#Repeat.\#Loop = \#Repeat.\#Loop-1
if \#Contains(dorep, `OVER') then call value \#Identity.\#Loop,
\#OverArray{[}\#OverArray\textasciitilde items - \#Repeat.\#Loop{]} end
if \#Contains (do specification, forexpr) then do if \#For.\#Loop = 0
then leave \#For.\#Loop = \#For.\#Loop - 1 end if \#Contains (do
specification, whileexpr) then do Value = \#Evaluate(whileexp,
expression)

if Value == `0' then leave if Value == `1' then call \#Raise `SYNTAX',
34.3, Value end \#Execute (do instruction, instruction list) TraceOfEnd:
call \#Goto \#Iterate.\#Loop /* to IterateLabel */ LeaveLabel:

\#Loop = \#Loop - 1

\hypertarget{do-loop-tracing}{%
\subsection{DO loop tracing}\label{do-loop-tracing}}

When clauses are being traced by \#TraceSource, due to
pos(\#Tracing.\#Level, `AIR') \textgreater{} 0, the DO instruction shall
be traced when it is encountered and again each time the IterateLabel
(see nnn) is encountered. The END instruction shall be traced when the
TraceOfEnd label is encountered.

When expressions or intermediates are being traced they shall be traced
in the order specified by nnn. Hence, in the absence of conditions
arising, those executed prior to the first execution of OnceLabel shall
be shown once per execution of the DO instruction; others shall be shown
depending on the outcome of the tests.

The code in the DO description: t = value (\#Identity. \#Loop) if
\#Indicator == `D' then call \#Raise `NOVALUE', \#Identity.\#Loop call
value \#Identity.\#Loop, t + \#By.\#Loop

represents updating the control variable of the loop. That assignment is
subject to tracing, and other expressions involving state variables are
not. When tracing intermediates, the BY value will have a tag of
`\textgreater+\textgreater{}',

\hypertarget{drop}{%
\subsection{DROP}\label{drop}}

For a definition of the syntax of this instruction, see nnn.

The DROP instruction restores variables to an uninitialized state.

The words of the variable\_list are processed from left to right.

A word which is a VAR\_SYMBOL, not contained in parentheses, specifies a
variable to be dropped. If VAR\_SYMBOL does not contain a period, or has
only a single period as its last character, the variable

associated with VAR\_SYMBOL by the variable pool is dropped:

\#Response = Var Drop (\#Pool,VAR\_ SYMBOL, `0')

If VAR\_SYMBOL has a period other than as the last character, the
variable associated with VAR\_SYMBOL by the variable pool is dropped by:

\#Response = Var Drop (\#Pool,VAR SYMBOL, `1')

If the word of the variable\_list is a VAR\_SYMBOL enclosed in
parentheses then the value of the

VAR\_SYMBOL is processed. The value is considered in uppercase: \#Value
= Config Upper (\#Value)

Each word in that value found by the WORD built-in function, from left
to right, is subjected to this process:

If the word does not have the syntax of VAR\_SYMBOL a condition is
raised: call \#Raise `SYNTAX', 20.1, word

Otherwise the VAR\_SYMBOL indicated by the word is dropped, as if that
VAR\_SYMBOL were a word of the variable\_list.

\hypertarget{exit}{%
\subsection{EXIT}\label{exit}}

For a definition of the syntax of this instruction, see nnn.

The EXIT instruction is used to unconditionally complete execution of a
program.

Any expression is evaluated:

if \#Contains(exit, expression) then Value = \#Evaluate(exit,
expression) \#Level = 1

\#Pool = \#Pooll

The opportunity is provided for a final trap. \#API Enabled = `1' call
Var\_Reset \#Pool call Config Termination \#API Enabled = `0'

The processing of the program is complete. See nnn for what API Start
returns as the result.

If the normal sequence of execution ``falls through'' the end of the
program; that is, would execute a further statement if one were appended
to the program, then the program is terminated in the same manner as an
EXIT instruction with no argument.

\hypertarget{expose}{%
\subsection{EXPOSE}\label{expose}}

The expose instruction identifies variables that are not local to the
method.

We need a check that this starts method; similarities with PROCEDURE.

For a definition of the syntax of this instruction, see nnn.

It is used at the start of a method, after method initialization, to
make variables in the receiver's pool

accessible: if \#AllowExpose then call \#Raise `SYNTAX', 17.2

The words of the variable\_list are processed from left to right.

A word which is a VAR\_SYMBOL, not contained in parentheses, specifies a
variable to be made accessible. If VAR\_SYMBOL does not contain a
period, or has only a single period as its last character, the variable
associated with VAR\_SYMBOL by the variable pool (as a non-tailed name)
is given the

attribute `exposed'. call Var\_ Expose \#Pool, VAR SYMBOL, `0'

If VAR\_SYMBOL has a period other than as last character, the variable
associated with VAR\_SYMBOL

in the variable pool ( by the name derived from VAR\_SYMBOL, see nnn) is
given the attribute `exposed'. call Var\_ Expose \#Pool, Derived Name,
`1'

If the word from the variable\_list is a VAR\_SYMBOL enclosed in
parentheses then the VAR\_SYMBOL is exposed, as if that VAR\_SYMBOL was
a word in the variable\_list. The value of the VAR\_SYMBOL is

processed. The value is considered in uppercase: \#Value = Config Upper
(\#Value)

Each word in that value found by the WORD built-in function, from left
to right, is subjected to this process:

If the word does not have the syntax of VAR\_SYMBOL a condition is
raised: call \#Raise `SYNTAX', 20.1, word

Otherwise the VAR\_SYMBOL indicated by the word is exposed, as if that
VAR\_SYMBOL were a word of the variable\_list.

\hypertarget{forward}{%
\subsection{FORWARD}\label{forward}}

For a definition of the syntax of this instruction, see nnn.

The FORWARD instruction is used to send a message based on the current
message. if \#Contains (forward, `ARRAY') \& \#Contains(forward,
`ARGUMENTS') then call \#Raise `SYNTAX', nn.n

\hypertarget{guard}{%
\subsection{GUARD}\label{guard}}

For a definition of the syntax of this instruction, see nnn.

The GUARD instruction is used to conditionally delay the execution of a
method. do forever Value = \#Evaluate( guard, expression)

if Value == `1' then leave

if Value == `0' then call \#Raise `SYNTAX', 34.7, Value Drop exclusive
access and wait for change

end \#\#\# IF

For a definition of the syntax of this instruction, see nnn.

The IF instruction is used to conditionally execute an instruction, or
to select between two alternatives. The expression is evaluated. If the
value is neither `0' nor `1' error 34.1 occurs. If the value is `1', the
instruction in the then is executed. If the value is `0' and e/se is
specified, the instruction in the else is executed.

In the former case, if tracing clauses, the clause consisting of the
THEN keyword shall be traced in addition to the instructions.

In the latter case, if tracing clauses, the clause consisting of the
ELSE keyword shall be traced in addition to the instructions.

\hypertarget{interpret}{%
\subsection{INTERPRET}\label{interpret}}

For a definition of the syntax of this instruction, see nnn.

The INTERPRET instruction is used to execute instructions that have been
built dynamically by evaluating an expression.

The expression is evaluated.

The HALT condition is tested for, and may be raised, in the same way it
is tested at clause termination, see nnn.

The process of syntactic recognition described in clause 6 is applied,
with Config\_SourceChar obtaining its results from the characters of the
value, in left-to-right order, without producing any EOL or EOS events.
When the characters are exhausted, the event EOL occurs, followed by the
event EOS.

If that recognition would produce any message then the interpret raises
the corresponding `SYNTAX' condition.

If the program recognized contains any LABELs then the interpret raises
a condition: call \#Raise `SYNTAX',47.1,Label

where Label is the first LABEL in the program.

Otherwise the instruction\_list in the program is executed.

\hypertarget{iterate}{%
\subsection{ITERATE}\label{iterate}}

For a definition of the syntax of this instruction, see nnn.

The ITERATE instruction is used to alter the flow of control within a
repetitive DO. For a definition of the nesting correction, see nnn.

\#Loop = \#Loop - NestingCorrection call \#Goto \#Iterate.\#Loop

\hypertarget{execution-of-labels}{%
\subsection{Execution of labels}\label{execution-of-labels}}

The execution of a label has no effect, other than clause termination
activity and any tracing. if \#Tracing.\#Level==`L' then call
\#TraceSource

\hypertarget{leave}{%
\subsection{LEAVE}\label{leave}}

For a definition of the syntax of this instruction, see nnn.

The LEAVE instruction is used to immediately exit one or more repetitive
DOs. For a definition of the nesting correction, see nnn.

\#Loop = \#Loop - NestingCorrection call \#Goto \#Leave.\#Loop

\hypertarget{message-term}{%
\subsection{Message term}\label{message-term}}

We can do this by reference to method invokation, just as we do CALL by
reference to invoking a function.

\hypertarget{loop}{%
\subsection{LOOP}\label{loop}}

Shares most of it's definition with repetitive DO.

\hypertarget{nop}{%
\subsection{NOP}\label{nop}}

For a definition of the syntax of this instruction, see nnn. The NOP
instruction has no effect other than the effects associated with all
instructions.

\hypertarget{numeric}{%
\subsection{NUMERIC}\label{numeric}}

For a definition of the syntax of this instruction, see nnn. The NUMERIC
instruction is used to change the way in which arithmetic operations are
carried out.

\hypertarget{numeric-digits}{%
\subsubsection{NUMERIC DIGITS}\label{numeric-digits}}

For a definition of the syntax of this instruction, see nnn.

NUMERIC DIGITS controls the precision under which arithmetic operations
and arithmetic built-in functions will be evaluated.

if \#Contains(numericdigits, expression) then

Value = \#Evaluate(numericdigits, expression) else Value = 9

if \datatype(Value,`W') then

call \#Raise `SYNTAX',26.5,Value Value = Value + 0 if
Value\textless=\#Fuzz.\#Level then

call \#Raise `SYNTAX',33.1,Value if Value\textgreater\#Limit Digits then

call \#Raise `SYNTAX',33.2,Value \#Digits.\#Level = Value

\hypertarget{numeric-form}{%
\subsubsection{NUMERIC FORM}\label{numeric-form}}

For a definition of the syntax of this instruction, see nnn.

NUMERIC FORM controls which form of exponential notation is to be used
for the results of operations and arithmetic built-in functions.

The value of form is either taken directly from the SCIENTIFIC or
ENGINEERING keywords, or by

evaluating valueexp . if \#Contains (numeric,numericsuffix) then

Value = `SCIENTIFIC' else if \#Contains (numericformsuffix,
`SCIENTIFIC') then Value = `SCIENTIFIC' else if \#Contains
(numericformsuffix, `ENGINEERING') then Value = `ENGINEERING' else do
Value = \#Evaluate (numericformsuffix,valueexp) Value = translate (left
(Value,1)) select when Value == `S' then Value = `SCIENTIFIC' when Value
== `E' then Value = `ENGINEERING' otherwise call \#Raise
`SYNTAX',33.3,Value end end \#Form.\#Level = Value

\hypertarget{numeric-fuzz}{%
\subsubsection{NUMERIC FUZZ}\label{numeric-fuzz}}

For a definition of the syntax of this instruction, see nnn. NUMERIC
FUZZ controls how many digits, at full precision, will be ignored during
a numeric comparison.

If \#Contains (numericfuzz,expression) then Value = \#Evaluate
(numericfuzz,expression) else Value = 0 If \datatype(Value,`W') then
call \#Raise `SYNTAX',26.6,Value Value = Value+0 If Value \textless{} 0
then call \#Raise `SYNTAX',26.6,Value If Value \textgreater=
\#Digits.\#Level then call \#Raise `SYNTAX',33.1,\#Digits.\#Level,Value
\#Fuzz.\#Level = Value

\hypertarget{options}{%
\subsubsection{OPTIONS}\label{options}}

For a definition of the syntax of this instruction, see nnn.

The OPTIONS instruction is used to pass special requests to the language
processor.

The expression is evaluated and the value is passed to the language
processor. The language processor treats the value as a series of blank
delimited words. Any words in the value that are not recognized by

the language processor are ignored without producing an error. call
Config Options (Expression)

\hypertarget{parse}{%
\subsubsection{PARSE}\label{parse}}

For a definition of the syntax of this instruction, see nnn.

The PARSE instruction is used to assign data from various sources to
variables.

The purpose of the PARSE instruction is to select substrings of the
parse\_fype under control of the template\_list. \textbar f the
template\_list is omitted, or a template in the list is omitted, then a
template which is the null string is implied.

Processing for the PARSE instruction begins by constructing a value, the
source to be parsed.

ArgNum = 0 select when \#Contains (parse type, `ARG') then do ArgNum = 1
ToParse = \#Arg.\#Level.ArgNum end when \#Contains (parse type,
`LINEIN') then ToParse = linein('`) when \#Contains (parse type, 'PULL')
then do /* Acquire from external queue or default input. */ \#Response =
Config Pull() if left(\#Response, 1) == `F' then call Config Default
Input ToParse = \#Outcome end when \#Contains (parse type, `SOURCE')
then ToParse = \#Configuration \#HowInvoked \#Source when \#Contains
(parse type, `VALUE') then if \#Contains(parse value, expression) then
ToParse = '\,' else ToParse = \#Evaluate(parse value, expression) when
\#Contains (parse type, `VAR') then ToParse = \#Evaluate (parse var,
VAR\_SYMBOL) when \#Contains (parse type, `VERSION') then ToParse =
\#Version end Uppering = \#Contains(parse, `UPPER') The first template
is associated with this source. If there are further templates, they are
matched against null strings unless `ARG' is specified, when they are
matched against further arguments. The parsing process is defined by the
following routine, ParseData. The template\_list is accessed by
ParseData as a stemmed variable. This variable Template. has elements
which are null strings except

for any elements with tails 1,2,3,\ldots{} corresponding to the tokens
of the template\_list from left to right.

ParseData: /* Targets will be flagged as the template is examined.
\emph{/ Target.=`0' /} Token is a cursor on the components of the
template, moved by FindNextBreak. \emph{/ Token = 1 /} Tok ig a cursor
on the components of the template moved through the target variables by
routine WordParse. */ Tok = 1

do forever /* Until commas dealt with. \emph{/ /} BreakStart and
BreakEnd indicate the position in the source string where there is a
break that divides the source. When the break is a pattern they are the
start of the pattern and the position just beyond it. */ BreakStart =
BreakEnd = 1 SourceEnd = length(ToParse) + 1 If Uppering then ToParse =
translate (ToParse)

do while Template.Tok == '\,' \& Template.Tok == `,'

/* Isolate the data to be processed on this iteration. \emph{/ call
FindNextBreak /} Also marks targets. */

/* Results have been set in DataStart which indicates the start of the
isolated data and BreakStart and BreakEnd which are ready for the next
iteration. Tok has not changed. */

/* If a positional takes the break leftwards from the end of the
previous selection, the source selected is the rest of the string, */

if BreakEnd \textless= DataStart then DataEnd = SourceEnd

else DataEnd = BreakStart

/* Isolated data, to be assigned from: */

Data=substr (ToParse,DataStart, DataEnd-DataStart) call WordParse /*
Does the assignments. */

end /* while \emph{/ if Template.Tok == `,' then leave /} Continue with
next source. */ Token=Token+1 Tok=Token if ArgNum
\textless\textgreater{} 0 then do ArgNum = ArgNum+1 ToParse =
\#Arg.ArgNum end else ToParse='\,' end

return /* from ParseData */

FindNextBreak: do while Template.Token == '\,' \& Template.Token == `,'

Type=left (Template.Token,1) /* The source data to be processed next
will normally start at the end of the break that ended the previous
piece. (However, the relative positionals alter this.) */ DataStart =
BreakEnd select

when Type=`\,``' \textbar{} Type=''\,''' \textbar{} Type=`(' then do if
Type=`(' then do /* A parenthesis introduces a pattern which is not a
constant. */ Token = Token+1 Pattern = value(Template.Token)

if \#Indicator == `D' then call \#Raise `NOVALUE', Template.Token Token
= Token+1 end

else

/* The following removes the outer quotes from the literal pattern
\emph{/ interpret ``Pattern=''Template.Token Token = Token+1 /} Is that
pattern in the remaining source? \emph{/ PatternPos=pos (Pattern,
ToParse,DataStart) if PatternPos\textgreater0 then do /} Selected source
runs up to the pattern. \emph{/ BreakStart=PatternPos
BreakEnd=PatternPos+length (Pattern) return end leave /} The rest of the
source is selected. */ end

when datatype(Template.Token,`W') \textbar{} pos(Type,`+-=')
\textgreater{} 0 then do /* A positional specifies where the relevant
piece of the subject ends. \emph{/ if pos (Type, `+-=') = 0 then do /}
Whole number positional \emph{/ BreakStart = Template.Token Token =
Token+1 end else do /} Other forms of positional. \emph{/
Direction=Template.Token Token = Token + 1 /} For a relative positional,
the position is relative to the start of the previous trigger, and the
source segment starts there. \emph{/ if Direction == `=' then DataStart
= BreakStart /} The adjustment can be given as a number or a variable in
parentheses. */ if Template.Token =`(' then do Token=Token + 1
BreakStart = value(Template. Token) if \#Indicator == `D' then call
\#Raise `NOVALUE', Template.Token Token=Token + 1 end else BreakStart =
Template.Token

if

if

el

\datatype (BreakStart,`W')

then call \#Raise `SYNTAX', 26.4,BreakStart Token = Token+1

Direction=`+'

then BreakStart=DataStart+BreakStart se if Direction=`-'

then BreakStart=DataStart-BreakStart

end /* Adjustment should remain within the ToParse \emph{/ BreakStart =
max(1, BreakStart) BreakStart = min(SourceEnd, BreakStart) BreakEnd =
BreakStart /} No actual literal marks the boundary. */ return

end

when Template.Token == `.' \& pos(Type, `0123456789.')\textgreater0 then
/* A number that isn't a whole number. */

call

\#Raise `SYNTAX', 26.4, Template.Token

/* Raise will not return */

otherwise do /* It is a target, not a pattern */ Target.Token=`1' Token
= Token+1

end

end /*

select */

end /* while \emph{/ /} When no more explicit breaks, break is at the
end of the source. */ DataStart=BreakEnd BreakStart=SourceEnd
BreakEnd=SourceEnd

return

WordParse: /* The names in the template are assigned blank-delimited
values from the source string. */

/* From FindNextBreak */

do while Target.Tok /* Until no more targets for this data. */

/* Last target gets all the residue of the Data.

Next Tok

= Tok + 1

if \Target.NextTok then do call Assign (Data)

leave end /* Not 1 Data = s if Data else do Word =

ast target; assign a word. */ trip (Data, `L') == '\,' then call
Assign('\,')

word (Data,1)

call Assign Word

Data =

substr(Data,length(Word) + 1)

*/

/* The word terminator is not part of the residual data: */ if Data ==
'\,' then Data = substr (Data, 2)

end

Tok = Tok + 1

*/

end Tok=Token /* Next time start on new part of template. \emph{/ return
Assign: if Template.Tok==`.' then Tag=`\textgreater.\textgreater{}' else
do Tag=`\textgreater=\textgreater{}' call value Template.Tok,arg(1) end
/} Arg(1) is an implied argument of the tracing. if \#Tracing.\#Level ==
`R' \textbar{} \#Tracing.\#Level == `I'

return

\hypertarget{procedure}{%
\subsection{PROCEDURE}\label{procedure}}

For a definition of the syntax of this instruction, see nnn.

then call \#Trace Tag The PROCEDURE instruction is used within an
internal routine to protect all the existing variables by making them
unknown to following instructions. Selected variables may be exposed.

It is used at the start of a routine, after routine initialization:

if \#AllowProcedure.\#Level then call \#Raise `SYNTAX', 17.1

\#AllowProcedure.\#Level = 0

/* It introduces a new variable pool: */

call \#Config ObjectNew

call var\_set (\#Outcome,`\#UPPER', `0', \#Pool) /* Previous \#Pool is
upper from the new \#Pool. */

\#Pool=\#O0Outcome

IsProcedure.\#Level=`1'

call Var\_Empty \#Pool

If there is a variable\_list, it provides access to a previous variable
pool.

The words of the variable\_list are processed from left to right.

A word which is a VAR\_SYMBOL, not contained in parentheses, specifies a
variable to be made accessible. If VAR\_SYMBOL does not contain a
period, or has only a single period as its last character, the variable
associated with VAR\_SYMBOL by the variable pool (as a non-tailed name)
is given the

attribute `exposed'. call Var\_ Expose \#Pool, VAR SYMBOL, `0'

If VAR\_SYMBOL has a period other than as last character, the variable
associated with VAR\_SYMBOL

in the variable pool ( by the name derived from VAR\_SYMBOL, see nnn) is
given the attribute `exposed'. call Var\_ Expose \#Pool, Derived Name,
`1'

If the word from the variable\_list is a VAR\_SYMBOL enclosed in
parentheses then the VAR\_SYMBOL is exposed, as if that VAR\_SYMBOL was
a word in the variable\_list. Tne value of the VAR\_SYMBOL is

processed. The value is considered in uppercase: \#Value = Config Upper
(\#Value)

Each word in that value found by the WORD built-in function, from left
to right, is subjected to this process:

If the word does not have the syntax of VAR\_SYMBOL a condition is
raised: call \#Raise `SYNTAX', 20.1, word

Otherwise the VAR\_SYMBOL indicated by the word is exposed, as if that
VAR\_SYMBOL were a word of the variable\_list.

\hypertarget{pull}{%
\subsection{PULL}\label{pull}}

For a definition of the syntax of this instruction, see nnn.

A PULL instruction is a shorter form of the equivalent instruction:
PARSE UPPER PULL template list

\hypertarget{push}{%
\subsection{PUSH}\label{push}}

For a definition of the syntax of this instruction, see nnn. The PUSH
instruction is used to place a value on top of the stack.

If \#Contains(push,expression) then Value = \#Evaluate (push,
expression) else Value = '\,' call Config Push Value

\hypertarget{queue}{%
\subsection{QUEUE}\label{queue}}

For a definition of the syntax of this instruction, see nnn. The QUEUE
instruction is used to place a value on the bottom of the stack.

If \#Contains (queue,expression) then Value = \#Evaluate (queue,
expression) else Value = '\,' call Config Queue Value

\hypertarget{raise}{%
\subsection{RAISE}\label{raise}}

The RAISE instruction returns from the current method or routine and
raises a condition.

\hypertarget{reply}{%
\subsection{REPLY}\label{reply}}

The REPLY instruction is used to allow both the invoker of a method, and
the replying method, to continue executing.

Must set up for error of expression on subsequent RETURN. \#\#\# RETURN
For a definition of the syntax of this instruction, see nnn. The RETURN
instruction is used to return control and possibly a result from a
program or internal routine to the point of its invocation. The RETURN
keyword may be followed by an optional expression, which will be
evaluated and returned as a result to the caller of the routine. Any
expression is evaluated: if \#Contains(return,expression) then \#Outcome
= \#Evaluate(return, expression)

else if \#IsFunction.\#Level then call \#Raise `SYNTAX', 45.1,
\#Name.\#Level

At this point the clause termination occurs and then the following:

If the routine started with a PROCEDURE instruction then the associated
pool is taken out of use: if \#IsProcedure.\#Level then \#Pool = \#Upper
A RETURN instruction which is interactively entered at a pause point
leaves the pause point. if \#Level = \#AtPause then \#AtPause = 0 The
activity at this level is complete: \#Level = \#Level-1 \#NewLevel =
\#Level+1 If \#Level is not zero, the processing of the RETURN
instruction and the invocation is complete. Otherwise processing of the
program is completed:

The opportunity is provided for a final trap. \#API Enabled = `1'

call Var\_Reset \#Pool

call Config Termination

\#API Enabled = `0'

The processing of the program is complete. See nnn for what API Start
returns as the result. \#\#\# SAY

For a definition of the syntax of this instruction, see nnn.

The SAY instruction is used to write a line to the default output
stream.

If \#Contains(say,expression) then Value = Evaluate (say, expression)
else Value = '\,' call Config Default Output Value

\hypertarget{select}{%
\subsection{SELECT}\label{select}}

For a definition of the syntax of this instruction, see nnn.

The SELECT instruction is used to conditionally execute one of several
alternative instructions. When tracing, the clause containing the
keyword SELECT is traced at this point.

The \#Contains(select\_body, when) test in the following description
refers to the items of the optional when repetition in order:

LineNum = \#LineNumber Ending = \#Clause (EndLabel) Value=\#Evaluate
(select body, expression) /* In the required WHEN \emph{/ if Value ==
`1' \& Value == `0' then call \#Raise `SYNTAX',34.2,Value If Value==`1'
then call \#Execute when, instruction else do do while \#Contains
(select body, when) Value = \#Evaluate (when, expression) If Value==`1'
then do call \#Execute when, instruction call \#Goto Ending end if Value
== `0' then call \#Raise `SYNTAX', 34.2, Value end /} Of each when */

If \#Contains(select body, `OTHERWISE') then call \#Raise `SYNTAX', 7.3,
LineNum If \#Contains (select body, instruction list) then call
\#Execute select body, instruction list end EndLabel:

When tracing, the clause containing the END keyword is traced at this
point.

\hypertarget{signal}{%
\subsection{SIGNAL}\label{signal}}

For a definition of the syntax of this instruction, see nnn.

The SIGNAL instruction is used to cause a change in the flow of control
or is used with the ON and OFF keywords to control the trapping of
conditions.

If \#Contains (signal,signal spec) then do Condition = \#Instance(signal
spec,condition)

\#Instruction.Condition.\#Level = `SIGNAL' If \#Contains (signal spec,
`OFF') then \#Enabling.Condition.\#Level = `OFF' else
\#Enabling.Condition.\#Level = `ON'

If Contains (signal spec,taken constant) then Name = \#Instance
(condition, taken constant)

else

Name = Condition \#TrapName.Condition.\#Level = Name end

If there was a signal\_spec this complete the processing of the signal
instruction. Otherwise: if \#Contains (signal, valueexp)

then Name \#Evaluate(valueexp, expression)

else Name \#Instance(signal,taken constant)

The Name matches the first LABEL in the program which has that value.
The comparison is made with the `==``' operator.

If no label matches then a condition is raised:

call \#Raise `SYNTAX',16.1, Name

If the name is a trace-only label then a condition is raised: call
\#Raise `SYNTAX', 16.2, Name

If the name matches a label, execution continues at that label after
these settings: \#Loop.\#Level = 0

/* A SIGNAL interactively entered leaves the pause point. */

if \#Level = \#AtPause then \#AtPause = 0

\hypertarget{trace}{%
\subsection{TRACE}\label{trace}}

For a definition of the syntax of this instruction, see nnn.

The TRACE instruction is used to control the trace setting which in turn
controls the tracing of execution of the program.

The TRACE instruction is ignored if it occurs within the program (as
opposed to source obtained by

Config\_Trace\_Input) and interactive trace is requested
(\#Interactive.\#Level = `1'). Otherwise: \#TraceInstruction = `1' value
= '\,' if \#Contains(trace, valueexp) then Value = \#Evaluate(valueexp,
expression) if \#Contains (trace, taken constant) then Value =
\#Instance (trace, taken constant) if datatype(Value) == `NUM' \&
\datatype(Value,`W') then call \#Raise `SYNTAX', 26.7, Value if
datatype(Value,`W') then do /* Numbers are used for skipping. \emph{/ if
Value\textgreater=0 then \#InhibitPauses = Value else \#InhibitTrace =
-Value end else do if length(Value) = 0 then do \#Interactive.\#Level =
`0' Value = `N' end /} Each question mark toggles the interacting. */ do
while left(Value,1)==`?' \#Interactive.\#Level = \#Interactive.\#Level
Value = substr(Value,2) end if length(Value) = 0 then do Value =
translate( left(Value,1) ) if verify(Value, `ACEFILNOR') \textgreater{}
0 then call \#Raise `SYNTAX', 24.1, Value if Value==`0O' then
\#Interactive.\#Level=`0' end \#Tracing.\#Level = Value end

\hypertarget{trace-output}{%
\subsection{Trace output}\label{trace-output}}

If \#NoSource is `1' there is no trace output.

The routines \#TraceSource and \#Trace specify the output that results
from the trace settings. That output is presented to the configuration
by Config\_Trace\_Output as lines. Each line has a clause identifier at
the left, followed by a blank, followed by a three character tag,
followed by a blank, followed by the trace data.

The width of the clause identifier shall be large enough to hold the
line number of the last line in the program, and no larger. The clause
identifier is the source program line number, or all blank if the line
number is the same as the previous line number indicated and no
execution with trace Off has occurred since. The line number is
right-aligned with leading zeros replaced by blank characters.

When input at a pause is being executed (\#AtPause = 0 ), \#Trace does
nothing when the tag is not `+++'.

When input at a pause is being executed, \#TraceSource does nothing. If
\#InhibitTrace is greater than zero, \#TraceSource does nothing except
decrement \#InhibitTrace. Otherwise, unless the current clause is a null
clause, \#TraceSource outputs all lines of the source program which
contain any part of the current clause, with any characters in those
lines which are not part of the current clause and not
other\_blank\_characters replaced by blank characters. The possible
replacement of other\_blank\_characters is defined by the configuration.
The tag is '\emph{-}'', or if the line is not the first line of the
clause. ™,*'. \#Trace output also has a clause identifier and has a tag
which is the argument to the \#Trace invocation. The data is truncated,
if necessary, to \#Limit\_TraceData characters. The data is enclosed by
quotation marks and the quoted data preceded by two blanks. If the data
is truncated, the trailing quote has the three characters `\ldots{}'
appended. \_ when \#Tracing.\#Level is `C' or `E' or `F' or `N' or `A'
and the tag is `\textgreater\textgreater\textgreater{}' then the data is
the value of the command passed to the environment; \_ when the tag is
`+++' then the data is the four characters `RC concatenated with the
character ``\,``; \_ when \#Tracing.\#Level is 'l' or `R' the data is
the most recently evaluated value. Trace output can also appear as the
result of a'SYNTAX' condition occurring, irrespective of the trace
setting. If a'SYNTAX' condition occurs and it is not trapped by SIGNAL
ON SYNTAX, then the clause in error shall be traced, along with a
traceback. A traceback is a display of each active CALL and INTERPRET
instruction, and function invocation, displayed in reverse order of
execution, each with a tag of `+4+'. \#\#\# USE For a definition of the
syntax of this instruction, see nnn. The USE instruction assigns the
values of arguments to variables. Better not say copies since COPY
method has different semantics. The optional VAR\_SYMBOL positions,
positions 1, 2, \ldots, of the instruction are considered from left to
right. If the position has a VAR\_SYMBOL then its value is assigned to:

if \#ArgExists.Position then call Value VAR\_SYMBOL, \#Arg.Position else

Messy because VALUE bif won't DROP and var\_drop needs to know if
compound.

\hypertarget{conditions-and-messages}{%
\section{Conditions and Messages}\label{conditions-and-messages}}

When an error occurs during execution of a program, an error number and
message are associated with it. The error number has two parts, the
error code and the error subcode. These are the integer and decimal
parts of the error number. Subcodes beginning or ending in zero are not
used.

Error codes in the range 1 to 90 and error subcodes up to .9 are
reserved for errors described here and for future extensions of this
standard.

concatenated with \#RC

Error number 3 is available to report error conditions occuring during
the initialization phase; error number 2 is available to report error
conditions during the termination phase. These are error conditions
recognized by the language processor, but the circumstances of their
detection is outside of the scope of this standard.

The ERRORTEXT built-in function returns the text as initialized in nnn
when called with the `Standard' option. When the `Standard' option is
omitted, implementation-dependent text may be returned.

When messages are issued any message inserts are replaced by actual
values.

The notation for detection of a condition is:

call \#Raise Condition, Arg2, Arg3, Arg4, Arg5, Arg6é

Some of the arguments may be omitted. In the case of condition `SYNTAX'
the arguments are the message number and the inserts for the message. In
other cases the argument is a further description of the condition.

The action of the program as a result of a condition is dependent on any
signa/\_spec and callon\_spec in the program.

\hypertarget{raising-of-conditions}{%
\subsection{Raising of conditions}\label{raising-of-conditions}}

The routine \#Raise corresponds to raising a condition. In the following
definition, the instructions containing SIGNAL VALUE and INTERPRET
denote transfers of control in the program being processed. The
instruction EXIT denotes termination. If not at an interactive pause,
this will be termination of the program, see nnn, and there will be
output by Config\_Trace\_Output of the message (with prefix \_ see nnn)
and tracing (see nnn). If at an interactive pause (\#AtPause = 0), this
will be termination of the interpretation of the interactive input;
there will be output by Config\_Trace\_Output of the message (without
traceback) before continuing. The description of the continuation is in
nnn after the ``interpret \#Outcome'' instruction.

The instruction ``interpret `CALL' \#TrapName.\#Condition.\#Level''
below does not set the variables RESULT and .RESULT; any result returned
is discarded.

\#Raise: /* If there is no argument, this is an action which has been
delayed from the time the condition occurred until an appropriate clause
boundary. */ if \arg(1,`E') then do Description = \#PendingDescription.
\#Condition. \#Level Extra = \#PendingExtra.\#Condition. \#Level

end else do \#Condition = arg(1) if \#Condition == `SYNTAX' then do

Description = arg(2) Extra = arg(3) end else do Description =
\#Message(arg(2),arg(3),arg(4),arg(5)) call Var Set \#ReservedPool,
`.MN', 0, arg(2) Extra = '! end end

/* The events for disabled conditions are ignored or cause termination.
*/

if \#Enabling.\#Condition.\#Level == `OFF' \textbar{} \#AtPause = 0 then
do if \#Condition == `SYNTAX' \& \#Condition == `HALT' then return /* To
after use of \#Raise. \emph{/ if \#Condition == `HALT' then Description
= \#Message(4.1, Description) exit /} Terminate with Description as the
message. */ end

/* SIGNAL actions occur as soon as the condition is raised. */

if \#Instruction.\#Condition.\#Level == ```SIGNAL' then do
\#ConditionDescription.\#Level = Description \#ConditionExtra.\#Level =
Extra \#ConditionInstruction.\#Level = `SIGNAL'
\#Enabling.\#Condition.\#Level = `OFF' signal value
\#TrapName.\#Condition.\#Level end

/* All CALL actions are initially delayed until a clause boundary. */

if arg(1,`E') then do /* Events within the handler are not stacked up,
except for one extra HALT while a first is being handled. \emph{/
EventLevel = \#Level if \#Enabling.\#Condition.\#Level == `DELAYED' then
do if \#Condition == `HALT' then return EventLevel =
\#EventLevel.\#Condition. \#Level if \#PendingNow.\#Condition.EventLevel
then return /} Setup a HALT to come after the one being handled. \emph{/
end /} Record a delayed event. \emph{/
\#PendingNow.\#Condition.EventLevel = `1'
\#PendingDescription.\#Condition.EventLevel = Description
\#PendingExtra.\#Condition.EventLevel = Extra
\#Enabling.\#Condition.EventLevel = `DELAYED' return end /} Here for
CALL action after delay. \emph{/ /} Values for the CONDITION built-in
function. \emph{/ \#Condition.\#NewLevel = \#Condition
\#ConditionDescription.\#NewLevel = \#PendingDescription. \#Condition.
\#Level \#ConditionExtra.\#NewLevel = \#PendingExtra.\#Condition.
\#Level \#ConditionInstruction.\#NewLevel = `CALL' interpret `CALL'
\#TrapName.\#Condition.\#Level \#Enabling.\#Condition.\#Level = `ON'
return /} To clause termination */

\hypertarget{messages-during-execution}{%
\subsection{Messages during execution}\label{messages-during-execution}}

The state function \#Message corresponds to constructing a message.

This definition is for the message text in nnn. Translations in which
the message inserts are in a different order are permitted.

In addition to the result defined below, the values of MsgNumber and
\#LineNumber shall be shown when a message is output. Also there shall
be an indication of whether the error occurred in code executed at an
interactive pause, see nnn.

Messages are shown by writing them to the default error stream.

\#Message: MsgNumber = arg(1) if \#NoSource then MsgNumber = MsgNumber
\% 1 /* And hence no inserts */ Text = \#ErrorText.MsgNumber Expanded =
'\,'

do Index = 2 parse var Text Begin `\textless{}' Insert `\textgreater{}'
+1 Text

if Insert = '\,' then leave Insert = arg(Index) if length(Insert)
\textgreater{} \#Limit MessageInsert then Insert = left(Insert,\#Limit
MessageInsert)`\ldots{}'

Expanded = Expanded \textbar\textbar{} Begin \textbar\textbar{} Insert

end

Expanded = Expanded \textbar\textbar{} Begin

say Expanded return

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/provided_classes.tex
complete, but very deficient, extended standard

\hypertarget{provided-classes}{%
\chapter{Provided classes}\label{provided-classes}}

(Informative)

\hypertarget{notation}{%
\section{Notation}\label{notation}}

The provided classes are defined mainly through code. \#\# The
Collection Classes

\hypertarget{collection-class-routines}{%
\subsection{Collection Class Routines}\label{collection-class-routines}}

These routines are used in the definition of the collection classes
::routine CommonxXor /* Returns a new collection that contains all items
from self and the argument except that all indexes that appear in both
collections are removed. \emph{/ /} When the target is a bag, there may
be an index in the bag that is duplicated and the same value as an index
in the argument. Should one copy of the index survive in the bag?
\emph{/ vel if (arg(1)\textasciitilde class==.Set \&
arg(2)\textasciitilde class==.Bag) then v=2 if
(arg(1)\textasciitilde class==.Table \&
arg(2)\textasciitilde class==.Bag) then v=2 if
(arg(1)\textasciitilde class==.Table \&
arg(2)\textasciitilde class==.Relation) then v=2 if
(arg(1)\textasciitilde class==.Directory \&
arg(2)\textasciitilde class==.Bag) then v=2 if
(arg(1)\textasciitilde class==.Directory \&
arg(2)\textasciitilde class==.Relation) then v=2 /} This version it
does: \emph{/ if v=1 then do This = arg(1) /} self of caller */
r=This\textsubscript{class}new ab=MayEnBag (arg (2) )
ss=This\textasciitilde supplier do while ss\textasciitilde available
r{[}ss\textasciitilde index{]} =ss\textasciitilde item
ss\textasciitilde next end cs=ab\textasciitilde supplier do while
cs\textasciitilde available if r\textsubscript{hasindex(cs}index) then
r\textasciitilde remove (cs\textasciitilde index) else
r{[}cs\textasciitilde index{]} =cs\textasciitilde item
cs\textasciitilde next end return r end

/* But following matches practice on Set\textasciitilde XOR(bag) etc. */

This = arg(1) /* self of caller */

r=This\textsubscript{class}new

ab=MayEnBag (arg (2) )

ss=This\textasciitilde supplier

do while ss\textasciitilde available if
\ab\textsubscript{hasindex(ss}index) then r{[}ss\textasciitilde index{]}
=ss\textasciitilde item ss\textasciitilde next end

cs=ab\textasciitilde supplier

do while cs\textasciitilde available if
\This\textsubscript{hasindex(cs}index) then
r{[}cs\textasciitilde index{]} =cs\textasciitilde item es-next end

return r

::routine CommoniIntersect /* Returns a new collection of the same class
as SELF that contains the items from SELF that have indexes also in the
argument. \emph{/ /} Actually an index in SELF can only be `matched'
with one in the argument once. Hence copy and removal. \emph{/ This =
arg(1) /} self of caller */ w= .Bag\textasciitilde new
sc=This\textasciitilde supplier do while sc\textasciitilde available
wlsc\textasciitilde index{]} =sc\textasciitilde index sc-next end
r=This\textsubscript{class}new

cs=MayEnBag (arg(2))\textasciitilde supplier do while
cs\textasciitilde available i=cs\textasciitilde index if
w\textasciitilde hasindex(i) then do r{[}i{]}=This {[}i{]}
w\textasciitilde remove (i) end cs\textasciitilde next end return r

::routine CommonUnion /* Returns a new collection of the same class as
SELF that contains all the items from SELF and items from the argument
that have an index not in the first. \emph{/ /} Best to add them all. By
adding non-receiver first we ensure that receiver takes priority when
same indexes. \emph{/ This = arg(1) /} self of caller */
r=This\textsubscript{class}new cs=MayEnBag
(arg(2))\textasciitilde supplier do while cs\textasciitilde available
r{[}cs\textasciitilde index{]} =cs\textasciitilde item
cs\textasciitilde next end cs=This\textasciitilde supplier do while
cs\textasciitilde available r{[}cs\textasciitilde index{]}
=cs\textasciitilde item cs\textasciitilde next end return r

::routine CommonDifference /* Returns a new collection containing only
those index-item pairs from the SELF whose indexes the other collection
does not contain. \emph{/ This = arg(1) /} self of caller */
r=This\textsubscript{class}new cs=This\textasciitilde supplier do while
cs\textasciitilde available r{[}cs\textasciitilde index{]}
=cs\textasciitilde item es-next end cs=MayEnBag
(arg(2))\textasciitilde supplier do while cs\textasciitilde available
r\textasciitilde remove (cs\textasciitilde index) es-next end return r

::routine MayEnBag

/* For List and Queue the indexes are dropped. */ rearg(1) if r-clags ==
.List \textbar{} r-class == .Queue then r=EnBag(r) return r

::routine EnBag r=.Bag\textasciitilde new
s=arg(1)\textasciitilde supplier do while s\textasciitilde available if
arg(1)\textasciitilde class == .List \textbar{}
arg(1)\textasciitilde class == .Queue then r{[}s\textasciitilde item{]}
=s\textasciitilde item else /* This case is when the receiver is a Bag.
*/ r{[}s\textasciitilde index{]} =s\textasciitilde index
s\textasciitilde-next end return r

\hypertarget{the-collection-class}{%
\subsection{The collection class}\label{the-collection-class}}

::Cclass `Collection'

\hypertarget{init}{%
\subsubsection{INIT}\label{init}}

::method init

expose a /* A collection is modelled as using 3 slots in an array for
each element. The first slot holds the item, the second the index, and
the third is used by particular types of collection. This order of slots
is arbitary, chosen to match order of arguments for PUT and
SUPPLIER\textasciitilde NEW. \emph{/ /} The first set of 3 slots is
reserved for other purposes, to avoid having separate variables which
the subclassing would need to access. */

a=.array\textasciitilde new

a{[}1{]} /\emph{ItemsCount}/=0

a{[}2{]}/\emph{Unique}/=0

return self

\hypertarget{exposed}{%
\subsubsection{EXPOSED}\label{exposed}}

::method exposed private expose a

/* This method allows subclasses to get at the implementation of
Collection. */ return a

\hypertarget{findindex}{%
\subsubsection{FINDINDEX}\label{findindex}}

::method findindex private expose a /* Returns array index if the
collection contains any item associated with the index specified or
returns 0 otherwise. \emph{/ do j=4 by 3 to
1+3}a{[}1{]}/\emph{ItemsCount}/ if alj+l{]}==arg(1) then return j end j
return 0

\hypertarget{at}{%
\subsubsection{AT}\label{at}}

::method at /* vANY \emph{/ expose a /} Returns the item associated with
the specified index. */ j=self\textasciitilde findindex(arg(1)) if j=0
then return .nil return a{[}j{]}

\hypertarget{section}{%
\subsubsection{{[}{]}}\label{section}}

::method `{[}{]}' /* Synonym for the AT method. */ forward message `AT'

\hypertarget{put}{%
\subsubsection{PUT}\label{put}}

::method put /* rANY rANy \emph{/ expose a use arg item, index /}
Replaces any existing item associated with the specified index with the
new item. Otherwise adds the item-index pair. \emph{/
j=self\textasciitilde findindex (index) if j\textgreater0 then do
alj{]}=item return end a{[}1{]} /}ItemsCount\emph{/=a {[}1{]}
/}ItemsCount\emph{/+1 j=1+3}al{[}1{]} /\emph{ItemsCount}/ alj{]}=item
a{[}lj+1{]} =index a{[}j+2{]}=0 return /* Error 91 in OOI if context
requiring result. */

\hypertarget{section-1}{%
\subsubsection{{[}{]}=}\label{section-1}}

:smethod `{[}{]}=' /* Synonym for the PUT method. */ forward message
`PUT'

\hypertarget{hasindex}{%
\subsubsection{HASINDEX}\label{hasindex}}

::method hasindex /* vANY \emph{/ /} Returns 1 (true) if the collection
contains any item associated with the index specified or returns 0
(false) otherwise. */

return self\textasciitilde findindex (arg(1))\textgreater0

\hypertarget{items}{%
\subsubsection{ITEMS}\label{items}}

::method items expose a

/* Returns the number of items in the collection. \emph{/ return
a{[}1{]}/}ItemsCount*/

\hypertarget{remove}{%
\subsubsection{REMOVE}\label{remove}}

::method remove /* vANY */

expose a /* Returns and removes from a collection the member item with
the specified index. */

j=self\textasciitilde findindex(arg(1))

if j=0 then return .nil

r=al{[}j{]}

self\textasciitilde removeit (j)

return r

\hypertarget{removeit}{%
\subsubsection{REMOVEIT}\label{removeit}}

::method removeit private

expose a

use arg j

/* Remove relevant slots from the array, with compaction. */

do j=j+3 by 3 to 14+3\emph{a{[}1{]}/}ItemsCount*/
alj-3l=saljl;alj-2{]}=al{[}j+1{]};alj-ll=alj+2{]} end j

a{[}1{]} /\emph{ItemsCount}/=a {[}1{]} /\emph{ItemsCount}/-1

return

\hypertarget{makearrya}{%
\subsubsection{MAKEARRYA}\label{makearrya}}

::method makearray

expose a /* Returns a single-index array containing the receiver list
items. \emph{/ r= .array\textasciitilde new /} To build result in. */

do j=4 by 3 to 1+3\emph{a{[}1{]}/}ItemsCount*/
r{[}r\textasciitilde dimension(1)+1{]}=a{[}j{]} end j

return r

\hypertarget{makearrayx}{%
\subsubsection{MAKEARRAYX}\label{makearrayx}}

::method makearrayx private expose a /* Returns a single-index array
containing the receiver index items. \emph{/ r=
.array\textasciitilde new /} To build result in. \emph{/ do j=4 by 3 to
1+3}a{[}1{]}/\emph{ItemsCount}/
r{[}r\textasciitilde dimension(1)+1{]}=a{[}j+1{]} end j return r

\hypertarget{supplier}{%
\subsubsection{SUPPLIER}\label{supplier}}

::method supplier expose a

/* Returns a supplier object for the list. */ return
.supplier\textsubscript{new(self}makearray: .collection,
self\textasciitilde makearrayx)

\hypertarget{class-list}{%
\subsection{Class list}\label{class-list}}

::¢class `List' subclass Collection

\hypertarget{put-1}{%
\subsubsection{PUT}\label{put-1}}

::method put /* vANY rANY */ use arg item, index
a=self\textasciitilde exposed

/* PUT for a List must not be an insertion. */
j=self\textasciitilde findindex (index) if j=0 then call Raise
`Syntax',93.918 alj{]}=item return

\hypertarget{of}{%
\subsubsection{OF}\label{of}}

::method of class /* 1 or more oANY Are they omittable? Not in I00
\emph{/ /} Returns a newly created list containing the specified value
objects in the order specified. */ r= self \textasciitilde{} new do j =
1 to arg() r\textasciitilde{} insert (arg(j)) end j return r

\hypertarget{insert}{%
\subsubsection{INSERT}\label{insert}}

::method insert /* rANY oANY */

use arg item, index

a=self\textasciitilde exposed /* Returns a list-supplied index for a new
item, of specified value, which is added to the list. The new item
follows the existing item with the specified index in the list ordering.
\emph{/ /} Establish the index of what preceeds the new element. \emph{/
/} If there was no index given, the new item becomes the last on list.
\emph{/ /} .nil argument means first */

if arg(2,`E') then p=arg(2)

else p=self\textasciitilde last

/* Convert from list index to underlying array index. */

if p==.nil then j=1

else j=self\textasciitilde findindex(p)

if j=0 then call Raise `Syntax',93.918

j=j+3 /* Where new entry will be. \emph{/ /} Move space to required
place. */

a{[}1{]} /\emph{ItemsCount}/=a {[}1{]} /\emph{ItemsCount}/+1

do k=1+3\emph{a{[}1{]}/}ItemsCount*/ by -3 to j+3

a{[}k{]} =a{[}k-3{]} ;a{[}k+1{]} =a{[}k-2{]} ;a{[}k{]} =a{[}k-3{]} end

/* Insert new element. */

alj{]}=item /* A new, unique, index is needed. \emph{/ /} The basic
requirement is for something unique, so this would be correct:

is.object\textasciitilde new /* a unique object, used as a key (the
index on the list) \emph{/ }

/ /* However, a number can be used. (At risk of the user thinking it is
sensible to do arithmetic on it.) */

a{[}j+1{]}=a{[}2{]}/\emph{Unique}/;a{[}2{]} /\emph{Unique}/=a{[}2{]}
/\emph{Unique}/+1

a{[}j+2{]}=0

return a{[}j+1{]}

\hypertarget{first}{%
\subsubsection{FIRST}\label{first}}

::method first a=self\textasciitilde exposed

/* Returns the index of the first item in the list. \emph{/ if
a{[}1{]}/}ItemsCount*/=0 then return .nil

141 return a{[}5{]}

\hypertarget{last}{%
\subsubsection{LAST}\label{last}}

::smethod last a=self\textasciitilde exposed

/* Returns the index of the last item in the list. \emph{/ if
a{[}1{]}/}ItemsCount*/=0 then return .nil return
a{[}3\emph{a{[}1{]}/}ItemsCount*/+2{]}

\hypertarget{firstitem}{%
\subsubsection{FIRSTITEM}\label{firstitem}}

::method firstitem a=self\textasciitilde exposed

/* Returns the first item in the list. \emph{/ if
a{[}1{]}/}ItemsCount*/=0 then return .nil return a{[}4{]}

\hypertarget{lastitem}{%
\subsubsection{LASTITEM}\label{lastitem}}

::method lastitem a=self\textasciitilde exposed

/* Returns the last item in the list. \emph{/ if
a{[}1{]}/}ItemsCount*/=0 then return .nil return
a{[}3\emph{a{[}1{]}/}ItemsCount*/+1{]}

\hypertarget{next}{%
\subsubsection{NEXT}\label{next}}

::method next /* vANY \emph{/ a=self\textasciitilde exposed /} Returns
the index of the item that follows the list item having the specified
index. \emph{/ j=self\textasciitilde findindex(arg(1)) if j=0 then call
Raise `Syntax',93.918 j=j+3 if
j\textgreater3}a{[}1{]}/\emph{ItemsCount}/ then return .nil /* Next of
last was requested. */ return a{[}j+1{]}

\hypertarget{previous}{%
\subsubsection{PREVIOUS}\label{previous}}

::method previous /* vANY */

a=self\textasciitilde exposed /* Returns the index of the item that
precedes the list item having the specified index. */

j=self\textasciitilde findindex(arg(1))

if j=0 then call Raise `Syntax',93.918

j=j-3

if j\textless4 then return .nil /* Previous of first was requested. */

return a{[}j+1{]}

\hypertarget{section-2}{%
\subsubsection{SECTION}\label{section-2}}

::method section /* rANY oWHOLE\textgreater=0 */

a=self\textasciitilde exposed /* Returns a new list containing selected
items from the receiver list. The first item in the new list is the item
corresponding to the index specified, in the receiver list. */

j=self\textasciitilde findindex(arg(1))

if j=0 then call Raise `Syntax',93.918

r= .list\textasciitilde new /* To build result in. \emph{/ if arg(2,`E')
then s = arg(2) else s = self\textasciitilde items; do s
r\textasciitilde insert (a{[}j{]}) j=j+3 if
j\textgreater1+3}a{[}1{]}/\emph{ItemsCount}/ then leave end return r

\hypertarget{class-queue}{%
\subsection{Class queue}\label{class-queue}}

::class `Queue' subclass Collection

/* A queue is a sequenced collection with whole-number indexes. The

indexes specify the position of an item relative to the head (first
item) of the queue. Adding or removing an item changes the association
of an index to its queue item. */

\hypertarget{push}{%
\subsubsection{PUSH}\label{push}}

::method push /* vANY */

/* Adds the object value to the queue at its head. \emph{/
a=self\textasciitilde exposed a{[}1{]} /}ItemCount\emph{/=a {[}1{]}
/}ItemCount*/+1

/* Slide along to make a space. \emph{/ do
j=1+3}a{[}1{]}/\emph{ItemCount}/ by -3 to 7

aljl=alj-3{]} alj+l{]}=a{[}j-2{]}+1; /* Index changes \emph{/ end j
a{[}4{]}=arg(1) a{[}5J=1 return 12.2.4.2 PULL{]}=a{[}j+4{]}-1; /} Index
changes */ end j return r

\hypertarget{queue}{%
\subsubsection{QUEUE}\label{queue}}

::method queue /* vANY \emph{/ /} Adds the object value to the queue at
its tail. \emph{/ a=self\textasciitilde exposed a{[}1{]}
/}ItemCount\emph{/=a {[}1{]} /}ItemCount\emph{/+1 a{[}1+3\emph{a{[}1{]}
/}ItemCount*/{]}=arg(1) a{[}2+3\emph{a{[}1{]}
/}ItemCount*/{]}=a{[}1{]}/}ItemCount*/ return

\hypertarget{peek}{%
\subsubsection{PEEK}\label{peek}}

::method peek

a=self\textasciitilde exposed /* Returns the item at the head of the
queue. The collection remains unchanged. */

return a{[}4{]}

\hypertarget{remove-1}{%
\subsubsection{REMOVE}\label{remove-1}}

::method remove /* rcWHOLE\textgreater O \emph{/ /} Returns and removes
from a collection the member item with the specified index. \emph{/
a=self\textasciitilde exposed if
a{[}1{]}/}ItemCount\emph{/\textless arg(1) then return .nil
r=self\textasciitilde remove: super (arg(1)) /} Reset the indexes.
\emph{/ k=0 do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ k=k+1
alj+l{]}=k end j return r

\hypertarget{class-table}{%
\subsection{Class table}\label{class-table}}

::Class `Table' subclass Collection

\hypertarget{makearray}{%
\subsubsection{MAKEARRAY}\label{makearray}}

::method makearray /* Returns a single-index array containing the index
objects. \emph{/ /} This is different from Collection MAKEARRAY where
items rather than indexes are in the returned array. */ forward message
`MAKEARRAYX'

\hypertarget{union}{%
\subsubsection{UNION}\label{union}}

::method union /* rCOLLECTION */ return CommonUnion (self,arg(1))

\hypertarget{intersection}{%
\subsubsection{INTERSECTION}\label{intersection}}

::method intersection /* rCOLLECTION */ return CommoniIntersect
(self,arg(1))

\hypertarget{xor}{%
\subsubsection{XOR}\label{xor}}

::method xor /* rCOLLECTION */ return CommonXor (self,arg(1))

\hypertarget{difference}{%
\subsubsection{DIFFERENCE}\label{difference}}

::method difference /* rCOLLECTION */ return
CommonDifference(self,arg(1))

\hypertarget{subset}{%
\subsubsection{SUBSET}\label{subset}}

::method subset /* rCOLLECTION */ return
self\textsubscript{difference(arg(1))}items = 0

\hypertarget{class-set}{%
\subsubsection{Class set}\label{class-set}}

::¢class `Set' subclass table

/* A set is a collection that restricts the member items to have a value
that is the same as the index. Any object can be placed in a set. There
can be only

one occurrence of any object in a set. */

\hypertarget{put-2}{%
\subsubsection{PUT}\label{put-2}}

/* Second arg same as first. Committee has dropped second? \emph{/
::method put /} YANY oANY \emph{/ /} Makes the object value a member
item of the collection and associates it with specified index. \emph{/
if arg(2,`E') then if arg(2)==arg(1) then signal error /} 949 */
self\textasciitilde put:super(arg(1),arg(1))

\hypertarget{of-1}{%
\subsubsection{OF}\label{of-1}}

::method of class /* 1 or more rANYy \emph{/ /} Returns a newly created
set containing the specified value objects. */ r=self\textasciitilde new
do j=1 to arg() r\textasciitilde put (arg(j)) end j return r

\hypertarget{union-1}{%
\subsubsection{UNION}\label{union-1}}

::method union /* rCOLLECTION */ return CommonUnion (self,
EnBag(arg(1)))

\hypertarget{intersection-1}{%
\subsubsection{INTERSECTION}\label{intersection-1}}

::method intersection /* rCOLLECTION */ return CommoniIntersect
(self,EnBag(arg(1)))

\hypertarget{xor-1}{%
\subsubsection{XOR}\label{xor-1}}

::method xor /* rCOLLECTION */ return CommonXor (self, EnBag(arg(1)))

\hypertarget{difference-1}{%
\subsubsection{DIFFERENCE}\label{difference-1}}

::method difference /* rCOLLECTION */ return CommonDifference (self,
EnBag(arg(1)))

\hypertarget{class-relation}{%
\subsection{Class relation}\label{class-relation}}

::Class `Relation' subclass Collection

\hypertarget{put-3}{%
\subsubsection{PUT}\label{put-3}}

::method put /* vANY rANY */

use arg item, index

a=self\textasciitilde exposed /* Makes the object value a member item of
the relation and associates it with the specified index. If the relation
already contains any items with the specified index, this method adds a
new member item value with the same index, without removing any existing
members */

a{[}1{]} /\emph{ItemsCount}/=a {[}1{]} /\emph{ItemsCount}/+1

j=1+3\emph{al{[}1{]} /}ItemsCount*/

alj{]}=item

a{[}lj+1{]} =index

a{[}j+2{]}=0

return /* Error 91 in OOI if context requiring result. */

\hypertarget{items-1}{%
\subsubsection{ITEMS}\label{items-1}}

::method items /* oANY \emph{/ a=self\textasciitilde exposed /} Returns
the number of relation items with the specified index. If you specify no
index, this method returns the total number of items associated with all
indexes in the relation. \emph{/ if \arg(1,`E') then return
a{[}1{]}/}ItemsCount\emph{/ n=0 do j=4 by 3 to
1+3}a{[}1{]}/\emph{ItemsCount}/ if arg(1)==a{[}j+1{]} then n=n+1 end j
return n

\hypertarget{makearray-1}{%
\subsubsection{MAKEARRAY}\label{makearray-1}}

::method makearray forward message `MAKEARRAYX'

\hypertarget{supplier-1}{%
\subsubsection{SUPPLIER}\label{supplier-1}}

::method supplier /* oANY \emph{/ a=self\textasciitilde exposed /}
Returns a supplier object for the collection. If an index is specified,
the supplier enumerates all of the items in the relation with the
specified index. \emph{/ m=.array\textasciitilde new /} For the items
\emph{/ r=.array\textasciitilde new /} For the indexes \emph{/ do j=4 by
3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if arg(1,`E') then if
arg(1)=s=a{[}j+1{]} then iterate n=r\textasciitilde dimension(1)+1

m{[}n{]} =a{[}j{]} r{[}n{]} =a{[}j+1{]}

end j return .supplier\textasciitilde new(m,r) 12.2.7.5 UNION ::method
union /* rCOLLECTION */

/* Union for a relation is just all of both. */
r=self\textsubscript{class}new cs=self\textasciitilde supplier do while
cs\textasciitilde available

r{[}cs\textasciitilde index{]} =cs\textasciitilde item es-next end
cs=MayEnBag (arg(1))\textasciitilde supplier do while
cs\textasciitilde available r{[}cs\textasciitilde index{]}
=cs\textasciitilde item es-next end return r

\hypertarget{intersection-2}{%
\subsubsection{INTERSECTION}\label{intersection-2}}

::method intersection /* rCOLLECTION */

/* Intersection for a relation requires the items as well as the keys to
match. */

r=self\textsubscript{class}new sc=self\textsubscript{class}new
cs=self\textasciitilde supplier do while cs\textasciitilde available sc
{[}cs\textasciitilde index{]} =cs\textasciitilde item
cs\textasciitilde next end cs=MayEnBag (arg(1))\textasciitilde supplier
do while cs\textasciitilde available if
sc\textsubscript{hasitem(cs}item,cs\textasciitilde index) then r
{[}ecs\textasciitilde index{]} =sc\textsubscript{removeitem(cs}item,
cs\textasciitilde next end return r

cs\textasciitilde index)

\hypertarget{xor-2}{%
\subsubsection{XOR}\label{xor-2}}

::method xor /* rCOLLECTION \emph{/ /} Returns a new relation that
contains all items from self and

the argument except that all index-item pairs that appear in both
collections are removed. */

r=self\textsubscript{class}new cs=self\textasciitilde supplier do while
cs\textasciitilde available r{[}cs\textasciitilde index{]}
=cs\textasciitilde item cs\textasciitilde next end cs=MayEnBag
(arg(1))\textasciitilde supplier do while cs\textasciitilde available if
self\textsubscript{hasitem(cs}item,cs\textasciitilde index) then
r\textsubscript{removeitem(cs}item, cs\textasciitilde index) else
r{[}cs\textasciitilde index{]} =cs\textasciitilde item
cs\textasciitilde next end return r

\hypertarget{difference-2}{%
\subsubsection{DIFFERENCE}\label{difference-2}}

::method difference /* rCOLLECTION */

/* Returns a new relation containing only those index-item pairs from
the g Y P SELF whose indexes the other collection does not contain. */
r=self\textsubscript{class}new cs=self\textasciitilde supplier

do while cs\textasciitilde available r{[}cs\textasciitilde index{]}
=cs\textasciitilde item cs\textasciitilde next end

cs=MayEnBag (arg(1))\textasciitilde supplier

do while cs\textasciitilde available r\textsubscript{removeitem(cs}item,
cs\textasciitilde index) cs\textasciitilde next end

return r

\hypertarget{subset-1}{%
\subsubsection{SUBSET}\label{subset-1}}

::method subset /* rCOLLECTION */ return
self\textsubscript{difference(arg(1))}items = 0

\hypertarget{removeitem}{%
\subsubsection{REMOVEITEM}\label{removeitem}}

::method removeitem /* YANY rANY \emph{/ a=self\textasciitilde exposed
/} Returns and removes from a relation the member item value (associated
with the specified index). If value is not a member item associated with
index index, this method returns the NIL object and removes no item.
\emph{/ do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if alj{]}==arg(1)
\& a{[}j+1l{]}==arg(2) then do self\textasciitilde removeit (j) return
arg(1) end end j return .nil

\hypertarget{index}{%
\subsubsection{INDEX}\label{index}}

::method index /* vANY \emph{/ a=self\textasciitilde exposed /} Returns
the index for the specified item. If there is more than one index
associated with the specified item, the one this method returns is not
defined. \emph{/ do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if
arg(1)==a{[}j{]} then return a{[}j+1{]} end j return .nil

\hypertarget{allat}{%
\subsubsection{ALLAT}\label{allat}}

::method allat /* vANY \emph{/ a=self\textasciitilde exposed /} Returns
a single-index array containing all the items associated with the
specified index. \emph{/ r=.array\textasciitilde new do j=4 by 3 to
1+3}a{[}1{]}/\emph{ItemsCount}/ if arg(1)==a{[}j+1{]} then
r{[}r\textasciitilde dimension(1)+1{]} = aljl end j return r

\hypertarget{hasitem}{%
\subsubsection{HASITEM}\label{hasitem}}

::method hasitem /* YANY rANY \emph{/ a=self\textasciitilde exposed /}
Returns 1 (true) if the relation contains the member item value
(associated with specified index). Returns 0 (false) otherwise. \emph{/
do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if alj{]}==arg(1) \&
a{[}j+l{]}==arg(2) then return 1 end j return 0

\hypertarget{allindex}{%
\subsubsection{ALLINDEX}\label{allindex}}

::method allindex /* vANY */

a=self\textasciitilde exposed /* Returns a single-index array containing
all indexes for the specified item. \emph{/ r=.array\textasciitilde new
do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if alj{]}==arg(1) then do
r{[}r-\textasciitilde dimension (1)+1{]} =a{[}j+1{]} end end j return r

\#\#3 The bag class

::class `Bag' subclass relation

/* A bag is a collection that restricts the member items to having a
value that is the same as the index. Any object can be placed in a bag,
and the same object can be placed in a bag multiple times. */

\hypertarget{of-2}{%
\subsubsection{OF}\label{of-2}}

::method of class /* 1 or more rANYy \emph{/ /} Returns a newly created
bag containing the specified value objects. */ r=self\textasciitilde new
do j=1 to arg() r\textasciitilde put (arg(j)) end j return r

\hypertarget{put-4}{%
\subsubsection{PUT}\label{put-4}}

::method put /* vANY oANY \emph{/ /} Committee does away with second
argument? \emph{/ /} Makes the object value a member item of the
collection and associates it with the specified index. If you specify
index, it must be the same as value. */ if arg(2,`E') then if
arg(2)=s=arg(1) then signal error self\textasciitilde put: super
(arg(1),arg(1))

\hypertarget{union-2}{%
\subsubsection{UNION}\label{union-2}}

::method union /* rCOLLECTION */ return CommonUnion (self,
EnBag(arg(1)))

\hypertarget{intersection-3}{%
\subsubsection{INTERSECTION}\label{intersection-3}}

::method intersection /* rCOLLECTION */ return CommoniIntersect
(self,EnBag(arg(1)))

\hypertarget{xor-3}{%
\subsubsection{XOR}\label{xor-3}}

::method xor /* rCOLLECTION */ return CommonXor (self, EnBag(arg(1)))

\hypertarget{difference-3}{%
\subsubsection{DIFFERENCE}\label{difference-3}}

::method difference /* rCOLLECTION */ return CommonDifference (self,
EnBag(arg(1)))

\hypertarget{the-directory-class}{%
\subsection{The directory class}\label{the-directory-class}}

::class `Directory' subclass Collection

\hypertarget{at-1}{%
\subsubsection{AT}\label{at-1}}

::method at /* vANY */ a=self\textasciitilde exposed

/* Returns the item associated with the specified index. */
j=self\textasciitilde findindex(arg(1))

if j=0 then return .nil

/* Run the method if there is one. */ if a{[}j+2{]} then return
self\textasciitilde run(a{[}j{]}) return a{[}j{]}

\hypertarget{put-5}{%
\subsubsection{PUT}\label{put-5}}

::method put /* vANY rANY \emph{/ a=self\textasciitilde exposed /} Makes
the object value a member item of the collection and associates it with
the specified index. */ if
\arg(2)\textasciitilde hasmethod(`MAKESTRING') then call Raise `Syntax',
93.938 self\textasciitilde put:super(arg(1),arg(2)
\textasciitilde makestring) return

\hypertarget{makearray-2}{%
\subsubsection{MAKEARRAY}\label{makearray-2}}

::method makearray forward message `MAKEARRAYX'

\hypertarget{supplier-2}{%
\subsubsection{SUPPLIER}\label{supplier-2}}

::method supplier a=self\textasciitilde exposed /* Returns a supplier
object for the directory. \emph{/ /} Check out what happens to the
SETENTRY fields. \emph{/ r=.array\textasciitilde new /} For items
\emph{/ do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/
r{[}r\textasciitilde dimension(1)+1{]}=a{[}j{]}

end j return .supplier\textsubscript{new(r,self}makearray) \#\#\#\#
UNION ::method union /* rCOLLECTION */

return CommonUnion (self,arg(1))

\hypertarget{intersection-4}{%
\subsubsection{INTERSECTION}\label{intersection-4}}

::method intersection /* rCOLLECTION */ return CommoniIntersect
(self,arg(1))

\hypertarget{xor-4}{%
\subsubsection{XOR}\label{xor-4}}

::method xor /* rCOLLECTION */ return CommonXor (self,arg(1))

\hypertarget{difference-4}{%
\subsubsection{DIFFERENCE}\label{difference-4}}

::method difference /* rCOLLECTION */ return
CommonDifference(self,arg(1))

\hypertarget{subset-2}{%
\subsubsection{SUBSET}\label{subset-2}}

::method subset /* rCOLLECTION */ return
self\textsubscript{difference(arg(1))}items = 0

\hypertarget{setentry}{%
\subsubsection{SETENTRY}\label{setentry}}

::smethod setentry /* rSTRING oANY */

a=self\textasciitilde exposed /* Sets the directory entry with the
specified name (translated to uppercase) to the second argument,
replacing any existing entry or method for the specified name. */

n=translate(arg(1))

j=self\textasciitilde findindex(n)

if j=0 \& \arg(2,`E') then return

if \arg(2,`E') then do /* Removal */

self\textasciitilde removeit (j)

return end

if j=0 then do /* It's new \emph{/
a{[}1{]}/}ItemsCount\emph{/=a{[}1{]}/}ItemsCount\emph{/ +1
j=1+3}al{[}1{]} /\emph{ItemsCount}/ alj+l{]}=n end

alj{]}=arg(2)

a{[}4j+2{]}=0

return

\hypertarget{entry}{%
\subsubsection{ENTRY}\label{entry}}

::smethod entry /* rSTRING \emph{/ a=self\textasciitilde exposed /}
Returns the directory entry with the specified name (translated to
uppercase). \emph{/ n=translate(arg(1))
j=self\textasciitilde findindex(n) /}if j=0 then signal error according
to online \emph{/ /} Online has something about running UNKNOWN. \emph{/
if j=0 then return .nil /} If there is an entry decide whether to invoke
it. */ if a\textasciitilde hasindex(j) then do if \al[j+2] then return
al{[}j{]} return self\textasciitilde run(al{[}jl{]}) end

\hypertarget{hasentry}{%
\subsubsection{HASENTRY}\label{hasentry}}

::method hasentry /* rSTRING \emph{/ /} Returns 1 (true) if the
directory has an entry or a method for the specified name (translated to
uppercase) or 0 (false) otherwise. */

return self\textasciitilde findindex (translate(arg(1)))\textgreater0

\hypertarget{setmethod}{%
\subsubsection{SETMETHOD}\label{setmethod}}

::method setmethod /* rSTRING oMETHOD \emph{/
a=self\textasciitilde exposed /} Associates entry with the specified
name (translated to uppercase) with method method. Thus, the language
processor returns the result of running method when you access this
entry. \emph{/ /} (Part of METHOD checking converts string or array to
actual method.) \emph{/ n=translate(arg(1))
j=self\textasciitilde findindex(n) if j=0 \& \arg(2,`E') then return if
\arg(2,`E') then do self\textasciitilde removeit (j) return end if j=0
then do /} It's new \emph{/
a{[}1{]}/}ItemsCount\emph{/=a{[}1{]}/}ItemsCount\emph{/ +1
j=1+3}al{[}1{]} /\emph{ItemsCount}/ alj+l{]}=n end alj{]}=arg(2)
a{[}j+2{]}=1 return

\hypertarget{unknown}{%
\subsubsection{UNKNOWN}\label{unknown}}

::method unknown /* rSTRING rARRAY */

/* Runs either the ENTRY or SETENTRY method, depending on whether the
message name supplied ends with an equal sign. If the message name does
not end with an equal sign, this method runs the ENTRY method, passing
the message name as its argument. */

if right (arg(1),1)=s=`=' then

return self\textasciitilde entry(arg(1)) /* 2?? Not clear whether second
argument is mandatory. */ t=.nil

if arg(2,`E') then t=arg(2) {[}1{]} self\textasciitilde setentry (left
(arg(1),length(arg(1))-1),t) 12.3 The stem class For some reason, the
stem class doesn't have PUT and AT methods, which stops us having a
general rule about {[}{]} synonyms AT, {[}{]}= synonyms PUT. Anyway,
committee doing without this class as such.

Here is temporary stuff showing how to use algebra in the collection
coding.

/* This 1998 version uses Rony's rules for XOR and INTERSECTION based on
UNION and DIFFERENCE */

/* Test Set-Operator-Methods on different collection objects */

/* This top part has some rough parts - not meant for standard. */

/* The dumps put out results sorted, so that comparisons can be made
between implementations that keep collections in different orders. */

/* Invocation example: settest.cmd 1\textgreater{} tmp.res
2\textgreater{} tmp.err */

/* Jnitial verification that new definitions are in effect */ J18list =
.List\textasciitilde new if \J18list\textasciitilde hasmethod(``J18'')
then signal error

/* Input collections used for the tests */

coll.1 = .array\textasciitilde of(1, 2,, 4)

coll.2 = list\textasciitilde of(2, 3, 6)

coll.3 = .queue\textsubscript{newPUSH(2)PUSH(3)}\textasciitilde PUSH(7)

coll.4 = .directory\textsubscript{new}\textasciitilde setentry(1,
``eins'')\textasciitilde\textasciitilde setentry(3, ``drei'')

coll.5 = .bag\textsubscript{newput(2)put(3)put(5)}\textasciitilde put(2)

coll.6 =
.relation\textsubscript{new}\textasciitilde{}``{[}J=''(''zwei'',
2)\sout{''{[}{]}=''('\,''drei'', 3)}''{[}J=''('vier'',
8)\textasciitilde\textasciitilde''J=''C''drei'',3) coll.7 =
.set\textasciitilde of(2, 3, 9){]}=tmpSupp\textasciitilde ITEM
tmpSupp\textasciitilde NEXT END do until hope hope=1 do j=1 to
k\textasciitilde dimension(1)-1 if
k{[}j{]}\textsubscript{string\textgreater k{[}j+1{]}}string \textbar,
(k{[}j{]}\textsubscript{string=k{[}j+1{]}}string \&
i{[}j{]}\textsubscript{string\textless i{[}j+1{]}}string) then do{]}=t
hope=0 end end end if O=collection\textasciitilde items then say'' The
result is empty!'' else do j=1 to k\textasciitilde dimension(1) SAY ''
'' ``index'' pp(k{[}j{]}) ``item'' ppd{[}j){]}=alj+1{]};a{[}j- 1
{]}=alj+2{]} end j

a{[}1{]}/\emph{ItemsCount}/=a{[} 1 {]}/\emph{ItemsCount}/-1

return

::method makearray expose a /* Returns a single-index array containing
the receiver list items. \emph{/ r=.array\textasciitilde new = /} To
build result in. \emph{/ do j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/
r{[}r\textasciitilde dimension(1)+1{]}=a{[}j{]} end j returm r

::method makearrayx private expose a

/* Returns a single-index array containing the receiver index items.
\emph{/ r=.array\textasciitilde new = /} To build result in. \emph{/ do
j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/

r{[}r\textasciitilde dimension(1)+1{]}=a{[}j+1{]} end j return r

:imethod supplier expose a /* Returns a supplier object for the list. */
return
.supplier\textsubscript{new(self}makearray:.collection,self\textasciitilde makearrayx)

::class `List' subclass Collection

zimethod J18 = /* Here to demonstrate .LIST is replaced */ return

/* List and Queue are special because there is an order to their
elements. */

::method put /* rANY rANY */ use arg item, index
a=self\textasciitilde exposed

/* PUT for a List must not be an insertion. */
j=self\textasciitilde findindex(index) if j=0 then call Raise
`Syntax',93.918 a{[}j{]}=item retum

zimethod of class /* 1 or more oANY Are they omittable? Not in IOO
\emph{/ /} Returns a newly created list containing the specified value
objects in the order specified. */ r=self \textasciitilde{} new do j = 1
to argQ) r \textasciitilde{} insert(arg(j)) end j return r

zimethod insert /* rANY oANY */

use arg item, index

a=self\textasciitilde exposed /* Returns a list-supplied index for a new
item, of specified value, which is added to the list. The new item
follows the existing item with the specified index in the list ordering.
\emph{/ /} Establish the index of what preceeds the new element. \emph{/
/} Tf there was no index given, the new item becomes the last on list.
\emph{/ /} mil argument means first */

if arg(2,`E') then p=arg(2)

else p=self\textasciitilde last

/* Convert from list index to underlying array index. */

if p==.nil then j=1 else j=self\textasciitilde findindex(p)

if j=0 then call Raise `Syntax',93.918

j=J+3 /* Where new entry will be. \emph{/ /} Move space to required
place. */

a{[}1{]}/\emph{ItemsCount}/=a{[}1 {]}/\emph{ItemsCount}/+1

do k=1+3\emph{a{[}1{]}/}ItemsCount*/ by -3 to j+3

a{[}k{]}=a{[}k-3{]};a{[}k+1 {]}=a{[}k-2{]};a{[}k{]}=a{[}k-3{]} end

/* Insert new element. */

a{[}j{]}=item /* A new, unique, index is needed. \emph{/ /} The basic
requirement is for something unique, so this would be correct:

i=.object\textasciitilde new /* a unique object, used as a key (the
index on the list) \emph{/ }/ /* However, a number can be used. (At risk
of the user thinking it is sensible to do arithmetic on it.) */

a{[}j+1{]}=a{[}2{]}/\emph{Unique}/;a{[}2{]}/\emph{Unique}/=a{[}2
{]}/\emph{Unique}/+1

a{[}j+2{]}=0

return a{[}j+1{]}

:smethod first a=self\textasciitilde exposed

/* Returns the index of the first item in the list. \emph{/ if
a{[}1{]}/}ItemsCount*/=0 then return .nil return a{[}5{]}

::method last a=self\textasciitilde exposed

/* Returns the index of the last item in the list. \emph{/ if
a{[}1{]}/}ItemsCount*/=0 then return .nil return
a{[}3\emph{a{[}1{]}/}ItemsCount*/+2{]}\textbar{}

::method firstitem a=self\textasciitilde exposed

/* Returns the first item in the list. \emph{/ if
a{[}1{]}/}ItemsCount*/=0 then return .nil return a{[}4{]}

::method lastitem a=self\textasciitilde exposed

/* Returns the last item in the list. \emph{/ if
a{[}1{]}/}ItemsCount\emph{/=0 then return .nil return
a{[}3\emph{a{[}1{]}/}ItemsCount*/+1 \textbar{]}
a{[}j+1{]}=al{[}j+4{]}-1; /} Index changes */ end j returm r

zimethod queue =/* rANY */

/* Adds the object value to the queue at its tail. \emph{/
a=self\textasciitilde exposed a{[}1{]}/}ItemCount\emph{/=a{[}1
\textbar/}ItemCount*/+1
a{[}\{1+3\emph{a{[}1{]}/}ItemCount*/\textbar=arg(1)
a{[}2+3\emph{a{[}1{]}/}ItemCount\emph{/\textbar=al{[} 1 {]}/}ItemCount*/
return{]} r{[}n{]}=a{[}j+1{]} end j return
.supplier\textasciitilde new(m,r){]}=1

returm

:imethod unknown =/* rSTRING rARRAY */

/* Runs either the ENTRY or SETENTRY method, depending on whether the
message name supplied ends with an equal sign. If the message name does
not end with an equal sign, this method runs the ENTRY method, passing
the message name as its argument. */ if right(arg(1),1)==`=``' then

return self\textasciitilde entry(arg(1)) /* 22 Not clear whether second
argument is mandatory. */ t=.nil

if arg(2,`E') then t=arg(2){[}1{]}
self\textasciitilde setentry(left(arg(1),length(arg(1))-1),t)

routine CommonXor /* Returns a new collection that contains all items
from self and the argument except that all indexes that appear in both
collections are removed. \emph{/ /} When the target is a bag, there may
be an index in the bag that is duplicated and the same value as an index
in the argument. Should one copy of the index survive in the bag? */
Ihs=arg(1)\textasciitilde difference(arg(2))
rhs=Cast(arg(1),MayEnBag(arg(2)))\textasciitilde difference(arg(1))
return lhs\textasciitilde union(rhs)

nroutine CommonUnion /* Returns a new collection of the same class as
SELF that contains all the items from SELF and items from the argument
that have an index not in the first. \emph{/ /} Best to add them all. By
adding non-receiver first we ensure that receiver takes priority when
same indexes. \emph{/ This = arg(1) /} self of caller */
r=This\textsubscript{class}new
cs=MayEnBag(arg(2))\textasciitilde supplier do while
cs\textasciitilde available
r{[}cs\textasciitilde index{]}=cs\textasciitilde item
cs\textasciitilde next end cs=This\textasciitilde supplier do while
cs\textasciitilde available
r{[}cs\textasciitilde index{]}=cs\textasciitilde item
cs\textasciitilde next end return r

nroutine CommonDifference /* Returns a new collection containing only
those index-item pairs from the SELF whose indexes the other collection
does not contain. \emph{/ This = arg(1) /} self of caller */
r=This\textsubscript{class}new cs=This\textasciitilde supplier do while
cs\textasciitilde available
r{[}cs\textasciitilde index{]}=cs\textasciitilde item
cs\textasciitilde next end cs=MayEnBag(arg(2))\textasciitilde supplier

do while cs\textasciitilde available r\textsubscript{remove(cs}index)
cs\textasciitilde next end

return r

routine MayEnBag

/* For List and Queue the indexes are dropped. */ r=arg(1) if
r\textasciitilde class == .List \textbar{} r\textasciitilde class ==
.Queue then r=EnBag(r) retum r

routine EnBag r=.Bag\textasciitilde new s=arg(1)\textasciitilde supplier
do while s\textasciitilde available if arg(1)\textasciitilde class ==
.List \textbar{} arg(1)\textasciitilde class == .Queue then
1{[}s\textasciitilde item{]}=s\textasciitilde item else /* This case is
when the receiver is a Bag. */
1{[}s\textasciitilde index{]}=s\textasciitilde index
s\textasciitilde next end return r

/* This Cast routine commented away, since replaced by Oct 98 Rony
version. routine Cast public use arg Target, Other TmpColl =
Target\textsubscript{class}new /* Create an instance of type Target
\emph{/ TmpSupp = Other\textasciitilde supplier /} Get supplier from
Other */ signal on syntax do while TmpSupp\textasciitilde available
TmpColl{[}TmpSupp\textasciitilde index{]} = TmpSupp\textasciitilde item
TmpSupp\textasciitilde next end return TmpColl

/* Tf syntax error 93.949, then target is an index-only collection like
a set.*/ syntax:

if condition( ``O'' )\textasciitilde code = ``93.949'' then signal
IndexOnly

raise propagate /* Unhandled syntax error, raise in caller */

IndexOnly: /* This for index-only collections. */ do while
TmpSupp\textasciitilde available
TmpColl{[}TmpSupp\textasciitilde index{]} = TmpSupp\textasciitilde index
TmpSupp\textasciitilde next

end return TmpColl End commented away */

/* 98-09-24, ---ref; CAST2.CMD return a collection of type ``target''
which collected all item/index pairs of the argument ``other'' */

2: ROUTINE cast PUBLIC USE ARG target, other

SIGNAL ON SYNTAX IF ~other \textasciitilde{} HASMETHOD( ``SUPPLIER'' )
THEN RAISE SYNTAX 98.907 ARRAY (``COLLECTION
(i.e.~argument2=`other'-object must have a `SUPPLIER'-method)'' )

tmpColl = target \textasciitilde{} CLASS \textasciitilde{} NEW /* create
a an instance of type target \emph{/ tmpSupp = other\textasciitilde{}
SUPPLIER /} get supplier from other */

/* is index of ``other'' usable ? */ bIndexUsable = other
\textasciitilde{} HASMETHOD( ``UNION'' )

IF .Debug = .true THEN IF ~bIndexUsable THEN SAY'' /// index of `other'
not usable for setlike-operations''

/* possible syntax-error, if index and item must have the same value,
e.g.~for sets/bags \emph{/ SIGNAL ON SYNTAX NAME INDEX ONLY target
\textasciitilde{} CLASS \textasciitilde{} NEW \textasciitilde{} PUT(
1,2) /} test, if target-type is index-only */

SIGNAL ON SYNTAX DO WHILE tmpSupp \textasciitilde{} AVAILABLE IF
bIndexUsable THEN tmpColl{[} tmpSupp \textasciitilde{} INDEX {]} =
tmpSupp \textasciitilde{} ITEM ELSE tmpColl{[} tmpSupp \textasciitilde{}
ITEM {]} = tmpSupp \textasciitilde{} ITEM tmpSupp \textasciitilde{} NEXT
END RETURN tmpColl

INDEX\_ONLY : /* this is for index-only collections (e.g.~sets, bags) */
SIGNAL ON SYNTAX IF .Debug = .true THEN SAY'' \textbackslash{}`target'
is an index-only collection (index==item)'' DO WHILE tmpSupp
\textasciitilde{} AVAILABLE

IF bIndexUsable THEN tmpColl{[} tmpSupp \textasciitilde{} INDEX {]} =
tmpSupp \textasciitilde{} INDEX ELSE tmpColl{[} tmpSupp
\textasciitilde{} ITEM {]} = tmpSupp \textasciitilde{} ITEM tmpSupp
\textasciitilde{} NEXT END RETURN tmpColl

SYNTAX: RAISE PROPAGATE /* raise error in caller */

\hypertarget{the-stream-class}{%
\section{The stream class}\label{the-stream-class}}

The stream class provides input/output on external streams. ::class
stream

::method init /* rString \emph{/ Initializes a stream object for a
stream named name, but does not open the stream. ::smethod query /}
keywords */

There is also QUERY as command with method COMMAND.

Used with options, the QUERY method returns specific information about a
stream. ::method charin

::smethod charout

::method chars

::method linein

::method lineout

::method lines

::method qualify

::method command /* rString */

Returns a string after performing the specified stream command. ::method
open

There is also OPEN as command with method COMMAND.

Opens the stream to which you send the message and returns ``READY:''.

Committee dropping OPEN POSITION QUERY SEEK as methods in favour of
command use. ::method state

Returns a string that indicates the current state of the specified
stream. ::method say

::method uninit

::method position /* Ugh \emph{/ POSITION is a synonym for SEEK.
::method seek /} Ugh */

Sets the read or write position a specified number (offset) within a
persistent stream.

::method flush

Returns ``READY:''. Forces any data currently buffered for writing to be
written to the stream receiving the message.

There is also FLUSH as command with method COMMAND.

Committee dropping FLUSH.

::method close

Closes the stream that receives the message.

There is also CLOSE as command with method COMMAND.

Semantics are `seen by other thread'. ::method string

::method makearray /* rCHARLINE */ Returns a fixed array that contains
the data from the stream in line or character format, starting from the

current read position. ::method supplier

Returns a supplier object for the stream. ::method description

::smethod arrayin /* rCHARLINE */

Mixed case value works on OOI.

Committee dropping Arrayin \& Arrayout. Arrayin == MakeArray

Returns a fixed array that contains the data from the stream in line or
character format, starting from the

current read position. ::smethod arrayout /* rARRAY rCHARLINE */

Returns a stream object that contains the data from array.

\hypertarget{the-alarm-class}{%
\section{The alarm class}\label{the-alarm-class}}

::class alarm

::method init /* Time, Msg */ Sets up an alarm for a future time atime.
::method cancel

Cancels the pending alarm request represented by the receiver. This
method takes no action if the specified time has already been reached.

\hypertarget{the-monitor-class}{%
\section{The monitor class}\label{the-monitor-class}}

The Monitor class forwards messages to a destination object.

-local {[}`OUTPUT'{]} = .monitor\textasciitilde new(.output)

::class monitor

\hypertarget{init-1}{%
\subsection{INIT}\label{init-1}}

Initializes the newly created monitor object.

::method init /* oDESTINATION */ expose Destination Destination =
.queue\textasciitilde new if arg(1,`E') then
Destination\textasciitilde push (arg(1)) return

\hypertarget{current}{%
\subsection{CURRENT}\label{current}}

Returns the current destination object.

::smethod current expose Destination return Destination {[}1{]}

\hypertarget{destination}{%
\subsection{DESTINATION}\label{destination}}

Returns a new destination object.

::method destination /* oDESTINATION */ expose Destination if arg(1,`E')
then Destination\textasciitilde push (arg(1)) else
Destination\textasciitilde pull return Destination {[}1{]}

\hypertarget{unknown-1}{%
\subsection{UNKNOWN}\label{unknown-1}}

Reissues or forwards to the current monitor destination all unknown
messages sent to a monitor object

::method unknown expose Destination

Extra parens needed here in original OREXX syntax

forward to destination{[}1{]} message arg(1) arguments arg(2) return

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/rationale.tex
complete, but very deficient, extended standard

\hypertarget{rationale}{%
\chapter{Rationale}\label{rationale}}

This annex explains some of the decisions made by the committee that
drafted this standard, and assists in the understanding of this
document. Some of the statements made here are opinions rather than
facts. These should be interpreted as if prefixed by ``In the opinion of
the X3J18 committee\ldots{}''.

The language described in this standard is, almost entirely, a
compatible extension of the language described by the third reference of
Annex C, which we call ``Classic Rexx''.

The extension allows programs to be written in a less monolithic
fashion; ``Directives'' are introduced to allow one file to contain
several executable units and to allow a program to be written as several
files. The functional extension centers on the addition of objects.
Unlike the individual strings which are the data of Classic Rexx, an
object may be composite. The use of identifiers to reference objects is
an indirect reference, that is two identifiers may refer to the same
object. Classic Rexx avoided any aliasing, even to the extent having
by-reference parameters, to promote simple error free programming. In
the years since Rexx originated the problems tackled by programmers have
become more complex and data structures larger, so that the benefit of
simplicity is outweighed by the power of assignment semantics that are
not simply copying all the data.

Even with the addition of references Rexx remains a typeless language,
in the sense that the programmer need not consider underlying hardware
formats such as LONG or FLOAT representations. Object Rexx does have
classes, which are the hardware independent analogy to types. The class
of an object corresponds to the operations that can be performed upon
it.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/scope_purpose_and_application.tex
complete, but very deficient, extended standard

\hypertarget{scope-purpose-and-application}{%
\chapter{Scope, purpose, and
application}\label{scope-purpose-and-application}}

\hypertarget{scope}{%
\section{Scope}\label{scope}}

This standard specifies the semantics and syntax of the programming
language Rexx by specifying requirements for a conforming language
processor. The scope of this standard includes

\begin{itemize}
\item
  the syntax and constraints of the Rexx language;
\item
  the semantic rules for interpreting Rexx programs;
\item
  the restrictions and limitations that a conforming language processor
  may impose;
\item
  the semantics of configuration interfaces.
\end{itemize}

This standard does not specify

\begin{itemize}
\item
  the mechanism by which Rexx programs are transformed for use by a
  data-processing system;
\item
  the mechanism by which Rexx programs are invoked for use by a
  data-processing system;
\item
  the mechanism by which input data are transformed for use by a Rexx
  program;
\item
  the mechanism by which output data are transformed after being
  produced by a Rexx program;
\item
  the encoding of Rexx programs;
\item
  the encoding of data to be processed by Rexx programs;
\item
  the encoding of output produced by Rexx programs;
\item
  the size or complexity of a program and its data that will exceed the
  capacity of any specific data-processing system or the capacity of a
  particular language processor;
\item
  all minimal requirements of a data-processing system that is capable
  of supporting a conforming language processor;
\item
  the syntax of the configuration interfaces.
\end{itemize}

\hypertarget{purpose}{%
\section{Purpose}\label{purpose}}

The purpose of this standard is to facilitate portability of Rexx
programs for use on a wide variety of configurations.

\hypertarget{application}{%
\section{Application}\label{application}}

This standard is applicable to Rexx language processors.

\hypertarget{recommendation}{%
\section{Recommendation}\label{recommendation}}

It is recommended that before detailed reading of this standard, a
reader should first be familiar with the Rexx language, for example
through reading one of the books about Rexx. It is also recommended that
the annexes should be read in conjunction with this standard.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/syntax_constructs.tex
complete, but very deficient, extended standard

\hypertarget{syntax-constructs}{%
\chapter{Syntax constructs}\label{syntax-constructs}}

\hypertarget{notation}{%
\section{Notation}\label{notation}}

\hypertarget{backus-naur-form-bnf}{%
\subsection{Backus-Naur Form (BNF)}\label{backus-naur-form-bnf}}

The syntax constructs in this standard are defined in Backus-Naur Form
(BNF). The syntax used in these BNF productions has

\begin{itemize}
\item
  a left-hand side (called identifier);
\item
  the characters ':=``;
\item
  a right-hand side (called bnf\_expression). The left-hand side
  identifies syntactic constructs. The right-hand side describes valid
  ways of writing a specific syntactic construct. The right-hand side
  consists of operands and operators, and may be grouped.
\end{itemize}

\hypertarget{operands}{%
\subsection{Operands}\label{operands}}

Operands may be terminals or non-terminals. If an operand appears as
identifier in some other production it is called a non-terminal,
otherwise it is called a terminal. Terminals are either literal or
symbolic. Literal terminals are enclosed in quotes and represent
literally (apart from case) what must be present in the source being
described. Symbolic terminals formed with lower case characters
represent something which the configuration may, or may not, allow in
the source program, see nnn, nnn, nnn, nnn. Symbolic terminals formed
with uppercase characters represent events and tokens, see nnn and nnn.
\#\#\# Operators The following lists the valid operators, their meaning,
and their precedence; the operator listed first has the highest
precedence; apart from precedence recognition is from left to right:

\begin{itemize}
\item
  the postfix plus operator specifies one or more repetitions of the
  preceding construct;
\item
  abuttal specifies that the preceding and the following construct must
  appear in the given order;
\item
  the operator `\textbar{}' specifies alternatives between the preceding
  and the following constructs.
\end{itemize}

\hypertarget{grouping}{%
\subsection{Grouping}\label{grouping}}

Parentheses and square brackets are used to group constructs.
Parentheses are used for the purpose of grouping only. Square brackets
specify that the enclosed construct is optional.

\hypertarget{bnf-syntax-definition}{%
\subsection{BNF syntax definition}\label{bnf-syntax-definition}}

The BNF syntax, described in BNF, is:

production := identifier `:=' bnf expression

bnf expression t= abuttal \textbar{} bnf expression `\textbar{}' abuttal

abuttal t= {[}abuttal{]} bnf primary

bnf primary := `{[}T' bnf expression '{]}' \textbar{} `(' bnf expression
`)' \textbar{} literal \textbar{}

identifier \textbar{} message identifier \textbar{} bnf primary `+'

\hypertarget{syntactic-errors}{%
\subsection{Syntactic errors}\label{syntactic-errors}}

The syntax descriptions (see nnn and nnn) make use of
message\_identifiers which are shown as Msgnn.nn or Msgnn, where nn is a
number. These actions produce the correspondingly numbered error
messages (see nnn and nnn).

\hypertarget{lexical}{%
\section{Lexical}\label{lexical}}

The lexical level processes the source and provides tokens for further
recognition by the top syntax level.

\hypertarget{lexical-elements}{%
\subsection{Lexical elements}\label{lexical-elements}}

\hypertarget{events}{%
\subsubsection{Events}\label{events}}

The fully-capitalized identifiers in the BNF syntax (see nnn) represent
events. An event is either supplied by the configuration or occurs as
result of a look-ahead in left-to-right parsing. The following events
are defined:

\begin{itemize}
\item
  EOL occurs at the end of a line of the source. It is provided by
  Config\_SourceChar, see nnn;
\item
  EOS occurs at the end of the source program. It is provided by
  Config\_SourceChar;
\item
  RADIX occurs when the character about to be scanned is `X' or `x' or
  `B' or `b' not followed by a general\_letter, or a digit, or'.';
\item
  CONTINUE occurs when the character about to be scanned is `,', and the
  characters after the `,' up to EOL represent a repetition of comment
  or blank, and the EOL is not immediately followed by an EOS;
\item
  EXPONENT\_SIGN occurs when the character about to be scanned is `+' or
  `-', and the characters to the left of the sign, currently parsed as
  part of Const\_symbol, represent a plain\_number followed by `E' or
  `e', and the characters to the right of the sign represent a
  repetition of digit not followed by a general\_letter or'.'.
\item
  would put ASSIGN here for the leftmost `=' in a clause that is not
  within parentheses or brackets. But Simon not

  happy with message term being an assignment? \#\#\#\# Actions and
  tokens Mixed case identifiers with an initial capital letter cause an
  action when they appear as operands ina production. These actions
  perform further tests and create tokens for use by the top syntax
  level. The following actions are defined:
\item
  Special supplies the source recognized as special to the top syntax
  level;
\item
  Eol supplies a semicolon to the top syntax level;
\item
  Eos supplies an end of source indication to the top syntax level;
\item
  Var\_symbol supplies the source recognized as Var\_symbol to the top
  syntax level, as keywords or VAR\_SYMBOL tokens, see nnn. The
  characters in a Var\_symbol are converted by Config\_Upper to
  uppercase. Msg30.1 shall be produced if Var\_symbo/ contains more than
  \#Limit\_Name characters, see nnn;
\item
  Const\_symbol supplies the source recognized as Const\_symbo! to the
  top syntax level. If itis a number it is passed as a NUMBER token,
  otherwise it is passed as a CONST\_SYMBOL token. The characters in a
  Const\_symbol are converted by Config\_Upper to become the characters
  that comprise that NUMBER or CONST\_SYMBOL. Msg30.1 shall be produced
  if Const\_symbo! contains more than \#Limit\_Name characters;
\item
  Embedded\_quotation\_mark records an occurrence of two consecutive
  quotation marks within a string delimited by quotation marks for
  further processing by the String action;
\item
  Embedded\_apostrophe records an occurrence of two consecutive
  apostrophes within a string delimited by apostrophes for further
  processing by the String action;
\item
  String supplies the source recognized as String to the top syntax
  level as a STRING token. Any occurrence of Embedded\_quotation\_mark
  or Embedded\_apostrophe is replaced by a single quotation mark or
  apostrophe, respectively. Msg30.2 shall be produced if the resulting
  string contains more than \#Limit\_Literal characters;
\item
  Binary\_string supplies the converted binary string to the top syntax
  level as a STRING token, after checking conformance to the
  binary\_string syntax. If the binary\_string does not contain any
  occurrence of a binary\_digit, a string of length 0 is passed to the
  top syntax level. The occurrences of binary\_digit are concatenated to
  form a number in radix 2. Zero or 4 digits are added at the left if
  necessary to make the number of digits a multiple of 8. If the
  resulting number of digits exceeds 8 times \#Limit\_Literal then
  Msg30.2 shall be produced. The binary digits are converted to an
  encoding, see nnn. The encoding is supplied to the top syntax level as
  a STRING token;
\item
  Hex\_string supplies the converted hexadecimal string to the top
  syntax level as a STRING token, after checking conformance to the
  hex\_string syntax. If the hex\_string does not contain any occurrence
  of a hex\_digit, a string of length 0 is passed to the top syntax
  level. The occurrences of hex\_digit are each converted to a number
  with four binary digits and concatenated. 0 to 7 digits are added at
  the left if necessary to make the number of digits a multiple of 8. If
  the resulting number of digits exceeds 8 times \#Limit\_Literal then
  Msg30.2 shall be produced. The binary digits are converted to an
  encoding. The encoding is supplied to the top syntax level as a STRING
  token;
\item
  Operator supplies the source recognized as Operator (excluding
  characters that are not operator\_char ) to the top syntax level. Any
  occurrence of an ofher\_negator within Operator is supplied as '\,';
\item
  Blank records the presence of a blank. This may subsequently be tested
  (see nnn). Constructions of type Number, Const\_symbol, Var\_symbol or
  String are called operands. 6.2.1.3 Source characters The source is
  obtained from the configuration by the use of Config\_SourceChar (see
  nnn). If no character is available because the source is not a correct
  encoding of characters, message Msg22.1 shall be produced. The terms
  extra\_letter, other\_blank\_character, other\_negator, and
  other\_character used in the productions of the lexical level refer to
  characters of the groups extra\_letters (see nnn),
\end{itemize}

other\_blank\_characters (see nnn), other\_negators (see nnn) and
other\_characters (see nnn), respectively.

\hypertarget{rules}{%
\subsubsection{Rules}\label{rules}}

In scanning, recognition that causes an action (see nnn) only occurs if
no other recognition is possible, except that Embedded\_apostrophe and
Embedded\_quotation\_mark actions occur wherever possible.

\hypertarget{lexical-level}{%
\subsection{Lexical level}\label{lexical-level}}

\hypertarget{interaction-between-levels-of-syntax}{%
\subsection{Interaction between levels of
syntax}\label{interaction-between-levels-of-syntax}}

When the lexical process recognizes tokens to be supplied to the top
level, there can be changes made or tokens added. Recognition is
performed by the lexical process and the top level process ina
synchronized way. The tokens produced by the lexical level can be
affected by what the top level syntax has recognized. Those tokens will
affect subsequent recognition by the top level. Both processes operate
on the characters and the tokens in the order they are produced. The
term ``context'' refers to the progress of the recognition at some
point, without consideration of unprocessed characters and tokens. If a
token which is `+', `-', ' or `(' appears in a lexical level context
(other than after the keyword `PARSE') where the keyword `VALUE' could
appear in the corresponding top level context, then `VALUE' is passed to
the top level before the token is passed. If an `=' operator\_char
appears in a lexical level context where it could be the `=' of an
assignment or message\_instruction in the corresponding top level
context then it is recognized as the `=``' of that instruction. (It will
be outside of brackets and parentheses, and any Var\_symbo/ immediately
preceding it is passed as a VAR\_SYMBOL). If an operand is followed by a
colon token in the lexical level context then the operand only is passed
to the top level syntax as a LABEL, provided the context permits a
LABEL. Except where the rules above determine the token passed, a
Var\_symbol is passed as a terminal (a keyword) rather than as a
VAR\_SYMBOL under the following circumstances:

\begin{itemize}
\item
  if the symbol is spelled `WHILE' or `UNTIL' it is a keyword wherever a
  VAR\_SYMBOL would be part of an expression within a do\_specification,
\item
  if the symbol is spelled `TO' , `BY', or `FOR' it is a keyword
  wherever a VAR\_SYMBOL would be part of an expression within a
  do\_rep;
\item
  if the symbol is spelled `WITH' it is a keyword wherever a VAR\_SYMBOL
  would be part of a parsevalue, or part of an expression or
  taken\_constant within address;
\item
  if the symbol is spelled `THEN' it is keyword wherever a VAR\_SYMBOL
  would be part of an expression immediately following the keyword `IF'
  or `WHEN'. Except where the rules above determine the token passed, a
  Var\_symbol is passed as a keyword if the spelling of it matches a
  keyword which the top level syntax recognizes in its current context,
  otherwise the Var\_symbol is passed as a VAR\_SYMBOL token. In a
  context where the top level syntax could accept a `\textbar\textbar{}'
  token as the next token, a'\textbar\textbar' operator ora'\,' operator
  may be inferred and passed to the top level provided that the next
  token from the lexical level is a left parenthesis or an operand that
  is not a keyword. If the blank action has recorded the presence of one
  or more blanks to the left of the next token then the '\,' operator is
  inferred. Otherwise, a'\textbar\textbar' operator is inferred, except
  if the next token is a left parenthesis following an operand (see
  nnn); in this case no operator is inferred. When any of the keywords
  `OTHERWISE', `THEN', or `ELSE' is recognized, a semicolon token is
  supplied as the following token. A semicolon token is supplied as the
  previous token when the `THEN' keyword is recognized. A semicolon
  token is supplied as the token following a LABEL.
\end{itemize}

\hypertarget{reserved-symbols}{%
\subsubsection{Reserved symbols}\label{reserved-symbols}}

A Const\_symbol which starts with a period and is not a Number shall be
spelled .MN, .RESULT, .RC, .RS, or .SIGL otherwise Msg50.1 is issued.

\hypertarget{function-name-syntax}{%
\subsubsection{Function name syntax}\label{function-name-syntax}}

A symbol which is the leftmost component of a function shall not end
with a period, otherwise Msg51.1 is issued.

\hypertarget{syntax}{%
\section{Syntax}\label{syntax}}

\hypertarget{syntax-elements}{%
\subsection{Syntax elements}\label{syntax-elements}}

The tokens generated by the actions described in nnn form the basis for
recognizing larger constructs.

\hypertarget{syntax-level}{%
\subsection{Syntax level}\label{syntax-level}}

starter:=x3j18

x3j18:=program Eos \textbar{} Msg35.1

program := {[}label list{]} {[}ncl{]} {[}requires+{]}
{[}prolog\_instruction+{]} (class definition {[}requires+{]})+ requires
:= `REQUIRES' ( taken constant \textbar{} Mgg19.8 ) ```;'+ prolog
instruction:= (package \textbar{} import \textbar{} options) nel package
= `PACKAGE'( NAME \textbar{} Msgnn ) import = `IMPORT' ( NAME \textbar{}
Msgnn ) {[}`.'{]} options := `OPTIONS' ( symbol+ \textbar{} Msgnn ) nel
:= null \_clause+ \textbar{} Msg21.1 null clause = `;' {[}label list{]}
label list = (LABEL `;')+ class definition = class {[}property info{]}
{[}method definition+{]} class = `CLASS' ( taken constant \textbar{}
Msg19.12 ) {[}class\_option+{]}

{[}`INHERIT' ( taken constant \textbar{} Msg19.13 )+{]} nel visibility
\textbar{} modifier \textbar{} `BINARY' \textbar{} `DEPRECATED'
`EXTENDS' ( NAME \textbar{} Msgnn ) `USES' ( NAMElist \textbar{} Msgnn )
\textbar{} `IMPLEMENTS' ( NAMElist \textbar{} Msgnn )

class option

numeric digits:= `DIGITS' {[}expression{]}

external \textbar{} metaclass \textbar{} submix /* \textbar{} 'PUBLIC!
*/

external = `EXTERNAL' (STRING \textbar{} Msg19.14) metaclass =
`METACLASS' ( taken constant \textbar{} Msg19.15 ) submix = `MIXINCLASS'
( taken constant \textbar{} Msg19.16 ) \textbar{} `SUBCLASS' ( taken
constant \textbar{} Msg19.17 ) visibility = `PUBLIC' \textbar{}
`PRIVATE' modifier = `ABSTRACT' \textbar{} `FINAL' \textbar{}
`INTERFACE' \textbar{} `ADAPTER'! NAMElist = NAME {[}(`,' ( NAME
\textbar{} Msgnn ) )+{]} property \_info = numeric \textbar{} property
assignment \textbar{} properties \textbar{} trace numeric = `NUMERIC'
(numeric digits \textbar{} numeric form \textbar{} Msg25.15) numeric
form = `FORM' {[}`ENGINEERING' \textbar{} `SCIENTIFIC'{]} property
agsignment := NAME \textbar{} assignment properties := `PROPERTIES' (
properties option+ \textbar{} Msgnn) properties option := properties
vigibility \textbar{} properties modifier properties visibility :=
`INHERITABLE' \textbar{} `PRIVATE' \textbar{} `PUBLIC' \textbar{}
`INDIRECT' properties modifier := `CONSTANT' \textbar{} `STATIC'
\textbar{} `VOLATILE' \textbar{} `TRANSIENT' trace := `TRACE' {[}`ALL'
\textbar{} `METHODS' \textbar{} `OFF' \textbar{} `RESULTS'{]} method
definition = (method {[}expose ncl{]}\textbar{} routine) balanced expose
= `EXPOSE' variable list method = `METHOD' (taken constant \textbar{}
Msg19.9) {[} `(' assigncommalist \textbar{} Msgnn ( `)' \textbar{} Msgnn
){]} {[}method option+{]} nel assigncommalist assignment {[}(`,' (
assignment \textbar{} Msgnn ) )+{]}

method visibility \textbar{} method modifier \textbar{} `PROTECT'
\textbar{} `RETURNS' ( term \textbar{} Msgnn ) \textbar{} `SIGNAL' (
termcommalist \textbar{} Msgnn )

method option

`DEPRECATED'! `CLASS' \textbar{} `ATTRIBUTE' \textbar{} /\emph{`PRIVATE'
\textbar{} }/ guarded guarded := `GUARDED' \textbar{} `UNGUARDED! method
visibility := 'INHERITABLE' \textbar{} `PRIVATE' \textbar{} `PUBLIC'
\textbar{} `SHARED!' method modifier := `ABSTRACT' \textbar{} `CONSTANT'
\textbar{} `FINAL' \textbar{} `NATIVE' \textbar{} `STATIC' termcommalist
:= term {[}(`,' ( term \textbar{} Msgnn ) )+{]} routine := `ROUTINE' (
taken constant \textbar{} Msg19.11 ) {[}`PUBLIC'{]} nel balanced:=
instruction list {[}`END' Msgl0.1{]} instruction list:= instruction+

/* The second part ig about groups */

instruction = group \textbar{} gingle instruction nel group = do ncl
\textbar{} if \textbar{} loop nel \textbar{} select nel do = do
specification nel {[}instruction+{]} {[}group\_handler{]} (`END'
{[}NAME{]} \textbar{} Eos Msg14.1 \textbar{} Msg35.1) group option :=
`LABEL' ( NAME \textbar{} Msgnn ) \textbar{} `PROTECT' ( term \textbar{}
Msgnn )

group handler catch \textbar{} finally catch finally

`CATCH' {[} NAME `=' {]} ( NAME \textbar{} Msgnn) nel {[}instruction+{]}

catch = /* FINALLY implies a semicolon. */ finally = `FINALLY' nel (
instruction+ \textbar{} Msgnn ) if := `IF' expression {[}ncl{]} (then
\textbar{} Msg18.1) {[}else{]} then := `THEN' nel (instruction
\textbar{} EOS Msg14.3 \textbar{} `END' Msg10.5) else := `ELSE' nel
(instruction \textbar{} EOS Msg14.4 \textbar{} `END' Msg10.6) loop :=
`LOOP' {[}group\_option+{]} {[}repetitor{]} {[}conditional{]} nel

instruction+ {[}group\_handler{]} loop ending

loop ending `END' {[}VAR SYMBOL{]} \textbar{} EOS Msg14.n \textbar{}
Msg35.1

conditional = `WHILE' whileexpr \textbar{} `UNTIL' untilexpr untilexpr =
expression whileexpr = expression repetitor = assignment {[}count
option+{]} \textbar{} expression \textbar{} over `FOREVER' count option
= loopt \textbar{} loopb \textbar{} loopf loopt = `TO' expression loopb
= `BY' expression loopf = `FOR' expression over = VAR SYMBOL `OVER'
expression

NUMBER `OVER' Msg31.1\\
CONST SYMBOL `OVER' (Msg31.2 \textbar{} Msg31.3)

select := `SELECT' {[}group option+{]} ncl select \_ body
{[}group\_handler{]} (`END' {[}NAME Msg10.4{]} \textbar{} EOS Msgl14.2
\textbar{} Msg7.2)

(when \textbar{} Msg7.1) {[}when+{]} {[}otherwise{]}

`WHEN' expresgion {[}ncl{]} (then \textbar{} Msg18.2)

`OTHERWISE' ncl {[}instruction+{]}

select body when otherwise

/* Third part is for single instructions. */ single instruction:=
assignment \textbar{} message instruction \textbar{} keyword instruction
\textbar{} command assignment := VAR SYMBOL `\#' expression NUMBER `\#'
Msg31.1 CONST SYMBOL `\#!' (Msg31.2 \textbar{} Mgg31.3)

message instruction := message term \textbar{} message term `\#'
expression keyword instruction:= address \textbar{} arg \textbar{} call
\textbar{} drop \textbar{} exit interpret \textbar{} iterate \textbar{}
leave

nop \textbar{} numeric \textbar{} options parse \textbar{} procedure
\textbar{} pull \textbar{} push \textbar{} queue

raise \textbar{} reply \textbar{} return \textbar{} say \textbar{}
signal \textbar{} trace \textbar{} use `THEN' Msg8.1 `ELSE' Msg8.2
`WHEN' Msg9.1 \textbar{} `OTHERWISE' Msg9.2 command = expression address
= `ADDRESS' {[}(taken constant {[}expression{]}{]}

taken constant symbol \textbar{} STRING

valueexp = `VALUE' expression connection = ad option+ ad option = error
\textbar{} input \textbar{} output \textbar{} Msg25.5 error = `ERROR!'
(resourceo \textbar{} Msg25.14) input = `INPUT' (resourcei \textbar{}
Msg25.6) resourcei := resources \textbar{} `NORMAL' output = `OUTPUT'
(resourceo \textbar{} Mgg25.7) resourceo := `APPEND' (resources
\textbar{} Msg25.8) \textbar{} `REPLACE' (resources \textbar{} Msg25.9)
\textbar{} resources \textbar{} `NORMAL' resources := `STREAM'
(VAR\_SYMBOL \textbar{} Msg53.1) \textbar{} `STEM' (VAR\_SYMBOL
\textbar{} Msg53.2) vref := `(' var symbol (`)' \textbar{} Msg46.1) var
symbol = VAR\_SYMBOL \textbar{} Msg20.1

arg := `ARG' {[}template list{]}

eall := `CALL' (callon\_spec \textbar{} (taken constant \textbar{} vref
\textbar{} Msg19.2) {[}expression list{]} ) callon spec := `ON'
(callable condition \textbar{} Msg25.1)

{[}`NAME' (symbol constant term \textbar{} Msg19.3){]} \textbar{} `OFF'
(callable condition \textbar{} Msg25.2)

symbol constant term := term

callable condition:= `ANY' \textbar{} `ERROR' \textbar{} `FAILURE'
\textbar{} `HALT' \textbar{} `NOTREADY' \textbar{} `USER' ( symbol
constant term \textbar{} Msg19.18 )

condition := callable condition \textbar{} `LOSTDIGITS' \textbar{}
`NOMETHOD' \textbar{} `NOSTRING' \textbar{} `NOVALUE' \textbar{}
'SYNTAX!

expression list do specification do simple

expr \textbar{} {[}expr{]} `,' {[}expression list{]} do simple
\textbar{} do repetitive `DO' {[}group\_option+{]}

do repetitive = do simple (dorep \textbar{} conditional \textbar{} dorep
conditional) dorep = `FOREVER' \textbar{} repetitor drop = `DROP'
variable list variable list = (vref \textbar{} var\_symbol)+ exit =
`EXIT' {[}expression{]} forward = `FORWARD' {[}forward \_option+
\textbar{} Msg25.18{]} forward option = `CONTINUE' \textbar{}
ArrayArgOption \textbar{} MessageOption \textbar{} ClassOption
\textbar{} ToOption ArrayArgOption:=`ARRAY' arguments \textbar{}
`ARGUMENTS' term MessageOption :=`MESSAGE' term ClassOption =`CLASS'
term ToOption =`TO' term guard = `GUARD' (`ON' \textbar{} Msg25.22)
{[}(`WHEN' \textbar{} Msg25.21) expression{]} \textbar{} ( `OFF'
\textbar{} Msg25.19) {[}(`WHEN' \textbar{} Msg25.21) expression{]}
interpret = `INTERPRET' expression iterate = `ITERATE' {[}VAR SYMBOL
\textbar{} Msg20.2{]} leave = `LEAVE' {[}VAR SYMBOL \textbar{}
Msg20.2{]} nop = `NOP!' numeric = `NUMERIC' (numeric digits \textbar{}
numeric form

numeric fuzz \textbar{} Msg25.15)

numeric digits `DIGITS' {[}expression{]} numeric form `FORM' {[}numeric
form suffix{]} numeric form suffix:=(`ENGINEERING'
\textbar{}`SCIENTIFIC'\textbar valueexp \textbar{} Msg25.11) numeric
fuzz `FUZZ' {[}expression{]}

options = `OPTIONS' expression parse = `PARSE' {[}translations{]} (parse
type \textbar Msg25.12) {[}template list{]} translations := `CASELESS'
{[}`UPPER' \textbar{} `LOWER'{]} \textbar{} (`UPPER' \textbar{} `LOWER')
{[}`CASELESS'{]} parse type = parse key \textbar{} parse value
\textbar{} parse var \textbar{} term parse key = `ARG' \textbar{} `PULL'
\textbar{} `SOURCE' \textbar{} `LINEIN' \textbar{} `VERSION! parse value
= 'VALUE' {[}expression{]} (`WITH' \textbar{} Mgg38.3) parse var = `VAR'
var symbol template := NAME {[}( {[}pattern{]} NAME) +{]} pattern:=
STRING \textbar{} {[}indicator{]} NUMBER \textbar{} {[}indicator{]} `('
symbol `)' indicator := `+' \textbar{} `-' \textbar{} `s! procedure =
'PROCEDURE' {[}expose \textbar{} Msg25.17{]} pull = `PULL' {[}template
list{]} push = `PUSH' {[}expression{]} queue = `QUEUE' {[}expression{]}
raise = `RAISE' conditions (raise option \textbar{} Msg25.24) conditions
= `ANY' \textbar{} `ERROR' term \textbar{} `FAILURE' term \textbar{}
`HALT'\textbar{} `LOSTDIGITS' \textbar{} `NOMETHOD' \textbar{}
`NOSTRING' \textbar{} ``NOTREADY' \textbar{} `NOVALUE' \textbar{}
`PROPAGATE' \textbar{} `SYNTAX' term \textbar{} `USER' ( symbol constant
term \textbar{} Msg19.18) \textbar{} Msg25.23 raise option :=
ExitRetOption \textbar{} Description \textbar{} ArrayOption

ExitRetOption := `EXIT! {[}term{]} \textbar{} 'RETURN' {[}term{]}
Description =`DESCRIPTION' term ArrayOption = `ADDITIONAL' term
\textbar{} ```ARRAY' arguments reply = `REPLY' {[} expression{]} return
= `RETURN' {[}expression{]} say = `SAY' {[}expression{]} signal =
`SIGNAL' (signal spec \textbar{} valueexp

symbol constant term \textbar{} Msg19.4)

signal spec := `ON' (condition \textbar{} Msg25.3) {[}`NAME' (symbol
constant term \textbar{} Msg19.3){]} \textbar{} `OFF' (condition
\textbar{} Msg25.4)

trace = `TRACE' {[}(taken\_constant \textbar{} Msgl19.6) \textbar{}
valueexp{]} use = `USE' (`ARG' \textbar{} Msg25.26) {[}use list{]} use
list = VAR\_SYMBOL \textbar{} {[}VAR SYMBOL{]} `,' {[}use list{]}

/* Note: The next part describes templates. */ template list template
\textbar{} {[}template{]} `,' {[}template list{]}

template = (trigger \textbar{} target \textbar{} Msg38.1)+ target =
VAR\_SYMBOL \textbar{} `.!' trigger = pattern \textbar{} positional
pattern = STRING \textbar{} vrefp vrefp = `(' (VAR\_SYMBOL \textbar{}
Msgl19.7) (`)' \textbar{} Msg46.1) positional = absolute positional
\textbar{} relative positional absolute positional:= NUMBER `=' position
pogition := NUMBER \textbar{} vrefp \textbar{} Msg38.2 relative
posgitional:= (`+' \textbar{} `-``') position

/* Note: The final part specifies the various forms of symbol, and
expression. */

symbol = VAR\_SYMBOL \textbar{} CONST SYMBOL \textbar{} NUMBER
expression = expr {[}(`,' Msg37.1) \textbar{} (`)' Msg37.2 ){]} expr =
expr alias

expr alias and expression

expr alias or operator and expression

or operator := `\textbar{}' \textbar{} `\&\&! and expression :=
comparison \textbar{} and expression'\&' comparison comparison :=
concatenation \textbar{} comparison comparison operator concatenation
comparison \_operator:= normal compare \textbar{} strict compare normal
compare:= `=! \textbar{} ``\s! \textbar{} hep! \textbar{} tact
\textbar{} tot \textbar{} tet \textbar{} toet len! W\textgreater!
\textless! strict\_compare:='==' \textbar{} `\s=' \textbar{}
`\textgreater\textgreater!' \textbar{} `\textless\textless!' \textbar{}
`\textgreater\textgreater e! \textbar{}'\textless\textless s! \textbar{}
```\e\textgreater t \textbar{}'\textless c! concatenation := addition
\textbar{} concatenation (' ' \textbar{} `\textbar\textbar{}') addition
addition := multiplication \textbar{} addition additive operator
multiplication additive operator:= `+' \textbar{} '-! multiplication t=
power expression

multiplication multiplicative operator

power expression multiplicative operator:= '*' \textbar{} `/' \textbar{}
`//' \textbar{} `\%! power expression := prefix expression \textbar{}
power expression'**' prefix expression prefix expression := (`+'
\textbar{} `-' \textbar{} '\,') prefix expression term \textbar{}
Msg35.1 /* ``Stub'' has to be identified semantically? */

term = simple term {[} `.' ( term \textbar{} Msgnn ){]} simple term :=
symbol \textbar{} STRING \textbar{} invoke \textbar{} indexed `('
expression ( `)' \textbar{} Msg36 ) initializer \textbar{} message term
`\#\#! message term:= term ('\textasciitilde' \textbar{}
`\textasciitilde\textasciitilde{}') method name {[}arguments{]} term
`{[}'{[} expression list {]} ('{]}' \textbar{} Msg36.2)

method name:=(taken constant \textbar{} Msg19.19) {[}`:' ( VAR\_SYMBOL
\textbar{} Msg19.21 ){]} /* Method-call without arguments ig
syntactically like symbol. \emph{/ /} Editor - not sure of my notes
about here. */

invoke := (symbol \textbar{} STRING) arguments arguments := `\#('
{[}expression list{]} (`)' \textbar{} Meg36) expression list :=
expregsion \textbar{} {[}expression{]} `,' {[}expression list{]} indexed
= (symbol \textbar{} STRING) indices indices = `\#{[}' {[}expression
list{]} ('{]}' \textbar{} Msg36.n) initializer = `{[}'expression list
('{]}' \textbar{} Msg36.n)

\hypertarget{syntactic-information}{%
\section{Syntactic information}\label{syntactic-information}}

\hypertarget{var_symbol-matching}{%
\subsection{VAR\_SYMBOL matching}\label{var_symbol-matching}}

Any VAR\_SYMBOL in a do\_ending must be matched by the same VAR\_SYMBOL
occurring at the start of an assignment contained in the
do\_specification of the do that contains both the do\_specification and
the do\_ending, as described in nnn. If there is a VAR\_SYMBOL in a
do\_ending for which there is no assignment in the corresponding
do\_specification then message Msg10.3 is produced and no further
activity is defined. If there is a VAR\_SYMBOL in a do\_ending which
does not match the one occurring in the assignment then message Msg10.2
is produced and no further activity is defined. An iterate or leave must
be contained in the instruction\_list of some do with a
do\_specification which is do\_repetitive, otherwise a message (Msg28.2
or Msg28.1 respectively) is produced and no further activity is defined.
If an iterate or leave contains a VAR\_SYMBOL there must be a matching
VAR\_SYMBOL ina do\_specification, otherwise a message (Msg28.1,
Msg28.2, Msg28.3 or Msg28.4 appropriately) is produced and no further
activity is defined. The matching VAR\_SYMBOL will occur at the start of
an assignment in the do\_specification. Tne do\_specification will be
associated with a do by nnn. The /ferafe or leave will be a single
instruction in an instruction\_list associated with a do by nnn. These
two dos shall be the same, or the latter nested one or more levels
within the former. The number of levels is called the
nesting\_correction and influences the semantics of the iterafe or
leave. It is zero if the two dos are the same. The nesting\_correction
for /ferates or leaves that do not contain VAR\_SYMBOL is zero.

\hypertarget{trace-only-labels}{%
\subsection{Trace-only labels}\label{trace-only-labels}}

Instances of LABEL which occur within a grouping\_instruction and are
not in a nc/ at the end of that grouping\_instruction are instances of
trace-only labels.

\hypertarget{clauses-and-line-numbers}{%
\subsection{Clauses and line numbers}\label{clauses-and-line-numbers}}

The activity of tracing execution is defined in terms of clauses. A
program consists of clauses, each clause ended by a semicolon special
token. The semicolon may be explicit in the program or inferred. The
line number of a clause is one more than the number of EOL events
recognized before the first token of the clause was recognized.

\hypertarget{nested-if-instructions}{%
\subsection{Nested IF instructions}\label{nested-if-instructions}}

The syntax specification nnn allows `IF' instructions to be nested and
does not fully specify the association of an `ELSE' keyword with an `IF'
keyword. An `ELSE' associates with the closest prior `IF' that it can
associate with in conformance with the syntax.

\hypertarget{choice-of-messages}{%
\subsection{Choice of messages}\label{choice-of-messages}}

The specifications nnn and nnn permit two alternative messages in some
circumstances. The following rules apply:

\begin{itemize}
\item
  Msg15.1 shall be preferred to Msg15.3 if the choice of Msg15.3 would
  result in the replacement for the insertion being a blank character;
\item
  Msg15.2 shall be preferred to Msg15.4 if the choice of Msg15.4 would
  result in the replacement for the insertion being a blank character;
\item
  Msg31.3 shall be preferred to Msg31.2 if the replacement for the
  insertion in the message starts with a period;
\item
  Preference is given to the message that appears later in the list:
  Msg21.1, Msg27.1, Msg25.16, Msg36, Msg38.3, Msg35.1, other messages.
\end{itemize}

\hypertarget{creation-of-messages}{%
\subsection{Creation of messages}\label{creation-of-messages}}

The message\_identifiers in clause 6 correlate with the tails of stem
\#ErrorText., which is initialized in nnn to identify particular
messages. The action of producing an error message will replace any
insertions in the message text and present the resulting text, together
with information on the origin of the error, to the configuration by
writing on the default error stream. Further activity by the language
processor is permitted, but not defined by this standard. The effect of
an error during the writing of an error message is not defined.

\hypertarget{error-message-prefix}{%
\subsubsection{Error message prefix}\label{error-message-prefix}}

The error message selected by the message number is preceded by a
prefix. The text of the prefix is \#ErrorText.0.1 except when the error
is in source that execution of an interactive trace interpret
instruction (see nnn) is processing, in which case the text is
\#ErrorText.0.2. The insert called in these texts is the message number.
The insert called is the line number of the error. The line number of
the error is one more than the number of EOL events encountered before
the error was detectable, except for messages Msg6.1, Msg14, Msg14.1,
Msg14.2, Msg14.3, and Msg14.4. For Msg6.1 it is one more than the number
of EOL events encountered before the line containing the unmatched '/*'.
For the others, it is the line number of the clause containing the
keyword referenced in the message text. The insert called is the value
provided on the API\_ Start function which started processing of the
program, see nnn.

\hypertarget{replacement-of-insertions}{%
\section{Replacement of insertions}\label{replacement-of-insertions}}

Within the text of error messages, an insertion consists of the
characters `\textless{}', `\textgreater{}', and what is between those
characters. There will be a word in the insertion that specifies the
replacement text, with the following meaning:

\begin{itemize}
\item
  if the word is `hex-encoding' and the message is not Msg23.1 then the
  replacement text is the value of the leftmost character which caused
  the source to be syntactically incorrect. The value is in hexadecimal
  notation;
\item
  if the word is `token' then the replacement text is the part of the
  source program which was recognized as the detection token, or in the
  case of Msg31.1 and Msg31.2, the token before the detection token. The
  detection token is the leftmost token for which the program up to and
  including the token could not be parsed as the left part of a program
  without causing a message. If the detection token is a semicolon that
  was not present in the source but was supplied during recognition then
  the replacement is the previous token;
\item
  if the word is `position' then the replacement text is a number
  identifying the detection character. The detection character is the
  leftmost character in the hex\_string or binary\_string which did not
  match the required syntax. The number is a count of the characters in
  the string which preceded the detection character, including the
  initial quote or apostrophe. In deciding the leftmost blank in a
  quoted string of radix `X' or `B' that is erroneous not that: A blank
  as the first character of the quoted string is an error. The leftmost
  embedded sequence of blanks can validly follow any number of non-blank
  characters. Otherwise a blank run that follows an odd numbered
  sequence of non-blanks (or a number not a multiple of four in the case
  of radix `B') is not valid. If the string is invalid for a reason not
  described above, the leftmost blank of the rightmost sequence of
  blanks is the invalid blank to be referenced in the message;
\item
  if the word is `char' then the replacement text is the detection
  character;
\item
  if the word is `linenumber' then the replacement text is the line
  number of a clause associated with the error. The wording of the
  message text specifies which clause that is;
\item
  if the word is `keywords' then the replacement text is a list of the
  keywords that the syntax would allow at the context where the error
  occurred. If there are two keywords they shall be separated by the
  four characters ' or `. If more, the last shall be preceded by the
  three characters 'or' and the others shall be followed by the two
  characters `,'. The keywords will be uppercased and in alphabetical
  order.
\end{itemize}

Replacement text is truncated to \#Limit\_Messagelnsert characters if it
would otherwise be longer than that, except for a keywords replacement.
When an insert is both truncated and appears within quotes in the
message, the three characters `\ldots{}' are inserted in the message
after the trailing quote.

\hypertarget{syntactic-equivalence}{%
\section{Syntactic equivalence}\label{syntactic-equivalence}}

If a message\_term contains a `{[}' it is regarded as an equivalent
message\_term without a `{[}', for execution. The equivalent is
term\textasciitilde{}`{[}{]}'(expression\_list). See nnn. If a
message\_instruction has the construction message\_term `=' expression
it is regarded as equivalent to a message\_term with the same components
as the message\_term left of the `=', except that the taken\_constant
has an `=' character appended and arguments has the expression from the
right of the `=' as an extra first argument. See nnn.

### 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/to_be_processed.tex
complete, but very deficient, extended standard

\hypertarget{to-be-processed}{%
\chapter{To be processed:}\label{to-be-processed}}

The following decisons are abstracted from minutes. We need to ensure
they are covered in the main standard and their rationale appropriately
reworded for this annex.

Aliasing. Assignment is viewed as making the target reference the same
object as the source. Hence the object (and changes to it) may be
accessed through more than one name. For `immutable' objects a changed
version of an object can only be produced by creating a new object. For
compatibility with classic Rexx, strings are immutable objects.
Non-strings may or may not be immutable. Note that there is an
alternative model in which distinction is made between assignments which
copy values and assignments which copy references. This alternative was
not chosen; the committee prefered the model in which all data names are
naming references (which may be implicitly followed to values).

Arguments `by-reference'. The introduction of aliasing makes this
natural although the detail has simple-versus-general contentions. (Is
it necessary for simple strings to be passed by reference.
Encapsulation. An object may `own' some variables and access to those
may be limited (so that re-implementation of the object could use
different variables without upsetting the usage of the object).
Classess. There will be `factory' objects capable of creating multiple
new objects which have common characteristics about how they can be
used.

Inheritance and hierachy. The semantics of a clas may be specified by
adding to the semantics of another class. This relation is used to form
a tree. We prefer a singly rooted tree, rooted in the class `Object'
which is built-in to the language. Other classes will also be built-in.
Experience with OOM and other languages is that unrestricted inheritance
by one class from multiple classes does not work in the way the coder
intended (the implementations of the classes do not combine
successfully). If multiple inheritance is added to Rexx at all, it will
be in the cautious `MIXIN' flavor of OOI.

Messaging: Executing some labelled code which is associated with objects
of a given class is a form of invocation that is sufficiently different
from classic Rexx to justify a new syntax construct. The new syntax is
Receiver \textasciitilde{} MethodName(Arguments) and implies both a
different search for the method to be invoked and a special role for the
receiver as opposed to the other arguments of the invocation.

Packaging: In principle a `program builder' could be used in developing
Rexx programs with many classes and methods, and that builder could hide
from the coder the details of how the configuration held the methods.
However, rather than define a program builder we are choosing to define
a simple method of holding multiple classes \& mthods (with
specification of their hierarchy) within a single text file. The
non-executable dividers in such a file are known as directives. The
files are known as packages and a package may specify (by directive)
that it requires another package in order to function correctly. There
are questions about when initialization of required packages occurs; we
intend to find a solution that does not require the complete graph of
requirements to be initialized before other code is executed.

A note on the syntax of directives. When no special token (eg ::) is
used to introduce directives the directives are be recognizable by the
spelling of the keyword. (CLASS REQUIRES etc.) The purpose of the
special token is emphasis of directives rather than implementation ease
in ``pre-processing'' the directives.

Packages in non-Rexx. It is necessary to exploit packages that are not
written in Rexx. To invoke their methods it is necessary that the
package makes known to the Rexx method search the names of the classes
and their methods. To do more than invoke the methods (eg to subclass
the the external classes) requires complicated mechanisms and may not be
a requirement.

External procedures. To allow Classic internal procedures to be
separated into different files with undue change of semantics, the
PROCEDURE statement will be permiteed as the first statement of a
routine which is in a separate file.

Concurrency will be added, that is multiple execution cursors
progressing through one program. The mechanism for creating multiple
cursors will be the ``early reply'' where one cursor becomes two; one of
two progresses by ``falling through'' the early reply and the other
starts its progress after the site of the current invocation. Multiple
cursors carry the risk of execution interleaving in a way which negates
the coder's intentions in writing so that clauses would execute
sequentially. The language definition will be tightened to ensure
atomicity of string assignment etc. Additionally, a set of rules about
allowing two cursors on the same method at the same time will provide a
reduction of the risk. Since in many cases the data which have to be
maintained consistent will reside in a single object the rules are
object-based. In general a cursor on a method executing against a
particular object will delay any other cursor from executing methods
against that object.

This rule provides sensible synchronization without much effort from the
programmer but other controls may be provided:

\begin{enumerate}
\def\labelenumi{\alph{enumi})}
\item
  Stronger control, eg only one cursor within the methods of a set of
  objects.
\item
  More detailed control, eg division of a method into sections which
  allow/disallow other cursors into the section.
\end{enumerate}

Extended Variable Pools. The API for variable pools will need to be
extended to reflect the model in which the named content in a pool is
always a reference (and the reference is followed when the value of a
string is required.) We note that OO! adopts a convention that names
starting with `!' (shriek) name objects that are not intended for access
by the coder. These objects will not be standardized. Additionally some
objects without shriek names are not candidates for standardising, eg
SYSTEM, .KERNEL.

A model is needed for whether changes made to methods are seen by
objects created before the changes. Changes that are seen are preferable
where a long-lived object is being brought up-to-date. Changes that only
apply to future objects are preferable if avoiding failure of what
``used to work'' is the priority. In view of OOM experience the standard
should allow both, on a method by method choice. (eg perhaps a bug fix
applied retrospectively but not an enhancement.)

Multiple inheritance. Study of the `method search' algorithm, see later,
shows that this is an ``add-on'' that could readily be retained or
omitted. That argues in principle for retention, since the non-user of
multiple inheritance would not suffer from it. On the other hand it adds
complexity and can be misused even in the conservative form that OOI has
it.

Signature-based method search. This is not in OOI but is in languages
such as Java.

Subclassing of imported classes. It is our intention to say that
imported classes can be used in all the same ways as builtin classes.
Because this may be impractical to implement with some external classes,
a conforming language processor will have a list (which may be empty) of
external classes it supports. (And hence nothing of the current SOM
interface will be part of the standard.)

Persistent objects. It is our belief that support for very-long-running
programs is required. It is a moot point whether the ENVIRONMENT
directory is enough.

If persistent objects are to converted to a form which is platform
independent, (``pickling''), there are difficulties in deciding what
pointers should be followed and further objects included, as opposed to
objects being assumed available on all platforms. This topic is defered.

Locking across a set of objects. In OOI this can only be done by locking
the events serially, which has more risk of deadlock than locking them
simultaneously. The decision was made not to add simultaneous locking.

Critical sections. The GUARD mechanism can be used in a `critical
section' style. Nothing will be added to the definition.

Old objects seeing new changed methods. When bugs in long running
programs are fixed, there can be a benefit if old objects see the
corrected methods. It seems practical to offer a variation of DEFINE for
this - see method lookup discussion.

The committee does not find the current OOI approach to merging
`classic' stems with OO stems satisfactory. It invalidates some existing
programs. (A warning about this was put in A8.3.3 of X3.274.) It
produces surprises for OO programmers, eg a==b after
a=.stem\textasciitilde new; b=.stem\textasciitilde new. The proposed
alternative is to make the presence/absence of a dot at the end of the
name determine whether coercion to string is done. The `classic' meaning
of A.=B. would be restored but AA=BB, AA==BB etc. would have their OO
meanings. The meaning of USE ARG with a dotted name would be defined to
allow `by reference' passing of a stem. Square brackets could be used
with both dotted and undotted names. A further proposal is to note that
this leaves few differences between the DIRECTORY class and the
non-dotted STEM class so that it might be a further improvement if the
DIRECTORY class was extended to the extent that the STEM class was
unnecessary.

There is a potential problem which the committee has not fully analysed
in the OO! treatment of SAY and streams. OOI has made features (of the
STREAM bif) that were configuration determined in X3-274 into OO
language methods, and has made SAY a method (undocumented?). Full
analysis may show that more of I/O could (\& should?) be made standard
or may show that some OO! I/O language should not be standardized.

The committee discussed what parts of the OOI implementation were
suitable to be defined in a standard. Potentially, all the builtin
classes and objects (which are reachable from .ENVIRONMENT) might be
standardized. However, names which start with an exclamation mark denote
unsuitable things. The committee also thought the following unsuitable:

\begin{itemize}
\tightlist
\item
  Anything specific to SOM. - RX\_QUEUE - Stream\_Supplier - Parts of
  LOCAL other than direct reference to the default streams. There is a
  naming problem with this. The names in OOI are STDIN, STDOUT and
  STDERR. We would prefer INPUT, OUTPUT, and ERROR to be\_ consistent
  with the keywords. OOI has used those names for something else. We
  will work on the proposal that we use the prefered names and the
  MONITOR class is dropped. (Users who want the monitor function can get
  it with a few lines of directive.)
\end{itemize}

The committee feels that OOI over-specifies the index of an item in a
LIST. In OO it is a count giving the sequence over time of the
insertions in the list. The risk in using numbers is that they may be
(wrongly) used as positions, and arithmetic done on them. It is proposed
that the index of a list item be of class OBJECT rather than of class
STRING.

In OOI, the .ENVIRONMENT is global, not read-only, and contains builtin
objects such as .TRUE and .FALSE. The committee regards this as too
risky - suppose that .TRUE was accidentally or maliciously revalued as
0! It seems sufficient to add read-only as a characteristic of
directories. (This characteristic at the element level might be
expensive to implement.) Reserved symbols (X8-274 clause 6.2.3.1) also
provide a mechanism for preventing the override of builtin names. It
won't be possible for a standard to exactly define in a
system-independent way the scopes/lifetimes of -ENVIRONMENT and .LOCAL
but (as with OOl) the .LOCAL will relate to ``One API\_START'' and
-ENVIRONMENT will have a wider scope. (Power on to power off of some
system'). The proposed ``search order'' is:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Things provided by the system which no user is expected to want to
  override. Perhaps .TRUE .FALSE NIL.
\item
  The .LOCAL read/write directory, initialized with the default streams,
  changable by the user for individual program executions. Perhaps
  METHODS here.
\item
  The read-only part of the environment, that is the builtin classes and
  objects. Also .SYSTEM perhaps.
\item
  The read/write ENVIRONMENT directory. Changable by programmers
  co-operating at the system level. Final placement of all builtins
  needs discussion, but the read-only true\&false requirement will be
  met. Note that the algorithm of method lookup does not change if ``old
  objects see newest methods'' is desired. What changes is whether the
  method tables are updated in place or copied-and-updated when they are
  changed.
\item
  There have been sugestions to allow the REQUIRES directive appear in
  more places. The committee agrees with this and proposes:
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\Alph{enumi})}
\tightlist
\item
  All REQUIRES directives must appear together in the file. B) These
  directives may appear anywhere the OOI implementation currently allows
  them to appear.
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\item
  Message numbers and prose are now allocated to messages detected by
  the syntax, additional to the messages known to the first standard.
  Most messages simply involve new minor codes sequential beyond those
  defined in the first standard.
\item
  Proposed language, eg FORWARD, METHOD, and CLASS clauses, allow for
  many options which can appear in any order. These can be written in
  the BNF (in the manner that TO BY FOR were handled in the first
  standard) but it is neater to extend the BNF metalanguage.
\item
  The OOI syntax used in the FORWARD instruction has examples of the
  `argument' construct, which is either a symbol-or-string taken as a
  constant or is an expression in parentheses. The committee will define
  `term' to be allowed in such places. This is a change to the OOI for
  valid programs only in the case where a MESSAGE option used a symbol
  intending it to be `taken as a constant'. (As opposed to taken as a
  variable with the value defaulting to its name when uninitialized.)
\item
  In a similar vein to 4 above, some other positions where the
  ``variable reference'' notation is used (or proposed) will be changed.
  It would be nice to allow ``term'' in all these places but ambiguity
  consideration means some will be ``sub-expression'', ie parenthesed
  expression, notation.
\item
  The colon used for superclass specification will allow
  symbol-or-string to follow. DATA:
\item
  The model of data used in defining the first standard needs changing
  for OO, to:
\end{enumerate}

\begin{itemize}
\item
  Variable pools are objects, objects are variable pools.
\item
  Variable pool contents are references to objects, not values of
  strings.
\item
  Pools are not numbered, they are referenced.
\item
  The state variables (those with names beginning `\#' used to define
  processing in the standard) are present in all pools, as opposed to
  being in a separate pool.
\end{itemize}

This data model gives a natural interpretation to the variable pool API
applied to local pools. (Local pools may access non-local pool items by
reason of EXPOSE.) In principle this leads to different threads of
execution (resulting from REPLY) being able to execute the API. (In
practice OOI has a restriction to executing the API only on the `main'
thread and the committee needs to know if this is due to a generally
applicable difficulty.)

The committee considered the relevance of IBM's ``Object Rexx
Programming Guide'' G25H-7597-1 to the Configuration section of the
standard. The material there in Appendix A under headings External
Function Interface, System Exit Interface, and Variable Pool Interface
was deemed material for inclusion, and the rest not. This is similar to
the first standard, although there will be an extra trap, for method
calls. The committee considered the relevance of the STREAM section of
IBM's ``Object Rexx Reference'', G25H-7598-0. That stream class brings
into the language more I/O than the original Rexx, eg an explicit CLOSE.
The new standard will partially follow this trend also.

PEEK on queue unnecessary - same as AT{[}1{]}?

Also need to resolve the issues on Monitor class and on run time
inspection.

### 2023-06-02 ARB/standards/work-in-progress/document/tex/standard/provided_classes.tex
add cyrillic.rexx, rename unicode.md to Unicode_Requirements.md

@@
 
-aljl=alj-3{]} alj+l{]}=a{[}j-2{]}+1; /* Index changes \emph{/ end j
-a{[}4{]}=arg(1) a{[}5J=1 return 12.2.4.2 PULL{]}=a{[}j+4{]}-1; /} Index
-changes */ end j return r
+aljl=alj-3{]} alj+l{]}=a{[}j-2{]}+1; /* Index changes */ end j
+a{[}4{]}=arg(1) a{[}5J=1 return 12.2.4.2 PULL
+
+::method pull /* Returns and removes the item at the head of the queue.
+\emph{/ a=self\textasciitilde exposed if a{[}1{]}/}ItemCount\emph{/=0
+then return .nil /} Stays empty \emph{/ r=a{[}4{]} a{[}1{]}
+/}ItemCount\emph{/=a {[}1{]} /}ItemCount\emph{/-1 do j=4 by 3 to
+1+3}a{[}1{]}/\emph{ItemCount}/ aljl=a{[}j+3{]} alj+l{]}=a{[}j+4{]}-1; /*
+Index changes */ end j return r
 
@@
 coll.6 =
-.relation\textsubscript{new}\textasciitilde{}``{[}J=''(''zwei'',
-2)\sout{''{[}{]}=''('\,''drei'', 3)}''{[}J=''('vier'',
-8)\textasciitilde\textasciitilde''J=''C''drei'',3) coll.7 =
-.set\textasciitilde of(2, 3, 9){]}=tmpSupp\textasciitilde ITEM
+.relation\textsubscript{new}\textasciitilde{}``{[}J=''(``zwei'',
+2)\st{``{[}{]}=''('``drei'', 3)}``{[}J=''('vier'',
+8)\textasciitilde\textasciitilde{}``J=''C''drei'',3) coll.7 =
+.set\textasciitilde of(2, 3, 9)
+
+coll.8 = .table\textsubscript{new}\textasciitilde{}``{[}{]}=''(``zwei'',
+2)\st{``{[}J=''(``drei'', 3)}``{[}T{]}J=''C''vier'', 10)
+
+coll.0 = 8
+
+message. 1 = ``UNION''
+
+message.2 = ``INTERSECTION'' \_\_/* index the same in both \emph{/
+message.3 = ``DIFFERENCE'' /} if index only in Ist collection \emph{/
+message.4 = ``XOR'' /} unique index among both collections \emph{/
+message.5 = ``SUBSET'' /} target is subset of other collection */
+
+message. = 5
+
+target. = coll. hstart = 4 istart = \textbar{} jstart = 1
+
+output = 1 setOfTargets = .set\textasciitilde new
+
+SAY ``Test Results of Set Operations on Collection Classes -- dated''
+date(`U')
+
+SAY
+
+DO h= hstart TO target.0 /* loop over target targetID =
+target.h\textsubscript{class}id IF
+\setOfTargets\textasciitilde hasindex(targetID) THEN DO SAY SAY
+CENTER('' Target:'' targetID ````, 70,''=``)
+setOfTargets\textasciitilde put(targetID) output = 1 END
+
+DO i= istart TO coll.0 /* loop over other collections if output then do
+output = 0 argumentID = coll.i\textsubscript{class}id SAY SAY CENTER(''
+argument:'' argumentID ````, 65,''=``) SAY SAY''INPUT:'' SAY ``contents
+of'' pp(targetID) ``target:'' CALL dump\_collection target.h SAY
+
+SAY ``contents of'' pp(argumentID) ``argument:''
+
+CALL dump\_collection colli
+
+SAY
+
+SAY CENTER('' start set operators ``, 65,''-``) end
+
+DO j =jstart TO message.0 /* loop over set operators tmpString
+\textbar{} = RIGHT(``h'' pp(h) ``i'' ppG) ``j'' ppG), 65)
+
+*/
+
+*/
+
+*/
+
+tmpString2 = pp(targetID ``\textasciitilde{}'' message.j
+\textbar\textbar{} ``('' argumentID ``)'')
+
+SAY OVERLAY ( tmpString2, tmpString1 )
+
+/* set resume parameter in case of error*/
+
+jstart = j+1 IF jstart\textgreater message.0 THEN DO istart = i+{]}
+
+IF istart\textgreater coll.0 THEN DO hstart = h+1 istart = 1
+
+END jstart = 1 output = \textbar{} END /* process method invocation */
+IF target.h\textasciitilde hasmethod(message.j) THEN DO tmp =
+.message\textasciitilde new(target.h, message.j, ``I'',
+coll.i)\textasciitilde send if ``The String
+class''=tmp\textsubscript{class}defaultname then do if
+datatype(tmp,``B'') then do
+
+if tmp then SAY'' Result is TRUE'' else SAY'' Result is FALSE'' end end
+else CALL dump\_collection tmp END ELSE
+
+SAY pp(targetID) ``does not have method \textasciitilde{}''
+pp(message.j)
+
+SAY LEFT(````, 40,''-``) END jstart = 1 END jstart = 1 istart = 1 output
+= 1 END
+
+RETURN
+
+dump\_collection:procedure USE ARG collection k =
+.array\textasciitilde new i= .array\textasciitilde new tmpSupp =
+collection\textasciitilde supplier DO WHILE
+tmpSupp\textasciitilde AVAILABLE
+k{[}k\textasciitilde dimension(1)+1{]}=tmpSupp\textasciitilde INDEX
+i\{i\textsubscript{dimension(1)+1{]}=tmpSupp}ITEM
 tmpSupp\textasciitilde NEXT END do until hope hope=1 do j=1 to
@@
 (k{[}j{]}\textsubscript{string=k{[}j+1{]}}string \&
-i{[}j{]}\textsubscript{string\textless i{[}j+1{]}}string) then do{]}=t
+i{[}j{]}\textsubscript{string\textless i{[}j+1{]}}string) then do
+
+t=k{[}j{]}sk(jJ=kUj+ 1 )sk{[}j+1{]}=t t=i{[}j)siG {]}=if+ Usifj+t{]}=t
 hope=0 end end end if O=collection\textasciitilde items then say'' The
 result is empty!'' else do j=1 to k\textasciitilde dimension(1) SAY ''
-'' ``index'' pp(k{[}j{]}) ``item'' ppd{[}j){]}=alj+1{]};a{[}j- 1
-{]}=alj+2{]} end j
+'' ``index'' pp(k{[}j{]}) ``item'' ppd{[}j)
+
+end RETURN /* Auxiliary routines \emph{/ pp: RETURN ``{[}''
+\textbar\textbar{} ARG(1)\textasciitilde string \textbar\textbar{} ``J''
+{[}}
+
+/* X3J18 Rexx Language Standard proposal for the Set-like operations on
+the Collection classes */
+
+/* Tn the same way that the first standard uses BIFs which are defined
+using other BIFs and ultimately the arithmetic and character operators,
+the second standard can define classes using other classes and some
+fundamental basis.
+
+This program gives the definition of some other classes, in a form which
+(when thoroughly developed) might be part of the second standard. It
+also has a testing mechanism, which will not be part of a standard.
+
+This particular program is implementing collections on top of array. */
+
+/* The class Collection is not one builtin, but is used to simplify the
+definition. */
+
+::class `Collection'
+
+:imethod init expose a /* A collection is modelled as using 3 slots in
+an array for each element. The first slot holds the item, the second the
+index, and the third is used by particular types of collection. This
+order of slots is arbitary, chosen to match order of arguments for PUT
+and SUPPLIER\textasciitilde NEW. \emph{/ /} The first set of 3 slots is
+reserved for other purposes, to avoid having separate variables which
+the subclassing would need to access. \emph{/
+a=.alray\textasciitilde new a{[}1{]}/}ItemsCount\emph{/=0
+a{[}2{]}/}Unique*/=0
+
+return self
+
+::method exposed private expose a
+
+/* This method allows subclasses to get at the implementation of
+Collection. */ retum a
+
+::method findindex private expose a /* Returns array index if the
+collection contains any item associated with the index specified or
+returns 0 otherwise. \emph{/ do j=4 by 3 to
+143}a{[}1{]}/\emph{ItemsCount}/ if a{[}j+1{]}==arg(1) then return j end
+j return 0
+
+::method at /* rANY \emph{/ expose a /} Returns the item associated with
+the specified index. */ j=self\textasciitilde findindex(arg(1)) if j=0
+then return .nil return a{[}j{]}
+
+::method `{[}{]}' /* Synonym for the AT method. */ forward message `AT'
+
+::method put /* rANY rANY \emph{/ expose a use arg item, index /}
+Replaces any existing item associated with the specified index with the
+new item. Otherwise adds the item-index pair. \emph{/
+j=self\textasciitilde findindex(index) if j\textgreater0 then do
+a{[}j{]}=item returm end a{[}1{]}/}ItemsCount\emph{/=a{[}1
+{]}/}ItemsCount\emph{/+1 j=lt3}a{[}1{]}/\emph{ItemsCount}/ a{[}j{]}=item
+a{[}j+1{]}=index a{[}j+2{]}=0 return /* Error 91 in OOI if context
+requiring result. */
+
+:method `{[}{]}=' /* Synonym for the PUT method. */
+
+forward message `PUT'
+
+z:method hasindex /* rANY \emph{/ /} Returns \textbar{} (true) if the
+collection contains any item associated with the index specified or
+returns O (false) otherwise. */
+
+return self\textasciitilde findindex(arg(1))\textgreater0
+
+::method items expose a
+
+/* Returns the number of items in the collection. \emph{/ return
+a{[}1{]}/}ItemsCount*/
+
+zimethod remove /* rANY */
+
+expose a /* Returns and removes from a collection the member item with
+the specified index. */
+
+j=self\textasciitilde findindex(arg(1))
+
+if j=0 then return .nil
+
+r=alj{]}
+
+self\textasciitilde removeit(j)
+
+returm r
+
+::method removeit private
+
+expose a
+
+use arg j
+
+/* Remove relevant slots from the array, with compaction. */
+
+do j=j+3 by 3 to 14+3\emph{a{[}1{]}/}ItemsCount*/
+a{[}j-3{]}=alj{]};alj-2{]}=alj+1{]};a{[}j- 1 {]}=alj+2{]} end j
 
@@
 a{[}1{]}/}ItemsCount\emph{/=0 then return .nil return
-a{[}3\emph{a{[}1{]}/}ItemsCount*/+1 \textbar{]}
-a{[}j+1{]}=al{[}j+4{]}-1; /} Index changes */ end j returm r
+a{[}3}a{[}1{]}/\emph{ItemsCount}/+1 \textbar{}
+
+::method next P* rANY \emph{/ a=self\textasciitilde exposed /} Returns
+the index of the item that follows the list item having the specified
+index. */
+
+j=self\textasciitilde findindex(arg(1))
+
+if j=0 then call Raise `Syntax',93.918
+
+jait3
+
+if j\textgreater3\emph{a{[}1{]}/}ItemsCount\emph{/ then return .nil /}
+Next of last was requested. */
+
+return a{[}j+1{]}
+
+zimethod previous /* rANY */
+
+a=self\textasciitilde exposed /* Returns the index of the item that
+precedes the list item having the specified index. */
+
+j=self\textasciitilde findindex(arg(1))
+
+if j=0 then call Raise `Syntax',93.918
+
+ja-3
+
+if j\textless4 then return .nil /* Previous of first was requested. */
+
+return a{[}j+1{]}
+
+zimethod section /* rANY OoWHOLE\textgreater=0 \emph{/
+a=self\textasciitilde exposed /} Returns a new list containing selected
+items from the receiver list. The first item in the new list is the item
+corresponding to the index specified, in the receiver list. \emph{/
+j=self\textasciitilde findindex(arg(1)) if j=0 then call Raise
+`Syntax',93.918 r=.list\textasciitilde new /} To build result in.
+\emph{/ if arg(2,`E') then s = arg(2) else s =
+self\textasciitilde items; do s r\textasciitilde insert(a{[}j{]}) jait3
+if j\textgreater1+3}a{[}1{]}/\emph{ItemsCount}/ then leave end retum r
+
+::class `Queue' subclass Collection /* A queue is a sequenced collection
+with whole-number indexes. The indexes specify the position of an item
+relative to the head (first item) of
+
+the queue. Adding or removing an item changes the association of an
+index to its queue item. */
+
+::method push /* rANY */
+
+/* Adds the object value to the queue at its head. \emph{/
+a=self\textasciitilde exposed a{[}1{]}/}ItemCount\emph{/=a{[}1
+\textbar/}ItemCount*/+1
+
+/* Slide along to make a space. \emph{/ do
+j=1+3}a{[}1{]}/\emph{ItemCount}/ by -3 to 7
+
+aljl=alj-3{]} a{[}j+1{]}=a{[}j-2{]}+1; /* Index changes */ end j
+a{[}4{]}=arg(1) a{[}5{]}=1 return ::method pull
+
+/* Returns and removes the item at the head of the queue. \emph{/
+a=self\textasciitilde exposed if a{[}1{]}/}ItemCount\emph{/=0 then
+return .nil /} Stays empty \emph{/ r=a{[}4{]}
+a{[}1{]}/}ItemCount\emph{/=a{[}1{]}/}ItemCount\emph{/- 1 do j=4 by 3 to
+14+3}a{[}1{]}/\emph{ItemCount}/ alj\}=alj+3{]} a{[}j+1{]}=al{[}j+4{]}-1;
+/* Index changes */ end j returm r
 
@@
 a=self\textasciitilde exposed a{[}1{]}/}ItemCount\emph{/=a{[}1
-\textbar/}ItemCount*/+1
-a{[}\{1+3\emph{a{[}1{]}/}ItemCount*/\textbar=arg(1)
+\textbar/}ItemCount\emph{/+1
+a{[}\{1+3}a{[}1{]}/\emph{ItemCount}/\textbar=arg(1)
 a{[}2+3\emph{a{[}1{]}/}ItemCount\emph{/\textbar=al{[} 1 {]}/}ItemCount*/
-return{]} r{[}n{]}=a{[}j+1{]} end j return
-.supplier\textasciitilde new(m,r){]}=1
+return
+
+:imethod peek
+
+a=self\textasciitilde exposed /* Returns the item at the head of the
+queue. The collection remains unchanged. */
+
+return a{[}4{]}
+
+zimethod remove /* rWHOLE\textgreater O \emph{/ /} Returns and removes
+from a collection the member item with the specified index. \emph{/
+a=self\textasciitilde exposed if
+a{[}1{]}/}ItemCount\emph{/\textless arg(1) then return .nil
+r=self\textasciitilde remove:super(arg(1)) /} Reset the indexes. \emph{/
+k=0 do j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/ k=k+1 a\{j+1J=k
+
+end j return r
+
+::class ``Table' subclass Collection
+
+::method makearray /* Returns a single-index array containing the index
+objects. \emph{/ /} This is different from Collection MAKEARRAY where
+items rather than indexes are in the returned array. */ forward message
+`MAKEARRAYX'
+
+::method union /* rCOLLECTION */ return CommonUnion(self,arg(1))
+
+::method intersection /* rCOLLECTION \emph{/ /} Returns a new collection
+of the same class as SELF that contains the items from SELF that have
+indexes also in the argument. \emph{/ /} Actually an index in SELF can
+only be `matched' with one in the argument once. */
+
+return self\textsubscript{difference(self}difference(arg(1)))
+
+::method xor /* rCOLLECTION \emph{/ /} Returns a new relation that
+contains all items from self and the argument except that all index-item
+pairs that appear in both collections are removed. */ return
+CommonXor(self,arg(1))
+
+::method difference /* rCOLLECTION */ return
+CommonDifference(self,arg(1))
+
+::method subset = /* rCOLLECTION */ return
+self\textsubscript{difference(arg(1))}items = 0
+
+::class `Set' subclass table
+
+/* A set is a collection that restricts the member items to have a value
+that is the same as the index. Any object can be placed in a set. There
+can be only one occurrence of any object in a set. */
+
+/* Second arg same as first. Committee has dropped second? \emph{/
+:rmethod put /} rANY oANY \emph{/ /} Makes the object value a member
+item of the collection and associates it with specified index. \emph{/
+if arg(2,`E') then if arg(2)==arg(1) then signal error /} 949 */
+self\textasciitilde put:super(arg(1),arg(1))
+
+zimethod of class /* 1 or more rANY \emph{/ /} Returns a newly created
+set containing the specified value objects. */ r=self\textasciitilde new
+do j=1 to arg() r\textasciitilde put(arg(j)) end j retum r
+
+::method union /* rCOLLECTION */ return CommonUnion(self,EnBag(arg(1)))
+
+::method intersection /* rCOLLECTION */ return
+self\textsubscript{difference(self}difference(arg(1)))
+
+::method xor /* rCOLLECTION */ return CommonXor(self,EnBag(arg(1)))
+
+::method difference /* rCOLLECTION */ return
+CommonDifference(self,EnBag(arg(1)))
+
+::class `Relation' subclass Collection
+
+::method put /* rANY rANY */
+
+use arg item, index
+
+a=self\textasciitilde exposed /* Makes the object value a member item of
+the relation and associates it with the specified index. If the relation
+already contains any items with the specified index, this method adds a
+new member item value with the same index, without removing any existing
+members */
+
+a{[}1{]}/\emph{ItemsCount}/=a{[}1 {]}/\emph{ItemsCount}/+1
+
+j=lt3\emph{a{[}1{]}/}ItemsCount*/
+
+a{[}j{]}=item
+
+a{[}j+1{]}=index
+
+a{[}j+2{]}=0
+
+return /* Error 91 in OOI if context requiring result. */
+
+:imethod items /* oANY \emph{/ a=self\textasciitilde exposed /} Returns
+the number of relation items with the specified index. If you specify
+
+no index, this method returns the total number of items associated with
+all indexes in the relation. */
+
+if \arg(1,'E') then return a{[}1{]}/\emph{ItemsCount}/
+
+n=0
+
+do j=4 by 3 to 143\emph{a{[}1{]}/}ItemsCount*/
+
+if arg(1)==a{[}j+1{]} then n=n+1 end j retum n
+
+::method makearray forward message `MAKEARRAYX'
+
+zimethod supplier /* oANY \emph{/ a=self\textasciitilde exposed /}
+Returns a supplier object for the collection. If an index is specified,
+the supplier enumerates all of the items in the relation with the
+specified index. \emph{/ m=.array\textasciitilde new /} For the items
+\emph{/ r=.array\textasciitilde new /} For the indexes \emph{/ do j=4 by
+3 to 143}a{[}1{]}/\emph{ItemsCount}/ if arg(1,`E') then if
+arg(1)==a{[}j+1{]} then iterate n=r\textasciitilde dimension(1)+1
+m{[}n{]}=alj{]} r{[}n{]}=a{[}j+1{]} end j return
+.supplier\textasciitilde new(m,r)
+
+::method union /* rCOLLECTION \emph{/ /} Union for a relation is just
+all of both. */ r=self\textsubscript{class}new
+cs=self\textasciitilde supplier do while cs\textasciitilde available
+r{[}cs\textasciitilde index{]}=cs\textasciitilde item
+cs\textasciitilde next end cs=MayEnBag(arg(1))\textasciitilde supplier
+do while cs\textasciitilde available
+r{[}cs\textasciitilde index{]}=cs\textasciitilde item
+cs\textasciitilde next end return r
+
+::method intersection /* rCOLLECTION */ return
+self\textsubscript{difference(self}difference(arg(1)))
+
+::method xor /* rCOLLECTION */ return CommonXor(self,arg(1))
+
+zimethod difference /* rCOLLECTION \emph{/ /} Returns a new relation
+containing only those index-item pairs from the
+
+163 SELF whose indexes the other collection does not contain. */
+r=self\textsubscript{class}new cs=self\textasciitilde supplier do while
+cs\textasciitilde available
+r{[}cs\textasciitilde index{]}=cs\textasciitilde item
+cs\textasciitilde next end cs=MayEnBag(arg(1))\textasciitilde supplier
+do while cs\textasciitilde available
+r\textsubscript{removeitem(cs}item,cs\textasciitilde index)
+cs\textasciitilde next end return r
+
+::method subset = /* rCOLLECTION */ return
+self\textsubscript{difference(arg(1))}items = 0
+
+zimethod removeitem /* rANY rANY \emph{/ a=self\textasciitilde exposed
+/} Returns and removes from a relation the member item value (associated
+with the specified index). If value is not a member item associated with
+index index, this method returns the NIL object and removes no item.
+\emph{/ do j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/ if a\{jJ==arg(1)
+\& a{[}j+1{]}==arg(2) then do self\textasciitilde removeit(j) return
+arg(1) end end j return .nil
+
+:rmethod index /* rANY \emph{/ a=self\textasciitilde exposed /} Returns
+the index for the specified item. If there is more than one index
+associated with the specified item, the one this method returns is not
+defined. \emph{/ do j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/ if
+arg(1)==a{[}j{]} then return a{[}j+1{]} end j return .nil
+
+::method allat /* rANY */
+
+a=self\textasciitilde exposed /* Returns a single-index array containing
+all the items associated with the specified index. */
+
+r=.array\textasciitilde new
+
+do j=4 by 3 to 143\emph{a{[}1{]}/}ItemsCount*/
+
+if arg(1)==a{[}j+1{]} then r{[}r\textasciitilde dimension(1)+1{]} =
+a{[}j{]} end j return r
+
+::method hasitem /* rANY rANY \emph{/ a=self\textasciitilde exposed /}
+Returns 1 (true) if the relation contains the member item value
+(associated with specified index). Returns 0 (false) otherwise. \emph{/
+do j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/ if a\{jJ==arg(1) \&
+a{[}j+1{]}==arg(2) then return \textbar{} end j return 0
+
+z:method allindex /* rANY \emph{/ a=self\textasciitilde exposed /}
+Returns a single-index array containing all indexes for the specified
+item. \emph{/ r=.array\textasciitilde new do j=4 by 3 to
+143}a{[}1{]}/\emph{ItemsCount}/ if a{[}j{]}==arg(1) then do
+r{[}r\textasciitilde dimension(1)+1{]}=a{[}j+1{]} end end j returm r
+
+::class `Bag' subclass relation
+
+/* A bag is a collection that restricts the member items to having a
+value that is the same as the index. Any object can be placed in a bag,
+and the same object can be placed in a bag multiple times. */
+
+zimethod of class /* 1 or more rANY \emph{/ /} Returns a newly created
+bag containing the specified value objects. */ r=self\textasciitilde new
+do j=1 to arg() r\textasciitilde put(arg(j)) end j retum r
+
+:rmethod put /* rANY oANY \emph{/ /} Committee does away with second
+argument? \emph{/ /} Makes the object value a member item of the
+collection and associates it with the specified index. If you specify
+index, it must be the same as value. */ if arg(2,`E') then if
+arg(2)==arg(1) then signal error
+self\textasciitilde put:super(arg(1),arg(1))
+
+/* Bag may be a subclass of relation but many methods have different
+semantics. */
+
+::method union /* rCOLLECTION */ return CommonUnion(self,EnBag(arg(1)))
+
+::method intersection /* rCOLLECTION */ return
+self\textsubscript{difference(self}difference(arg(1)))
+
+::method xor /* rCOLLECTION */ return CommonXor(self,EnBag(arg(1)))
+
+::method difference /* rCOLLECTION */ return
+CommonDifference(self,EnBag(arg(1)))
+
+::class `Directory' subclass Collection
+
+/* Later we take three array elements for each element in the directory,
+one for the item, one to contain the index, one to say if the item is a
+method
+
+to be run or not. */
+
+::method at /* rANY \emph{/ a=self\textasciitilde exposed /} Returns the
+item associated with the specified index. \emph{/
+j=self\textasciitilde findindex(arg(1)) if j=0 then return .nil /} Run
+the method if there is one. */ if a{[}j+2{]} then return
+self\textasciitilde run(a{[}j{]}) return a{[}j{]}
+
+::method put /* rANY rANY \emph{/ a=self\textasciitilde exposed /} Makes
+the object value a member item of the collection and associates it with
+the specified index. */ if
+\arg(2)\textasciitilde hasmethod(MAKESTRING'\,') then call Raise
+`Syntax', 93.938 self\textsubscript{put:super(arg(1),arg(2)}makestring)
+return
+
+::method makearray forward message `MAKEARRAYX'
+
+:imethod supplier a=self\textasciitilde exposed
+
+/* Returns a supplier object for the directory. */
+
+/* Check out what happens to the SETENTRY fields. \emph{/
+r=.array\textasciitilde new /} For items */
+
+do j=4 by 3 to 143\emph{a{[}1{]}/}ItemsCount*/
+r{[}r\textasciitilde dimension(1)+1{]}=a{[}j{]} end j
+
+return .supplier\textsubscript{new(r,self}makearray)
+
+::method union /* rCOLLECTION */ return CommonUnion(self,arg(1))
+
+::method intersection /* rCOLLECTION */ return
+self\textsubscript{difference(self}difference(arg(1)))
+
+::method xor /* rCOLLECTION */ return CommonXor(self,arg(1))
+
+::method difference /* rCOLLECTION */ return
+CommonDifference(self,arg(1))
+
+::method subset = /* rCOLLECTION */ return
+self\textsubscript{difference(arg(1))}items = 0
+
+z:method setentry /* rSTRING oANY \emph{/ a=self\textasciitilde exposed
+/} Sets the directory entry with the specified name (translated to
+uppercase) to the second argument, replacing any existing entry or
+method for the specified name. \emph{/ n=translate(arg(1))
+j=self\textasciitilde findindex(n) if j=0 \& \arg(2,`E') then return if
+\arg(2,`E') then do /} Removal \emph{/ self\textasciitilde removeit(j)
+returm end if j=0 then do /} It's new \emph{/
+a{[}1{]}/}ItemsCount\emph{/=a{[}1{]}/}ItemsCount\emph{/ +1
+j=lt3}a{[}1{]}/\emph{ItemsCount}/ a{[}j+1{]}J=n end a{[}j{]}=arg(2)
+a{[}j+2{]}=0
+
+returm
+
+::method entry /* rSTRING */
+
+a=self\textasciitilde exposed /* Returns the directory entry with the
+specified name (translated to uppercase). */
+
+n=translate(arg(1))
+
+j=self\textasciitilde findindex(n) /\emph{if j=0 then signal error
+according to online }/ /* Online has something about running UNKNOWN.
+\emph{/ if j=0 then return .nil /} Tf there is an entry decide whether
+to invoke it. */ if a\textasciitilde hasindex(j) then do if \a[j+2] then
+return a{[}j{]} return self\textasciitilde run(a{[}j{]}) end
+
+z:method hasentry /* rSTRING \emph{/ /} Returns 1\textbar{} (true) if
+the directory has an entry or a method for the specified name
+(translated to uppercase) or O (false) otherwise. */
+
+return self\textasciitilde findindex(translate(arg(1)))\textgreater0
+
+z:method setmethod /* rSTRING oMETHOD \emph{/
+a=self\textasciitilde exposed /} Associates entry with the specified
+name (translated to uppercase) with method method. Thus, the language
+processor returns the result of running method when you access this
+entry. \emph{/ /} (Part of METHOD checking converts string or array to
+actual method.) \emph{/ n=translate(arg(1))
+j=self\textasciitilde findindex(n) if j=0 \& \arg(2,`E') then return if
+\arg(2,'E') then do self\textasciitilde removeit(j) returm end if j=0
+then do /} It's new \emph{/
+a{[}1{]}/}ItemsCount\emph{/=a{[}1{]}/}ItemsCount\emph{/ +1
+j=lt3}a{[}1{]}/\emph{ItemsCount}/ a{[}j+1{]}J=n end a{[}j{]}=arg(2)
+a\{j+2{]}=1
 

### 2023-06-02 ARB/standards/work-in-progress/document/tex/standard/syntax_constructs.tex
add cyrillic.rexx, rename unicode.md to Unicode_Requirements.md

@@
   characters that are not operator\_char ) to the top syntax level. Any
-  occurrence of an ofher\_negator within Operator is supplied as '\,';
+  occurrence of an ofher\_negator within Operator is supplied as
+  '\textquotesingle;
 \item
@@
 point, without consideration of unprocessed characters and tokens. If a
-token which is `+', `-', ' or `(' appears in a lexical level context
-(other than after the keyword `PARSE') where the keyword `VALUE' could
-appear in the corresponding top level context, then `VALUE' is passed to
-the top level before the token is passed. If an `=' operator\_char
-appears in a lexical level context where it could be the `=' of an
-assignment or message\_instruction in the corresponding top level
-context then it is recognized as the `=``' of that instruction. (It will
-be outside of brackets and parentheses, and any Var\_symbo/ immediately
-preceding it is passed as a VAR\_SYMBOL). If an operand is followed by a
-colon token in the lexical level context then the operand only is passed
-to the top level syntax as a LABEL, provided the context permits a
-LABEL. Except where the rules above determine the token passed, a
-Var\_symbol is passed as a terminal (a keyword) rather than as a
-VAR\_SYMBOL under the following circumstances:
+token which is `+', `-', \textquotesingle{} or `(' appears in a lexical
+level context (other than after the keyword `PARSE') where the keyword
+`VALUE' could appear in the corresponding top level context, then
+`VALUE' is passed to the top level before the token is passed. If an `='
+operator\_char appears in a lexical level context where it could be the
+`=' of an assignment or message\_instruction in the corresponding top
+level context then it is recognized as the `=``' of that instruction.
+(It will be outside of brackets and parentheses, and any Var\_symbo/
+immediately preceding it is passed as a VAR\_SYMBOL). If an operand is
+followed by a colon token in the lexical level context then the operand
+only is passed to the top level syntax as a LABEL, provided the context
+permits a LABEL. Except where the rules above determine the token
+passed, a Var\_symbol is passed as a terminal (a keyword) rather than as
+a VAR\_SYMBOL under the following circumstances:
 
@@
 `WHEN' Msg9.1 \textbar{} `OTHERWISE' Msg9.2 command = expression address
-= `ADDRESS' {[}(taken constant {[}expression{]}{]}
+= `ADDRESS' {[}(taken constant {[}expression{]}
+
+Msg19.1 \textbar{} valueexp) {[} `WITH' connection{]} {]}
 
@@
 power expression'**' prefix expression prefix expression := (`+'
-\textbar{} `-' \textbar{} '\,') prefix expression term \textbar{}
-Msg35.1 /* ``Stub'' has to be identified semantically? */
+\textbar{} `-' \textbar{} '\textquotesingle) prefix expression term
+\textbar{} Msg35.1 /* ``Stub'' has to be identified semantically? */
 

### 2023-08-02 ARB/standards/work-in-progress/unicode/Unicode.Implimentation.tex
Add files via upload
Working paper Unicode implementation written in LaTeX anud compiled with lualatex in two passes.

% \documentclass[draft]{article}
\documentclass[b4paper]{article}
\usepackage{enumitem}
\usepackage{eqparbox}
\usepackage{expl3}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{listings}         %Requires luatex engine to handle UTF-8
                              %Consider using minted, but ...
\usepackage{upquote}

\newlist{definition}{description}{2} %% <- pick a larger number if you want to nest these
\setlist[definition]{
    labelwidth=\eqboxwidth{listlabel@\EnumitemId},
    leftmargin=!,
    format=\definitionlabel,
}
\newcommand\definitionlabel[2][l]{\eqmakebox[listlabel@\EnumitemId][#1]{#2}}

\newlist{resource}{description}{2} %% <- pick a larger number if you want to nest these
\setlist[resource]{
    labelwidth=\eqboxwidth{listlabel@\EnumitemId},
    leftmargin=!,
    format=\resourcelabel,
}
\newcommand\resourcelabel[2][l]{\eqmakebox[listlabel@\EnumitemId][#1]{#2}}

\ExplSyntaxOn
\NewDocumentCommand\enginedetails{}%
  {
    \c_sys_engine_exec_str
    \c_space_tl
    \c_sys_engine_version_str
    \c_space_tl
    (\c_sys_engine_format_str)
  }
\ExplSyntaxOff

\title{Unicode Implementation Issues%
  \thanks
    {
    This is a working document for the Rexx Language Association (RexxLA).
    There will be one or more separate requirements documents.
    It is written in \LaTeX\ \fmtversion\ and was rendered \today\ using \enginedetails.
    }
  }
\author{Shmuel (Seymour J.) Metz}

\begin{document}
\lstset{language=Rexx, extendedchars=true, frame=trbl, escapechar={!}, linewidth=20em}

\maketitle

\begin{abstract}
Classic Rexx supports character code pages in which every character can be represented in a single octet,
and treats only ASCII letters as alphabetic.
Unicode is nominally a 32-bit character set, with code points restricted to 17 16-bit planes,
i.e., code points U+00 through U+10FFFF.
This document describes basic issues and solutions for extending Rexx to support Unicode.
\end{abstract}

\section{Background}
Many modern languages support Unicode encoded as
\href{https://datatracker.ietf.org/doc/rfc3629/}{UTF-8} (\cite{RFC 3629}),
and the \href{https://www.ietf.org/}{IETF}, in \cite{RFC 5198},
has mandated the use of UTF-8 with NFC in new protocols.
Rexx will need to support UTF-8 if it is not to become a backwater,
and that support should not break existing code.

\section{Scope}
This document only describes issues related to expanded alphabets and multi-octet representations of characters;
it does not address such issues as bidirectional text.
It presents suggested solutions for requirements that are presented in other documents.
Also, most of the details are under active discussion, so everything here is provisional.

\section{Nomenclature}

The definitions given in \cite{Glossary} and \cite{Unicode} take
precedence over those given here. Quoted text is taken from those sources.
Except for definitions taken from official IETF and Unicode documents,
the nomenclature used here is illustrative rather than normative;
language design teams will formally define, e.g., method names, encoding of parameters.
Most of the sections assume that there will be destinct string types
for extended grapheme clusters, Unicode code points, legacy code pages
and raw octets. However, there has been discussion of including
metadata, in which case come of the classes might be merged.

\begin{definition}
\item [.Legacy] The string class for strings in legacy code pages.
\item [.octets] The string class for uninterprted octet strings.
\item [.Ucp] The string class for strings of Unicode code points.
\item [.Ugc] The string class for strings of Unicod extended grapheme clusters.
\item [\textbf{\'{ }}] U+0301, COMBINING ACUTE ACCENT
\\*
Set in bold to distinguish it from U+27, APOSTROPHE.
\item [Bidi] "Abbreviation of bidirectional, in reference to mixed left-to-right and right-to-left text."
\item [BIF] Built In Function
\item [BIM] Built In Method
\item [Block] "A grouping of characters within the Unicode encoding
space used for organizing code charts. Each block is a uniquely named,
continuous, non-overlapping range of code points, containing a
multiple of 16 code points, and starting at a location that is a
multiple of 16. A block may contain unassigned code points, which are
reserved."
\item [BMP] Basic Multilingual Plane
\\*
The first 64 Ki code points of Unicode, from U+0000 to U+FFFF.
\item [BOM] Byte Order Mark: U+FEFF, ZERO WIDTH NON-BREAKING SPACE (ZWNBSP)
\\*
Used as the first character to indicate byte order for UCS-2, UTF-16 and UCS-4;
\\*
Optional as the first character for UTF-8.
\item [\'{E}] U+C9, LATIN CAPITAL LETTER E WITH ACUTE
\\*
Set in normal weight: U+45 U+0301 will be shown as "E\textbf{\'{ }}".
\item [\'{e}] U+E9, LATIN SMALL LETTER E WITH ACUTE
\\*
Set in normal weight: U+65 U+0301 will be shown as "e\textbf{\'{ }}".
\item [encoded character] The smallest constituant of a Unicode string.
"The Unicode Standard does not define what is and is not a text
element in different processes; instead, it defines elements called
encoded characters. An encoded character is represented by a number
from 0 to 10FFFF\textsubscript{16} called a code point."
\item [EGC] Extended grapheme cluster
\item [GCGID] "Acronym for Graphic Character Global Identifier.
These are listed in the IBM document
Character Data Representation Architecture, Level 1, Registry SC09-1391."
See \url{https://www.ibm.com/downloads/cas/G01BQVRV}.
\item [grapheme cluster]
"A grapheme cluster consists of a base character followed by any
number of continuing characters, where acontinuing character may
include any nonspacing combining mark, certain spacing combining
marks, or a join control."
{\cite{Annex29}} defines two types of grapheme clusters;
"An extended grapheme cluster is the same as a legacy grapheme cluster,
with the addition of some other characters.
The continuing characters are extended to include all spacing combining marks,
such as the spacing (but dependent) vowel signs in Indic scripts."
\item [high surrogate] A code point in the range U+D800-U+DBFF, used
as the first half of a surrogate pair.
\item [IETF] \href{https://ietf.org}{Internet Engineering Task Force}
\item [introducer] The first octet in the UTF-8 encoding of a Unicode character beyond U+7F
\item [low surrogate] A code point in the range U+DC00-U+DFFF, used
as the second half of a surrogate pair.
\item [NFC]  Unicode
"Normalization Form C (NFC). A normalization form that erases any
canonical differences, and generally produces a composed result. For
example, a + umlaut is converted to \symbol{"00E4} in this form. This form most
closely matches legacy usage. The formal definition is D120 in Section
3.11, Normalization Forms."
\\*
the normalization endorsed by the IETF
\item [NFD] Unicode
"Normalization Form D (NFD). A normalization form that erases any
canonical differences, and produces a decomposed result. For example,
\symbol{"00E4} is converted to a + umlaut in this form. This form is most often
used in internal processing, such as in collation. The formal
definition is D118 in Section 3.11, Normalization Forms."

\item [NFKC] Unicode
"Normalization Form KC (NFKC). A normalization form that erases both
canonical and compatibility differences, and generally produces a
composed result: for example, the single \symbol{"01C6} character is converted to d
+ \symbol{"017E} in this form. This form is commonly used in matching. The formal
definition is D121 in Section 3.11, Normalization Forms."

\item [NFKD] Unicode
"Normalization Form KD (NFKD). A normalization form that erases both
canonical and compatibility differences, and produces a decomposed
result: for example, the single \symbol{"01C6} character is converted to d + z +
caron in this form. The formal definition is D119 in Section 3.11,
Normalization Forms."

\item [octet] 8-bit byte
\item [Plane] "A range of 65,536 (1000016) contiguous Unicode code
points, where the first code point is an integer multipleof 65,536
(1000016). Planes are numbered from 0 to 16, with the number being the
first code point of the plane divided by 65,536. Thus Plane 0 is
U+0000..U+FFFF, Plane 1 is U+10000..U+1FFFF, ..., and Plane 16 (1016)
is U+100000..10FFFF. (Note that ISO/IEC 10646 uses hexadecimal
notation for the plane numbers-for example, Plane B instead of Plane
11). (See Basic Multilingual Plane and supplementary planes.)"
\item [RFC] Request For Comments
\\*
A formal document published by the IETF defining, e.g., a protocol.
RFC documents contain technical specifications and organizational notes for the Internet.
\begin{itemize}
  \item Best Current Practice (BCP)
  \item Experimental
  \item Informational
  \item Proposed Standard
  \item Internet Standard (STD)
  \item Historic
\end{itemize}
\item [surrogate] A code point in the range U+D800-U+DFFF used to
encode 21-bit code points into pairs of 16-bit bytes.
\item [surrogate pair] A high surrogate (in the range U+D800-U+DBFF)
followed by a low surrogate (in the range U+DC00-U+DFFF),
collectively representing a 21-bit code point.
\item [TBD] To Be Determined.
\item [UCS] Universal Character Set, ISO 10646, roughly equivalent to Unicode
\item [UCS-2] A 16 bit subset of Unicode, containing only the BMP.
\item [The Unicode Consortium] A non-profit corporation devoted to
developing, maintaining, and promoting software internationalization standards and data.
\item [UTF-8] UCS Transformation Format 8 (\cite{RFC 3629}).
\\*
The encoding of Unicode endorsed by the IETF
\item [UTF-8 octet sequence]
The sequence of octets representing a single Unicode code point.
It may consist of a single ASCII character padded on the left with a zero bit,
or of a one octet introducers followed by a 1-3 octet tail.

\begin{tabular}{| r | l |}
\hline
Code points & UTF-8 octet sequence \\
\hline
U+0000 - U+007F    & 0xxxxxxx \\
\hline
U+0080 - U+07FF    & 110xxxxx 10xxxxxx \\
\hline
U+0800 - U+FFFF    & 1110xxxx 10xxxxxx 10xxxxxx \\
\hline
U+10000 - U+10FFFF & 11110xxx 10xxxxxx 10xxxxxx 10xxxxxx \\
\hline
\end{tabular}
\item [UTF-16] UCS Transformation Format 16 (\cite{RFC 2781}).
\end{definition}

Some code samples use ooRexx notation. Those samples are illustrative rather than normative.

\section{Statement of problem}

\subsection{Primary Problem}

In Rexx, everything is a string; there are no classes, declarations or types.

While the ANSI standard does not mandate any particular character width for \textbf{Config\_C2B()},
most if not all implementations use a width of 8,
and the length of \textbf{c2x(}\textit{foo}\textbf{)}
is twice the length of \textit{foo}.

A large body of existing code operates on binary data from external files,
or accessed from memory via the \textbf{storage()} BIF,
under the assumption that, e.g., \textbf{c2x()}, left(), \textbf{right()}, \textbf{substr()}, operate on octets.
A UTF-8 introducer is treated the same as any other value.

A large body of existing code operates on text under the assumption that those facilities operate on characters.

The two categories overlap.

There is no conflict as long as each character is contained within a single octet.
However, Unicode has code points beyond U+FF,
and UTF-8 encoding of non-ASCII Unicode characters will
require more than a single octet even for code points less than U+100 if they are beyond U+7F.

For example, the Unicode string 'Caf\'{e}' has 4 code points, 4
grapheme clusters and 5 octets in UTF-8 encoding while
'Cafe\textbf{\'{ }}' has 5 code points, 4 grapheme clusters and 6 octets
in UTF-8 encoding, yet many text processing applications need to treat
them as equivalent.

\subsection{Secondary Problems}
Rexx has built-in case conversion, but it is based on ASCII and
can't even convert the accented letters found in, e.g., ISO-8859-1
(Latin-1), ISO-8859-15 (Latin-9), much less all those found in Unicode.

\begin{lstlisting}
    foo = 'Ren!\'{e}!'
    parse upper var foo bar
\end{lstlisting}

sets bar to REN\'{e} rather than to REN\'{E}.

In current implementations the range of characters is extremely small,
and thus the \textbf{xrange()} only returns short strings.
With Unicode the range expands to U+00 through U+10FFFF.
That is far too large to allow as either an explict or implicit range in \textbf{xrange()}.

The current definition restricts symbols to ASCII characters,
which excludes letters used in many languages. If the rules are extended to include
Unicode letters and digits beyond ASCII, the rules for equality of symbols must be addressed.
Are composed and decomposed strings identical?
Are base characters identical to their compatibility alterrnates?
Are subscript and superscript digits distinct from their ASCII
counterparts?

When dealing with Unicode, different code sequences may have identical
rendering due to the existence of both fully composed and combining
code points and the exixtence of compatibility code points.
Rexx needs a way to test two Unicode strings for equivalence.

There has been some discussion claiming a need to continue supporting legacy code pages.

There has been some discussion of the need to detect unassigned code points.

\section{Solutions}

Most of this section assumes that there will be three distict string
types, and that some methods will not exist in all three types, or
will behave differently.
The type names used here are placeholders, and will be replaced once
there is consensus on what to call them.

\subsection{\textbf{OPTIONS} statement or directive}

Define an option on the OPTIONS statement, or on a similar directive,
to specify either raw octets or Unicode characters.
This breaks programs that operate on both binary data and text.

OPTIONS mighy also specify a source encoding parameter, overriding any code page in an environment variable or file metadata.

\subsection{Types}

Allow Rexx variables to contain three distinct types of data: raw octets, Unicode code points and text,
Define \textbf{storage()} as returning raw octets,
and provide conversion functions ("casts").
There is an ongoing discussion as to whether all three are needed.

There should be no implicit conversion between \textbf{.octets} and Unicode strings.

There should be implicit conversion between Unicode types, but it need not be reversible, i.e.,
a conversion from .Ucp to .Ugc and back need not produce the same code points as the original

If support for legacy code pages is needed, a fourth string type could be defined.
Strings of this type could include a code page attribute.

\subsection{Width parameter}

Add a width parameter to \textbf{c2x()} for Unicode code point string
(raw octet strings may require input and output widths for UCS-2 AND UTF-16 data, and the
utility of \textbf{c2x()} for grapheme clusters needs more
analysis), and raise conditions with distinct error codes if any code
point is out of range or if an invalid UTF-8 sequence is detected.  A
case could be made for using either the bit size or the digit size as
the width.

Assuming that the width is in in terms of octets
\begin{lstlisting}[linewidth=35em]
    foo = 'Ren!\'{e}!'
    bar = foo~UTF-8
    /* assumes that width parameter is in octets */
    say foo~c2x(2) /* Displays 00520065006E00E9  */
    say bar~c2x    /* Displays 52656EC3A9        */
\end{lstlisting}

There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
\begin{lstlisting}[linewidth=35em]
    foo = .octets~new
    foo = 'Ren!\'{e}!'
    !\textellipsis!
    parse var foo ASCII !{\textquotesingle}\'{e}{\textquotesingle}! .
    say ASCII                     /* Displays Ren */
\end{lstlisting}

If there is no raw octet string added to the language,
then \textbf{c2x()} might behave differently depending on whether an explicit width is provided.

\subsection{Constants}
Other languages allow specifying Unicode characters using either
the hexadecimal value of the code point or the assigned name of the code point,
e.g., U+E9 might be coded as {\textbackslash}u\{E9\} or {\textbackslash}u[LATIN SMALL LETTER E WITH ACUTE].
The syntax used for such constants should be consisten across Rexx variants,
including rules for optional spaces between (hex) digits,
and should take into account the recommendations in \cite{RFC 5137}.
Implementations should use the machine readable data bases published by
the Unicode Consortium in order to ease migration to new versions of Unicode.

The form {\textquotesingle}U+digits U+digits ...{\textquotesingle}U is clearer,
but may break code that abuts the variable U with a string literal.
The form {\textquotesingle}U+digits U+digits ...{\textquotesingle}X is acceptable.
There should be a discussion of syntax for named Unicode constants, e.g.,
{\textquotesingle}[COMBINING ACUTE ACCENT]{\textquotesingle}U is
equivalent to {\textquotesingle}U+0301{\textquotesingle}U.
There has been some discussion of using the form
{\textquotesingle}\textellipsis{\textquotesingle}T and of escape conventions
used in other languages.

Binary and hexadecimal literals are of type \textbf{.octets} and cannot be implicitly coerced to Unicode strings.
Other string literals are legacy or Unicode and cannot be implictly coerced to \textbf{.octets}.

An alternative is to add a new notation using the ASCII \verb|`| as a
framing character. This has the potential issue that it may be
difficult to visually distinguish  \verb|`| from  \verb|'|.

Another alternative is to use literals of the forms
{\textquotesingle}\textellipsis{\textquotesingle}type:codepage,
{\textquotesingle}\textellipsis{\textquotesingle}U:codepoints and
{\textquotesingle}\textellipsis{\textquotesingle}U:clusters. However, that makes it incompatinle with ooRexx.

The forms {\textquotesingle}\textellipsis{\textquotesingle}\{modifier\} and
{\textquotesingle}\textellipsis{\textquotesingle}type\{modifier\}
have no obvious conflict with existing syntax and has no abutment issues.

The following table is intended as a discussion point and should be updated whenever a consencuss is reached.

\begin{tabular}{| l | l | l |}
\hline
Example & Type & Semantics \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}              & Ucp & Unicode text by code point \\
                                                          & Ugc & Unicode text by grapheme cluster (TBD) \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}R               & Ucp & Unicode text by code point \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}Ucp             & Ucp & Unicode text by code point \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}Ugc             & Ucp & Unicode text by grapheme cluster  \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ucp\}       & Ucp & Unicode text by code point \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ugc\}       & Ucp & Unicode text by grapheme cluster  \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ISO8859-1\} & Legacy & ISO 8859-1 legacy text \\
\hline
\end{tabular}

\subsection{Error detection}
The condition names given below are illustrative and not normative.
Unicode BIF/BIMs should detect invalid input and signal the following conditions, with unique error codes.

\begin{definition}
\item [INVALIDCODEPOINT]%
But carefully read 3.2 Conformance Requirements,
Code Points Unassigned to Abstract Characters, p. 77, in \cite{Unicode}.
\item[INVALIDUTF]%
An invalid UTF-8 octet sequence or invalid use of a UTF-16 surrogate.
\item [NOENCODING]%
The operation requested requires a code page name.
\item [NOTEXT]%
An operation was requested that is not valid for a binary file.
\item [RANGE]%
A code point or other numeric entioty excees the permitted range.
This might occur, e.g., when an application requuires that characters be limited to the BMP.
\end{definition}

\subsection{I/O}

The \textbf{command} method of the stream classes should support a \textbf{CODEPAGE} option;
attempting to set a codepage for a binary file should raise a \textbf{NOTTEXT} condition with a distinct error code.
There should be an option or suboption controling whether to use Unicode if an apparent BOM is detected.

The default should probably be UTF-8 or ISO 8859-1 with switching to Unicode if a BOM is detected.

A discussion is needed on whether and when to create or discard byte order marks.

\subsection{Case folding}

The rules for case folding should take into account mathematical usage.
\begin{itemize}
\item U+1D400 through U+1D7FF are semantically distinct from other letters.
\item Superscripts and subscripts have semantic significance.
\end{itemize}

\subsection{Raw octet strings}
The methods should include those of the ooRexx .string class, except that
\begin{itemize}
\item The unit of operation is the uninterpreted octet.
\item The \textbf{makeString} method requires an encoding parameter.
\item Parameters are raw octet strings.
\item There are no caseless methods.
\end{itemize}

\subsection{Legacy strings}

If there is a .Legacy string type then the methods should have the
same semantics as the existing methods for the .string class.
If the new standard has a code page attribute then there should be an
access method for it and the init or new method should allow
specifying the code page.

The following Additional methods should be defined:
\begin{definition}
\item [iconv] %
Convert from one code page to another and return a legacy string.
\item [makeCodePageString] %
Return a Unicode string in which individual code points can be accessed
\item [makeGraphemClusterString] %
Return a Unicode string in which only complete grapheme clusters can be accessed
\end{definition}

\subsection{Unicode strings}

There should be subtypes depending on whether the
unit of operation is the Unicode code point or the grapheme cluster.
The methods should be those of the ooRexx .string class except:
\begin{itemize}
\item
The caseless comparisons will use the \cite{UCD}.
There will be variants to preserve or remove accents\footnote{Is Enye (\~n) considered an accented letter?}.
\item There should be no \textbf{bit\textellipsis} methods.
\item Add an optional third parameter to the \textbf{upper} and \textbf{lower} methods
to control whether to translate mon-ASCII characters and whether to strip accents.
\item
The following Additional methods should be defined:
\begin{definition}
\item [makeCodePageString]
Return a Unicode string in which individual code points can be accessed
\item [makeGraphemClusterString]
Return a Unicode string in which only complete grapheme clusters can be accessed
\item [makeLegacyString]
Return a legacy string for the specified code page in which individual octets can be accessed
\item [makeOctets]
Return a raw octet string using a specified encoding.\footnote{default
to UTF-8 or raise NOENCODING if no encoding specified?}
\item [NFC] Return a string normalized with Normalization Form C
\item [NFD] Return a string normalized with Normalization Form D
\end{definition}
\end{itemize}

There should be a description of how to handle conversion from Unicode code
points and grapheme clusters that do not exist in the target code page.
\begin{itemize}
\item Replace with U+1A (SUB) or other specified character
\item Raise SYNTAX with a unique error code
\item Raise a new condition name with a unique error code
\end{itemize}

\subsection{Methods}
The following table summarizes some of the methods that differ among string types.

\begin{tabular}{| l | c | c | c | c |}
\hline
Method               & \textbf{.Legacy} & \textbf{.octets} & \textbf{.Ucp}  & \textbf{.Ugc} \\
\hline
\textbf{{[}{]}}      & by octet         & by octet         & by code point  & by cluster    \\
\textbf{\&}          & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
\textbf{|}           & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
\textbf{\&\&}        & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
\textbf{=}           & by octet         & by octet         & by code point  & by cluster    \\
\textbf{{\neg}=}     & by octet         & by octet         & by code point  & by cluster    \\
\textbf{><}          & by octet         & by octet         & by code point  & by cluster    \\
\textbf{<>}          & by octet         & by octet         & by code point  & by cluster    \\
\textbf{<}           & by octet         & by octet         & by code point  & by cluster    \\
\textbf{<=}          & by octet         & by octet         & by code point  & by cluster    \\
\textbf{>}           & by octet         & by octet         & by code point  & by cluster    \\
\textbf{{\neg}>}     & by octet         & by octet         & by code point  & by cluster    \\
\textbf{abbrev}      & by octet         & by octet         & by code point  & by cluster    \\
\textbf{bitand}      & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
\textbf{bitor}       & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
\textbf{bitxor}      & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
\textbf{c2b}         & 8 bits           & default 8 bits   & Explicit width & \textbf{n/a}  \\
\textbf{c2x}         & 8 bits           & default 8 bits   & Explicit width & \textbf{n/a}  \\
\textbf{center}      & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
\textbf{change}      & \textbf{.Legacy} & \textbf{.octets} & by code point  & by cluster    \\
\textbf{find}        & by octet         & by octet         & by code point  & by cluster    \\
\textbf{index}       & by octet         & by octet         & by code point  & by cluster    \\
\textbf{left}        & by octet         & by octet         & by code point  & by cluster    \\
\textbf{makeArray}   & by octet         & by octet         & by code point  & by cluster    \\
\textbf{makeString}  & by octet         & by octet         & by code point  & by cluster    \\
\textbf{pos}         & by octet         & by octet         & by code point  & by cluster    \\
\textbf{right}       & by octet         & by octet         & by code point  & by cluster    \\
\textbf{strip}       & by octet         & by octet         & by code point  & by cluster    \\
\textbf{substr}      & by octet         & by octet         & by code point  & by cluster    \\
\textbf{verify}      & by octet         & by octet         & by code point  & by cluster    \\
\hline
\end{tabular}
\\*
\\*
The backslash (``\textbackslash'') may be used in place of the Logical Not (``\neg'').

In addition, the makeString and makeString methods should allow caseless option
parameters to control the class and attributes of the returned array
elements or string, including:

\begin{definition}
\item [Clusters] Unicode text with extended grapeme clusters as the abstract units.
\item [Codepoints] Unicode text with Unicode scalars as the abstract units.
\item [cp=legacy code page] Legacy or octet string with specified encoding.
\item [cp=UTF-8]
Legacy or octet string with UTF-8 encoding;
functions like center and left will give unexpected results.
\item [Legacy] Legacy string with default encoding unless cp= is also specified.
\item [raw] Raw octet string; cp= must be specified.
\end{definition}

\subsection{Coercions}
In addition to explicit casts via BIF/BIM, the following promotions
from code-point strings to grapheme-cluster strings will be automatic

\begin{itemize}
\item Concatenation of code-point strings with grapheme-cluster strings.
will be grapheme-cluster strings.
\item Code point search arguments for grapheme-cluster strings.
will be coerced to grapheme-cluster strings.
\item Legacy strings tagged to a code page, in a context that requires Unicode,
will be converted to the appropriate Unicode string type.
\end{itemize}

There will be no automatic promotion for octet strings, Unicode to
legacy nor for legacy strings without code page tagging.

\section{Bibliography}

\begin{thebibliography}{9}
\bibitem[Glossary of Unicode Terms]{Glossary}
\href{https://unicode.org/glossary/}%
{\textit{Glossary of Unicode Terms}}

\bibitem[ISO/IEC 8859-1:1997 (E)]{ISO 8859-1}
\href{https://www.open-std.org/JTC1/sc2/wg3/docs/n411.pdf}%
{\textit{Final Text of DIS 8859-1, 8-bit single-byte coded graphic character sets-Part 1: Latin alphabet No.1}}

\bibitem[ISO/IEC 10646:2020]{ISO 10646}
\href{https://www.iso.org/standard/76835.html}%
{\textit{Information technology - Universal coded character set (UCS)}}

\bibitem[RFC 2781]{RFC 2781}
\href{https://datatracker.ietf.org/doc/rfc2781/}%
{\textit{UTF-16, a transformation format of ISO 10646}}

\bibitem[RFC 3629, STD 63]{RFC 3629}
\href{https://datatracker.ietf.org/doc/rfc3629/}%
{\textit{UTF-8, a transformation format of ISO 10646}}

\bibitem[RFC 5137, BCP 137]{RFC 5137}
\href{https://datatracker.ietf.org/doc/rfc5137/}%
{\textit{ASCII Escaping of Unicode Characters}}

\bibitem[RFC 5198]{RFC 5198}
\href{https://datatracker.ietf.org/doc/rfc5198/}%
{\textit{Unicode Format for Network Interchange}}

\bibitem[Unicode]{Unicode}
\href{https://www.unicode.org/standard/standard.html}%
{\textit{The Unicode Standard}}

\bibitem[Unicode Character Database]{UCD}
\href{https://www.unicode.org/ucd/}%
{\textit{Unicode Character Database (UCD)}}

\bibitem[Unicode CLDR Project]{CLDR}
\href{https://cldr.unicode.org/}%
{\textit{Unicode Common Locale Data Repository (CLDR)}}

\bibitem[Unicode{\textregistered} Standard Annex 29]{Annex29}
\href{https://www.unicode.org/reports/tr29/}%
{\textit{Unicode{\textregistered} Standard Annex 29\\ %
Unicode Text Segmentation}}

\end{thebibliography}

\section{Resources}
There are some useful tools available on the WWW:

\begin{resource}
\item[Compart Unicode]%
\href{https://www.compart.com/en/unicode/}%
{Look up, e.g., code point.}
\end{resource}

\end{document}

### 2023-09-02 ARB/standards/work-in-progress/unicode/UnicodeTools/doc/js/search-data.js
Release 0.4 with a full tokenizer

@@ -129,60 +129,6 @@ var searchData = [
             id: "NameOf", type: "Attribute", name: "NameOf", info: "Class Unicode.Property", href: "Properties.cls.html#NameOf"
             },
     {
-        id: "Rexx.Tokenizer.cls", type: "Package", name: "Rexx.Tokenizer.cls", info: null,  href: "Rexx.Tokenizer.cls.html"
-    },
-        {
-        id: "ANSI.Rexx.Tokenizer", type: "Class", name: "ANSI.Rexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ANSI.Rexx.Tokenizer"
-        },
-        {
-        id: "ANSI.Rexx.Unicode.Tokenizer", type: "Class", name: "ANSI.Rexx.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ANSI.Rexx.Unicode.Tokenizer"
-        },
-        {
-        id: "ooRexx.Tokenizer", type: "Class", name: "ooRexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ooRexx.Tokenizer"
-        },
-        {
-        id: "ooRexx.Unicode.Tokenizer", type: "Class", name: "ooRexx.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ooRexx.Unicode.Tokenizer"
-        },
-        {
-        id: "Regina.Tokenizer", type: "Class", name: "Regina.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Regina.Tokenizer"
-        },
-        {
-        id: "Regina.Unicode.Tokenizer", type: "Class", name: "Regina.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Regina.Unicode.Tokenizer"
-        },
-        {
-        id: "Rexx.Tokenizer", type: "Class", name: "Rexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Rexx.Tokenizer"
-        },
-            {
-            id: "init-3", type: "Method", name: "init", info: "Class Rexx.Tokenizer", href: "Rexx.Tokenizer.cls.html#init-3"
-            },
[…]

### 2023-09-02 ARB/standards/work-in-progress/unicode/UnicodeTools/parser/doc/js/search-data.js

@@ -1,0 +1,71 @@
+var searchData = [
+        {
+        id: "Rexx.Tokenizer.cls", type: "Package", name: "Rexx.Tokenizer.cls", info: null,  href: "Rexx.Tokenizer.cls.html"
+    },
+        {
+        id: "ANSI.Rexx.Tokenizer", type: "Class", name: "ANSI.Rexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ANSI.Rexx.Tokenizer"
+        },
+        {
+        id: "ANSI.Rexx.Unicode.Tokenizer", type: "Class", name: "ANSI.Rexx.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ANSI.Rexx.Unicode.Tokenizer"
+        },
+        {
+        id: "ooRexx.Tokenizer", type: "Class", name: "ooRexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ooRexx.Tokenizer"
+        },
+        {
+        id: "ooRexx.Unicode.Tokenizer", type: "Class", name: "ooRexx.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ooRexx.Unicode.Tokenizer"
+        },
+        {
+        id: "Regina.Tokenizer", type: "Class", name: "Regina.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Regina.Tokenizer"
+        },
+        {
+        id: "Regina.Unicode.Tokenizer", type: "Class", name: "Regina.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Regina.Unicode.Tokenizer"
+        },
+        {
+        id: "Rexx.Tokenizer", type: "Class", name: "Rexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Rexx.Tokenizer"
+        },
+            {
+            id: "init", type: "Method", name: "init", info: "Class Rexx.Tokenizer", href: "Rexx.Tokenizer.cls.html#init"
+            },
+            {
[…]

### 2023-09-26 ARB/standards/work-in-progress/unicode/Unicode.Implimentation.tex
Move implementation document to its proper place

@@
+%Compile with latexmk -lualatex
+%
+% Changes
+% Table of contents
+% \underline{default}
+% Split nomenclature section
+% Link to 3.9 Unicode Encoding Forms
+% Codepoints/EGCs option in support of charin()
+% Correct and expand constants example
+% Index
+% stream - seek and update
+% TUTOR
+% Reference ARB GitHub issues
+%
 % \documentclass[draft]{article}
@@
 \usepackage{expl3}
+\usepackage{fontspec}
 \usepackage{geometry}
@@
                               %Consider using minted, but ...
+\usepackage{makeidx}
 \usepackage{upquote}
+\usepackage{cleveref}
 
-\newlist{definition}{description}{2} %% <- pick a larger number if you want to nest these
+\makeindex
+\setmainfont{CMU Serif}
+% texlive-cm-unicode or retrieve cm-unicode with MikTex Console
+
+\newlist{definition}{description}{2} % <- pick a larger number if you want to nest these
 \setlist[definition]{
@@
 
-\newlist{resource}{description}{2} %% <- pick a larger number if you want to nest these
+\newlist{issue}{enumerate}{2} % <- pick a larger number if you want to nest these
+\crefname{issuei}{issue}{issues}
+\Crefname{issuei}{Issue}{Issues}
+\crefname{issueii}{issue}{issues}
+\Crefname{issueii}{Issue}{Issues}
+
+\newlist{resource}{description}{2} % <- pick a larger number if you want to nest these
 \setlist[resource]{
@@
 
+\tableofcontents
+
 \section{Background}
+The Architecture Review board is investigating areas in which Rexx may be
+in need of enhancements, some of which are listed in \cite{Req}.
 Many modern languages support Unicode encoded as
@@
 and that support should not break existing code.
+Elaboration of this issue may be added to \cite{Req} \cref{issue:MB,issue:UTF-8}.
 
 \section{Scope}
-This document only describes issues related to expanded alphabets and multi-octet representations of characters;
-it does not address such issues as bidirectional text.
-It presents suggested solutions for requirements that are presented in other documents.
-Also, most of the details are under active discussion, so everything here is provisional.
+This document only describes issues related to \cref{issue:EA,issue:MB}, expanded alphabet and
+multi-octet representations of characters in Unicode \cref{issue:MB,issue:UTF-8}; it does not
+address such issues as bidirectional text, nor does it address legacy
+DBCS support. It presents suggested solutions for requirements that
+are presented in other documents.  Also, most of the details are under
+active discussion, so everything here is provisional.
+
+This document concentrates on classic Rexx, ANSI Rexx and ooRexx; the
+implementations most relevant are ooRexx and Regina.  However, cREXX
+and NetRexx must be taken into account in order to avoid unnecessary
+differences.
 
@@
 Most of the sections assume that there will be destinct string types
-for extended grapheme clusters, Unicode code points, legacy code pages
-and raw octets. However, there has been discussion of including
+for extended grapheme clusters (\cref{issue:NT,issue:c2u}), Unicode code points (\cref{issue:NT,issue:c2u}),
+legacy code pages (\cref{issue:Leg,issue:mLS,issue:iconv})
+and raw octets (\cref{issue:ROS,issue:NT,issue:mO}). However, there has been discussion of including
 metadata, in which case come of the classes might be merged.
 
+Some of the nomenclature used in this document, including
+the illustrative class names \textbf{BYTES}, \textbf{CODEPOINTS} and \textbf{TEXT}
+is taken from the RXU preprocessor of The Unicode Tools of Rexx (TUTOR).
+It is planned to encorporate more vocabulary from RXU in subsequent revisions.
+
+The word \textbf{text} has its generic meaning unless given in upper case.
+
+An \underline{underscored} term in a list of options is the default.
+Terms separated by the Or symbol (|) are mutually exclussive alternatives.
+Terms in brackets are optional.
+
+The bold character \textbf{\'{ }} in examples refers to
+U+0301, COMBINING ACUTE ACCENT.
+The characters \textbf{\'{E}} and \textbf{\'{e}} in examples refer to the
+composed characters U+C9, LATIN CAPITAL LETTER E WITH ACUTE and
+U+E9, LATIN SMALL LETTER E WITH ACUTE, not to the decomposed clusters
+<U+0045, U+0301> and <U+0065, U+0301>.
+
+Some code samples use ooRexx notation. Those samples are illustrative rather than normative.
+
+References to requirments in \cref{ARBiss} \nameref{ARBiss} are by number.
+
+\section{Statement of problems}
+
+\subsection{Character width}
+
+In Rexx, everything is a string; there are no classes, declarations or types. (\cref{issue:Leg})
+
+While the ANSI standard does not mandate any particular character width for \textbf{Config\_C2B()},
+most if not all implementations use a width of 8,
+and the length of \textbf{c2x(}\textit{foo}\textbf{)}
+is twice the length of \textit{foo}. (\cref{issue:c2x,issue:c2xW})
+
+A large body of existing code operates on binary data from external files,
+or accessed from memory via the \textbf{storage()} BIF,
+under the assumption that, e.g., \textbf{c2x()}, left(), \textbf{right()}, \textbf{substr()}, operate on octets(]cref{issue:ROS,issue:mO}.
+A UTF-8 introducer is treated the same as any other value.
+
+A large body of existing code operates on text under the assumption
+that those facilities operate on characters. (\cref{issue:US})
+
+The two categories overlap.
+
+There is no conflict as long as each character is contained within a single octet.
+However, Unicode has code points beyond U+FF,
+and UTF-8 encoding of non-ASCII Unicode characters will
+require more than a single octet even for code points less than U+100 if they are beyond U+7F.
+
+For example, the Unicode string "Caf\'{e}" has 4 code points, 4
+grapheme clusters and 5 octets in UTF-8 encoding while
+"Cafe\textbf{\'{ }}" has 5 code points, 4 grapheme clusters and 6 octets
+in UTF-8 encoding (\cref{issue:NT}), yet many text processing applications need to treat
+them as equivalent.
+
+\subsection{Case conversion}
+Rexx has built-in case conversion, but it is based on ASCII and
+can't even convert the accented letters found in, e.g., ISO-8859-1
+(Latin-1), ISO-8859-15 (Latin-9), much less all those found in Unicode.
+
+\begin{lstlisting}
+    foo = 'Ren!\'{e}!'
+    parse upper var foo bar
+\end{lstlisting}
+
+sets bar to REN\'{e} rather than to REN\'{E}.
+
+\subsection{\textbf{xrange()}}
+In current implementations the range of characters is extremely small,
+and thus the \textbf{xrange()} only returns short strings. (\cref{issue:xr})
+With Unicode the range expands to U+00 through U+10FFFF.
+That is far too large to allow as either an explict or implicit range in \textbf{xrange()}.
+
+\subsection{Identifiers, labels and procedure names}
+The current definition restricts symbols to ASCII characters, which
+excludes letters used in many languages. If the rules are extended, in
+accordance with \cite{Annex31}.  to include Unicode letters and digits
+beyond ASCII, the rules for equality of symbols must be addressed.
+Are composed and decomposed strings identical? Are base characters
+identical to their compatibility alterrnates? Are subscript and
+superscript digits distinct from their ASCII counterparts?
+
+When a numeric value is used as a label or procedure name, a precise
+definition of the semantics is needed.  While the definition
+"Translate Binary to a coded string." of Config\_B2C in ANSI X3J18-199X
+is rather vague, it is adequate when the character set is restricted
+to an 8-bit character set.  The definition must be replaced by one
+that is unamgiguous when applied to Unicode.  (\ref{issue:IAF})
+There are several
+obvious options.
+\begin{enumerate}
+\item Treat every 32 bits as a Unicode scalar
+\item Treat every 8 bits as part of a UTF-8 sequence
+\item Treat every 8 bits as a character in a legacy code page.
+\end{enumerate}
+Each of these options has disadvantages.
+The same issue exists for Config\_C2B.
+
+When dealing with Unicode, different code sequences may have identical
+rendering due to the existence of both fully composed and combining
+code points and the exixtence of compatibility code points.
+Rexx needs a way to test two Unicode strings for equivalence. (\cref{issue:SNO,issue:NFCD,issue:NFKCD})
+
+\subsection{Status of legacy code pages}
+There has been some discussion claiming a need to continue supporting legacy code pages. (\cref{issue:LCPS,issue:mLS,issue:iconv})
+
+\subsection{Unassigned and invalid code points}
+There has been some discussion of the need to detect unassigned code points. (\cref{issue:UICP,issue:ED})
+
+\subsection{New notation}
+Unicode introduces new notation, which Rexx currently doesn't support, e.g., U+xxxx, <codepoint, ...>.
+
+\section{ARB Issues}
+\label{ARBiss}
+
+This section lists all issues in \cite{Req}, whether or not they relate to Unicode.
+
+\begin{issue}[label=I\arabic*,ref=\fbox{I\arabic*}]
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730857}%
+{Expanded Alphabets}}
+\label{issue:EA}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730866}%
+{Multi-Byte Representations}}
+\label{issue:MB}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730964}%
+{Unicode Strings}}
+\label{issue:US}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730903}%
+{Unassigned \& Invalid Code Points}}
+\label{issue:UICP}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730928}%
+{Error Detection}}
+\label{issue:ED}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730846}%
+{UTF-8 Support}}
+\label{issue:UTF-8}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730880}%
+{Identifiers \& Allowable Formats}}
+\label{issue:IAF}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730910}%
+{New Notation (e.g., U+xxxx)}}
+\label{issue:NN}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730970}%
+{Constants for Unicode Characters}}
+\label{issue:CUC}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730937}%
+{Case Folding}}
+\label{issue:CF}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731043}%
+{Caseless Comparisons}}
+\label{issue:CComp}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730872}%
+{Case Conversion}}
+\label{issue:CConv}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730976}%
+{Coercions}}
+\label{issue:C}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731061}%
+{Strict / Non-strict operators}}
+\label{issue:SNO}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730954}%
+{Legacy Strings}}
+\label{issue:Leg}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730887}%
+{Legacy Code Pages Support}}
+\label{issue:LCPS}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730947}%
+{Raw Octet Strings}}
+\label{issue:ROS}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730934}%
+{I/O Methods}}
+\label{issue:I/O}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35753880}%
+{IO Random Access}}
+\label{issue:RA}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35801621}%
+{New Text Functions}}
+\label{issue:NT}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731024}%
+{NFC \& NFD Functions}}
+\label{issue:NFCD}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731034}%
+{NFKC \& NFKD Functions}}
+\label{issue:NFKCD}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730987}%
+{c2x() Function}}
+\label{issue:c2x}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731048}%
+{Width Parameters for c2x()}}
+\label{issue:c2xW}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730995}%
+{c2u() Function}}
+\label{issue:c2u}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731000}%
+{makeCodePointString Function}}
+\label{issue:mCPS}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731003}%
+{makeGraphemClusterString Function}}
+\label{issue:mGCS}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731012}%
+{makeLegacyString Function}}
+\label{issue:mLS}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731020}%
+{makeOctets Function}}
+\label{issue:mO}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731039}%
+{iconv Function}}
+\label{issue:iconv}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730875}%
+{xrange() Function}}
+\label{issue:xr}
+
+\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731055}%
+{PARSE Instruction}}
+\label{issue:P}
+
+\end{issue}
+
+\section{Solutions}
+
+Most of this section assumes that there will be three distict string
+types, (\cref{issue:mCPS,issue:mGCS,issue:mO}),
+and that some methods will not exist in all three types, or
+will behave differently.
+The type names used here are placeholders, and will be replaced once
+there is consensus on what to call them.
+
+\subsection{\textbf{OPTIONS} statement or directive}
+
+Define an option on the OPTIONS statement, or on a similar directive,
+to specify either raw octets or Unicode characters.
+This breaks programs that operate on both binary data and text.
+
+OPTIONS might also specify a source encoding parameter, overriding any
+code page in an environment variable or file metadata.
+
 \begin{definition}
+\item[SOURCECP=] Encoding of the Rexx source code
+\item[LITERALS=] Type of string constant with no suffix
+\begin{definition}
+\item[BYTES] Raw octets
+\item[CODEPOINTS] Unicode by code point
+\item[LEGACY] Legacy 8-bit characters
+\item[\underline{TEXT}] Unicode by extended grapheme cluster
+\end{definition}
+\end{definition}
+
+\subsection{Classes and types}
+
+Allow Rexx variables to contain three distinct types of data: raw octets, Unicode code points and text,
+(\cref{issue:mCPS,issue:mGCS,issue:mO}),
+
+Define \textbf{storage()} as returning raw octets,
+and provide conversion functions ("casts").
+There is an ongoing discussion as to whether all three are needed.
+
+There should be no implicit conversion between \textbf{.BYTES} and Unicode strings.
+
+There should be implicit conversion between Unicode types, but it need not be reversible, i.e.,
+a conversion from .CODEPOINTS to .TEXT and back need not produce the same code points as the original
+
+If support for legacy code pages is needed, a fourth string type could be defined.
+Strings of this type could include a code page attribute.
+
+For ooRexx a single class with distinguishing attributes could be
+used, or a separate class for each type.
+
+\subsection{Width parameters for c2x()}
+
+Add a width parameter to \textbf{c2x()} for Unicode code point string
+(\cref{issue:c2xW})
+(raw octet strings may require input and output widths for UCS-2 AND UTF-16 data, and the
+utility of \textbf{c2x()} for grapheme clusters needs more
+analysis), and raise conditions with distinct error codes if any code
+point is out of range or if an invalid UTF-8 sequence is detected.  A
+case could be made for using either the bit size or the digit size as
+the width.
+
+Assuming that the width is in in terms of octets
+\begin{lstlisting}[linewidth=35em]
+    foo = 'Ren!\'{e}!'
+    bar = foo~UTF-8
+    /* assumes that width parameter is in octets */
+    say foo~c2x(2) /* Displays 00520065006E00E9  */
+    say bar~c2x    /* Displays 52656EC3A9        */
+\end{lstlisting}
+
+There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
+\begin{lstlisting}[linewidth=35em]
+    foo = .BYTES~new
+    foo = 'Ren!\'{e}!'
+    !\textellipsis!
+    parse var foo ASCII !{\textquotesingle}\'{e}{\textquotesingle}! .
+    say ASCII                     /* Displays Ren */
+\end{lstlisting}
+
+If there is no raw octet string added to the language,
+then \textbf{c2x()} might behave differently depending on whether an explicit width is provided.
+
+\subsection{Constants}
+Other languages allow specifying Unicode characters using either
+the hexadecimal value of the code point or the assigned name of the code point,
+(\cref{issue:CUC})
+e.g., U+E9
+(\cref{issue:NN})
+might be coded as {\textbackslash}u\{E9\} or {\textbackslash}u[LATIN SMALL LETTER E WITH ACUTE].
+The syntax used for such constants should be consisten across Rexx variants,
+including rules for optional spaces between (hex) digits,
+and should take into account the recommendations in \cite{RFC 5137}.
+Implementations should use the machine readable data bases published by
+the Unicode Consortium in order to ease migration to new versions of Unicode.
+
+The form {\textquotesingle}U+digits U+digits ...{\textquotesingle}U is clearer,
+but may break code that abuts the variable U with a string literal.
+The form {\textquotesingle}U+digits U+digits ...{\textquotesingle}X is acceptable.
+There should be a discussion of syntax for named Unicode constants, e.g.,
+{\textquotesingle}[COMBINING ACUTE ACCENT]{\textquotesingle}U is
+equivalent to {\textquotesingle}U+0301{\textquotesingle}U.
+There has been some discussion of using the form
+{\textquotesingle}\textellipsis{\textquotesingle}T and of escape conventions
+used in other languages.
+
+Binary and hexadecimal literals are of type \textbf{.BYTES} and cannot be implicitly coerced to Unicode strings.
+Other string literals are legacy or Unicode and cannot be implictly coerced to \textbf{.BYTES}.
+
+An alternative is to add a new notation using the ASCII \verb|`| as a
+framing character. This has the potential issue that it may be
+difficult to visually distinguish  \verb|`| from  \verb|'|.
+
+Another alternative is to use literals of the forms
+{\textquotesingle}\textellipsis{\textquotesingle}type:codepage,
+{\textquotesingle}\textellipsis{\textquotesingle}U:codepoints and
+{\textquotesingle}\textellipsis{\textquotesingle}U:clusters. However,
+that makes it incompatinle with ooRexx.
+
+The forms {\textquotesingle}\textellipsis{\textquotesingle}\{modifier\} and
+{\textquotesingle}\textellipsis{\textquotesingle}type\{modifier\}
+have no obvious conflict with existing syntax and has no abutment issues.
+
+The following table is intended as a discussion point and should be updated whenever a consencuss is reached.
+
+\begin{tabular}{| l | l | l |}
+\hline
+Example & Type & Semantics \\
+\hline
+{\textquotesingle}Ren\'{e}{\textquotesingle}               & CODEPOINTS & Unicode text by code point             \\
+                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
+\hline
+{\textquotesingle}U+00E9{\textquotesingle}R                & CODEPOINTS & Unicode text by code point             \\
+\hline
+{\textquotesingle}U+00E9{\textquotesingle}CODEPOINTS       & CODEPOINTS & Unicode text by code point             \\
+\hline
+{\textquotesingle}U+00E9{\textquotesingle}TEXT             & TEXT       & Unicode text by grapheme cluster       \\
+\hline
+{\textquotesingle}Ren\'{e}{\textquotesingle}\{CODEPOINTS\} & CODEPOINTS & Unicode text by code point             \\
+\hline
+{\textquotesingle}Ren\'{e}{\textquotesingle}\{TEXT\}       & TEXT       & Unicode text by grapheme cluster       \\
+\hline
+{\textquotesingle}Ren\'{e}{\textquotesingle}\{ISO8859-1\}  & Legacy     & ISO 8859-1 legacy text                 \\
+\hline                                                     & CODEPOINTS & Unicode text by code point             \\
+{\verb|`|}Ren\'{e}{\verb|`|}
+                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
+\hline
+{\verb|`|}Ren\'{e}{\verb|`|}C                              & CODEPOINTS & Unicode text by code point             \\
+\hline
+{\verb|`|}Ren\'{e}{\verb|`|}T                              & TEXT       & Unicode text by grapheme cluster       \\
+\hline
+\end{tabular}
+
+An issue that must be resolved is whether string literals should have
+binary fidelity or visual fidelity to the target code page, i.e.,
+should the quoted source characters be given in the source code page
+or in the target code page.  E.g., should the word "foo" in ISO 8859-1
+and in EBCDIC be coded as
+{\textquotesingle}foo{\textquotesingle}{ISO8859-1} and
+{\textquotesingle}foo{\textquotesingle}{EBCDIC}, requiring that Rexx
+translate the text, or should the second be entered with the EBCDIC
+code points 86, 96, 96, making it harder to read and edit?
+
+\subsection{Error detection}
+The Rexx standard should specify ehat to do when a string being transformed
+contains invalid code points or octet seqiences. (\cref{issue:ED})
+The condition names given below are illustrative and not normative.
+Unicode BIF/BIMs should detect invalid input and signal the following conditions, with unique error codes.
+It is TBD whether some or all of these should be folded into SYNTAX.
+
+\begin{definition}
+\item [INVALIDCODEPOINT]%
+But carefully read 3.2 Conformance Requirements,
+Code Points Unassigned to Abstract Characters, p. 77, in \cite{Unicode}.
+\item[INVALIDUTF]%
+An invalid UTF-8 octet sequence or invalid use of a UTF-16 surrogate.
+\item [NOENCODING]%
+The operation requested requires a code page name.
+\item [NOTEXT]%
+An operation was requested that is not valid for a binary file.
+\item [RANGE]%
+A code point or other numeric entioty excees the permitted range.
+This might occur, e.g., when an application requuires that characters be limited to the BMP.
+\end{definition}
+
+\subsection{I/O}
+
+Stream input/output will need additional support for Unicode.
+\cref{issue:I/O}
+The \textbf{command}  and \textbf{open} methods of the stream classes
+should support a \textbf{CODEPAGE} option; attempting to set a
+codepage for a binary file should raise a \textbf{NOTTEXT} condition
+with a distinct error code.  There should be an option or suboption
+controling whether to use Unicode if an apparent BOM is detected.
+
+The default should probably be UTF-8 or ISO 8859-1 with switching to
+Unicode if a BOM is detected.
+
+Similarly, there should be an option to distinguish between reading
+individual code points and reading EGC; this is needed so that the
+\textbf{charin()} BIM can correctly interpret the character count for Unicode input.
+
+Handling seek for variable-width encodings and EGCs is problematical.
+\cref{issue:RA}
+There are several viable approaches, each with disadvantages.
+
+\begin{itemize}
+\item Use octet count, as at present. \\
+This risks a program doing a seek within a UTF-8 sequence,
+within a UCS-2/UCS-4 character,
+within a UTF-16 surrogate pair,
+or, for EGC access, withing an EGC.
+Strategies to ameliorate this include
+\begin{definition}
+\item [UTF-8]
+Read the octet at the specified offset and verify that either the high
+bit is 0 or the two high order bits are 1.
+\item [UCS-2]
+Verify that the offset is a multiple of 2.
+\item [UTF-16]
+Verify that the offset is a multiple of 2.  Read two octets at the
+specified offset and verify that is not a low surrogate (U+DC00-U+DFFF),
+\item [UCS-4]
+Verify that the offset is a multiple of 2.
+Read four bytes and verify that they do not exceed U+10FFFF.
+\end{definition}
+\item Use character or EGC count \\
+This is expensive
+\item Define a cursor type. \\
+This violates "Everything is a string" in classic Rexx.
+\end{itemize}
+
+Handling update in the middle of a file with variable-width encodings
+or EGC access has similar issues, Absent an acceptable way to deal
+with it, prohibit update in the middle of a file for anything but
+binary and legacy characters.
+
+A discussion is needed on whether and when to create or discard byte order marks.
+
+\subsection{Case folding}
+
+Case folding of symbols and caseless comparisons will use the \cite{UCD}.
+\cref{issue:CF,issue:CComp,issue:CConv})
+The rules for case folding should take into account mathematical usage.
+\begin{itemize}
+\item U+1D400 through U+1D7FF are semantically distinct from other letters.
+\item Superscripts and subscripts have semantic significance.
+\end{itemize}
+
+\subsection{Raw octet strings}
+Many exist Rexx programs deal with binary data,
+and Rexx has conversion functions to deal with those data,
+e.g., c2x, x2c.
+(\cref{issue:ROS})
+The new version of Rexx will need methods to continue dealing with binary data.
+The methods should include those of the ooRexx .string class, except that
+\begin{itemize}
+\item The unit of operation is the uninterpreted octet.
+\item The \textbf{makeString} method requires an encoding parameter.
+\item Parameters are raw octet strings.
+\item There are no caseless methods.
+\end{itemize}
+
+Additional conversion functions will be needed (\cref{issue:NT,issue:mLS,issue:mO}).
+
+\subsection{Legacy strings}
+
+If there is a .Legacy string type then the methods should have the
+same semantics as the existing methods for the .string class.
+If the new standard has a code page attribute then there should be an
+access method for it and the init or new method should allow
+specifying the code page.
+
+The following Additional functions and methods should be defined:
+\begin{definition}
+\item [iconv] %
+Convert from one code page to another and return a legacy string. (\cref{issue:iconv})
+\item [makeCodePageString] %
+Return a Unicode string in which individual code points can be accessed. (\cref{issue:mCPS})
+\item [makeGraphemClusterString] %
+Return a Unicode string in which only complete grapheme clusters can be accessed. (\cref{issue:mGCS})
+\end{definition}
+
+\subsection{Unicode strings}
+
+There should be subtypes depending on whether the
+unit of operation is the Unicode code point or the grapheme cluster. (\cref{issue:US})
+The methods should be those of the ooRexx .string class except:
+\begin{itemize}
+\item there should be a BIF for each in order to allow use in classic Rexx.
+\item Discuss how and whether to introduce equivalent operators.
+\item discuss whether \textbf{=} should ignore all leading and trailing whitespace
+or only leading and trailing blanks. Discuss need for other equality tests.
+\item \textbf{==} tests for absolute equality for CODEPOINTS but
+EGC equivalence for TEXT.
+Discuss which of the Unicode equality tests to use and,
+if more than one, which is primary.
+\item
+The caseless comparisons will use the \cite{UCD}.
+(\cref{issue:CComp})
+There will be variants to preserve or remove accents\footnote{Is Enye (\~n) considered an accented letter?}.
+\item There should be no \textbf{bit\textellipsis} methods.
+\item Add an optional third parameter to the \textbf{upper} and \textbf{lower} methods
+to control whether to translate mon-ASCII characters and whether to strip accents.
+\item
+The following Additional functions and methods should be defined:
+\begin{definition}
+\item[c2u] Return code points in Unicode notation. Possible options include
+\begin{itemize}
+\item U+xxxx or bare hexadecimal
+\item frame in <> and separate with commas
+\item indicate EGC boundaries
+\item indicate properties
+\end{itemize}
+\item [makeCodePageString]
+Return a Unicode string in which individual code points can be accessed (\cref{issue:mCPS})
+\item [makeGraphemClusterString]
+Return a Unicode string in which only complete grapheme clusters can be accessed (\cref{issue:mGCS})
+\item [makeLegacyString]
+Return a legacy string for the specified code page in which individual octets can be accessed (\cref{issue:LCPS,issue:mLS})
+\item [makeOctets]
+Return a raw octet string using a specified encoding.\footnote{default
+to UTF-8 or raise NOENCODING if no encoding specified?} (\cref{issue:mO})
+\item [NFC] Return a string normalized with Normalization Form C (\cref{issue:NFCD})
+\item [NFD] Return a string normalized with Normalization Form D (\cref{issue:NFCD})
+\end{definition}
+\end{itemize}
+
+There should be a description of how to handle conversion from Unicode code
+points and grapheme clusters that do not exist in the target code page.
+(\cref{issue:ED,issue:Leg,issue:iconv})
+\begin{itemize}
+\item Replace with U+1A (SUB) or other specified character
+\item Raise SYNTAX with a unique error code
+\item Raise a new condition name with a unique error code
+\end{itemize}
+
+\subsection{Functions and methods}
+The following table summarizes some of the methods that differ among string types.
+With the exceptions of \textbf{{[}{]}}, \textbf{makeArray} and \textbf{makeString},
+there is an operator or polymorphic BIF for each BIM listed.
+
+\begin{tabular}{| l | c | c | c | c |}
+\hline
+Method               & \textbf{.Legacy} & \textbf{.BYTES}  & \textbf{.CODEPOINTS}  & \textbf{TEXT} \\
+\hline
+\textbf{{[}{]}}      & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{\&}          & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
+\textbf{|}           & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
+\textbf{\&\&}        & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
+\textbf{=}           & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{==}          & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{{\neg}=}     & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{><}          & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{<>}          & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{<}           & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{<=}          & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{>}           & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{{\neg}>}     & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{abbrev}      & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{bitand}      & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
+\textbf{bitor}       & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
+\textbf{bitxor}      & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
+\textbf{c2b}         & 8 bits           & default 8 bits   & Explicit width & \textbf{n/a}  \\
+\textbf{c2x}         & 8 bits           & default 8 bits   & Explicit width & \textbf{n/a}  \\
+\textbf{center}      & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
+\textbf{change}      & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{find}        & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{index}       & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{left}        & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{makeArray}   & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{makeString}  & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{pos}         & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{right}       & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{strip}       & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{substr}      & by octet         & by octet         & by code point  & by cluster    \\
+\textbf{verify}      & by octet         & by octet         & by code point  & by cluster    \\
+\hline
+\end{tabular}
+\\*
+\\*
+The backslash ("\textbackslash") may be used in place of the Logical Not ("\neg").
+
+In addition, the makeString and makeString methods should allow caseless option
+parameters to control the class and attributes of the returned array
+elements or string, including:
+
+\begin{definition}
+\item [Clusters] Unicode text with extended grapeme clusters as the abstract units.
+\item [Codepoints] Unicode text with Unicode scalars as the abstract units.
+\item [cp=legacy code page] Legacy or octet string with specified encoding.
+\item [cp=UTF-8]
+Legacy or octet string with UTF-8 encoding;
+functions like center and left will give unexpected results.
+\item [Legacy] Legacy string with default encoding unless cp= is also specified.
+\item [raw] Raw octet string; cp= must be specified.
+\end{definition}
+
+\subsection{Coercions}
+In addition to explicit casts via BIF/BIM, the following promotions
+from code-point strings to grapheme-cluster strings will be automatic
+\cref{issue:C})
+
+\begin{itemize}
+\item Concatenation of code-point strings with grapheme-cluster strings.
+will be grapheme-cluster strings.
+\item Code point search arguments for grapheme-cluster strings.
+will be coerced to grapheme-cluster strings.
+\item Legacy strings tagged to a code page, in a context that requires Unicode,
+will be converted to the appropriate Unicode string type.
+\end{itemize}
+
+There will be no automatic promotion for octet strings, Unicode to
+legacy nor for legacy strings without code page tagging.
+
+\section{Bibliography}
+
+\subsection{GitHub ARB documents}
+\renewcommand{\refname}{ARB}
+
+\begin{thebibliography}{20}
+\bibitem[New Types]{New Types}
+\href{https://github.com/RexxLA/rexx-repository/edit/master/ARB/standards/work-in-progress/unicode/UnicodeTools/doc/string-types.md}%
+{\textit{New types of strings}}
+
+\bibitem[Requirements]{Req}
+\href{https://github.com/users/RexxLA/projects/2/views/6}%
+{\textit{Unicode Standard Development - requirements}}
+
+\bibitem[Requirements Status]{ReqStat}
+\href{https://github.com/users/RexxLA/projects/2/views/1}%
+{\textit{Unicode Standard Development - requirements status}}
+
+\bibitem[RXU]{RXU}
+\href{https://github.com/RexxLA/rexx-repository/blob/master/ARB/standards/work-in-progress/unicode/UnicodeTools/rxu.md}%
+{\textit{The RXU Rexx Preprocessor for Unicode}}
+
+\end{thebibliography}
+
+\subsection{Official documents}
+\renewcommand{\refname}{References}
+
+\begin{thebibliography}{20}
+\bibitem[Glossary of Unicode Terms]{Glossary}
+\href{https://unicode.org/glossary/}%
+{\textit{Glossary of Unicode Terms}}
+
+\bibitem[ISO/IEC 8859-1:1997 (E)]{ISO 8859-1}
+\href{https://www.open-std.org/JTC1/sc2/wg3/docs/n411.pdf}%
+{\textit{Final Text of DIS 8859-1, 8-bit single-byte coded graphic character sets-Part 1: Latin alphabet No.1}}
+
+\bibitem[ISO/IEC 10646:2020]{ISO 10646}
+\href{https://www.iso.org/standard/76835.html}%
+{\textit{Information technology - Universal coded character set (UCS)}}
+
+\bibitem[RFC 2781]{RFC 2781}
+\href{https://datatracker.ietf.org/doc/rfc2781/}%
+{\textit{UTF-16, a transformation format of ISO 10646}}
+
+\bibitem[RFC 3629, STD 63]{RFC 3629}
+\href{https://datatracker.ietf.org/doc/rfc3629/}%
+{\textit{UTF-8, a transformation format of ISO 10646}}
+
+\bibitem[RFC 5137, BCP 137]{RFC 5137}
+\href{https://datatracker.ietf.org/doc/rfc5137/}%
+{\textit{ASCII Escaping of Unicode Characters}}
+
+\bibitem[RFC 5198]{RFC 5198}
+\href{https://datatracker.ietf.org/doc/rfc5198/}%
+{\textit{Unicode Format for Network Interchange}}
+
+\bibitem[Unicode]{Unicode}
+\href{https://www.unicode.org/standard/standard.html}%
+{\textit{The Unicode Standard}}
+
+\bibitem[Unicode]{Unicode3.9}
+\href{https://www.unicode.org/versions/Unicode15.0.0/ch03.pdf#page=50}%
+{\textit{3.9 Unicode Encoding Forms}}
+
+\bibitem[Unicode Character Database]{UCD}
+\href{https://www.unicode.org/ucd/}%
+{\textit{Unicode Character Database (UCD)}}
+
+\bibitem[Unicode CLDR Project]{CLDR}
+\href{https://cldr.unicode.org/}%
+{\textit{Unicode Common Locale Data Repository (CLDR)}}
+
+\bibitem[Unicode{\textregistered} Standard Annex 29]{Annex29}
+\href{https://www.unicode.org/reports/tr29/}%
+{\textit{Unicode{\textregistered} Standard Annex 29\\ %
+Unicode Text Segmentation}}
+
+\bibitem[Unicode{\textregistered} Standard Annex 31]{Annex31}
+\href{https://www.unicode.org/reports/tr31/}%
+{\textit{Unicode{\textregistered} Standard Annex 31\\ %
+Unicode Identifier and Pattern Syntax}}
+
+\bibitem[Unicode{\textregistered} Standard Annex 42]{Annex42}
+\href{https://www.unicode.org/reports/tr42/tr42-34.html}%
+{\textit{Unicode{\textregistered} Standard Annex 42\\ %
+Unicode Character Database in XML}}
+
+\bibitem[Unicode{\textregistered} Standard Annex 44]{Annex44}
+\href{https://www.unicode.org/reports/tr44/tr44-32.html}%
+{\textit{Unicode{\textregistered} Standard Annex 44\\ %
+Unicode Character Database}}
+
+\end{thebibliography}
+
+\section{Glossary}
+
+The definitions given in \cite{Glossary} and \cite{Unicode} take
+precedence over those given here. Quoted text is taken from those sources, or from IETF documents.
+Except for definitions taken from official IETF and Unicode documents,
+the nomenclature used here is illustrative rather than normative;
+language design teams will formally define, e.g., method names, encoding of parameters.
+
+\begin{definition}
+\item [.BYTES] The string class for uninterprted octet strings.
+\item [.CODEPOINTS] The string class for strings of Unicode code points.
 \item [.Legacy] The string class for strings in legacy code pages.
-\item [.octets] The string class for uninterprted octet strings.
-\item [.Ucp] The string class for strings of Unicode code points.
-\item [.Ugc] The string class for strings of Unicod extended grapheme clusters.
-\item [\textbf{\'{ }}] U+0301, COMBINING ACUTE ACCENT
-\\*
-Set in bold to distinguish it from U+27, APOSTROPHE.
+May be merged with .BYTES.
+\item [.TEXT] The string class for strings of Unicod extended grapheme clusters.
+\item[ARB] RexxLA Architecture Review Board
 \item [Bidi] "Abbreviation of bidirectional, in reference to mixed left-to-right and right-to-left text."
@@
 reserved."
-\item [BMP] Basic Multilingual Plane
+\item [BMP] Basic Multilingual Plane\index{BMP!definition}index{BMP!Basic Multilingual Plane!definition}
 \\*
@@
 Optional as the first character for UTF-8.
-\item [\'{E}] U+C9, LATIN CAPITAL LETTER E WITH ACUTE
-\\*
-Set in normal weight: U+45 U+0301 will be shown as "E\textbf{\'{ }}".
-\item [\'{e}] U+E9, LATIN SMALL LETTER E WITH ACUTE
-\\*
-Set in normal weight: U+65 U+0301 will be shown as "e\textbf{\'{ }}".
 \item [encoded character] The smallest constituant of a Unicode string.
@@
 \item [UCS-2] A 16 bit subset of Unicode, containing only the BMP.
-\item [The Unicode Consortium] A non-profit corporation devoted to
+\item [Unicode Consortium] A non-profit corporation devoted to
 developing, maintaining, and promoting software internationalization standards and data.
-\item [UTF-8] UCS Transformation Format 8 (\cite{RFC 3629}).
+\item [UTF-8] UCS Transformation Format 8
+(\cite{Unicode3.9} and \cite{RFC 3629}).
 \\*
@@
 \end{tabular}
-\item [UTF-16] UCS Transformation Format 16 (\cite{RFC 2781}).
+\item [UTF-16] UCS Transformation Format 16
+(\cite{Unicode3.9} and \cite{RFC 2781}).
 \end{definition}
-
-Some code samples use ooRexx notation. Those samples are illustrative rather than normative.
-
-\section{Statement of problem}
-
-\subsection{Primary Problem}
-
-In Rexx, everything is a string; there are no classes, declarations or types.
-
-While the ANSI standard does not mandate any particular character width for \textbf{Config\_C2B()},
-most if not all implementations use a width of 8,
-and the length of \textbf{c2x(}\textit{foo}\textbf{)}
-is twice the length of \textit{foo}.
-
-A large body of existing code operates on binary data from external files,
-or accessed from memory via the \textbf{storage()} BIF,
-under the assumption that, e.g., \textbf{c2x()}, left(), \textbf{right()}, \textbf{substr()}, operate on octets.
-A UTF-8 introducer is treated the same as any other value.
-
-A large body of existing code operates on text under the assumption that those facilities operate on characters.
-
-The two categories overlap.
-
-There is no conflict as long as each character is contained within a single octet.
-However, Unicode has code points beyond U+FF,
-and UTF-8 encoding of non-ASCII Unicode characters will
-require more than a single octet even for code points less than U+100 if they are beyond U+7F.
-
-For example, the Unicode string 'Caf\'{e}' has 4 code points, 4
-grapheme clusters and 5 octets in UTF-8 encoding while
-'Cafe\textbf{\'{ }}' has 5 code points, 4 grapheme clusters and 6 octets
-in UTF-8 encoding, yet many text processing applications need to treat
-them as equivalent.
-
-\subsection{Secondary Problems}
-Rexx has built-in case conversion, but it is based on ASCII and
-can't even convert the accented letters found in, e.g., ISO-8859-1
-(Latin-1), ISO-8859-15 (Latin-9), much less all those found in Unicode.
-
-\begin{lstlisting}
-    foo = 'Ren!\'{e}!'
-    parse upper var foo bar
-\end{lstlisting}
-
-sets bar to REN\'{e} rather than to REN\'{E}.
-
-In current implementations the range of characters is extremely small,
-and thus the \textbf{xrange()} only returns short strings.
-With Unicode the range expands to U+00 through U+10FFFF.
-That is far too large to allow as either an explict or implicit range in \textbf{xrange()}.
-
-The current definition restricts symbols to ASCII characters,
-which excludes letters used in many languages. If the rules are extended to include
-Unicode letters and digits beyond ASCII, the rules for equality of symbols must be addressed.
-Are composed and decomposed strings identical?
-Are base characters identical to their compatibility alterrnates?
-Are subscript and superscript digits distinct from their ASCII
-counterparts?
-
-When dealing with Unicode, different code sequences may have identical
-rendering due to the existence of both fully composed and combining
-code points and the exixtence of compatibility code points.
-Rexx needs a way to test two Unicode strings for equivalence.
-
-There has been some discussion claiming a need to continue supporting legacy code pages.
-
-There has been some discussion of the need to detect unassigned code points.
-
-\section{Solutions}
-
-Most of this section assumes that there will be three distict string
-types, and that some methods will not exist in all three types, or
-will behave differently.
-The type names used here are placeholders, and will be replaced once
-there is consensus on what to call them.
-
-\subsection{\textbf{OPTIONS} statement or directive}
-
-Define an option on the OPTIONS statement, or on a similar directive,
-to specify either raw octets or Unicode characters.
-This breaks programs that operate on both binary data and text.
-
-OPTIONS mighy also specify a source encoding parameter, overriding any code page in an environment variable or file metadata.
-
-\subsection{Types}
-
-Allow Rexx variables to contain three distinct types of data: raw octets, Unicode code points and text,
-Define \textbf{storage()} as returning raw octets,
-and provide conversion functions ("casts").
-There is an ongoing discussion as to whether all three are needed.
-
-There should be no implicit conversion between \textbf{.octets} and Unicode strings.
-
-There should be implicit conversion between Unicode types, but it need not be reversible, i.e.,
-a conversion from .Ucp to .Ugc and back need not produce the same code points as the original
-
-If support for legacy code pages is needed, a fourth string type could be defined.
-Strings of this type could include a code page attribute.
-
-\subsection{Width parameter}
-
-Add a width parameter to \textbf{c2x()} for Unicode code point string
-(raw octet strings may require input and output widths for UCS-2 AND UTF-16 data, and the
-utility of \textbf{c2x()} for grapheme clusters needs more
-analysis), and raise conditions with distinct error codes if any code
-point is out of range or if an invalid UTF-8 sequence is detected.  A
-case could be made for using either the bit size or the digit size as
-the width.
-
-Assuming that the width is in in terms of octets
-\begin{lstlisting}[linewidth=35em]
-    foo = 'Ren!\'{e}!'
-    bar = foo~UTF-8
-    /* assumes that width parameter is in octets */
-    say foo~c2x(2) /* Displays 00520065006E00E9  */
-    say bar~c2x    /* Displays 52656EC3A9        */
-\end{lstlisting}
-
-There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
-\begin{lstlisting}[linewidth=35em]
-    foo = .octets~new
-    foo = 'Ren!\'{e}!'
-    !\textellipsis!
-    parse var foo ASCII !{\textquotesingle}\'{e}{\textquotesingle}! .
-    say ASCII                     /* Displays Ren */
-\end{lstlisting}
-
-If there is no raw octet string added to the language,
-then \textbf{c2x()} might behave differently depending on whether an explicit width is provided.
-
-\subsection{Constants}
-Other languages allow specifying Unicode characters using either
-the hexadecimal value of the code point or the assigned name of the code point,
-e.g., U+E9 might be coded as {\textbackslash}u\{E9\} or {\textbackslash}u[LATIN SMALL LETTER E WITH ACUTE].
-The syntax used for such constants should be consisten across Rexx variants,
-including rules for optional spaces between (hex) digits,
-and should take into account the recommendations in \cite{RFC 5137}.
-Implementations should use the machine readable data bases published by
-the Unicode Consortium in order to ease migration to new versions of Unicode.
-
-The form {\textquotesingle}U+digits U+digits ...{\textquotesingle}U is clearer,
-but may break code that abuts the variable U with a string literal.
-The form {\textquotesingle}U+digits U+digits ...{\textquotesingle}X is acceptable.
-There should be a discussion of syntax for named Unicode constants, e.g.,
-{\textquotesingle}[COMBINING ACUTE ACCENT]{\textquotesingle}U is
-equivalent to {\textquotesingle}U+0301{\textquotesingle}U.
-There has been some discussion of using the form
-{\textquotesingle}\textellipsis{\textquotesingle}T and of escape conventions
-used in other languages.
-
-Binary and hexadecimal literals are of type \textbf{.octets} and cannot be implicitly coerced to Unicode strings.
-Other string literals are legacy or Unicode and cannot be implictly coerced to \textbf{.octets}.
-
-An alternative is to add a new notation using the ASCII \verb|`| as a
-framing character. This has the potential issue that it may be
-difficult to visually distinguish  \verb|`| from  \verb|'|.
-
-Another alternative is to use literals of the forms
-{\textquotesingle}\textellipsis{\textquotesingle}type:codepage,
-{\textquotesingle}\textellipsis{\textquotesingle}U:codepoints and
-{\textquotesingle}\textellipsis{\textquotesingle}U:clusters. However, that makes it incompatinle with ooRexx.
-
-The forms {\textquotesingle}\textellipsis{\textquotesingle}\{modifier\} and
-{\textquotesingle}\textellipsis{\textquotesingle}type\{modifier\}
-have no obvious conflict with existing syntax and has no abutment issues.
-
-The following table is intended as a discussion point and should be updated whenever a consencuss is reached.
-
-\begin{tabular}{| l | l | l |}
-\hline
-Example & Type & Semantics \\
-\hline
-{\textquotesingle}Ren\'{e}{\textquotesingle}              & Ucp & Unicode text by code point \\
-                                                          & Ugc & Unicode text by grapheme cluster (TBD) \\
-\hline
-{\textquotesingle}U+00E9{\textquotesingle}R               & Ucp & Unicode text by code point \\
-\hline
-{\textquotesingle}U+00E9{\textquotesingle}Ucp             & Ucp & Unicode text by code point \\
-\hline
-{\textquotesingle}U+00E9{\textquotesingle}Ugc             & Ucp & Unicode text by grapheme cluster  \\
-\hline
-{\textquotesingle}Ren\'{e}{\textquotesingle}\{ucp\}       & Ucp & Unicode text by code point \\
-\hline
-{\textquotesingle}Ren\'{e}{\textquotesingle}\{ugc\}       & Ucp & Unicode text by grapheme cluster  \\
-\hline
-{\textquotesingle}Ren\'{e}{\textquotesingle}\{ISO8859-1\} & Legacy & ISO 8859-1 legacy text \\
-\hline
-\end{tabular}
-
-\subsection{Error detection}
-The condition names given below are illustrative and not normative.
-Unicode BIF/BIMs should detect invalid input and signal the following conditions, with unique error codes.
-
-\begin{definition}
-\item [INVALIDCODEPOINT]%
-But carefully read 3.2 Conformance Requirements,
-Code Points Unassigned to Abstract Characters, p. 77, in \cite{Unicode}.
-\item[INVALIDUTF]%
-An invalid UTF-8 octet sequence or invalid use of a UTF-16 surrogate.
-\item [NOENCODING]%
-The operation requested requires a code page name.
-\item [NOTEXT]%
-An operation was requested that is not valid for a binary file.
-\item [RANGE]%
-A code point or other numeric entioty excees the permitted range.
-This might occur, e.g., when an application requuires that characters be limited to the BMP.
-\end{definition}
-
-\subsection{I/O}
-
-The \textbf{command} method of the stream classes should support a \textbf{CODEPAGE} option;
-attempting to set a codepage for a binary file should raise a \textbf{NOTTEXT} condition with a distinct error code.
-There should be an option or suboption controling whether to use Unicode if an apparent BOM is detected.
-
-The default should probably be UTF-8 or ISO 8859-1 with switching to Unicode if a BOM is detected.
-
-A discussion is needed on whether and when to create or discard byte order marks.
-
-\subsection{Case folding}
-
-The rules for case folding should take into account mathematical usage.
-\begin{itemize}
-\item U+1D400 through U+1D7FF are semantically distinct from other letters.
-\item Superscripts and subscripts have semantic significance.
-\end{itemize}
-
-\subsection{Raw octet strings}
-The methods should include those of the ooRexx .string class, except that
-\begin{itemize}
-\item The unit of operation is the uninterpreted octet.
-\item The \textbf{makeString} method requires an encoding parameter.
-\item Parameters are raw octet strings.
-\item There are no caseless methods.
-\end{itemize}
-
-\subsection{Legacy strings}
-
-If there is a .Legacy string type then the methods should have the
-same semantics as the existing methods for the .string class.
-If the new standard has a code page attribute then there should be an
-access method for it and the init or new method should allow
-specifying the code page.
-
-The following Additional methods should be defined:
-\begin{definition}
-\item [iconv] %
-Convert from one code page to another and return a legacy string.
-\item [makeCodePageString] %
-Return a Unicode string in which individual code points can be accessed
-\item [makeGraphemClusterString] %
-Return a Unicode string in which only complete grapheme clusters can be accessed
-\end{definition}
-
-\subsection{Unicode strings}
-
-There should be subtypes depending on whether the
-unit of operation is the Unicode code point or the grapheme cluster.
-The methods should be those of the ooRexx .string class except:
-\begin{itemize}
-\item
-The caseless comparisons will use the \cite{UCD}.
-There will be variants to preserve or remove accents\footnote{Is Enye (\~n) considered an accented letter?}.
-\item There should be no \textbf{bit\textellipsis} methods.
-\item Add an optional third parameter to the \textbf{upper} and \textbf{lower} methods
-to control whether to translate mon-ASCII characters and whether to strip accents.
-\item
-The following Additional methods should be defined:
-\begin{definition}
-\item [makeCodePageString]
-Return a Unicode string in which individual code points can be accessed
-\item [makeGraphemClusterString]
-Return a Unicode string in which only complete grapheme clusters can be accessed
-\item [makeLegacyString]
-Return a legacy string for the specified code page in which individual octets can be accessed
-\item [makeOctets]
-Return a raw octet string using a specified encoding.\footnote{default
-to UTF-8 or raise NOENCODING if no encoding specified?}
-\item [NFC] Return a string normalized with Normalization Form C
-\item [NFD] Return a string normalized with Normalization Form D
-\end{definition}
-\end{itemize}
-
-There should be a description of how to handle conversion from Unicode code
-points and grapheme clusters that do not exist in the target code page.
-\begin{itemize}
-\item Replace with U+1A (SUB) or other specified character
-\item Raise SYNTAX with a unique error code
-\item Raise a new condition name with a unique error code
-\end{itemize}
-
-\subsection{Methods}
-The following table summarizes some of the methods that differ among string types.
-
-\begin{tabular}{| l | c | c | c | c |}
-\hline
-Method               & \textbf{.Legacy} & \textbf{.octets} & \textbf{.Ucp}  & \textbf{.Ugc} \\
-\hline
-\textbf{{[}{]}}      & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{\&}          & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
-\textbf{|}           & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
-\textbf{\&\&}        & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
-\textbf{=}           & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{{\neg}=}     & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{><}          & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{<>}          & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{<}           & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{<=}          & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{>}           & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{{\neg}>}     & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{abbrev}      & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{bitand}      & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
-\textbf{bitor}       & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
-\textbf{bitxor}      & \textbf{n/a}     & by octet         & \textbf{n/a}   & \textbf{n/a}  \\
-\textbf{c2b}         & 8 bits           & default 8 bits   & Explicit width & \textbf{n/a}  \\
-\textbf{c2x}         & 8 bits           & default 8 bits   & Explicit width & \textbf{n/a}  \\
-\textbf{center}      & by octet         & \textbf{n/a}     & by code point  & by cluster    \\
-\textbf{change}      & \textbf{.Legacy} & \textbf{.octets} & by code point  & by cluster    \\
-\textbf{find}        & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{index}       & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{left}        & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{makeArray}   & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{makeString}  & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{pos}         & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{right}       & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{strip}       & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{substr}      & by octet         & by octet         & by code point  & by cluster    \\
-\textbf{verify}      & by octet         & by octet         & by code point  & by cluster    \\
-\hline
-\end{tabular}
-\\*
-\\*
-The backslash (``\textbackslash'') may be used in place of the Logical Not (``\neg'').
-
-In addition, the makeString and makeString methods should allow caseless option
-parameters to control the class and attributes of the returned array
-elements or string, including:
-
-\begin{definition}
-\item [Clusters] Unicode text with extended grapeme clusters as the abstract units.
-\item [Codepoints] Unicode text with Unicode scalars as the abstract units.
-\item [cp=legacy code page] Legacy or octet string with specified encoding.
-\item [cp=UTF-8]
-Legacy or octet string with UTF-8 encoding;
-functions like center and left will give unexpected results.
-\item [Legacy] Legacy string with default encoding unless cp= is also specified.
-\item [raw] Raw octet string; cp= must be specified.
-\end{definition}
-
-\subsection{Coercions}
-In addition to explicit casts via BIF/BIM, the following promotions
-from code-point strings to grapheme-cluster strings will be automatic
-
-\begin{itemize}
-\item Concatenation of code-point strings with grapheme-cluster strings.
-will be grapheme-cluster strings.
-\item Code point search arguments for grapheme-cluster strings.
-will be coerced to grapheme-cluster strings.
-\item Legacy strings tagged to a code page, in a context that requires Unicode,
-will be converted to the appropriate Unicode string type.
-\end{itemize}
-
-There will be no automatic promotion for octet strings, Unicode to
-legacy nor for legacy strings without code page tagging.
-
-\section{Bibliography}
-
-\begin{thebibliography}{9}
-\bibitem[Glossary of Unicode Terms]{Glossary}
-\href{https://unicode.org/glossary/}%
-{\textit{Glossary of Unicode Terms}}
-
-\bibitem[ISO/IEC 8859-1:1997 (E)]{ISO 8859-1}
-\href{https://www.open-std.org/JTC1/sc2/wg3/docs/n411.pdf}%
-{\textit{Final Text of DIS 8859-1, 8-bit single-byte coded graphic character sets-Part 1: Latin alphabet No.1}}
-
-\bibitem[ISO/IEC 10646:2020]{ISO 10646}
-\href{https://www.iso.org/standard/76835.html}%
-{\textit{Information technology - Universal coded character set (UCS)}}
-
-\bibitem[RFC 2781]{RFC 2781}
-\href{https://datatracker.ietf.org/doc/rfc2781/}%
-{\textit{UTF-16, a transformation format of ISO 10646}}
-
-\bibitem[RFC 3629, STD 63]{RFC 3629}
-\href{https://datatracker.ietf.org/doc/rfc3629/}%
-{\textit{UTF-8, a transformation format of ISO 10646}}
-
-\bibitem[RFC 5137, BCP 137]{RFC 5137}
-\href{https://datatracker.ietf.org/doc/rfc5137/}%
-{\textit{ASCII Escaping of Unicode Characters}}
-
-\bibitem[RFC 5198]{RFC 5198}
-\href{https://datatracker.ietf.org/doc/rfc5198/}%
-{\textit{Unicode Format for Network Interchange}}
-
-\bibitem[Unicode]{Unicode}
-\href{https://www.unicode.org/standard/standard.html}%
-{\textit{The Unicode Standard}}
-
-\bibitem[Unicode Character Database]{UCD}
-\href{https://www.unicode.org/ucd/}%
-{\textit{Unicode Character Database (UCD)}}
-
-\bibitem[Unicode CLDR Project]{CLDR}
-\href{https://cldr.unicode.org/}%
-{\textit{Unicode Common Locale Data Repository (CLDR)}}
-
-\bibitem[Unicode{\textregistered} Standard Annex 29]{Annex29}
-\href{https://www.unicode.org/reports/tr29/}%
-{\textit{Unicode{\textregistered} Standard Annex 29\\ %
-Unicode Text Segmentation}}
-
-\end{thebibliography}
 
@@
 \begin{resource}
+\item[CLDR]%
+\href{https://cldr.unicode.org/index/downloads}%
+{CLDR Releases/Downloads}
+
 \item[Compart Unicode]%
@@
 {Look up, e.g., code point.}
+
+\item[TUTOR]%
+\href{https://github.com/RexxLA/rexx-repository/tree/master/ARB/standards/work-in-progress/unicode/UnicodeTools}%
+{The Unicode Tools Of Rexx}
+
+\item[UCD]%
+\href{https://www.unicode.org/Public/UCD/latest/}%
+{Unicode{\textregistered} Character Database (UCD)}
+
+\item[UCD index]%
+\href{https://www.unicode.org/charts/charindex.html}%
+{Unicode{\textregistered} Character Name Index}
 \end{resource}
 
+\printindex
+
 \end{document}

### 2023-10-02 ARB/standards/work-in-progress/unicode/UnicodeTools/doc/js/search-data.js
Moce docs fro stream.cls to md format

@@ -68,31 +68,4 @@ var searchData = [
             {
             id: "NameOf", type: "Attribute", name: "NameOf", info: "Class Unicode.Property", href: "Properties.cls.html#NameOf"
             },
-    {
-        id: "Stream.cls", type: "Package", name: "Stream.cls", info: null,  href: "Stream.cls.html"
-    },
-        {
-        id: "%21CharIn", type: "Routine", name: "!CharIn", info: "Stream.cls",  href: "Stream.cls.html#%21CharIn"
-        },
-        {
-        id: "%21CharOut", type: "Routine", name: "!CharOut", info: "Stream.cls",  href: "Stream.cls.html#%21CharOut"
-        },
-        {
-        id: "%21Chars", type: "Routine", name: "!Chars", info: "Stream.cls",  href: "Stream.cls.html#%21Chars"
-        },
-        {
-        id: "%21Linein", type: "Routine", name: "!Linein", info: "Stream.cls",  href: "Stream.cls.html#%21Linein"
-        },
-        {
-        id: "%21LineOut", type: "Routine", name: "!LineOut", info: "Stream.cls",  href: "Stream.cls.html#%21LineOut"
-        },
-        {
-        id: "%21Lines", type: "Routine", name: "!Lines", info: "Stream.cls",  href: "Stream.cls.html#%21Lines"
-        },
-        {
-        id: "%21Stream", type: "Routine", name: "!Stream", info: "Stream.cls",  href: "Stream.cls.html#%21Stream"
-        },
-        {
-        id: "NormalizedStreamName", type: "Routine", name: "NormalizedStreamName", info: "Stream.cls",  href: "Stream.cls.html#NormalizedStreamName"
-        },
 ];

### 2023-10-02 ARB/standards/work-in-progress/unicode/UnicodeTools/doc/js/search-data.js
Move docs for Properties.cls to readme.md

@@ -1,71 +1,0 @@
-var searchData = [
-        {
-        id: "Properties.cls", type: "Package", name: "Properties.cls", info: null,  href: "Properties.cls.html"
-    },
-        {
-        id: "MultiStageTable", type: "Class", name: "MultiStageTable", info: "Properties.cls", href: "Properties.cls.html#MultiStageTable"
-        },
-            {
-            id: "init", type: "Method", name: "init", info: "Class MultiStageTable", href: "Properties.cls.html#init"
-            },
-            {
-            id: "%5B%5D-1", type: "Method", name: "[]", info: "Class MultiStageTable", href: "Properties.cls.html#%5B%5D-1"
-            },
-            {
-            id: "Compress", type: "Method", name: "Compress", info: "Class MultiStageTable", href: "Properties.cls.html#Compress"
-            },
-        {
-        id: "PersistentStringTable", type: "Class", name: "PersistentStringTable", info: "Properties.cls", href: "Properties.cls.html#PersistentStringTable"
-        },
-            {
-            id: "Load", type: "Method", name: "Load", info: "Class PersistentStringTable", href: "Properties.cls.html#Load"
-            },
-            {
-            id: "Save", type: "Method", name: "Save", info: "Class PersistentStringTable", href: "Properties.cls.html#Save"
-            },
-        {
-        id: "Unicode.Property", type: "Class", name: "Unicode.Property", info: "Properties.cls", href: "Properties.cls.html#Unicode.Property"
-        },
-            {
-            id: "%5B%5D", type: "Method", name: "[]", info: "Class Unicode.Property", href: "Properties.cls.html#%5B%5D"
-            },
-            {
-            id: "Activate", type: "Method", name: "Activate", info: "Class Unicode.Property", href: "Properties.cls.html#Activate"
-            },
-            {
-            id: "BinaryFile", type: "Method", name: "BinaryFile", info: "Class Unicode.Property", href: "Properties.cls.html#BinaryFile"
-            },
-            {
-            id: "BinFile.Qualify", type: "Method", name: "BinFile.Qualify", info: "Class Unicode.Property", href: "Properties.cls.html#BinFile.Qualify"
-            },
-            {
-            id: "getPersistent", type: "Method", name: "getPersistent", info: "Class Unicode.Property", href: "Properties.cls.html#getPersistent"
-            },
-            {
[…]

### 2023-10-02 ARB/standards/work-in-progress/unicode/UnicodeTools/parser/doc/js/search-data.js
Move docs for parser to readme.md

@@ -1,71 +1,0 @@
-var searchData = [
-        {
-        id: "Rexx.Tokenizer.cls", type: "Package", name: "Rexx.Tokenizer.cls", info: null,  href: "Rexx.Tokenizer.cls.html"
-    },
-        {
-        id: "ANSI.Rexx.Tokenizer", type: "Class", name: "ANSI.Rexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ANSI.Rexx.Tokenizer"
-        },
-        {
-        id: "ANSI.Rexx.Unicode.Tokenizer", type: "Class", name: "ANSI.Rexx.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ANSI.Rexx.Unicode.Tokenizer"
-        },
-        {
-        id: "ooRexx.Tokenizer", type: "Class", name: "ooRexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ooRexx.Tokenizer"
-        },
-        {
-        id: "ooRexx.Unicode.Tokenizer", type: "Class", name: "ooRexx.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#ooRexx.Unicode.Tokenizer"
-        },
-        {
-        id: "Regina.Tokenizer", type: "Class", name: "Regina.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Regina.Tokenizer"
-        },
-        {
-        id: "Regina.Unicode.Tokenizer", type: "Class", name: "Regina.Unicode.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Regina.Unicode.Tokenizer"
-        },
-        {
-        id: "Rexx.Tokenizer", type: "Class", name: "Rexx.Tokenizer", info: "Rexx.Tokenizer.cls", href: "Rexx.Tokenizer.cls.html#Rexx.Tokenizer"
-        },
-            {
-            id: "init", type: "Method", name: "init", info: "Class Rexx.Tokenizer", href: "Rexx.Tokenizer.cls.html#init"
-            },
-            {
[…]

### 2023-10-02 ARB/standards/work-in-progress/unicode/UnicodeTools/parser/doc/js/typeahead-bootstrap.js

@@ -1,709 +1,0 @@
-/* =============================================================
- * bootstrap3-typeahead.js v4.0.2
- * https://github.com/bassjobsen/Bootstrap-3-Typeahead
- * =============================================================
- * Original written by @mdo and @fat
- * =============================================================
- * Copyright 2014 Bass Jobsen @bassjobsen
- *
- * Licensed under the Apache License, Version 2.0 (the 'License');
- * you may not use this file except in compliance with the License.
- * You may obtain a copy of the License at
- *
- * http://www.apache.org/licenses/LICENSE-2.0
- *
- * Unless required by applicable law or agreed to in writing, software
- * distributed under the License is distributed on an 'AS IS' BASIS,
- * WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
- * See the License for the specific language governing permissions and
- * limitations under the License.
- * ============================================================ */
-
-
-(function (root, factory) {
-
-  'use strict';
-
-  // CommonJS module is defined
-  if (typeof module !== 'undefined' && module.exports) {
-    module.exports = factory(require('jquery'));
-  }
-  // AMD module is defined
-  else if (typeof define === 'function' && define.amd) {
-    define(['jquery'], function ($) {
-      return factory ($);
-    });
-  } else {
-    factory(root.jQuery);
-  }
-
-}(this, function ($) {
-
-  'use strict';
-  // jshint laxcomma: true
-
-
- /* TYPEAHEAD PUBLIC CLASS DEFINITION
-  * ================================= */
-
-  var Typeahead = function (element, options) {
-    this.$element = $(element);
-    this.options = $.extend({}, Typeahead.defaults, options);
-    this.matcher = this.options.matcher || this.matcher;
-    this.sorter = this.options.sorter || this.sorter;
-    this.select = this.options.select || this.select;
-    this.autoSelect = typeof this.options.autoSelect == 'boolean' ? this.options.autoSelect : true;
-    this.highlighter = this.options.highlighter || this.highlighter;
-    this.render = this.options.render || this.render;
-    this.updater = this.options.updater || this.updater;
-    this.displayText = this.options.displayText || this.displayText;
-    this.itemLink = this.options.itemLink || this.itemLink;
-    this.itemTitle = this.options.itemTitle || this.itemTitle;
-    this.followLinkOnSelect = this.options.followLinkOnSelect || this.followLinkOnSelect;
-    this.source = this.options.source;
[…]

### 2023-10-03 ARB/standards/work-in-progress/unicode/UnicodeTools/setenv.sh
Simplify setenv

@@ -1,3 +1,3 @@
 # Call it using ". ./setenv.sh", else export will not work!
 echo "Setting env"
-export PATH=$(pwd):$(pwd)/properties:$PATH
+export PATH=$(pwd):$PATH

### 2023-10-04 ARB/standards/work-in-progress/unicode/Unicode.Implimentation.tex
Add files via upload
Externalize bibliography for Unicode.tex

@@
 \documentclass[b4paper]{article}
+\usepackage[
+    backend=biber,%
+    backref=true,%
+    indexing=true,%
+    sorting=anyt,%,
+    style=alphabetic%
+    ]%
+    {biblatex}
+\addbibresource{ARB.bib}
+\addbibresource{FAQ.bib}
+\addbibresource{IETF.bib}
+\addbibresource{ISO.bib}
+\addbibresource{Unicode.bib}
 \usepackage{enumitem}
@@
 \usepackage{hyperref}
+\usepackage{ifthen}
+\usepackage{imakeidx}
 \usepackage{listings}         %Requires luatex engine to handle UTF-8
@@
 \usepackage{makeidx}
+\usepackage{tocbibind}
 \usepackage{upquote}
@@
 
-\makeindex
+\makeindex[intoc]
 \setmainfont{CMU Serif}
@@
 
+\newcommand \fullcref [1]
+   { \ifthenelse {\equal{\nameref{#1}}{}}
+       {\cref{#1}}
+       {\cref{#1} (\nameref{#1})}
+   }
+
 \title{Unicode Implementation Issues%
@@
 The Architecture Review board is investigating areas in which Rexx may be
-in need of enhancements, some of which are listed in \cite{Req}.
+in need of enhancements, some of which are listed in \cite{ARB:Issues}.
 Many modern languages support Unicode encoded as
-\href{https://datatracker.ietf.org/doc/rfc3629/}{UTF-8} (\cite{RFC 3629}),
-and the \href{https://www.ietf.org/}{IETF}, in \cite{RFC 5198},
+\href{https://datatracker.ietf.org/doc/rfc3629/}{UTF-8} (\cite{RFC:3629}),
+and the \href{https://www.ietf.org/}{IETF}, in \cite{RFC:5198},
 has mandated the use of UTF-8 with NFC in new protocols.
@@
 and that support should not break existing code.
-Elaboration of this issue may be added to \cite{Req} \cref{issue:MB,issue:UTF-8}.
+Elaboration of this issue may be added to \cite{ARB:Issues} \cref{issue:MB,issue:UTF-8}.
 
@@
 are presented in other documents.  Also, most of the details are under
-active discussion, so everything here is provisional.
+active discussion, so everything here is provisional. Information on Unicode may be found at \citetitle{Unicode} and \citetitle{UFAQs}.
 
@@
 
-The definitions given in \cite{Glossary} and \cite{Unicode} take
+The definitions given in \cite{USTD:Gloss} and \cite{Unicode} take
 precedence over those given here. Quoted text is taken from those sources.
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[caption={Non-ASCII upper casing in classic Rexx}]
     foo = 'Ren!\'{e}!'
@@
 
-\subsection{Identifiers, labels and procedure names}
-The current definition restricts symbols to ASCII characters, which
-excludes letters used in many languages. If the rules are extended, in
-accordance with \cite{Annex31}.  to include Unicode letters and digits
-beyond ASCII, the rules for equality of symbols must be addressed.
-Are composed and decomposed strings identical? Are base characters
-identical to their compatibility alterrnates? Are subscript and
-superscript digits distinct from their ASCII counterparts?
-
-When a numeric value is used as a label or procedure name, a precise
-definition of the semantics is needed.  While the definition
-"Translate Binary to a coded string." of Config\_B2C in ANSI X3J18-199X
-is rather vague, it is adequate when the character set is restricted
-to an 8-bit character set.  The definition must be replaced by one
-that is unamgiguous when applied to Unicode.  (\ref{issue:IAF})
-There are several
-obvious options.
+\subsection{Numeric string conversion to characters}
+\label{problem:numeric}
+The definitions of binary and hexadecimal literals, and of the b2c()
+and x2c() functions, in \cite{ANSI:J18}, depend on Config\_B2C. While the
+definition"Translate Binary to a coded string." of Config\_B2C in ANSI
+X3J18-1996 is rather vague, it is adequate when the character set is
+restricted to an 8-bit character set.  The definition must be replaced
+by one that is unamgiguous when applied to Unicode.
+(\cref{issue:CUC}) There are several obvious options.
+
 \begin{enumerate}
@@
 \end{enumerate}
+
 Each of these options has disadvantages.
-The same issue exists for Config\_C2B.
+
+The converse issue exists for Config\_C2B.
+
+\subsection{Identifiers, labels and procedure names}
+The current definition restricts symbols to ASCII characters, which
+excludes letters used in many languages. If the rules are extended, in
+accordance with \cite{UAX31}. to include Unicode letters and digits
+beyond ASCII, the rules for equality of symbols must be addressed.
+Are composed and decomposed strings identical? Are base characters
+identical to their compatibility alterrnates? Are subscript and
+superscript digits distinct from their ASCII counterparts?
+
+When a numeric value is used as a label or procedure name, a precise
+definition of the semantics is needed.  This is essentially the same
+issue as \fullcref{problem:numeric} above.
 
@@
 code points and the exixtence of compatibility code points.
-Rexx needs a way to test two Unicode strings for equivalence. (\cref{issue:SNO,issue:NFCD,issue:NFKCD})
+Rexx needs a way to test two Unicode strings for equivalence.
+(\cref{issue:SNO,issue:NFCD,issue:NFKCD})
+
+\subsection{PARSE}
+The PARSE tempates currently support spliiting text based on a
+restricted set of whitespace characters.  A decision must be made
+whether to only support space and tab as whitespace for PARSE, or to
+use one of the whitespace definitions in \cite{UAX31}.
 
@@
 \subsection{Unassigned and invalid code points}
-There has been some discussion of the need to detect unassigned code points. (\cref{issue:UICP,issue:ED})
+There has been some discussion of the need to detect
+unassigned\footnote{Unsigned is not the same as noncharacter.}
+code points. (\cref{issue:UICP,issue:ED})
 
@@
 
-\section{ARB Issues}
+\subsection{Tools}
+There are tools available that manipilate Rexx source code in various
+ways, e.g., code completion, cross referencing, navigation, prettyprinting, syntax highligting.
+These tools may not handle new syntax added in support of Unicode. Examples include
+\begin{itemize}
+\item The IBM Z\textregistered Open Editor \\
+A Rexx Language Support plugin for Visual Studio Code
+\item The \LaTeX\  package \href{https://www.ctan.org/pkg/listings}{\textbf{listings}}
+\item The \LaTeX \ package \href{https://www.ctan.org/pkg/minted}{\textbf{minted}}
+\item The Python package Pygments
+\end{itemize}
+
+\section{ARB GitHub Issues}
 \label{ARBiss}
 
-This section lists all issues in \cite{Req}, whether or not they relate to Unicode.
-
-\begin{issue}[label=I\arabic*,ref=\fbox{I\arabic*}]
+This section lists all issues in \cite{ARB:Issues}, whether or not they relate to Unicode.
+
+\begin{issue}[label=I.\arabic*,ref=\fbox{I.\arabic*}]
 \item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730857}%
@@
 
+The RXU preprocessor accepts \textbf{OPTIONS DEFAULTSTRING} \textit{default}, where default can be one of \textbf{ BYTES},\textbf{ CODEPOINTS}, \textbf{TEXT} OR \textbf{NONE}.
+
 \subsection{Classes and types}
@@
 used, or a separate class for each type.
+
+RXU uses BYTES for both octet strings and legacy strings.
 
@@
 Assuming that the width is in in terms of octets
-\begin{lstlisting}[linewidth=35em]
+\begin{lstlisting}[
+    caption={c2x() with width parameter},%
+    linewidth=35em%
+    ]
     foo = 'Ren!\'{e}!'
@@
 There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
-\begin{lstlisting}[linewidth=35em]
+\begin{lstlisting}[
+    caption={Implicit coercion},%
+    linewidth=35em]%
+    ]
     foo = .BYTES~new
@@
 including rules for optional spaces between (hex) digits,
-and should take into account the recommendations in \cite{RFC 5137}.
+and should take into account the recommendations in \cite{RFC:5137}.
 Implementations should use the machine readable data bases published by
@@
 \end{tabular}
-
+\\
 An issue that must be resolved is whether string literals should have
@@
 code points 86, 96, 96, making it harder to read and edit?
+
+Another issue is whether numeric constants used as labels and call
+targets should be treated as strings of legacy code points, UTF-8
+sequences or as strings of Unicode scalars.
+
+RXU has five tpes of string literals:
+\begin{definition}
+\item['string'] The type is determined by OPTIONS.
+\item['string'P] Unicode by code point
+\item['string'T] Unicode by EGC
+\item['space separated tokens'U] Unicode code points by alias, name or value
+\item['string'Y] Legacy bYte string
+\end{definition}
 
@@
 within a UTF-16 surrogate pair,
-or, for EGC access, withing an EGC.
+or, for EGC access, within an EGC.
 Strategies to ameliorate this include
@@
 \item Define a cursor type. \\
-This violates "Everything is a string" in classic Rexx.
+This violates "Everything is a string" in classic Rexx, but is reasonable for ooRexx.
 \end{itemize}
@@
 
-Case folding of symbols and caseless comparisons will use the \cite{UCD}.
+Case folding of symbols and caseless comparisons will use the \citetitle{UCD}.
 \cref{issue:CF,issue:CComp,issue:CConv})
@@
 
-\section{Bibliography}
-
-\subsection{GitHub ARB documents}
-\renewcommand{\refname}{ARB}
-
-\begin{thebibliography}{20}
-\bibitem[New Types]{New Types}
-\href{https://github.com/RexxLA/rexx-repository/edit/master/ARB/standards/work-in-progress/unicode/UnicodeTools/doc/string-types.md}%
-{\textit{New types of strings}}
-
-\bibitem[Requirements]{Req}
-\href{https://github.com/users/RexxLA/projects/2/views/6}%
-{\textit{Unicode Standard Development - requirements}}
-
-\bibitem[Requirements Status]{ReqStat}
-\href{https://github.com/users/RexxLA/projects/2/views/1}%
-{\textit{Unicode Standard Development - requirements status}}
-
-\bibitem[RXU]{RXU}
-\href{https://github.com/RexxLA/rexx-repository/blob/master/ARB/standards/work-in-progress/unicode/UnicodeTools/rxu.md}%
-{\textit{The RXU Rexx Preprocessor for Unicode}}
-
-\end{thebibliography}
-
-\subsection{Official documents}
-\renewcommand{\refname}{References}
-
-\begin{thebibliography}{20}
-\bibitem[Glossary of Unicode Terms]{Glossary}
-\href{https://unicode.org/glossary/}%
-{\textit{Glossary of Unicode Terms}}
-
-\bibitem[ISO/IEC 8859-1:1997 (E)]{ISO 8859-1}
-\href{https://www.open-std.org/JTC1/sc2/wg3/docs/n411.pdf}%
-{\textit{Final Text of DIS 8859-1, 8-bit single-byte coded graphic character sets-Part 1: Latin alphabet No.1}}
-
-\bibitem[ISO/IEC 10646:2020]{ISO 10646}
-\href{https://www.iso.org/standard/76835.html}%
-{\textit{Information technology - Universal coded character set (UCS)}}
-
-\bibitem[RFC 2781]{RFC 2781}
-\href{https://datatracker.ietf.org/doc/rfc2781/}%
-{\textit{UTF-16, a transformation format of ISO 10646}}
-
-\bibitem[RFC 3629, STD 63]{RFC 3629}
-\href{https://datatracker.ietf.org/doc/rfc3629/}%
-{\textit{UTF-8, a transformation format of ISO 10646}}
-
-\bibitem[RFC 5137, BCP 137]{RFC 5137}
-\href{https://datatracker.ietf.org/doc/rfc5137/}%
-{\textit{ASCII Escaping of Unicode Characters}}
-
-\bibitem[RFC 5198]{RFC 5198}
-\href{https://datatracker.ietf.org/doc/rfc5198/}%
-{\textit{Unicode Format for Network Interchange}}
-
-\bibitem[Unicode]{Unicode}
-\href{https://www.unicode.org/standard/standard.html}%
-{\textit{The Unicode Standard}}
-
-\bibitem[Unicode]{Unicode3.9}
-\href{https://www.unicode.org/versions/Unicode15.0.0/ch03.pdf#page=50}%
-{\textit{3.9 Unicode Encoding Forms}}
-
-\bibitem[Unicode Character Database]{UCD}
-\href{https://www.unicode.org/ucd/}%
-{\textit{Unicode Character Database (UCD)}}
-
-\bibitem[Unicode CLDR Project]{CLDR}
-\href{https://cldr.unicode.org/}%
-{\textit{Unicode Common Locale Data Repository (CLDR)}}
-
-\bibitem[Unicode{\textregistered} Standard Annex 29]{Annex29}
-\href{https://www.unicode.org/reports/tr29/}%
-{\textit{Unicode{\textregistered} Standard Annex 29\\ %
-Unicode Text Segmentation}}
-
-\bibitem[Unicode{\textregistered} Standard Annex 31]{Annex31}
-\href{https://www.unicode.org/reports/tr31/}%
-{\textit{Unicode{\textregistered} Standard Annex 31\\ %
-Unicode Identifier and Pattern Syntax}}
-
-\bibitem[Unicode{\textregistered} Standard Annex 42]{Annex42}
-\href{https://www.unicode.org/reports/tr42/tr42-34.html}%
-{\textit{Unicode{\textregistered} Standard Annex 42\\ %
-Unicode Character Database in XML}}
-
-\bibitem[Unicode{\textregistered} Standard Annex 44]{Annex44}
-\href{https://www.unicode.org/reports/tr44/tr44-32.html}%
-{\textit{Unicode{\textregistered} Standard Annex 44\\ %
-Unicode Character Database}}
-
-\end{thebibliography}
-
 \section{Glossary}
 
-The definitions given in \cite{Glossary} and \cite{Unicode} take
+The definitions given in \cite{USTD:Gloss} and \cite{Unicode} take
 precedence over those given here. Quoted text is taken from those sources, or from IETF documents.
@@
 marks, or a join control."
-{\cite{Annex29}} defines two types of grapheme clusters;
+{\cite{UAX29}} defines two types of grapheme clusters;
 "An extended grapheme cluster is the same as a legacy grapheme cluster,
@@
 \item [UTF-8] UCS Transformation Format 8
-(\cite{Unicode3.9} and \cite{RFC 3629}).
+(\cite{USTD:3.9} and \cite{RFC:3629}).
 \\*
@@
 \item [UTF-16] UCS Transformation Format 16
-(\cite{Unicode3.9} and \cite{RFC 2781}).
+(\cite{USTD:3.9} and \cite{RFC:2781}).
 \end{definition}
@@
 
+\section{Bibliography}
+This version of the document prints all included bibliography entries, even those note cited.
+\nocite{*}
+
+\printbibliography[%
+    keyword=ARB,%
+    title={GitHub ARB documents}%
+    ]
+\printbibliography[%
+    keyword=FAQ,%
+    title=FAQs%
+    ]
+\printbibliography[%
+    keyword=standard,%
+    notkeyword=FAQ,%
+    title={Official documents}%
+    ]
+
 \printindex

### 2023-10-25 Unicode.Implimentation.tex
Add files via upload
Add initial RXU information

--- 2023-10-04 ARB/standards/work-in-progress/unicode/Unicode.Implimentation.tex
@@
 
+\defbibcheck{standard}{\iffieldequalstr{type}{standard}{}{\skipentry}}
+
 \makeindex[intoc]
+
 \setmainfont{CMU Serif}
@@
 are presented in other documents.  Also, most of the details are under
-active discussion, so everything here is provisional. Information on Unicode may be found at \citetitle{Unicode} and \citetitle{UFAQs}.
+active discussion, so everything here is provisional.
+Information on Unicode may be found at \citetitle{Unicode} and \citetitle{UFAQs}.
 
@@
 
-The RXU preprocessor accepts \textbf{OPTIONS DEFAULTSTRING} \textit{default}, where default can be one of \textbf{ BYTES},\textbf{ CODEPOINTS}, \textbf{TEXT} OR \textbf{NONE}.
+The RXU preprocessor accepts \textbf{OPTIONS DEFAULTSTRING} \textit{default},
+where default can be one of \textbf{ BYTES},\textbf{ CODEPOINTS}, \textbf{TEXT} OR \textbf{NONE}.
 
@@
 Assuming that the width is in in terms of octets
-\begin{lstlisting}[
+\begin{lstlisting}[%
     caption={c2x() with width parameter},%
@@
 then \textbf{c2x()} might behave differently depending on whether an explicit width is provided.
+
+\subsection{RXU c2x() and c2u()}
+In RXU, \textbf{c2x()} for Unicode code point strings returns the
+hexadecimal values of UTF-8 sequences rather than the hexadecimal
+values of unicode scalars. The second parameter of \textbf{c2u()}
+controls the output format, and \textbf{c2x(foo,UTF-32)} returns space
+separated Unicode scalars.
 
@@
 
-RXU has five tpes of string literals:
+RXU has five types of string literals:
 \begin{definition}
@@
 \item['string'T] Unicode by EGC
-\item['space separated tokens'U] Unicode code points by alias, name or value
+\item['space separated tokens'U]
+Unicode code points by alias, name or hexadecimal value;
+the string is BYTES rather than CODEPOINTS or TEXT.
 \item['string'Y] Legacy bYte string
@@
 \printbibliography[%
-    keyword=standard,%
-    notkeyword=FAQ,%
+    check={standard},%
+    notkeyword={FAQ},%
     title={Official documents}%

### 2023-10-27 ARB/standards/work-in-progress/unicode/Unicode.Implimentation.tex
Add files via upload
More cross-references to RXU

@@
 
-The RXU preprocessor accepts \textbf{OPTIONS DEFAULTSTRING} \textit{default}, where default can be one of \textbf{ BYTES},\textbf{ CODEPOINTS}, \textbf{TEXT} OR \textbf{NONE}.
+\subsubsection{RXU OPTIONS values}
+
+The RXU preprocessor accepts these\cite{RXU:OPT} values on the OPTIONS statement:
+
+\begin{definition}
+\item[DEFAULTSTRING \textit{default}] where \textit{default} determines the type of a string literal with no suffix:
+\begin{definition}
+\item[\underline{BYTES}]Y
+\item[CODEPOINTS]P
+\item[NONE]
+\item[TEXT]T
+\end{definition}
+\item[COERCIONS \textit{ucoercion}] where \textit{ucoercion} determines the result of an operation on strings of mixed string types:
+\begin{definition}
+\item[DEMOTE]Type of lowest operand
+\item[LEFT]Type of left operand
+\item[NONE]Raise SYNTAX
+\item[\underline{PROMOTE}]Type of highest operand
+\item[RIGHT]Type of right operand
+\end{definition}
+\end{definition}
 
@@
 
-RXU has five tpes of string literals:
+\subsubsection{RXU string types}
+
+RXU has five types of string literals\cite{RXU:NewTypes}:
 \begin{definition}
@@
 \item['string'T] Unicode by EGC
-\item['space separated tokens'U] Unicode code points by alias, name or value
-\item['string'Y] Legacy bYte string
+\item['space separated tokens'U]
+Unicode code points by alias, name or hexadecimal value;
+the string is BYTES rather than CODEPOINTS or TEXT.
+\item['string'Y] Legacy bYte string,can only be promoted if valid UTF-8 sequences
 \end{definition}

### 2023-10-31 Unicode.Implimentation.tex
Add files via upload

@@
 \item The \LaTeX \ package \href{https://www.ctan.org/pkg/minted}{\textbf{minted}}
-\item The Python package Pygments
+\item The Python package \href{https://pygments.org/}{\textbf{Pygments}}
 \end{itemize}
@@
 
-The RXU preprocessor accepts \textbf{OPTIONS DEFAULTSTRING} \textit{default},
-where default can be one of \textbf{ BYTES},\textbf{ CODEPOINTS}, \textbf{TEXT} OR \textbf{NONE}.
+\subsubsection{RXU OPTIONS values}
+
+The RXU preprocessor accepts these\cite{RXU:OPT} values on the OPTIONS statement:
+
+\begin{definition}
+\item[DEFAULTSTRING \textit{default}] where \textit{default} determines the type of a string literal with no suffix:
+\begin{definition}
+\item[\underline{BYTES}]Y
+\item[CODEPOINTS]P
+\item[NONE]
+\item[TEXT]T
+\end{definition}
+\item[COERCIONS \textit{ucoercion}] where \textit{ucoercion} determines the result of an operation on strings of mixed string types:
+\begin{definition}
+\item[DEMOTE]Type of lowest operand
+\item[LEFT]Type of left operand
+\item[NONE]Raise SYNTAX
+\item[\underline{PROMOTE}]Type of highest operand
+\item[RIGHT]Type of right operand
+\end{definition}
+\end{definition}
 
@@
 
-RXU has five types of string literals:
+\subsubsection{RXU string types}
+
+RXU has five types of string literals\cite{RXU:NewTypes}:
 \begin{definition}
@@
 the string is BYTES rather than CODEPOINTS or TEXT.
-\item['string'Y] Legacy bYte string
+\item['string'Y] Legacy bYte string,can only be promoted if valid UTF-8 sequences
 \end{definition}
@@
 
+\subsubsection{RXU I/O}
+
+in RXU, stream I/O\cite{RXU:STREAM} defaults to legacy (BYTES).
+
+RXU adds an \textbf{ENCODING} option to \textbf{STREAM(}\textit{streamid},\textbf{OPEN }\textit{options}\textbf{)}.
+When this option is used the stream is Unicode enabled, and some stream functions are not allowed.
+\textbf{ENCODING} must be followed by the name of then encoding, which for Unicode may in turn be followed by \textit{error} and \textit{target} options:
+
+\begin{itemize}
+  \item \underline{\textbf{REPLACE}} \\
+  Replace invalid characters with the Unicode Replacement Character (U+FFFd).
+  \item \textbf{SYNTAX} \\
+  Raise SYNTAX for invalid characters.
+  \item \textbf{CODEPOINTS} \\
+  Units of codepoints.
+  \item \underline{\textbf{TEXT}} \\
+  Units of extended grapheme clusters.
+\end{itemize}
+
+RXU adds  options to  \textbf{STREAM(}\textit{streamid},\textbf{QUERY}\textit{options}\textbf{)}
+
+\begin{itemize}
+\item \textbf{Query Encoding Name}
+\item \textbf{Query Encoding Target} \\
+CODEPOINTS  or TEXT
+\item \textbf{Query Encoding Error} \\
+REPLACE or SYNTAX
+\item \textbf{Query Encoding LastError}
+\item \textbf{Query Encoding \\
+Everything but LastError}
+\end{itemize}
+
 \subsection{Case folding}
 
-Case folding of symbols and caseless comparisons will use the \citetitle{UCD}.
+Case folding of symbols and caseless comparisons will use the \citetitle{UCD}ENCODING.
 \cref{issue:CF,issue:CComp,issue:CConv})
@@
 \end{definition}
+
+\subsubsection{RXU functions and methods}
+
+RXU adds the \textbf{DECODE} and \textbf{ENCODE} methods in the Unicode class. These provide an easy way to implement the proposed \textbf{iconv} function.
+
+RXU adds these\cite{RXU:NewBIFs} additional built-in functions:
+
+\begin{itemize}
+\item \textbf{BYTES}
+\item \textbf{C2U}
+\item \textbf{CODEPOINTS}
+\item \textbf{GRAPHEMES}
+\item \textbf{N2P}
+\item \textbf{P2N}
+\item \textbf{STRINGTYPE}
+\item \textbf{TEXT}
+\item \textbf{UNICODE(}\textit{string},\textit{function}\textbf{)} \\
+Performs the named function on \textit{string}:
+\begin{description}
+\item[isNFD] Returns 1 if string is in normalized form D
+\item[toLowercase]
+Returns upper case translatio.
+\item[toUpercase]
+Returns lower case translation
+\item[toNFD] Normalizes to form D (decomposed)
+\end{description}
+\item \textbf{UNICODE(}\textit{codepoint},\textbf{PROPERTY},\textit{property}\textbf{)} \\
+Retrieves the value of the named property
+\item \textbf{UTF8} \\
+Validates a UTF-8 string and optionally converts it to a different Unicode transform.
+\end{itemize}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/built-in_classes.tex
add cyrillic.rexx, rename unicode.md to Unicode_Requirements.md
upd

@@
-\hypertarget{built-in-classes}{%
-\chapter{Built-in classes}\label{built-in-classes}}
+%preprocessed texin
+\chapter{Built-in classes}\label{built-in-classes}
 
-\hypertarget{notation}{%
-\section{Notation}\label{notation}}
+\section{Notation}\label{notation}
 
@@
 
-\hypertarget{object-class-and-method}{%
-\section{Object, class and method}\label{object-class-and-method}}
+\section{Object, class and method}\label{object-class-and-method}
 
@@
 
-\hypertarget{the-object-class}{%
-\subsection{The object class}\label{the-object-class}}
+\subsection{The object class}\label{the-object-class}
 
@@
 
-\hypertarget{the-class-class}{%
-\subsection{The class class}\label{the-class-class}}
+\subsection{The class class}\label{the-class-class}
 
@@
 
-::method `' We can do all the operators by appeal to classic section 7.
-::method'-' ::method `-'
+::method `\textquotesingle{} We can do all the operators by appeal to
+classic section 7. ::method'-' ::method `-'
 
@@
 
-\hypertarget{the-supplier-class}{%
-\section{The supplier class}\label{the-supplier-class}}
+\section{The supplier class}\label{the-supplier-class}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/built-in_functions.tex
work in standard book production process including readme.md
upd

@@
-\hypertarget{built-in-functions}{%
-\chapter{Built-in functions}\label{built-in-functions}}
-
-\hypertarget{notation}{%
-\section{Notation}\label{notation}}
+%preprocessed texin
+\chapter{Built-in functions}\label{built-in-functions}
+
+\section{Notation}\label{notation}
 
@@
 
-\hypertarget{routines-used-by-built-in-functions}{%
 \section{Routines used by built-in
-functions}\label{routines-used-by-built-in-functions}}
+functions}\label{routines-used-by-built-in-functions}
 
@@
 
-\hypertarget{argument-checking}{%
-\subsection{Argument checking}\label{argument-checking}}
+\subsection{Argument checking}\label{argument-checking}
 
@@
 
-\hypertarget{raising-the-syntax-condition}{%
 \subsection{Raising the SYNTAX
-condition}\label{raising-the-syntax-condition}}
+condition}\label{raising-the-syntax-condition}
 
@@
 
-\hypertarget{character-built-in-functions}{%
 \section{Character built-in
-functions}\label{character-built-in-functions}}
+functions}\label{character-built-in-functions}
 
@@
 
-\hypertarget{abbrev}{%
-\subsection{ABBREV}\label{abbrev}}
+\subsection{ABBREV}\label{abbrev}
 
@@
 
-\hypertarget{center}{%
-\subsection{CENTER}\label{center}}
+\subsection{CENTER}\label{center}
 
@@
 
-\hypertarget{centre}{%
-\subsection{CENTRE}\label{centre}}
+\subsection{CENTRE}\label{centre}
 
@@
 
-\hypertarget{changestr}{%
-\subsection{CHANGESTR}\label{changestr}}
+\subsection{CHANGESTR}\label{changestr}
 
@@
 
-\hypertarget{compare}{%
-\subsection{COMPARE}\label{compare}}
+\subsection{COMPARE}\label{compare}
 
@@
 
-\hypertarget{copies}{%
-\subsection{COPIES}\label{copies}}
+\subsection{COPIES}\label{copies}
 
@@
 
-\hypertarget{countstr}{%
-\subsection{COUNTSTR}\label{countstr}}
+\subsection{COUNTSTR}\label{countstr}
 
@@
 
-\hypertarget{datatype}{%
-\subsection{DATATYPE}\label{datatype}}
+\subsection{DATATYPE}\label{datatype}
 
@@
 
-\hypertarget{delstr}{%
-\subsection{DELSTR}\label{delstr}}
+\subsection{DELSTR}\label{delstr}
 
@@
 
-\hypertarget{delword}{%
-\subsection{DELWORD}\label{delword}}
+\subsection{DELWORD}\label{delword}
 
@@
 
-\hypertarget{insert}{%
-\subsection{INSERT}\label{insert}}
+\subsection{INSERT}\label{insert}
 
@@
 
-\hypertarget{lastpos}{%
-\subsection{LASTPOS}\label{lastpos}}
+\subsection{LASTPOS}\label{lastpos}
 
@@
 
-\hypertarget{left}{%
-\subsection{LEFT}\label{left}}
+\subsection{LEFT}\label{left}
 
@@
 
-\hypertarget{overlay}{%
-\subsection{OVERLAY}\label{overlay}}
+\subsection{OVERLAY}\label{overlay}
 
@@
 
-\hypertarget{reverse}{%
-\subsection{REVERSE}\label{reverse}}
+\subsection{REVERSE}\label{reverse}
 
@@
 
-\hypertarget{right}{%
-\subsection{RIGHT}\label{right}}
+\subsection{RIGHT}\label{right}
 
@@
 
-\hypertarget{space}{%
-\subsection{SPACE}\label{space}}
+\subsection{SPACE}\label{space}
 
@@
 
-\hypertarget{strip}{%
-\subsection{STRIP}\label{strip}}
+\subsection{STRIP}\label{strip}
 
@@
 
-\hypertarget{substr}{%
-\subsection{SUBSTR}\label{substr}}
+\subsection{SUBSTR}\label{substr}
 
@@
 
-\hypertarget{subword}{%
-\subsection{SUBWORD}\label{subword}}
+\subsection{SUBWORD}\label{subword}
 
@@
 
-\hypertarget{translate}{%
-\subsection{TRANSLATE}\label{translate}}
+\subsection{TRANSLATE}\label{translate}
 
@@
 
-\hypertarget{verify}{%
-\subsection{VERIFY}\label{verify}}
+\subsection{VERIFY}\label{verify}
 
@@
 
-\hypertarget{word}{%
-\subsection{WORD}\label{word}}
+\subsection{WORD}\label{word}
 
@@
 
-\hypertarget{wordindex}{%
-\subsection{WORDINDEX}\label{wordindex}}
+\subsection{WORDINDEX}\label{wordindex}
 
@@
 
-\hypertarget{wordlength}{%
-\subsection{WORDLENGTH}\label{wordlength}}
+\subsection{WORDLENGTH}\label{wordlength}
 
@@
 
-\hypertarget{wordpos}{%
-\subsection{WORDPOS}\label{wordpos}}
+\subsection{WORDPOS}\label{wordpos}
 
@@
 
-\hypertarget{words}{%
-\subsection{WORDS}\label{words}}
+\subsection{WORDS}\label{words}
 
@@
 
-\hypertarget{xrange}{%
-\subsection{XRANGE}\label{xrange}}
+\subsection{XRANGE}\label{xrange}
 
@@
 
-\hypertarget{arithmetic-built-in-functions}{%
 \section{Arithmetic built-in
-functions}\label{arithmetic-built-in-functions}}
+functions}\label{arithmetic-built-in-functions}
 
@@
 
-\hypertarget{abs}{%
-\subsection{ABS}\label{abs}}
+\subsection{ABS}\label{abs}
 
@@
 
-\hypertarget{format}{%
-\subsection{FORMAT}\label{format}}
+\subsection{FORMAT}\label{format}
 
@@
 
-\hypertarget{max}{%
-\subsection{MAX}\label{max}}
+\subsection{MAX}\label{max}
 
@@
 
-\hypertarget{min}{%
-\subsection{MIN}\label{min}}
+\subsection{MIN}\label{min}
 
@@
 
-\hypertarget{sign}{%
-\subsection{SIGN}\label{sign}}
+\subsection{SIGN}\label{sign}
 
@@
 
-\hypertarget{trunc}{%
-\subsection{TRUNC}\label{trunc}}
+\subsection{TRUNC}\label{trunc}
 
@@
 
-\hypertarget{state-built-in-functions}{%
-\section{State built-in functions}\label{state-built-in-functions}}
+\section{State built-in functions}\label{state-built-in-functions}
 
@@
 
-\hypertarget{address}{%
-\subsection{ADDRESS}\label{address}}
+\subsection{ADDRESS}\label{address}
 
@@
 
-\hypertarget{arg}{%
-\subsection{ARG}\label{arg}}
+\subsection{ARG}\label{arg}
 
@@
 
-\hypertarget{condition}{%
-\subsection{CONDITION}\label{condition}}
+\subsection{CONDITION}\label{condition}
 
@@
 
-\hypertarget{errortext}{%
-\subsection{ERRORTEXT}\label{errortext}}
+\subsection{ERRORTEXT}\label{errortext}
 
@@
 
-\hypertarget{form}{%
-\subsection{FORM}\label{form}}
+\subsection{FORM}\label{form}
 
@@
 
-\hypertarget{fuzz}{%
-\subsection{FUZZ}\label{fuzz}}
+\subsection{FUZZ}\label{fuzz}
 
@@
 
-\hypertarget{sourceline}{%
-\subsection{SOURCELINE}\label{sourceline}}
+\subsection{SOURCELINE}\label{sourceline}
 
@@
 
-\hypertarget{trace}{%
-\subsection{TRACE}\label{trace}}
+\subsection{TRACE}\label{trace}
 
@@
 
-\hypertarget{b2xx}{%
-\subsection{B2xX}\label{b2xx}}
+\subsection{B2xX}\label{b2xx}
 
@@
 
-\hypertarget{bitand}{%
-\subsection{BITAND}\label{bitand}}
+\subsection{BITAND}\label{bitand}
 
@@
 
-\hypertarget{bitor}{%
-\subsection{BITOR}\label{bitor}}
+\subsection{BITOR}\label{bitor}
 
@@
 
-\hypertarget{bitxor}{%
-\subsection{BITXOR}\label{bitxor}}
+\subsection{BITXOR}\label{bitxor}
 
@@
 
-\hypertarget{c2d}{%
-\subsection{C2D}\label{c2d}}
+\subsection{C2D}\label{c2d}
 
@@
 
-\hypertarget{d2c}{%
-\subsection{D2C}\label{d2c}}
+\subsection{D2C}\label{d2c}
 
@@
 
-\hypertarget{d2x}{%
-\subsection{D2X}\label{d2x}}
+\subsection{D2X}\label{d2x}
 
@@
 
-\hypertarget{x2b}{%
-\subsection{X2B}\label{x2b}}
+\subsection{X2B}\label{x2b}
 
@@
 
-\hypertarget{x2c}{%
-\subsection{X2C}\label{x2c}}
+\subsection{X2C}\label{x2c}
 
@@
 
-\hypertarget{x2d}{%
-\subsection{X2D}\label{x2d}}
+\subsection{X2D}\label{x2d}
 
@@
 
-\hypertarget{inputoutput-built-in-functions}{%
 \section{Input/Output built-in
-functions}\label{inputoutput-built-in-functions}}
+functions}\label{inputoutput-built-in-functions}
 
@@
 
-\hypertarget{charout}{%
-\subsection{CHAROUT}\label{charout}}
+\subsection{CHAROUT}\label{charout}
 
@@
 
-\hypertarget{chars}{%
-\subsection{CHARS}\label{chars}}
+\subsection{CHARS}\label{chars}
 
@@
 
-\hypertarget{linein}{%
-\subsection{LINEIN}\label{linein}}
+\subsection{LINEIN}\label{linein}
 
@@
 
-\hypertarget{lineout}{%
-\subsection{LINEOUT}\label{lineout}}
+\subsection{LINEOUT}\label{lineout}
 
@@
 
-\hypertarget{qualify}{%
-\subsection{QUALIFY}\label{qualify}}
+\subsection{QUALIFY}\label{qualify}
 
@@
 
-\hypertarget{stream}{%
-\subsection{STREAM}\label{stream}}
+\subsection{STREAM}\label{stream}
 
@@
 
-\hypertarget{other-built-in-functions}{%
-\section{Other built-in functions}\label{other-built-in-functions}}
-
-\hypertarget{date}{%
-\subsection{DATE}\label{date}}
+\section{Other built-in functions}\label{other-built-in-functions}
+
+\subsection{DATE}\label{date}
 
@@
 
-\hypertarget{queued}{%
-\subsection{QUEUED}\label{queued}}
+\subsection{QUEUED}\label{queued}
 
@@
 
-\hypertarget{random}{%
-\subsection{RANDOM}\label{random}}
+\subsection{RANDOM}\label{random}
 
@@
 
-\hypertarget{symbol}{%
-\subsection{SYMBOL}\label{symbol}}
+\subsection{SYMBOL}\label{symbol}
 
@@
 
-\hypertarget{time}{%
-\subsection{TIME}\label{time}}
+\subsection{TIME}\label{time}
 
@@
 
-\hypertarget{queued-1}{%
-\subsection{QUEUED}\label{queued-1}}
+\subsection{QUEUED}\label{queued-1}
 
@@
 
-\hypertarget{random-1}{%
-\subsection{RANDOM}\label{random-1}}
+\subsection{RANDOM}\label{random-1}
 
@@
 
-\hypertarget{symbol-1}{%
-\subsection{SYMBOL}\label{symbol-1}}
+\subsection{SYMBOL}\label{symbol-1}
 
@@
 
-\hypertarget{time-1}{%
-\subsection{TIME}\label{time-1}}
+\subsection{TIME}\label{time-1}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/configuration.tex
work in standard book production process including readme.md
upd

@@
-\hypertarget{configuration}{%
-\chapter{Configuration}\label{configuration}}
+%preprocessed texin
+\chapter{Configuration}\label{configuration}
 
@@
 
-\hypertarget{notation}{%
-\section{Notation}\label{notation}}
+\section{Notation}\label{notation}
 
@@
 
-\hypertarget{notation-for-completion-response-and-conditions}{%
 \subsection{Notation for completion response and
-conditions}\label{notation-for-completion-response-and-conditions}}
+conditions}\label{notation-for-completion-response-and-conditions}
 
@@
 
-\hypertarget{processing-initiation}{%
-\section{Processing initiation}\label{processing-initiation}}
+\section{Processing initiation}\label{processing-initiation}
 
@@
 
-\hypertarget{api-start}{%
-\subsection{API Start}\label{api-start}}
+\subsection{API Start}\label{api-start}
 
@@
 
-\hypertarget{source-programs-and-character-sets}{%
 \section{Source programs and character
-sets}\label{source-programs-and-character-sets}}
+sets}\label{source-programs-and-character-sets}
 
@@
 
-\hypertarget{syntactic_characters}{%
-\subsection{Syntactic\_characters}\label{syntactic_characters}}
+\subsection{Syntactic\_characters}\label{syntactic_characters}
 
@@
 
-\hypertarget{extra_letters}{%
-\subsection{Extra\_letters}\label{extra_letters}}
+\subsection{Extra\_letters}\label{extra_letters}
 
@@
 
-\hypertarget{other_blank_characters}{%
-\subsection{Other\_blank\_characters}\label{other_blank_characters}}
+\subsection{Other\_blank\_characters}\label{other_blank_characters}
 
@@
 
-\hypertarget{other_negators}{%
-\subsection{Other\_negators}\label{other_negators}}
+\subsection{Other\_negators}\label{other_negators}
 
@@
 \begin{itemize}
-\item
-  \begin{itemize}
 \tightlist
 \item
+  \begin{itemize}
+  \tightlist
+  \item
     circumflex accent, caret;
@@
 
-\hypertarget{other_characters}{%
-\subsection{Other\_characters}\label{other_characters}}
+\subsection{Other\_characters}\label{other_characters}
 
@@
 
-\hypertarget{configuration-characters-and-encoding}{%
 \section{Configuration characters and
-encoding}\label{configuration-characters-and-encoding}}
+encoding}\label{configuration-characters-and-encoding}
 
@@
 
-\hypertarget{config_sourcechar}{%
-\subsection{Config\_SourceChar}\label{config_sourcechar}}
+\subsection{Config\_SourceChar}\label{config_sourcechar}
 
@@
 
-\hypertarget{config_otherblankcharacters}{%
-\subsection{Config\_OtherBlankCharacters}\label{config_otherblankcharacters}}
+\subsection{Config\_OtherBlankCharacters}\label{config_otherblankcharacters}
 
@@
 
-\hypertarget{config_upper}{%
-\subsection{Config\_Upper}\label{config_upper}}
+\subsection{Config\_Upper}\label{config_upper}
 
@@
 
-\hypertarget{config_lower}{%
-\subsection{Config\_Lower}\label{config_lower}}
+\subsection{Config\_Lower}\label{config_lower}
 
@@
 
-\hypertarget{config_compare}{%
-\subsection{Config\_Compare}\label{config_compare}}
+\subsection{Config\_Compare}\label{config_compare}
 
@@
 
-\hypertarget{config-_b2c}{%
-\subsection{Config \_B2C}\label{config-_b2c}}
+\subsection{Config \_B2C}\label{config-_b2c}
 
@@
 
-\hypertarget{config_c2b}{%
-\subsection{Config\_C2B}\label{config_c2b}}
+\subsection{Config\_C2B}\label{config_c2b}
 
@@
 
-\hypertarget{config_substr}{%
-\subsection{Config\_Substr}\label{config_substr}}
+\subsection{Config\_Substr}\label{config_substr}
 
@@
 
-\hypertarget{config-length}{%
-\subsection{Config Length}\label{config-length}}
+\subsection{Config Length}\label{config-length}
 
@@
 
-\hypertarget{config_xrange}{%
-\subsection{Config\_Xrange}\label{config_xrange}}
+\subsection{Config\_Xrange}\label{config_xrange}
 
@@
 
-\hypertarget{objects}{%
-\section{Objects}\label{objects}}
+\section{Objects}\label{objects}
 
@@
 
-\hypertarget{config_objectnew}{%
-\subsection{Config\_ObjectNew}\label{config_objectnew}}
+\subsection{Config\_ObjectNew}\label{config_objectnew}
 
@@
 
-\hypertarget{config_array_size}{%
-\subsection{Config\_Array\_Size}\label{config_array_size}}
+\subsection{Config\_Array\_Size}\label{config_array_size}
 
@@
 
-\hypertarget{config_array_put}{%
-\subsection{Config\_Array\_Put}\label{config_array_put}}
+\subsection{Config\_Array\_Put}\label{config_array_put}
 
@@
 
-\hypertarget{config_array_at}{%
-\subsection{Config\_Array\_At}\label{config_array_at}}
+\subsection{Config\_Array\_At}\label{config_array_at}
 
@@
 
-\hypertarget{config_array_hasindex}{%
-\subsection{Config\_Array\_Hasindex}\label{config_array_hasindex}}
+\subsection{Config\_Array\_Hasindex}\label{config_array_hasindex}
 
@@
 
-\hypertarget{config_array_remove}{%
-\subsection{Config\_Array\_Remove}\label{config_array_remove}}
+\subsection{Config\_Array\_Remove}\label{config_array_remove}
 
@@
 
-\hypertarget{commands}{%
-\section{Commands}\label{commands}}
+\section{Commands}\label{commands}
 
@@
 
-\hypertarget{config-_command}{%
-\subsection{Config \_Command}\label{config-_command}}
+\subsection{Config \_Command}\label{config-_command}
 
@@
 
-\hypertarget{external-routines}{%
-\section{External routines}\label{external-routines}}
+\section{External routines}\label{external-routines}
 
@@
 
-\hypertarget{config_externalroutine}{%
-\subsection{Config\_ExternalRoutine}\label{config_externalroutine}}
+\subsection{Config\_ExternalRoutine}\label{config_externalroutine}
 
@@
 
-\hypertarget{config_externalmethod}{%
-\subsection{Config\_ExternalMethod}\label{config_externalmethod}}
+\subsection{Config\_ExternalMethod}\label{config_externalmethod}
 
@@
 
-\hypertarget{config-push}{%
-\subsection{Config Push}\label{config-push}}
+\subsection{Config Push}\label{config-push}
 
@@
 
-\hypertarget{contig_queue}{%
-\subsection{Contig\_Queue}\label{contig_queue}}
+\subsection{Contig\_Queue}\label{contig_queue}
 
@@
 
-\hypertarget{config_pull}{%
-\subsection{Config\_Pull}\label{config_pull}}
+\subsection{Config\_Pull}\label{config_pull}
 
@@
 
-\hypertarget{contig_queued}{%
-\subsection{Contig\_Queued}\label{contig_queued}}
+\subsection{Contig\_Queued}\label{contig_queued}
 
@@
 
-\hypertarget{streams}{%
-\section{Streams}\label{streams}}
+\section{Streams}\label{streams}
 
@@
 
-\hypertarget{config-_stream_charin}{%
-\subsection{Config \_Stream\_Charin}\label{config-_stream_charin}}
+\subsection{Config \_Stream\_Charin}\label{config-_stream_charin}
 
@@
 
-\hypertarget{config_stream_position}{%
-\subsection{Config\_Stream\_Position}\label{config_stream_position}}
+\subsection{Config\_Stream\_Position}\label{config_stream_position}
 
@@
 
-\hypertarget{config-_stream_command}{%
-\subsection{Config \_Stream\_Command}\label{config-_stream_command}}
+\subsection{Config \_Stream\_Command}\label{config-_stream_command}
 
@@
 
-\hypertarget{config_stream_state}{%
-\subsection{Config\_Stream\_State}\label{config_stream_state}}
+\subsection{Config\_Stream\_State}\label{config_stream_state}
 
@@
 
-\hypertarget{config_stream_charout}{%
-\subsection{Config\_Stream\_Charout}\label{config_stream_charout}}
+\subsection{Config\_Stream\_Charout}\label{config_stream_charout}
 
@@
 
-\hypertarget{config_stream_qualified}{%
-\subsection{Config\_Stream\_Qualified}\label{config_stream_qualified}}
+\subsection{Config\_Stream\_Qualified}\label{config_stream_qualified}
 
@@
 
-\hypertarget{config_stream_unique}{%
-\subsection{Config\_Stream\_Unique}\label{config_stream_unique}}
+\subsection{Config\_Stream\_Unique}\label{config_stream_unique}
 
@@
 
-\hypertarget{config_stream_query}{%
-\subsection{Config\_Stream\_Query}\label{config_stream_query}}
+\subsection{Config\_Stream\_Query}\label{config_stream_query}
 
@@
 
-\hypertarget{config_stream_close}{%
-\subsection{Config\_Stream\_Close}\label{config_stream_close}}
+\subsection{Config\_Stream\_Close}\label{config_stream_close}
 
@@
 
-\hypertarget{config-_stream_count}{%
-\subsection{Config \_Stream\_Count}\label{config-_stream_count}}
+\subsection{Config \_Stream\_Count}\label{config-_stream_count}
 
@@
 
-\hypertarget{external-variable-pools}{%
-\section{External variable pools}\label{external-variable-pools}}
+\section{External variable pools}\label{external-variable-pools}
 
@@
 
-\hypertarget{config-get}{%
-\subsection{Config Get}\label{config-get}}
+\subsection{Config Get}\label{config-get}
 
@@
 
-\hypertarget{config-set}{%
-\subsection{Config Set}\label{config-set}}
+\subsection{Config Set}\label{config-set}
 
@@
 
-\hypertarget{configuration-characteristics}{%
 \section{Configuration
-characteristics}\label{configuration-characteristics}}
+characteristics}\label{configuration-characteristics}
 
@@
 
-\hypertarget{config_constants}{%
-\subsection{Config\_Constants}\label{config_constants}}
+\subsection{Config\_Constants}\label{config_constants}
 
@@
 
-\hypertarget{configuration-routines}{%
-\section{Configuration routines}\label{configuration-routines}}
+\section{Configuration routines}\label{configuration-routines}
 
@@
 
-\hypertarget{config_trace_query}{%
-\subsection{Config\_Trace\_Query}\label{config_trace_query}}
+\subsection{Config\_Trace\_Query}\label{config_trace_query}
 
@@
 
-\hypertarget{config_trace_input}{%
-\subsection{Config\_Trace\_Input}\label{config_trace_input}}
+\subsection{Config\_Trace\_Input}\label{config_trace_input}
 
@@
 
-\hypertarget{config_trace_output}{%
-\subsection{Config\_Trace\_Output}\label{config_trace_output}}
+\subsection{Config\_Trace\_Output}\label{config_trace_output}
 
@@
 
-\hypertarget{config-_default_input}{%
-\subsection{Config \_Default\_Input}\label{config-_default_input}}
+\subsection{Config \_Default\_Input}\label{config-_default_input}
 
@@
 
-\hypertarget{config_default_output}{%
-\subsection{Config\_Default\_Output}\label{config_default_output}}
+\subsection{Config\_Default\_Output}\label{config_default_output}
 
@@
 
-\hypertarget{config_initialization}{%
-\subsection{Config\_Initialization}\label{config_initialization}}
+\subsection{Config\_Initialization}\label{config_initialization}
 
@@
 
-\hypertarget{config_termination}{%
-\subsection{Config\_Termination}\label{config_termination}}
+\subsection{Config\_Termination}\label{config_termination}
 
@@
 
-\hypertarget{config_halt_query}{%
-\subsection{Config\_Halt\_Query}\label{config_halt_query}}
+\subsection{Config\_Halt\_Query}\label{config_halt_query}
 
@@
 
-\hypertarget{config-_nosource}{%
-\subsection{Config \_NoSource}\label{config-_nosource}}
+\subsection{Config \_NoSource}\label{config-_nosource}
 
@@
 
-\hypertarget{config_time}{%
-\subsection{Config\_Time}\label{config_time}}
+\subsection{Config\_Time}\label{config_time}
 
@@
 
-\hypertarget{config_random_seed}{%
-\subsection{Config\_Random\_Seed}\label{config_random_seed}}
+\subsection{Config\_Random\_Seed}\label{config_random_seed}
 
@@
 
-\hypertarget{config_random_next}{%
-\subsection{Config\_Random\_Next}\label{config_random_next}}
+\subsection{Config\_Random\_Next}\label{config_random_next}
 
@@
 
-\hypertarget{config_options}{%
-\subsection{Config\_Options}\label{config_options}}
+\subsection{Config\_Options}\label{config_options}
 
@@
 
-\hypertarget{traps}{%
-\section{Traps}\label{traps}}
+\section{Traps}\label{traps}
 
@@
 
-\hypertarget{variable-pool}{%
-\section{Variable pool}\label{variable-pool}}
+\section{Variable pool}\label{variable-pool}
 
@@
 
-\hypertarget{api-set}{%
-\subsection{API Set}\label{api-set}}
+\subsection{API Set}\label{api-set}
 
@@
 
-\hypertarget{api-value}{%
-\subsection{API Value}\label{api-value}}
+\subsection{API Value}\label{api-value}
 
@@
 
-\hypertarget{api_drop}{%
-\subsection{API\_Drop}\label{api_drop}}
+\subsection{API\_Drop}\label{api_drop}
 
@@
 
-\hypertarget{api-setdirect}{%
-\subsection{API SetDirect}\label{api-setdirect}}
+\subsection{API SetDirect}\label{api-setdirect}
 
@@
 
-\hypertarget{api_valuedirect}{%
-\subsection{API\_ValueDirect}\label{api_valuedirect}}
+\subsection{API\_ValueDirect}\label{api_valuedirect}
 
@@
 
-\hypertarget{api-dropdirect}{%
-\subsection{API DropDirect}\label{api-dropdirect}}
+\subsection{API DropDirect}\label{api-dropdirect}
 
@@
 
-\hypertarget{api-valueother}{%
-\subsection{API ValueOther}\label{api-valueother}}
+\subsection{API ValueOther}\label{api-valueother}
 
@@
 
-\hypertarget{api-next}{%
-\subsection{API Next}\label{api-next}}
+\subsection{API Next}\label{api-next}
 
@@
 
-\hypertarget{api-nextvariable}{%
-\subsection{API NextVariable}\label{api-nextvariable}}
+\subsection{API NextVariable}\label{api-nextvariable}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/definitions_and_document_notation.tex
work in standard book production process including readme.md
upd

@@
-\hypertarget{definitions-and-document-notation}{%
+%preprocessed texin
 \chapter{Definitions and document
-notation}\label{definitions-and-document-notation}}
+notation}\label{definitions-and-document-notation}
 
@@
 
-\hypertarget{definitions}{%
-\section{Definitions}\label{definitions}}
+\section{Definitions}\label{definitions}
 
-\hypertarget{application-programming-interface}{%
-\subsection{application programming
-interface:}\label{application-programming-interface}}
-
+\begin{description}
+\tightlist
+\item[application programming interface]
 A set of functions which allow access to some Rexx facilities from
 non-Rexx programs.
-
-\hypertarget{arguments}{%
-\subsection{arguments:}\label{arguments}}
-
+\item[arguments]
 The expressions (separated by commas) between the parentheses of a
@@
 however invoked.
-
-\hypertarget{built-in-function}{%
-\subsection{built-in function:}\label{built-in-function}}
-
+\item[built-in function]
 A function (which may be called as a subroutine) that is defined in
 section nnn of this standard and can be used directly from a program.
-
-\hypertarget{character-string}{%
-\subsection{character string:}\label{character-string}}
-
+\item[character string]
 A sequence of zero or more characters.
-
-\hypertarget{clause}{%
-\subsection{clause:}\label{clause}}
-
+\item[clause]
 A section of the program, ended by a semicolon. The semicolon may be
 implied by the end of a line or by some other constructs.
-
-\hypertarget{coded}{%
-\subsection{coded:}\label{coded}}
-
+\item[coded]
 A coded string is a string which is not necessarily comprised of
@@
 functions, such as D2C.
-
-\hypertarget{command}{%
-\subsection{command:}\label{command}}
-
+\item[command]
 A clause consisting of just an expression is an instruction known as a
 command. The expression is evaluated and the result is passed as a
-command string to some external environment. \#\#\# condition:
-
+command string to some external environment.
+\item[condition]
 A specific event, or state, which can be trapped by CALL ON or SIGNAL
 ON.
-
-\hypertarget{configuration}{%
-\subsection{configuration:}\label{configuration}}
-
+\item[configuration]
 Any data-processing system, operating system and software used to
 operate a language processor.
-
-\hypertarget{conforming-language-processor}{%
-\subsection{conforming language
-processor:}\label{conforming-language-processor}}
-
+\item[conforming language processor]
 A language processor which obeys all the provisions of this standard.
-
-\hypertarget{construct}{%
-\subsection{construct:}\label{construct}}
-
+\item[construct]
 A named syntax grouping, for example ``expression'', ``do\_
 specification''.
-
-\hypertarget{default-error-stream}{%
-\subsection{default error stream:}\label{default-error-stream}}
-
+\item[default error stream]
 An output stream, determined by the configuration, on which error
 messages are written.
-
-\hypertarget{default-input-stream}{%
-\subsection{default input stream:}\label{default-input-stream}}
-
+\item[default input stream]
 An input stream having a name which is the null string. The use of this
 stream may be implied.
-
-\hypertarget{default-output-stream}{%
-\subsection{default output stream:}\label{default-output-stream}}
-
+\item[default output stream]
 An output stream having a name which is the null string. The use of this
 stream may be implied.
-
-\hypertarget{direct-symbol}{%
-\subsection{direct symbol:}\label{direct-symbol}}
-
+\item[direct symbol]
 A symbol which, without any modification, names a variable in a variable
 pool.
-
-\hypertarget{directive}{%
-\subsection{directive:}\label{directive}}
-
+\item[directive]
 Clauses which begin with two colons are directives. Directives are not
@@
 also be written with the two colons implied.
-
-\hypertarget{dropped}{%
-\subsection{dropped:}\label{dropped}}
-
+\item[dropped]
 A symbol which is in an unitialized state, as opposed to having had a
@@
 pool have an attribute of `dropped' or `not-dropped'.
-
-\hypertarget{encoding}{%
-\subsection{encoding:}\label{encoding}}
-
+\item[encoding]
 The relation between a character string anda corresponding number. The
 encoding of character strings is determined by the configuration.
-
-\hypertarget{end-of-line}{%
-\subsection{end-of-line:}\label{end-of-line}}
-
+\item[end-of-line]
 An event that occurs during the scanning of a source program. Normally
@@
 source program.
-
-\hypertarget{environment}{%
-\subsection{environment:}\label{environment}}
-
+\item[environment]
 The context in which a command may be executed. This is comprised of the
@@
 command.
-
-\hypertarget{environment-name}{%
-\subsection{environment name:}\label{environment-name}}
-
+\item[environment name]
 The name of an external procedure or process that can execute commands.
@@
 externally but then alterable by using the ADDRESS instruction.
-
-\hypertarget{error-number}{%
-\subsection{error number:}\label{error-number}}
-
+\item[error number]
 A number which identifies a particular situation which has occurred
@@
 defined by this standard.
-
-\hypertarget{exposed}{%
-\subsection{exposed:}\label{exposed}}
-
+\item[exposed]
 Normally, a symbol refers to a variable in the most recently established
@@
 an exposed variable.
-
-\hypertarget{expression}{%
-\subsection{expression:}\label{expression}}
-
+\item[expression]
 The most general of the constructs which can be evaluated to produce a
 single string value.
-
-\hypertarget{external-data-queue}{%
-\subsection{external data queue:}\label{external-data-queue}}
-
+\item[external data queue]
 A queue of strings that is external to REXX programs in that other
-programs may have access
-
-to the queue whenever REXX relinquishes control to some other program.
-
-\hypertarget{external-routine}{%
-\subsection{external routine:}\label{external-routine}}
-
+programs may have access to the queue whenever REXX relinquishes control
+to some other program.
+\item[external routine]
 A function or subroutine that is neither built-in nor in the same
 program as the CALL instruction or function call that invokes it.
-
-\hypertarget{external-variable-pool}{%
-\subsection{external variable pool:}\label{external-variable-pool}}
-
+\item[external variable pool]
 A named variable pool supplied by the configuration which can be
 accessed by the VALUE built-in function.
-
-\hypertarget{function}{%
-\subsection{function:}\label{function}}
-
+\item[function]
 Some processing which can be invoked by name and will produce a result.
@@
 provided by the configuration (see n).
-
-\hypertarget{identifier}{%
-\subsection{identifier:}\label{identifier}}
-
+\item[identifier]
 The name of a construct.
-
-\hypertarget{implicit-variable}{%
-\subsection{implicit variable:}\label{implicit-variable}}
-
+\item[implicit variable]
 A tailed variable which is in a variable pool solely as a result of an
@@
 `implicit' or `not-implicit'.
-
-\hypertarget{instruction}{%
-\subsection{instruction:}\label{instruction}}
-
+\item[instruction]
 One or more clauses that describe some course of action to be taken by
 the language processor.
-
-\hypertarget{internal-routine}{%
-\subsection{internal routine:}\label{internal-routine}}
-
+\item[internal routine]
 A function or subroutine that is in the same program as the CALL
 instruction or function call that invokes it.
-
-\hypertarget{keyword}{%
-\subsection{keyword:}\label{keyword}}
-
+\item[keyword]
 This standard specifies special meaning for some tokens which consist of
@@
 Such tokens, in these contexts, are keywords.
-
-\hypertarget{label}{%
-\subsection{label:}\label{label}}
-
+\item[label]
 A clause that consists of a single symbol or a literal followed by a
 colon.
-
-\hypertarget{language-processor}{%
-\subsection{language processor:}\label{language-processor}}
-
+\item[language processor]
 Compiler, translator or interpreter working in combination with a
 configuration.
-
-\hypertarget{notation-function}{%
-\subsection{notation function:}\label{notation-function}}
-
+\item[notation function]
 A function with the sole purpose of providing a notation for describing
@@
 function.
-
-\hypertarget{null-clause}{%
-\subsection{null clause:}\label{null-clause}}
-
+\item[null clause]
 A clause which has no tokens.
-
-\hypertarget{null-string}{%
-\subsection{null string:}\label{null-string}}
-
+\item[null string]
 A character string with no characters, that is, a string of length zero.
-
-\hypertarget{production}{%
-\subsection{production:}\label{production}}
-
+\item[production]
 The definition of a construct, in Backus-Naur form.
-
-\hypertarget{return-code}{%
-\subsection{return code:}\label{return-code}}
-
+\item[return code]
 A string that conveys some information about the command that has been
@@
 command but can also be used to represent other information.
-
-\hypertarget{routine}{%
-\subsection{routine:}\label{routine}}
-
+\item[routine]
 Some processing which can be invoked by name.
-
-\hypertarget{state-variable}{%
-\subsection{state variable:}\label{state-variable}}
-
+\item[state variable]
 A component of the state of progress in processing a program, described
@@
 access a state variable.
-
-\hypertarget{stem}{%
-\subsection{stem:}\label{stem}}
-
+\item[stem]
 If a symbol naming a variable contains a period which is not the first
@@
 is the stem.
-
-\hypertarget{stream}{%
-\subsection{stream:}\label{stream}}
-
+\item[stream]
 Named streams are used as the sources of input and the targets of
@@
 example the input from a keyboard.
-
-\hypertarget{string}{%
-\subsection{string:}\label{string}}
-
+\item[string]
 For many operations the unit of data is a string. It may, or may not, be
@@
 individually.
-
-\hypertarget{subcode}{%
-\subsection{subcode:}\label{subcode}}
-
+\item[subcode]
 The decimal part of an error number.
-
-\hypertarget{subroutine}{%
-\subsection{subroutine:}\label{subroutine}}
-
+\item[subroutine]
 An internal, built-in, or external routine that may or may not return a
@@
 result string the subroutine can also be invoked by a function call, in
-which case it is being called as a function. \#\#\# symbol:
-
+which case it is being called as a function.
+\item[symbol]
 A sequence of characters used as a name, see nnn. Symbols are used to
 name variables, functions, etc.
-
-\hypertarget{tailed-name}{%
-\subsection{tailed name:}\label{tailed-name}}
-
+\item[tailed name]
 The names in a variable pool have an attribute of `tailed' or
@@
 of a symbol, the part that follows a stem.
-
-\hypertarget{token}{%
-\subsection{token:}\label{token}}
-
+\item[token]
 The unit of low-level syntax from which high-level constructs are built.
 Tokens are literal strings, symbols, operators, or special characters.
-
-\hypertarget{trace}{%
-\subsection{trace:}\label{trace}}
-
+\item[trace]
 A description of some or all of the clauses of a program, produced as
 each is executed.
-
-\hypertarget{trap}{%
-\subsection{trap:}\label{trap}}
-
+\item[trap]
 A function provided by the user which replaces or augments some normal
 function of the language processor.
+\item[variable pool]
+A collection of the names of variables and their associated values.
+\end{description}
 
-\hypertarget{variable-pool}{%
-\subsection{variable pool:}\label{variable-pool}}
+\section{Document notation}\label{document-notation}
 
-A collection of the names of variables and their associated values.
-
-\hypertarget{document-notation}{%
-\section{Document notation}\label{document-notation}}
-
-\hypertarget{rexx-code}{%
-\subsection{Rexx Code}\label{rexx-code}}
+\subsection{Rexx Code}\label{rexx-code}
 
@@
 
-\hypertarget{italics}{%
-\subsection{Italics}\label{italics}}
+\subsection{Italics}\label{italics}
 
 Throughout this standard, except in Rexx code, references to the
-constructs defined in section nnn are italicized.
+constructs defined in section nnn are \emph{italicized}.

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/evaluation.tex
work in standard book production process including readme.md
upd

@@
-\hypertarget{evaluation}{%
-\chapter{Evaluation}\label{evaluation}}
+%preprocessed texin
+\chapter{Evaluation}\label{evaluation}
 
@@
 
-\hypertarget{variables}{%
-\section{Variables}\label{variables}}
+\section{Variables}\label{variables}
 
@@
 
-\hypertarget{var_empty}{%
-\subsection{Var\_Empty}\label{var_empty}}
+\subsection{Var\_Empty}\label{var_empty}
 
@@
 
-\hypertarget{var-set}{%
-\subsection{Var Set}\label{var-set}}
+\subsection{Var Set}\label{var-set}
 
@@
 
-\hypertarget{var_value}{%
-\subsection{Var\_Value}\label{var_value}}
+\subsection{Var\_Value}\label{var_value}
 
@@
 
-\hypertarget{var_drop}{%
-\subsection{Var\_Drop}\label{var_drop}}
+\subsection{Var\_Drop}\label{var_drop}
 
@@
 
-\hypertarget{var_expose}{%
-\subsection{Var\_Expose}\label{var_expose}}
+\subsection{Var\_Expose}\label{var_expose}
 
@@
 
-\hypertarget{var_reset}{%
-\subsection{Var\_Reset}\label{var_reset}}
+\subsection{Var\_Reset}\label{var_reset}
 
@@
 
-\hypertarget{symbols}{%
-\section{Symbols}\label{symbols}}
+\section{Symbols}\label{symbols}
 
@@
 
-\hypertarget{value-of-a-variable}{%
-\section{Value of a variable}\label{value-of-a-variable}}
+\section{Value of a variable}\label{value-of-a-variable}
 
@@
 
-\hypertarget{value-of-a-reserved-symbol}{%
 \subsection{Value of a reserved
-symbol}\label{value-of-a-reserved-symbol}}
+symbol}\label{value-of-a-reserved-symbol}
 
@@
 
-\hypertarget{expressions-and-operators}{%
-\section{Expressions and operators}\label{expressions-and-operators}}
+\section{Expressions and operators}\label{expressions-and-operators}
 
@@
 
-\hypertarget{the-value-of-a-term}{%
-\subsection{The value of a term}\label{the-value-of-a-term}}
+\subsection{The value of a term}\label{the-value-of-a-term}
 
-See nnn for the syntax of a ferm.
+See nnn for the syntax of a term.
 
@@
 
-\hypertarget{the-value-of-a-prefix_expression}{%
 \subsection{The value of a
-prefix\_expression}\label{the-value-of-a-prefix_expression}}
+prefix\_expression}\label{the-value-of-a-prefix_expression}
 
@@
 
-\hypertarget{the-value-of-a-power_expression}{%
 \subsection{The value of a
-power\_expression}\label{the-value-of-a-power_expression}}
+power\_expression}\label{the-value-of-a-power_expression}
 
@@
 
-\begin{verbatim}
-if datatype(lhs)\=='NUM' then call #Raise 'SYNTAX',41.1,lhs,'**'
-
-if \datatype(rhs,'W') then call #Raise 'SYNTAX',26.1,rhs,'**'
-\end{verbatim}
-
-The value of the power\_expression is
+\lstinputlisting[language=rexx,label=evaluation-datatype.rexx,caption=evaluation-datatype.rexx]{evaluation-datatype.rexx}
+power\_expression is
 
@@
 
-\hypertarget{the-value-of-a-multiplication}{%
 \subsection{The value of a
-multiplication}\label{the-value-of-a-multiplication}}
+multiplication}\label{the-value-of-a-multiplication}
 
@@
 
-\hypertarget{the-value-of-an-addition}{%
-\subsection{The value of an addition}\label{the-value-of-an-addition}}
+\subsection{The value of an addition}\label{the-value-of-an-addition}
 
@@
 
-\hypertarget{the-value-of-a-concatenation}{%
 \subsection{The value of a
-concatenation}\label{the-value-of-a-concatenation}}
+concatenation}\label{the-value-of-a-concatenation}
 
@@
 
-\hypertarget{the-value-of-a-comparison}{%
-\subsection{The value of a comparison}\label{the-value-of-a-comparison}}
+\subsection{The value of a comparison}\label{the-value-of-a-comparison}
 
@@
 then the variable \#Test is set as follows:
+\lstinputlisting[language=rexx,label=evaluation-comparison.rexx,caption=evaluation-comparison.rexx]{evaluation-comparison.rexx}
 
-\begin{verbatim}
-if datatype(lhs)\== 'NUM' | datatype(rhs)\== 'NUM' then do
-/* Non-numeric non-strict comparison */
-lhs=strip(lhs, 'B', ' ') /* ExtraBlanks not stripped */
-rhs=strip(rhs, 'B', ' ')
-
-if length(lhs)>length(rhs) then rhs=left (rhs, length (lhs) )
-else lhs=left (lhs, length (rhs) )
-if lhs>>rhs then #Test='G'
-else if lhs<<rhs then #Test='L'
-else #Test='E'
-
-end
-else do /* Numeric comparison */
-if left(-lhs,1) == '-' & left(+rhs,1) \== '-' then #Test='G!
-else if left(-rhs,1) == '-' & left(+lhs,1) \== '-' then #Test='L'
-else do
-Difference=lhs - rhs /* Will never raise an arithmetic condition. */
-if Difference > 0 then #Test='G'
-else if Difference < 0 then #Test='L'
-else #Test='E'
-end
-end
-The value of #Test, in conjunction with the operator in the comparison, determines the value of the
-comparison.
-The value of the comparison is '1' if
-- #Test is 'E' and the operator is one of '="", '=="", '>=', <=", ‘\>', '\<', 'p>=', '<<=', \>>', or <<)
-
-- #Test is 'G' and the operator is one of '>', '>=", ‘\<', ‘\=', '<>', '><', Nes", ‘>>! ‘p>’, or <<")
-- #Test is 'L' and the operator is one of '<', <=", \>', \=', '<>', '><', \==', '<<', *<<=', or \>>'.
-In all other cases the value of the comparison is '0'.
-
-If the comparison is not a concatenation then:
-if #Tracing.#Level == 'I' then call #Trace '>0O>'
-\end{verbatim}
-
-\hypertarget{the-value-of-an-and_expression}{%
 \subsection{The value of an
-and\_expression}\label{the-value-of-an-and_expression}}
+and\_expression}\label{the-value-of-an-and_expression}
 
@@
 
-\hypertarget{the-value-of-an-expression}{%
 \subsection{The value of an
-expression}\label{the-value-of-an-expression}}
+expression}\label{the-value-of-an-expression}
 
@@
 
-\hypertarget{arithmetic-operations}{%
-\subsection{Arithmetic operations}\label{arithmetic-operations}}
+\subsection{Arithmetic operations}\label{arithmetic-operations}
 
@@
 high setting of numeric digits to avoid exponential notation.
+\lstinputlisting[language=rexx,label=evaluation\_arithmetic.rexx,caption=evaluation\_arithmetic.rexx]{evaluation\_arithmetic.rexx}
 
-\begin{verbatim}
-ArithoOp:
+\section{Functions}\label{functions}
 
-arg Numberl, Operator, Number2
-/* The Operator will be applied to Numberl and Number2 under the numeric
-settings #Digits.#Level, #Form.#Level, #Fuzz.#Level */
-
-/* The result is the result of the operation, or the raising of a 'SYNTAX' or
-'LOSTDIGITS' condition. */
-
-/* Variables with digit 1 in their names refer to the first argument of the
-operation. Variables with digit 2 refer to the second argument. Variables
-with digit 3 refer to the result. */
-
-/* The quotations and page numbers are from the first reference in
-Annex C of this standard. */
-
-/* The operands are prepared first. (Page 130) Function Prepare does this,
-separating sign, mantissa and exponent. */
-
-v = Prepare (Numberl, #Digits.#Level)
-parse var v Signl Mantissal Exponentl
-
-v = Prepare (Number2, #Digits.#Level)
-parse var v Sign2 Mantissa2 Exponent2
-
-/* The calculation depends on the operator. The routines set Sign3
-Mantissa3 and Exponent3. */
-
-Comparator = ''
-
-select
-when Operator == '*' then call Multiply
-
-when Operator
-when Operator
-
-' then call DivType
-*' then call Power
-when Operator %' then call DivType
-when Operator '//" then call DivType
-otherwise call AddSubComp
-
-end
-
-call PostOp /* Assembles Number3 */
-
-if Comparator \== '' then do
-/* Comparison requires the result of subtraction made into a logical */
-/* value. */
-t = '0'
-select
-when left (Number3,1) == '-' then
-if wordpos(Comparator,'< <= <> >< \= \>') > 0 then t = '1'
-when Number3 \== '0' then
-if wordpos(Comparator,'> >= <> >< \= \<') > 0 then t = '1'
-otherwise
-if wordpos(Comparator,'>= = =< \< \>') > 0 then t = '1!
-end
-Number3 = t
-end
-return Number3 /* From ArithOp */
-/* Activity before every operation: */
-Prepare: /* Returns Sign Mantissa and Exponent */
-/* Preparation of operands, Page 130 */
-/* ",...terms being operated upon have leading zeros removed (noting the
-
-position of any decimal point, and leaving just one zero if all the digits in
-
-the number are zeros) and are then truncated to DIGITS+1 significant digits
-(if necessary)..." */
-
-arg Number, Digits
-
-/* Blanks are not significant. */
-
-/* The exponent is separated */
-
-parse upper value space(Number,0) with Mantissa 'E' Exponent
-if Exponent == '' then Exponent = '0'
-
-/* The sign is separated and made explicit. */
-
-Sign = '+' /* By default */
-if left(Mantissa,1) == '-' then Sign = '-'
-if verify (left (Mantissa,1),'+-') = 0 then Mantissa = substr(Mantissa,2)
-
-/* Make the decimal point implicit; remove any actual Point from the
-Mantissa. */
-Pp = pos('.',Mantissa)
-if p > 0 then Mantissa = delstr(Mantissa,p,1)
-else p = 1+length (Mantissa)
-
-/* Drop the leading zeros */
-do q = 1 to length(Mantissa) - 1
-
-if substr(Mantissa,g,1) \== '0' then leave
-p=p-l
-end q
-
-Mantissa = substr(Mantissa,q)
-
-/* Detect if Mantissa suggests more significant digits than DIGITS
-caters for. */
-do j = Digits+1 to length (Mantissa)
-if substr(Mantissa,j,1) \== '0' then call #Raise 'LOSTDIGITS', Number
-end j
-
-/* Combine exponent with decimal point position, Page 127 */
-/* "Exponential notation means that the number includes a power of ten
-
-following an 'E' that indicates how the decimal point will be shifted. Thus
-
-4E9 is just a shorthand way of writing 4000000000 " */
-
-/* Adjust the exponent so that decimal point would be at right of
-the Mantissa. */
-Exponent = Exponent - (length(Mantissa) - p + 1)
-
-/* Truncate if necessary */
-
-t = length(Mantissa) - (Digits+1)
-
-if t > 0 then do
-Exponent = Exponent + t
-Mantissa = left (Mantissa,Digits+1)
-end
-
-if Mantissa == '0' then Exponent = 0
-
-return Sign Mantissa Exponent
-
-/* Activity after every operation. */
-/* The parts of the value are composed into a single string, Number3. */
-PostOp:
-
-/* Page 130 */
-/* 'traditional' rounding */
-t = length(Mantissa3) - #Digits.#Level
-if t > 0 then do
-/* 'traditional' rounding */
-Mantissa3 = left (Mantissa3,#Digits.#Level+1) + 5
-if length(Mantissa3) > #Digits.#Level+1 then
-/* There was 'carry' */
-
-Exponent3 = Exponent3 + 1
-
-Mantissa3 = left (Mantissa3,#Digits.#Level)
-
-Exponent3 = Exponent3 + t
-
-end
-/* "A result of zero is always expressed as a single character '0' "*/
-if verify (Mantissa3,'0') = 0 then Number3 = '0'
-else do
-
-if Operator == '/' | Operator == '**' then do
-
-/* Page 130 "For division, insignificant trailing zeros are removed
-after rounding." */
-
-/* Page 133 "... insignificant trailing zeros are removed." */
-do q = length(Mantissa3) by -1 to 2
-if substr(Mantissa3,q,1) \== '0' then leave
-Exponent3 = Exponent3 + 1
-end q
-Mantissa3 = substr(Mantissa3,1,q)
-end
-if Floating() == 'E' then do /* Exponential format */
-
-Exponent3 = Exponent3 + (length(Mantissa3)-1)
-
-/* Page 136 "Engineering notation causes powers of ten to be expressed as a
-
-multiple of 3 - the integer part may therefore range from 1 through
-999." */
-
-g=l
-
-if #Form.#Level == 'E' then do
-
-/* Adjustment to make exponent a multiple of 3 */
-g = Exponent3//3 /* Recursively using ArithOp as
-an external routine. */
-if g < 0 then g =g+ 3
-Exponent3 = Exponent3 - g
-geaqgitl
-if length(Mantissa3) < g then
-Mantissa3 = left (Mantissa3,g,'0')
-end /* Engineering */
-
-/* Exact check on the exponent. */
-if Exponent3 > #Limit ExponentDigits then
-
-call #Raise 'SYNTAX', 42.1, Numberl, Operator, Number2
-if -#Limit ExponentDigits > Exponent3 then
-
-call #Raise 'SYNTAX', 42.2, Numberl, Operator, Number2
-
-/* Insert any decimal [point. */
-
-if length(Mantissa3) \= g then Mantissa3 = insert('.',Mantissa3,g)
-/* Insert the E */
-if Exponent3 >= 0 then Number3
-else Number3
-end /* Exponent format */
-else do /* 'pure number' notation */
-p = length(Mantissa3) + Exponent3 /* Position of the point within
-Mantissa */
-/* Add extra zeros needed on the left of the point. */
-if p < 1 then do
-Mantissa3 = copies('0',1 - p)| |Mantissa3
-p=il
-end
-/* Add needed zeros on the right. */
-if p > length(Mantissa3) then
-Mantissa3 = Mantissa3||copies('0',p-length (Mantissa3) )
-/* Format with decimal point. */
-Number3 = Mantissa3
-
-Mantissa3'E+'Exponent3
-Mantissa3'E'Exponent3
-
-if p < length(Number3) then Number3 = insert('.',Mantissa3,p)
-else Number3 = Mantissa3
-end /* pure */
-if Sign3 == '-' then Number3 = '-'Number3
-end /* Non-Zero */
-return
-/* This tests whether exponential notation is needed. */
-Floating:
-/* The rule in the reference has been improved upon. */
-Ct = ter
-if Exponent3+length(Mantissa3) > #Digits.#Level then t = 'E'
-if length(Mantissa3) + Exponent3 < -5 then t = 'E'
-return t
-/* Add, Subtract and Compare. */
-
-AddSubComp: /* Page 130 */
-/* This routine is used for comparisons since comparison is
-defined in terms of subtraction. Page 134 */
-/* "Numeric comparison is affected by subtracting the two numbers (calculating
-
-the difference) and then comparing the result with '0'." */
-NowDigits = #Digits.#Level
-if Operator \=='+' & Operator \== '-' then do
-
-Comparator = Operator
-
-/* Page 135 "The effect of NUMERIC FUZZ is to temporarily reduce the value
-of NUMERIC DIGITS by the NUMERIC FUZZ value for each numeric comparison" */
-NowDigits = NowDigits - #Fuzz.#Level
-
-end
-/* Page 130 "If either number is zero then the other number ... is used as
-the result (with sign adjustment as appropriate). */
-if Mantissa2 == '0' then do /* Result is the lst operand */
-Sign3=Signl; Mantissa3 = Mantissal; Exponent3 = Exponentl
-return ''
-end
-
-if Mantissal
-Sign3 = Sign
-
-== '0' then do /* Result is the 2nd operand */
-2
-if Operator \
-
-; Mantissa3 = Mantissa2; Exponent3 = Exponent2
-== '+' then if Sign3 = '+' then Sign3 rat
-else Sign3
-
-bat
-return ''
-end
-
-/* The numbers may need to be shifted into alignment. */
-
-/* Change to make the exponent to reflect a decimal point on the left,
-
-so that right truncation/extension of mantissa doesn't alter exponent. */
-Exponentl = Exponentl + length (Mantissal1)
-Exponent2 = Exponent2 + length (Mantissa2)
-
-/* Deduce the implied zeros on the left to provide alignment. */
-
-Aligni = 0
-
-Align2 = Exponentl - Exponent2
-
-if Align2 > 0 then do /* Arg 1 provides a more significant digit */
-Align2 = min(Align2,NowDigits+1) /* No point in shifting further. */
-/* Shift to give Arg2 the same exponent as Argl */
-
-Mantissa2 = copies('0',Align2) || Mantissa2
-Exponent2 = Exponentl
-end
-
-if Align2 < 0 then do /* Arg 2 provides a more significant digit */
-/* Shift to give Argl the same exponent as Arg2 */
-
-Aligni = -Align2
-
-Alignl = min(Align1l,NowDigits+1) /* No point in shifting further. */
-Align2 = 0
-
-Mantissal = copies('0',Alignl) || Mantissal
-
-Exponentl = Exponent2
-
-end
-
-/* Maximum working digits is NowDigits+1. Footnote 41. */
-
-SigDigits
-SigDigits
-
-max (length (Mantissal) , length (Mantissaz2) )
-min (SigDigits,NowDigits+1)
-
-/* Extend a mantissa with right zeros, if necessary. */
-Mantissal = left (Mantissal,SigDigits,'0')
-
-Mantissa2 = left (Mantissa2,SigDigits,'0')
-
-/* The exponents are adjusted so that
-
-the working numbers are integers, ie decimal point on the right. */
-
-Exponent3 = Exponentl-SigDigits
-Exponentl = Exponent3
-Exponent2 = Exponent3
-if Operator = '+' then
-Mantissa3 = (Signl || Mantissal) + (Sign2 || Mantissa2)
-
-else Mantissa3 (Signl || Mantigsal) - (Sign2 || Mantissa2)
-
-/* Separate the sign */
-if Mantissa3 < 0 then do
-
-Sign3 = '-'
-Mantissa3 = substr (Mantissa3,2)
-end
-
-else Sign3 = '+'
-
-/* "The result is then rounded to NUMERIC DIGITS digits if necessary,
-taking into account any extra (carry) digit on the left after addition,
-but otherwise counting from the position corresponding to the most
-significant digit of the terms being added or subtracted." */
-
-if length(Mantissa3) > SigDigits then SigDigits = SigDigits+l
-d = SigDigits - NowDigits /* Digits to drop. */
-if d <= 0 then return
-t = length(Mantissa3) - d /* Digits to keep. */
-/* Page 130. "values of 5 through 9 are rounded up, values of 0 through 4 are
-rounded down." */
-if t > 0 then do
-/* 'traditional' rounding */
-Mantissa3 = left(Mantissa3, t +1) +5
-if length(Mantissa3) > t+1 then
-/* There was 'carry' */
-/* Keep the extra digit unless it takes us over the limit. */
-if t < NowDigits then t = t+l
-else Exponent3 = Exponent3+1
-Mantissa3 = left (Mantissa3,t)
-Exponent3 = Exponent3 + d
-end /* Rounding */
-else Mantissa3 = '0'
-return /* From AddSubComp */
-
-/* Multiply operation: */
-
-Multiply: /* p 131 */
-
-/* Note the sign of the result */
-
-if Signl == Sign2 then Sign3 = '+'
-else Sign3 = '-'
-/* Note the exponent */
-Exponent3 = Exponentl + Exponent2
-if Mantissal == '0' then do
-Mantissa3 = '0'
-return
-end
-/* Multiply the Mantissas */
-Mantissa3 = ''
-
-do q=1 to length (Mantissa2)
-Mantissa3 = Mantissa3'0'
-do substr(Mantissa2,q,1)
-Mantissa3 = Mantissa3 + Mantissal
-end
-end q
-return /* From Multiply */
-
-/* Types of Division: */
-
-DivType: /* p 131 */
-/* Check for divide-by-zero */
-
-if Mantissa2 == '0' then call #Raise 'SYNTAX',
-
-/* Note the exponent of the result */
-Exponent3 = Exponentl - Exponent2
-
-/* Compute (one less than) how many digits will be in the integer
-
-part of the result. */
-
-IntDigits = length(Mantissal) - Length(Mantissa2) + Exponent3
-/* In some cases, the result is known to be zero.
-if Mantigsal = 0 | (IntDigits < 0 & Operator
-
-Mantissa3 = 0
-Sign3 = '+'
-Exponent3 = 0
-return
-end
-/* In some cases, the result is known to be to be the first argument.
-if IntDigits < 0 & Operator == '//' then do
-Mantissa3 = Mantissal
-Sign3 = Signl
-Exponent3 = Exponentl
-return
-end
-/* Note the sign of the result. */
-if Signl == Sign2 then Sign3 = '+'
-else Sign3 = '-'
-
-/* Make Mantissal at least as large as Mantissa2 so Mantissa2 can be
-subtracted without causing leading zero to result. Page 131 */
-
-az 0
-do while Mantissa2 > Mantissal
-
-Mantissal = Mantissal'0'
-Exponent3 = Exponent3 - 1
-aztadtl
-end
-/* Traditional divide */
-Mantissa3 = ''
-
-/* Subtract from part of Mantissal that has length of Mantissa2 */
-
-left (Mantissal,length(Mantissa2) )
-substr (Mantissal, length (Mantissa2)+1)
-o forever
-
-x
-Y
-d
-
-/* Develop a single digit in z by repeated subtraction.
-
-ze=O0
-do forever
-xX = kK - Mantissa2
-if left(x,1) == '-' then leave
-Zeze¢t+tl
-end
-
-x = x + Mantissa2 /* Recover from over-subtraction */
-/* The digit becomes part of the result */
-
-Mantissa3 = Mantissa3 || z
-
-if Mantissa3 == '0' then Mantissa3 = '' /* A single leading
-
-##
-
-*f
-'%') then do
-
-*/
-zero can happen. */
-/* x||y is the current residue */
-if y == '' then if x = 0 then leave /* Remainder is zero */
-
-if length(Mantissa3) > #Digits.#Level then leave /* Enough digits
-in the result */
-
-/* Check type of division */
-if Operator \== '/' then do
-if IntDigits = 0 then leave
-IntDigits = IntDigits - 1
-end
-/* Prepare for next digit */
-/* Digits come from y, until that is exhausted. */
-/* When y is exhausted an extra zero is added to Mantissal */
-if y == '' then do
-y = ror
-Exponent3 = Exponent3 - 1
-aztadtl
-end
-xX = xX | | left (y,1)
-y = substr(y,2)
-end /* Iterate for next digit. */
-
-Remainder = x || y
-Exponent3 = Exponent3 + length(y) /* The loop may have been left early.
-/* Leading zeros are taken off the Remainder. */
-do while length(Remainder) > 1 & Left (Remainder,1) == '0'
-Remainder = substr (Remainder, 2)
-end
-if Operator \== '/' then do
-/* Check whether % would fail, even if operation is // */
-
-if Floating() 'E' then do
-if Operator '%' then MsgNum
-else MsgNum
-
-/* Page 133. % could fail by needing exponential notation */
-
-26.11
-26.12
-
-call #Raise 'SYNTAX', MsgNum, Numberl , Number2, #Digits.#Level
-
-end
-end
-if Operator == '//' then do
-/* We need the remainder */
-Sign3 = Signl
-
-Mantissa3 = Remainder
-Exponent3 = Exponentl - a
-end
-
-return /* From DivType */
-
-/* The Power operation: */
-
-Power: /* page 132 */
-/* The second argument should be an integer */
-if \WholeNumber2() then call #Raise 'SYNTAX', 26.8, Number2
-/* Lhs to power zero is always 1 */
-if Mantissa2 == '0' then do
-Sign3 = '+'
-Mantissa3
-Exponent3
-return
-end
-
-i
-ror
-
-/* Pages 132-133 The Power algorithm */
-Rhs = left (Mantissa2,length(Mantissa2)+Exponent2,'0')/* Explicit
-integer form */
-L length (Rhs)
-b X2B(D2X(Rhs)) /* Makes Rhs in binary notation */
-/* Ignore initial zeros */
-do q=l1byl
-if substr(b,q,1) \== '0' then leave
-end q
-ael
-do forever
-/* Page 133 "Using a precision of DIGITS+L+1" */
-if substr(b,q,1) == '1' then do
-a = Recursion('*',Signl || Mantissal'E'Exponent1)
-
-*/
-if left(a,2) == 'MN' then signal PowerFailed
-end
-
-/* Check for finished */
-
-if q = length(b) then leave
-
-/* Square a */
-
-a = Recursion('*',a)
-if left(a,2) == 'MN' then signal PowerFailed
-q=qrtil
-end
-/* Divide into one for negative power */
-if Sign2 == '-' then do
-Sign2 = '+'
-a = Recursion('/')
-if left(a,2) == 'MN' then signal PowerFailed
-end
-
-/* Split the value up so that PostOp can put it together with rounding */
-Parse value Prepare(a,#Digits.#Level+L+1) with Sign3 Mantissa3 Exponent3
-return
-
-PowerFailed:
-/* Distinquish overflow and underflow */
-ReWas = substr(a,4)
-if Sign2 = '-' then if ReWas == '42.1' then RcWas
-else RcWas
-call #Raise 'SYNTAX', RcWas, Numberl, '**', Number2
-/* No return */
-
-"42.2!
-"42.1!
-
-WholeNumber2:
-numeric digits Digits
-if #Form.#Level == 'S' then numeric form scientific
-
-else numeric form engineering
-return datatype (Number2, 'W')
-
-Recursion: /* Called only from '**! */
-numeric digits #Digits.#Level + L + 1
-signal on syntax name Overflowed
-/* Uses ArithOp again under new numeric settings. */
-if arg(1) == '/' then return 1l/a
-else return a * arg(2)
-Over flowed:
-return 'MN '.MN
-\end{verbatim}
-
-\hypertarget{functions}{%
-\section{Functions}\label{functions}}
-
-\hypertarget{invocation}{%
-\subsection{Invocation}\label{invocation}}
+\subsection{Invocation}\label{invocation}
 
@@
 
-\hypertarget{evaluation-of-arguments}{%
-\subsection{Evaluation of arguments}\label{evaluation-of-arguments}}
+\subsection{Evaluation of arguments}\label{evaluation-of-arguments}
 
@@
 
-\hypertarget{the-value-of-a-label}{%
-\subsection{The value of a label}\label{the-value-of-a-label}}
+\subsection{The value of a label}\label{the-value-of-a-label}
 
@@
 
-\hypertarget{the-value-of-a-function}{%
-\subsection{The value of a function}\label{the-value-of-a-function}}
+\subsection{The value of a function}\label{the-value-of-a-function}
 
@@
 
-\hypertarget{the-value-of-a-method}{%
-\subsection{The value of a method}\label{the-value-of-a-method}}
+\subsection{The value of a method}\label{the-value-of-a-method}
 
@@
 
-\hypertarget{the-value-of-a-message-term}{%
 \subsection{The value of a message
-term}\label{the-value-of-a-message-term}}
+term}\label{the-value-of-a-message-term}
 
@@
 method invocation which the subject message term is running under. */
+\lstinputlisting[language=rexx,label=evaluation-selectmethod,caption=evaluation-selectmethod]{evaluation-selectmethod}
 
-\begin{verbatim}
-SelectMethod:
-
-/* If symbol given, receiver must be self. */
-if arg(3,'E') then if arg(1)\==#Self then signal error /* syntax number? */
-
-t arg(2) /* Will have been uppercased, unless a literal. */
-
-x arg(1) /* Cursor through places to look for the method. */
-
-Mixing 1 /* Off for potential mixins ignored because symbol given. */
-Mixins -array~new /* to note any Mixins involved. */
-
-/* Look in the method table of the object, if no 'symbol' given. */
-if arg(3,'E') then do
-Mixing = 0
-
-end
-else do
-m = x~#MethodTable[t]
-if m \== .nil then return m
-end
-
-do until x==.object
-/* Follow the class hierarchy. */
-x = x-class
-/* Note any mixins for later reference. */
-Mix = x~Inherited /* An array, ordered as the directive left-to-right. */
-
-if Mix \== .nil then /* Append to the record. */
-do j=1 to Mix~dimension (1)
-Mixins [Mixins~dimension(1)+1] = Mix[j]
-end
-
-if Mixing do
-/* Consider mixins only for superclasses of 'symbol'. */
-do j=1 to Mixins~dimension (1)
-/* Look at the baseclass of each. */
-/* That is closest superclass not a mixin. */
-s = Mixins[j]~class
-do while s~Mixin /* Assert stop at .object if not before. */
-s=s~class
-end
-if s==x then do
-m=Mixins [j]~#MethodTable[t]
-if m \== .nil then return m
-end
-end j
-end /* Mixing */
-if arg(3,'E') then if arg(3)==x then do
-Mixing=1
-end
-if Mixing do
-/* Consider non-Mixins */
-m= x-#InstanceMethodTable[t]
-
-if m \== .nil then return m
-end
-
-x=x~superclass
-
-end
-
-/* Try for UNKNOWN instead */
-if t == 'UNKNOWN' then return .nil
-if \arg(3,'E') then return SelectMethod arg(1),'UNKNOWN'
-else return SelectMethod arg(1),'UNKNOWN',arg(3)
-\end{verbatim}
-
-\hypertarget{use-of-config_externalroutine}{%
 \subsection{Use of
-Config\_ExternalRoutine}\label{use-of-config_externalroutine}}
+Config\_ExternalRoutine}\label{use-of-config_externalroutine}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/instructions.tex
work in standard book production process including readme.md
upd

@@
-\hypertarget{instructions}{%
-\chapter{Instructions}\label{instructions}}
+%preprocessed texin
+\chapter{Instructions}\label{instructions}
 
@@
 
-\hypertarget{routine-initialization}{%
-\section{Routine initialization}\label{routine-initialization}}
+\section{Routine initialization}\label{routine-initialization}
 
@@
 
-\hypertarget{clause-initialization}{%
-\section{Clause initialization}\label{clause-initialization}}
+\section{Clause initialization}\label{clause-initialization}
 
@@
 
-\hypertarget{clause-termination}{%
-\section{Clause termination}\label{clause-termination}}
+\section{Clause termination}\label{clause-termination}
 
@@
 
-\hypertarget{instruction}{%
-\section{Instruction}\label{instruction}}
-
-\hypertarget{address}{%
-\subsection{ADDRESS}\label{address}}
+\section{Instruction}\label{instruction}
+
+\subsection{ADDRESS}\label{address}
 
@@
 
-\hypertarget{assignment}{%
-\subsection{Assignment}\label{assignment}}
+\subsection{Assignment}\label{assignment}
 
@@
 
-\hypertarget{call}{%
-\subsection{CALL}\label{call}}
+\subsection{CALL}\label{call}
 
@@
 
-\hypertarget{command-to-the-configuration}{%
 \subsection{Command to the
-configuration}\label{command-to-the-configuration}}
+configuration}\label{command-to-the-configuration}
 
@@
 
-\hypertarget{do}{%
-\subsection{DO}\label{do}}
+\subsection{DO}\label{do}
 
@@
 
-\hypertarget{do-loop-tracing}{%
-\subsection{DO loop tracing}\label{do-loop-tracing}}
+\subsection{DO loop tracing}\label{do-loop-tracing}
 
@@
 
-\hypertarget{drop}{%
-\subsection{DROP}\label{drop}}
+\subsection{DROP}\label{drop}
 
@@
 
-\hypertarget{exit}{%
-\subsection{EXIT}\label{exit}}
+\subsection{EXIT}\label{exit}
 
@@
 
-\hypertarget{expose}{%
-\subsection{EXPOSE}\label{expose}}
+\subsection{EXPOSE}\label{expose}
 
@@
 
-\hypertarget{forward}{%
-\subsection{FORWARD}\label{forward}}
+\subsection{FORWARD}\label{forward}
 
@@
 
-\hypertarget{guard}{%
-\subsection{GUARD}\label{guard}}
+\subsection{GUARD}\label{guard}
 
@@
 
-\hypertarget{interpret}{%
-\subsection{INTERPRET}\label{interpret}}
+\subsection{INTERPRET}\label{interpret}
 
@@
 
-\hypertarget{iterate}{%
-\subsection{ITERATE}\label{iterate}}
+\subsection{ITERATE}\label{iterate}
 
@@
 
-\hypertarget{execution-of-labels}{%
-\subsection{Execution of labels}\label{execution-of-labels}}
+\subsection{Execution of labels}\label{execution-of-labels}
 
@@
 
-\hypertarget{leave}{%
-\subsection{LEAVE}\label{leave}}
+\subsection{LEAVE}\label{leave}
 
@@
 
-\hypertarget{message-term}{%
-\subsection{Message term}\label{message-term}}
+\subsection{Message term}\label{message-term}
 
@@
 
-\hypertarget{loop}{%
-\subsection{LOOP}\label{loop}}
+\subsection{LOOP}\label{loop}
 
@@
 
-\hypertarget{nop}{%
-\subsection{NOP}\label{nop}}
+\subsection{NOP}\label{nop}
 
@@
 
-\hypertarget{numeric}{%
-\subsection{NUMERIC}\label{numeric}}
+\subsection{NUMERIC}\label{numeric}
 
@@
 
-\hypertarget{numeric-digits}{%
-\subsubsection{NUMERIC DIGITS}\label{numeric-digits}}
+\subsubsection{NUMERIC DIGITS}\label{numeric-digits}
 
@@
 
-\hypertarget{numeric-form}{%
-\subsubsection{NUMERIC FORM}\label{numeric-form}}
+\subsubsection{NUMERIC FORM}\label{numeric-form}
 
@@
 
-\hypertarget{numeric-fuzz}{%
-\subsubsection{NUMERIC FUZZ}\label{numeric-fuzz}}
+\subsubsection{NUMERIC FUZZ}\label{numeric-fuzz}
 
@@
 
-\hypertarget{options}{%
-\subsubsection{OPTIONS}\label{options}}
+\subsubsection{OPTIONS}\label{options}
 
@@
 
-\hypertarget{parse}{%
-\subsubsection{PARSE}\label{parse}}
+\subsubsection{PARSE}\label{parse}
 
@@
 
-\hypertarget{procedure}{%
-\subsection{PROCEDURE}\label{procedure}}
+\subsection{PROCEDURE}\label{procedure}
 
@@
 
-\hypertarget{pull}{%
-\subsection{PULL}\label{pull}}
+\subsection{PULL}\label{pull}
 
@@
 
-\hypertarget{push}{%
-\subsection{PUSH}\label{push}}
+\subsection{PUSH}\label{push}
 
@@
 
-\hypertarget{queue}{%
-\subsection{QUEUE}\label{queue}}
+\subsection{QUEUE}\label{queue}
 
@@
 
-\hypertarget{raise}{%
-\subsection{RAISE}\label{raise}}
+\subsection{RAISE}\label{raise}
 
@@
 
-\hypertarget{reply}{%
-\subsection{REPLY}\label{reply}}
+\subsection{REPLY}\label{reply}
 
@@
 
-\hypertarget{select}{%
-\subsection{SELECT}\label{select}}
+\subsection{SELECT}\label{select}
 
@@
 
-\hypertarget{signal}{%
-\subsection{SIGNAL}\label{signal}}
+\subsection{SIGNAL}\label{signal}
 
@@
 
-\hypertarget{trace-output}{%
-\subsection{Trace output}\label{trace-output}}
+\subsection{Trace output}\label{trace-output}
 
@@
 
-\hypertarget{conditions-and-messages}{%
-\section{Conditions and Messages}\label{conditions-and-messages}}
+\section{Conditions and Messages}\label{conditions-and-messages}
 
@@
 
-\hypertarget{raising-of-conditions}{%
-\subsection{Raising of conditions}\label{raising-of-conditions}}
+\subsection{Raising of conditions}\label{raising-of-conditions}
 
@@
 
-\hypertarget{messages-during-execution}{%
-\subsection{Messages during execution}\label{messages-during-execution}}
+\subsection{Messages during execution}\label{messages-during-execution}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/provided_classes.tex
work in standard book production process including readme.md
upd

@@
-\hypertarget{provided-classes}{%
-\chapter{Provided classes}\label{provided-classes}}
+%preprocessed texin
+\chapter{Provided classes}\label{provided-classes}
 
@@
 
-\hypertarget{collection-class-routines}{%
-\subsection{Collection Class Routines}\label{collection-class-routines}}
+\subsection{Collection Class Routines}\label{collection-class-routines}
 
@@
 
-\hypertarget{the-collection-class}{%
-\subsection{The collection class}\label{the-collection-class}}
+\subsection{The collection class}\label{the-collection-class}
 
@@
 
-\hypertarget{init}{%
-\subsubsection{INIT}\label{init}}
+\subsubsection{INIT}\label{init}
 
@@
 
-\hypertarget{exposed}{%
-\subsubsection{EXPOSED}\label{exposed}}
+\subsubsection{EXPOSED}\label{exposed}
 
@@
 
-\hypertarget{findindex}{%
-\subsubsection{FINDINDEX}\label{findindex}}
+\subsubsection{FINDINDEX}\label{findindex}
 
@@
 
-\hypertarget{at}{%
-\subsubsection{AT}\label{at}}
+\subsubsection{AT}\label{at}
 
@@
 
-\hypertarget{section}{%
-\subsubsection{{[}{]}}\label{section}}
+\subsubsection{{[}{]}}\label{section}
 
@@
 
-\hypertarget{put}{%
-\subsubsection{PUT}\label{put}}
+\subsubsection{PUT}\label{put}
 
@@
 
-\hypertarget{section-1}{%
-\subsubsection{{[}{]}=}\label{section-1}}
+\subsubsection{{[}{]}=}\label{section-1}
 
@@
 
-\hypertarget{hasindex}{%
-\subsubsection{HASINDEX}\label{hasindex}}
+\subsubsection{HASINDEX}\label{hasindex}
 
@@
 
-\hypertarget{items}{%
-\subsubsection{ITEMS}\label{items}}
+\subsubsection{ITEMS}\label{items}
 
@@
 
-\hypertarget{remove}{%
-\subsubsection{REMOVE}\label{remove}}
+\subsubsection{REMOVE}\label{remove}
 
@@
 
-\hypertarget{removeit}{%
-\subsubsection{REMOVEIT}\label{removeit}}
+\subsubsection{REMOVEIT}\label{removeit}
 
@@
 
-\hypertarget{makearrya}{%
-\subsubsection{MAKEARRYA}\label{makearrya}}
+\subsubsection{MAKEARRYA}\label{makearrya}
 
@@
 
-\hypertarget{makearrayx}{%
-\subsubsection{MAKEARRAYX}\label{makearrayx}}
+\subsubsection{MAKEARRAYX}\label{makearrayx}
 
@@
 
-\hypertarget{supplier}{%
-\subsubsection{SUPPLIER}\label{supplier}}
+\subsubsection{SUPPLIER}\label{supplier}
 
@@
 
-\hypertarget{class-list}{%
-\subsection{Class list}\label{class-list}}
+\subsection{Class list}\label{class-list}
 
@@
 
-\hypertarget{put-1}{%
-\subsubsection{PUT}\label{put-1}}
+\subsubsection{PUT}\label{put-1}
 
@@
 
-\hypertarget{of}{%
-\subsubsection{OF}\label{of}}
+\subsubsection{OF}\label{of}
 
@@
 
-\hypertarget{insert}{%
-\subsubsection{INSERT}\label{insert}}
+\subsubsection{INSERT}\label{insert}
 
@@
 
-\hypertarget{first}{%
-\subsubsection{FIRST}\label{first}}
+\subsubsection{FIRST}\label{first}
 
@@
 
-\hypertarget{last}{%
-\subsubsection{LAST}\label{last}}
+\subsubsection{LAST}\label{last}
 
@@
 
-\hypertarget{firstitem}{%
-\subsubsection{FIRSTITEM}\label{firstitem}}
+\subsubsection{FIRSTITEM}\label{firstitem}
 
@@
 
-\hypertarget{lastitem}{%
-\subsubsection{LASTITEM}\label{lastitem}}
+\subsubsection{LASTITEM}\label{lastitem}
 
@@
 
-\hypertarget{next}{%
-\subsubsection{NEXT}\label{next}}
+\subsubsection{NEXT}\label{next}
 
@@
 
-\hypertarget{previous}{%
-\subsubsection{PREVIOUS}\label{previous}}
+\subsubsection{PREVIOUS}\label{previous}
 
@@
 
-\hypertarget{section-2}{%
-\subsubsection{SECTION}\label{section-2}}
+\subsubsection{SECTION}\label{section-2}
 
@@
 
-\hypertarget{class-queue}{%
-\subsection{Class queue}\label{class-queue}}
+\subsection{Class queue}\label{class-queue}
 
@@
 
-\hypertarget{push}{%
-\subsubsection{PUSH}\label{push}}
+\subsubsection{PUSH}\label{push}
 
@@
 
-\hypertarget{queue}{%
-\subsubsection{QUEUE}\label{queue}}
+\subsubsection{QUEUE}\label{queue}
 
@@
 
-\hypertarget{peek}{%
-\subsubsection{PEEK}\label{peek}}
+\subsubsection{PEEK}\label{peek}
 
@@
 
-\hypertarget{remove-1}{%
-\subsubsection{REMOVE}\label{remove-1}}
+\subsubsection{REMOVE}\label{remove-1}
 
@@
 
-\hypertarget{class-table}{%
-\subsection{Class table}\label{class-table}}
+\subsection{Class table}\label{class-table}
 
@@
 
-\hypertarget{makearray}{%
-\subsubsection{MAKEARRAY}\label{makearray}}
+\subsubsection{MAKEARRAY}\label{makearray}
 
@@
 
-\hypertarget{union}{%
-\subsubsection{UNION}\label{union}}
+\subsubsection{UNION}\label{union}
 
@@
 
-\hypertarget{intersection}{%
-\subsubsection{INTERSECTION}\label{intersection}}
+\subsubsection{INTERSECTION}\label{intersection}
 
@@
 
-\hypertarget{xor}{%
-\subsubsection{XOR}\label{xor}}
+\subsubsection{XOR}\label{xor}
 
@@
 
-\hypertarget{difference}{%
-\subsubsection{DIFFERENCE}\label{difference}}
+\subsubsection{DIFFERENCE}\label{difference}
 
@@
 
-\hypertarget{subset}{%
-\subsubsection{SUBSET}\label{subset}}
+\subsubsection{SUBSET}\label{subset}
 
@@
 
-\hypertarget{class-set}{%
-\subsubsection{Class set}\label{class-set}}
+\subsubsection{Class set}\label{class-set}
 
@@
 
-\hypertarget{put-2}{%
-\subsubsection{PUT}\label{put-2}}
+\subsubsection{PUT}\label{put-2}
 
@@
 
-\hypertarget{of-1}{%
-\subsubsection{OF}\label{of-1}}
+\subsubsection{OF}\label{of-1}
 
@@
 
-\hypertarget{union-1}{%
-\subsubsection{UNION}\label{union-1}}
+\subsubsection{UNION}\label{union-1}
 
@@
 
-\hypertarget{intersection-1}{%
-\subsubsection{INTERSECTION}\label{intersection-1}}
+\subsubsection{INTERSECTION}\label{intersection-1}
 
@@
 
-\hypertarget{xor-1}{%
-\subsubsection{XOR}\label{xor-1}}
+\subsubsection{XOR}\label{xor-1}
 
@@
 
-\hypertarget{difference-1}{%
-\subsubsection{DIFFERENCE}\label{difference-1}}
+\subsubsection{DIFFERENCE}\label{difference-1}
 
@@
 
-\hypertarget{class-relation}{%
-\subsection{Class relation}\label{class-relation}}
+\subsection{Class relation}\label{class-relation}
 
@@
 
-\hypertarget{put-3}{%
-\subsubsection{PUT}\label{put-3}}
+\subsubsection{PUT}\label{put-3}
 
@@
 
-\hypertarget{items-1}{%
-\subsubsection{ITEMS}\label{items-1}}
+\subsubsection{ITEMS}\label{items-1}
 
@@
 
-\hypertarget{makearray-1}{%
-\subsubsection{MAKEARRAY}\label{makearray-1}}
+\subsubsection{MAKEARRAY}\label{makearray-1}
 
@@
 
-\hypertarget{supplier-1}{%
-\subsubsection{SUPPLIER}\label{supplier-1}}
+\subsubsection{SUPPLIER}\label{supplier-1}
 
@@
 
-\hypertarget{intersection-2}{%
-\subsubsection{INTERSECTION}\label{intersection-2}}
+\subsubsection{INTERSECTION}\label{intersection-2}
 
@@
 
-\hypertarget{xor-2}{%
-\subsubsection{XOR}\label{xor-2}}
+\subsubsection{XOR}\label{xor-2}
 
@@
 
-\hypertarget{difference-2}{%
-\subsubsection{DIFFERENCE}\label{difference-2}}
+\subsubsection{DIFFERENCE}\label{difference-2}
 
@@
 
-\hypertarget{subset-1}{%
-\subsubsection{SUBSET}\label{subset-1}}
+\subsubsection{SUBSET}\label{subset-1}
 
@@
 
-\hypertarget{removeitem}{%
-\subsubsection{REMOVEITEM}\label{removeitem}}
+\subsubsection{REMOVEITEM}\label{removeitem}
 
@@
 
-\hypertarget{index}{%
-\subsubsection{INDEX}\label{index}}
+\subsubsection{INDEX}\label{index}
 
@@
 
-\hypertarget{allat}{%
-\subsubsection{ALLAT}\label{allat}}
+\subsubsection{ALLAT}\label{allat}
 
@@
 
-\hypertarget{hasitem}{%
-\subsubsection{HASITEM}\label{hasitem}}
+\subsubsection{HASITEM}\label{hasitem}
 
@@
 
-\hypertarget{allindex}{%
-\subsubsection{ALLINDEX}\label{allindex}}
+\subsubsection{ALLINDEX}\label{allindex}
 
@@
 
-\hypertarget{of-2}{%
-\subsubsection{OF}\label{of-2}}
+\subsubsection{OF}\label{of-2}
 
@@
 
-\hypertarget{put-4}{%
-\subsubsection{PUT}\label{put-4}}
+\subsubsection{PUT}\label{put-4}
 
@@
 
-\hypertarget{union-2}{%
-\subsubsection{UNION}\label{union-2}}
+\subsubsection{UNION}\label{union-2}
 
@@
 
-\hypertarget{intersection-3}{%
-\subsubsection{INTERSECTION}\label{intersection-3}}
+\subsubsection{INTERSECTION}\label{intersection-3}
 
@@
 
-\hypertarget{xor-3}{%
-\subsubsection{XOR}\label{xor-3}}
+\subsubsection{XOR}\label{xor-3}
 
@@
 
-\hypertarget{difference-3}{%
-\subsubsection{DIFFERENCE}\label{difference-3}}
+\subsubsection{DIFFERENCE}\label{difference-3}
 
@@
 
-\hypertarget{the-directory-class}{%
-\subsection{The directory class}\label{the-directory-class}}
+\subsection{The directory class}\label{the-directory-class}
 
@@
 
-\hypertarget{at-1}{%
-\subsubsection{AT}\label{at-1}}
+\subsubsection{AT}\label{at-1}
 
@@
 
-\hypertarget{put-5}{%
-\subsubsection{PUT}\label{put-5}}
+\subsubsection{PUT}\label{put-5}
 
@@
 
-\hypertarget{makearray-2}{%
-\subsubsection{MAKEARRAY}\label{makearray-2}}
+\subsubsection{MAKEARRAY}\label{makearray-2}
 
@@
 
-\hypertarget{supplier-2}{%
-\subsubsection{SUPPLIER}\label{supplier-2}}
+\subsubsection{SUPPLIER}\label{supplier-2}
 
@@
 
-\hypertarget{intersection-4}{%
-\subsubsection{INTERSECTION}\label{intersection-4}}
+\subsubsection{INTERSECTION}\label{intersection-4}
 
@@
 
-\hypertarget{xor-4}{%
-\subsubsection{XOR}\label{xor-4}}
+\subsubsection{XOR}\label{xor-4}
 
@@
 
-\hypertarget{difference-4}{%
-\subsubsection{DIFFERENCE}\label{difference-4}}
+\subsubsection{DIFFERENCE}\label{difference-4}
 
@@
 
-\hypertarget{subset-2}{%
-\subsubsection{SUBSET}\label{subset-2}}
+\subsubsection{SUBSET}\label{subset-2}
 
@@
 
-\hypertarget{setentry}{%
-\subsubsection{SETENTRY}\label{setentry}}
+\subsubsection{SETENTRY}\label{setentry}
 
@@
 
-\hypertarget{entry}{%
-\subsubsection{ENTRY}\label{entry}}
+\subsubsection{ENTRY}\label{entry}
 
@@
 
-\hypertarget{hasentry}{%
-\subsubsection{HASENTRY}\label{hasentry}}
+\subsubsection{HASENTRY}\label{hasentry}
 
@@
 
-\hypertarget{setmethod}{%
-\subsubsection{SETMETHOD}\label{setmethod}}
+\subsubsection{SETMETHOD}\label{setmethod}
 
@@
 
-\hypertarget{unknown}{%
-\subsubsection{UNKNOWN}\label{unknown}}
+\subsubsection{UNKNOWN}\label{unknown}
 
@@
 
-\hypertarget{the-stream-class}{%
-\section{The stream class}\label{the-stream-class}}
+\section{The stream class}\label{the-stream-class}
 
@@
 
-\hypertarget{the-alarm-class}{%
-\section{The alarm class}\label{the-alarm-class}}
+\section{The alarm class}\label{the-alarm-class}
 
@@
 
-\hypertarget{the-monitor-class}{%
-\section{The monitor class}\label{the-monitor-class}}
+\section{The monitor class}\label{the-monitor-class}
 
@@
 
-\hypertarget{init-1}{%
-\subsection{INIT}\label{init-1}}
+\subsection{INIT}\label{init-1}
 
@@
 
-\hypertarget{current}{%
-\subsection{CURRENT}\label{current}}
+\subsection{CURRENT}\label{current}
 
@@
 
-\hypertarget{destination}{%
-\subsection{DESTINATION}\label{destination}}
+\subsection{DESTINATION}\label{destination}
 
@@
 
-\hypertarget{unknown-1}{%
-\subsection{UNKNOWN}\label{unknown-1}}
+\subsection{UNKNOWN}\label{unknown-1}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/syntax_constructs.tex
work in standard book production process including readme.md
upd

@@
-\hypertarget{syntax-constructs}{%
-\chapter{Syntax constructs}\label{syntax-constructs}}
+%preprocessed texin
+\chapter{Syntax constructs}\label{syntax-constructs}
 
-\hypertarget{notation}{%
-\section{Notation}\label{notation}}
+\section{Notation}\label{notation}
 
-\hypertarget{backus-naur-form-bnf}{%
-\subsection{Backus-Naur Form (BNF)}\label{backus-naur-form-bnf}}
+\subsection{Backus-Naur Form (BNF)}\label{backus-naur-form-bnf}
 
@@
 
-\hypertarget{operands}{%
-\subsection{Operands}\label{operands}}
+\subsection{Operands}\label{operands}
 
@@
 
-\hypertarget{grouping}{%
-\subsection{Grouping}\label{grouping}}
+\subsection{Grouping}\label{grouping}
 
@@
 
-\hypertarget{bnf-syntax-definition}{%
-\subsection{BNF syntax definition}\label{bnf-syntax-definition}}
+\subsection{BNF syntax definition}\label{bnf-syntax-definition}
 
@@
 
-\hypertarget{syntactic-errors}{%
-\subsection{Syntactic errors}\label{syntactic-errors}}
+\subsection{Syntactic errors}\label{syntactic-errors}
 
@@
 
-\hypertarget{lexical}{%
-\section{Lexical}\label{lexical}}
+\section{Lexical}\label{lexical}
 
@@
 
-\hypertarget{lexical-elements}{%
-\subsection{Lexical elements}\label{lexical-elements}}
+\subsection{Lexical elements}\label{lexical-elements}
 
-\hypertarget{events}{%
-\subsubsection{Events}\label{events}}
+\subsubsection{Events}\label{events}
 
@@
 
-\hypertarget{rules}{%
-\subsubsection{Rules}\label{rules}}
+\subsubsection{Rules}\label{rules}
 
@@
 
-\hypertarget{lexical-level}{%
-\subsection{Lexical level}\label{lexical-level}}
+\subsection{Lexical level}\label{lexical-level}
 
-\hypertarget{interaction-between-levels-of-syntax}{%
 \subsection{Interaction between levels of
-syntax}\label{interaction-between-levels-of-syntax}}
+syntax}\label{interaction-between-levels-of-syntax}
 
@@
 
-\hypertarget{reserved-symbols}{%
-\subsubsection{Reserved symbols}\label{reserved-symbols}}
+\subsubsection{Reserved symbols}\label{reserved-symbols}
 
@@
 
-\hypertarget{function-name-syntax}{%
-\subsubsection{Function name syntax}\label{function-name-syntax}}
+\subsubsection{Function name syntax}\label{function-name-syntax}
 
@@
 
-\hypertarget{syntax}{%
-\section{Syntax}\label{syntax}}
+\section{Syntax}\label{syntax}
 
-\hypertarget{syntax-elements}{%
-\subsection{Syntax elements}\label{syntax-elements}}
+\subsection{Syntax elements}\label{syntax-elements}
 
@@
 
-\hypertarget{syntax-level}{%
-\subsection{Syntax level}\label{syntax-level}}
+\subsection{Syntax level}\label{syntax-level}
 
@@
 
-\hypertarget{syntactic-information}{%
-\section{Syntactic information}\label{syntactic-information}}
+\section{Syntactic information}\label{syntactic-information}
 
-\hypertarget{var_symbol-matching}{%
-\subsection{VAR\_SYMBOL matching}\label{var_symbol-matching}}
+\subsection{VAR\_SYMBOL matching}\label{var_symbol-matching}
 
@@
 
-\hypertarget{trace-only-labels}{%
-\subsection{Trace-only labels}\label{trace-only-labels}}
+\subsection{Trace-only labels}\label{trace-only-labels}
 
@@
 
-\hypertarget{clauses-and-line-numbers}{%
-\subsection{Clauses and line numbers}\label{clauses-and-line-numbers}}
+\subsection{Clauses and line numbers}\label{clauses-and-line-numbers}
 
@@
 
-\hypertarget{nested-if-instructions}{%
-\subsection{Nested IF instructions}\label{nested-if-instructions}}
+\subsection{Nested IF instructions}\label{nested-if-instructions}
 
@@
 
-\hypertarget{choice-of-messages}{%
-\subsection{Choice of messages}\label{choice-of-messages}}
+\subsection{Choice of messages}\label{choice-of-messages}
 
@@
 
-\hypertarget{creation-of-messages}{%
-\subsection{Creation of messages}\label{creation-of-messages}}
+\subsection{Creation of messages}\label{creation-of-messages}
 
@@
 
-\hypertarget{error-message-prefix}{%
-\subsubsection{Error message prefix}\label{error-message-prefix}}
+\subsubsection{Error message prefix}\label{error-message-prefix}
 
@@
 
-\hypertarget{replacement-of-insertions}{%
-\section{Replacement of insertions}\label{replacement-of-insertions}}
+\section{Replacement of insertions}\label{replacement-of-insertions}
 
@@
 
-\hypertarget{syntactic-equivalence}{%
-\section{Syntactic equivalence}\label{syntactic-equivalence}}
+\section{Syntactic equivalence}\label{syntactic-equivalence}
 

### 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/evaluation.tex
upd

@@
 
+The value of \#Test, in conjunction with the operator in the comparison,
+determines the value of the comparison. The value of the comparison is
+`1' if - \#Test is `E' and the operator is one of `=``\,``,'==````,
+`\textgreater=', \textless='', `\textgreater{}', `\textless{}',
+`p\textgreater=', `\textless\textless=', \textgreater\textgreater', or
+\textless\textless)
+
+\begin{itemize}
+\tightlist
+\item
+  \#Test is `G' and the operator is one of `\textgreater{}',
+  `\textgreater=``, `\textless{}', `=', `\textless\textgreater{}',
+  `\textgreater\textless{}', Nes'', `\textgreater\textgreater!
+  `p\textgreater{}', or \textless\textless``)
+\item
+  \#Test is `L' and the operator is one of `\textless{}', \textless=``,
+  \textgreater{}`, =', `\textless\textgreater{}',
+  `\textgreater\textless{}', ==`,'\textless\textless',
+  *\textless\textless=`, or \textgreater\textgreater{}'. In all other
+  cases the value of the comparison is `0'.
+\end{itemize}
+
+If the comparison is not a concatenation then: if \#Tracing.\#Level ==
+`I' then call \#Trace `\textgreater0O\textgreater{}'
+
 \subsection{The value of an
@@
 high setting of numeric digits to avoid exponential notation.
-\lstinputlisting[language=rexx,label=evaluation\_arithmetic.rexx,caption=evaluation\_arithmetic.rexx]{evaluation\_arithmetic.rexx}
+\lstinputlisting[language=rexx,label=evaluation-arithmetic.rexx,caption=evaluation-arithmetic.rexx]{evaluation-arithmetic.rexx}
 
@@
 if \#Tracing.\#Level == `I' then call \#Trace
-`\textgreater F\textgreater{}' Invocation of a message\_term may result
-in a value, in which case: if \#Tracing.\#Level == `I' then call \#Trace
+`\textgreater F\textgreater{}'
+
+Invocation of a message\_term may result in a value, in which case:
+
+if \#Tracing.\#Level == `I' then call \#Trace
 `\textgreater M\textgreater{}'

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Create reply-to-jlf-review.md (first paragraphs)

# Reply to Jean Louis Faucher review of the TUTOR article

I am replying in markdown format, because pdf is not very useable to have a dialog.

Please note that the version you have reviewed is not final. I have uploaded the final, published, versions of the articles in the Unicode/publications/35 subdirectory.

I will not copy some of your comments in the pdf, and intersperse my own comments in between. Page references are to the version you reviewed, not to the final document.

(p. 33)
>Inheritance from String.
>
>Each time a new instance of Byte, Codepoints, Graphemes or Text is created, a new copy of the original string is created. That can be confirmed by displaying the identityhash of the ~string.
>
>Not a problem for a prototype, but would be a problem for a real implementation. As written elsewhere, these 4 instances are a view of the same string. So it would be better to share this same string between the 4 views.
>
>Executor provides only Text, which correspond roughly to Tutor’s Graphemes (no normalization, but also no automatic conversion to UTF-8).  The original string is referenced (delegation), not inherited.
>That doesnt forbid to offer the same interface as .String.

OK. As you say, TUTOR is a prototype; I have not strived for efficiency. I am also under the impression that the code is much simpler if I use inheritance instead of delegation (but I might be wrong).

(p. 34)
>Executor applies NFC normalization on the fly when strict comparison of text.
>
>The difference between strict and non-strict is the management of blanks and numeric values.
>
>Rexxref: The two strings must be identical (character by character) and of the same length to be considered strictly equal. 
>No padding, no attempt to perform a numeric comparison

My interpretation of strict equality forbids any normalization. I understand "character by character" as _binary_ equality. 
If you normalize on the fly, you lose a first-class operator to determine whether two strings are binary identical.

My impression is that normalization pertains to _non-strict_ equality. "Normal" equality, that is, non-strict equality, is, somehow, the default
one: blanks are ignored, numbers have to be equivalent, etc., ... and "e" + acute accent is the same as "é".

OTOH, "==", I think, should mean "identical", that is, binary identical.

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md: some more paras

@@
 OTOH, "==", I think, should mean "identical", that is, binary identical.
+
+(p. 43)
+>Can’t resist to remind my point of view, but feel free to ignore, was already discussed at one moment and written somewhere in the draft documents.
+>
+>It’s well defined: it’s what the interpreter exposes to you via the native C2X.
+>
+>Netrexx: UCS-2 (16 bits)
+>
+>Crexx: UTF-8
+>
+>all other interpreters: byte, whatever the encoding.
+>
+>You defined several views. Just apply these views on the native C2X. That will be exactly the same hex digits than the native C2X, but grouped according the view.
+
+I think the final version refers to it (not sure about the version you have commented): it appears that recent versions of Java store Unicode strings as ISO-8859-1 when all 
+codepoints are < "FF"X. This means that the notion of "internal representation" does no longer make sense (or varies according to whether a string contains a single
+codepoint >= "100"X, which is ridiculous), as it is the remnant of a situation which is no longer true: when strings were manageable by a simple adlen pair, and were
+composed of bytes.
+
+(Ibid.)
+>Hum… Don’t you have another BIF that convert to  a target encoding?
+
+Yes. The addition of a second, encoding parameter to C2X is a compatibility aid: if you want NetRexx behaviour, for example, you just add "UTF-16" and you are done.
+
+(p. 45)
+> This is not very usual to return 2 results at once. Sounds like an internal optimization for your implementation.
+> 
+> UTF8()) returns a stem.
+
+You are right, there are several details in the current version that are internal optimization leftovers. Will try to fix in a subsequent release.

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md

@@
 
-You are right, there are several details in the current version that are internal optimization leftovers. Will try to fix in a subsequent release.
+You are right, there are several details in the current version that are internal optimization leftovers. Will try to fix in a subsequent release. https://github.com/RexxLA/rexx-repository/issues/10
+
+(Ibid.)
+>This decode/encode reminds me Python.
+>
+>But your functions seems to be still a draft…
+>
+>with decode, you can transcode from a source encoding to a target encoding, where the target encoding is limited to UTF-8/16/32?
+
+Only -8 and -32. This is another leftover. I need the -32 version internally, but I should most probably not expose that as an API.
+
+(cont.)
+>encode is also doing a transcoding… Only the target encoding is specified.
+>The source encoding is UTF-8/16/32? Why is it not needed to specify it here?
+
+My way of understanding the standard is the following: Unicode is the _lingua franca_ of all encodings. You only need a DECODE function that
+specifies an encoding E (this converts from E to Unicode, our _lingua franca_), and a ENCODE function: this converts from Unicode,
+our _lingua franca_, to a specified encoding E. You do not need a source encoding for ENCODE, since your source string is always (abstract)
+Unicode.
+
+If you want to transcode, say, from UTF-32 to another encoding E, you should first DECODE from UTF-32, and then ENCODE to E.
+
+Most probably, the fact that DECODE allows a limited set of target Unicode encodings is misleading, and therefore a bad design decision. Thanks for bringing
+this aspect to my attention.

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md

@@
 this aspect to my attention.
+
+(p. 47)
+>I’m not aware of WTF32, and I don’t see how it could be needed. There is no low/high surrogates in UTF32.
+>Do you have links?
+
+I can't seem to be able to recover the links I once had, my bad; I have not invented the term. Just because UTF-32 does not allow
+lone surrogates, we _need_ WTF-32. For example, when handling WTF-8, we may find lone surrogates. Either we forbid transforming
+WTF-8 to a four-bytes per codepoint format (bad), or we have to implement WTF-32.
+
+(p. 48)
+>U2C() is cited once elsewhere in this document. But not yet implemented? or discarded?
+
+Forgot to document it :/ It was implemented as a method, but not as a (new) BIF.
+Fixed: https://github.com/RexxLA/rexx-repository/blob/master/ARB/standards/work-in-progress/unicode/UnicodeTools/doc/new-functions.md#u2c-unicode-to-character

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md

@@
 Fixed: https://github.com/RexxLA/rexx-repository/blob/master/ARB/standards/work-in-progress/unicode/UnicodeTools/doc/new-functions.md#u2c-unicode-to-character
+
+(Ibid.)
+>Should I provide U2C? it would support only the U+ notation.
+>Yes, having C2U/U2C would be consistent with C2X/X2C.
+>The target encoding would be the default encoding.
+
+I can see that this is a note to yourself. Just wanted to add that this is just the idea: consistency with C2X/X2C. My U2C allows names, aliases and labels, in addition to codepoints and U+codepoints.
+
+(Ibid.)
+>Low level, so maybe you need it for your implementation.
+>
+>But from an end-user perspective, don’t you have a BIF that returns all the properties of a unicode character? including its name and aliases. Yes: UNICODE()
+
+I think that in many cases it may be very convenient to specify a character by name. Not so much low-level, I think.
+
+The idea of subsuming this in UNICODE is great. https://github.com/RexxLA/rexx-repository/issues/11

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md

@@
 The idea of subsuming this in UNICODE is great. https://github.com/RexxLA/rexx-repository/issues/11
+
+(p. 50)
+>You could also support a Unicode character name, but I understand that it collides with the other form UNICODE(string…).
+
+Again a great idea. https://github.com/RexxLA/rexx-repository/issues/12
+
+(p. 55)
+>Under macOS, and probably under Linux:
+>
+>  rlwrap rexx rxutry
+>
+>to have an history of commands.
+
+Thanks for the tip! I will document it in the following release.
+
+(p. 56)
+>Something specific to ooRexx (and maybe Netrexx?): the caseless methods.
+
+Have not delved into case folding still. Thanks for the info!

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md

@@
 
-Please note that the version you have reviewed is not final. I have uploaded the final, published, versions of the articles in the Unicode/publications/35 subdirectory.
+Please note that the version you have reviewed is not the final one. I have uploaded the final, published, versions of the articles in the Unicode/publications/35 subdirectory.
 
-I will not copy some of your comments in the pdf, and intersperse my own comments in between. Page references are to the version you reviewed, not to the final document.
+I will now copy some of your comments in the pdf, and intersperse my own comments in between. Page references are to the version you reviewed, not to the final document.
 

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md

@@
 
-OTOH, "==", I think, should mean "identical", that is, binary identical.
+"==", I think, should mean "identical", that is, binary identical.
 
@@
 codepoints are < "FF"X. This means that the notion of "internal representation" does no longer make sense (or varies according to whether a string contains a single
-codepoint >= "100"X, which is ridiculous), as it is the remnant of a situation which is no longer true: when strings were manageable by a simple adlen pair, and were
-composed of bytes.
+codepoint >= "100"X, which is ridiculous), as it is the remnant of a situation which is no longer true: when strings were manageable by a simple adlen pair and a chunk of bytes.
 

### 2024-03-23 ARB/standards/work-in-progress/unicode/Symposium 2024 review/reply-to-jlf-review.md
Update reply-to-jlf-review.md

@@
 My way of understanding the standard is the following: Unicode is the _lingua franca_ of all encodings. You only need a DECODE function that
-specifies an encoding E (this converts from E to Unicode, our _lingua franca_), and a ENCODE function: this converts from Unicode,
+specifies an encoding E (this converts from E to Unicode, our _lingua franca_), and an ENCODE function: this converts from Unicode,
 our _lingua franca_, to a specified encoding E. You do not need a source encoding for ENCODE, since your source string is always (abstract)

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/built-in_classes.tex
add filenames

@@
 
-::class object
+\lstinputlisting[language=rexx,label=object.rexx,caption=object.rexx]{object.rexx}
 
-::method new class Returns a new instance of the receiver class.
+Returns a new instance of the receiver class.
 
-call Config ObjectNew return \#Outcome
+\lstinputlisting[language=rexx,label=config.rexx,caption=config.rexx]{config.rexx}
 
-::method `==' `==' with no argument gives a hash value in OOI.
+\emph{\texttt{\textquotesingle{}==\textquotesingle{}} with no argument
+gives a hash value in OOI.}
 
-call Config ObjectCompare \#Receiver, \#Arg.1
+\lstinputlisting[language=rexx,label=doubleequals.rexx,caption=doubleequals.rexx]{doubleequals.rexx}
 
-if \#Outcome == `equal' then return `1'
+Returns a copy of the receiver object. The copied object has the same
+methods as the receiver object and an equivalent set of object
+variables, with the same values.
 
-else return `0'
+\lstinputlisting[language=rexx,label=objectcopy.rexx,caption=objectcopy.rexx]{objectcopy.rexx}
 
-::method `\textless\textgreater!'
+\emph{Since we have \texttt{var\_empty} we could save a primitive by
+rendering `new' as `copy' plus `empty'.}
 
-use arg a
-
-return \self==a
-
-::method `\textgreater\textless{}' forward message
-`\textless\textgreater{}'
-
-::method `=' forward message `=='
-
-::method `=! forward message'\textless\textgreater'
-
-::method `==' forward message `\textless\textgreater{}'
-
-::method copy Returns a copy of the receiver object. The copied object
-has the same methods as the receiver object and an equivalent set of
-object variables, with the same values.
-
-call Config ObjectCopy \#Receiver
-
-return \#Outcome
-
-Since we have var\_empty we could save a primitive by rendering `new' as
-`copy' plus `empty'. ::method defaultname
+\lstinputlisting[language=rexx,label=defaultname.rexx,caption=defaultname.rexx]{defaultname.rexx}
 
@@
 
-call var\_value \#Receiver, `\#Human', `0'
+\lstinputlisting[language=rexx,label=receiver.rexx,caption=receiver.rexx]{receiver.rexx}
 
-return \#Outcome
+\emph{This field would have been filled in at `NEW' time.}
 
-This field would have been filled in at `NEW' time.
-
-::method `OBJECTNAMES' /* rvSTRING */
+\lstinputlisting[language=rexx,label=objectnames.rexx,caption=objectnames.rexx]{objectnames.rexx}
 
@@
 
-call var\_set \#Receiver, \#ObjectName, `0', \#Arg.1 return
+\lstinputlisting[language=rexx,label=varset.rexx,caption=varset.rexx]{varset.rexx}
 
-Initialized to \#Human? Or ObjectName does forwarding until assigned to?
+\emph{Initialized to \#Human? Or ObjectName does forwarding until
+assigned to?}
 
-::method objectname
+\lstinputlisting[language=rexx,label=objectname,caption=objectname]{objectname}
 
-Returns the receiver object's name (which the OBJECTNAME= method sets).
-call var\_value \#Receiver, \#ObjectName, `0' return \#Outcome
+Returns the receiver object's name (which the \texttt{OBJECTNAME=}
+method sets).
 
-::method string
+\lstinputlisting[language=rexx,label=receiverObjectName.rexx,caption=receiverObjectName.rexx]{receiverObjectName.rexx}
 
-Returns a human-readable string representation for the object. return
-\#Receiver\textasciitilde ObjectName
+Returns a human-readable string representation for the object.
 
-::method class Returns the class object that received the message that
-created the object.
+\lstinputlisting[language=rexx,label=methodclass.rexx,caption=methodclass.rexx]{methodclass.rexx}
 
-call var value \#Receiver, \#IsA, `0' return \#Outcome
+Returns the class object that received the message that created the
+object.
 
-::method setmethod /* rSTRING oSTRING/METHOD/ARRAY */ Adds a method to
-the receiver object's collection of object methods.
+\lstinputlisting[language=rexx,label=setmethod.rexx,caption=setmethod.rexx]{setmethod.rexx}
 
-Is `object methods' what is intended; you add to a class without adding
-to its instance methods? Yes. if \#ArgExists.2 then m = Arg.2
+Adds a method to the receiver object's collection of object methods.
 
-else m = .NIL call set\_var \#Receiver, `METHODS.'\#Arg.1, `1', m return
-::method hasmethod /* rSTRING */
+\emph{Is `object methods' what is intended; you add to a class without
+adding to its instance methods? Yes.}
+
+\lstinputlisting[language=rexx,label=hasmethod.rexx,caption=hasmethod.rexx]{hasmethod.rexx}
 
@@
 
-This presumably means inherited as well as SETMETHOD ones. What about
-ones set to .NIL?
+\emph{This presumably means inherited as well as \texttt{SETMETHOD}
+ones. What about ones set to \texttt{.NIL}?}
 
-Need to use the same search as for sending.
+\emph{Need to use the same search as for sending.}
 
-::method unsetmethod private
+\lstinputlisting[language=rexx,label=unsetmethod.rexx,caption=unsetmethod.rexx]{unsetmethod.rexx}
 
 Removes a method from the receiver object's collection of object
-methods. Use var\_drop
+methods.
 
-Private means Receiver = Self check.
+\emph{Use \texttt{var\_drop}}
 
-::method request /* rSTRING */ Returns an object of the specified class,
-or the NIL object if the request cannot be satisfied. t = `MAKE'\#Arg.1
+\emph{Private means Receiver = Self check.}
 
-if \#Receiver\textasciitilde hasmethod(t) then return .NIL forward
-message(t) array ()
+\lstinputlisting[language=rexx,label=requestmethod.rexx,caption=requestmethod.rexx]{requestmethod.rexx}
 
-::method run private /* rMETHOD Ugh keyoptions */ Runs the specified
-method. The method has access to the object variables of the receiver
-object, just as if the receiver object had defined the method by using
-SETMETHOD.
+Returns an object of the specified class, or the \texttt{NIL} object if
+the request cannot be satisfied.
 
-::smethod startat Undocumented?
+\lstinputlisting[language=rexx,label=runmethod.rexx,caption=runmethod.rexx]{runmethod.rexx}
 
-::method start /* rMESSAGE oArglist */
+Runs the specified method. The method has access to the object variables
+of the receiver object, just as if the receiver object had defined the
+method by using \texttt{SETMETHOD}.
 
-Returns a message object and sends it a START message to start
-concurrent processing. ::method init
+\lstinputlisting[language=rexx,label=startat.rexx,caption=startat.rexx]{startat.rexx}
+
+Returns a message object and sends it a \texttt{START} message to start
+concurrent processing.
+
+\lstinputlisting[language=rexx,label=init.rexx,caption=init.rexx]{init.rexx}
 
@@
 
-::class class
+\lstinputlisting[language=rexx,label=classclass.rexx,caption=classclass.rexx]{classclass.rexx}
 
-Lots of these methods are both class and instance. \textbar{} don't know
-whether to list them twice. ::method new class /* OARGLIST */
+\emph{Lots of these methods are both class and instance. I don't know
+whether to list them twice.}
+
+\lstinputlisting[language=rexx,label=newclass.rexx,caption=newclass.rexx]{newclass.rexx}
 
@@
 the instance methods of the class. This method initializes a new
-instance by running its INIT methods.
+instance by running its \texttt{INIT} methods.
 
-::method subclass class
+\lstinputlisting[language=rexx,label=subclassclass.rexx,caption=subclassclass.rexx]{subclassclass.rexx}
 
@@
 
-::method subclasses class
+\lstinputlisting[language=rexx,label=immediatesubclass,caption=immediatesubclass]{immediatesubclass}
 
@@
 
-::method define class /* rSTRING oMETHOD */
+\lstinputlisting[language=rexx,label=defineclass.rexx,caption=defineclass.rexx]{defineclass.rexx}
 
@@
 
-::method delete
+\lstinputlisting[language=rexx,label=deleteclass.rexx,caption=deleteclass.rexx]{deleteclass.rexx}
 
@@
 
-Builtin classes cannot be altered.
+\emph{Builtin classes cannot be altered.}
 
-::method method class /* rSTRING */
+\lstinputlisting[language=rexx,label=methodmethod.rexx,caption=methodmethod.rexx]{methodmethod.rexx}
 
@@
 
-Do we have to keep saying ``method object'' as opposed to ``method''
-because ``method name'' exists? ::method querymixinclass
+\emph{Do we have to keep saying ``method object'' as opposed to
+``method'' because ``method name'' exists?}
+
+\lstinputlisting[language=rexx,label=querymixin.rexx,caption=querymixin.rexx]{querymixin.rexx}
 
@@
 
-::method mixinclass class /* 3 of em */
+\lstinputlisting[language=rexx,label=newmixinclass.rexx,caption=newmixinclass.rexx]{newmixinclass.rexx}
 
@@
 
-::method inherit class /* rCLASS oCLASS */
+\lstinputlisting[language=rexx,label=inheritclass.rexx,caption=inheritclass.rexx]{inheritclass.rexx}
 
 Causes the receiver class to inherit the instance and class methods of
-the class object specified. The
+the class object specified. The optional class is a class object that
+specifies the position of the new superclass in the list of
+superclasses.
 
-optional class is a class object that specifies the position of the new
-superclass in the list of superclasses. ::method uninherit class /*
-rCLASSOBJ */
+\lstinputlisting[language=rexx,label=uninheritclass.rexx,caption=uninheritclass.rexx]{uninheritclass.rexx}
 
-Nullifies the effect of any previous INHERIT message sent to the
-receiver for the class specified. ::smethod enhanced class /*
-rCOLLECTION oArgs */
+Nullifies the effect of any previous \texttt{INHERIT} message sent to
+the receiver for the class specified.
+
+\lstinputlisting[language=rexx,label=enhancedclass.rexx,caption=enhancedclass.rexx]{enhancedclass.rexx}
 
 Returns an enhanced new instance of the receiver class, with object
-methods that are the instance
+methods that are the instance methods of the class enhanced by the
+methods in the specified collection of methods.
 
-methods of the class enhanced by the methods in the specified collection
-of methods. ::method baseclass class
+\lstinputlisting[language=rexx,label=baseclassclass.rexx,caption=baseclassclass.rexx]{baseclassclass.rexx}
 
@@
 mixin class. If the class is not a mixin class, then the base class is
-the class receiving the BASECLASS message.
+the class receiving the \texttt{BASECLASS} message.
 
-::method superclasses class Returns the immediate superclasses of the
-receiver class in the form of a single-index array of the
+\lstinputlisting[language=rexx,label=superclassclass.rexx,caption=superclassclass.rexx]{superclassclass.rexx}
 
-required size. ::method id class
+Returns the immediate superclasses of the receiver class in the form of
+a single-index array of the required size.
 
-Returns a string that is the class identity (instance SUBCLASS and
-MIXINCLASS methods.)
+\lstinputlisting[language=rexx,label=methodid.rexx,caption=methodid.rexx]{methodid.rexx}
 
-::method metaclass class
+Returns a string that is the class identity (instance \texttt{SUBCLASS}
+and \texttt{MIXINCLASS} methods.)
 
-Returns the receiver class's default metaclass. ::method methods class
-/* oCLASSOBJECT */
+\lstinputlisting[language=rexx,label=metaclassclass.rexx,caption=metaclassclass.rexx]{metaclassclass.rexx}
+
+Returns the receiver class's default metaclass.
+
+\lstinputlisting[language=rexx,label=methodsclass.rexx,caption=methodsclass.rexx]{methodsclass.rexx}
 
 Returns a supplier object for all the instance methods of the receiver
-class and its superclasses, if no argument is specified. \#\#\# The
-method class
+class and its superclasses, if no argument is specified.
 
-::class method
+\subsection{The method class}\label{the-method-class}
 
-::method new class /* rSTRING rSOURCE */ Returns a new instance of
-method class, which is an executable representation of the code
-contained in
+\lstinputlisting[language=rexx,label=methodclass.rexx,caption=methodclass.rexx]{methodclass.rexx}
 
-the source. ::method setprivate
+Returns a new instance of method class, which is an executable
+representation of the code contained in the source.
 
-Specifies that a method is a private method. ::method setprotected
+\lstinputlisting[language=rexx,label=setprivatemethod.rexx,caption=setprivatemethod.rexx]{setprivatemethod.rexx}
 
-::method setsecuritymanager
+Specifies that a method is a private method.
 
-::method setguarded Reverses any previous SETUNGUARDED messages,
-restoring the receiver to the default guarded
+\lstinputlisting[language=rexx,label=setvarious.rexx,caption=setvarious.rexx]{setvarious.rexx}
 
-status. ::method setunguarded
+Reverses any previous \texttt{SETUNGUARDED} messages, restoring the
+receiver to the default guarded status.
+
+\lstinputlisting[language=rexx,label=setunguarded.rexx,caption=setunguarded.rexx]{setunguarded.rexx}
 
 Lets an object run a method even when another method is active on the
-same object. If a method object
+same object. If a method object does not receive a \texttt{SETUNGUARDED}
+message, it requires exclusive use of its object variable pool.
 
-does not receive a SETUNGUARDED message, it requires exclusive use of
-its object variable pool. ::method source
+\lstinputlisting[language=rexx,label=methodsource.rexx,caption=methodsource.rexx]{methodsource.rexx}
 
 Returns the method source code as a single index array of source lines.
-::method interface
 
-::method setinterface
+\lstinputlisting[language=rexx,label=methodinterface,caption=methodinterface]{methodinterface}
 
-11.3 The string class The string class provides conventional strings and
-numbers.
+\subsection{The string class}\label{the-string-class}
 
-Some differences from REXX class of NetRexx. ::class string
+The string class provides conventional strings and numbers.
 
-::method new class
+\emph{Some differences from REXX class of NetRexx.}
 
-::method `\textquotesingle{} We can do all the operators by appeal to
-classic section 7. ::method'-' ::method `-'
+\lstinputlisting[language=rexx,label=stringclass.rexx,caption=stringclass.rexx]{stringclass.rexx}
 
-use arg a return \a General problem of making the error message come
-right.
+\emph{We can do all the operators by appeal to classic section 7.}
 
-:method :method :method :method :method :method :method :method :method
-:method :method :method :method :method :method :method :method :method
-:method :method :method :method :method :method :method :method :method
-:method :method
+\lstinputlisting[language=rexx,label=stringmethods,caption=stringmethods]{stringmethods}
 
-:method :method :method :method :method :method :method :method :method
-:method :method :method :method :method :method :method :method :method
-:method :method :method :method :method :method :method :method :method
-:method :method :method :method :method :method :method :method :method
-:method :method :method :method :method
+\emph{General problem of making the error message come right.}
 
-All
+\lstinputlisting[language=rexx,label=stringoperatorsandmethods,caption=stringoperatorsandmethods]{stringoperatorsandmethods}
 
-centre center changestr compare copies counstr datatype delstr delword
-insert lastpos left length overlay pos reverse right space strip substr
-subword translate verify word wordindex wordlength wordpos words abs
-format max
-
-min
-
-sign trunc B2x bitand bitor bitxor C2D
-
-C2x
-
-D2xX ::smethod D2C ::smethod X2B ::smethod X2C ::smethod X2D ::method
-string
-
-::method makestring \#\#\# The array class
+\subsection{The array class}\label{the-array-class}
 
@@
 
-To be done. Dimensionality set at first use. Count commas, not classic
-arg().
+\emph{To be done. Dimensionality set at first use. Count commas, not
+classic \texttt{arg()}.}
 
-::class array
-
-::method new class /* 0 or more WHOLE\textgreater=0 */
+\lstinputlisting[language=rexx,label=arrayclass.rexx,caption=arrayclass.rexx]{arrayclass.rexx}
 
@@
 
-::method of class /* 0 or more ANY */
+\lstinputlisting[language=rexx,label=returnsingleindexarray.rexx,caption=returnsingleindexarray.rexx]{returnsingleindexarray.rexx}
 
 Returns a newly created single-index array containing the specified
-value objects. ::method put /* rANY one or more WHOLE\textgreater0 */
+value objects.
+
+\lstinputlisting[language=rexx,label=makearraymemberitem.rexx,caption=makearraymemberitem.rexx]{makearraymemberitem.rexx}
 
 Makes the object value a member item of the array and associates it with
-the specified index or indexes. ::method ' {[}{]}=' /* 1 or more
-WHOLE\textgreater0O */
+the specified index or indexes.
 
-This method is the same as the PUT method.
+\lstinputlisting[language=rexx,label=operatorputmethodarray.rexx,caption=operatorputmethodarray.rexx]{operatorputmethodarray.rexx}
 
-::method at /* 1 or more WHOLE\textgreater0O */
+This method is the same as the \texttt{PUT} method.
+
+\lstinputlisting[language=rexx,label=arraymethodat.rexx,caption=arraymethodat.rexx]{arraymethodat.rexx}
 
@@
 
-::method `{[}{]}' /* 1 or more WHOLE\textgreater0O */
+\lstinputlisting[language=rexx,label=arraymethodatoperator.rexx,caption=arraymethodatoperator.rexx]{arraymethodatoperator.rexx}
 
@@
 
-::method remove /* 1 or more WHOLE\textgreater0O */
+\lstinputlisting[language=rexx,label=arraymethodremove.rexx,caption=arraymethodremove.rexx]{arraymethodremove.rexx}
 
 Returns and removes the member item with the specified index or indexes
-from the array. ::method hasindex /* 1 or more WHOLE\textgreater0O */
+from the array.
+
+\lstinputlisting[language=rexx,label=arraymethodhasindex.rexx,caption=arraymethodhasindex.rexx]{arraymethodhasindex.rexx}
 
@@
 
-::method items /* (None) \emph{/ Returns the number of items in the
-collection. ::method dimension /} OWHOLE\textgreater0O */
+\lstinputlisting[language=rexx,label=arraymethoditems.rexx,caption=arraymethoditems.rexx]{arraymethoditems.rexx}
+
+Returns the number of items in the collection.
+
+\lstinputlisting[language=rexx,label=arraymethoddimension.rexx,caption=arraymethoddimension.rexx]{arraymethoddimension.rexx}
 
@@
 
-::method size /* (None) */
+\lstinputlisting[language=rexx,label=arraymethodsize.rexx,caption=arraymethodsize.rexx]{arraymethodsize.rexx}
 
@@
 
-::method first /* (None) */
+\lstinputlisting[language=rexx,label=arraymethodfirst.rexx,caption=arraymethodfirst.rexx]{arraymethodfirst.rexx}
 
-Returns the index of the first item in the array, or the NIL object if
-the array is empty.
+Returns the index of the first item in the array, or the \texttt{NIL}
+object if the array is empty.
 
-::method last /* (None) */
+\lstinputlisting[language=rexx,label=arraymethodlast,caption=arraymethodlast]{arraymethodlast}
 
-Returns the index of the last item in the array, or the NIL object if
-the array is empty.
+Returns the index of the last item in the array, or the \texttt{NIL}
+object if the array is empty.
 
-::method next /* rcWHOLE\textgreater O */
+\lstinputlisting[language=rexx,label=arraymethodnext.rexx,caption=arraymethodnext.rexx]{arraymethodnext.rexx}
 
 Returns the index of the item that follows the array item having the
-specified index or returns the NIL object if the item having that index
-is last in the array.
+specified index or returns the \texttt{NIL} object if the item having
+that index is last in the array.
 
-::method previous /* rcWHOLE\textgreater O */
+\lstinputlisting[language=rexx,label=arraymethodprevious.rexx,caption=arraymethodprevious.rexx]{arraymethodprevious.rexx}
 
 Returns the index of the item that precedes the array item having index
-index or the NIL object if the item having that index is first in the
-array.
+index or the \texttt{NIL} object if the item having that index is first
+in the array.
 
-::method makearray /* (None) */
+\lstinputlisting[language=rexx,label=arraymethodmakearray.rexx,caption=arraymethodmakearray.rexx]{arraymethodmakearray.rexx}
 
@@
 is the item corresponding to index start (the first argument) in the
-receiver
+receiver array.
 
-array. ::method supplier /* (None) \emph{/ Returns a supplier object for
-the collection. ::method section /} rcWHOLE\textgreater0O
-oOWHOLE\textgreater=0 */
+\lstinputlisting[language=rexx,label=arraymethodsupplier,caption=arraymethodsupplier]{arraymethodsupplier}
 
-\section{The supplier class}\label{the-supplier-class}
+Returns a supplier object for the collection.
+
+\lstinputlisting[language=rexx,label=arraymethodsection,caption=arraymethodsection]{arraymethodsection}
+
+\subsection{The supplier class}\label{the-supplier-class}
 
 A supplier object enumerates the items a collection contained at the
-time of the supplier's creation. ::class supplier
+time of the supplier's creation.
 
-::method new class /* rANYARRAY rINDEXARRAY */ Returns a new supplier
-object.
+\lstinputlisting[language=rexx,label=supplierclass.rexx,caption=supplierclass.rexx]{supplierclass.rexx}
 
-::method index
+Returns a new supplier object.
 
-Returns the index of the current item in the collection. ::method next
+\lstinputlisting[language=rexx,label=suppliermethodindex.rexx,caption=suppliermethodindex.rexx]{suppliermethodindex.rexx}
+
+Returns the index of the current item in the collection.
+
+\lstinputlisting[language=rexx,label=suppliermethodnext.rexx,caption=suppliermethodnext.rexx]{suppliermethodnext.rexx}
 
@@
 
-::method item
+\lstinputlisting[language=rexx,label=suppliermethoditem.rexx,caption=suppliermethoditem.rexx]{suppliermethoditem.rexx}
 
@@
 
-::method available
+\lstinputlisting[language=rexx,label=supplierindexavailable.rexx,caption=supplierindexavailable.rexx]{supplierindexavailable.rexx}
 
 Returns 1 (true) if an item is available from the supplier (that is, if
-the ITEM method would return a value). Returns 0 (false) otherwise.
+the \texttt{ITEM} method would return a value). Returns 0 (false)
+otherwise.
 
-11.5 The message class ::class message
+\subsection{The message class}\label{the-message-class}
 
-::method init class /* Ugh */
+\lstinputlisting[language=rexx,label=messageclass.rexx,caption=messageclass.rexx]{messageclass.rexx}
 
@@
 
-::method completed
+\lstinputlisting[language=rexx,label=messagecompletedmethod.rexx,caption=messagecompletedmethod.rexx]{messagecompletedmethod.rexx}
 
 Returns 1 if the message object has completed its message; returns 0
-otherwise. ::method notify /* rMESSAGE */
+otherwise.
+
+\lstinputlisting[language=rexx,label=messagebotifymethod.rexx,caption=messagebotifymethod.rexx]{messagebotifymethod.rexx}
 
 Requests notification about the completion of processing for the message
-SEND or START sends. ::method start /* oANY */
+\texttt{SEND} or \texttt{START} sends.
+
+\lstinputlisting[language=rexx,label=messagestartmethod.rexx,caption=messagestartmethod.rexx]{messagestartmethod.rexx}
 
 Sends the message for processing concurrently with continued processing
-of the sender. ::method send /* oANY */
+of the sender.
+
+\lstinputlisting[language=rexx,label=messagesendmethod.rexx,caption=messagesendmethod.rexx]{messagesendmethod.rexx}
 
@@
 
-::method result
+\lstinputlisting[language=rexx,label=methodresult.rexx,caption=methodresult.rexx]{methodresult.rexx}
 
-Returns the result of the message SEND or START sends.
+Returns the result of the message \texttt{SEND} or \texttt{START} sends.

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/built-in_functions.tex
add filenames

@@
 arg(2,`E')=0 then returm /} Default the first argument. \emph{/ if
-arg(1,`E') then Option = translate(left(arg(1),1)) else Option = `N'' /}
-Tf there is no second argument, the current time is returned. \emph{/ if
-arg(2,`E') = 0 then if arg(1,`E') then return ``TIME'(arg(1)) else
+arg(1,`E') then Option = translate(left(arg(1),1)) else Option = `N'\,'
+/} Tf there is no second argument, the current time is returned. \emph{/
+if arg(2,`E') = 0 then if arg(1,`E') then return ``TIME'(arg(1)) else
 return `TIME'() /} One cannot convert to elapsed times. */ if
@@
 
-when InOption == `L'' \textbar{} InOption == `N' then parse var InValue
-HH `:' MM `:' SS when InOption == `M' then MM = InValue otherwise SS =
-InValue end if datatype(HH,`W')=0 \textbar{} datatype(MM,`W')=0
-\textbar{} datatype(SS,`N')=0 then returm HH = HH + HourAdjust /*
-Convert to microseconds \emph{/ Micro = trunc((((HH } 60) + MM) * 60 +
-SS) * 1000000) /* There is no special message for time-out-of-range; the
-bad-format message is used. \emph{/ if Micro\textless0O \textbar{} Micro
-\textgreater{} 24}3600* 1000000 then returm /* Reconvert to further
-check the original. */ if TimeFormat(Micro,InOption) == InValue then
-return TimeFormat(Micro, Option) return
+when InOption == `L'\,' \textbar{} InOption == `N' then parse var
+InValue HH `:' MM `:' SS when InOption == `M' then MM = InValue
+otherwise SS = InValue end if datatype(HH,`W')=0 \textbar{}
+datatype(MM,`W')=0 \textbar{} datatype(SS,`N')=0 then returm HH = HH +
+HourAdjust /* Convert to microseconds \emph{/ Micro = trunc((((HH } 60)
++ MM) * 60 + SS) * 1000000) /* There is no special message for
+time-out-of-range; the bad-format message is used. \emph{/ if
+Micro\textless0O \textbar{} Micro \textgreater{} 24}3600* 1000000 then
+returm /* Reconvert to further check the original. */ if
+TimeFormat(Micro,InOption) == InValue then return TimeFormat(Micro,
+Option) return
 
@@
 Length = \#Bif Arg.3 else Length = max(length (String) +1-Num, 0) if
-\#Bif\_ArgExists.4 then Pad = \#Bif\_ Arg.4 else Pad = ' `! Output =''
+\#Bif\_ArgExists.4 then Pad = \#Bif\_ Arg.4 else Pad = ' `! Output ='\,'
 do Length \#Response Config Substr(String,Num) /* Attempt to fetch

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/configuration.tex
add filenames

@@
 
-We don't want to add undue ``magic'' to this section. It seems we will
-need the concept of a ``reference'' (equivalent to a machine address) so
-that this section can at least have composite objects as arguments. (As
-it already does but these are not Rexx objects)
-
-Possibly we could unify ``reference'' with ``variable pool number''
-since object one-to-one with its variable pool is a fair model. That way
-we don't need a new primitive for comparison of two references.
-
-JAVA is only a ``reference'' for NetRexx so some generalized JAVA-like
-support is needed for that. It would provide the answers to what classes
-were in the context, what their method signatures were etc.
+\emph{We don't want to add undue ``magic'' to this section. It seems we
+will need the concept of a ``reference'' (equivalent to a machine
+address) so that this section can at least have composite objects as
+arguments. (As it already does but these are not Rexx objects)}
+
+\emph{Possibly we could unify ``reference'' with ``variable pool
+number'' since object one-to-one with its variable pool is a fair model.
+That way we don't need a new primitive for comparison of two
+references.}
+
+\emph{JAVA is only a ``reference'' for NetRexx so some generalized
+JAVA-like support is needed for that. It would provide the answers to
+what classes were in the context, what their method signatures were
+etc.}
 
@@
 \item
-  Config\_ is used only from the language processor when processing
-  programs;
-\item
-  API\_is part of the application programming interface and is
+  \texttt{Config\_} is used only from the language processor when
+  processing programs;
+\item
+  \texttt{API\_} is part of the application programming interface and is
   accessible from programs which are not written in the Rexx language;
 \item
-  Trap\_ is not provided by the language processor but may be invoked by
-  the language processor. As its result, each function shall return a
-  completion Response. This is a string indicating how the function
-  behaved. The completion response may be the character `N' indicating
-  the normal behavior occurred; otherwise the first character is an
-  indicator of a different behavior and the remainder shall be suitable
-  as a human-readable description of the function's behavior.
-\end{itemize}
-
-This standard defines any additional results from Config\_ functions as
-made available to the language processor in variables. This does not
-constrain how a particular implementation should return these results.
+  \texttt{Trap\_} is not provided by the language processor but may be
+  invoked by the language processor.
+\end{itemize}
+
+As its result, each function shall return a completion
+\texttt{Response}. This is a string indicating how the function behaved.
+The completion response may be the character
+\texttt{\textquotesingle{}N\textquotesingle{}} indicating the normal
+behavior occurred; otherwise the first character is an indicator of a
+different behavior and the remainder shall be suitable as a
+human-readable description of the function's behavior.
+
+This standard defines any additional results from \texttt{Config\_}
+functions as made available to the language processor in variables. This
+does not constrain how a particular implementation should return these
+results.
 
@@
 
-As alternatives to the normal indicator `N', each function may return a
-completion response with indicator 'X' or `S'; other possible indicators
-are described for each function explicitly. The indicator `X' means that
-the function failed because resources were exhausted. The indicator `S'
-shows that the configuration was unable to perform the function. Certain
-indicators cause conditions to be raised. The possible raising of these
-conditions is implicit in the use of the function; it is not shown
-explicitly when the functions are used in this standard. The implicit
-action is call \#Raise `SYNTAX', Message, Description where: \#Raise
-raises the condition, see nnn. Message is determined by the indicator in
-the completion response. If the indicator is `X' then Message is 5.1. If
-the indicator is `S' then Message is 48.1. Description is the
-description in the completion response. The `SYNTAX' condition 5.1 can
-also be raised by any other activity of the language processor.
+As alternatives to the normal indicator
+\texttt{\textquotesingle{}N\textquotesingle{}}, each function may return
+a completion response with indicator
+\texttt{\textquotesingle{}X\textquotesingle{}} or
+\texttt{\textquotesingle{}S\textquotesingle{}}; other possible
+indicators are described for each function explicitly. The indicator
+\texttt{\textquotesingle{}X\textquotesingle{}} means that the function
+failed because resources were exhausted. The indicator
+\texttt{\textquotesingle{}S\textquotesingle{}} shows that the
+configuration was unable to perform the function.
+
+Certain indicators cause conditions to be raised. The possible raising
+of these conditions is implicit in the use of the function; it is not
+shown explicitly when the functions are used in this standard.
+
+The implicit action is
+
+\lstinputlisting[language=rexx,label=raisesyntax.rexx,caption=raisesyntax.rexx]{raisesyntax.rexx}
+
+where:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{\#Raise} raises the condition, see nnn.
+\item
+  \texttt{Message} is determined by the indicator in the completion
+  response. If the indicator is
+  \texttt{\textquotesingle{}X\textquotesingle{}} then \texttt{Message}
+  is \texttt{5.1}. If the indicator is
+  \texttt{\textquotesingle{}S\textquotesingle{}} then Message is
+  \texttt{48.1}.
+\item
+  \texttt{Description} is the description in the completion response.
+\end{itemize}
+
+The \texttt{\textquotesingle{}SYNTAX\textquotesingle{}} condition
+\texttt{5.1} can also be raised by any other activity of the language
+processor.
 
@@
 The processing initiation interface consists of a function which the
-configuration shall provide to invoke the language processor. We could
-do REQUIRES in a macro-expansion way by adding an argument to
-Contig\_SourceChar to specify the source file. However, I'm assuming we
-will prefer to recursively ``run'' each required file. One of the
-results of that will be the classes and methods made public by that
-REQUIRES subject.
+configuration shall provide to invoke the language processor.
+
+\emph{We could do \texttt{REQUIRES} in a macro-expansion way by adding
+an argument to \texttt{Config\_SourceChar} to specify the source file.
+However, I'm assuming we will prefer to recursively ``run'' each
+required file. One of the results of that will be the classes and
+methods made public by that \texttt{REQUIRES} subject.}
 
@@
 
-API Start(How, Source, Environment, Arguments, Streams, Traps, Provides)
-
-where: How is one of `COMMAND', `FUNCTION', or `SUBROUTINE' and
-indicates how the program is invoked.
-
-What does OOI say for How when running REQUIREd files?
-
-Source is an identification of the source of the program to be
-processed.
-
-Environment is the initial value of the environment to be used in
-processing commands. This has components for the name of the environment
-and how the input and output of commands is to be directed.
-
-Arguments is the initial argument list to be used in processing. This
-has components to specify the number of arguments, which arguments are
-omitted, and the values of arguments that are not omitted.
-
-Streams has components for the default input stream to be used and the
-default output streams to be used.
-
-Traps is the list of traps to be used in processing (see nnn). This has
-components to specify whether each trap is omitted or not.
-
-Semantics:
+\lstinputlisting[language=rexx,label=rexxapistart.rexx,caption=rexxapistart.rexx]{rexxapistart.rexx}
+
+where:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{How} is one of
+  \texttt{\textquotesingle{}COMMAND\textquotesingle{}},
+  \texttt{\textquotesingle{}FUNCTION\textquotesingle{}}, or
+  \texttt{\textquotesingle{}SUBROUTINE\textquotesingle{}} and indicates
+  how the program is invoked.
+\end{itemize}
+
+\emph{What does OOI say for How when running REQUIREd files?}
+
+\begin{itemize}
+\item
+  \texttt{Source} is an identification of the source of the program to
+  be processed.
+\item
+  \texttt{Environment} is the initial value of the environment to be
+  used in processing commands. This has components for the name of the
+  environment and how the input and output of commands is to be
+  directed.
+\item
+  \texttt{Arguments} is the initial argument list to be used in
+  processing. This has components to specify the number of arguments,
+  which arguments are omitted, and the values of arguments that are not
+  omitted.
+\item
+  \texttt{Streams} has components for the default input stream to be
+  used and the default output streams to be used.
+\item
+  \texttt{Traps} is the list of traps to be used in processing (see
+  nnn). This has components to specify whether each trap is omitted or
+  not.
+\end{itemize}
+
+\subsubsection{Semantics:}\label{semantics}
 
@@
 
-If the program was terminated due to a RETURN or EXIT instruction
-without an expression the completion response is `N'.
-
-If the program was terminated due to a RETURN or EXIT instruction with
-an expression the indicator in the completion response is `R' and the
-description of the completion response is the value of the expression.
+If the program was terminated due to a \texttt{RETURN} or \texttt{EXIT}
+instruction without an expression the completion response is
+\texttt{\textquotesingle{}N\textquotesingle{}}.
+
+If the program was terminated due to a \texttt{RETURN} or \texttt{EXIT}
+instruction with an expression the indicator in the completion response
+is \texttt{\textquotesingle{}R\textquotesingle{}} and the description of
+the completion response is the value of the expression.
 
 If the program was terminated due to an error the indicator in the
-completion response is `E' and the description in the completion
-response comprises information about the error that terminated
-processing.
-
-If How was `REQUIRED' and the completion response was not `E', the
-Provides argument is set to reference classes made available. See nnn
-for the semantics of these classes.
+completion response is \texttt{\textquotesingle{}E\textquotesingle{}}
+and the description in the completion response comprises information
+about the error that terminated processing.
+
+If How was \texttt{\textquotesingle{}REQUIRED\textquotesingle{}} and the
+completion response was not
+\texttt{\textquotesingle{}E\textquotesingle{}}, the \texttt{Provides}
+argument is set to reference classes made available. See nnn for the
+semantics of these classes.
 
@@
 \begin{itemize}
-\item
-  syntactic\_characters;
-\item
-  extra\_letters;
-\item
-  other\_blank\_characters;
-\item
-  other\_negators;
-\item
-  other\_characters.
+\tightlist
+\item
+  \emph{syntactic\_characters};
+\item
+  \emph{extra\_letters};
+\item
+  \emph{other\_blank\_characters};
+\item
+  \emph{other\_negators};
+\item
+  \emph{other\_characters}.
 \end{itemize}
@@
 The following characters represent the category of characters called
-syntactic\_characters, identified by their names. The glyphs used to
-represent them in this document are also shown. Syntactic\_characters
-shall be available in every configuration:
+\emph{syntactic\_characters}, identified by their names. The glyphs used
+to represent them in this document are also shown.
+\emph{Syntactic\_characters} shall be available in every configuration:
 
@@
 \item
-  apostrophe, single quotation mark, single quote;
+  ' apostrophe, single quotation mark, single quote;
+\item
+  \begin{itemize}
+  \tightlist
   \item
     asterisk, star;
+  \end{itemize}
 \item
@@
 \item
-  colon;
+  : colon;
 \item
@@
 \item
-  exclamation point, exclamation mark;
-\item
-  \begin{quote}
-  greater-than sign;
-  \end{quote}
-\end{itemize}
-
+  ! exclamation point, exclamation mark;
+\item
+  \textgreater{} greater-than sign;
+\item
+  \begin{itemize}
+  \tightlist
+  \item
     hyphen, minus sign;
-
+  \end{itemize}
+\item
   \textless{} less-than sign;
-
-\begin{itemize}
-\item
-  {[} left bracket, left square bracket; (\_ left parenthesis;
+\item
+  {[} left bracket, left square bracket;
+\item
+  ( left parenthesis;
 \item
@@
 \item
+  \begin{itemize}
+  \tightlist
+  \item
     plus sign;
+  \end{itemize}
 \item
@@
 \item
-  '' quotation mark, double quote; ~reverse slant, reverse solidus,
-  backslash; {]} right bracket, right square bracket;
-\item
-  ) right parenthesis; ; semicolon; /\_ slant, solidus, slash; a-z small
-  letters a through z;
+  '' quotation mark, double quote;
+\item
+  ~reverse slant, reverse solidus, backslash;
+\item
+  {]} right bracket, right square bracket;
+\item
+  ) right parenthesis;
+\item
+  ; semicolon;
+\item
+  / slant, solidus, slash;
+\item
+  a-z small letters a through z;
 \item
@@
 A configuration may have a category of characters in source programs
-called extra\_letters. Extra\_letters are determined by the
-configuration.
+called \emph{extra\_letters}. \emph{Extra\_letters} are determined by
+the configuration.
 
@@
 A configuration may have a category of characters in source programs
-called other\_blank\_characters. Other\_blank\_characters are determined
-by the configuration. Only the following characters represent possible
-characters of this category:
-
-\begin{itemize}
+called \emph{other\_blank\_characters}. \emph{Other\_blank\_characters}
+are determined by the configuration. Only the following characters
+represent possible characters of this category:
+
+\begin{itemize}
+\tightlist
 \item
@@
 A configuration may have a category of characters in source programs
-called other\_negators. Other\_negators are determined by the
-configuration. Only the following characters represent possible
+called \emph{other\_negators}. \emph{Other\_negators} are determined by
+the configuration. Only the following characters represent possible
 characters of this category. The glyphs used to represent them in this
@@
 \item
-  \begin{itemize}
-  \tightlist
-  \item
-    circumflex accent, caret;
-  \end{itemize}
-\item
-  --- not sign.
+  \^{} circumflex accent, caret;
+\item
+  ¬ not sign.
 \end{itemize}
@@
 A configuration may have a category of characters in source programs
-called other\_characters. Other\_characters are determined by the
-configuration.
+called \emph{other\_characters}. \emph{Other\_characters} are determined
+by the configuration.
 
@@
 functions which the configuration shall provide which are concerned with
-the encoding of characters. The following functions shall be provided:
-
-\begin{itemize}
-\item
-  Config\_SourceChar;
-\item
-  Config\_OtherBlankCharacters;
-\item
-  Config\_Upper;
-\item
-  Config\_Compare;
-\item
-  Config\_B2C;
-\item
-  Config\_C2B;
-\item
-  Config\_Substr;
-\item
-  Config\_Length;
-\item
-  Config\_Xrange.
+the encoding of characters.
+
+The following functions shall be provided:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Config\_SourceChar};
+\item
+  \texttt{Config\_OtherBlankCharacters};
+\item
+  \texttt{Config\_Upper};
+\item
+  \texttt{Config\_Compare};
+\item
+  \texttt{Config\_B2C};
+\item
+  \texttt{Config\_C2B};
+\item
+  \texttt{Config\_Substr};
+\item
+  \texttt{Config\_Length};
+\item
+  \texttt{Config\_Xrange}.
 \end{itemize}
@@
 
-Syntax:
-
-Config SourceChar ()
-
-Semantics: Supply the characters of the source program in sequence,
-together with the EOL and EOS events. The EOL event represents the end
-of a line. The EOS event represents the end of the source program. The
-EOS event must only occur immediately after an EOL event. Either a
-character or an event is supplied on each invocation, by setting
-\#Outcome.
+\subsubsection{Syntax:}\label{syntax}
+
+\lstinputlisting[language=rexx,label=configsourcechar.rexx,caption=configsourcechar.rexx]{configsourcechar.rexx}
+
+\subsubsection{Semantics:}\label{semantics-1}
+
+Supply the characters of the source program in sequence, together with
+the \emph{EOL} and \emph{EOS} events. The \emph{EOL} event represents
+the end of a line. The \emph{EOS} event represents the end of the source
+program. The \emph{EOS} event must only occur immediately after an
+\emph{EOL} event. Either a character or an event is supplied on each
+invocation, by setting \texttt{\#Outcome}.
 
@@
 program encoding is incorrect the indicator of the completion response
-is `E' and the description of the completion information is the encoding
-which is incorrect, in hexadecimal notation.
+is \texttt{\textquotesingle{}E\textquotesingle{}} and the description of
+the completion information is the encoding which is incorrect, in
+hexadecimal notation.
 
@@
 
-Syntax: Config OtherBlankCharacters ()
-
-Semantics: Get other\_blank\_characters (see nnn). Set \#Outcome to a
-string of zero or more distinct characters in arbitrary order. Each
-character is one that the configuration considers equivalent to the
-character Blank for the purposes of parsing.
+\subsubsection{Syntax:}\label{syntax-1}
+
+\lstinputlisting[language=rexx,label=configotherblankcharacters.rexx,caption=configotherblankcharacters.rexx]{configotherblankcharacters.rexx}
+
+\subsubsection{Semantics:}\label{semantics-2}
+
+Get \emph{other\_blank\_characters} (see nnn).
+
+Set \texttt{\#Outcome} to a string of zero or more distinct characters
+in arbitrary order. Each character is one that the configuration
+considers equivalent to the character \emph{Blank} for the purposes of
+parsing.
 
@@
 
-Syntax: Config Upper (Character)
-
-where: Character is the character to be translated to uppercase.
-Semantics: Translate Character to uppercase. Set \#Outcome to the
-translated character. Characters which have been subject to this
-translation are referred to as being in uppercase. Config\_Upper applied
-to a character in uppercase must not change the character.
+\subsubsection{Syntax:}\label{syntax-2}
+
+Config\_Upper(Character)
+
+where:
+
+Character is the character to be translated to uppercase.
+
+\subsubsection{Semantics:}\label{semantics-3}
+
+Translate Character to uppercase. Set \#Outcome to the translated
+character. Characters which have been subject to this translation are
+referred to as being in uppercase. Config\_Upper applied to a character
+in uppercase must not change the character.
 
@@
 
-Syntax:
-
-Config Lower (Character) where: Character is the character to be
-translated to lowercase. Semantics: Translate Character to lowercase.
-Set \#Outcome to the translated character. Characters which have been
-subject to this translation are referred to as being in lowercase.
-Config\_Lower applied to a character in lowercase must not change the
-character. Config\_Upper of the outcome of Config\_Lower(Character)
-shall be the original character.
+\subsubsection{Syntax:}\label{syntax-3}
+
+Config\_Lower(Character)
+
+where:
+
+Character is the character to be translated to lowercase.
+
+\subsubsection{Semantics:}\label{semantics-4}
+
+Translate Character to lowercase. Set \#Outcome to the translated
+character. Characters which have been subject to this translation are
+referred to as being in lowercase. Config\_Lower applied to a character
+in lowercase must not change the character. Config\_Upper of the outcome
+of Config\_Lower(Character) shall be the original character.
 
@@
 
-Syntax: Config Compare(Characterl, Character2)
-
-where:
-
-Character1 is the character to be compared with Character2. Character2
-is the character to be compared with Character1.
-
-Semantics:
+\subsubsection{Syntax:}\label{syntax-4}
+
+Config Compare(Characterl, Character2)
+
+where:
+
+Character1 is the character to be compared with Character2.
+
+Character2 is the character to be compared with Character1.
+
+\subsubsection{Semantics:}\label{semantics-5}
 
@@
 \item
-  `lesser' if Character' is less than Character2. The function shall
-  exhibit the following characteristics. If Config \_Compare(a,b)
-  produces
+  `lesser' if Character' is less than Character2.
+\item
+  The function shall exhibit the following characteristics. If Config
+  \_Compare(a,b) produces
 \item
@@
   `equal' then Config\_Compare(a,c) and Config\_Compare(b,c) produce the
-  same value. Syntactic characters which are different characters shall
-  not compare equal by Config\_Compare, see nnn.
+  same value.
+\item
+  Syntactic characters which are different characters shall not compare
+  equal by Config\_Compare, see nnn.
 \end{itemize}
@@
 
-Syntax: Config B2C (Binary) where: Binary is a sequence of digits, each
-`0' or `1'. The number of digits shall be a multiple of eight.
-Semantics:
+\subsubsection{Syntax:}\label{syntax-5}
+
+Config B2C (Binary)
+
+where:
+
+Binary is a sequence of digits, each `0' or `1'. The number of digits
+shall be a multiple of eight.
+
+\subsubsection{Semantics:}\label{semantics-6}
 
@@
 
-Syntax: Config C2B (String)
-
-where: String is a string. Semantics: Translate String to a sequence of
-digits, each `0' or `1'. Set \#Outcome to the result. This function is
-the inverse of Config\_B2C.
+\subsubsection{Syntax:}\label{syntax-6}
+
+Config C2B (String)
+
+where:
+
+String is a string.
+
+\subsubsection{Semantics:}\label{semantics-7}
+
+Translate String to a sequence of digits, each `0' or `1'. Set \#Outcome
+to the result. This function is the inverse of Config\_B2C.
 
@@
 
-Syntax: Config Substr(String, n)
-
-where: String is a string. nis an integer identifying a position within
-String. Semantics: Copy the n-th character from String. The leftmost
-character is the first character. Set Outcome to the resulting
-character. If this function is unable to supply a character because
-there is no n-th character in String the indicator of the completion
-response is `M'. If this function is unable to supply a character
-because the encoding of String is incorrect the indicator of the
-completion response is `E' and the description of the completion
-information is the encoding which is incorrect, in hexadecimal notation.
+\subsubsection{Syntax:}\label{syntax-7}
+
+Config Substr(String, n)
+
+where:
+
+String is a string.
+
+n is an integer identifying a position within String.
+
+\subsubsection{Semantics:}\label{semantics-8}
+
+Copy the n-th character from String. The leftmost character is the first
+character. Set Outcome to the resulting character.
+
+If this function is unable to supply a character because there is no
+n-th character in String the indicator of the completion response is
+`M'.
+
+If this function is unable to supply a character because the encoding of
+String is incorrect the indicator of the completion response is `E' and
+the description of the completion information is the encoding which is
+incorrect, in hexadecimal notation.
 
@@
 
-Syntax: Config Length (String)
-
-where: String is a string. Semantics: Set \#Outcome to the length of the
-string, that is, the number of characters in the string. If this
-function is unable to determine a length because the encoding of String
-is incorrect, the indicator of the completion response is `E' and the
-description of the completion information is the encoding which is
+\subsubsection{Syntax:}\label{syntax-8}
+
+Config Length (String)
+
+where:
+
+String is a string.
+
+\subsubsection{Semantics:}\label{semantics-9}
+
+Set \#Outcome to the length of the string, that is, the number of
+characters in the string.
+
+If this function is unable to determine a length because the encoding of
+String is incorrect, the indicator of the completion response is `E' and
+the description of the completion information is the encoding which is
 incorrect, in hexadecimal notation.
@@
 
-Syntax:
-
-Config Xrange(Characterl, Character2) where: Character1 is the null
-string, or a single character. Character2 is the null string, or a
-single character. Semantics: If Character1 is the null string then let
-LowBound be a lowest ranked character in the character set according to
-the ranking order provided by Config\_Compare; otherwise let LowBound be
-Character1. If Character2 is the null string then let HighBound be a
-highest ranked character in the character set according to the ranking
-order provided by Config\_Compare; otherwise let HighBound be
-Character2lf \#Outcome after Config\_Compare(LowBound,HighBound) has a
-value of
+\subsubsection{Syntax:}\label{syntax-9}
+
+Config Xrange(Characterl, Character2)
+
+where:
+
+Character1 is the null string, or a single character.
+
+Character2 is the null string, or a single character.
+
+\subsubsection{Semantics:}\label{semantics-10}
+
+If Character1 is the null string then let LowBound be a lowest ranked
+character in the character set according to the ranking order provided
+by Config\_Compare; otherwise let LowBound be Character1.
+
+If Character2 is the null string then let HighBound be a highest ranked
+character in the character set according to the ranking order provided
+by Config\_Compare; otherwise let HighBound be Character2lf \#Outcome
+after Config\_Compare(LowBound,HighBound) has a value of
 
@@
 
-Syntax: Config ObjectNew
-
-Semantics:
+\subsubsection{Syntax:}\label{syntax-10}
+
+Config ObjectNew
+
+\subsubsection{Semantics:}\label{semantics-11}
 
@@
 
-Syntax: Config Array Size(Object, size) where: Object is an object. Size
-is an integer greater or equal to 0. Semantics: The configuration should
-prepare to deal efficiently with the object as an array with indexes
-having values up to the value of size.
+\subsubsection{Syntax:}\label{syntax-11}
+
+Config Array Size(Object, size)
+
+where:
+
+Object is an object.
+
+Size is an integer greater or equal to 0.
+
+\subsubsection{Semantics:}\label{semantics-12}
+
+The configuration should prepare to deal efficiently with the object as
+an array with indexes having values up to the value of size.
 
@@
 
-Syntax: Config Array Put(Array, Item, Index)
-
-where: Array is an array. Item is an object Index is an integer greater
-or equal to 1. Semantics: The configuration shall record that the array
-has Item associated with Index.
+\subsubsection{Syntax:}\label{syntax-12}
+
+Config Array Put(Array, Item, Index)
+
+where:
+
+Array is an array.
+
+Item is an object
+
+Index is an integer greater or equal to 1.
+
+\subsubsection{Semantics:}\label{semantics-13}
+
+The configuration shall record that the array has Item associated with
+Index.
 
@@
 
-Syntax: Config Array At(Array, Index)
-
-where: Array is an array. Index is an integer greater or equal to 1.
-Semantics: The configuration shall return the item that the array has
-associated with Index.
+\subsubsection{Syntax:}\label{syntax-13}
+
+Config Array At(Array, Index)
+
+where:
+
+Array is an array.
+
+Index is an integer greater or equal to 1.
+
+\subsubsection{Semantics:}\label{semantics-14}
+
+The configuration shall return the item that the array has associated
+with Index.
 
@@
 
-Syntax: Config Array At(Array, Index) where: Array is an array. Index is
-an integer greater or equal to 1. Semantics: Return `1' if there is an
-item in Array associated with Index, `0' otherwise.
+\subsubsection{Syntax:}\label{syntax-14}
+
+Config Array At(Array, Index)
+
+where:
+
+Array is an array.
+
+Index is an integer greater or equal to 1.
+
+\subsubsection{Semantics:}\label{semantics-15}
+
+Return `1' if there is an item in Array associated with Index, `0'
+otherwise.
 
@@
 
-Syntax: Config Array At(Array, Index)
-
-where: Array is an array. Index is an integer greater or equal to 1.
-Semantics: After this operation, no item is associated with the Index in
-the Array.
+\subsubsection{Syntax:}\label{syntax-15}
+
+Config Array At(Array, Index)
+
+where:
+
+Array is an array.
+
+Index is an integer greater or equal to 1.
+
+\subsubsection{Semantics:}\label{semantics-16}
+
+After this operation, no item is associated with the Index in the Array.
 
@@
 
-Syntax:
-
-Config Command(Environment, Command) where: Environment is the
-environment to be addressed. It has components for:
+\subsubsection{Syntax:}\label{syntax-16}
+
+Config Command(Environment, Command)
+
+where:
+
+Environment is the environment to be addressed. It has components for:
 
@@
   stream. There is an indication of whether writing is to APPEND or
-  REPLACE. Command is the command to be executed. Semantics: Perform a
-  command.
+  REPLACE.
+\end{itemize}
+
+Command is the command to be executed.
+
+\subsubsection{Semantics:}\label{semantics-17}
+
+Perform a command.
+
+\begin{itemize}
 \item
@@
 The external routines interface consists of a function which the
-configuration shall provide to invoke external routines. See nnn and nnn
-for a description of the language features that use external routines.
+configuration shall provide to invoke external routines.
+
+See nnn and nnn for a description of the language features that use
+external routines.
 
@@
 
-Syntax: Config ExternalRoutine(How, NameType, Name, Environment,
-Arguments, Streams, Traps) where: How is one of `FUNCTION' or
-`SUBROUTINE' and indicates how the external routine is to be invoked.
+\subsubsection{Syntax:}\label{syntax-17}
+
+Config ExternalRoutine(How, NameType, Name, Environment, Arguments,
+Streams, Traps) where:
+
+How is one of `FUNCTION' or `SUBROUTINE' and indicates how the external
+routine is to be invoked.
 
@@
 
-Semantics:
+\subsubsection{Semantics:}\label{semantics-18}
 
@@
 the ultimate external call is to a method. Perhaps combine
-Config\_ExternalRoutine with Config\_ExternalMethod. There is a
-terminology clash on ``environment''. Perhaps easiest to change the
-classic to ``address\_environment''. (And make it part of new
-``environment''?) There are terminology decisions to make about
-``files'', ``programs'', and ``packages''. Possibly ``program'' is the
-thing you run (and we don't say what it means physically), ``file'' is a
-unit of scope (ROUTINEs in current file before those in REQUIREd), and
-``package'' we don't use (since a software package from a shop would
-probably have several files but not everything to run a program.) Using
-``file'' this way may not be too bad since we used ``stream'' rather
-than ``tile'' in the classic definition. The How parameter will need
-`METHOD' as a value. Should API\_Start also allow 'METHOD''. If we pass
-the new Environment we don't have to pass Streams separately.
-
-Text of Config\_ExternalMethod waiting on such decisions. Syntax:
+Config\_ExternalRoutine with Config\_ExternalMethod.
+
+There is a terminology clash on ``environment''. Perhaps easiest to
+change the classic to ``address\_environment''. (And make it part of new
+``environment''?)
+
+There are terminology decisions to make about ``files'', ``programs'',
+and ``packages''. Possibly ``program'' is the thing you run (and we
+don't say what it means physically), ``file'' is a unit of scope
+(ROUTINEs in current file before those in REQUIREd), and ``package'' we
+don't use (since a software package from a shop would probably have
+several files but not everything to run a program.) Using ``file'' this
+way may not be too bad since we used ``stream'' rather than ``tile'' in
+the classic definition.
+
+The How parameter will need `METHOD' as a value. Should API\_Start also
+allow 'METHOD''. If we pass the new Environment we don't have to pass
+Streams separately.
+
+Text of Config\_ExternalMethod waiting on such decisions.
+
+\subsubsection{Syntax:}\label{syntax-18}
 
 Config ExternalMethod (How, NameType, Name, Environment, Arguments,
-Streams, Traps) where: How is one of `FUNCTION' or `SUBROUTINE' and
-indicates how the external routine is to be invoked. NameType is a
-specification of whether the name was provided as a symbol or as a
-string literal. Name is the name of the routine to be invoked.
+Streams, Traps)
+
+where:
+
+How is one of `FUNCTION' or `SUBROUTINE' and indicates how the external
+routine is to be invoked.
+
+NameType is a specification of whether the name was provided as a symbol
+or as a string literal.
+
+Name is the name of the routine to be invoked.
+
 Environment is an environment value with the same components as on API\_
-Start. Arguments is a specification of the arguments to the routine,
-with the same components as on API\_Start. Streams is a specification of
-the default streams, with the same components as on API\_ Start. Traps
-is the list of traps to be used in processing, with the same components
-as on API\_ Start. Semantics: Invoke an external routine. Set \{Outcome
-to the result of the external routine, or set the indicator of the
-completion response to `D' if the external routine did not provide a
-result. If this function is unable to locate the routine the indicator
-of the completion response is `U'. As a result SYNTAX condition 43.1 is
-raised implicitly. If How indicated that a result from the routine was
-required but the routine did not provide one the indicator of the
-completion response is `H'. As a result SYNTAX condition 44.1 is raised
-implicitly. If How indicated that a result from the routine was required
-but the routine provided one that was too long (see \#Limit\_String in
-nnn) the indicator of the completion response is `L'. As a result SYNTAX
-condition 52 is raised implicitly. If the routine failed in a way not
-indicated by some other indicator the indicator of the completion
-response is `F'. As a result SYNTAX condition 40.1 is raised implicitly.
-5.8 External data queue The external data queue interface consists of
-functions which the configuration shall provide to manipulate an
-external data queue mechanism. See nnn, nnn, nnn, nnn, and nnn for a
-description of language features that use the external data queue. The
-configuration shall provide an external data queue mechanism. The
-following functions shall be provided:
+Start.
+
+Arguments is a specification of the arguments to the routine, with the
+same components as on API\_Start.
+
+Streams is a specification of the default streams, with the same
+components as on API\_ Start.
+
+Traps is the list of traps to be used in processing, with the same
+components as on API\_ Start.
+
+\subsubsection{Semantics:}\label{semantics-19}
+
+Invoke an external routine. Set \{Outcome to the result of the external
+routine, or set the indicator of the completion response to `D' if the
+external routine did not provide a result.
+
+If this function is unable to locate the routine the indicator of the
+completion response is `U'. As a result SYNTAX condition 43.1 is raised
+implicitly.
+
+If How indicated that a result from the routine was required but the
+routine did not provide one the indicator of the completion response is
+`H'. As a result SYNTAX condition 44.1 is raised implicitly.
+
+If How indicated that a result from the routine was required but the
+routine provided one that was too long (see \#Limit\_String in nnn) the
+indicator of the completion response is `L'. As a result SYNTAX
+condition 52 is raised implicitly.
+
+If the routine failed in a way not indicated by some other indicator the
+indicator of the completion response is `F'. As a result SYNTAX
+condition 40.1 is raised implicitly.
+
+\section{External data queue}\label{external-data-queue}
+
+The external data queue interface consists of functions which the
+configuration shall provide to manipulate an external data queue
+mechanism.
+
+See nnn, nnn, nnn, nnn, and nnn for a description of language features
+that use the external data queue. The configuration shall provide an
+external data queue mechanism. The following functions shall be
+provided:
 
@@
 
-Syntax: Config Push(String)
-
-where: String is the value to be retained in the external data queue.
-Semantics: Add String as an item to the end of the external data queue
+\subsubsection{Syntax:}\label{syntax-19}
+
+Config Push(String)
+
+where:
+
+String is the value to be retained in the external data queue.
+
+\subsubsection{Semantics:}\label{semantics-20}
+
+Add String as an item to the end of the external data queue from which
+Config\_Pull will remove an item.
+
+\subsection{Contig\_Queue}\label{contig_queue}
+
+\subsubsection{Syntax:}\label{syntax-20}
+
+Config Queue (String)
+
+where:
+
+String is the value to be retained in the external data queue.
+
+\subsubsection{Semantics:}\label{semantics-21}
+
+Add String as an item to the opposite end of the external data queue
 from which Config\_Pull will remove an item.
 
-\subsection{Contig\_Queue}\label{contig_queue}
-
-Syntax: Config Queue (String)
-
-where: String is the value to be retained in the external data queue.
-Semantics: Add String as an item to the opposite end of the external
-data queue from which Config\_Pull will remove an item.
-
 \subsection{Config\_Pull}\label{config_pull}
 
-Syntax: Config Pull()
-
-Semantics: Retrieve an item from the end of the external data queue to
-which Config\_Push adds an element to the list. Set \#Outcome to the
-value of the retrieved item. If no item could be retrieved the indicator
-of the completion response is `F'.
+\subsubsection{Syntax:}\label{syntax-21}
+
+Config Pull()
+
+\subsubsection{Semantics:}\label{semantics-22}
+
+Retrieve an item from the end of the external data queue to which
+Config\_Push adds an element to the list. Set \#Outcome to the value of
+the retrieved item.
+
+If no item could be retrieved the indicator of the completion response
+is `F'.
 
@@
 
-Syntax: Config Queued ()
-
-Semantics:
+\subsubsection{Syntax:}\label{syntax-22}
+
+Config Queued ()
+
+\subsubsection{Semantics:}\label{semantics-23}
 
@@
 shall provide to manipulate streams. See nnn, nnn, and nnn for a
-description of language features which use streams. Streams are
-identified by names and provide for the reading and writing of data.
-They shall support the concepts of characters, lines, positioning,
-default input stream and default output stream. The concept of a
-persistent stream shall be supported and the concept of a transient
-stream may be supported. A persistent stream is one where the content is
-not expected to change except when the stream is explicitly acted on. A
-transient stream is one where the data available is expected to vary
-with time.
+description of language features which use streams.
+
+Streams are identified by names and provide for the reading and writing
+of data. They shall support the concepts of characters, lines,
+positioning, default input stream and default output stream.
+
+The concept of a persistent stream shall be supported and the concept of
+a transient stream may be supported. A persistent stream is one where
+the content is not expected to change except when the stream is
+explicitly acted on. A transient stream is one where the data available
+is expected to vary with time.
 
 The concepts of binary and character streams shall be supported. The
-content of a character stream is expected to be characters. The null
-string is used as a name for both the default input stream and the
-default output stream. The null string names the default output stream
-only when it is an argument to the Config\_Stream\_Charout operation.
+content of a character stream is expected to be characters.
+
+The null string is used as a name for both the default input stream and
+the default output stream. The null string names the default output
+stream only when it is an argument to the Config\_Stream\_Charout
+operation.
 
@@
 \item
-  Config\_Stream\_Count. The results of these functions are described in
-  terms of the following stems with tails which are stream names:
+  Config\_Stream\_Count.
+\item
+  The results of these functions are described in terms of the following
+  stems with tails which are stream names:
 \item
@@
 
-Syntax:
-
-Config Stream Charin(Stream, OperationType) where: Stream is the name of
-the stream to be processed. OperationType is one of `CHARIN', `LINEIN',
-or `NULL. Semantics: Read from a stream. Increase
-\#Linein\_Position.Stream by one when the end-of-line indication is
-encountered. Increase \#Charin\_Position.Stream when the indicator will
-be 'N'. If OperationType is `CHARIN' the state variables describing the
-stream will be affected as follows: - when the configuration is able to
-provide data from a transient stream or the character at position
+\subsubsection{Syntax:}\label{syntax-23}
+
+Config Stream Charin(Stream, OperationType)
+
+where:
+
+Stream is the name of the stream to be processed.
+
+OperationType is one of `CHARIN', `LINEIN', or 'NULL.
+
+\subsubsection{Semantics:}\label{semantics-24}
+
+Read from a stream. Increase \#Linein\_Position.Stream by one when the
+end-of-line indication is encountered. Increase
+\#Charin\_Position.Stream when the indicator will be `N'.
+
+If OperationType is `CHARIN' the state variables describing the stream
+will be affected as follows: - when the configuration is able to provide
+data from a transient stream or the character at position
 \#Charin\_Position.Stream of a persistent stream then \#Outcome shall be
-set to contain the data. The indicator of the response shall be `N';
+set to contain the data.
+
+The indicator of the response shall be `N';
 
@@
   or `1'. The choice is decided by the configuration. The eight
-  character sequence indicates a binary stream, see nnn. If
-  OperationType is `LINEIN' then the action is the same as if Operation
+  character sequence indicates a binary stream, see nnn.
+\end{itemize}
+
+If OperationType is `LINEIN' then the action is the same as if Operation
 had been `CHARIN' with the following additional possibility. If
@@
 an embedded indication of the end-of-line is skipped. The characters
-  skipped contribute to the change of \#Charin\_Position.Stream.
-  \#Outcome is the null string. If OperationType is `NULL' then the
-  stream is accessed but no data is read.
-\end{itemize}
+skipped contribute to the change of \#Charin\_Position.Stream. \#Outcome
+is the null string.
+
+If OperationType is `NULL' then the stream is accessed but no data is
+read.
 
@@
 
-Syntax:
-
-Config Stream Position(Stream, OperationType, Position) where: Stream is
-the name of the stream to be processed. Operation is `CHARIN', `LINEIN',
-`CHAROUT', or `LINEOUT'. Position indicates where to position the
-stream. Semantics: If the operation is `CHARIN' or `CHAROUT' then
-Position is a character position, otherwise Position is a line position.
+\subsubsection{Syntax:}\label{syntax-24}
+
+Config Stream Position(Stream, OperationType, Position)
+
+where:
+
+Stream is the name of the stream to be processed.
+
+Operation is `CHARIN', `LINEIN', `CHAROUT', or `LINEOUT'.
+
+Position indicates where to position the stream.
+
+\subsubsection{Semantics:}\label{semantics-25}
+
+If the operation is `CHARIN' or `CHAROUT' then Position is a character
+position, otherwise Position is a line position.
+
 If Operation is `CHARIN' or `LINEIN' and the Position is beyond the
@@
 Position. Set \#Linein\_Position.Stream to correspond with this
-position. If Operation is `CHAROUT' or `LINEOQUT' and the Position is
-more than one beyond the limit of existing data then the indicator of
-the response shall be `R'. Otherwise if Operation is `CHAROUT' or
-`LINEOUT' then \#Charout\_Position.Stream is set to the position at
-which the next Config\_Stream\_Charout on the stream shall write, as
-indicated by Position. Set \#Lineout\_Position.Stream to correspond with
-this position. If this function is unable to position the stream because
-the stream is transient then the indicator of the completion response
-shall be `T'.
+position.
+
+If Operation is `CHAROUT' or `LINEOQUT' and the Position is more than
+one beyond the limit of existing data then the indicator of the response
+shall be `R'. Otherwise if Operation is `CHAROUT' or `LINEOUT' then
+\#Charout\_Position.Stream is set to the position at which the next
+Config\_Stream\_Charout on the stream shall write, as indicated by
+Position. Set \#Lineout\_Position.Stream to correspond with this
+position.
+
+If this function is unable to position the stream because the stream is
+transient then the indicator of the completion response shall be `T'.
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-25}
 
@@
 stream to be processed. Command is a configuration-specific command to
-be performed against the stream. Semantics: Issue a
-configuration-specific command against a stream. This may affect all
-state variables describing Stream which hold position information. It
-may alter the effect of any subsequent operation on the specified
+be performed against the stream.
+
+\subsubsection{Semantics:}\label{semantics-26}
+
+Issue a configuration-specific command against a stream. This may affect
+all state variables describing Stream which hold position information.
+It may alter the effect of any subsequent operation on the specified
 stream. If the indicator is set to `N', \#Outcome shall be set to
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-26}
 
 Config Stream State (Stream) where: Stream is the name of the stream to
-be queried. Semantics: Set the indicator to reflect the state of the
-stream. Return an indicator equal to the indicator that an immediately
-subsequent Config\_Stream\_Charin(Stream, `CHARIN') would return.
-Alternatively, return an indicator of `U'.
+be queried.
+
+\subsubsection{Semantics:}\label{semantics-27}
+
+Set the indicator to reflect the state of the stream. Return an
+indicator equal to the indicator that an immediately subsequent
+Config\_Stream\_Charin(Stream, `CHARIN') would return. Alternatively,
+return an indicator of `U'.
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-27}
 
@@
 will be eight characters, each `0' or `1', otherwise Data will be a
-single character. Semantics: When Data is the null string, no data is
-written. Otherwise write to the stream. The state variables describing
-the stream will be affected as follows:
+single character.
+
+\subsubsection{Semantics:}\label{semantics-28}
+
+When Data is the null string, no data is written. Otherwise write to the
+stream. The state variables describing the stream will be affected as
+follows:
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-28}
 
 Config Stream Qualified (Stream) where: Stream is the name of the stream
-to be processed. Semantics: Set \#Outcome to some name which identifies
-Stream. Return a completion response with indicator `B' if the argument
-is not acceptable to the configuration as identifying a stream.
+to be processed.
+
+\subsubsection{Semantics:}\label{semantics-29}
+
+Set \#Outcome to some name which identifies Stream. Return a completion
+response with indicator `B' if the argument is not acceptable to the
+configuration as identifying a stream.
 
@@
 
-Syntax:
-
-Config Stream Unique () Semantics: Set \#Outcome to a name that the
-configuration recognizes as a stream name. The name shall not be a name
-that the configuration associates with any existing data.
+\subsubsection{Syntax:}\label{syntax-29}
+
+Config Stream Unique ()
+
+\subsubsection{Semantics:}\label{semantics-30}
+
+Set \#Outcome to a name that the configuration recognizes as a stream
+name. The name shall not be a name that the configuration associates
+with any existing data.
 
@@
 
-Syntax: Config Stream Query (Stream) where: Stream is the name of the
-stream to be queried. Semantics: Set \#Outcome to `B' if the stream is a
-binary stream, or to `C' if it is a character stream.
+\subsubsection{Syntax:}\label{syntax-30}
+
+Config Stream Query (Stream) where: Stream is the name of the stream to
+be queried.
+
+\subsubsection{Semantics:}\label{semantics-31}
+
+Set \#Outcome to `B' if the stream is a binary stream, or to `C' if it
+is a character stream.
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-31}
 
 Config Stream Close (Stream) where: Stream is the name of the stream to
-be closed. Semantics: \#Charout\_Position.Stream and
-\#Lineout\_Position.Stream are set to 1 unless the stream has existing
-data, in which case they are set ready to write immediately after the
-existing data. If this function is unable to position the stream because
-the stream is transient then the indicator of the completion response
-shall be `T'.
+be closed.
+
+\subsubsection{Semantics:}\label{semantics-32}
+
+\#Charout\_Position.Stream and \#Lineout\_Position.Stream are set to 1
+unless the stream has existing data, in which case they are set ready to
+write immediately after the existing data. If this function is unable to
+position the stream because the stream is transient then the indicator
+of the completion response shall be `T'.
 
@@
 
-Syntax: Config Stream Count (Stream, Operation, Option) where: Stream is
-the name of the stream to be counted. Operation is `CHARS', or `LINES'.
-
-Option is `N' or `C'. Semantics: If the option is `N', \#Outcome is set
-to zero if:
+\subsubsection{Syntax:}\label{syntax-32}
+
+Config Stream Count (Stream, Operation, Option) where: Stream is the
+name of the stream to be counted. Operation is `CHARS', or `LINES'.
+
+Option is `N' or `C'.
+
+\subsubsection{Semantics:}\label{semantics-33}
+
+If the option is `N', \#Outcome is set to zero if:
 
@@
 configuration shall provide to manipulate variables in external variable
-pools. See nnn for the VALUE built-in function which uses external
-variable pools. The configuration shall provide an external variable
-pools mechanism. The following functions shall be provided:
+pools.
+
+See nnn for the VALUE built-in function which uses external variable
+pools.
+
+The configuration shall provide an external variable pools mechanism.
+The following functions shall be provided:
 
@@
 
-Syntax: Config Get (Poolid, Name) where: Poolid is an identification of
-the external variable pool. Name is the name of a variable. Semantics:
+\subsubsection{Syntax:}\label{syntax-33}
+
+Config Get (Poolid, Name)
+
+where:
+
+Poolid is an identification of the external variable pool.
+
+Name is the name of a variable.
+
+\subsubsection{Semantics:}\label{semantics-34}
+
 Get the value of a variable with name Name in the external variable pool
-Poolid. Set Outcome to this value. If Poolid does not identify an
-external pool provided by this configuration, the indicator of the
-completion response is `P'. If Name is not a valid name of a variable in
-the external pool, the indicator of the completion response is `F'.
+Poolid. Set Outcome to this value.
+
+If Poolid does not identify an external pool provided by this
+configuration, the indicator of the completion response is `P'.
+
+If Name is not a valid name of a variable in the external pool, the
+indicator of the completion response is `F'.
 
@@
 
-Syntax: Config Set (Poolid, Name, Value)
-
-where: Poolid is an identification of the external variable pool. Name
-is the name of a variable. Value is the value to be assigned to the
-variable. Semantics: Set a variable with name Name in the external
-variable pool Poolid to Value. If Poolid does not identify an external
-pool provided by this configuration, the indicator of the completion
-response is `P'. If Name is not a valid name of a variable in the
-external pool, the indicator of the completion response is `F'.
+\subsubsection{Syntax:}\label{syntax-34}
+
+Config Set (Poolid, Name, Value)
+
+where:
+
+Poolid is an identification of the external variable pool.
+
+Name is the name of a variable.
+
+Value is the value to be assigned to the variable.
+
+\subsubsection{Semantics:}\label{semantics-35}
+
+Set a variable with name Name in the external variable pool Poolid to
+Value.
+
+If Poolid does not identify an external pool provided by this
+configuration, the indicator of the completion response is `P'.
+
+If Name is not a valid name of a variable in the external pool, the
+indicator of the completion response is `F'.
 
@@
 
-Syntax:
-
-Config Constants () Semantics: Set the values of the following state
-variables:
+\subsubsection{Syntax:}\label{syntax-35}
+
+Config Constants ()
+
+\subsubsection{Semantics:}\label{semantics-36}
+
+Set the values of the following state variables:
 
@@
 configuration shall provide which provide functions for a language
-processor. The following functions shall be provided:
-
-\begin{itemize}
+processor.
+
+The following functions shall be provided:
+
+\begin{itemize}
+\tightlist
 \item
@@
 
-Syntax: Config Trace Query ()
-
-Semantics: Indicate whether external activity is requesting interactive
-tracing. Set \#Outcome to `Yes' if interactive tracing is currently
-requested. Otherwise set \#Outcome to `No'.
+\subsubsection{Syntax:}\label{syntax-36}
+
+Config Trace Query ()
+
+\subsubsection{Semantics:}\label{semantics-37}
+
+Indicate whether external activity is requesting interactive tracing.
+Set \#Outcome to `Yes' if interactive tracing is currently requested.
+Otherwise set \#Outcome to `No'.
 
@@
 
-Syntax: Config Trace Input ()
-
-Semantics: Set \#Outcome to a value from the source of trace input. The
-source of trace input is determined by the configuration.
+\subsubsection{Syntax:}\label{syntax-37}
+
+Config Trace Input ()
+
+\subsubsection{Semantics:}\label{semantics-38}
+
+Set \#Outcome to a value from the source of trace input. The source of
+trace input is determined by the configuration.
 
@@
 
-Syntax: Config Trace Output (Line)
-
-where: Line is a string. Semantics:
+\subsubsection{Syntax:}\label{syntax-38}
+
+Config Trace Output (Line)
+
+where:
+
+Line is a string.
+
+\subsubsection{Semantics:}\label{semantics-39}
 
@@
 
-Syntax: Config Default Input ()
-
-Semantics: Set \#Outcome to the value that LINEIN( ) would return.
+\subsubsection{Syntax:}\label{syntax-39}
+
+Config Default Input ()
+
+\subsubsection{Semantics:}\label{semantics-40}
+
+Set \#Outcome to the value that LINEIN( ) would return.
 
@@
 
-Syntax: Config Default Output (Line)
-
-where: Line is a string. Semantics: Write the string as a line in the
-manner of LINEOUT( ,Line).
+\subsubsection{Syntax:}\label{syntax-40}
+
+Config Default Output (Line)
+
+where:
+
+Line is a string.
+
+\subsubsection{Semantics:}\label{semantics-41}
+
+Write the string as a line in the manner of LINEOUT( ,Line).
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-41}
 
@@
 
-Semantics: This function is provided only as a counterpart to
-Trap\_Initialization; in itself it does nothing except return the
-response. An indicator of `F' gives rise to Msg3.1.
+\subsubsection{Semantics:}\label{semantics-42}
+
+This function is provided only as a counterpart to Trap\_Initialization;
+in itself it does nothing except return the response. An indicator of
+`F' gives rise to Msg3.1.
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-42}
 
@@
 
-Semantics: This function is provided only as a counterpart to
-Trap\_Termination; in itself it does nothing except return the response.
-An indicator of `F' gives rise to Msg2.1.
+\subsubsection{Semantics:}\label{semantics-43}
+
+This function is provided only as a counterpart to Trap\_Termination; in
+itself it does nothing except return the response. An indicator of `F'
+gives rise to Msg2.1.
 
@@
 
-Syntax: Config Halt Query ()
-
-Semantics: Indicate whether external activity has requested a HALT
-condition to be raised. Set \#Outcome to `Yes if HALT is requested.
-Otherwise set \#Outcome to 'No'.
-
-5.12.9 Config \_Halt\_Reset
-
-Syntax: Config Halt Reset ()
-
-Semantics:
+\subsubsection{Syntax:}\label{syntax-43}
+
+Config Halt Query ()
+
+\subsubsection{Semantics:}\label{semantics-44}
+
+Indicate whether external activity has requested a HALT condition to be
+raised. Set \#Outcome to `Yes if HALT is requested. Otherwise set
+\#Outcome to 'No'.
+
+\subsection{Config \_Halt\_Reset}\label{config-_halt_reset}
+
+\subsubsection{Syntax:}\label{syntax-44}
+
+Config Halt Reset ()
+
+\subsubsection{Semantics:}\label{semantics-45}
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-45}
 
@@
 
-Semantics: Indicate whether the source of the program may or may not be
-output by the language processor. Set \#NoSource to `1' to indicate that
-the source of the program may not be output by the language processor,
-at various points in processing where it would otherwise be output.
-Otherwise, set \#NoSource to `0'. A configuration shall allow any
-program to be processed in such a way that Config\_NoSource() sets
-\#NoSource to `0'. A configuration may allow any program to be processed
-in such a way that Config\_NoSource() sets \#NoSource to `1'.
+\subsubsection{Semantics:}\label{semantics-46}
+
+Indicate whether the source of the program may or may not be output by
+the language processor. Set \#NoSource to `1' to indicate that the
+source of the program may not be output by the language processor, at
+various points in processing where it would otherwise be output.
+Otherwise, set \#NoSource to `0'.
+
+A configuration shall allow any program to be processed in such a way
+that Config\_NoSource() sets \#NoSource to `0'. A configuration may
+allow any program to be processed in such a way that Config\_NoSource()
+sets \#NoSource to `1'.
 
@@
 
-Syntax:
+\subsubsection{Syntax:}\label{syntax-46}
 
@@
 
-Semantics: Get a time stamp. Set \#Time to a string whose value is the
-integer number of microseconds that have elapsed between 00:00:00 on
-January first 0001 and the time that Config\_Time is called, at
-longitude zero. Values sufficient to allow for any date in the year 9999
-shall be supported. The value returned may be an approximation but shall
-not be smaller than the value returned by a previous use of the
-function.
+\subsubsection{Semantics:}\label{semantics-47}
+
+Get a time stamp. Set \#Time to a string whose value is the integer
+number of microseconds that have elapsed between 00:00:00 on January
+first 0001 and the time that Config\_Time is called, at longitude zero.
+Values sufficient to allow for any date in the year 9999 shall be
+supported. The value returned may be an approximation but shall not be
+smaller than the value returned by a previous use of the function.
 
@@
 
-Syntax: Config Random Seed (Seed)
-
-where: Seed is a sequence of up to \#Bif\_Digits. RANDOM digits.
-Semantics: Set a seed, so that subsequent uses of Config\_Random\_Next
-will reproducibly return quasi-random numbers.
+\subsubsection{Syntax:}\label{syntax-47}
+
+Config Random Seed (Seed)
+
+where:
+
+Seed is a sequence of up to \#Bif\_Digits. RANDOM digits.
+
+\subsubsection{Semantics:}\label{semantics-48}
+
+Set a seed, so that subsequent uses of Config\_Random\_Next will
+reproducibly return quasi-random numbers.
 
@@
 
-Syntax:
-
-Config Random Next (Min, Max) where: Min is the lower bound, inclusive,
-on the number returned in \#Outcome. Max is the upper bound, inclusive,
-on the number returned in \#Outcome. Semantics: Set \#Outcome to a
-quasi-random nonnegative integer in the range Min to Max.
+\subsubsection{Syntax:}\label{syntax-48}
+
+Config Random Next (Min, Max)
+
+where:
+
+Min is the lower bound, inclusive, on the number returned in \#Outcome.
+
+Max is the upper bound, inclusive, on the number returned in \#Outcome.
+
+\subsubsection{Semantics:}\label{semantics-49}
+
+Set \#Outcome to a quasi-random nonnegative integer in the range Min to
+Max.
 
@@
 
-Syntax: Config Options (String) where: String is a string. Semantics: No
-effect beyond the effects common to all Config\_ invocations. The value
-of the string will have come from an OPTIONS instruction, see nnn.
+\subsubsection{Syntax:}\label{syntax-49}
+
+Config Options (String)
+
+where:
+
+String is a string.
+
+\subsubsection{Semantics:}\label{semantics-50}
+
+No effect beyond the effects common to all Config\_ invocations. The
+value of the string will have come from an OPTIONS instruction, see nnn.
 
@@
 invoked; otherwise the invocation of the trap replaces the potential
-invocation of the Config\_ function. In the rest of this standard, the
-trapping mechanism is not shown explicitly. It is implied by the use of
-a Config\_ function. The names of the traps are
-
-\begin{itemize}
+invocation of the Config\_ function.
+
+In the rest of this standard, the trapping mechanism is not shown
+explicitly. It is implied by the use of a Config\_ function.
+
+The names of the traps are
+
+\begin{itemize}
+\tightlist
 \item
@@
 
-Syntax: API Set(Symbol, Value) where: Symbol is a symbol. Value is the
-string whose value is to be assigned to the variable. Semantics: Assign
-the value of Value to the variable identified by Symbol. If Symbol
-contains no periods or contains one period as its last character: Var\_
-Set(\#Pool, Symbol, `0', Value) Otherwise: Var \_Set(\#Pool, \#Symbol,
-`1', Value) where: \#Symbol is Symbol after any replacements in the tail
-as described by nnn.
+\subsubsection{Syntax:}\label{syntax-50}
+
+API Set(Symbol, Value)
+
+where:
+
+Symbol is a symbol.
+
+Value is the string whose value is to be assigned to the variable.
+
+\subsubsection{Semantics:}\label{semantics-51}
+
+Assign the value of Value to the variable identified by Symbol. If
+Symbol contains no periods or contains one period as its last character:
+Var\_ Set(\#Pool, Symbol, `0', Value) Otherwise: Var \_Set(\#Pool,
+\#Symbol, `1', Value) where: \#Symbol is Symbol after any replacements
+in the tail as described by nnn.
 
@@
 
-Syntax: API Value (Symbol)
-
-where: Symbol is a symbol. Semantics: Return the value of the variable
-identified by Symbol. If Symbol contains no periods or contains one
+\subsubsection{Syntax:}\label{syntax-51}
+
+API Value (Symbol)
+
+where:
+
+Symbol is a symbol.
+
+\subsubsection{Semantics:}\label{semantics-52}
+
+Return the value of the variable identified by Symbol. If Symbol
+contains no periods or contains one
 
@@
 Otherwise the value of \#Outcome after: Var Value(\#Pool, \#Symbol, `1')
-where: \#Symbol is Symbol after any replacements in the tail as
-described by nnn.
+
+where:
+
+\#Symbol is Symbol after any replacements in the tail as described by
+nnn.
 
@@
 
-Syntax: API Drop (Symbol)
-
-where: Symbol is a symbol. Semantics: Drop the variable identified by
-Symbol. If Symbol contains no periods or contains one period as its last
-character: Var Drop(\#Pool, Symbol, `0') Otherwise: Var Drop(\#Pool,
-\#Symbol, `1') where: \#Symbol is Symbol after any replacements in the
-tail as described by nnn.
+\subsubsection{Syntax:}\label{syntax-52}
+
+API Drop (Symbol)
+
+where:
+
+Symbol is a symbol.
+
+\subsubsection{Semantics:}\label{semantics-53}
+
+Drop the variable identified by Symbol. If Symbol contains no periods or
+contains one period as its last character: Var Drop(\#Pool, Symbol, `0')
+Otherwise: Var Drop(\#Pool, \#Symbol, `1') where: \#Symbol is Symbol
+after any replacements in the tail as described by nnn.
 
@@
 
-Syntax: API SetDirect (Symbol, Value)
+\subsubsection{Syntax:}\label{syntax-53}
+
+API SetDirect (Symbol, Value)
 
 where: Symbol is a direct symbol. Value is the string whose value is to
-be assigned to the variable. Semantics:
+be assigned to the variable.
+
+\subsubsection{Semantics:}\label{semantics-54}
 
@@
 
-Syntax: API ValueDirect (Symbol)
-
-where: Symbol is a direct symbol. Semantics: Return the value of the
-variable identified by Symbol. If the Symbol contains no period: Var
-\_Value(\#Pool, Symbol, `0') Otherwise: Var \_Value(\#Pool, Symbol, `1')
+\subsubsection{Syntax:}\label{syntax-54}
+
+API ValueDirect (Symbol)
+
+where: Symbol is a direct symbol.
+
+\subsubsection{Semantics:}\label{semantics-55}
+
+Return the value of the variable identified by Symbol. If the Symbol
+contains no period: Var \_Value(\#Pool, Symbol, `0') Otherwise: Var
+\_Value(\#Pool, Symbol, `1')
 
@@
 
-Syntax: API DropDirect (Symbol)
-
-where: Symbol is a direct symbol. Semantics:
+\subsubsection{Syntax:}\label{syntax-55}
+
+API DropDirect (Symbol)
+
+where: Symbol is a direct symbol.
+
+\subsubsection{Semantics:}\label{semantics-56}
 
@@
 
-Syntax: API ValueOther (Qualifier) where: Qualifier is an indication
+\subsubsection{Syntax:}\label{syntax-56}
+
+API ValueOther (Qualifier) where: Qualifier is an indication
 distinguishing the result to be returned including any necessary further
-qualification. Semantics: Return characteristics of the program,
-depending on the value of Qualifier. The possibilities for the value to
-be returned are:
+qualification.
+
+\subsubsection{Semantics:}\label{semantics-57}
+
+Return characteristics of the program, depending on the value of
+Qualifier. The possibilities for the value to be returned are:
 
@@
 
-Syntax: API Next ()
-
-Semantics:
+\subsubsection{Syntax:}\label{syntax-57}
+
+API Next ()
+
+\subsubsection{Semantics:}\label{semantics-58}
 
@@
 
-\subsection{API NextVariable}\label{api-nextvariable}
-
-Syntax: API NextVariable()
-
-Semantics:
+\subsection{API\_NextVariable}\label{api_nextvariable}
+
+\subsubsection{Syntax:}\label{syntax-58}
+
+\lstinputlisting[language=rexx,label=apinextvariable.rexx,caption=apinextvariable.rexx]{apinextvariable.rexx}
+
+\subsubsection{Semantics:}\label{semantics-59}
 

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/conformance.tex
work in standard book production process including readme.md
upd
add filenames

@@
-\hypertarget{conformance}{%
-\chapter{Conformance}\label{conformance}}
+%preprocessed texin
+\chapter{Conformance}\label{conformance}
 
-\hypertarget{conformance-1}{%
-\section{Conformance}\label{conformance-1}}
+\section{Conformance}\label{conformance-1}
 
@@
 \item
-  alist of all definitions or values for the features in this standard
+  a list of all definitions or values for the features in this standard
   which are specified to be dependent on the configuration.
@@
 
-\hypertarget{limits}{%
-\section{Limits}\label{limits}}
+\section{Limits}\label{limits}
 
@@
 three. Other limits shall be one of the numbers one, five or twenty
-five, or any of these multiplied by some power of ten. Limitations that
-conforming language processors may impose are:
+five, or any of these multiplied by some power of ten.
+
+Limitations that conforming language processors may impose are:
 
@@
 \item
-  NUMERIC DIGITS values shall be supported up to a value of at least
-  nine hundred and ninety nine.
+  \texttt{NUMERIC\ DIGITS} values shall be supported up to a value of at
+  least nine hundred and ninety nine.
 \item

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/definitions_and_document_notation.tex
add filenames

@@
 
-Lots more for NetRexx
+\emph{Lots more for NetRexx}
 
@@
 The expressions (separated by commas) between the parentheses of a
-function call or following the name on a CALL instruction. Also the
-corresponding values which may be accessed by a function or routine,
+function call or following the name on a \texttt{CALL} instruction. Also
+the corresponding values which may be accessed by a function or routine,
 however invoked.
@@
 of external routines and commands, and the results of some built-in
-functions, such as D2C.
+functions, such as \texttt{D2C}.
 \item[command]
@@
 \item[condition]
-A specific event, or state, which can be trapped by CALL ON or SIGNAL
-ON.
+A specific event, or state, which can be trapped by \texttt{CALL\ ON} or
+\texttt{SIGNAL\ ON}.
 \item[configuration]
@@
 \item[construct]
-A named syntax grouping, for example ``expression'', ``do\_
-specification''.
+A named syntax grouping, for example ``\emph{expression}'',
+``\emph{do\_specification}''.
 \item[default error stream]
@@
 \item[encoding]
-The relation between a character string anda corresponding number. The
+The relation between a character string and a corresponding number. The
 encoding of character strings is determined by the configuration.
@@
 Commands are sent to the current named environment, initially selected
-externally but then alterable by using the ADDRESS instruction.
+externally but then alterable by using the \texttt{ADDRESS} instruction.
 \item[error number]
@@
 A named variable pool supplied by the configuration which can be
-accessed by the VALUE built-in function.
+accessed by the \texttt{VALUE} built-in function.
 \item[function]
@@
 \item[internal routine]
-A function or subroutine that is in the same program as the CALL
-instruction or function call that invokes it.
+A function or subroutine that is in the same program as the
+\texttt{CALL} instruction or function call that invokes it.
 \item[keyword]

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/directives.tex
work in standard book production process including readme.md
upd
add filenames

@@
-\hypertarget{directives}{%
-\chapter{Directives}\label{directives}}
+%preprocessed texin
+\chapter{Directives}\label{directives}
 
@@
 
-The BNF\_primary referenced may be directly in the production or in some
-component referenced in the
+The \texttt{BNF\_primary} referenced may be directly in the production
+or in some component referenced in the \emph{production}, recursively.
+The components are considered in left to right order.
 
-production, recursively. The components are considered in left to right
-order. \#Contains (Identifier, BNF primary)
+\lstinputlisting[language=rexx,label=directivescontainsidentifier.rexx,caption=directivescontainsidentifier.rexx]{directivescontainsidentifier.rexx}
 
@@
 
-Identifier is an identifier in a production (see nnn) defined in nnn.
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Identifier} is an \emph{identifier} in a \emph{production}
+  (see nnn) defined in nnn.
+\item
+  \texttt{BNF\_primary} is a \emph{bnf\_primary} (see nnn) in a
+  \emph{production} defined in nnn.
+\end{itemize}
 
-BNF\_primary is a bnf\_primary (see nnn) in a production defined in nnn.
-Return `1' if the production identitied by identifier contained a
-bnf\_primary identified by BNF\_primary, otherwise return `0'.
+Return \texttt{\textquotesingle{}1\textquotesingle{}} if the
+\emph{production} identitied by \emph{Identifier} contained a
+\emph{bnf\_primary} identified by \texttt{BNF\_primary}, otherwise
+return \texttt{\textquotesingle{}0\textquotesingle{}}.
 
-\#Instance (Identifier, BNF primary) where: Identifier is an identifier
-in a production defined in nnn. BNF\_primary is a Onf\_primary in a
-production defined in nnn. Returns the content of the particular
-instance of the BNF\_primary. If the BNF\_primary is a VAR\_SYMBOL this
-is referred to as the symbol ``taken as a constant.''
+\lstinputlisting[language=rexx,label=directivesinstance.rexx,caption=directivesinstance.rexx]{directivesinstance.rexx}
 
-\#Evaluate (Identifier, BNF primary) where: Identifier is an identifier
-in a production defined in nnn. BNF\_primary is a Onf\_primary in a
-production defined in nnn. Return the value of the BNF\_primary in the
-production identified by Identifier.
+where:
 
-\#Execute (Identifier, BNF primary) where: Identifier is an identifier
-in a production defined in nnn. BNF\_primary is a Onf\_primary in a
-production defined in nnn. Perform the instructions identified by the
-BNF\_primary in the production identified by Identifier.
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Identifier} is an \emph{identifier} in a \emph{production}
+  defined in nnn.
+\item
+  \texttt{BNF\_primary} is a \emph{bnf\_primary} in a \emph{production}
+  defined in nnn.
+\end{itemize}
 
-\#Parses (Value, BNF primary) where: Value is a string BNF\_primary is a
-Onf\_primary in a production defined in nnn. Return `1' if Value matches
-the definition of the BNF\_primary, by the rules of clause 6, `0'
-otherwise.
+Returns the content of the particular instance of the
+\texttt{BNF\_primary}. If the \texttt{BNF\_primary} is a
+\texttt{VAR\_SYMBOL} this is referred to as the symbol ``taken as a
+constant.''
 
-\#Clause (Label) where: Label is a label in code used by this standard
-to describe processing. Return an identification of that label. The
-value of this identification is used only by the \#Goto notation
-function.
+\lstinputlisting[language=rexx,label=directivesevaluate.rexx,caption=directivesevaluate.rexx]{directivesevaluate.rexx}
 
-\#Goto (Value) where:
+where:
 
-Value identifies a label in code used by this standard to describe
-processing. The description of processing continues at the identified
-label.
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Identifier} is an \emph{identifier} in a \emph{production}
+  defined in nnn.
+\item
+  \texttt{BNF\_primary} is a \emph{bnf\_primary} in a \emph{production}
+  defined in nnn.
+\end{itemize}
 
-\#Retry ()
+Return the value of the \texttt{BNF\_primary} in the \emph{production}
+identified by \texttt{Identifier}.
+
+\lstinputlisting[language=rexx,label=directivesexecute.rexx,caption=directivesexecute.rexx]{directivesexecute.rexx}
+
+where:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Identifier} is an \emph{identifier} in a \emph{production}
+  defined in nnn.
+\item
+  \texttt{BNF\_primary} is a \emph{bnf\_primary} in a \emph{production}
+  defined in nnn.
+\end{itemize}
+
+Perform the instructions identified by the \texttt{BNF\_primary} in the
+\emph{production} identified by \texttt{Identifier}.
+
+\lstinputlisting[language=rexx,label=directivesparses.rexx,caption=directivesparses.rexx]{directivesparses.rexx}
+
+where:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Value} is a string
+\item
+  \texttt{BNF\_primary} is a \emph{bnf\_primary} in a \emph{production}
+  defined in nnn.
+\end{itemize}
+
+Return \texttt{\textquotesingle{}1\textquotesingle{}} if Value matches
+the definition of the \texttt{BNF\_primary}, by the rules of clause 6,
+\texttt{\textquotesingle{}0\textquotesingle{}} otherwise.
+
+\lstinputlisting[language=rexx,label=directivesclause.rexx,caption=directivesclause.rexx]{directivesclause.rexx}
+
+where:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Label} is a label in code used by this standard to describe
+  processing.
+\end{itemize}
+
+Return an identification of that label. The value of this identification
+is used only by the \texttt{\#Goto} notation function.
+
+\lstinputlisting[language=rexx,label=directivesgoto.rexx,caption=directivesgoto.rexx]{directivesgoto.rexx}
+
+where:
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{Value} identifies a label in code used by this standard to
+  describe processing.
+\end{itemize}
+
+The description of processing continues at the identified label.
+
+\lstinputlisting[language=rexx,label=directivesrexxtry.rexx,caption=directivesrexxtry.rexx]{directivesrexxtry.rexx}
 
@@
 effect of transferring execution to the beginning of that clause, with
-state variable \#Loop set to the value it had when that clause was
-previously executed.
+state variable \texttt{\#Loop} set to the value it had when that clause
+was previously executed.
 
-\hypertarget{initializing}{%
-\section{Initializing}\label{initializing}}
+\section{Initializing}\label{initializing}
 
-Some of the initializing, now grouped in classic section 8.2.1 will have
-to come here so that we have picked up anything from the START\_API that
-needs to be passed on to the execution of REQUIRES subject.
+\emph{Some of the initializing, now grouped in classic section 8.2.1
+will have to come here so that we have picked up anything from the
+\texttt{START\_API} that needs to be passed on to the execution of
+REQUIRES subject. We will be using some operations that are forward
+reference to what was section nnn.}
 
-We will be using some operations that are forward reference to what was
-section nnn.
+\subsection{Program initialization and message
+texts}\label{program-initialization-and-message-texts}
 
-\hypertarget{program-initialization-and-message-texts}{%
-\subsection{Program initialization and message
-texts}\label{program-initialization-and-message-texts}}
+Processing of a program begins when \texttt{API\_Start} is executed. A
+pool becomes current for the reserved variables.
 
-Processing of a program begins when API\_Start is executed. A pool
-becomes current for the reserved
+\lstinputlisting[language=rexx,label=initialization.rexx,caption=initialization.rexx]{initialization.rexx}
 
-variables. call Config ObjectNew \#ReservedPool = \#Outcome \#Pool =
-\#ReservedPool Is it correct to make the reserved variables and the
-builtin objects in the same pool?
+\emph{Is it correct to make the reserved variables and the builtin
+objects in the same pool?}
 
 Some of the values which affect processing of the program are parameters
-of API\_ Start:
+of \texttt{API\_Start}: \texttt{\#Howlnvoked} is set to
+\texttt{\textquotesingle{}COMMAND\textquotesingle{}},
+\texttt{\textquotesingle{}FUNCTION\textquotesingle{}} or
+\texttt{\textquotesingle{}SUBROUTINE\textquotesingle{}} according to the
+first parameter of \texttt{API\_Start}.
 
-\#Howlnvoked is set to `COMMAND', `FUNCTION' or `SUBROUTINE' according
-to the first parameter of APL Start.
+\texttt{\#Source} is set to the value of the second parameter of
+\texttt{API\_Start}.
 
-\#Source is set to the value of the second parameter of API\_ Start.
+The third parameter of \texttt{API\_Start} is used to determine the
+initial active environment.
 
-The third parameter of API\_Start is used to determine the initial
-active environment.
-
-The fourth parameter of API Start is used to determine the arguments.
-For each argument position \#ArgExists.1.ArgNumber is set `1' if there
-is an argument present, `0' if not. ArgNumber is the number of the
-argument position, counting from 1. If \#ArgExists.1.ArgNumber is `1'
-then \#Arg.1.ArgNumber is set to the value of the corresponding
-argument. If \#ArgExists.1.ArgNumber is `0' then \#Arg.1.Arg is set to
-the null string. \#ArgExists.1.0 is set to the largest n for which
-\#ArgExists.1.n is `1', or to zero if there is no such value of n.
+The fourth parameter of \texttt{API\_Start} is used to determine the
+arguments. For each argument position \texttt{\#ArgExists.1.ArgNumber}
+is set \texttt{\textquotesingle{}1\textquotesingle{}} if there is an
+argument present, \texttt{\textquotesingle{}0\textquotesingle{}} if not.
+\texttt{ArgNumber} is the number of the argument position, counting from
+\texttt{1}. If \texttt{\#ArgExists.1.ArgNumber} is
+\texttt{\textquotesingle{}1\textquotesingle{}} then
+\texttt{\#Arg.1.ArgNumber} is set to the value of the corresponding
+argument. If \texttt{\#ArgExists.1.ArgNumber} is
+\texttt{\textquotesingle{}0\textquotesingle{}} then \texttt{\#Arg.1.Arg}
+is set to the null string. \texttt{\#ArgExists.1.0} is set to the
+largest \texttt{n} for which \texttt{\#ArgExists.1.n} is
+\texttt{\textquotesingle{}1\textquotesingle{}}, or to zero if there is
+no such value of \texttt{n}.
 
 Some of the values which affect processing of the program are provided
-by the configuration: call Config OtherBlankCharacters
+by the configuration:
 
-\#A11Blanks\textless Index ``\#Al1Blanks'' \# ``\,'' \textgreater{} = '
-'\#Outcome /* ``Real'' blank concatenated with others */
+\lstinputlisting[language=rexx,label=configotherblankcharacters.rexx,caption=configotherblankcharacters.rexx]{configotherblankcharacters.rexx}
+our model are only distinquished by the values within their pool so we
+can construct the builtin classes incomplete and then complete them with
+directives.}
 
-\#Bif Digits. = 9
+\emph{Can we initialize the methods of .nil by directives?}
 
-call Config Constants
-
--true = `1'
-
--false = `0O'
-
-Objects in our model are only distinquished by the values within their
-pool so we can construct the builtin classes incomplete and then
-complete them with directives.
-
-Can we initialize the methods of .nil by directives?
-
-call Config ObjectNew
-
--List = \#Outcome
-
-call var\_set .List, \#IsClass, `0', `1'
-
-call var\_set .List, \#ID, `0', `List'
+\lstinputlisting[language=rexx,label=configobjectnew.rexx,caption=configobjectnew.rexx]{configobjectnew.rexx}
 
@@
 the text of error messages. The relation between message numbers and
-message text is defined by the following list, where the message
+message text is defined by the following list, where the message number
+appears immediately before an
+\texttt{\textquotesingle{}=\textquotesingle{}} and the message text
+follows in quotes.
 
-number appears immediately before an `=' and the message text follows in
-quotes.
+\lstinputlisting[language=rexx,label=configerrortexts1.rexx,caption=configerrortexts1.rexx]{configerrortexts1.rexx}
 
-\#ErrorText. stl
+\emph{Unsound now we are using `term'?}
 
-\#ErrorText.0.1 = `Error running , line :'
-
-\#ErrorText.0.2 = `Error in interactive trace:'
-
-\#ErrorText.0.3 = `Interactive trace. ``Trace Off'' to end debug.',
-``ENTER to continue.'
-
-\#ErrorText.2 = `Failure during finalization'
-
-\#ErrorText.2.1 = `Failure during finalization: '
-
-\#ErrorText.3 \#ErrorText.3.1
-
-`Failure during initialization' `Failure during initialization: '
-
-\#ErrorText.4 \#ErrorText.4.1
-
-`Program interrupted' `Program interrupted with HALT condition: '
-
-\#ErrorText.5 \#ErrorText.5.1
-
-`System resources exhausted' `System resources exhausted: '
-
-\#ErrorText.6 \#ErrorText.6.1
-
-'Unmatched ``/\emph{'' or quote! 'Unmatched comment delimiter (''/}'')!
-
-\#ErrorText.6.2 \#ErrorText.6.3
-
-\#ErrorText.7 = \#ErrorText.7.1 =
-
-\#ErrorText.7.2 =
-
-\#ErrorText.7.3 =
-
-\#ErrorText.8 = \#ErrorText.8.1 \#ErrorText.8.2
-
-\#ErrorText.9 = \#ErrorText.9.1 \#ErrorText.9.2
-
-\#ErrorText.10 \#ErrorText.10.1= \#ErrorText.10.2=
-
-\#ErrorText.10.3=
-
-\#ErrorText.10.4= \#ErrorText.10.5= \#ErrorText.10.6=
-
-\#ErrorText.13 = \#ErrorText.13.1=
-
-\#ErrorText.14 = \#ErrorText.14.1= \#ErrorText.14.2= \#ErrorText.14.3=
-\#ErrorText.14.4=
-
-\#ErrorText.15 = \#ErrorText.15.1=
-
-\#ErrorText.15.2= \#ErrorText.15.3= \#ErrorText.15.4= \#ErrorText.16 =
-\#ErrorText.16.1= \#ErrorText.16.2= \#ErrorText.16.3=
-
-\#ErrorText.17 = \#ErrorText.17.1=
-
-\#ErrorText.17.2=
-
-\#ErrorText.18 =
-
-\#ErrorText.18.1=
-
-\#ErrorText.18.2=
-
-``Unmatched single quote (`)'' `Unmatched double quote (``)'
-
-`WHEN or OTHERWISE expected' ``SELECT on line `found ``''\,'
-
-``SELECT on line `or END; found ``''\,'!
-
-requires WHEN;',
-
-`All WHEN expressions of SELECT on line are',
-
-`false; OTHERWISE expected'
-
-``Unexpected THEN or ELSE'
-
-`THEN has no corresponding IF or WHEN clause' `ELSE has no corresponding
-THEN clause'
-
-``Unexpected WHEN or OTHERWISE' `WHEN has no corresponding SELECT'
-`OTHERWISE has no corresponding SELECT'
-
-`Unexpected or unmatched END'
-
-`END has no corresponding DO or SELECT'
-
-`END corresponding to DO on line ', `must have a symbol following that
-matches', `the control variable (or no symbol);',
-
-`found ``''\,'!
-
-`END corresponding to DO on line ', `must not have a symbol following it
-because', `there is no control variable;',
-
-`found ``''\,'!
-
-`END corresponding to SELECT on line ', `must not have a symbol
-following;',
-
-`found ``''\,'!
-
-`END must not immediately follow THEN'
-
-`END must not immediately follow ELSE'
-
-`Invalid character in program' `Invalid character ``('`X)'' in program'
-
-``Incomplete DO/SELECT/IF'
-
-`DO instruction requires a matching END' `SELECT instruction requires a
-matching END' `THEN requires a following instruction' `ELSE requires a
-following instruction'
-
-`Invalid hexadecimal or binary string'
-
-`Invalid location of blank in position', ` in hexadecimal string'
-
-`Invalid location of blank in position', ` in binary string'
-
-`Only 0-9, a-f, A-F, and blank are valid in a', ``hexadecimal string;
-found''``\,'
-
-`Only 0, 1, and blank are valid in a',
-
-`binary string; found ``''\,'
-
-`Label not found'
-
-`Label ``'' not found'
-
-`Cannot SIGNAL to label ``'' because it is', `inside an IF, SELECT or DO
-group'
-
-`Cannot invoke label ``'' because it is', `inside an IF, SELECT or DO
-group'
-
-`Unexpected PROCEDURE'
-
-``PROCEDURE is valid only when it is the first', `instruction executed
-after an internal CALL', `or function invocation'
-
-`The EXPOSE `instruction executed after a method invocation!' `THEN
-expected'
-
-`IF keyword on line requires', `matching THEN clause; found ``''\,'
-
-`WHEN keyword on line requires',
-
-requires WHEN, OTHERWISE,',
-
-instruction is valid only when it is the first', \#ErrorText.19 =
-\#ErrorText.19.1=
-
-\#ErrorText.19.2= \#ErrorText.19.3= \#ErrorText.19.4= \#ErrorText.19.6=
-\#ErrorText.19.7= \#ErrorText.19.8=
-
-\#ErrorText.19.9=
-
-\#ErrorText.19.11='String or symbol
-
-\#ErrorText.19.12='String or symbol
-
-\#ErrorText.19.13=
-
-\#ErrorText.19.15='String or symbol
-
-\#ErrorText.19.16='String or symbol
-
-\#ErrorText.19.17=
-
-Unsound now we are using ``term'?
-
-65
-
-\#ErrorText.20 = \#ErrorText.20.1= \#ErrorText.20.2= \#ErrorText.20.3=
-
-\#ErrorText.21 \#ErrorText.21.1
-
-\#ErrorText.22 = \#ErrorText.22.1=
-
-\#ErrorText.23 = \#ErrorText.23.1=
-
-\#ErrorText.24 = \#ErrorText.24.1= \#ErrorText.25 = \#ErrorText.25.1=
-\#ErrorText.25.2= \#ErrorText.25.3= \#ErrorText.25.4= \#ErrorText.25.5=
-\#ErrorText.25.6= \#ErrorText.25.7=
-
-\#ErrorText.25.8=
-
-`matching THEN clause; found ``''\,' `String or symbol expected' `String
-or symbol expected after ADDRESS keyword;', ``found''``! `String or
-symbol expected after CALL keyword;',''found ``''! `String or symbol
-expected after NAME keyword;', ``found''``! `String or symbol expected
-after SIGNAL keyword;',''found ``''! `String or symbol expected after
-TRACE keyword;', ``found''``! `Symbol expected in parsing
-pattern;',''found ``''! `String or symbol expected after REQUIRES;',
-``found''``! `String or symbol expected after METHOD;',''found ``''!
-
-expected after ROUTINE;', ``found''``!
-
-expected after CLASS;', ``found''``! `String or symbol expected after
-INHERIT;',''found ``''!
-
-expected after METACLASS;', ``found''``!
-
-expected after MIXINCLASS;`, ``found''``! `String or symbol expected
-after SUBCLASS;',''found ``''! 'Name expected' `Name required; found
-``''\,'
-
-`Found ``'' where only a name is valid'
-
-'Found ``'' where only a name or
-
-`Invalid data on end of clause' `The clause ended at an unexpected
-token;',
-
-`found ``''\,'
-
-'Invalid ``Invalid
-
-'Invalid ``Invalid
-
-'Invalid
-
-character string' character string
-
-data string' data string
-
-TRACE request'
-
-is valid'
-
-''xX''
-
-`'! xX''
-
-`TRACE request letter must be one of',
-
-' ``ACEFILNOR'';
-
-found
-
-``evalue\textgreater{}'''
-
-`Invalid sub-keyword found' `CALL ON must be followed by one of the',
-
-'keywords ;
-
-found
-
-``etoken\textgreater{}''!
-
-`CALL OFF must be followed by one of the',
-
-'keywords ;
-
-found
-
-``etoken\textgreater{}''!
-
-``SIGNAL ON must be followed by one of the',
-
-'keywords ;
-
-found
-
-``etoken\textgreater{}''!
-
-`SIGNAL OFF must be followed by one of the',
-
-'keywords ;
-
-found
-
-``etoken\textgreater{}''!
-
-``ADDRESS WITH must be followed by one of the',
-
-'keywords ;
-
-found
-
-``etoken\textgreater{}''!
-
-``INPUT must be followed by one of the',
-
-'keywords ; 'OUTPUT must be followed by 'keywords ; ``APPEND must be
-followed by 'keywords ;
-
-found
-
-found
-
-found
-
-``etoken\textgreater{}''! one of the', ``etoken\textgreater{}''! one of
-the', ``etoken\textgreater{}''!
-
-\#ErrorText.25.9=
-
-\#ErrorText.25.11= \#ErrorText.25.12= \#ErrorText.25.13=
-\#ErrorText.25.14= \#ErrorText.25.15= \#ErrorText.25.16=
-\#ErrorText.25.17=
-
-\#ErrorText.25.18=
-
-\#ErrorText.26 \#ErrorText.26.1=
-
-\#ErrorText.26.2=
-
-\#ErrorText.26.3=
-
-\#ErrorText.26.4= \#ErrorText.26.5=
-
-\#ErrorText.26.6=
-
-\#ErrorText.26.7= \#ErrorText.26.8=
-
-\#ErrorText.26.11
-
-\#ErrorText.26.12
-
-\#ErrorText.27 \#ErrorText.27.1=
-
-\#ErrorText.28 \#ErrorText.28.1= \#ErrorText.28.2= \#ErrorText.28.3=
-
-\#ErrorText.28.4=
-
-\#ErrorText.29 \#ErrorText.29.1=
-
-\#ErrorText.30 \#ErrorText.30.1= \#ErrorText.30.2=
-
-\#ErrorText.31 \#ErrorText.31.1=
-
-\#ErrorText.31.2=
-
-\#ErrorText.31.3=
-
-``REPLACE must be followed by one of the', `keywords ; found
-``''\,'\,''NUMERIC FORM must be followed by one of the', `keywords ;
-found ``''\,'
-
-``PARSE must be followed by one of the', `keywords ; found ``''\,'
-
-``UPPER must be followed by one of the', `keywords ; found ``''\,'
-
-``ERROR must be followed by one of the', `keywords ; found
-``''\,'\,''NUMERIC must be followed by one of the', `keywords ; found
-``''\,' `FOREVER must be followed by one of the', ``keywords or nothing;
-found''``'''PROCEDURE must be followed by the keyword', ``EXPOSE or
-nothing; found''``\,'
-
-``FORWARD must be followed by one of the the keywords',
-
-`; found ``''\,'
-
-`Invalid whole number'
-
-`Whole numbers must fit within current DIGITS', `setting(); found
-``''\,'
-
-`Value of repetition count expression in DO instruction',
-
-`must be zero or a positive whole number;', `found ``''\,'
-
-`Value of FOR expression in DO instruction', `must be zero or a positive
-whole number;', `found ``''\,'
-
-``Positional pattern of parsing template', `must be a whole number;
-found ``''\,'\,''NUMERIC DIGITS value',
-
-`must be a positive whole number; ``NUMERIC FUZZ value',
-
-`must be zero or a positive whole number;',
-
-`found ``''\,'
-
-`Number used in TRACE setting',
-
-`must be a whole number; found ``''\,'
-
-'Operand to right of the power operator (``**``)`, 'must be a whole
-number; found ``''\,'
-
-``Result of \% operation would need',
-
-found ``'''
-
-`exponential notation at current NUMERIC DIGITS '
-
-``Result of \% operation used for // `, `operation would need',
-
-`exponential notation at current NUMERIC DIGITS '
-
-`Invalid DO syntax' `Invalid use of keyword ``'' in DO clause'
-
-`Invalid LEAVE or ITERATE'
-
-`LEAVE is valid only within a repetitive DO loop' `ITERATE is valid only
-within a repetitive DO loop' `Symbol following LEAVE (``'') must',
-
-`either match control variable of a current',
-
-`DO loop or be omitted'
-
-`Symbol following ITERATE (``'') must', `either match control variable
-of a current',
-
-`DO loop or be omitted'
-
-`Environment name too long' `Environment name exceeds',
-
-\#Limit EnvironmentName `characters; found ``''\,' `Name or string too
-long'
-
-`Name exceeds' \#Limit Name `characters' ``Literal string exceeds'
-\#Limit Literal `characters' `Name starts with number or ``.''\,'!
-
-`A value cannot be assigned to a number;',
-
-`found ``''\,'
-
-`Variable symbol must not start with a number;', `found ``''\,'
-
-`Variable symbol must not start with a ``.'';',
-
-`found ``''\,' \#ErrorText.33 = \#ErrorText.33.1=
-
-\#ErrorText.33.2=
-
-\#ErrorText.33.3=
-
-\#ErrorText.34 = \#ErrorText.34.1=
-
-`Invalid expression result' `Value of NUMERIC DIGITS (``'')',
-
-'must exceed value of NUMERIC FUZZ
-
-``(\textless¢value\textgreater)'''!
-
-`Value of NUMERIC DIGITS (``'')',
-
-`must not exceed'
-
-\#Limit Digits
-
-``Result of expression following NUMERIC FORM',
-
-'must start with ``E''
-
-or ``S''; found ``'''
-
-``Logical value not''0'' or ``1''!
-
-'Value of expression
-
-following IF keyword',
-
-`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.2= `Value
-of expression following WHEN keyword',
-
-`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.3= `Value
-of expression following WHILE keyword',
-
-`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.4= `Value
-of expression following UNTIL keyword',
-
-`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.5= `Value
-of expression to left',
-
-`of logical operator ``''\,',
-
-`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.34.6= `Value
-of expression to right',
-
-`of logical operator ``''\,',
-
-`must be exactly ``0'' or ``1''; found ``''\,' \#ErrorText.35 = `Invalid
-expression' \#ErrorText.35.1= `Invalid expression detected at ``''\,'
-
-\#ErrorText.36
-
-`Unmatched ``('' in expression'
-
-\#ErrorText.37 = \#ErrorText.37.1= \#ErrorText.37.2=
-
-'Unexpected n a n ny nt
-
-`Unexpected ``,''! `Unmatched ``)'' in expression'
-
-or
-
-\#ErrorText.38 = \#ErrorText.38.1= \#ErrorText.38.2= \#ErrorText.38.3=
-
-`Invalid template or pattern'
-
-`Invalid parsing template detected at ``''\,' `Invalid parsing position
-detected at ``''\,' ``PARSE VALUE instruction requires WITH keyword'
-
-``Incorrect call to routine'
-
-`External routine ``'' failed'
-
-`Not enough arguments in invocation of ;', `minimum expected is '
-
-`Too many arguments in invocation of ;', `maximum expected is '
-
-`Missing argument in invocation of ;', `argument is required'
-`ebif\textgreater{} argument ',
-
-`exponent exceeds' \#Limit ExponentDigits ``found''``\,'
-
-` argument ',
-
-`must be a number; found ``''\,'! ` argument ',
-
-\#ErrorText.40 = \#ErrorText.40.1= \#ErrorText.40.3= \#ErrorText.40.4=
-\#ErrorText.40.5=
-
-\#ErrorText.40.9= `digits;',
-
-\#ErrorText.40.11=
-
-\#ErrorText.40.12=
-
-`must be a whole number; found ``''\,' \#ErrorText.40.13=` argument ',
-
-`must be zero or positive; found ``''\,' \#ErrorText.40.14=` argument ',
-
-`must be positive; found ``''\,'
-
-\#ErrorText.40.17=` argument 1', `must have an integer part in the range
-0:90 and a', `decimal part no larger than .9; found ``''\,'
-\#ErrorText.40.18=` conversion must',
-
-``have a year in the range 0001 to 9999! \#ErrorText.40.19=` argument 2,
-``'', is not in the format',
-
-`described by argument 3, ``''\,'
-
-\#ErrorText.40.21=` argument must not be null' \#ErrorText.40.23=`
-argument ',
-
-`must be a single character; found ``''\,' \#ErrorText.40.24=` argument
-1',
-
-`must be a binary string; found ``''\,' \#ErrorText.40.25=` argument 1',
-
-`must be a hexadecimal string; found ``''\,' \#ErrorText.40.26=`
-argument 1',
-
-`must be a valid symbol; found ``''\,' \#ErrorText.40.27=` argument 1',
-
-`must be a valid stream name; found ``''\,' \#ErrorText.40.28=` argument
-,',
-
-`option must start with one of ``'';',
-
-`found ``''\,'
-
-\#ErrorText.40.29=` conversion to format ``'' is not allowed'
-\#ErrorText.40.31=` argument 1 (``'') must not exceed 100000'
-\#ErrorText.40.32=` the difference between argument 1 (``'') and',
-
-`argument 2 (``'') must not exceed 100000' \#ErrorText.40.33=` argument
-1 (``'') must be less than',
-
-`or equal to argument 2 (``'')' \#ErrorText.40.34=` argument 1 (``'')
-must be less than',
-
-`or equal to the number of lines',
-
-`in the program (\textless sourceline()\textgreater)'
-\#ErrorText.40.35=` argument 1 cannot be expressed as a whole number;',
-
-`found ``''\,'
-
-\#ErrorText.40.36=` argument 1', `must be the name of a variable in the
-pool;', `found ``''\,'
-
-\#ErrorText.40.37=` argument 3',
-
-`must be the name of a pool; found ``''\,' \#ErrorText.40.38=` argument
-',
-
-`is not large enough to format ``''\,' \#ErrorText.40.39=` argument 3 is
-not zero or one; found ``''\,' \#ErrorText.40.41=` argument ',
-
-`must be within the bounds of the stream;',
-
-`found ``''\,'
-
-\#ErrorText.40.42=` argument 1; cannot position on this stream;', `found
-``''\,' \#ErrorText.40.45=` argument must be a single',
-
-`non-alphanumeric character or the null string;',
-
-' ``found '''
-
-\#ErrorText.40.46=` argument 3, ``'', is a format incompatible',
-
-`with separator specified in argument '
-
-\#ErrorText.41 = `Bad arithmetic conversion' \#ErrorText.41.1=
-`Non-numeric value (``'')', `to left of arithmetic operation ``''\,'
-
-\#ErrorText.41.2= `Non-numeric value (``'')', `to right of arithmetic
-operation ``''\,' \#ErrorText.41.3= `Non-numeric value (``'')',
-
-`used with prefix operator ``''\,'
-
-\#ErrorText.41.4= `Value of TO expression in DO instruction', `must be
-numeric; found ``''\,'!
-
-\#ErrorText.41.5= `Value of BY expression in DO instruction', `must be
-numeric; found ``''\,'!
-
-\#ErrorText.41.6= `Value of control variable expression of DO
-instruction', `must be numeric; found ``''\,'!
-
-\#ErrorText.41.7= `Exponent exceeds' \#Limit ExponentDigits `digits;',
-`found ``''\,'
-
-\#ErrorText.42 = `Arithmetic overflow/underflow' \#ErrorText.42.1l=
-`Arithmetic overflow detected at', `Nevalue\textgreater{} ``;',
-`exponent of result requires more than', \#Limit ExponentDigits `digits'
-\#ErrorText.42.2= `Arithmetic underflow detected at',
-`Nevalue\textgreater{} ``;`, `exponent of result requires more than',
-\#Limit ExponentDigits `digits'\,''Arithmetic overflow; divisor must not
-be zero'
-
-\#ErrorText.42.3
-
-``Routine not found' `Could not find routine ``''\,'
-
-\#ErrorText.43 = \#ErrorText.43.1= \#ErrorText.44 = `Function did not
-return data'
-
-\#ErrorText.44.1= `No data returned from function ``''\,'
-
-\#ErrorText.45 = `No data specified on function RETURN'
-\#ErrorText.45.1= `Data expected on RETURN instruction because',
-`routine ``'' was called as a function'
-
-\#ErrorText.46 = `Invalid variable reference' \#ErrorText.46.1= `Extra
-token (``'') found in variable', `reference; ``)'' expected'
-
-\#ErrorText.47 = `Unexpected label' \#ErrorText.47.1l= `INTERPRET data
-must not contain labels;', 'found ``''!
-
-\#ErrorText.48 = `Failure in system service' \#ErrorText.48.1= `Failure
-in system service: '
-
-\#ErrorText.49 = `Interpretation Error' \#ErrorText.49.1=
-`Interpretation Error: ' \#ErrorText.50 = `Unrecognized reserved symbol'
-
-\#ErrorText.50.1= `Unrecognized reserved symbol ``''\,'
-
-\#ErrorText.51 = `Invalid function name'
-
-\#ErrorText.51.1= `Unquoted function names must not end with a period;',
-`found ``''\,'
-
-\#ErrorText.52
-
-``Result returned by'''' is longer than', \#Limit String `characters'
-
-\#ErrorText.53 = `Invalid option' \#ErrorText.53.1l= `Variable reference
-expected', `after STREAM keyword; found ``''\,' \#ErrorText.53.2=
-`Variable reference expected', `after STEM keyword; found ``''\,'
-\#ErrorText.53.3= `Argument to STEM must have one period,', `as its last
-character; found ``''\,' \#ErrorText.54 = `Invalid STEM value'
-\#ErrorText.54.1= `For this use of STEM, the value of ``'' must be a',
-`count of lines; found: ``''\,'
+\lstinputlisting[language=rexx,label=errortexts2.rexx,caption=errortexts2.rexx]{errortexts2.rexx}
 
@@
 
-call Config NoSource
+\lstinputlisting[language=rexx,label=confignosource.rexx,caption=confignosource.rexx]{confignosource.rexx}
 
-If Config\_NoSource has set \#NoSource to `0' the lines of source
-processed by clause 6 are copied to \#SourceLine. , with \#SourceLine.O
-being a count of the lines and \#SourceLine.n for n=1 to \#SourceLine.0
+If \texttt{Config\_NoSource} has set \texttt{\#NoSource} to
+\texttt{\textquotesingle{}0\textquotesingle{}} the lines of source
+processed by clause 6 are copied to \texttt{\#SourceLine.} , with
+\texttt{\#SourceLine.O} being a count of the lines and
+\texttt{\#SourceLine.n} for \texttt{n}=1 to \texttt{\#SourceLine.0}
 being the source lines in order.
 
-If Config\_NoSource has set \#NoSource to `1' then \#SourceLine.0 is set
-to 0. The following state variables affect tracing:
+If \texttt{Config\_NoSource} has set \texttt{\#NoSource} to
+\texttt{\textquotesingle{}1\textquotesingle{}} then
+\texttt{\#SourceLine.0} is set to \texttt{0}.
 
-\#InhibitPauses = 0
+The following state variables affect tracing:
 
-\#InhibitTrace = 0
+\lstinputlisting[language=rexx,label=configtracesettings.rexx,caption=configtracesettings.rexx]{configtracesettings.rexx}
 
-\#AtPause = 0 /* Off until interactive input being received. */
+An initial variable pool is established:
 
-\#Trace QueryPrior = `No' An initial variable pool is established:
-
-call Config ObjectNew
-
-\#Pool = \#Outcome
-
-\#P0011 = \#Pool
-
-call Var\_Empty \#Pool
-
-call Var\_Reset \#Pool
-
-\#Level = 1 /* Level of invocation */ \#NewLevel = 2
-\#IsFunction.\#Level = (\#HowInvoked == `FUNCTION')
+\lstinputlisting[language=rexx,label=configinitialvarpool.rexx,caption=configinitialvarpool.rexx]{configinitialvarpool.rexx}
 
@@
 
-\#Digits.\#Level = 9 /* Numeric Digits \emph{/ \#Form.\#Level =
-`SCIENTIFIC' /} Numeric Form \emph{/ \#Fuzz.\#Level = 0 /} Numeric Fuzz
-\emph{/ \#StartTime.\#Level = '\,' /} Elapsed time boundary */
-\#LineNumber = '\,'
+\lstinputlisting[language=rexx,label=configdigits.rexx,caption=configdigits.rexx]{configdigits.rexx}
 
-\#Tracing.\#Level = `N'
+An environment is provided by the \texttt{API\_Start} to become the
+initial active environment to which commands will be addressed. The
+alternate environment is made the same:
 
-\#Interactive.\#Level = `0'
-
-69 An environment is provided by the API\_ Start to become the initial
-active environment to which commands will be addressed. The alternate
-environment is made the same:
-
-/* Call the environments ACTIVE, ALTERNATE, TRANSIENT where these are
-never-initialized state variables.
-
-Similarly call the redirections I O and E */
-
-call EnvAssign ALTERNATE, \#Level, ACTIVE, \#Level
+\lstinputlisting[language=rexx,label=configenvironments.rexx,caption=configenvironments.rexx]{configenvironments.rexx}
 
@@
 
-\#Enabling.SYNTAX.\#Level = `OFF' \#Enabling.HALT.\#Level = `OFF'
-\#Enabling.ERROR.\#Level = `OFF' \#Enabling.FAILURE.\#Level = `OFF'
-\#Enabling.NOTREADY.\#Level = `OFF' \#Enabling.NOVALUE.\#Level = `OFF'
-\#Enabling.LOSTDIGITS.\#Level = `OFF'
+\lstinputlisting[language=rexx,label=configdisablesconditions.rexx,caption=configdisablesconditions.rexx]{configdisablesconditions.rexx}
 
-\#PendingNow.HALT.\#Level = 0 \#PendingNow.ERROR.\#Level = 0
-\#PendingNow.FAILURE.\#Level = 0 \#PendingNow.NOTREADY.\#Level = 0 /*
-The following field corresponds to the results from the CONDITION
-built-in function. */ \#Condition.\#Level = '\,' The opportunity is
-provided for a trap to initialize the pool. \#API Enabled = `1' call
-Var\_Reset \#Pool call Config Initialization \#API Enabled = `0' \#\#
-REQUIRES For each requires in order of appearence: A use of Start\_API
-with \#instance(requires, taken\_constant). Msg40.1 or a new if
-completion `E'. Add Provides to an ordered collection. Not cyclic
-because .LIST can be defined without defining REQUIRES but a fairly
-profound forward reference. \#\# CLASS For each class in order of
-appearence: \#ClassName = \#Instance(class, taken constant) call
-var\_value \#ReservedPool, `\#CLASSES.'ClassName, '1' if \#Indicator ==
-`D' then do call Config ObjectNew \#Class = \#Outcome call var\_set
-\#ReservedPool, `\#CLASSES.'ClassName, '1', \#Class end else call
-\#Raise `SYNTAX', nn.nn, \#ClassName
+The opportunity is provided for a trap to initialize the pool.
 
-New instance of CLASS class added to list. Msg ``Duplicate ::CLASS
-directive instruction''(?) \#\# METHOD
+\lstinputlisting[language=rexx,label=configinitializepooltrap.rexx,caption=configinitializepooltrap.rexx]{configinitializepooltrap.rexx}
 
-For each method in order of appearence: call Config ObjectNew \#Po00ol =
-\#Outcome call Config ObjectSource (\#Pool) \#MethodName =
-\#Instance(method, taken constant) call var\_value \#Class,
-`\#METHODS.'\#MethodName, `1' if \#Indicator == `D' then call var set
-\#Class, `\#METHODS.'\#MethodName, `1', \#Pool else call \#Raise
-`SYNTAX', nn.nn, \#MethodName, \#ClassName
+\section{REQUIRES}\label{requires}
 
-GUARDED \& public is default. if \#contains(method, `PRIVATE') then
-m\textasciitilde setprivate; if \#contains(method, 'UNGUARDED)) then
-m\textasciitilde setunguarded
+For each requires in order of appearence:
 
-Why is there a keyword for GUARDED but not for PUBLIC here?
+\emph{A use of Start\_API with \#instance(requires, taken\_constant).
+Msg40.1 or a new if completion `E'. Add Provides to an ordered
+collection. Not cyclic because .LIST can be defined without defining
+REQUIRES but a fairly profound forward reference.}
 
-Does CLASS option mean ENHANCE with Class class methods?
+\section{CLASS}\label{class}
 
-\#CurrentClass \textasciitilde class(\#instance(method,
-taken\_constant), m)
+For each class in order of appearence:
 
+\lstinputlisting[language=rexx,label=classnamesorderofappearance.rexx,caption=classnamesorderofappearance.rexx]{classnamesorderofappearance.rexx}
+
+\emph{New instance of CLASS class added to list. Msg ``Duplicate ::CLASS
+directive instruction''(?)} \#\# METHOD
+
+For each method in order of appearence:
+
+\lstinputlisting[language=rexx,label=methodsorderofappearance.rexx,caption=methodsorderofappearance.rexx]{methodsorderofappearance.rexx}
+
+\emph{GUARDED \& public is default. if \#contains(method, `PRIVATE')
+then m\textasciitilde setprivate; if \#contains(method, 'UNGUARDED))
+then m\textasciitilde setunguarded\\
+Why is there a keyword for GUARDED but not for PUBLIC here?\\
+Does CLASS option mean ENHANCE with Class class methods?\\
+\#CurrentClass\textasciitilde class(\#instance(method, taken\_constant),
+m)\\
 For ATTRIBUTE, should we actually construct source for two methods?
-ATTRIBUTE case needs test of null body. OO! doesn't have source (because
-it actually traps UNKNOWN?).
-
+ATTRIBUTE case needs test of null body. OOI doesn't have source (because
+it actually traps UNKNOWN?).\\
 For EXTERNAL test for null body. Simon Nash doc says ``Accessibility to
@@
 doesn't even tell us about search order. We will need a
-Config\_ExternalClass to import the public names of the class.
+Config\_ExternalClass to import the public names of the class.}
 
-\hypertarget{routine}{%
-\section{ROUTINE}\label{routine}}
+\section{ROUTINE}\label{routine}
 
@@
 
-Add name (with duplicate check) to list for this file.
-
+\emph{Add name (with duplicate check) to list for this file.\\
 Extra step needed in the invocation search order. Although this is
@@
 (Except perhaps when the routine was made available by a REQUIRES; in
-that case the PARSE SOURCE answer has to change.)
-
-have the builtins-defined-by-directives elsewhere; it would make sense
-if they wound up about here.
+that case the PARSE SOURCE answer has to change.)\\
+I have the builtins-defined-by-directives elsewhere; it would make sense
+if they wound up about here.}

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/evaluation.tex
add filenames

@@
 
-This section refers to the DATATYPE built-in function when checking
-operands, see nnn. Except for considerations of limits on the values of
-exponents, the test:
+This section refers to the \texttt{DATATYPE} built-in function when
+checking operands, see nnn. Except for considerations of limits on the
+values of exponents, the test:
 
-datatype (Subject) == `NUM' is equivalent to testing whether the subject
-matches the syntax: num := {[}blank+{]} {[}`+' \textbar{} `-'{]}
-{[}blank+{]} number {[}blank+{]}
+\lstinputlisting[language=rexx,label=datatypenum.rexx,caption=datatypenum.rexx]{datatypenum.rexx}
 
-For the syntax of number see nnn.
+is equivalent to testing whether the subject matches the syntax:
 
-When the matching subject does not include a `-' the value is the value
-of the number in the match, otherwise the value is the value of the
-expression (0 - number).
+\lstinputlisting[language=rexx,label=ebfdatatype.ebnf,caption=ebfdatatype.ebnf]{ebfdatatype.ebnf}
+
+For the syntax of \emph{number} see nnn.
+
+When the matching subject does not include a
+\texttt{\textquotesingle{}-\textquotesingle{}} the value is the value of
+the number in the match, otherwise the value is the value of the
+expression \texttt{(0\ -\ number)}.
 
@@
 
-datatype (Subject , `W')
+\lstinputlisting[language=rexx,label=datatypesubject.rexx,caption=datatypesubject.rexx]{datatypesubject.rexx}
 
-is a test that the Subject matches that syntax and also has a value that
-is ``whole'', that is has no non-zero fractional part.
+is a test that the \texttt{Subject} matches that syntax and also has a
+value that is ``whole'', that is has no non-zero fractional part.
 
-When these two tests are made and the Subject matches the constraints
-but has an exponent that is not
+When these two tests are made and the \texttt{Subject} matches the
+constraints but has an exponent that is not in the correct range of
+values then a condition is raised:
 
-in the correct range of values then a condition is raised: call \#Raise
-`SYNTAX', 41.7, Subject
+\lstinputlisting[language=rexx,label=callraisesyntax.rexx,caption=callraisesyntax.rexx]{callraisesyntax.rexx}
 
-This possibility is implied by the uses of DATATYPE and not shown
-explicitly in the rest of this section nnn.
+This possibility is implied by the uses of \texttt{DATATYPE} and not
+shown explicitly in the rest of this section nnn.
 
@@
 The notation used here is the same as that defined in sections nnn and
-nnn, including the fact that the Var\_ routines may return an indicator
-of `N', `S' or `X'.
+nnn, including the fact that the \texttt{Var\_} routines may return an
+indicator of \texttt{\textquotesingle{}N\textquotesingle{}},
+\texttt{\textquotesingle{}S\textquotesingle{}} or
+\texttt{\textquotesingle{}X\textquotesingle{}}.
 
@@
 non-tailed name; names with different qualification and the same
-spelling are different items in the pool. For those Var\_ functions with
-a third argument this argument indicates the qualification; it is `1'
-when addressing tailed names or `0' when addressing non-tailed names.
+spelling are different items in the pool. For those \texttt{Var\_}
+functions with a third argument this argument indicates the
+qualification; it is \texttt{\textquotesingle{}1\textquotesingle{}} when
+addressing tailed names or
+\texttt{\textquotesingle{}0\textquotesingle{}} when addressing
+non-tailed names.
 
@@
 A variable pool is associated with a reference denoted by the first
-argument, with name Pool. The value of Pool may alter during execution.
-The same name, in conjunction with different values of Pool, can
-correspond to different values.
+argument, with name \texttt{Pool}. The value of \texttt{Pool} may alter
+during execution. The same name, in conjunction with different values of
+\texttt{Pool}, can correspond to different values.
 
@@
 
-Var\_Empty (Pool)
+\lstinputlisting[language=rexx,label=varpoolempty.rexx,caption=varpoolempty.rexx]{varpoolempty.rexx}
 
@@
 
-Var \_Set(Pool, Name, `0', Value)
+\lstinputlisting[language=rexx,label=varsetzero.rexx,caption=varsetzero.rexx]{varsetzero.rexx}
 
@@
 The name is a non-tailed name. If the specified name has the `exposed'
-attribute then Var\_Set operates on the variable pool referenced by
-\#Upper in this pool and this rule is applied to that pool. When the
-pool with attribute `not-exposed' for this name is determined the
-specified value is associated with the specified name. It also
-associates the attributes `not-dropped' and `not-implicit'. If that
+attribute then \texttt{Var\_Set} operates on the variable pool
+referenced by \texttt{\#Upper} in this pool and this rule is applied to
+that pool. When the pool with attribute `not-exposed' for this name is
+determined the specified value is associated with the specified name. It
+also associates the attributes `not-dropped' and `not-implicit'. If that
 attribute was previously `not-dropped' then the indicator returned is
-`R'. The name is a stem if it contains just one period, as its rightmost
-character. When the name is a stem Var\_Set(Pool, TailedName, `1',Value)
-is executed for all possible valid tailed names which
+\texttt{\textquotesingle{}R\textquotesingle{}}. The name is a stem if it
+contains just one period, as its rightmost character. When the name is a
+stem
+\texttt{Var\_Set(Pool,\ TailedName,\ \textquotesingle{}1\textquotesingle{},Value)}
+is executed for all possible valid tailed names which have Name as their
+stem, and then those tailed-names are given the attribute `implicit'.
 
-have Name as their stem, and then those tailed-names are given the
-attribute `implicit'. Var \_Set(Pool, Name, `1', Value)
+\lstinputlisting[language=rexx,label=varsetone.rexx,caption=varsetone.rexx]{varsetone.rexx}
 
@@
 stem is determined the name is considered in that pool. If the name has
-the `exposed' attribute then the
-
-variable pool referenced by \#Upper in the pool is considered and this
-rule applied to that pool. When the pool with attribute `not-exposed' is
-determined the specified value is associated with the specified name. It
-also associates the attributes `not-dropped' and `not-implicit' . If
-that attribute was previously `not-dropped' then the indicator returned
-is `R'.
+the `exposed' attribute then the variable pool referenced by \#Upper in
+the pool is considered and this rule applied to that pool. When the pool
+with attribute `not-exposed' is determined the specified value is
+associated with the specified name. It also associates the attributes
+`not-dropped' and `not-implicit' . If that attribute was previously
+`not-dropped' then the indicator returned is
+\texttt{\textquotesingle{}R\textquotesingle{}}.
 
@@
 
-Var \_Value(Pool, Name, `0')
+\lstinputlisting[language=rexx,label=varvaluezero.rexx,caption=varvaluezero.rexx]{varvaluezero.rexx}
 
@@
 attribute then Var\_Value operates on the variable pool referenced by
-\#Upper in this pool and this rule is applied to that pool. When the
-pool with attribute `not-exposed' for this name is determined the
-indicator returned is `D' if the name has `dropped' associated, `N'
-otherwise. In the former case \#Outcome is set equal to Name, in the
-latter case \#Outcome is set to the value most
+\texttt{\#Upper} in this pool and this rule is applied to that pool.
+When the pool with attribute `not-exposed' for this name is determined
+the indicator returned is `D' if the name has `dropped' associated,
+\texttt{\textquotesingle{}N\textquotesingle{}} otherwise.
 
-recently associated with the name by Var\_Set. Var \_Value(Pool, Name,
-`1')
+In the former case \texttt{\#Outcome} is set equal to \texttt{Name}, in
+the latter case \texttt{\#Outcome} is set to the value most recently
+associated with the name by \texttt{Var\_Set}.
+
+\lstinputlisting[language=rexx,label=varvalueone.rexx,caption=varvalueone.rexx]{varvalueone.rexx}
 
@@
 including the first period, is the stem. The stem is a non-tailed name.
-If the specified stem has the `exposed' attribute then Var\_Value
-operates on the variable pool referenced by \#Upper in this pool and
-this rule is applied to that pool. When the pool with attribute
-`not-exposed' for the stem is determined the name is considered in that
-pool. If the name has the `exposed' attribute then the variable pool
-referenced by \#Upper in the pool is considered and this rule applied to
-that pool. When the pool with attribute `not-exposed' is determined the
-indicator returned is `D' if the name has `dropped' associated, `N'
-otherwise. In the former case \#Outcome is set equal to Name, in the
-latter case \#Outcome is set to the value most recently associated with
-the name by Var\_Set.
+If the specified stem has the `exposed' attribute then
+\texttt{Var\_Value} operates on the variable pool referenced by
+\texttt{\#Upper} in this pool and this rule is applied to that pool.
+When the pool with attribute `not-exposed' for the stem is determined
+the name is considered in that pool. If the name has the `exposed'
+attribute then the variable pool referenced by \#Upper in the pool is
+considered and this rule applied to that pool. When the pool with
+attribute `not-exposed' is determined the indicator returned is
+\texttt{\textquotesingle{}D\textquotesingle{}} if the name has `dropped'
+associated, \texttt{\textquotesingle{}N\textquotesingle{}} otherwise. In
+the former case \texttt{\#Outcome} is set equal to \texttt{Name}, in the
+latter case \texttt{\#Outcome} is set to the value most recently
+associated with the name by \texttt{Var\_Set}.
 
@@
 
-Var \_Drop(Pool, Name, `0')
+\lstinputlisting[language=rexx,label=vardropzero.rexx,caption=vardropzero.rexx]{vardropzero.rexx}
 
@@
 The name is a non-tailed name. If the specified name has the `exposed'
-attribute then Var\_Drop operates on the variable pool referenced by
-\#Upper in this pool and this rule is applied to that pool. When the
-pool with attribute `not-exposed' for this name is determined the
-attribute `dropped' is associated with the specified name. Also, when
-the name is a stem, Var\_Drop(Pool,TailedName,`1') is executed for all
-possible valid tailed names which have Name as astem.
+attribute then \texttt{Var\_Drop} operates on the variable pool
+referenced by \texttt{\#Upper} in this pool and this rule is applied to
+that pool. When the pool with attribute `not-exposed' for this name is
+determined the attribute `dropped' is associated with the specified
+name. Also, when the name is a stem,
+\texttt{Var\_Drop(Pool,TailedName,\textquotesingle{}1\textquotesingle{})}
+is executed for all possible valid tailed names which have \texttt{Name}
+as a stem.
 
-Var \_Drop(Pool, Name, `1')
+\lstinputlisting[language=rexx,label=vardropone.rexx,caption=vardropone.rexx]{vardropone.rexx}
 
@@
 pool. If the name has the `exposed' attribute then the variable pool
-referenced by \#Upper in the pool is considered and this rule applied to
-that pool. When the pool with attribute `not-exposed' is determined the
-attribute `dropped' is associated with the specified name.
+referenced by \texttt{\#Upper} in the pool is considered and this rule
+applied to that pool. When the pool with attribute `not-exposed' is
+determined the attribute `dropped' is associated with the specified
+name.
 
@@
 
-Var\_Expose (Pool, Name, `0')
+\lstinputlisting[language=rexx,label=varexposezero.rexx,caption=varexposezero.rexx]{varexposezero.rexx}
 
@@
 with the specified name. Also, when the name is a stem,
-Var\_Expose(Pool, TailedName,`1') is executed for all possible valid
-tailed names which have Name as a stem.
+\texttt{Var\_Expose(Pool,\ TailedName,\textquotesingle{}1\textquotesingle{})}
+is executed for all possible valid tailed names which have Name as a
+stem.
 
-Var\_Expose (Pool, Name, `1')
+\lstinputlisting[language=rexx,label=varexpose1.rexx,caption=varexpose1.rexx]{varexpose1.rexx}
 
@@
 
-Var\_ Reset (Pool)
+\lstinputlisting[language=rexx,label=varresetpool.rexx,caption=varresetpool.rexx]{varresetpool.rexx}
 
 The function operates on the variable pool with the specified reference.
-It establishes the effect of subsequent API\_Next and API\_NextVariable
-functions (see sections nnn and nnn). A Var\_Reset is implied by any
-API\_ operation other than API\_Next and API\_NextVariable.
+It establishes the effect of subsequent \texttt{API\_Next} and
+\texttt{API\_NextVariable} functions (see sections nnn and nnn). A
+\texttt{Var\_Reset} is implied by any \texttt{API\_} operation other
+than \texttt{API\_Next} and \texttt{API\_NextVariable}.
 
@@
 
-The value of a symbol which is a NUMBER or a CONST\_SYMBOL which is not
-a reserved symbol is the content of the appropriate token.
+The value of a symbol which is a \emph{NUMBER} or a \emph{CONST\_SYMBOL}
+which is not a reserved symbol is the content of the appropriate token.
 
-The value of a VAR\_SYMBOL which is ``taken as a constant'' is the
-VAR\_SYMBOL itself, otherwise the VAR\_SYMBOL identifies a variable and
-its value may vary during execution.
+The value of a \emph{VAR\_SYMBOL} which is ``taken as a constant'' is
+the \emph{VAR\_SYMBOL} itself, otherwise the \_VAR\_SYMBOL: identifies a
+variable and its value may vary during execution.
 
 Accessing the value of a symbol which is not ``taken as a constant''
-shall result in trace output, see nnn: if \#Tracing.\#Level == `I' then
-call \#Trace Tag
+shall result in trace output, see nnn:
 
-where Tag is `\textgreater L\textgreater{}' unless the symbol is a
-VAR\_SYMBOL which, when used as an argument to Var\_Value, does not
-yield an indicator `D'. In that case, the Tag is
-'\textgreater V\textgreater``.
+\lstinputlisting[language=rexx,label=tracinglevel.rexx,caption=tracinglevel.rexx]{tracinglevel.rexx}
+
+where \texttt{Tag} is
+\texttt{\textquotesingle{}\textgreater{}L\textgreater{}\textquotesingle{}}
+unless the symbol is a \emph{VAR\_SYMBOL} which, when used as an
+argument to \texttt{Var\_Value}, does not yield an indicator
+\texttt{\textquotesingle{}D\textquotesingle{}}. In that case, the Tag is
+\texttt{\textquotesingle{}\textgreater{}V\textgreater{}\textquotesingle{}}.
 
@@
 
-If VAR\_SYMBOL does not contain a period, or contains only one period as
-its last character, the value of
+If \emph{VAR\_SYMBOL} does not contain a period, or contains only one
+period as its last character, the value of the variable is the value
+associated with \emph{VAR\_SYMBOL} in the variable pool, that is
+\texttt{\#Outcome} after
+\texttt{Var\_Value(Pool,\ VAR\_SYMBOL,\ \textquotesingle{}0\textquotesingle{})}.
 
-the variable is the value associated with VAR\_SYMBOL in the variable
-pool, that is \#Outcome after Var\_ Value (Pool, VAR SYMBOL, `0')
+If the indicator is \texttt{\textquotesingle{}D\textquotesingle{}},
+indicating the variable has the `dropped' attribute, the
+\texttt{NOVALUE} condition is raised; see nnn and nnn for exceptions to
+this.
 
-If the indicator is `D', indicating the variable has the `dropped'
-attribute, the NOVALUE condition is raised; see nnn and nnn for
-exceptions to this. \#Response = Var Value(Pool, VAR SYMBOL, `0') if
-left(\#Response,1) == `D' then call \#Raise `NOVALUE', VAR\_SYMBOL, '\,'
-If VAR\_SYMBOL contains a period which is not its last character, the
-value of the variable is the value associated with the derived name.
-7.3.1 Derived names A derived name is derived from a VAR\_SYMBOL as
-follows:
+\lstinputlisting[language=rexx,label=varpoolsymbol.rexx,caption=varpoolsymbol.rexx]{varpoolsymbol.rexx}
 
-VAR SYMBOL := Stem Tail
+If \emph{VAR\_SYMBOL} contains a period which is not its last character,
+the value of the variable is the value associated with the derived name.
 
-Stem := PlainSymbol `.'!
+\subsection{Derived names}\label{derived-names}
 
-Tail i= (PlainSymbol \textbar{} `.' {[}PlainSymbol{]}) {[}`.'
-{[}PlainSymbol1{]}{]}+ PlainSymbol := (general letter \textbar{} digit)+
+A derived name is derived from a \emph{VAR\_SYMBOL} as follows:
 
-The derived name is the concatenation of: - the Stem, without further
-evaluation; - the Tail, with the PlainSymbols replaced by the values of
-the symbols. The value of a PlainSymbol which does not start with a
-digit is \#Outcome after Var\_ Value (Pool, PlainSymbol,`0') These
-values are obtained without raising the NOVALUE condition.
+\lstinputlisting[language=rexx,label=derivednames.rexx,caption=derivednames.rexx]{derivednames.rexx}
+
+The derived name is the concatenation of:
+
+\begin{itemize}
+\tightlist
+\item
+  the \emph{Stem}, without further evaluation;
+\item
+  the \emph{Tail}, with the \emph{PlainSymbols} replaced by the values
+  of the symbols. The value of a \emph{PlainSymbol} which does not start
+  with a digit is \texttt{\#Outcome} after
+\end{itemize}
+
+\lstinputlisting[language=rexx,label=varvalueplainsymbol.rexx,caption=varvalueplainsymbol.rexx]{varvalueplainsymbol.rexx}
+
+These values are obtained without raising the \texttt{NOVALUE}
+condition.
 
@@
 then the variable \#Test is set as follows:
+
 \lstinputlisting[language=rexx,label=evaluation-comparison.rexx,caption=evaluation-comparison.rexx]{evaluation-comparison.rexx}
@@
 determines the value of the comparison. The value of the comparison is
-`1' if - \#Test is `E' and the operator is one of `=``\,``,'==````,
+`1' if - \#Test is `E' and the operator is one of `=``\,``,'==``\,``,
 `\textgreater=', \textless='', `\textgreater{}', `\textless{}',
@@
   `\textgreater=``, `\textless{}', `=', `\textless\textgreater{}',
-  `\textgreater\textless{}', Nes'', `\textgreater\textgreater!
+  `\textgreater\textless{}', Nes'','\textgreater\textgreater!
   `p\textgreater{}', or \textless\textless``)
@@
 
-The code of ArithOp itself is assumed to operate under a sufficiently
-high setting of numeric digits to avoid exponential notation.
+The code of \texttt{ArithOpp} itself is assumed to operate under a
+sufficiently high setting of numeric digits to avoid exponential
+notation.
+
 \lstinputlisting[language=rexx,label=evaluation-arithmetic.rexx,caption=evaluation-arithmetic.rexx]{evaluation-arithmetic.rexx}
@@
 
-Invocation occurs when a function or a message\_term or a callis
-evaluated. Invocation of a function may result in a value, in which
+Invocation occurs when a \emph{function} or a \emph{message\_term} or a
+\emph{call} is evaluated. Invocation of a function may result in a
+value, in which case:
+
+\lstinputlisting[language=rexx,label=tracinglevelF.rexx,caption=tracinglevelF.rexx]{tracinglevelF.rexx}
+
+Invocation of a \emph{message\_term} may result in a value, in which
 case:
 
-if \#Tracing.\#Level == `I' then call \#Trace
-`\textgreater F\textgreater{}'
-
-Invocation of a message\_term may result in a value, in which case:
-
-if \#Tracing.\#Level == `I' then call \#Trace
-`\textgreater M\textgreater{}'
+\lstinputlisting[language=rexx,label=tracinglevelM.rexx,caption=tracinglevelM.rexx]{tracinglevelM.rexx}
 
@@
 
-The argument positions are the positions in the exoression\_list where
-syntactically an expression occurs or could have occurred. Let ArgNumber
-be the number of an argument position, counting from 1 at the left; the
-range of ArgNumber is all whole numbers greater than zero.
+The argument positions are the positions in the \emph{expression\_list}
+where syntactically an \emph{expression} occurs or could have occurred.
+Let \texttt{ArgNumber} be the number of an argument position, counting
+from \texttt{1} at the left; the range of \texttt{ArgNumber} is all
+whole numbers greater than zero.
 
-For each value of ArgNumber, \#ArgExists.\#NewLevel.ArgNumber is set `1'
-if there is an expression present, `O' if not.
+For each value of \texttt{ArgNumber},
+\texttt{\#ArgExists.\#NewLevel.ArgNumber} is set
+\texttt{\textquotesingle{}1\textquotesingle{}} if there is an expression
+present, \texttt{\textquotesingle{}O\textquotesingle{}} if not.
 
-From the left, if \#ArgExists.\#NewLevel.ArgNumber is `1' then
-\#Arg.\#NewLevel.ArgNumber is set to the value of the corresponding
-expression. If \#ArgExists.\#NewLevel.ArgNumber is `0' then
-\#Arg.\#NewLevel.ArgNumber is set to the null string.
+From the left, if \texttt{\#ArgExists.\#NewLevel.ArgNumber} is
+\texttt{\textquotesingle{}1\textquotesingle{}} then
+\texttt{\#Arg.\#NewLevel.ArgNumber} is set to the value of the
+corresponding expression. If \texttt{\#ArgExists.\#NewLevel.ArgNumber}
+is \texttt{\textquotesingle{}0\textquotesingle{}} then
+\texttt{\#Arg.\#NewLevel.ArgNumber} is set to the null string.
 
-\#ArgExists.\#NewLevel.0 is set to the largest ArgNumber for which
-\#ArgExists.\#NewLevel.ArgNumber is `1', or to zero if there is no such
-value of ArgNumber.
+\texttt{\#ArgExists.\#NewLevel.0} is set to the largest
+\texttt{ArgNumber} for which \texttt{\#ArgExists.\#NewLevel.ArgNumber}
+is \texttt{\textquotesingle{}1\textquotesingle{}}, or to zero if there
+is no such value of \texttt{ArgNumber}.
 
@@
 
-The value of a LABEL, or of the taken\_constant in the function or
-call\_instruction, is taken as a constant, see nnn. If the
-taken\_constant is not a string\_literal it is a reference to the first
-LABEL in the program which has the same value. The comparison is made
-with the `==' operator.
+The value of a \emph{LABEL}, or of the \emph{taken\_constant} in the
+function or \emph{call\_instruction}, is taken as a constant, see nnn.
+If the \emph{taken\_constant} is not a \emph{string\_literal} it is a
+reference to the first \emph{LABEL} in the program which has the same
+value. The comparison is made with the `==' operator.
 
 If there is such a matching label and the label is trace-only (see nnn)
-then a condition is raised: call \#Raise `SYNTAX', 16.3, taken constant
+then a condition is raised:
+
+\lstinputlisting[language=rexx,label=callraisesyntax.rexx,caption=callraisesyntax.rexx]{callraisesyntax.rexx}
 
@@
 
-If there is no such matching label, or if the taken\_constant is a
-string\_literal, further comparisons are made.
+If there is no such matching label, or if the \emph{taken\_constant} is
+a \emph{string\_literal}, further comparisons are made.
 
-If the value of the taken\_constant matches the name of some built-in
-function then that built-in function is invoked. The names of the
-built-in functions are defined in section nnn and are in uppercase.
+If the value of the \emph{taken\_constant} matches the name of some
+built-in function then that built-in function is invoked. The names of
+the built-in functions are defined in section nnn and are in uppercase.
 
 If the value does not match any built-in function name,
-Config\_ExternalRoutine is used to invoke an external routine.
+\texttt{Config\_ExternalRoutine} is used to invoke an external routine.
 
-Whenever a matching label is found, the variables SIGL and .SIGL are
-assigned the value of the line number of the clause which caused the
-search for the label. In the case of an invocation resulting from a
+Whenever a matching label is found, the variables \texttt{SIGL} and
+\texttt{.SIGL} are assigned the value of the line number of the clause
+which caused the search for the label. In the case of an invocation
+resulting from a condition occurring that shall be the clause in which
+the condition occurred.
 
-condition occurring that shall be the clause in which the condition
-occurred. Var \_ Set(\#Pool, `SIGL', `0', \#LineNumber) var Set(0 ,
-`.SIGL', `0', \#LineNumber)
+\lstinputlisting[language=rexx,label=varset.rexx,caption=varset.rexx]{varset.rexx}
 
-The name used in the invocation is held in \#Name.\#Level for possible
-use in an error message from the RETURN clause, see nnn
+The name used in the invocation is held in \texttt{\#Name.\#Level} for
+possible use in an error message from the \texttt{RETURN} clause, see
+nnn
 
@@
 
-An internal routine completes when \#Level returns to the value it had
-when the routine was invoked. The value of the internal function is the
-value of the expression on the return which completed the routine. The
-value of an external function is determined by Config\_ExternalRoutine.
+An internal routine completes when \texttt{\#Level} returns to the value
+it had when the routine was invoked. The value of the internal function
+is the value of the \emph{expression} on the \emph{return} which
+completed the routine. The value of an external function is determined
+by \texttt{Config\_ExternalRoutine}.
 
@@
 
-An internal method completes when \#Level returns to the value it had
-when the routine was invoked. The value of the internal method is the
-value of the expression on the return which completed the method. The
-value of an external method is determined by Config\_ExternalMethod.
+An internal method completes when \texttt{\#Level} returns to the value
+it had when the routine was invoked. The value of the internal method is
+the value of the \emph{expression} on the \emph{return} which completed
+the method. The value of an external method is determined by
+\texttt{Config\_ExternalMethod}.
 
@@
 
-See nnn for the syntax of a message\_term. The value of the ferm within
-a message\_term is called the receiver.
+See nnn for the syntax of a \emph{message\_term}. The value of the
+\emph{term} within a \emph{message\_term} is called the receiver.
 
 The receiver and any arguments of the term are evaluated, in left to
-right order. r= \#evaluate(message term, term) If the message term
-contains `\textasciitilde\textasciitilde{}' the value of the message
-term is the receiver. Any effect on .Result? Otherwise the value of a
-message\_term is the value of the method it invokes. The method invoked
-is determined by the receiver and the taken\_constant and symbol. t =
-\#Instance(message term, taken constant) If there is a symbol, it is
-subject to a constraints. if \#contains (message term, symbol) then do
-if r \textless\textgreater{} \#Self then
+\lstinputlisting[language=rexx,label=evaluatemessageterm.rexx,caption=evaluatemessageterm.rexx]{evaluatemessageterm.rexx}
 
-call \#Raise `SYNTAX', nn.n
+If the message term contains
+\texttt{\textquotesingle{}\textasciitilde{}\textasciitilde{}\textquotesingle{}}
+the value of the message term is the receiver.
 
-/* OOI: ``Message search overrides can only be used from methods of the
-target object.'' */
+\emph{Any effect on .Result?}
+
+Otherwise the value of a \emph{message\_term} is the value of the method
+it invokes. The method invoked is determined by the receiver and the
+\emph{taken\_constant} and \emph{symbol}.
+
+\lstinputlisting[language=rexx,label=evaluatemessagetermconstant.rexx,caption=evaluatemessagetermconstant.rexx]{evaluatemessagetermconstant.rexx}
+
+If there is a \emph{symbol}, it is subject to a constraints.
+
+\lstinputlisting[language=rexx,label=evaluatemessagtermsymbol.rexx,caption=evaluatemessagtermsymbol.rexx]{evaluatemessagtermsymbol.rexx}
 
 The search will progress from the object to its class and superclasses.
-/* This is going to be circular because it describes the message lookup
-algorithm and also uses messages. However for the messages in this code
-the message names are chosen to be unique to a method so there is no
-need to use this algorithm in deciding which method is intended. */
 
-/* message term ::= receiver `\textasciitilde{}' taken constant `:'
-VAR\_SYMBOL arguments */
-
-/* This code reflects OOI - the arguments on the message don't affect
-the method choice. */
-
-/* This code selects a method based on its arguments, receiver,
-taken\_constant, and symbol. */
-
-/* This code is used in a context where \#Self is the receiver of the
-method invocation which the subject message term is running under. */
-\lstinputlisting[language=rexx,label=evaluation-selectmethod,caption=evaluation-selectmethod]{evaluation-selectmethod}
+\lstinputlisting[language=rexx,label=messagemethodsearch.rexx-\textgreater,caption=messagemethodsearch.rexx-\textgreater]{messagemethodsearch.rexx-\textgreater}
 
@@
 
-The values of the arguments to the use of Config\_ExternalRoutine, in
-order, are:
+The values of the arguments to the use of
+\texttt{Config\_ExternalRoutine}, in order, are:
 
-The argument How is `SUBROUTINE' if the invocation is from a call,
-```FUNCTION' if the invocation is from a function.
+The argument \texttt{How} is
+\texttt{\textquotesingle{}SUBROUTINE\textquotesingle{}} if the
+invocation is from a \emph{call},
+\texttt{\textquotesingle{}FUNCTION\textquotesingle{}} if the invocation
+is from a \emph{function}.
 
-The argument NameType is `1' if the taken\_constant is a
-string\_literal, `0' otherwise.
+The argument \texttt{NameType} is
+\texttt{\textquotesingle{}1\textquotesingle{}} if the
+\emph{taken\_constant} is a \emph{string\_literal},
+\texttt{\textquotesingle{}0\textquotesingle{}} otherwise.
 
-The argument Name is the value of the faken\_constant.
+The argument \texttt{Name} is the value of the \emph{taken\_constant}.
 
-The argument Environment is the value of this argument on the API\_
-Start which started this execution. The argument Arguments is the \#Arg.
-and \#ArgExists. data.
+The argument \texttt{Environment} is the value of this argument on the
+\texttt{API\_Start} which started this execution.
 
-The argument Streams is the value of this argument on the API\_Start
-which started this execution.
+The argument \texttt{Arguments} is the \texttt{\#Arg.} and
+\texttt{\#ArgExists.} data.
 
-The argument Traps is the value of this argument on the API\_Start which
-started this execution. Var\_Reset is invoked and \#API\_Enabled set to
-`1' before use of Config\_ExternalRoutine. \#API\_Enabled is set to `O'
-after.
+The argument \texttt{Streams} is the value of this argument on the
+\texttt{API\_Start} which started this execution.
 
-The response from Config\_ExternalRoutine is processed. If no conditions
-are (implicitly) raised, \#Outcome is the value of the function.
+The argument \texttt{Traps} is the value of this argument on the
+\texttt{API\_Start} which started this execution.
+
+\texttt{Var\_Reset} is invoked and \texttt{\#API\_Enabled} set to
+\texttt{\textquotesingle{}1\textquotesingle{}} before use of
+\texttt{Config\_ExternalRoutine}. \texttt{\#API\_Enabled} is set to
+\texttt{\textquotesingle{}O\textquotesingle{}} after.
+
+The response from \texttt{Config\_ExternalRoutine} is processed. If no
+conditions are (implicitly) raised, \texttt{\#Outcome} is the value of
+the function.

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/foreword.tex
work in standard book production process including readme.md
upd
add filenames

@@
-\hypertarget{foreword}{%
-\chapter{Foreword}\label{foreword}}
+%preprocessed texin
+\chapter{Foreword}\label{foreword}
 
-\hypertarget{purpose}{%
-\section{Purpose}\label{purpose}}
+\section{Purpose}\label{purpose}
 
@@
 language Rexx. Its purpose is to facilitate portability of Rexx programs
-for use on a wide variety of computer systems. History The computer
-programming language Rexx was designed by Mike Cowlishaw to satisfy the
-following principal aims:
+for use on a wide variety of computer systems.
+
+\section{History}\label{history}
+
+The computer programming language Rexx was designed by Mike Cowlishaw to
+satisfy the following principal aims:
 
@@
 to work with objects created in various ways, led to Rexx extensions and
-to a second X3J18 project which produced this standard. (Ed - hopefully)
+to a second X3J18 project which produced this standard. \emph{(Ed -
+hopefully)}
 
-\hypertarget{committee-lists}{%
-\section{Committee lists}\label{committee-lists}}
-
-(Here)
+Committee lists \emph{(Here)}
 
 This standard was prepared by the Technical Development Committee for
-Rexx, X3J18. There are annexes in this standard; they are informative
-and are not considered part of this standard.
+Rexx, X3J18.
+
+There are annexes in this standard; they are informative and are not
+considered part of this standard.
 
 Suggestions for improvement of this standard will be welcome. They
-should be sent to the
-
-Information Technology Industry Council, 1250 Eye Street, NW, Washington
-DC 20005-3922.
+should be sent to the Information Technology Industry Council, 1250 Eye
+Street, NW, Washington DC 20005-3922.
 
@@
 
-To be inserted The people who contributed to Technical Committee J18 on
-Rexx, which developed this standard, include:
+\emph{To be inserted}
+
+The people who contributed to Technical Committee J18 on Rexx, which
+developed this standard, include:

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/instructions.tex
upd
add filenames

@@
 
-If we left Routine initialization to here we can leave method
-initialization. \#\# Method initialization
+\emph{If we left Routine initialization to here we can leave method
+initialization.}
+
+\section{Method initialization}\label{method-initialization}
 
@@
 
-call Config ObjectNew
-
-\#Po00ol = \#Outcome
-
-Set self and super
+\lstinputlisting[language=rexx,label=localvariablepoolinit.rexx,caption=localvariablepoolinit.rexx]{localvariablepoolinit.rexx}
+
+\emph{Set self and super}
 
@@
 
-If the routine is invoked as a function, \#lsFunction.\#NewLevel shall
-be set to `1', otherwise to `0O'; this affects the processing of a
-subsequent RETURN instruction.
-
-\#AllowProcedure.\#NewLevel = `1'
+If the routine is invoked as a function,
+\texttt{\#IsFunction.\#NewLevel} shall be set to
+\texttt{\textquotesingle{}1\textquotesingle{}}, otherwise to
+\texttt{\textquotesingle{}0\textquotesingle{}}; this affects the
+processing of a subsequent RETURN instruction.
+
+\lstinputlisting[language=rexx,label=routineinit.rexx,caption=routineinit.rexx]{routineinit.rexx}
 
 Many of the initial values for a new invocation are inherited from the
-caller's values. \#Digits.\#NewLevel = \#Digits.\#Level
-
-\#Form.\#NewLevel = \#Form.\#Level
-
-\#Fuzz.\#NewLevel = \#Fuzz.\#Level
-
-\#StartTime.\#NewLevel = \#StartTime.\#Level
-
-\#Tracing.\#NewLevel = \#Tracing.\#Level \#Interactive.\#NewLevel =
-\#Interactive.\#Level
-
-call EnvAssign ACTIVE, \#NewLevel, ACTIVE, \#Level call EnvAssign
-ALTERNATE, \#NewLevel, ALTERNATE, \#Level
-
-do t=1 to 7 Condition = word(`SYNTAX HALT ERROR FAILURE NOTREADY NOVALUE
-LOSTDIGITS',t) \#Enabling.Condition.\#NewLevel =
-\#Enabling.Condition.\#Level
-
-\#Instruction.Condition.\#NewLevel = \#Instruction.Condition.\#Level
-\#TrapName.Condition.\#NewLevel = \#TrapName.Condition.\#Level
-\#EventLevel.Condition.\#NewLevel = \#EventLevel.Condition.\#Level end t
+caller's values.
+
+\lstinputlisting[language=rexx,label=callersvalues.rexx,caption=callersvalues.rexx]{callersvalues.rexx}
 
 If this invocation is not caused by a condition occurring, see nnn, the
-state variables for the CONDITION
-
-built-in function are copied. \#Condition.\#NewLevel =
-\#Condition.\#Level \#ConditionDescription.\#NewLevel =
-\#ConditionDescription.\#Level \#ConditionExtra.\#NewLevel =
-\#ConditionExtra.\#Level \#ConditionInstruction.\#NewLevel =
-\#ConditionInstruction.\#Level Execution of the initialized routine
-continues at the new level of invocation. \#Level = \#NewLevel
-\#NewLevel = \#Level + 1
+state variables for the CONDITION built-in function are copied.
+
+\lstinputlisting[language=rexx,label=conditionlevel.rexx,caption=conditionlevel.rexx]{conditionlevel.rexx}
+
+Execution of the initialized routine continues at the new level of
+invocation.
+
+\lstinputlisting[language=rexx,label=newlevel.rexx,caption=newlevel.rexx]{newlevel.rexx}
 
@@
 
-The clause is traced before execution: if pos(\#Tracing.\#Level, `AIR')
-\textgreater{} 0 then call \#TraceSource
+The clause is traced before execution:
+
+\lstinputlisting[language=rexx,label=traceclauseinit.rexx,caption=traceclauseinit.rexx]{traceclauseinit.rexx}
 
@@
 
-\#ClauseTime.\#Level = '\,' The state variable \#LineNumber is set to
-the line number of the clause, see nnn. A clause other than a null
-clause or label or procedure instruction sets:
-
-\#AllowProcedure.\#Level = `0' /* See message 17.1 */
+\lstinputlisting[language=rexx,label=clauseretaindatetime.rexx,caption=clauseretaindatetime.rexx]{clauseretaindatetime.rexx}
+
+The state variable \texttt{\#LineNumber} is set to the line number of
+the clause, see nnn.
+
+A clause other than a null clause or label or procedure instruction
+sets:
+
+\lstinputlisting[language=rexx,label=allowprocedure.rexx,caption=allowprocedure.rexx]{allowprocedure.rexx}
 
@@
 
-if \#InhibitTrace \textgreater{} 0 then \#InhibitTrace = \#InhibitTrace
-- 1 Polling for a HALT condition occurs:
-
-\#Response = Config Halt Query ()
-
-if \#0utcome == `Yes' then do call Config Halt Reset
-
-call \#Raise `HALT', substr(\#Response,2) /* May return */ end
+\lstinputlisting[language=rexx,label=inihibittrace.rexx,caption=inihibittrace.rexx]{inihibittrace.rexx}
+
+Polling for a HALT condition occurs:
+
+\lstinputlisting[language=rexx,label=haltquery.rexx,caption=haltquery.rexx]{haltquery.rexx}
 
 At the end of each clause there is a check for conditions which occurred
-and were delayed. It is acted on
-
-if this is the clause in which the condition arose. do t=1 to 4
-\#Condition=WORD(`HALT FAILURE ERROR NOTREADY',t) /* HALT can be
-established during HALT handling. */ do while
-\#PendingNow.\#Condition.\#Level \#PendingNow.\#Condition.\#Level = `0'
-call \#Raise end end
-
-Interactive tracing may be turned on via the configuration. Only a
-change in the setting is significant. call Config Trace Query
-
-if \#AtPause = 0 \& \#Outcome == `Yes' \& \#Trace QueryPrior == `No'
-then do /* External request for Trace `?R!' */ \#Interactive.\#Level =
-`1' \#Tracing.\#Level = `R' end
-
-\#TraceQueryPrior = \#Outcome
-
-Tracing just not the same with NetRexx.
+and were delayed. It is acted on if this is the clause in which the
+condition arose.
+
+\lstinputlisting[language=rexx,label=endclauseconditions.rexx,caption=endclauseconditions.rexx]{endclauseconditions.rexx}
+be turned on via the configuration. Only a change in the setting is
+significant.
+
+\lstinputlisting[language=rexx,label=configtracequery.rexx,caption=configtracequery.rexx]{configtracequery.rexx}
+
+\emph{Tracing just not the same with NetRexx.}
 
@@
 
-At a pause point: if \#AtPause = 0 \& \#Interactive.\#Level \&
-\#InhibitTrace = 0 then do if \#InhibitPauses \textgreater{} 0 then
-\#InhibitPauses = \#InhibitPauses-1 else do \#TraceInstruction = `0' do
-forever
-
-call Config Trace Query
-
-if \#Outcome == `No' \& \#Trace QueryPrior == `Yes' then do /* External
-request to stop tracing. */ \#Trace\_QueryPrior=\#Outcome
-\#Interactive.\#Level = `0' \#Tracing.\#Level = `N' leave end
-
-if \#Outcome == `Yes' \& \#Trace QueryPrior == `No' then do /* External
-request for Trace `?R!' */ \#Trace QueryPrior = \#Outcome
-\#Interactive.\#Level = `1' \#Tracing.\#Level = `R' leave end
-
-if \#Interactive.\#Level \textbar{} \#TraceInstruction then leave
-
-/* Accept input for immediate execution. */
-
-call Config Trace Input
-
-if length(\#0utcome) = 0 \textbar{} \#0Outcome == `=' then leave
-\#AtPause = \#Level
-
-interpret \#Outcome
-
-\#AtPause = 0 end /* forever loop \emph{/ if \#Outcome == `=' then call
-\#Retry /} With no return */ end end
+At a pause point:
+
+\lstinputlisting[language=rexx,label=pausepoint.rexx,caption=pausepoint.rexx]{pausepoint.rexx}
 
@@
 the commands input stream) whole number tails are used to order input
-for presentation to the submitted command. Stem.0 must contain a whole
-number indicating the number of compound variables to be presented, and
-stem. 1 through stem.n (where n=stem.0) are the compound variables to be
-presented to the submitted command.
+for presentation to the submitted command. \texttt{Stem.0} must contain
+a whole number indicating the number of compound variables to be
+presented, and \texttt{stem.\ 1} through \texttt{stem.n} (where
+\texttt{n=stem.0}) are the compound variables to be presented to the
+submitted command.
 
@@
 does not perform I/O in a manner compatible with the request, the value
-of \#Env\_Type. may be set to `UNUSED' as an alternative to `STEM' and
-`STREAM' where those values are assigned in the following code.
-
-In the following code the particular use of \#Contains(address,
-expression) refers to an expression immediately contained in the
-address.
-
-Addrinstr: /* If ADDRESS keyword alone, environments are swapped.
-\emph{/ if \#Contains (address, taken constant), \& \#Contains
-(address,valueexp), \& \#Contains (address, `WITH') then do call
-EnvAssign TRANSIENT, \#Level, ACTIVE, \#Level call EnvAssign ACTIVE,
-\#Level, ALTERNATE, \#Level call EnvAssign ALTERNATE, \#Level,
-TRANSIENT, \#Level return end /} The environment name will be explicitly
-specified. */ if \#Contains(address,taken constant) then Name =
-\#Instance(address, taken \_ constant) else Name = \#Evaluate(valueexp,
-expression) if length(Name) \textgreater{} \#LimitEnvironmentName then
-call \#Raise `SYNTAX', 29.1, Name
-
-if \#Contains(address,expression) then do /* The command is evaluated
-(but not issued) at this point. */ Command = \#Evaluate (address,
-expression)
-
-if \#Tracing.\#Level == `C' \textbar{} \#Tracing.\#Level == `A' then do
-call \#Trace '\textgreater\textgreater\textgreater! end
-
-end
-
-call AddressSetup /* Note what is specified on the ADDRESS instruction.
-\emph{/ /} If there is no command, the persistent environment is being
-set. */ if \#Contains(address,expression) then do
-
-call EnvAssign ACTIVE, \#Level, TRANSIENT, \#Level
-
-return
-
-end
-
-call CommandIssue Command /* See nnn \emph{/ return /} From Addrinstr */
-
-AddressSetup: /* Note what is specified on the ADDRESS instruction, into
-the TRANSIENT environment. \emph{/ EnvTail = `TRANSIENT.'\#Level /}
-Initialize with defaults. */ \#Env\_Name.EnvTail = '\,'
-
-\#Env\_ Type.I.EnvTail = `NORMAL' \#Env\_ Type.O.EnvTail = `NORMAL'
-\#Env\_ Type.E.EnvTail = `NORMAL'
-
-\#Env\_Resource.I.EnvTail = '\,' \#Env\_Resource.O.EnvTail = `!
-\#Env\_Resource.E.EnvTail ='' /* APPEND / REPLACE does not apply to
-input. */
-
-\#Env\_Position.I.EnvTail = `INPUT' \#Env\_Position.O.EnvTail =
-`REPLACE' \#Env\_Position.E.EnvTail = `REPLACE'
-
-/* If anything follows ADDRESS, it will include the command processor
-name.*/ \#Env\_Name.EnvTail = Name
-
-/* Connections may be explicitly specified. */ if \#Contains (address,
-connection) then do
-
-if \#Contains(connection,input) then do /* input redirection */ if
-\#Contains (resourcei, `STREAM') then do \#Env\_Type.I.EnvTail =
-`STREAM' \#Env\_Resource.1.EnvTail=\#Evaluate(resourcei, VAR\_SYMBOL)
-end if \#Contains (resourcei, `STEM') then do \#Env\_Type.I.EnvTail =
-`STEM'
-
-Temp=\#Instance (resourcei,VAR\_SYMBOL) if \#Parses(Temp, stem /* See
-nnn \emph{/) then call \#Raise `SYNTAX', 53.3, Temp
-\#Env\_Resource.I.EnvTail=Temp end end /} Input */
-
-if \#Contains(connection,output) then /* output redirection */ call
-NoteTarget O
-
-if \#Contains(connection,error) then /* error redirection \emph{/ /} The
-prose on the description of \#Contains specifies that the relevant
-resourceo is used in NoteTarget. */ call NoteTarget E
-
-end /* Connection */
-
-return /* from AddressSetup */
-
-NoteTarget:
-
-/* Note the characteristics of an output resource. */
-
-arg Which /* O or E */
-
-if \#Contains (resourceo, `STREAM') then do
-\#Env\_Type.Which.EnvTail=`STREAM'
-\#Env\_Resource.Which.EnvTail=\#Evaluate(resourceo, VAR\_SYMBOL) end
-
-if \#Contains(resourceo,`STEM') then do \#Env\_Type.Which.EnvTail=`STEM'
-Temp=\#Instance (resourceo, VAR\_SYMBOL) if \#Parses(Temp, stem /* See
-nnn */) then
-
-call \#Raise `SYNTAX', 53.3, Temp
-
-\#Env\_Resource.Which.EnvTail=Temp end
-
-if \#Contains (resourceo,append) then
-\#Env\_Position.Which.EnvTail=`APPEND' return /* From NoteTarget */
-
-EnvAssign: /* Copy the values that name an environment and describe its
-redirections. */ arg Lhs, LhsLevel, Rhs, RhsLevel
-\#Env\_Name.Lhs.LhsLevel = \#Env\_Name.Rhs.RhsLevel \#Env\_
-Type.I.Lhs.LhsLevel = \#Env\_Type.I.Rhs.RhsLevel \#Env\_
-Resource.I.Lhs.LhsLevel = \#Env\_Resource.I.Rhs.RhsLevel
-\#Env\_Position.I.Lhs.LhsLevel = \#Env\_Position.I.Rhs.RhsLevel \#Env\_
-Type.O.Lhs.LhsLevel = \#Env\_Type.O.Rhs.RhsLevel
-\#Env\_Resource.O.Lhs.LhsLevel = \#Env\_Resource.O.Rhs.RhsLevel
-\#Env\_Position.O.Lhs.LhsLevel = \#Env\_Position.O.Rhs.RhsLevel \#Env\_
-Type.E.Lhs.LhsLevel = \#Env\_Type.E.Rhs.RhsLevel
-\#Env\_Resource.E.Lhs.LhsLevel \#Env\_Resource.E.Rhs.RhsLevel
-\#Env\_Position.E.Lhs.LhsLevel \#Env\_Position.E.Rhs.RhsLevel return
+of \texttt{\#Env\_Type.} may be set to
+\texttt{\textquotesingle{}UNUSED\textquotesingle{}} as an alternative to
+\texttt{\textquotesingle{}STEM\textquotesingle{}} and
+\texttt{\textquotesingle{}STREAM\textquotesingle{}} where those values
+are assigned in the following code.
+
+In the following code the particular use of
+\texttt{\#Contains(address,\ expression)} refers to an expression
+immediately contained in the \emph{address}.
+
+\lstinputlisting[language=rexx,label=addressinstruct.rexx,caption=addressinstruct.rexx]{addressinstruct.rexx}
 
@@
 
-The ARG instruction is a shorter form of the equivalent instruction:
-PARSE UPPER ARG template list
+The ARG instruction is a shorter form of the equivalent instruction:\\
+\texttt{PARSE\ UPPER\ ARG}\emph{\texttt{template\ list}}
 
@@
 built-in function, or as part of the execution of a PARSE instruction.
-Assignment involves an expression and a VAR\_SYMBOL. The value of the
-expression is determined; see nnn.
-
-If the VAR\_SYMBOL does not contain a period, or contains only one
-period as its last character, the
-
-value is associated with the VAR\_SYMBOL: call Var Set \#Pool,VAR
-SYMBOL, `0',Value
+Assignment involves an \emph{expression} and a \emph{VAR\_SYMBOL}. The
+value of the \emph{expression} is determined; see nnn.
+
+If the \emph{VAR\_SYMBOL} does not contain a period, or contains only
+one period as its last character, the value is associated with the
+\emph{VAR\_SYMBOL}:
+
+\lstinputlisting[language=rexx,label=assignmentvarpoolzero.rexx,caption=assignmentvarpoolzero.rexx]{assignmentvarpoolzero.rexx}
 
 Otherwise, a name is derived, see nnn. The value is associated with the
-derived name: call Var Set \#Pool,Derived Name,`1',Value
+derived name:
+
+\lstinputlisting[language=rexx,label=assignmentvarpoolone.rexx,caption=assignmentvarpoolone.rexx]{assignmentvarpoolone.rexx}
 
@@
 The CALL instruction is used to invoke a routine, or is used to control
-trapping of conditions. lf a vrefis specified that value is the name of
-the routine to invoke:
-
-if \#Contains (call, vref) then Name = \#Evaluate(vref, var\_symbol)
-
-If a taken\_constant is specified, that name is used. if \#Contains
-(call, taken constant) then Name = \#Instance(call, taken constant)
+trapping of conditions.
+
+lf a \emph{vref} is specified that value is the name of the routine to
+invoke:
+
+\lstinputlisting[language=rexx,label=callcontains.rexx,caption=callcontains.rexx]{callcontains.rexx}
+
+If a \emph{taken\_constant} is specified, that name is used.
+
+\lstinputlisting[language=rexx,label=calltakenconstant.rexx,caption=calltakenconstant.rexx]{calltakenconstant.rexx}
 
 The name is used to invoke a routine, see nnn. If that routine does not
-return a result the RESULT and
-
--RESULT variables become uninitialized: call Var Drop \#Pool, `RESULT',
-`0! call Var Drop \#ReservedPool,'.RESULT', `0'
+return a result the RESULT and .RESULT variables become uninitialized:
+
+\lstinputlisting[language=rexx,label=callvardrop.rexx,caption=callvardrop.rexx]{callvardrop.rexx}
 
@@
 
-If the routine returns a result and the trace setting is `R' or `I' then
-a trace with that result and a tag
-'\textgreater\textgreater\textgreater'' shall be produced, associated
-with the call instruction.
-
-If a callon\_spec is specified: If \#Contains(call,callon spec) then do
-Condition = \#Instance(callon\_spec,callable condition)
-
-\#Instruction.Condition.\#Level = `CALL' If \#Contains(callon spec,
-`OFF') then \#Enabling.Condition.\#Level = `OFF' else
-\#Enabling.Condition.\#Level = `ON'
-
-/* Note whether NAME supplied. */ If Contains (callon spec,taken
-constant) then Name = \#Instance (callable condition, taken\_constant)
-
-else
-
-Name = Condition \#TrapName.Condition.\#Level = Name end
+If the routine returns a result and the trace setting is
+\texttt{\textquotesingle{}R\textquotesingle{}} or
+\texttt{\textquotesingle{}I\textquotesingle{}} then a trace with that
+result and a tag
+\texttt{\textquotesingle{}\textgreater{}\textgreater{}\textgreater{}\textquotesingle{}}
+shall be produced, associated with the call instruction.
+
+If a \emph{callon\_spec} is specified:
+
+\lstinputlisting[language=rexx,label=callcallonspec.rexx,caption=callcallonspec.rexx]{callcallonspec.rexx}
 
@@
 
-For a definition of the syntax of a command, see nnn. A command that is
-not part of an ADDRESS instruction is processed in the ACTIVE
-environment.
-
-Command = \#Evaluate(command, expression)
-
-if \#Tracing.\#Level == `C' \textbar{} \#Tracing.\#Level == `A' then
-call \#Trace `\textgreater\textgreater\textgreater!'
-
-call EnvAssign TRANSIENT, \#Level, ACTIVE, \#Level
-
-call CommandiIssue Command
-
-Commandlssue is also used to describe the ADDRESS instruction:
-
-CommandIssue: parse arg Cmd /* Issues the command, requested environment
-is TRANSIENT \emph{/ /} This description does not require the command
-processor to understand stems, so it uses an altered environment. */
-call EnvAssign PASSED, \#Level, TRANSIENT, \#Level EnvTail =
-`TRANSIENT.'\#Level
-
-/* Note the command input. \emph{/ if \#Env\_Type.I.EnvTail = `STEM'
-then do /} Check reasonableness of the stem. \emph{/ Stem =
-\#Env\_Resource.I.EnvTail Lines = value(Stem'0') if \datatype(Lines,`W')
-then call \#Raise `SYNTAX',54.1,Stem'0', Lines if Lines\textless0 then
-call \#Raise `SYNTAX',54.1,Stem'0', Lines /} Use a stream for the stem
-*/ \#Env\_ Type.I.PASSED.\#Level = `STREAM' call Config Stream Unique
-InputStream = \#Outcome \#Env\_Resource.1I.PASSED.\#Level = InputStream
-call charout InputStream , vl do j = 1 to Lines call lineout
-InputStream, value(Stem \textbar\textbar{} j) end j call lineout
-InputStream end
-
-/* Note the command output. */
-
-if \#Env\_Type.O.EnvTail = `STEM' then do Stem =
-\#Env\_Resource.O.EnvTail if \#Env\_Position.O.EnvTail == `APPEND' then
-do
-
-/* Check that Stem.0 will accept incrementing. \emph{/ Lines=value
-(Stem'0'); if \datatype(Lines,`W') then call \#Raise
-`SYNTAX',54.1,Stem'0', Lines if Lines\textless0 then call \#Raise
-`SYNTAX',54.1,Stem'0', Lines end else call value Stem'0',O /} Use a
-stream for the stem */ \#Env\_Type.O.PASSED.\#Level = `STREAM' call
-Config Stream Unique \#Env\_Resource.O.PASSED.\#Level = \#Outcome end
-
-/* Note the command error stream. */
-
-if \#Env\_Type.E.EnvTail = `STEM' then do Stem =
-\#Env\_Resource.E.EnvTail if \#Env\_Position.E.EnvTail == `APPEND' then
-do
-
-/* Check that Stem.0 will accept incrementing. */ Lines=value (Stem'0');
-if \datatype(Lines,`W') then call \#Raise `SYNTAX',54.1,Stem'0', Lines
-if Lines\textless0 then call \#Raise `SYNTAX',54.1,Stem'0', Lines
-
-end else call value Stem'0',0O /* Use a stream for the stem */ \#Env\_
-Type.E.PASSED.\#Level = `STREAM' call Config Stream Unique
-\#Env\_Resource.E.PASSED.\#Level = \#Outcome end
-
-\#API Enabled = `1'
-
-call Var\_Reset \#Pool
-
-/* Specifying PASSED here implies all the
-
-components of that environment. */
-
-\#Response = Config Command (PASSED, Cmd) \#Indicator = left
-(\#Response,1) Description = substr (\#Response, 2)
-
-\#API Enabled = `0'
-
-/* Recognize success and failure. */ if \#AtPause = 0 then do
-
-call value `RC', \#RC
-
-call var Set 0, `.RC', 0, \#RC end
-
-select when \#Indicator==`N' then Temp=0
-
-when \#Indicator==`F' then Temp=-1 /* Failure \emph{/ when
-\#Indicator==`E' then Temp=1 /} Error \emph{/ end call Var Set 0, `.RS',
-0, Temp /} Process the output \emph{/ if \#Env\_Type.O.EnvTail=`STEM'
-then do /} get output into stem. \emph{/ Stem =
-\#Env\_Resource.0O.EnvTail OutputStream =
-\#Env\_Resource.0O.PASSED.\#Level do while lines (OutputStream)
-\textgreater{} 0 call value Stem'0O',value(Stem'0')4+1 call value
-Stem\textbar{} \textbar value(Stem'0'),linein (OutputStream) end end /}
-Stemmed Output \emph{/ if \#Env\_Type.E.EnvTail=`STEM' then do /} get
-error output into stem. */ Stem = \#Env\_Resource.E.EnvTail OutputStream
-= \#Env\_Resource.E.PASSED.\#Level do while lines (OutputStream)
-\textgreater{} 0 call value Stem'0O',value(Stem'0')4+1 call value
-Stem\textbar{} \textbar value(Stem'0'),linein (OutputStream)
-
-end end /* Stemmed Error output */ if \#Indicator == `N' \&
-pos(\#Tracing.\#Level, `CAIR') \textgreater{} 0 then call \#Trace `+++'
-if (\#Indicator == `N' \& \#Tracing.\#Level==`E'), \textbar{}
-(\#Indicator==`F' \& (\#Tracing.\#Level==`F' \textbar{}
-\#Tracing.\#Level==`N')) then do
-
-call \#Trace `\textgreater\textgreater\textgreater!' call \#Trace `+++!'
-
-end \#Condition=`FAILURE' if \#Indicator=`F' \&
-\#Enabling.\#Condition.\#Level == `OFF' then call \#Raise `FAILURE' ,
-Cmd else if \#Indicator=`E' \textbar{} \#Indicator=`F' then call \#Raise
-`ERROR', Cmd
-
-return /* From CommandIssue */
+For a definition of the syntax of a command, see nnn.
+
+A command that is not part of an ADDRESS instruction is processed in the
+ACTIVE environment.
+
+\lstinputlisting[language=rexx,label=commandtoconfifuration.rexx,caption=commandtoconfifuration.rexx]{commandtoconfifuration.rexx}
+
+\texttt{CommandIssue} is also used to describe the ADDRESS instruction:
+
+\lstinputlisting[language=rexx,label=commandIssue.rexx,caption=commandIssue.rexx]{commandIssue.rexx}
 
@@
 The DO instructions is used to group instructions together and
-optionally to execute them repeatedly. Executing a do\_simple has the
-same effect as executing a nop, except in its trace output. Executing
-the do\_ending associated with a do\_simple has the same effect as
-executing a nop, except in its trace output.
-
-A do\_instruction that does not contain a do\_simple is equivalent,
-except for trace output, to a sequence of instructions in the following
-order.
-
-\#Loop = \#Loop+1 \#Iterate.\#Loop = \#Clause (IterateLabel)
-\#Once.\#Loop = \#Clause (OnceLabel) \#Leave.\#Loop = \#Clause
-(LeaveLabel) if \#Contains (do specification,assignment) then
-\#Identity.\#Loop = \#Instance(assignment, VAR SYMBOL) if \#Contains (do
-specification, repexpr) then if \datatype(repexpr,`W') then call \#Raise
-`SYNTAX', 26.2,repexpr else do \#Repeat.\#Loop = repexpr+0 if
-\#Repeat.\#Loop\textless0 then call \#Raise
-`SYNTAX',26.2,\#Repeat.\#Loop end if \#Contains (do
-specification,assignment) then do \#StartValue.\#Loop = \#Evaluate
-(assignment, expression) if datatype (\#StartValue.\#Loop) == `NUM' then
-call \#Raise `SYNTAX', 41.6, \#StartValue.\#Loop \#StartValue.\#Loop =
-\#StartValue.\#Loop + 0 if \#Contains (do specification,byexpr) then
-\#By.\#Loop = 1 end
-
-The following three assignments are made in the order in which `TO'``,
-`BY' and `FOR' appear in docount; see nnn.
-
-if \#Contains (do specification, toexpr) then do if datatype(toexpr) ==
-`NUM' then call \#Raise `SYNTAX', 41.4, toexpr \#To.\#LOop = toexpr+0 if
-\#Contains (do specification, byexpr) then do if datatype (byexpr)
-==`NUM' then call \#Raise `SYNTAX', 41.5, byexpr \#By.\#Loop = byexpr+0
-if \#Contains (do specification, forexpr) then do if \datatype(forexpr,
-`W') then call \#Raise `SYNTAX', 26.3, forexpr \#For.\#Loop = forexpr+0
-if \#For.\#Loop \textless0 then call \#Raise `SYNTAX', 26.3,
-\#For.\#Loop end if \#Contains (do specification,assignment) then do
-call value \#Identity.\#Loop, \#StartValue.\#Loop end if \#Contains (do
-specification, `OVER') then do Value = \#Evaluate(dorep, expression)
-\#OverArray.\#Loop = Value \textasciitilde{} makearray
-
-\#Repeat.\#Loop = \#OverArray\textasciitilde items /* Count this
-downwards as if repexpr. */ \#Iidentity.\#Loop = \#Instance(dorep, VAR
-SYMBOL) end
-
-call \#Goto \#Once.\#Loop /* to OnceLabel */
-
-IterateLabel:
-
-if \#Contains (do specification, untilexpr) then do Value =
-\#Evaluate(untilexp, expression)
-
-if Value == `1' then leave if Value == `0' then call \#Raise `SYNTAX',
-34.4, Value end
-
-if \#Contains (do specification, assignment) then do t = value
-(\#Identity. \#Loop)
-
-if \#Indicator == `D' then call \#Raise `NOVALUE', \#Identity.\#Loop
-call value \#Identity.\#Loop, t + \#By.\#Loop end
-
-OnceLabel:
-
-if \#Contains (do specification, toexpr) then do if
-\#By.\#Loop\textgreater=0 then do if value(\#Identity.\#Loop)
-\textgreater{} \#To.\#Loop then leave end else do if
-value(\#Identity.\#Loop) \textless{} \#To.\#Loop then leave end end
-
-if \#Contains(dorep, repexpr) \textbar{} \#Contains(dorep, `OVER') then
-do if \#Repeat.\#Loop = 0 then leave \#Repeat.\#Loop = \#Repeat.\#Loop-1
-if \#Contains(dorep, `OVER') then call value \#Identity.\#Loop,
-\#OverArray{[}\#OverArray\textasciitilde items - \#Repeat.\#Loop{]} end
-if \#Contains (do specification, forexpr) then do if \#For.\#Loop = 0
-then leave \#For.\#Loop = \#For.\#Loop - 1 end if \#Contains (do
-specification, whileexpr) then do Value = \#Evaluate(whileexp,
-expression)
-
-if Value == `0' then leave if Value == `1' then call \#Raise `SYNTAX',
-34.3, Value end \#Execute (do instruction, instruction list) TraceOfEnd:
-call \#Goto \#Iterate.\#Loop /* to IterateLabel */ LeaveLabel:
-
-\#Loop = \#Loop - 1
+optionally to execute them repeatedly.
+
+Executing a \emph{do\_simple} has the same effect as executing a
+\emph{nop}, except in its trace output. Executing the \emph{do\_ending}
+associated with a \emph{do\_simple} has the same effect as executing a
+\emph{nop}, except in its trace output.
+
+A \emph{do\_instruction} that does not contain a \emph{do\_simple} is
+equivalent, except for trace output, to a sequence of instructions in
+the following order.
+
+\lstinputlisting[language=rexx,label=doinstruction.rexx,caption=doinstruction.rexx]{doinstruction.rexx}
+
+The following three assignments are made in the order in which
+\texttt{\textquotesingle{}TO\textquotesingle{}},
+\texttt{\textquotesingle{}BY\textquotesingle{}} and
+\texttt{\textquotesingle{}FOR\textquotesingle{}} appear in
+\emph{docount}; see nnn.
+
+\lstinputlisting[language=rexx,label=doassignments.rexx,caption=doassignments.rexx]{doassignments.rexx}
 
@@
 
-When clauses are being traced by \#TraceSource, due to
-pos(\#Tracing.\#Level, `AIR') \textgreater{} 0, the DO instruction shall
-be traced when it is encountered and again each time the IterateLabel
-(see nnn) is encountered. The END instruction shall be traced when the
-TraceOfEnd label is encountered.
+When clauses are being traced by \texttt{\#TraceSource}, due to
+\texttt{pos(\#Tracing.\#Level,\ \textquotesingle{}AIR\textquotesingle{})\ \textgreater{}\ 0},
+the DO instruction shall be traced when it is encountered and again each
+time the \texttt{IterateLabel} (see nnn) is encountered. The END
+instruction shall be traced when the \texttt{TraceOfEnd} label is
+encountered.
 
@@
 in the order specified by nnn. Hence, in the absence of conditions
-arising, those executed prior to the first execution of OnceLabel shall
-be shown once per execution of the DO instruction; others shall be shown
-depending on the outcome of the tests.
-
-The code in the DO description: t = value (\#Identity. \#Loop) if
-\#Indicator == `D' then call \#Raise `NOVALUE', \#Identity.\#Loop call
-value \#Identity.\#Loop, t + \#By.\#Loop
+arising, those executed prior to the first execution of
+\texttt{OnceLabel} shall be shown once per execution of the DO
+instruction; others shall be shown depending on the outcome of the
+tests.
+
+The code in the DO description:
+
+\lstinputlisting[language=rexx,label=dolooptracing.rexx,caption=dolooptracing.rexx]{dolooptracing.rexx}
 
@@
 not. When tracing intermediates, the BY value will have a tag of
-`\textgreater+\textgreater{}',
+\texttt{\textquotesingle{}\textgreater{}+\textgreater{}\textquotesingle{}}.
 
@@
 
-The words of the variable\_list are processed from left to right.
+The words of the \emph{variable\_list} are processed from left to right.
 
@@
 variable to be dropped. If VAR\_SYMBOL does not contain a period, or has
-only a single period as its last character, the variable
-
-associated with VAR\_SYMBOL by the variable pool is dropped:
-
-\#Response = Var Drop (\#Pool,VAR\_ SYMBOL, `0')
+only a single period as its last character, the variable associated with
+VAR\_SYMBOL by the variable pool is dropped:
+
+\lstinputlisting[language=rexx,label=drop.rexx,caption=drop.rexx]{drop.rexx}
 
@@
 
-\#Response = Var Drop (\#Pool,VAR SYMBOL, `1')
-
-If the word of the variable\_list is a VAR\_SYMBOL enclosed in
-parentheses then the value of the
-
-VAR\_SYMBOL is processed. The value is considered in uppercase: \#Value
-= Config Upper (\#Value)
+\lstinputlisting[language=rexx,label=drop2.rexx,caption=drop2.rexx]{drop2.rexx}
+
+If the word of the \emph{variable\_list} is a VAR\_SYMBOL enclosed in
+parentheses then the value of the VAR\_SYMBOL is processed. The value is
+considered in uppercase:
+
+\lstinputlisting[language=rexx,label=dropconfigupper.rexx,caption=dropconfigupper.rexx]{dropconfigupper.rexx}
 
@@
 If the word does not have the syntax of VAR\_SYMBOL a condition is
-raised: call \#Raise `SYNTAX', 20.1, word
+raised:
+
+\lstinputlisting[language=rexx,label=dropnotsymbolcondition.rexx,caption=dropnotsymbolcondition.rexx]{dropnotsymbolcondition.rexx}
 
 Otherwise the VAR\_SYMBOL indicated by the word is dropped, as if that
-VAR\_SYMBOL were a word of the variable\_list.
+VAR\_SYMBOL were a word of the \emph{variable\_list}.
 
@@
 
-Any expression is evaluated:
-
-if \#Contains(exit, expression) then Value = \#Evaluate(exit,
-expression) \#Level = 1
-
-\#Pool = \#Pooll
-
-The opportunity is provided for a final trap. \#API Enabled = `1' call
-Var\_Reset \#Pool call Config Termination \#API Enabled = `0'
-
-The processing of the program is complete. See nnn for what API Start
-returns as the result.
+Any \emph{expression} is evaluated:
+
+\lstinputlisting[language=rexx,label=exitevaluate.rexx,caption=exitevaluate.rexx]{exitevaluate.rexx}
+
+The opportunity is provided for a final trap.
+
+\lstinputlisting[language=rexx,label=finaltrap.rexx,caption=finaltrap.rexx]{finaltrap.rexx}
+
+The processing of the program is complete. See nnn for what
+\texttt{API\_Start} returns as the result.
 
@@
 
-We need a check that this starts method; similarities with PROCEDURE.
+\emph{We need a check that this starts method; similarities with
+PROCEDURE.}
 
@@
 It is used at the start of a method, after method initialization, to
-make variables in the receiver's pool
-
-accessible: if \#AllowExpose then call \#Raise `SYNTAX', 17.2
-
-The words of the variable\_list are processed from left to right.
+make variables in the receiver's pool accessible:
+
+\lstinputlisting[language=rexx,label=allowexpose.rexx,caption=allowexpose.rexx]{allowexpose.rexx}
+
+The words of the \emph{variable\_list} are processed from left to right.
 
@@
 associated with VAR\_SYMBOL by the variable pool (as a non-tailed name)
-is given the
-
-attribute `exposed'. call Var\_ Expose \#Pool, VAR SYMBOL, `0'
+is given the attribute `exposed'.
+
+\lstinputlisting[language=rexx,label=exposenoperiod.rexx,caption=exposenoperiod.rexx]{exposenoperiod.rexx}
 
 If VAR\_SYMBOL has a period other than as last character, the variable
-associated with VAR\_SYMBOL
-
-in the variable pool ( by the name derived from VAR\_SYMBOL, see nnn) is
-given the attribute `exposed'. call Var\_ Expose \#Pool, Derived Name,
-`1'
-
-If the word from the variable\_list is a VAR\_SYMBOL enclosed in
+associated with VAR\_SYMBOL in the variable pool ( by the name derived
+from VAR\_SYMBOL, see nnn) is given the attribute `exposed'.
+
+\lstinputlisting[language=rexx,label=exposederivedname.rexx,caption=exposederivedname.rexx]{exposederivedname.rexx}
+
+If the word from the \emph{variable\_list} is a VAR\_SYMBOL enclosed in
 parentheses then the VAR\_SYMBOL is exposed, as if that VAR\_SYMBOL was
-a word in the variable\_list. The value of the VAR\_SYMBOL is
-
-processed. The value is considered in uppercase: \#Value = Config Upper
-(\#Value)
-
-Each word in that value found by the WORD built-in function, from left
-to right, is subjected to this process:
+a word in the variable\_list. The value of the VAR\_SYMBOL is processed.
+The value is considered in uppercase:
+
+\lstinputlisting[language=rexx,label=configupper.rexx,caption=configupper.rexx]{configupper.rexx}
+the WORD built-in function, from left to right, is subjected to this
+process:
 
 If the word does not have the syntax of VAR\_SYMBOL a condition is
-raised: call \#Raise `SYNTAX', 20.1, word
-
-Otherwise the VAR\_SYMBOL indicated by the word is exposed, as if that
-VAR\_SYMBOL were a word of the variable\_list.
+\lstinputlisting[language=rexx,label=raisesyntaxnotsymbol2.rexx,caption=raisesyntaxnotsymbol2.rexx]{raisesyntaxnotsymbol2.rexx}
+VAR\_SYMBOL indicated by the word is exposed, as if that VAR\_SYMBOL
+were a word of the \emph{variable\_list}.
 
@@
 The FORWARD instruction is used to send a message based on the current
-message. if \#Contains (forward, `ARRAY') \& \#Contains(forward,
-`ARGUMENTS') then call \#Raise `SYNTAX', nn.n
+message.
+
+\lstinputlisting[language=rexx,label=forward.rexx,caption=forward.rexx]{forward.rexx}
 
@@
 The GUARD instruction is used to conditionally delay the execution of a
-method. do forever Value = \#Evaluate( guard, expression)
-
-if Value == `1' then leave
-
-if Value == `0' then call \#Raise `SYNTAX', 34.7, Value Drop exclusive
-access and wait for change
-
-end \#\#\# IF
+method.
+
+\lstinputlisting[language=rexx,label=guardinstruction.rexx,caption=guardinstruction.rexx]{guardinstruction.rexx}
+\lstinputlisting[language=rexx,label=dropexclusive.rexx,caption=dropexclusive.rexx]{dropexclusive.rexx}
+
+\subsection{IF}\label{if}
 
@@
 The IF instruction is used to conditionally execute an instruction, or
-to select between two alternatives. The expression is evaluated. If the
-value is neither `0' nor `1' error 34.1 occurs. If the value is `1', the
-instruction in the then is executed. If the value is `0' and e/se is
-specified, the instruction in the else is executed.
+to select between two alternatives. The \emph{expression} is evaluated.
+If the value is neither \texttt{\textquotesingle{}0\textquotesingle{}}
+nor \texttt{\textquotesingle{}1\textquotesingle{}} error 34.1 occurs. If
+the value is `1', the \emph{instruction} in the \emph{then} is executed.
+If the value is \texttt{\textquotesingle{}0\textquotesingle{}} and
+\emph{else} is specified, the instruction in the \emph{else} is
+executed.
 
@@
 
-The expression is evaluated.
+The \emph{expression} is evaluated.
 
@@
 The process of syntactic recognition described in clause 6 is applied,
-with Config\_SourceChar obtaining its results from the characters of the
-value, in left-to-right order, without producing any EOL or EOS events.
-When the characters are exhausted, the event EOL occurs, followed by the
-event EOS.
-
-If that recognition would produce any message then the interpret raises
-the corresponding `SYNTAX' condition.
-
-If the program recognized contains any LABELs then the interpret raises
-a condition: call \#Raise `SYNTAX',47.1,Label
-
-where Label is the first LABEL in the program.
-
-Otherwise the instruction\_list in the program is executed.
+with \texttt{Config\_SourceChar} obtaining its results from the
+characters of the value, in left-to-right order, without producing any
+\texttt{EOL} or \texttt{EOS} events. When the characters are exhausted,
+the event \texttt{EOL} occurs, followed by the event \texttt{EOS}. If
+that recognition would produce any message then the \emph{interpret}
+raises the corresponding `SYNTAX' condition.
+
+If the program recognized contains any LABELs then the \emph{interpret}
+raises a condition:
+
+\lstinputlisting[language=rexx,label=raiselabel.rexx,caption=raiselabel.rexx]{raiselabel.rexx}
+LABEL in the \emph{program}.
+
+Otherwise the \emph{instruction\_list} in the \emph{program} is
+executed.
 
@@
 The ITERATE instruction is used to alter the flow of control within a
-repetitive DO. For a definition of the nesting correction, see nnn.
-
-\#Loop = \#Loop - NestingCorrection call \#Goto \#Iterate.\#Loop
+repetitive DO.
+
+For a definition of the nesting correction, see nnn.
+
+\lstinputlisting[language=rexx,label=instructioniterate.rexx,caption=instructioniterate.rexx]{instructioniterate.rexx}
 
@@
 The execution of a label has no effect, other than clause termination
-activity and any tracing. if \#Tracing.\#Level==`L' then call
-\#TraceSource
+activity and any tracing.
+
+\lstinputlisting[language=rexx,label=traceexecutelabel.rexx,caption=traceexecutelabel.rexx]{traceexecutelabel.rexx}
 
@@
 The LEAVE instruction is used to immediately exit one or more repetitive
-DOs. For a definition of the nesting correction, see nnn.
-
-\#Loop = \#Loop - NestingCorrection call \#Goto \#Leave.\#Loop
+DOs.
+
+For a definition of the nesting correction, see nnn.
+
+\lstinputlisting[language=rexx,label=instructionleave.rexx,caption=instructionleave.rexx]{instructionleave.rexx}
 
@@
 
-We can do this by reference to method invokation, just as we do CALL by
-reference to invoking a function.
+\emph{We can do this by reference to method invokation, just as we do
+CALL by reference to invoking a function.}
 
@@
 
-Shares most of it's definition with repetitive DO.
+\emph{Shares most of it's definition with repetitive DO.}
 
@@
 
-For a definition of the syntax of this instruction, see nnn. The NOP
-instruction has no effect other than the effects associated with all
-instructions.
+For a definition of the syntax of this instruction, see nnn.
+
+The NOP instruction has no effect other than the effects associated with
+all instructions.
 
@@
 
-For a definition of the syntax of this instruction, see nnn. The NUMERIC
-instruction is used to change the way in which arithmetic operations are
-carried out.
+For a definition of the syntax of this instruction, see nnn.
+
+The NUMERIC instruction is used to change the way in which arithmetic
+operations are carried out.
 
@@
 
-if \#Contains(numericdigits, expression) then
-
-Value = \#Evaluate(numericdigits, expression) else Value = 9
-
-if \datatype(Value,`W') then
-
-call \#Raise `SYNTAX',26.5,Value Value = Value + 0 if
-Value\textless=\#Fuzz.\#Level then
-
-call \#Raise `SYNTAX',33.1,Value if Value\textgreater\#Limit Digits then
-
-call \#Raise `SYNTAX',33.2,Value \#Digits.\#Level = Value
+\lstinputlisting[language=rexx,label=numericdigits.rexx,caption=numericdigits.rexx]{numericdigits.rexx}
 
@@
 The value of form is either taken directly from the SCIENTIFIC or
-ENGINEERING keywords, or by
-
-evaluating valueexp . if \#Contains (numeric,numericsuffix) then
-
-Value = `SCIENTIFIC' else if \#Contains (numericformsuffix,
-`SCIENTIFIC') then Value = `SCIENTIFIC' else if \#Contains
-(numericformsuffix, `ENGINEERING') then Value = `ENGINEERING' else do
-Value = \#Evaluate (numericformsuffix,valueexp) Value = translate (left
-(Value,1)) select when Value == `S' then Value = `SCIENTIFIC' when Value
-== `E' then Value = `ENGINEERING' otherwise call \#Raise
-`SYNTAX',33.3,Value end end \#Form.\#Level = Value
+ENGINEERING keywords, or by evaluating \emph{valueexp}.
+
+\lstinputlisting[language=rexx,label=numericform.rexx,caption=numericform.rexx]{numericform.rexx}
 
@@
 
-For a definition of the syntax of this instruction, see nnn. NUMERIC
-FUZZ controls how many digits, at full precision, will be ignored during
-a numeric comparison.
-
-If \#Contains (numericfuzz,expression) then Value = \#Evaluate
-(numericfuzz,expression) else Value = 0 If \datatype(Value,`W') then
-call \#Raise `SYNTAX',26.6,Value Value = Value+0 If Value \textless{} 0
-then call \#Raise `SYNTAX',26.6,Value If Value \textgreater=
-\#Digits.\#Level then call \#Raise `SYNTAX',33.1,\#Digits.\#Level,Value
-\#Fuzz.\#Level = Value
+For a definition of the syntax of this instruction, see nnn.
+
+NUMERIC FUZZ controls how many digits, at full precision, will be
+ignored during a numeric comparison.
+
+\lstinputlisting[language=rexx,label=numericfuzz.rexx,caption=numericfuzz.rexx]{numericfuzz.rexx}
 
@@
 
-The expression is evaluated and the value is passed to the language
-processor. The language processor treats the value as a series of blank
-delimited words. Any words in the value that are not recognized by
-
-the language processor are ignored without producing an error. call
-Config Options (Expression)
+The \emph{expression} is evaluated and the value is passed to the
+language processor. The language processor treats the value as a series
+of blank delimited words. Any words in the value that are not recognized
+by the language processor are ignored without producing an error.
+
+\lstinputlisting[language=rexx,label=configoptions.rexx,caption=configoptions.rexx]{configoptions.rexx}
 
@@
 The purpose of the PARSE instruction is to select substrings of the
-parse\_fype under control of the template\_list. \textbar f the
-template\_list is omitted, or a template in the list is omitted, then a
-template which is the null string is implied.
+\emph{parse\_fype} under control of the \emph{template\_list}. If the
+\emph{template\_list} is omitted, or a \emph{template} in the list is
+omitted, then a template which is the null string is implied.
 
@@
 
-ArgNum = 0 select when \#Contains (parse type, `ARG') then do ArgNum = 1
-ToParse = \#Arg.\#Level.ArgNum end when \#Contains (parse type,
-`LINEIN') then ToParse = linein('`) when \#Contains (parse type, 'PULL')
-then do /* Acquire from external queue or default input. */ \#Response =
-Config Pull() if left(\#Response, 1) == `F' then call Config Default
-Input ToParse = \#Outcome end when \#Contains (parse type, `SOURCE')
-then ToParse = \#Configuration \#HowInvoked \#Source when \#Contains
-(parse type, `VALUE') then if \#Contains(parse value, expression) then
-ToParse = '\,' else ToParse = \#Evaluate(parse value, expression) when
-\#Contains (parse type, `VAR') then ToParse = \#Evaluate (parse var,
-VAR\_SYMBOL) when \#Contains (parse type, `VERSION') then ToParse =
-\#Version end Uppering = \#Contains(parse, `UPPER') The first template
-is associated with this source. If there are further templates, they are
-matched against null strings unless `ARG' is specified, when they are
-matched against further arguments. The parsing process is defined by the
-following routine, ParseData. The template\_list is accessed by
-ParseData as a stemmed variable. This variable Template. has elements
-which are null strings except
-
-for any elements with tails 1,2,3,\ldots{} corresponding to the tokens
-of the template\_list from left to right.
-
-ParseData: /* Targets will be flagged as the template is examined.
-\emph{/ Target.=`0' /} Token is a cursor on the components of the
-template, moved by FindNextBreak. \emph{/ Token = 1 /} Tok ig a cursor
-on the components of the template moved through the target variables by
-routine WordParse. */ Tok = 1
-
-do forever /* Until commas dealt with. \emph{/ /} BreakStart and
-BreakEnd indicate the position in the source string where there is a
-break that divides the source. When the break is a pattern they are the
-start of the pattern and the position just beyond it. */ BreakStart =
-BreakEnd = 1 SourceEnd = length(ToParse) + 1 If Uppering then ToParse =
-translate (ToParse)
-
-do while Template.Tok == '\,' \& Template.Tok == `,'
-
-/* Isolate the data to be processed on this iteration. \emph{/ call
-FindNextBreak /} Also marks targets. */
-
-/* Results have been set in DataStart which indicates the start of the
-isolated data and BreakStart and BreakEnd which are ready for the next
-iteration. Tok has not changed. */
-
-/* If a positional takes the break leftwards from the end of the
-previous selection, the source selected is the rest of the string, */
-
-if BreakEnd \textless= DataStart then DataEnd = SourceEnd
-
-else DataEnd = BreakStart
-
-/* Isolated data, to be assigned from: */
-
-Data=substr (ToParse,DataStart, DataEnd-DataStart) call WordParse /*
-Does the assignments. */
-
-end /* while \emph{/ if Template.Tok == `,' then leave /} Continue with
-next source. */ Token=Token+1 Tok=Token if ArgNum
-\textless\textgreater{} 0 then do ArgNum = ArgNum+1 ToParse =
-\#Arg.ArgNum end else ToParse='\,' end
-
-return /* from ParseData */
-
-FindNextBreak: do while Template.Token == '\,' \& Template.Token == `,'
-
-Type=left (Template.Token,1) /* The source data to be processed next
-will normally start at the end of the break that ended the previous
-piece. (However, the relative positionals alter this.) */ DataStart =
-BreakEnd select
-
-when Type=`\,``' \textbar{} Type=''\,''' \textbar{} Type=`(' then do if
-Type=`(' then do /* A parenthesis introduces a pattern which is not a
-constant. */ Token = Token+1 Pattern = value(Template.Token)
-
-if \#Indicator == `D' then call \#Raise `NOVALUE', Template.Token Token
-= Token+1 end
-
-else
-
-/* The following removes the outer quotes from the literal pattern
-\emph{/ interpret ``Pattern=''Template.Token Token = Token+1 /} Is that
-pattern in the remaining source? \emph{/ PatternPos=pos (Pattern,
-ToParse,DataStart) if PatternPos\textgreater0 then do /} Selected source
-runs up to the pattern. \emph{/ BreakStart=PatternPos
-BreakEnd=PatternPos+length (Pattern) return end leave /} The rest of the
-source is selected. */ end
-
-when datatype(Template.Token,`W') \textbar{} pos(Type,`+-=')
-\textgreater{} 0 then do /* A positional specifies where the relevant
-piece of the subject ends. \emph{/ if pos (Type, `+-=') = 0 then do /}
-Whole number positional \emph{/ BreakStart = Template.Token Token =
-Token+1 end else do /} Other forms of positional. \emph{/
-Direction=Template.Token Token = Token + 1 /} For a relative positional,
-the position is relative to the start of the previous trigger, and the
-source segment starts there. \emph{/ if Direction == `=' then DataStart
-= BreakStart /} The adjustment can be given as a number or a variable in
-parentheses. */ if Template.Token =`(' then do Token=Token + 1
-BreakStart = value(Template. Token) if \#Indicator == `D' then call
-\#Raise `NOVALUE', Template.Token Token=Token + 1 end else BreakStart =
-Template.Token
-
-if
-
-if
-
-el
-
-\datatype (BreakStart,`W')
-
-then call \#Raise `SYNTAX', 26.4,BreakStart Token = Token+1
-
-Direction=`+'
-
-then BreakStart=DataStart+BreakStart se if Direction=`-'
-
-then BreakStart=DataStart-BreakStart
-
-end /* Adjustment should remain within the ToParse \emph{/ BreakStart =
-max(1, BreakStart) BreakStart = min(SourceEnd, BreakStart) BreakEnd =
-BreakStart /} No actual literal marks the boundary. */ return
-
-end
-
-when Template.Token == `.' \& pos(Type, `0123456789.')\textgreater0 then
-/* A number that isn't a whole number. */
-
-call
-
-\#Raise `SYNTAX', 26.4, Template.Token
-
-/* Raise will not return */
-
-otherwise do /* It is a target, not a pattern */ Target.Token=`1' Token
-= Token+1
-
-end
-
-end /*
-
-select */
-
-end /* while \emph{/ /} When no more explicit breaks, break is at the
-end of the source. */ DataStart=BreakEnd BreakStart=SourceEnd
-BreakEnd=SourceEnd
-
-return
-
-WordParse: /* The names in the template are assigned blank-delimited
-values from the source string. */
-
-/* From FindNextBreak */
-
-do while Target.Tok /* Until no more targets for this data. */
-
-/* Last target gets all the residue of the Data.
-
-Next Tok
-
-= Tok + 1
-
-if \Target.NextTok then do call Assign (Data)
-
-leave end /* Not 1 Data = s if Data else do Word =
-
-ast target; assign a word. */ trip (Data, `L') == '\,' then call
-Assign('\,')
-
-word (Data,1)
-
-call Assign Word
-
-Data =
-
-substr(Data,length(Word) + 1)
-
-*/
-
-/* The word terminator is not part of the residual data: */ if Data ==
-'\,' then Data = substr (Data, 2)
-
-end
-
-Tok = Tok + 1
-
-*/
-
-end Tok=Token /* Next time start on new part of template. \emph{/ return
-Assign: if Template.Tok==`.' then Tag=`\textgreater.\textgreater{}' else
-do Tag=`\textgreater=\textgreater{}' call value Template.Tok,arg(1) end
-/} Arg(1) is an implied argument of the tracing. if \#Tracing.\#Level ==
-`R' \textbar{} \#Tracing.\#Level == `I'
-
-return
-
-\subsection{PROCEDURE}\label{procedure}
-
-For a definition of the syntax of this instruction, see nnn.
-
-then call \#Trace Tag The PROCEDURE instruction is used within an
-internal routine to protect all the existing variables by making them
-unknown to following instructions. Selected variables may be exposed.
+\lstinputlisting[language=rexx,label=instructionsparse.rexx,caption=instructionsparse.rexx]{instructionsparse.rexx}
+
+The first template is associated with this source. If there are further
+templates, they are matched against null strings unless `ARG' is
+specified, when they are matched against further arguments.
+
+The parsing process is defined by the following routine,
+\texttt{ParseData}. The \emph{template\_list} is accessed by
+\texttt{ParseDat}a as a stemmed variable. This variable
+\texttt{Template.} has elements which are null strings except for any
+elements with tails 1,2,3,\ldots{} corresponding to the tokens of the
+\emph{template\_list} from left to right.
+
+\lstinputlisting[language=rexx,label=parsedata.rexx,caption=parsedata.rexx]{parsedata.rexx}
+
+For a definition of the syntax of this instruction, see nnn.
+
+The PROCEDURE instruction is used within an internal routine to protect
+all the existing variables by making them unknown to following
+instructions. Selected variables may be exposed.
 
@@
 
-if \#AllowProcedure.\#Level then call \#Raise `SYNTAX', 17.1
-
-\#AllowProcedure.\#Level = 0
-
-/* It introduces a new variable pool: */
-
-call \#Config ObjectNew
-
-call var\_set (\#Outcome,`\#UPPER', `0', \#Pool) /* Previous \#Pool is
-upper from the new \#Pool. */
-
-\#Pool=\#O0Outcome
-
-IsProcedure.\#Level=`1'
-
-call Var\_Empty \#Pool
-
-If there is a variable\_list, it provides access to a previous variable
-pool.
-
-The words of the variable\_list are processed from left to right.
+\lstinputlisting[language=rexx,label=procedure.rexx,caption=procedure.rexx]{procedure.rexx}
+
+If there is a \emph{variable\_list}, it provides access to a previous
+variable pool.
+
+The words of the \emph{variable\_list} are processed from left to right.
 
@@
 associated with VAR\_SYMBOL by the variable pool (as a non-tailed name)
-is given the
-
-attribute `exposed'. call Var\_ Expose \#Pool, VAR SYMBOL, `0'
+is given the attribute `exposed'.
+
+\lstinputlisting[language=rexx,label=exposepool.rexx,caption=exposepool.rexx]{exposepool.rexx}
 
 If VAR\_SYMBOL has a period other than as last character, the variable
-associated with VAR\_SYMBOL
-
-in the variable pool ( by the name derived from VAR\_SYMBOL, see nnn) is
-given the attribute `exposed'. call Var\_ Expose \#Pool, Derived Name,
-`1'
-
-If the word from the variable\_list is a VAR\_SYMBOL enclosed in
+associated with VAR\_SYMBOL in the variable pool (by the name derived
+from VAR\_SYMBOL, see nnn) is given the attribute `exposed'.
+
+\lstinputlisting[language=rexx,label=exposederived.rexx,caption=exposederived.rexx]{exposederived.rexx}
+
+If the word from the \emph{variable\_list} is a VAR\_SYMBOL enclosed in
 parentheses then the VAR\_SYMBOL is exposed, as if that VAR\_SYMBOL was
-a word in the variable\_list. Tne value of the VAR\_SYMBOL is
-
-processed. The value is considered in uppercase: \#Value = Config Upper
-(\#Value)
+a word in the \emph{variable\_list}. Tne value of the VAR\_SYMBOL is
+processed. The value is considered in uppercase:
+
+\lstinputlisting[language=rexx,label=configuppervalue.rexx,caption=configuppervalue.rexx]{configuppervalue.rexx}
 
@@
 If the word does not have the syntax of VAR\_SYMBOL a condition is
-raised: call \#Raise `SYNTAX', 20.1, word
+raised:
+
+\lstinputlisting[language=rexx,label=callraisesyntax.rexx,caption=callraisesyntax.rexx]{callraisesyntax.rexx}
 
 Otherwise the VAR\_SYMBOL indicated by the word is exposed, as if that
-VAR\_SYMBOL were a word of the variable\_list.
+VAR\_SYMBOL were a word of the \emph{variable\_list}.
 
@@
 A PULL instruction is a shorter form of the equivalent instruction:
-PARSE UPPER PULL template list
+
+\lstinputlisting[language=rexx,label=upperpull.rexx,caption=upperpull.rexx]{upperpull.rexx}
 
@@
 
-For a definition of the syntax of this instruction, see nnn. The PUSH
-instruction is used to place a value on top of the stack.
-
-If \#Contains(push,expression) then Value = \#Evaluate (push,
-expression) else Value = '\,' call Config Push Value
+For a definition of the syntax of this instruction, see nnn.
+
+The PUSH instruction is used to place a value on top of the stack.
+
+\lstinputlisting[language=rexx,label=push.rexx,caption=push.rexx]{push.rexx}
 
@@
 
-For a definition of the syntax of this instruction, see nnn. The QUEUE
-instruction is used to place a value on the bottom of the stack.
-
-If \#Contains (queue,expression) then Value = \#Evaluate (queue,
-expression) else Value = '\,' call Config Queue Value
+For a definition of the syntax of this instruction, see nnn.
+
+The QUEUE instruction is used to place a value on the bottom of the
+stack.
+
+\lstinputlisting[language=rexx,label=queue.rexx,caption=queue.rexx]{queue.rexx}
 
@@
 
-Must set up for error of expression on subsequent RETURN. \#\#\# RETURN
-For a definition of the syntax of this instruction, see nnn. The RETURN
-instruction is used to return control and possibly a result from a
-program or internal routine to the point of its invocation. The RETURN
-keyword may be followed by an optional expression, which will be
-evaluated and returned as a result to the caller of the routine. Any
-expression is evaluated: if \#Contains(return,expression) then \#Outcome
-= \#Evaluate(return, expression)
-
-else if \#IsFunction.\#Level then call \#Raise `SYNTAX', 45.1,
-\#Name.\#Level
+\emph{Must set up for error of expression on subsequent RETURN.} \#\#\#
+RETURN For a definition of the syntax of this instruction, see nnn.
+
+The RETURN instruction is used to return control and possibly a result
+from a program or internal routine to the point of its invocation.
+
+The RETURN keyword may be followed by an optional expression, which will
+be evaluated and returned as a result to the caller of the routine.
+
+Any \emph{expression} is evaluated:
+
+\lstinputlisting[language=rexx,label=return.rexx,caption=return.rexx]{return.rexx}
 
@@
 If the routine started with a PROCEDURE instruction then the associated
-pool is taken out of use: if \#IsProcedure.\#Level then \#Pool = \#Upper
+pool is taken out of use:
+
+\lstinputlisting[language=rexx,label=isprocedure.rexx,caption=isprocedure.rexx]{isprocedure.rexx}
+
 A RETURN instruction which is interactively entered at a pause point
-leaves the pause point. if \#Level = \#AtPause then \#AtPause = 0 The
-activity at this level is complete: \#Level = \#Level-1 \#NewLevel =
-\#Level+1 If \#Level is not zero, the processing of the RETURN
+leaves the pause point.
+
+\lstinputlisting[language=rexx,label=atpause.rexx,caption=atpause.rexx]{atpause.rexx}
+
+The activity at this level is complete:
+
+\lstinputlisting[language=rexx,label=newlevel.rexx,caption=newlevel.rexx]{newlevel.rexx}
+
+If \texttt{\#Level} is not zero, the processing of the RETURN
 instruction and the invocation is complete. Otherwise processing of the
@@
 
-The opportunity is provided for a final trap. \#API Enabled = `1'
-
-call Var\_Reset \#Pool
-
-call Config Termination
-
-\#API Enabled = `0'
-
-The processing of the program is complete. See nnn for what API Start
-returns as the result. \#\#\# SAY
+The opportunity is provided for a final trap.
+
+\lstinputlisting[language=rexx,label=apifinaltrap.rexx,caption=apifinaltrap.rexx]{apifinaltrap.rexx}
+
+The processing of the program is complete. See nnn for what
+\texttt{API\_Start} returns as the result.
+
+\subsection{SAY}\label{say}
 
@@
 
-If \#Contains(say,expression) then Value = Evaluate (say, expression)
-else Value = '\,' call Config Default Output Value
-
-\subsection{SELECT}\label{select}
+\lstinputlisting[language=rexx,label=say.rexx,caption=say.rexx]{say.rexx}
 
@@
 
-The \#Contains(select\_body, when) test in the following description
-refers to the items of the optional when repetition in order:
-
-LineNum = \#LineNumber Ending = \#Clause (EndLabel) Value=\#Evaluate
-(select body, expression) /* In the required WHEN \emph{/ if Value ==
-`1' \& Value == `0' then call \#Raise `SYNTAX',34.2,Value If Value==`1'
-then call \#Execute when, instruction else do do while \#Contains
-(select body, when) Value = \#Evaluate (when, expression) If Value==`1'
-then do call \#Execute when, instruction call \#Goto Ending end if Value
-== `0' then call \#Raise `SYNTAX', 34.2, Value end /} Of each when */
-
-If \#Contains(select body, `OTHERWISE') then call \#Raise `SYNTAX', 7.3,
-LineNum If \#Contains (select body, instruction list) then call
-\#Execute select body, instruction list end EndLabel:
+The \texttt{\#Contains(}\emph{\texttt{select\_body,\ when}}\texttt{)}
+test in the following description refers to the items of the optional
+\emph{when} repetition in order:
+
+\lstinputlisting[language=rexx,label=select.rexx,caption=select.rexx]{select.rexx}
 
@@
 
-If \#Contains (signal,signal spec) then do Condition = \#Instance(signal
-spec,condition)
-
-\#Instruction.Condition.\#Level = `SIGNAL' If \#Contains (signal spec,
-`OFF') then \#Enabling.Condition.\#Level = `OFF' else
-\#Enabling.Condition.\#Level = `ON'
-
-If Contains (signal spec,taken constant) then Name = \#Instance
-(condition, taken constant)
-
-else
-
-Name = Condition \#TrapName.Condition.\#Level = Name end
-
-If there was a signal\_spec this complete the processing of the signal
-instruction. Otherwise: if \#Contains (signal, valueexp)
-
-then Name \#Evaluate(valueexp, expression)
-
-else Name \#Instance(signal,taken constant)
+\lstinputlisting[language=rexx,label=signal.rexx,caption=signal.rexx]{signal.rexx}
+
+If there was a \emph{signal\_spec} this complete the processing of the
+signal instruction. Otherwise:
+
+\lstinputlisting[language=rexx,label=signalspec.rexx,caption=signalspec.rexx]{signalspec.rexx}
 
 The Name matches the first LABEL in the program which has that value.
-The comparison is made with the `==``' operator.
+The comparison is made with the
+\texttt{\textquotesingle{}==\textquotesingle{}} operator.
 
@@
 
-call \#Raise `SYNTAX',16.1, Name
-
-If the name is a trace-only label then a condition is raised: call
-\#Raise `SYNTAX', 16.2, Name
+\lstinputlisting[language=rexx,label=raisesyntaxname.rexx,caption=raisesyntaxname.rexx]{raisesyntaxname.rexx}
+
+If the name is a trace-only label then a condition is raised:
+
+\lstinputlisting[language=rexx,label=raisetraceonlysymbol.rexx,caption=raisetraceonlysymbol.rexx]{raisetraceonlysymbol.rexx}
 
 If the name matches a label, execution continues at that label after
-these settings: \#Loop.\#Level = 0
-
-/* A SIGNAL interactively entered leaves the pause point. */
-
-if \#Level = \#AtPause then \#AtPause = 0
+these settings:
+
+\lstinputlisting[language=rexx,label=executioncontinues.rexx,caption=executioncontinues.rexx]{executioncontinues.rexx}
 
@@
 The TRACE instruction is ignored if it occurs within the program (as
-opposed to source obtained by
-
-Config\_Trace\_Input) and interactive trace is requested
-(\#Interactive.\#Level = `1'). Otherwise: \#TraceInstruction = `1' value
-= '\,' if \#Contains(trace, valueexp) then Value = \#Evaluate(valueexp,
-expression) if \#Contains (trace, taken constant) then Value =
-\#Instance (trace, taken constant) if datatype(Value) == `NUM' \&
-\datatype(Value,`W') then call \#Raise `SYNTAX', 26.7, Value if
-datatype(Value,`W') then do /* Numbers are used for skipping. \emph{/ if
-Value\textgreater=0 then \#InhibitPauses = Value else \#InhibitTrace =
--Value end else do if length(Value) = 0 then do \#Interactive.\#Level =
-`0' Value = `N' end /} Each question mark toggles the interacting. */ do
-while left(Value,1)==`?' \#Interactive.\#Level = \#Interactive.\#Level
-Value = substr(Value,2) end if length(Value) = 0 then do Value =
-translate( left(Value,1) ) if verify(Value, `ACEFILNOR') \textgreater{}
-0 then call \#Raise `SYNTAX', 24.1, Value if Value==`0O' then
-\#Interactive.\#Level=`0' end \#Tracing.\#Level = Value end
+opposed to source obtained by \texttt{Config\_Trace\_Input}) and
+interactive trace is requested
+(\texttt{\#Interactive.\#Level\ =\ \textquotesingle{}1\textquotesingle{}}).
+Otherwise:
+
+\lstinputlisting[language=rexx,label=trace.rexx,caption=trace.rexx]{trace.rexx}
 
@@
 
-If \#NoSource is `1' there is no trace output.
-
-The routines \#TraceSource and \#Trace specify the output that results
-from the trace settings. That output is presented to the configuration
-by Config\_Trace\_Output as lines. Each line has a clause identifier at
-the left, followed by a blank, followed by a three character tag,
-followed by a blank, followed by the trace data.
+If \texttt{\#NoSource} is \texttt{\textquotesingle{}1\textquotesingle{}}
+there is no trace output.
+
+The routines \texttt{\#TraceSource} and \texttt{\#Trace} specify the
+output that results from the trace settings. That output is presented to
+the configuration by \texttt{Config\_Trace\_Output} as lines. Each line
+has a clause identifier at the left, followed by a blank, followed by a
+three character tag, followed by a blank, followed by the trace data.
 
@@
 
-When input at a pause is being executed (\#AtPause = 0 ), \#Trace does
-nothing when the tag is not `+++'.
-
-When input at a pause is being executed, \#TraceSource does nothing. If
-\#InhibitTrace is greater than zero, \#TraceSource does nothing except
-decrement \#InhibitTrace. Otherwise, unless the current clause is a null
-clause, \#TraceSource outputs all lines of the source program which
-contain any part of the current clause, with any characters in those
-lines which are not part of the current clause and not
-other\_blank\_characters replaced by blank characters. The possible
-replacement of other\_blank\_characters is defined by the configuration.
-The tag is '\emph{-}'', or if the line is not the first line of the
-clause. ™,*'. \#Trace output also has a clause identifier and has a tag
-which is the argument to the \#Trace invocation. The data is truncated,
-if necessary, to \#Limit\_TraceData characters. The data is enclosed by
-quotation marks and the quoted data preceded by two blanks. If the data
-is truncated, the trailing quote has the three characters `\ldots{}'
-appended. \_ when \#Tracing.\#Level is `C' or `E' or `F' or `N' or `A'
-and the tag is `\textgreater\textgreater\textgreater{}' then the data is
-the value of the command passed to the environment; \_ when the tag is
-`+++' then the data is the four characters `RC concatenated with the
-character ``\,``; \_ when \#Tracing.\#Level is 'l' or `R' the data is
-the most recently evaluated value. Trace output can also appear as the
-result of a'SYNTAX' condition occurring, irrespective of the trace
-setting. If a'SYNTAX' condition occurs and it is not trapped by SIGNAL
-ON SYNTAX, then the clause in error shall be traced, along with a
-traceback. A traceback is a display of each active CALL and INTERPRET
-instruction, and function invocation, displayed in reverse order of
-execution, each with a tag of `+4+'. \#\#\# USE For a definition of the
-syntax of this instruction, see nnn. The USE instruction assigns the
-values of arguments to variables. Better not say copies since COPY
-method has different semantics. The optional VAR\_SYMBOL positions,
-positions 1, 2, \ldots, of the instruction are considered from left to
-right. If the position has a VAR\_SYMBOL then its value is assigned to:
-
-if \#ArgExists.Position then call Value VAR\_SYMBOL, \#Arg.Position else
-
-Messy because VALUE bif won't DROP and var\_drop needs to know if
-compound.
+When input at a pause is being executed
+(\texttt{\#AtPause\ \textbackslash{}=\ 0}), \texttt{\#Trace} does
+nothing when the tag is not
+\texttt{\textquotesingle{}+++\textquotesingle{}}.
+
+When input at a pause is being executed, \texttt{\#TraceSource} does
+nothing.
+
+If \texttt{\#InhibitTrace} is greater than zero, \texttt{\#TraceSource}
+does nothing except decrement \texttt{\#InhibitTrace}. Otherwise, unless
+the current clause is a null clause, \texttt{\#TraceSource} outputs all
+lines of the source program which contain any part of the current
+clause, with any characters in those lines which are not part of the
+current clause and not \texttt{other\_blank\_characters} replaced by
+blank characters. The possible replacement of
+\texttt{other\_blank\_characters} is defined by the configuration. The
+tag is \texttt{\textquotesingle{}*-*\textquotesingle{}}, or if the line
+is not the first line of the clause.
+\texttt{\textquotesingle{}*,*\textquotesingle{}}.
+
+\texttt{\#Trace} output also has a clause identifier and has a tag which
+is the argument to the \texttt{\#Trace} invocation. The data is
+truncated, if necessary, to \texttt{\#Limit\_TraceData} characters. The
+data is enclosed by quotation marks and the quoted data preceded by two
+blanks. If the data is truncated, the trailing quote has the three
+characters \texttt{\textquotesingle{}...\textquotesingle{}} appended.
+
+\begin{itemize}
+\tightlist
+\item
+  when \texttt{\#Tracing.\#Level} is
+  \texttt{\textquotesingle{}C\textquotesingle{}} or
+  \texttt{\textquotesingle{}E\textquotesingle{}} or
+  \texttt{\textquotesingle{}F\textquotesingle{}} or
+  \texttt{\textquotesingle{}N\textquotesingle{}} or
+  \texttt{\textquotesingle{}A\textquotesingle{}} and the tag is
+  \texttt{\textquotesingle{}\textgreater{}\textgreater{}\textgreater{}\textquotesingle{}}
+  then the data is the value of the command passed to the environment;
+\item
+  when the tag is \texttt{\textquotesingle{}+++\textquotesingle{}} then
+  the data is the four characters
+  \texttt{\textquotesingle{}RC\ "\textquotesingle{}} concatenated with
+  the character \texttt{\textquotesingle{}"\textquotesingle{}};
+\item
+  when \texttt{\#Tracing.\#Level} is
+  \texttt{\textquotesingle{}l\textquotesingle{}} or
+  \texttt{\textquotesingle{}R\textquotesingle{}} the data is the most
+  recently evaluated value.
+\end{itemize}
+
+Trace output can also appear as the result of a'SYNTAX' condition
+occurring, irrespective of the trace setting. If a'SYNTAX' condition
+occurs and it is not trapped by SIGNAL ON SYNTAX, then the clause in
+error shall be traced, along with a traceback. A traceback is a display
+of each active CALL and INTERPRET instruction, and function invocation,
+displayed in reverse order of execution, each with a tag of
+\texttt{\textquotesingle{}+++\textquotesingle{}}.
+
+\subsection{USE}\label{use}
+
+For a definition of the syntax of this instruction, see nnn.
+
+The USE instruction assigns the values of arguments to variables.
+
+\emph{Better not say copies since COPY method has different semantics.}
+
+The optional VAR\_SYMBOL positions, positions 1, 2, \ldots, of the
+instruction are considered from left to right. If the position has a
+VAR\_SYMBOL then its value is assigned to:
+
+\lstinputlisting[language=rexx,label=use.rexx,caption=use.rexx]{use.rexx}
+
+\emph{Messy because VALUE bif won't DROP and var\_drop needs to know if
+compound.}
 
@@
 
-concatenated with \#RC
-
 Error number 3 is available to report error conditions occuring during
@@
 
-call \#Raise Condition, Arg2, Arg3, Arg4, Arg5, Arg6é
+\lstinputlisting[language=rexx,label=detectcondition.rexx,caption=detectcondition.rexx]{detectcondition.rexx}
 
@@
 The action of the program as a result of a condition is dependent on any
-signa/\_spec and callon\_spec in the program.
+\emph{signal\_spec} and \emph{callon\_spec} in the program.
 
@@
 
-The routine \#Raise corresponds to raising a condition. In the following
-definition, the instructions containing SIGNAL VALUE and INTERPRET
-denote transfers of control in the program being processed. The
-instruction EXIT denotes termination. If not at an interactive pause,
-this will be termination of the program, see nnn, and there will be
-output by Config\_Trace\_Output of the message (with prefix \_ see nnn)
-and tracing (see nnn). If at an interactive pause (\#AtPause = 0), this
-will be termination of the interpretation of the interactive input;
-there will be output by Config\_Trace\_Output of the message (without
-traceback) before continuing. The description of the continuation is in
-nnn after the ``interpret \#Outcome'' instruction.
-
-The instruction ``interpret `CALL' \#TrapName.\#Condition.\#Level''
+The routine \texttt{\#Raise} corresponds to raising a condition. In the
+following definition, the instructions containing SIGNAL VALUE and
+INTERPRET denote transfers of control in the program being processed.
+The instruction EXIT denotes termination. If not at an interactive
+pause, this will be termination of the program, see nnn, and there will
+be output by \texttt{Config\_Trace\_Output} of the message (with prefix
+\_ see nnn) and tracing (see nnn). If at an interactive pause
+(\texttt{\#AtPause\ \textbackslash{}=\ 0}), this will be termination of
+the interpretation of the interactive input; there will be output by
+\texttt{Config\_Trace\_Output} of the message (without traceback) before
+continuing. The description of the continuation is in nnn after the
+\texttt{"interpret\ \#Outcome”} instruction.
+
+The instruction
+\texttt{“interpret\ \textquotesingle{}CALL\textquotesingle{}\ \#TrapName.\#Condition.\#Level"}
 below does not set the variables RESULT and .RESULT; any result returned
@@
 
-\#Raise: /* If there is no argument, this is an action which has been
-delayed from the time the condition occurred until an appropriate clause
-boundary. */ if \arg(1,`E') then do Description = \#PendingDescription.
-\#Condition. \#Level Extra = \#PendingExtra.\#Condition. \#Level
-
-end else do \#Condition = arg(1) if \#Condition == `SYNTAX' then do
-
-Description = arg(2) Extra = arg(3) end else do Description =
-\#Message(arg(2),arg(3),arg(4),arg(5)) call Var Set \#ReservedPool,
-`.MN', 0, arg(2) Extra = '! end end
-
-/* The events for disabled conditions are ignored or cause termination.
-*/
-
-if \#Enabling.\#Condition.\#Level == `OFF' \textbar{} \#AtPause = 0 then
-do if \#Condition == `SYNTAX' \& \#Condition == `HALT' then return /* To
-after use of \#Raise. \emph{/ if \#Condition == `HALT' then Description
-= \#Message(4.1, Description) exit /} Terminate with Description as the
-message. */ end
-
-/* SIGNAL actions occur as soon as the condition is raised. */
-
-if \#Instruction.\#Condition.\#Level == ```SIGNAL' then do
-\#ConditionDescription.\#Level = Description \#ConditionExtra.\#Level =
-Extra \#ConditionInstruction.\#Level = `SIGNAL'
-\#Enabling.\#Condition.\#Level = `OFF' signal value
-\#TrapName.\#Condition.\#Level end
-
-/* All CALL actions are initially delayed until a clause boundary. */
-
-if arg(1,`E') then do /* Events within the handler are not stacked up,
-except for one extra HALT while a first is being handled. \emph{/
-EventLevel = \#Level if \#Enabling.\#Condition.\#Level == `DELAYED' then
-do if \#Condition == `HALT' then return EventLevel =
-\#EventLevel.\#Condition. \#Level if \#PendingNow.\#Condition.EventLevel
-then return /} Setup a HALT to come after the one being handled. \emph{/
-end /} Record a delayed event. \emph{/
-\#PendingNow.\#Condition.EventLevel = `1'
-\#PendingDescription.\#Condition.EventLevel = Description
-\#PendingExtra.\#Condition.EventLevel = Extra
-\#Enabling.\#Condition.EventLevel = `DELAYED' return end /} Here for
-CALL action after delay. \emph{/ /} Values for the CONDITION built-in
-function. \emph{/ \#Condition.\#NewLevel = \#Condition
-\#ConditionDescription.\#NewLevel = \#PendingDescription. \#Condition.
-\#Level \#ConditionExtra.\#NewLevel = \#PendingExtra.\#Condition.
-\#Level \#ConditionInstruction.\#NewLevel = `CALL' interpret `CALL'
-\#TrapName.\#Condition.\#Level \#Enabling.\#Condition.\#Level = `ON'
-return /} To clause termination */
+\lstinputlisting[language=rexx,label=interpretraise.rexx,caption=interpretraise.rexx]{interpretraise.rexx}
 
@@
 
-The state function \#Message corresponds to constructing a message.
+The state function \texttt{\#Message} corresponds to constructing a
+message.
 
@@
 
-In addition to the result defined below, the values of MsgNumber and
-\#LineNumber shall be shown when a message is output. Also there shall
-be an indication of whether the error occurred in code executed at an
-interactive pause, see nnn.
+In addition to the result defined below, the values of
+\texttt{MsgNumber} and \texttt{\#LineNumber} shall be shown when a
+message is output. Also there shall be an indication of whether the
+error occurred in code executed at an interactive pause, see nnn.
 
@@
 
-\#Message: MsgNumber = arg(1) if \#NoSource then MsgNumber = MsgNumber
-\% 1 /* And hence no inserts */ Text = \#ErrorText.MsgNumber Expanded =
-'\,'
-
-do Index = 2 parse var Text Begin `\textless{}' Insert `\textgreater{}'
-+1 Text
-
-if Insert = '\,' then leave Insert = arg(Index) if length(Insert)
-\textgreater{} \#Limit MessageInsert then Insert = left(Insert,\#Limit
-MessageInsert)`\ldots{}'
-
-Expanded = Expanded \textbar\textbar{} Begin \textbar\textbar{} Insert
-
-end
-
-Expanded = Expanded \textbar\textbar{} Begin
-
-say Expanded return
+\lstinputlisting[language=rexx,label=executionmessages.rexx,caption=executionmessages.rexx]{executionmessages.rexx}

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/provided_classes.tex
add filenames

--- 2023-05-02 ARB/standards/work-in-progress/document/tex/standard/provided_classes.tex
@@
 
-\hypertarget{collection-class-routines}{%
-\subsection{Collection Class Routines}\label{collection-class-routines}}
+\subsection{Collection Class Routines}\label{collection-class-routines}
 
 These routines are used in the definition of the collection classes
-::routine CommonxXor /* Returns a new collection that contains all items
-from self and the argument except that all indexes that appear in both
-collections are removed. \emph{/ /} When the target is a bag, there may
-be an index in the bag that is duplicated and the same value as an index
-in the argument. Should one copy of the index survive in the bag?
-\emph{/ vel if (arg(1)\textasciitilde class==.Set \&
-arg(2)\textasciitilde class==.Bag) then v=2 if
-(arg(1)\textasciitilde class==.Table \&
-arg(2)\textasciitilde class==.Bag) then v=2 if
-(arg(1)\textasciitilde class==.Table \&
-arg(2)\textasciitilde class==.Relation) then v=2 if
-(arg(1)\textasciitilde class==.Directory \&
-arg(2)\textasciitilde class==.Bag) then v=2 if
-(arg(1)\textasciitilde class==.Directory \&
-arg(2)\textasciitilde class==.Relation) then v=2 /} This version it
-does: \emph{/ if v=1 then do This = arg(1) /} self of caller */
-r=This\textsubscript{class}new ab=MayEnBag (arg (2) )
-ss=This\textasciitilde supplier do while ss\textasciitilde available
-r{[}ss\textasciitilde index{]} =ss\textasciitilde item
-ss\textasciitilde next end cs=ab\textasciitilde supplier do while
-cs\textasciitilde available if r\textsubscript{hasindex(cs}index) then
-r\textasciitilde remove (cs\textasciitilde index) else
-r{[}cs\textasciitilde index{]} =cs\textasciitilde item
-cs\textasciitilde next end return r end
 
-/* But following matches practice on Set\textasciitilde XOR(bag) etc. */
+\lstinputlisting[language=rexx,label=CommonXor.rexx,caption=CommonXor.rexx]{CommonXor.rexx}
 
-This = arg(1) /* self of caller */
+\subsection{The collection class}\label{the-collection-class}
 
-r=This\textsubscript{class}new
+\lstinputlisting[language=rexx,label=collectionclass.rexx,caption=collectionclass.rexx]{collectionclass.rexx}
 
-ab=MayEnBag (arg (2) )
+\subsubsection{INIT}\label{init}
 
-ss=This\textasciitilde supplier
+\lstinputlisting[language=rexx,label=collectioninitmethod.rexx,caption=collectioninitmethod.rexx]{collectioninitmethod.rexx}
 
-do while ss\textasciitilde available if
-\ab\textsubscript{hasindex(ss}index) then r{[}ss\textasciitilde index{]}
-=ss\textasciitilde item ss\textasciitilde next end
+\subsubsection{EXPOSED}\label{exposed}
 
-cs=ab\textasciitilde supplier
+\lstinputlisting[language=rexx,label=collectionexposedmethod.rexx,caption=collectionexposedmethod.rexx]{collectionexposedmethod.rexx}
 
-do while cs\textasciitilde available if
-\This\textsubscript{hasindex(cs}index) then
-r{[}cs\textasciitilde index{]} =cs\textasciitilde item es-next end
+\subsubsection{FINDINDEX}\label{findindex}
 
-return r
+\lstinputlisting[language=rexx,label=collectionfindindexmethod.rexx,caption=collectionfindindexmethod.rexx]{collectionfindindexmethod.rexx}
 
-::routine CommoniIntersect /* Returns a new collection of the same class
-as SELF that contains the items from SELF that have indexes also in the
-argument. \emph{/ /} Actually an index in SELF can only be `matched'
-with one in the argument once. Hence copy and removal. \emph{/ This =
-arg(1) /} self of caller */ w= .Bag\textasciitilde new
-sc=This\textasciitilde supplier do while sc\textasciitilde available
-wlsc\textasciitilde index{]} =sc\textasciitilde index sc-next end
-r=This\textsubscript{class}new
+\subsubsection{AT}\label{at}
 
-cs=MayEnBag (arg(2))\textasciitilde supplier do while
-cs\textasciitilde available i=cs\textasciitilde index if
-w\textasciitilde hasindex(i) then do r{[}i{]}=This {[}i{]}
-w\textasciitilde remove (i) end cs\textasciitilde next end return r
+\lstinputlisting[language=rexx,label=collectionatmethod.rexx,caption=collectionatmethod.rexx]{collectionatmethod.rexx}
 
-::routine CommonUnion /* Returns a new collection of the same class as
-SELF that contains all the items from SELF and items from the argument
-that have an index not in the first. \emph{/ /} Best to add them all. By
-adding non-receiver first we ensure that receiver takes priority when
-same indexes. \emph{/ This = arg(1) /} self of caller */
-r=This\textsubscript{class}new cs=MayEnBag
-(arg(2))\textasciitilde supplier do while cs\textasciitilde available
-r{[}cs\textasciitilde index{]} =cs\textasciitilde item
-cs\textasciitilde next end cs=This\textasciitilde supplier do while
-cs\textasciitilde available r{[}cs\textasciitilde index{]}
-=cs\textasciitilde item cs\textasciitilde next end return r
+\subsubsection{{[}{]}}\label{section}
 
-::routine CommonDifference /* Returns a new collection containing only
-those index-item pairs from the SELF whose indexes the other collection
-does not contain. \emph{/ This = arg(1) /} self of caller */
-r=This\textsubscript{class}new cs=This\textasciitilde supplier do while
-cs\textasciitilde available r{[}cs\textasciitilde index{]}
-=cs\textasciitilde item es-next end cs=MayEnBag
-(arg(2))\textasciitilde supplier do while cs\textasciitilde available
-r\textasciitilde remove (cs\textasciitilde index) es-next end return r
+\lstinputlisting[language=rexx,label=collectionatalt.rexx,caption=collectionatalt.rexx]{collectionatalt.rexx}
 
-::routine MayEnBag
+\subsubsection{PUT}\label{put}
 
-/* For List and Queue the indexes are dropped. */ rearg(1) if r-clags ==
-.List \textbar{} r-class == .Queue then r=EnBag(r) return r
+\lstinputlisting[language=rexx,label=collectionputmethod.rexx,caption=collectionputmethod.rexx]{collectionputmethod.rexx}
 
-::routine EnBag r=.Bag\textasciitilde new
-s=arg(1)\textasciitilde supplier do while s\textasciitilde available if
-arg(1)\textasciitilde class == .List \textbar{}
-arg(1)\textasciitilde class == .Queue then r{[}s\textasciitilde item{]}
-=s\textasciitilde item else /* This case is when the receiver is a Bag.
-*/ r{[}s\textasciitilde index{]} =s\textasciitilde index
-s\textasciitilde-next end return r
+\subsubsection{{[}{]}=}\label{section-1}
 
-\hypertarget{the-collection-class}{%
-\subsection{The collection class}\label{the-collection-class}}
+\lstinputlisting[language=rexx,label=collectionputalt.rexx,caption=collectionputalt.rexx]{collectionputalt.rexx}
 
-::Cclass `Collection'
+\subsubsection{HASINDEX}\label{hasindex}
 
-\hypertarget{init}{%
-\subsubsection{INIT}\label{init}}
+\lstinputlisting[language=rexx,label=collectionhasindexmethod.rexx,caption=collectionhasindexmethod.rexx]{collectionhasindexmethod.rexx}
 
-::method init
+\subsubsection{ITEMS}\label{items}
 
-expose a /* A collection is modelled as using 3 slots in an array for
-each element. The first slot holds the item, the second the index, and
-the third is used by particular types of collection. This order of slots
-is arbitary, chosen to match order of arguments for PUT and
-SUPPLIER\textasciitilde NEW. \emph{/ /} The first set of 3 slots is
-reserved for other purposes, to avoid having separate variables which
-the subclassing would need to access. */
+\lstinputlisting[language=rexx,label=collectionsitemsmethod.rexx,caption=collectionsitemsmethod.rexx]{collectionsitemsmethod.rexx}
 
-a=.array\textasciitilde new
+\subsubsection{REMOVE}\label{remove}
 
-a{[}1{]} /\emph{ItemsCount}/=0
+\lstinputlisting[language=rexx,label=collectionsremovemethod.rexx,caption=collectionsremovemethod.rexx]{collectionsremovemethod.rexx}
 
-a{[}2{]}/\emph{Unique}/=0
+\subsubsection{REMOVEIT}\label{removeit}
 
-return self
+\lstinputlisting[language=rexx,label=collectionsremoveitmethod.rexx,caption=collectionsremoveitmethod.rexx]{collectionsremoveitmethod.rexx}
 
-\hypertarget{exposed}{%
-\subsubsection{EXPOSED}\label{exposed}}
+\subsubsection{MAKEARRAY}\label{makearray}
 
-::method exposed private expose a
+\lstinputlisting[language=rexx,label=collectionsmakearraymethod.rexx,caption=collectionsmakearraymethod.rexx]{collectionsmakearraymethod.rexx}
 
-/* This method allows subclasses to get at the implementation of
-Collection. */ return a
+\subsubsection{MAKEARRAYX}\label{makearrayx}
 
-\hypertarget{findindex}{%
-\subsubsection{FINDINDEX}\label{findindex}}
+\lstinputlisting[language=rexx,label=collectionsmakearrayxmethod.rexx,caption=collectionsmakearrayxmethod.rexx]{collectionsmakearrayxmethod.rexx}
 
-::method findindex private expose a /* Returns array index if the
-collection contains any item associated with the index specified or
-returns 0 otherwise. \emph{/ do j=4 by 3 to
-1+3}a{[}1{]}/\emph{ItemsCount}/ if alj+l{]}==arg(1) then return j end j
-return 0
+\subsubsection{SUPPLIER}\label{supplier}
 
-\hypertarget{at}{%
-\subsubsection{AT}\label{at}}
+\lstinputlisting[language=rexx,label=collectionssuppliermethod.rexx,caption=collectionssuppliermethod.rexx]{collectionssuppliermethod.rexx}
 
-::method at /* vANY \emph{/ expose a /} Returns the item associated with
-the specified index. */ j=self\textasciitilde findindex(arg(1)) if j=0
-then return .nil return a{[}j{]}
+\subsection{Class list}\label{class-list}
 
-\hypertarget{section}{%
-\subsubsection{{[}{]}}\label{section}}
+\lstinputlisting[language=rexx,label=classlist.rexx,caption=classlist.rexx]{classlist.rexx}
 
-::method `{[}{]}' /* Synonym for the AT method. */ forward message `AT'
+\subsubsection{PUT}\label{put-1}
 
-\hypertarget{put}{%
-\subsubsection{PUT}\label{put}}
+\lstinputlisting[language=rexx,label=collectionsputmethod.rexx,caption=collectionsputmethod.rexx]{collectionsputmethod.rexx}
 
-::method put /* rANY rANy \emph{/ expose a use arg item, index /}
-Replaces any existing item associated with the specified index with the
-new item. Otherwise adds the item-index pair. \emph{/
-j=self\textasciitilde findindex (index) if j\textgreater0 then do
-alj{]}=item return end a{[}1{]} /}ItemsCount\emph{/=a {[}1{]}
-/}ItemsCount\emph{/+1 j=1+3}al{[}1{]} /\emph{ItemsCount}/ alj{]}=item
-a{[}lj+1{]} =index a{[}j+2{]}=0 return /* Error 91 in OOI if context
-requiring result. */
+\subsubsection{OF}\label{of}
 
-\hypertarget{section-1}{%
-\subsubsection{{[}{]}=}\label{section-1}}
+\lstinputlisting[language=rexx,label=collectionsofmethod.rexx,caption=collectionsofmethod.rexx]{collectionsofmethod.rexx}
 
-:smethod `{[}{]}=' /* Synonym for the PUT method. */ forward message
-`PUT'
+\subsubsection{INSERT}\label{insert}
 
-\hypertarget{hasindex}{%
-\subsubsection{HASINDEX}\label{hasindex}}
+\lstinputlisting[language=rexx,label=collectionsinsertmethod.rexx,caption=collectionsinsertmethod.rexx]{collectionsinsertmethod.rexx}
 
-::method hasindex /* vANY \emph{/ /} Returns 1 (true) if the collection
-contains any item associated with the index specified or returns 0
-(false) otherwise. */
+\subsubsection{FIRST}\label{first}
 
-return self\textasciitilde findindex (arg(1))\textgreater0
+\lstinputlisting[language=rexx,label=collectionsfirstmethod.rexx,caption=collectionsfirstmethod.rexx]{collectionsfirstmethod.rexx}
 
-\hypertarget{items}{%
-\subsubsection{ITEMS}\label{items}}
+\subsubsection{LAST}\label{last}
 
-::method items expose a
+\lstinputlisting[language=rexx,label=collectionslastmethod.rexx,caption=collectionslastmethod.rexx]{collectionslastmethod.rexx}
 
-/* Returns the number of items in the collection. \emph{/ return
-a{[}1{]}/}ItemsCount*/
+\subsubsection{FIRSTITEM}\label{firstitem}
 
-\hypertarget{remove}{%
-\subsubsection{REMOVE}\label{remove}}
+\lstinputlisting[language=rexx,label=collectionsfirstitemmethod.rexx,caption=collectionsfirstitemmethod.rexx]{collectionsfirstitemmethod.rexx}
 
-::method remove /* vANY */
+\subsubsection{LASTITEM}\label{lastitem}
 
-expose a /* Returns and removes from a collection the member item with
-the specified index. */
+\lstinputlisting[language=rexx,label=collectionslasttitemmethod.rexx,caption=collectionslasttitemmethod.rexx]{collectionslasttitemmethod.rexx}
 
-j=self\textasciitilde findindex(arg(1))
+\subsubsection{NEXT}\label{next}
 
-if j=0 then return .nil
+\lstinputlisting[language=rexx,label=collectionsnexttitemmethod.rexx,caption=collectionsnexttitemmethod.rexx]{collectionsnexttitemmethod.rexx}
 
-r=al{[}j{]}
+\subsubsection{PREVIOUS}\label{previous}
 
-self\textasciitilde removeit (j)
+\lstinputlisting[language=rexx,label=collectionsprevioustitemmethod.rexx,caption=collectionsprevioustitemmethod.rexx]{collectionsprevioustitemmethod.rexx}
 
-return r
+\subsubsection{SECTION}\label{section-2}
 
-\hypertarget{removeit}{%
-\subsubsection{REMOVEIT}\label{removeit}}
+\lstinputlisting[language=rexx ,label=collectionssectionmethod.rexx,caption=collectionssectionmethod.rexx]{collectionssectionmethod.rexx}
 
-::method removeit private
+\subsection{Class queue}\label{class-queue}
 
-expose a
+\lstinputlisting[language=rexx,label=collectionsqueueclass.rexx,caption=collectionsqueueclass.rexx]{collectionsqueueclass.rexx}
 
-use arg j
+\subsubsection{PUSH}\label{push}
 
-/* Remove relevant slots from the array, with compaction. */
+\lstinputlisting[language=rexx,label=queuepushmethod.rexx,caption=queuepushmethod.rexx]{queuepushmethod.rexx}
 
-do j=j+3 by 3 to 14+3\emph{a{[}1{]}/}ItemsCount*/
-alj-3l=saljl;alj-2{]}=al{[}j+1{]};alj-ll=alj+2{]} end j
+\subsubsection{PULL}\label{pull}
 
-a{[}1{]} /\emph{ItemsCount}/=a {[}1{]} /\emph{ItemsCount}/-1
+\lstinputlisting[language=rexx,label=collectionspullmethod.rexx,caption=collectionspullmethod.rexx]{collectionspullmethod.rexx}
 
-return
+\subsubsection{QUEUE}\label{queue}
 
-\hypertarget{makearrya}{%
-\subsubsection{MAKEARRYA}\label{makearrya}}
+\lstinputlisting[language=rexx,label=queuequeuemethod.rexx,caption=queuequeuemethod.rexx]{queuequeuemethod.rexx}
 
-::method makearray
+\subsubsection{PEEK}\label{peek}
 
-expose a /* Returns a single-index array containing the receiver list
-items. \emph{/ r= .array\textasciitilde new /} To build result in. */
+\lstinputlisting[language=rexx,label=queuepeekmethod.rexx,caption=queuepeekmethod.rexx]{queuepeekmethod.rexx}
 
-do j=4 by 3 to 1+3\emph{a{[}1{]}/}ItemsCount*/
-r{[}r\textasciitilde dimension(1)+1{]}=a{[}j{]} end j
+\subsubsection{REMOVE}\label{remove-1}
 
-return r
+\lstinputlisting[language=rexx,label=queueremovemethod.rexx,caption=queueremovemethod.rexx]{queueremovemethod.rexx}
 
-\hypertarget{makearrayx}{%
-\subsubsection{MAKEARRAYX}\label{makearrayx}}
+\subsection{Class table}\label{class-table}
 
-::method makearrayx private expose a /* Returns a single-index array
-containing the receiver index items. \emph{/ r=
-.array\textasciitilde new /} To build result in. \emph{/ do j=4 by 3 to
-1+3}a{[}1{]}/\emph{ItemsCount}/
-r{[}r\textasciitilde dimension(1)+1{]}=a{[}j+1{]} end j return r
+\lstinputlisting[language=rexx,label=tableclass.rexx,caption=tableclass.rexx]{tableclass.rexx}
 
-\hypertarget{supplier}{%
-\subsubsection{SUPPLIER}\label{supplier}}
+\subsubsection{MAKEARRAY}\label{makearray-1}
 
-::method supplier expose a
+\lstinputlisting[language=rexx,label=tablemakearray.rexx,caption=tablemakearray.rexx]{tablemakearray.rexx}
 
-/* Returns a supplier object for the list. */ return
-.supplier\textsubscript{new(self}makearray: .collection,
-self\textasciitilde makearrayx)
+\subsubsection{UNION}\label{union}
 
-\hypertarget{class-list}{%
-\subsection{Class list}\label{class-list}}
+\lstinputlisting[language=rexx,label=tableunionmethod.rexx,caption=tableunionmethod.rexx]{tableunionmethod.rexx}
 
-::¢class `List' subclass Collection
+\subsubsection{INTERSECTION}\label{intersection}
 
-\hypertarget{put-1}{%
-\subsubsection{PUT}\label{put-1}}
+\lstinputlisting[language=rexx,label=tableintersectionmethod.rexx,caption=tableintersectionmethod.rexx]{tableintersectionmethod.rexx}
 
-::method put /* vANY rANY */ use arg item, index
-a=self\textasciitilde exposed
+\subsubsection{XOR}\label{xor}
 
-/* PUT for a List must not be an insertion. */
-j=self\textasciitilde findindex (index) if j=0 then call Raise
-`Syntax',93.918 alj{]}=item return
+\lstinputlisting[language=rexx,label=tablexormethod.rexx,caption=tablexormethod.rexx]{tablexormethod.rexx}
 
-\hypertarget{of}{%
-\subsubsection{OF}\label{of}}
+\subsubsection{DIFFERENCE}\label{difference}
 
-::method of class /* 1 or more oANY Are they omittable? Not in I00
-\emph{/ /} Returns a newly created list containing the specified value
-objects in the order specified. */ r= self \textasciitilde{} new do j =
-1 to arg() r\textasciitilde{} insert (arg(j)) end j return r
+\lstinputlisting[language=rexx,label=tabledifferencemethod.rexx,caption=tabledifferencemethod.rexx]{tabledifferencemethod.rexx}
 
-\hypertarget{insert}{%
-\subsubsection{INSERT}\label{insert}}
+\subsubsection{SUBSET}\label{subset}
 
-::method insert /* rANY oANY */
+\lstinputlisting[language=rexx,label=tablesubsetmethod.rexx,caption=tablesubsetmethod.rexx]{tablesubsetmethod.rexx}
 
-use arg item, index
+\subsubsection{Class set}\label{class-set}
 
-a=self\textasciitilde exposed /* Returns a list-supplied index for a new
-item, of specified value, which is added to the list. The new item
-follows the existing item with the specified index in the list ordering.
-\emph{/ /} Establish the index of what preceeds the new element. \emph{/
-/} If there was no index given, the new item becomes the last on list.
-\emph{/ /} .nil argument means first */
+\lstinputlisting[language=rexx,label=classset.rexx,caption=classset.rexx]{classset.rexx}
 
-if arg(2,`E') then p=arg(2)
+\subsubsection{PUT}\label{put-2}
 
-else p=self\textasciitilde last
+\lstinputlisting[language=rexx,label=setputmethod.rexx,caption=setputmethod.rexx]{setputmethod.rexx}
 
-/* Convert from list index to underlying array index. */
+\subsubsection{OF}\label{of-1}
 
-if p==.nil then j=1
+\lstinputlisting[language=rexx,label=setofmethod.rexx,caption=setofmethod.rexx]{setofmethod.rexx}
 
-else j=self\textasciitilde findindex(p)
+\subsubsection{UNION}\label{union-1}
 
-if j=0 then call Raise `Syntax',93.918
+\lstinputlisting[language=rexx,label=setunion.rexx,caption=setunion.rexx]{setunion.rexx}
 
-j=j+3 /* Where new entry will be. \emph{/ /} Move space to required
-place. */
+\subsubsection{INTERSECTION}\label{intersection-1}
 
-a{[}1{]} /\emph{ItemsCount}/=a {[}1{]} /\emph{ItemsCount}/+1
+\lstinputlisting[language=rexx,label=setintersectionmethod.rexx,caption=setintersectionmethod.rexx]{setintersectionmethod.rexx}
 
-do k=1+3\emph{a{[}1{]}/}ItemsCount*/ by -3 to j+3
+\subsubsection{XOR}\label{xor-1}
 
-a{[}k{]} =a{[}k-3{]} ;a{[}k+1{]} =a{[}k-2{]} ;a{[}k{]} =a{[}k-3{]} end
+\lstinputlisting[language=rexx,label=setxormethod.rexx,caption=setxormethod.rexx]{setxormethod.rexx}
 
-/* Insert new element. */
+\subsubsection{DIFFERENCE}\label{difference-1}
 
-alj{]}=item /* A new, unique, index is needed. \emph{/ /} The basic
-requirement is for something unique, so this would be correct:
+\lstinputlisting[language=rexx,label=setdifferencemethod.rexx,caption=setdifferencemethod.rexx]{setdifferencemethod.rexx}
 
-is.object\textasciitilde new /* a unique object, used as a key (the
-index on the list) \emph{/ }
+\subsection{Class relation}\label{class-relation}
 
-/ /* However, a number can be used. (At risk of the user thinking it is
-sensible to do arithmetic on it.) */
+\lstinputlisting[language=rexx,label=relationclass.rexx,caption=relationclass.rexx]{relationclass.rexx}
 
-a{[}j+1{]}=a{[}2{]}/\emph{Unique}/;a{[}2{]} /\emph{Unique}/=a{[}2{]}
-/\emph{Unique}/+1
+\subsubsection{PUT}\label{put-3}
 
-a{[}j+2{]}=0
+\lstinputlisting[language=rexx,label=relputmethod.rexx,caption=relputmethod.rexx]{relputmethod.rexx}
 
-return a{[}j+1{]}
+\subsubsection{ITEMS}\label{items-1}
 
-\hypertarget{first}{%
-\subsubsection{FIRST}\label{first}}
+\lstinputlisting[language=rexx,label=relitemsmethod.rexx,caption=relitemsmethod.rexx]{relitemsmethod.rexx}
 
-::method first a=self\textasciitilde exposed
+\subsubsection{MAKEARRAY}\label{makearray-2}
 
-/* Returns the index of the first item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount*/=0 then return .nil
+\lstinputlisting[language=rexx,label=relmakearray method,caption=relmakearray method]{relmakearray method}
 
-141 return a{[}5{]}
+\subsubsection{SUPPLIER}\label{supplier-1}
 
-\hypertarget{last}{%
-\subsubsection{LAST}\label{last}}
+\lstinputlisting[language=rexx,label=relsuppliermethod.rexx,caption=relsuppliermethod.rexx]{relsuppliermethod.rexx}
 
-::smethod last a=self\textasciitilde exposed
+\subsubsection{UNION}\label{union-2}
 
-/* Returns the index of the last item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount*/=0 then return .nil return
-a{[}3\emph{a{[}1{]}/}ItemsCount*/+2{]}
+\lstinputlisting[language=rexx,label=relunionmethod.rexx,caption=relunionmethod.rexx]{relunionmethod.rexx}
 
-\hypertarget{firstitem}{%
-\subsubsection{FIRSTITEM}\label{firstitem}}
+\subsubsection{INTERSECTION}\label{intersection-2}
 
-::method firstitem a=self\textasciitilde exposed
+\lstinputlisting[language=rexx,label=relintersectionmethod.rexx,caption=relintersectionmethod.rexx]{relintersectionmethod.rexx}
 
-/* Returns the first item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount*/=0 then return .nil return a{[}4{]}
+\subsubsection{XOR}\label{xor-2}
 
-\hypertarget{lastitem}{%
-\subsubsection{LASTITEM}\label{lastitem}}
+\lstinputlisting[language=rexx,label=relxormethod.rexx,caption=relxormethod.rexx]{relxormethod.rexx}
 
-::method lastitem a=self\textasciitilde exposed
+\subsubsection{DIFFERENCE}\label{difference-2}
 
-/* Returns the last item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount*/=0 then return .nil return
-a{[}3\emph{a{[}1{]}/}ItemsCount*/+1{]}
+\lstinputlisting[language=rexx,label=reldifferencemethod.rexx,caption=reldifferencemethod.rexx]{reldifferencemethod.rexx}
 
-\hypertarget{next}{%
-\subsubsection{NEXT}\label{next}}
+\subsubsection{SUBSET}\label{subset-1}
 
-::method next /* vANY \emph{/ a=self\textasciitilde exposed /} Returns
-the index of the item that follows the list item having the specified
-index. \emph{/ j=self\textasciitilde findindex(arg(1)) if j=0 then call
-Raise `Syntax',93.918 j=j+3 if
-j\textgreater3}a{[}1{]}/\emph{ItemsCount}/ then return .nil /* Next of
-last was requested. */ return a{[}j+1{]}
+\lstinputlisting[language=rexx,label=relsubsetmethod.rexx,caption=relsubsetmethod.rexx]{relsubsetmethod.rexx}
 
-\hypertarget{previous}{%
-\subsubsection{PREVIOUS}\label{previous}}
+\subsubsection{REMOVEITEM}\label{removeitem}
 
-::method previous /* vANY */
+\lstinputlisting[language=rexx,label=relremoveitemmethod.rexx,caption=relremoveitemmethod.rexx]{relremoveitemmethod.rexx}
 
-a=self\textasciitilde exposed /* Returns the index of the item that
-precedes the list item having the specified index. */
+\subsubsection{INDEX}\label{index}
 
-j=self\textasciitilde findindex(arg(1))
+\lstinputlisting[language=rexx,label=relindexmethod.rexx,caption=relindexmethod.rexx]{relindexmethod.rexx}
 
-if j=0 then call Raise `Syntax',93.918
+\subsubsection{ALLAT}\label{allat}
 
-j=j-3
+\lstinputlisting[language=rexx,label=relAllAtmethod.rexx,caption=relAllAtmethod.rexx]{relAllAtmethod.rexx}
 
-if j\textless4 then return .nil /* Previous of first was requested. */
+\subsubsection{HASITEM}\label{hasitem}
 
-return a{[}j+1{]}
+\lstinputlisting[language=rexx,label=relhasitemmethod.rexx,caption=relhasitemmethod.rexx]{relhasitemmethod.rexx}
 
-\hypertarget{section-2}{%
-\subsubsection{SECTION}\label{section-2}}
+\subsubsection{ALLINDEX}\label{allindex}
 
-::method section /* rANY oWHOLE\textgreater=0 */
+\lstinputlisting[language=rexx,label=relallindexmethod.rexx,caption=relallindexmethod.rexx]{relallindexmethod.rexx}
 
-a=self\textasciitilde exposed /* Returns a new list containing selected
-items from the receiver list. The first item in the new list is the item
-corresponding to the index specified, in the receiver list. */
+\lstinputlisting[language=rexx,label=bagclass.rexx,caption=bagclass.rexx]{bagclass.rexx}
 
-j=self\textasciitilde findindex(arg(1))
+\subsubsection{OF}\label{of-2}
 
-if j=0 then call Raise `Syntax',93.918
+\lstinputlisting[language=rexx,label=bagofmethod.rexx,caption=bagofmethod.rexx]{bagofmethod.rexx}
 
-r= .list\textasciitilde new /* To build result in. \emph{/ if arg(2,`E')
-then s = arg(2) else s = self\textasciitilde items; do s
-r\textasciitilde insert (a{[}j{]}) j=j+3 if
-j\textgreater1+3}a{[}1{]}/\emph{ItemsCount}/ then leave end return r
+\subsubsection{PUT}\label{put-4}
 
-\hypertarget{class-queue}{%
-\subsection{Class queue}\label{class-queue}}
+\lstinputlisting[language=rexx,label=bagputmethod.rexx,caption=bagputmethod.rexx]{bagputmethod.rexx}
 
-::class `Queue' subclass Collection
+\subsubsection{UNION}\label{union-3}
 
-/* A queue is a sequenced collection with whole-number indexes. The
+\lstinputlisting[language=rexx,label=bagunionmethod.rexx,caption=bagunionmethod.rexx]{bagunionmethod.rexx}
 
-indexes specify the position of an item relative to the head (first
-item) of the queue. Adding or removing an item changes the association
-of an index to its queue item. */
+\subsubsection{INTERSECTION}\label{intersection-3}
 
-\hypertarget{push}{%
-\subsubsection{PUSH}\label{push}}
+\lstinputlisting[language=rexx,label=bagintersectionmethod.rexx,caption=bagintersectionmethod.rexx]{bagintersectionmethod.rexx}
 
-::method push /* vANY */
+\subsubsection{XOR}\label{xor-3}
 
-/* Adds the object value to the queue at its head. \emph{/
-a=self\textasciitilde exposed a{[}1{]} /}ItemCount\emph{/=a {[}1{]}
-/}ItemCount*/+1
+\lstinputlisting[language=rexx,label=bagxormethod.rexx,caption=bagxormethod.rexx]{bagxormethod.rexx}
 
-/* Slide along to make a space. \emph{/ do
-j=1+3}a{[}1{]}/\emph{ItemCount}/ by -3 to 7
+\subsubsection{DIFFERENCE}\label{difference-3}
 
-aljl=alj-3{]} alj+l{]}=a{[}j-2{]}+1; /* Index changes \emph{/ end j
-a{[}4{]}=arg(1) a{[}5J=1 return 12.2.4.2 PULL{]}=a{[}j+4{]}-1; /} Index
-changes */ end j return r
+\lstinputlisting[language=rexx,label=bagdifferencemethod.rexx,caption=bagdifferencemethod.rexx]{bagdifferencemethod.rexx}
 
-\hypertarget{queue}{%
-\subsubsection{QUEUE}\label{queue}}
+\subsection{The directory class}\label{the-directory-class}
 
-::method queue /* vANY \emph{/ /} Adds the object value to the queue at
-its tail. \emph{/ a=self\textasciitilde exposed a{[}1{]}
-/}ItemCount\emph{/=a {[}1{]} /}ItemCount\emph{/+1 a{[}1+3\emph{a{[}1{]}
-/}ItemCount*/{]}=arg(1) a{[}2+3\emph{a{[}1{]}
-/}ItemCount*/{]}=a{[}1{]}/}ItemCount*/ return
+\lstinputlisting[language=rexx,label=directoryclass.rexx,caption=directoryclass.rexx]{directoryclass.rexx}
 
-\hypertarget{peek}{%
-\subsubsection{PEEK}\label{peek}}
+\subsubsection{AT}\label{at-1}
 
-::method peek
+\lstinputlisting[language=rexx,label=diratmethod.rexx,caption=diratmethod.rexx]{diratmethod.rexx}
 
-a=self\textasciitilde exposed /* Returns the item at the head of the
-queue. The collection remains unchanged. */
+\subsubsection{PUT}\label{put-5}
 
-return a{[}4{]}
+\lstinputlisting[language=rexx,label=dirputmethod.rexx,caption=dirputmethod.rexx]{dirputmethod.rexx}
 
-\hypertarget{remove-1}{%
-\subsubsection{REMOVE}\label{remove-1}}
+\subsubsection{MAKEARRAY}\label{makearray-3}
 
-::method remove /* rcWHOLE\textgreater O \emph{/ /} Returns and removes
-from a collection the member item with the specified index. \emph{/
-a=self\textasciitilde exposed if
-a{[}1{]}/}ItemCount\emph{/\textless arg(1) then return .nil
-r=self\textasciitilde remove: super (arg(1)) /} Reset the indexes.
-\emph{/ k=0 do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ k=k+1
-alj+l{]}=k end j return r
+\lstinputlisting[language=rexx,label=diremakearray.rexx,caption=diremakearray.rexx]{diremakearray.rexx}
 
-\hypertarget{class-table}{%
-\subsection{Class table}\label{class-table}}
+\subsubsection{SUPPLIER}\label{supplier-2}
 
-::Class `Table' subclass Collection
+\lstinputlisting[language=rexx,label=dirsuppliermethod.rexx,caption=dirsuppliermethod.rexx]{dirsuppliermethod.rexx}
 
-\hypertarget{makearray}{%
-\subsubsection{MAKEARRAY}\label{makearray}}
+\subsubsection{UNION}\label{union-4}
 
-::method makearray /* Returns a single-index array containing the index
-objects. \emph{/ /} This is different from Collection MAKEARRAY where
-items rather than indexes are in the returned array. */ forward message
-`MAKEARRAYX'
+\lstinputlisting[language=rexx,label=dirunionmethod.rexx,caption=dirunionmethod.rexx]{dirunionmethod.rexx}
 
-\hypertarget{union}{%
-\subsubsection{UNION}\label{union}}
+\subsubsection{INTERSECTION}\label{intersection-4}
 
-::method union /* rCOLLECTION */ return CommonUnion (self,arg(1))
+\lstinputlisting[language=rexx,label=dirintersectionmethod.rexx,caption=dirintersectionmethod.rexx]{dirintersectionmethod.rexx}
 
-\hypertarget{intersection}{%
-\subsubsection{INTERSECTION}\label{intersection}}
+\subsubsection{XOR}\label{xor-4}
 
-::method intersection /* rCOLLECTION */ return CommoniIntersect
-(self,arg(1))
+\lstinputlisting[language=rexx,label=dirxormethod.rexx,caption=dirxormethod.rexx]{dirxormethod.rexx}
 
-\hypertarget{xor}{%
-\subsubsection{XOR}\label{xor}}
+\subsubsection{DIFFERENCE}\label{difference-4}
 
-::method xor /* rCOLLECTION */ return CommonXor (self,arg(1))
+\lstinputlisting[language=rexx,label=dirdifferencemethod.rexx,caption=dirdifferencemethod.rexx]{dirdifferencemethod.rexx}
 
-\hypertarget{difference}{%
-\subsubsection{DIFFERENCE}\label{difference}}
+\subsubsection{SUBSET}\label{subset-2}
 
-::method difference /* rCOLLECTION */ return
-CommonDifference(self,arg(1))
+\lstinputlisting[language=rexx,label=dirsubsetmethod.rexx,caption=dirsubsetmethod.rexx]{dirsubsetmethod.rexx}
 
-\hypertarget{subset}{%
-\subsubsection{SUBSET}\label{subset}}
+\subsubsection{SETENTRY}\label{setentry}
 
-::method subset /* rCOLLECTION */ return
-self\textsubscript{difference(arg(1))}items = 0
+\lstinputlisting[language=rexx,label=dirsetentrymethod.rexx,caption=dirsetentrymethod.rexx]{dirsetentrymethod.rexx}
 
-\hypertarget{class-set}{%
-\subsubsection{Class set}\label{class-set}}
+\subsubsection{ENTRY}\label{entry}
 
-::¢class `Set' subclass table
+\lstinputlisting[language=rexx,label=direntrymethod.rexx,caption=direntrymethod.rexx]{direntrymethod.rexx}
 
-/* A set is a collection that restricts the member items to have a value
-that is the same as the index. Any object can be placed in a set. There
-can be only
+\subsubsection{HASENTRY}\label{hasentry}
 
-one occurrence of any object in a set. */
+\lstinputlisting[language=rexx,label=dirhasentrymethod.rexx,caption=dirhasentrymethod.rexx]{dirhasentrymethod.rexx}
 
-\hypertarget{put-2}{%
-\subsubsection{PUT}\label{put-2}}
+\subsubsection{SETMETHOD}\label{setmethod}
 
-/* Second arg same as first. Committee has dropped second? \emph{/
-::method put /} YANY oANY \emph{/ /} Makes the object value a member
-item of the collection and associates it with specified index. \emph{/
-if arg(2,`E') then if arg(2)==arg(1) then signal error /} 949 */
-self\textasciitilde put:super(arg(1),arg(1))
+\lstinputlisting[language=rexx,label=dirsetmethod.rexxx,caption=dirsetmethod.rexxx]{dirsetmethod.rexxx}
 
-\hypertarget{of-1}{%
-\subsubsection{OF}\label{of-1}}
+\subsubsection{UNKNOWN}\label{unknown}
 
-::method of class /* 1 or more rANYy \emph{/ /} Returns a newly created
-set containing the specified value objects. */ r=self\textasciitilde new
-do j=1 to arg() r\textasciitilde put (arg(j)) end j return r
+\lstinputlisting[language=rexx,label=dirunknownmethod.rexx,caption=dirunknownmethod.rexx]{dirunknownmethod.rexx}
 
-\hypertarget{union-1}{%
-\subsubsection{UNION}\label{union-1}}
+\subsection{The stem class}\label{the-stem-class}
 
-::method union /* rCOLLECTION */ return CommonUnion (self,
-EnBag(arg(1)))
+\emph{For some reason, the stem class doesn't have PUT and AT methods,
+which stops us having a general rule about {[}{]} synonyms AT, {[}{]}=
+synonyms PUT.}
 
-\hypertarget{intersection-1}{%
-\subsubsection{INTERSECTION}\label{intersection-1}}
+\emph{Anyway, committee doing without this class as such.}
 
-::method intersection /* rCOLLECTION */ return CommoniIntersect
-(self,EnBag(arg(1)))
+\emph{Here is temporary stuff showing how to use algebra in the
+collection coding.}
 
-\hypertarget{xor-1}{%
-\subsubsection{XOR}\label{xor-1}}
+\lstinputlisting[language=rexx,label=stemclass.rexx,caption=stemclass.rexx]{stemclass.rexx}
 
-::method xor /* rCOLLECTION */ return CommonXor (self, EnBag(arg(1)))
+{[}The following code looks like a partial duplicate of the previous.
+Formatting pending -- JMB{]}
 
-\hypertarget{difference-1}{%
-\subsubsection{DIFFERENCE}\label{difference-1}}
+\lstinputlisting[language=rexx,label=setlikeoperations.rexx,caption=setlikeoperations.rexx]{setlikeoperations.rexx}
 
-::method difference /* rCOLLECTION */ return CommonDifference (self,
-EnBag(arg(1)))
+\section{The stream class}\label{the-stream-class}
 
-\hypertarget{class-relation}{%
-\subsection{Class relation}\label{class-relation}}
+The stream class provides input/output on external streams.
 
-::Class `Relation' subclass Collection
+\lstinputlisting[language=rexx,label=streamclass.rexx,caption=streamclass.rexx]{streamclass.rexx}
 
-\hypertarget{put-3}{%
-\subsubsection{PUT}\label{put-3}}
+Initializes a stream object for a stream named name, but does not open
+the stream.
 
-::method put /* vANY rANY */
+\lstinputlisting[language=rexx,label=streamquerymethod.rexx,caption=streamquerymethod.rexx]{streamquerymethod.rexx}
 
-use arg item, index
-
-a=self\textasciitilde exposed /* Makes the object value a member item of
-the relation and associates it with the specified index. If the relation
-already contains any items with the specified index, this method adds a
-new member item value with the same index, without removing any existing
-members */
-
-a{[}1{]} /\emph{ItemsCount}/=a {[}1{]} /\emph{ItemsCount}/+1
-
-j=1+3\emph{al{[}1{]} /}ItemsCount*/
-
-alj{]}=item
-
-a{[}lj+1{]} =index
-
-a{[}j+2{]}=0
-
-return /* Error 91 in OOI if context requiring result. */
-
-\hypertarget{items-1}{%
-\subsubsection{ITEMS}\label{items-1}}
-
-::method items /* oANY \emph{/ a=self\textasciitilde exposed /} Returns
-the number of relation items with the specified index. If you specify no
-index, this method returns the total number of items associated with all
-indexes in the relation. \emph{/ if \arg(1,`E') then return
-a{[}1{]}/}ItemsCount\emph{/ n=0 do j=4 by 3 to
-1+3}a{[}1{]}/\emph{ItemsCount}/ if arg(1)==a{[}j+1{]} then n=n+1 end j
-return n
-
-\hypertarget{makearray-1}{%
-\subsubsection{MAKEARRAY}\label{makearray-1}}
-
-::method makearray forward message `MAKEARRAYX'
-
-\hypertarget{supplier-1}{%
-\subsubsection{SUPPLIER}\label{supplier-1}}
-
-::method supplier /* oANY \emph{/ a=self\textasciitilde exposed /}
-Returns a supplier object for the collection. If an index is specified,
-the supplier enumerates all of the items in the relation with the
-specified index. \emph{/ m=.array\textasciitilde new /} For the items
-\emph{/ r=.array\textasciitilde new /} For the indexes \emph{/ do j=4 by
-3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if arg(1,`E') then if
-arg(1)=s=a{[}j+1{]} then iterate n=r\textasciitilde dimension(1)+1
-
-m{[}n{]} =a{[}j{]} r{[}n{]} =a{[}j+1{]}
-
-end j return .supplier\textasciitilde new(m,r) 12.2.7.5 UNION ::method
-union /* rCOLLECTION */
-
-/* Union for a relation is just all of both. */
-r=self\textsubscript{class}new cs=self\textasciitilde supplier do while
-cs\textasciitilde available
-
-r{[}cs\textasciitilde index{]} =cs\textasciitilde item es-next end
-cs=MayEnBag (arg(1))\textasciitilde supplier do while
-cs\textasciitilde available r{[}cs\textasciitilde index{]}
-=cs\textasciitilde item es-next end return r
-
-\hypertarget{intersection-2}{%
-\subsubsection{INTERSECTION}\label{intersection-2}}
-
-::method intersection /* rCOLLECTION */
-
-/* Intersection for a relation requires the items as well as the keys to
-match. */
-
-r=self\textsubscript{class}new sc=self\textsubscript{class}new
-cs=self\textasciitilde supplier do while cs\textasciitilde available sc
-{[}cs\textasciitilde index{]} =cs\textasciitilde item
-cs\textasciitilde next end cs=MayEnBag (arg(1))\textasciitilde supplier
-do while cs\textasciitilde available if
-sc\textsubscript{hasitem(cs}item,cs\textasciitilde index) then r
-{[}ecs\textasciitilde index{]} =sc\textsubscript{removeitem(cs}item,
-cs\textasciitilde next end return r
-
-cs\textasciitilde index)
-
-\hypertarget{xor-2}{%
-\subsubsection{XOR}\label{xor-2}}
-
-::method xor /* rCOLLECTION \emph{/ /} Returns a new relation that
-contains all items from self and
-
-the argument except that all index-item pairs that appear in both
-collections are removed. */
-
-r=self\textsubscript{class}new cs=self\textasciitilde supplier do while
-cs\textasciitilde available r{[}cs\textasciitilde index{]}
-=cs\textasciitilde item cs\textasciitilde next end cs=MayEnBag
-(arg(1))\textasciitilde supplier do while cs\textasciitilde available if
-self\textsubscript{hasitem(cs}item,cs\textasciitilde index) then
-r\textsubscript{removeitem(cs}item, cs\textasciitilde index) else
-r{[}cs\textasciitilde index{]} =cs\textasciitilde item
-cs\textasciitilde next end return r
-
-\hypertarget{difference-2}{%
-\subsubsection{DIFFERENCE}\label{difference-2}}
-
-::method difference /* rCOLLECTION */
-
-/* Returns a new relation containing only those index-item pairs from
-the g Y P SELF whose indexes the other collection does not contain. */
-r=self\textsubscript{class}new cs=self\textasciitilde supplier
-
-do while cs\textasciitilde available r{[}cs\textasciitilde index{]}
-=cs\textasciitilde item cs\textasciitilde next end
-
-cs=MayEnBag (arg(1))\textasciitilde supplier
-
-do while cs\textasciitilde available r\textsubscript{removeitem(cs}item,
-cs\textasciitilde index) cs\textasciitilde next end
-
-return r
-
-\hypertarget{subset-1}{%
-\subsubsection{SUBSET}\label{subset-1}}
-
-::method subset /* rCOLLECTION */ return
-self\textsubscript{difference(arg(1))}items = 0
-
-\hypertarget{removeitem}{%
-\subsubsection{REMOVEITEM}\label{removeitem}}
-
-::method removeitem /* YANY rANY \emph{/ a=self\textasciitilde exposed
-/} Returns and removes from a relation the member item value (associated
-with the specified index). If value is not a member item associated with
-index index, this method returns the NIL object and removes no item.
-\emph{/ do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if alj{]}==arg(1)
-\& a{[}j+1l{]}==arg(2) then do self\textasciitilde removeit (j) return
-arg(1) end end j return .nil
-
-\hypertarget{index}{%
-\subsubsection{INDEX}\label{index}}
-
-::method index /* vANY \emph{/ a=self\textasciitilde exposed /} Returns
-the index for the specified item. If there is more than one index
-associated with the specified item, the one this method returns is not
-defined. \emph{/ do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if
-arg(1)==a{[}j{]} then return a{[}j+1{]} end j return .nil
-
-\hypertarget{allat}{%
-\subsubsection{ALLAT}\label{allat}}
-
-::method allat /* vANY \emph{/ a=self\textasciitilde exposed /} Returns
-a single-index array containing all the items associated with the
-specified index. \emph{/ r=.array\textasciitilde new do j=4 by 3 to
-1+3}a{[}1{]}/\emph{ItemsCount}/ if arg(1)==a{[}j+1{]} then
-r{[}r\textasciitilde dimension(1)+1{]} = aljl end j return r
-
-\hypertarget{hasitem}{%
-\subsubsection{HASITEM}\label{hasitem}}
-
-::method hasitem /* YANY rANY \emph{/ a=self\textasciitilde exposed /}
-Returns 1 (true) if the relation contains the member item value
-(associated with specified index). Returns 0 (false) otherwise. \emph{/
-do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if alj{]}==arg(1) \&
-a{[}j+l{]}==arg(2) then return 1 end j return 0
-
-\hypertarget{allindex}{%
-\subsubsection{ALLINDEX}\label{allindex}}
-
-::method allindex /* vANY */
-
-a=self\textasciitilde exposed /* Returns a single-index array containing
-all indexes for the specified item. \emph{/ r=.array\textasciitilde new
-do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/ if alj{]}==arg(1) then do
-r{[}r-\textasciitilde dimension (1)+1{]} =a{[}j+1{]} end end j return r
-
-\#\#3 The bag class
-
-::class `Bag' subclass relation
-
-/* A bag is a collection that restricts the member items to having a
-value that is the same as the index. Any object can be placed in a bag,
-and the same object can be placed in a bag multiple times. */
-
-\hypertarget{of-2}{%
-\subsubsection{OF}\label{of-2}}
-
-::method of class /* 1 or more rANYy \emph{/ /} Returns a newly created
-bag containing the specified value objects. */ r=self\textasciitilde new
-do j=1 to arg() r\textasciitilde put (arg(j)) end j return r
-
-\hypertarget{put-4}{%
-\subsubsection{PUT}\label{put-4}}
-
-::method put /* vANY oANY \emph{/ /} Committee does away with second
-argument? \emph{/ /} Makes the object value a member item of the
-collection and associates it with the specified index. If you specify
-index, it must be the same as value. */ if arg(2,`E') then if
-arg(2)=s=arg(1) then signal error self\textasciitilde put: super
-(arg(1),arg(1))
-
-\hypertarget{union-2}{%
-\subsubsection{UNION}\label{union-2}}
-
-::method union /* rCOLLECTION */ return CommonUnion (self,
-EnBag(arg(1)))
-
-\hypertarget{intersection-3}{%
-\subsubsection{INTERSECTION}\label{intersection-3}}
-
-::method intersection /* rCOLLECTION */ return CommoniIntersect
-(self,EnBag(arg(1)))
-
-\hypertarget{xor-3}{%
-\subsubsection{XOR}\label{xor-3}}
-
-::method xor /* rCOLLECTION */ return CommonXor (self, EnBag(arg(1)))
-
-\hypertarget{difference-3}{%
-\subsubsection{DIFFERENCE}\label{difference-3}}
-
-::method difference /* rCOLLECTION */ return CommonDifference (self,
-EnBag(arg(1)))
-
-\hypertarget{the-directory-class}{%
-\subsection{The directory class}\label{the-directory-class}}
-
-::class `Directory' subclass Collection
-
-\hypertarget{at-1}{%
-\subsubsection{AT}\label{at-1}}
-
-::method at /* vANY */ a=self\textasciitilde exposed
-
-/* Returns the item associated with the specified index. */
-j=self\textasciitilde findindex(arg(1))
-
-if j=0 then return .nil
-
-/* Run the method if there is one. */ if a{[}j+2{]} then return
-self\textasciitilde run(a{[}j{]}) return a{[}j{]}
-
-\hypertarget{put-5}{%
-\subsubsection{PUT}\label{put-5}}
-
-::method put /* vANY rANY \emph{/ a=self\textasciitilde exposed /} Makes
-the object value a member item of the collection and associates it with
-the specified index. */ if
-\arg(2)\textasciitilde hasmethod(`MAKESTRING') then call Raise `Syntax',
-93.938 self\textasciitilde put:super(arg(1),arg(2)
-\textasciitilde makestring) return
-
-\hypertarget{makearray-2}{%
-\subsubsection{MAKEARRAY}\label{makearray-2}}
-
-::method makearray forward message `MAKEARRAYX'
-
-\hypertarget{supplier-2}{%
-\subsubsection{SUPPLIER}\label{supplier-2}}
-
-::method supplier a=self\textasciitilde exposed /* Returns a supplier
-object for the directory. \emph{/ /} Check out what happens to the
-SETENTRY fields. \emph{/ r=.array\textasciitilde new /} For items
-\emph{/ do j=4 by 3 to 1+3}a{[}1{]}/\emph{ItemsCount}/
-r{[}r\textasciitilde dimension(1)+1{]}=a{[}j{]}
-
-end j return .supplier\textsubscript{new(r,self}makearray) \#\#\#\#
-UNION ::method union /* rCOLLECTION */
-
-return CommonUnion (self,arg(1))
-
-\hypertarget{intersection-4}{%
-\subsubsection{INTERSECTION}\label{intersection-4}}
-
-::method intersection /* rCOLLECTION */ return CommoniIntersect
-(self,arg(1))
-
-\hypertarget{xor-4}{%
-\subsubsection{XOR}\label{xor-4}}
-
-::method xor /* rCOLLECTION */ return CommonXor (self,arg(1))
-
-\hypertarget{difference-4}{%
-\subsubsection{DIFFERENCE}\label{difference-4}}
-
-::method difference /* rCOLLECTION */ return
-CommonDifference(self,arg(1))
-
-\hypertarget{subset-2}{%
-\subsubsection{SUBSET}\label{subset-2}}
-
-::method subset /* rCOLLECTION */ return
-self\textsubscript{difference(arg(1))}items = 0
-
-\hypertarget{setentry}{%
-\subsubsection{SETENTRY}\label{setentry}}
-
-::smethod setentry /* rSTRING oANY */
-
-a=self\textasciitilde exposed /* Sets the directory entry with the
-specified name (translated to uppercase) to the second argument,
-replacing any existing entry or method for the specified name. */
-
-n=translate(arg(1))
-
-j=self\textasciitilde findindex(n)
-
-if j=0 \& \arg(2,`E') then return
-
-if \arg(2,`E') then do /* Removal */
-
-self\textasciitilde removeit (j)
-
-return end
-
-if j=0 then do /* It's new \emph{/
-a{[}1{]}/}ItemsCount\emph{/=a{[}1{]}/}ItemsCount\emph{/ +1
-j=1+3}al{[}1{]} /\emph{ItemsCount}/ alj+l{]}=n end
-
-alj{]}=arg(2)
-
-a{[}4j+2{]}=0
-
-return
-
-\hypertarget{entry}{%
-\subsubsection{ENTRY}\label{entry}}
-
-::smethod entry /* rSTRING \emph{/ a=self\textasciitilde exposed /}
-Returns the directory entry with the specified name (translated to
-uppercase). \emph{/ n=translate(arg(1))
-j=self\textasciitilde findindex(n) /}if j=0 then signal error according
-to online \emph{/ /} Online has something about running UNKNOWN. \emph{/
-if j=0 then return .nil /} If there is an entry decide whether to invoke
-it. */ if a\textasciitilde hasindex(j) then do if \al[j+2] then return
-al{[}j{]} return self\textasciitilde run(al{[}jl{]}) end
-
-\hypertarget{hasentry}{%
-\subsubsection{HASENTRY}\label{hasentry}}
-
-::method hasentry /* rSTRING \emph{/ /} Returns 1 (true) if the
-directory has an entry or a method for the specified name (translated to
-uppercase) or 0 (false) otherwise. */
-
-return self\textasciitilde findindex (translate(arg(1)))\textgreater0
-
-\hypertarget{setmethod}{%
-\subsubsection{SETMETHOD}\label{setmethod}}
-
-::method setmethod /* rSTRING oMETHOD \emph{/
-a=self\textasciitilde exposed /} Associates entry with the specified
-name (translated to uppercase) with method method. Thus, the language
-processor returns the result of running method when you access this
-entry. \emph{/ /} (Part of METHOD checking converts string or array to
-actual method.) \emph{/ n=translate(arg(1))
-j=self\textasciitilde findindex(n) if j=0 \& \arg(2,`E') then return if
-\arg(2,`E') then do self\textasciitilde removeit (j) return end if j=0
-then do /} It's new \emph{/
-a{[}1{]}/}ItemsCount\emph{/=a{[}1{]}/}ItemsCount\emph{/ +1
-j=1+3}al{[}1{]} /\emph{ItemsCount}/ alj+l{]}=n end alj{]}=arg(2)
-a{[}j+2{]}=1 return
-
-\hypertarget{unknown}{%
-\subsubsection{UNKNOWN}\label{unknown}}
-
-::method unknown /* rSTRING rARRAY */
-
-/* Runs either the ENTRY or SETENTRY method, depending on whether the
-message name supplied ends with an equal sign. If the message name does
-not end with an equal sign, this method runs the ENTRY method, passing
-the message name as its argument. */
-
-if right (arg(1),1)=s=`=' then
-
-return self\textasciitilde entry(arg(1)) /* 2?? Not clear whether second
-argument is mandatory. */ t=.nil
-
-if arg(2,`E') then t=arg(2) {[}1{]} self\textasciitilde setentry (left
-(arg(1),length(arg(1))-1),t) 12.3 The stem class For some reason, the
-stem class doesn't have PUT and AT methods, which stops us having a
-general rule about {[}{]} synonyms AT, {[}{]}= synonyms PUT. Anyway,
-committee doing without this class as such.
-
-Here is temporary stuff showing how to use algebra in the collection
-coding.
-
-/* This 1998 version uses Rony's rules for XOR and INTERSECTION based on
-UNION and DIFFERENCE */
-
-/* Test Set-Operator-Methods on different collection objects */
-
-/* This top part has some rough parts - not meant for standard. */
-
-/* The dumps put out results sorted, so that comparisons can be made
-between implementations that keep collections in different orders. */
-
-/* Invocation example: settest.cmd 1\textgreater{} tmp.res
-2\textgreater{} tmp.err */
-
-/* Jnitial verification that new definitions are in effect */ J18list =
-.List\textasciitilde new if \J18list\textasciitilde hasmethod(``J18'')
-then signal error
-
-/* Input collections used for the tests */
-
-coll.1 = .array\textasciitilde of(1, 2,, 4)
-
-coll.2 = list\textasciitilde of(2, 3, 6)
-
-coll.3 = .queue\textsubscript{newPUSH(2)PUSH(3)}\textasciitilde PUSH(7)
-
-coll.4 = .directory\textsubscript{new}\textasciitilde setentry(1,
-``eins'')\textasciitilde\textasciitilde setentry(3, ``drei'')
-
-coll.5 = .bag\textsubscript{newput(2)put(3)put(5)}\textasciitilde put(2)
-
-coll.6 =
-.relation\textsubscript{new}\textasciitilde{}``{[}J=''(''zwei'',
-2)\sout{''{[}{]}=''('\,''drei'', 3)}''{[}J=''('vier'',
-8)\textasciitilde\textasciitilde''J=''C''drei'',3) coll.7 =
-.set\textasciitilde of(2, 3, 9){]}=tmpSupp\textasciitilde ITEM
-tmpSupp\textasciitilde NEXT END do until hope hope=1 do j=1 to
-k\textasciitilde dimension(1)-1 if
-k{[}j{]}\textsubscript{string\textgreater k{[}j+1{]}}string \textbar,
-(k{[}j{]}\textsubscript{string=k{[}j+1{]}}string \&
-i{[}j{]}\textsubscript{string\textless i{[}j+1{]}}string) then do{]}=t
-hope=0 end end end if O=collection\textasciitilde items then say'' The
-result is empty!'' else do j=1 to k\textasciitilde dimension(1) SAY ''
-'' ``index'' pp(k{[}j{]}) ``item'' ppd{[}j){]}=alj+1{]};a{[}j- 1
-{]}=alj+2{]} end j
-
-a{[}1{]}/\emph{ItemsCount}/=a{[} 1 {]}/\emph{ItemsCount}/-1
-
-return
-
-::method makearray expose a /* Returns a single-index array containing
-the receiver list items. \emph{/ r=.array\textasciitilde new = /} To
-build result in. \emph{/ do j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/
-r{[}r\textasciitilde dimension(1)+1{]}=a{[}j{]} end j returm r
-
-::method makearrayx private expose a
-
-/* Returns a single-index array containing the receiver index items.
-\emph{/ r=.array\textasciitilde new = /} To build result in. \emph{/ do
-j=4 by 3 to 143}a{[}1{]}/\emph{ItemsCount}/
-
-r{[}r\textasciitilde dimension(1)+1{]}=a{[}j+1{]} end j return r
-
-:imethod supplier expose a /* Returns a supplier object for the list. */
-return
-.supplier\textsubscript{new(self}makearray:.collection,self\textasciitilde makearrayx)
-
-::class `List' subclass Collection
-
-zimethod J18 = /* Here to demonstrate .LIST is replaced */ return
-
-/* List and Queue are special because there is an order to their
-elements. */
-
-::method put /* rANY rANY */ use arg item, index
-a=self\textasciitilde exposed
-
-/* PUT for a List must not be an insertion. */
-j=self\textasciitilde findindex(index) if j=0 then call Raise
-`Syntax',93.918 a{[}j{]}=item retum
-
-zimethod of class /* 1 or more oANY Are they omittable? Not in IOO
-\emph{/ /} Returns a newly created list containing the specified value
-objects in the order specified. */ r=self \textasciitilde{} new do j = 1
-to argQ) r \textasciitilde{} insert(arg(j)) end j return r
-
-zimethod insert /* rANY oANY */
-
-use arg item, index
-
-a=self\textasciitilde exposed /* Returns a list-supplied index for a new
-item, of specified value, which is added to the list. The new item
-follows the existing item with the specified index in the list ordering.
-\emph{/ /} Establish the index of what preceeds the new element. \emph{/
-/} Tf there was no index given, the new item becomes the last on list.
-\emph{/ /} mil argument means first */
-
-if arg(2,`E') then p=arg(2)
-
-else p=self\textasciitilde last
-
-/* Convert from list index to underlying array index. */
-
-if p==.nil then j=1 else j=self\textasciitilde findindex(p)
-
-if j=0 then call Raise `Syntax',93.918
-
-j=J+3 /* Where new entry will be. \emph{/ /} Move space to required
-place. */
-
-a{[}1{]}/\emph{ItemsCount}/=a{[}1 {]}/\emph{ItemsCount}/+1
-
-do k=1+3\emph{a{[}1{]}/}ItemsCount*/ by -3 to j+3
-
-a{[}k{]}=a{[}k-3{]};a{[}k+1 {]}=a{[}k-2{]};a{[}k{]}=a{[}k-3{]} end
-
-/* Insert new element. */
-
-a{[}j{]}=item /* A new, unique, index is needed. \emph{/ /} The basic
-requirement is for something unique, so this would be correct:
-
-i=.object\textasciitilde new /* a unique object, used as a key (the
-index on the list) \emph{/ }/ /* However, a number can be used. (At risk
-of the user thinking it is sensible to do arithmetic on it.) */
-
-a{[}j+1{]}=a{[}2{]}/\emph{Unique}/;a{[}2{]}/\emph{Unique}/=a{[}2
-{]}/\emph{Unique}/+1
-
-a{[}j+2{]}=0
-
-return a{[}j+1{]}
-
-:smethod first a=self\textasciitilde exposed
-
-/* Returns the index of the first item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount*/=0 then return .nil return a{[}5{]}
-
-::method last a=self\textasciitilde exposed
-
-/* Returns the index of the last item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount*/=0 then return .nil return
-a{[}3\emph{a{[}1{]}/}ItemsCount*/+2{]}\textbar{}
-
-::method firstitem a=self\textasciitilde exposed
-
-/* Returns the first item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount*/=0 then return .nil return a{[}4{]}
-
-::method lastitem a=self\textasciitilde exposed
-
-/* Returns the last item in the list. \emph{/ if
-a{[}1{]}/}ItemsCount\emph{/=0 then return .nil return
-a{[}3\emph{a{[}1{]}/}ItemsCount*/+1 \textbar{]}
-a{[}j+1{]}=al{[}j+4{]}-1; /} Index changes */ end j returm r
-
-zimethod queue =/* rANY */
-
-/* Adds the object value to the queue at its tail. \emph{/
-a=self\textasciitilde exposed a{[}1{]}/}ItemCount\emph{/=a{[}1
-\textbar/}ItemCount*/+1
-a{[}\{1+3\emph{a{[}1{]}/}ItemCount*/\textbar=arg(1)
-a{[}2+3\emph{a{[}1{]}/}ItemCount\emph{/\textbar=al{[} 1 {]}/}ItemCount*/
-return{]} r{[}n{]}=a{[}j+1{]} end j return
-.supplier\textasciitilde new(m,r){]}=1
-
-returm
-
-:imethod unknown =/* rSTRING rARRAY */
-
-/* Runs either the ENTRY or SETENTRY method, depending on whether the
-message name supplied ends with an equal sign. If the message name does
-not end with an equal sign, this method runs the ENTRY method, passing
-the message name as its argument. */ if right(arg(1),1)==`=``' then
-
-return self\textasciitilde entry(arg(1)) /* 22 Not clear whether second
-argument is mandatory. */ t=.nil
-
-if arg(2,`E') then t=arg(2){[}1{]}
-self\textasciitilde setentry(left(arg(1),length(arg(1))-1),t)
-
-routine CommonXor /* Returns a new collection that contains all items
-from self and the argument except that all indexes that appear in both
-collections are removed. \emph{/ /} When the target is a bag, there may
-be an index in the bag that is duplicated and the same value as an index
-in the argument. Should one copy of the index survive in the bag? */
-Ihs=arg(1)\textasciitilde difference(arg(2))
-rhs=Cast(arg(1),MayEnBag(arg(2)))\textasciitilde difference(arg(1))
-return lhs\textasciitilde union(rhs)
-
-nroutine CommonUnion /* Returns a new collection of the same class as
-SELF that contains all the items from SELF and items from the argument
-that have an index not in the first. \emph{/ /} Best to add them all. By
-adding non-receiver first we ensure that receiver takes priority when
-same indexes. \emph{/ This = arg(1) /} self of caller */
-r=This\textsubscript{class}new
-cs=MayEnBag(arg(2))\textasciitilde supplier do while
-cs\textasciitilde available
-r{[}cs\textasciitilde index{]}=cs\textasciitilde item
-cs\textasciitilde next end cs=This\textasciitilde supplier do while
-cs\textasciitilde available
-r{[}cs\textasciitilde index{]}=cs\textasciitilde item
-cs\textasciitilde next end return r
-
-nroutine CommonDifference /* Returns a new collection containing only
-those index-item pairs from the SELF whose indexes the other collection
-does not contain. \emph{/ This = arg(1) /} self of caller */
-r=This\textsubscript{class}new cs=This\textasciitilde supplier do while
-cs\textasciitilde available
-r{[}cs\textasciitilde index{]}=cs\textasciitilde item
-cs\textasciitilde next end cs=MayEnBag(arg(2))\textasciitilde supplier
-
-do while cs\textasciitilde available r\textsubscript{remove(cs}index)
-cs\textasciitilde next end
-
-return r
-
-routine MayEnBag
-
-/* For List and Queue the indexes are dropped. */ r=arg(1) if
-r\textasciitilde class == .List \textbar{} r\textasciitilde class ==
-.Queue then r=EnBag(r) retum r
-
-routine EnBag r=.Bag\textasciitilde new s=arg(1)\textasciitilde supplier
-do while s\textasciitilde available if arg(1)\textasciitilde class ==
-.List \textbar{} arg(1)\textasciitilde class == .Queue then
-1{[}s\textasciitilde item{]}=s\textasciitilde item else /* This case is
-when the receiver is a Bag. */
-1{[}s\textasciitilde index{]}=s\textasciitilde index
-s\textasciitilde next end return r
-
-/* This Cast routine commented away, since replaced by Oct 98 Rony
-version. routine Cast public use arg Target, Other TmpColl =
-Target\textsubscript{class}new /* Create an instance of type Target
-\emph{/ TmpSupp = Other\textasciitilde supplier /} Get supplier from
-Other */ signal on syntax do while TmpSupp\textasciitilde available
-TmpColl{[}TmpSupp\textasciitilde index{]} = TmpSupp\textasciitilde item
-TmpSupp\textasciitilde next end return TmpColl
-
-/* Tf syntax error 93.949, then target is an index-only collection like
-a set.*/ syntax:
-
-if condition( ``O'' )\textasciitilde code = ``93.949'' then signal
-IndexOnly
-
-raise propagate /* Unhandled syntax error, raise in caller */
-
-IndexOnly: /* This for index-only collections. */ do while
-TmpSupp\textasciitilde available
-TmpColl{[}TmpSupp\textasciitilde index{]} = TmpSupp\textasciitilde index
-TmpSupp\textasciitilde next
-
-end return TmpColl End commented away */
-
-/* 98-09-24, ---ref; CAST2.CMD return a collection of type ``target''
-which collected all item/index pairs of the argument ``other'' */
-
-2: ROUTINE cast PUBLIC USE ARG target, other
-
-SIGNAL ON SYNTAX IF ~other \textasciitilde{} HASMETHOD( ``SUPPLIER'' )
-THEN RAISE SYNTAX 98.907 ARRAY (``COLLECTION
-(i.e.~argument2=`other'-object must have a `SUPPLIER'-method)'' )
-
-tmpColl = target \textasciitilde{} CLASS \textasciitilde{} NEW /* create
-a an instance of type target \emph{/ tmpSupp = other\textasciitilde{}
-SUPPLIER /} get supplier from other */
-
-/* is index of ``other'' usable ? */ bIndexUsable = other
-\textasciitilde{} HASMETHOD( ``UNION'' )
-
-IF .Debug = .true THEN IF ~bIndexUsable THEN SAY'' /// index of `other'
-not usable for setlike-operations''
-
-/* possible syntax-error, if index and item must have the same value,
-e.g.~for sets/bags \emph{/ SIGNAL ON SYNTAX NAME INDEX ONLY target
-\textasciitilde{} CLASS \textasciitilde{} NEW \textasciitilde{} PUT(
-1,2) /} test, if target-type is index-only */
-
-SIGNAL ON SYNTAX DO WHILE tmpSupp \textasciitilde{} AVAILABLE IF
-bIndexUsable THEN tmpColl{[} tmpSupp \textasciitilde{} INDEX {]} =
-tmpSupp \textasciitilde{} ITEM ELSE tmpColl{[} tmpSupp \textasciitilde{}
-ITEM {]} = tmpSupp \textasciitilde{} ITEM tmpSupp \textasciitilde{} NEXT
-END RETURN tmpColl
-
-INDEX\_ONLY : /* this is for index-only collections (e.g.~sets, bags) */
-SIGNAL ON SYNTAX IF .Debug = .true THEN SAY'' \textbackslash{}`target'
-is an index-only collection (index==item)'' DO WHILE tmpSupp
-\textasciitilde{} AVAILABLE
-
-IF bIndexUsable THEN tmpColl{[} tmpSupp \textasciitilde{} INDEX {]} =
-tmpSupp \textasciitilde{} INDEX ELSE tmpColl{[} tmpSupp
-\textasciitilde{} ITEM {]} = tmpSupp \textasciitilde{} ITEM tmpSupp
-\textasciitilde{} NEXT END RETURN tmpColl
-
-SYNTAX: RAISE PROPAGATE /* raise error in caller */
-
-\hypertarget{the-stream-class}{%
-\section{The stream class}\label{the-stream-class}}
-
-The stream class provides input/output on external streams. ::class
-stream
-
-::method init /* rString \emph{/ Initializes a stream object for a
-stream named name, but does not open the stream. ::smethod query /}
-keywords */
-
-There is also QUERY as command with method COMMAND.
+\emph{There is also QUERY as command with method COMMAND.}
 
 Used with options, the QUERY method returns specific information about a
-stream. ::method charin
+stream.
 
-::smethod charout
+\lstinputlisting[language=rexx,label=streamiomethods.rexx,caption=streamiomethods.rexx]{streamiomethods.rexx}
 
-::method chars
+Returns a string after performing the specified stream command.
 
-::method linein
+\lstinputlisting[language=rexx,label=steamopenmethod.rexx,caption=steamopenmethod.rexx]{steamopenmethod.rexx}
 
-::method lineout
-
-::method lines
-
-::method qualify
-
-::method command /* rString */
-
-Returns a string after performing the specified stream command. ::method
-open
-
-There is also OPEN as command with method COMMAND.
+\emph{There is also OPEN as command with method COMMAND.}
 
@@
 
-Committee dropping OPEN POSITION QUERY SEEK as methods in favour of
-command use. ::method state
+\emph{Committee dropping OPEN POSITION QUERY SEEK as methods in favour
+of command use.}
+
+\lstinputlisting[language=rexx,label=streamstatemethod.rexx,caption=streamstatemethod.rexx]{streamstatemethod.rexx}
 
 Returns a string that indicates the current state of the specified
-stream. ::method say
+stream.
 
-::method uninit
+\lstinputlisting[language=rexx,label=streamvariousmethods.rexx,caption=streamvariousmethods.rexx]{streamvariousmethods.rexx}
 
-::method position /* Ugh \emph{/ POSITION is a synonym for SEEK.
-::method seek /} Ugh */
+POSITION is a synonym for SEEK.
+
+\lstinputlisting[language=rexx,label=streamseekmethod.rexx,caption=streamseekmethod.rexx]{streamseekmethod.rexx}
 
@@
 
-::method flush
+\lstinputlisting[language=rexx,label=streamflushmethod.rexx,caption=streamflushmethod.rexx]{streamflushmethod.rexx}
 
@@
 
-There is also FLUSH as command with method COMMAND.
+\emph{There is also FLUSH as command with method COMMAND.}
 
-Committee dropping FLUSH.
+\emph{Committee dropping FLUSH.}
 
-::method close
+\lstinputlisting[language=rexx,label=streamclosemethod.rexx,caption=streamclosemethod.rexx]{streamclosemethod.rexx}
 
@@
 
-There is also CLOSE as command with method COMMAND.
+\emph{There is also CLOSE as command with method COMMAND.}
 
-Semantics are `seen by other thread'. ::method string
+\emph{Semantics are `seen by other thread'.}
 
-::method makearray /* rCHARLINE */ Returns a fixed array that contains
-the data from the stream in line or character format, starting from the
-
-current read position. ::method supplier
-
-Returns a supplier object for the stream. ::method description
-
-::smethod arrayin /* rCHARLINE */
-
-Mixed case value works on OOI.
-
-Committee dropping Arrayin \& Arrayout. Arrayin == MakeArray
+\lstinputlisting[language=rexx,label=streamstringmethod.rexx,caption=streamstringmethod.rexx]{streamstringmethod.rexx}
 
 Returns a fixed array that contains the data from the stream in line or
-character format, starting from the
+character format, starting from the current read position.
 
-current read position. ::smethod arrayout /* rARRAY rCHARLINE */
+\lstinputlisting[language=rexx,label=streamsuppliermethod.rexx,caption=streamsuppliermethod.rexx]{streamsuppliermethod.rexx}
+
+Returns a supplier object for the stream.
+
+\lstinputlisting[language=rexx,label=streamdescriptionmethod.rexx,caption=streamdescriptionmethod.rexx]{streamdescriptionmethod.rexx}
+
+\lstinputlisting[language=rexx,label=streamaaryinmethod.rexx,caption=streamaaryinmethod.rexx]{streamaaryinmethod.rexx}
+
+\emph{Mixed case value works on OOI.}
+
+\emph{Committee dropping Arrayin \& Arrayout. Arrayin == MakeArray}
+
+Returns a fixed array that contains the data from the stream in line or
+character format, starting from the current read position.
+
+\lstinputlisting[language=rexx,label=streamarrayoutmethod.rexx,caption=streamarrayoutmethod.rexx]{streamarrayoutmethod.rexx}
 
@@
 
-\hypertarget{the-alarm-class}{%
-\section{The alarm class}\label{the-alarm-class}}
+\section{The alarm class}\label{the-alarm-class}
 
-::class alarm
+\lstinputlisting[language=rexx,label=alarmclass,caption=alarmclass]{alarmclass}
 
-::method init /* Time, Msg */ Sets up an alarm for a future time atime.
-::method cancel
+Sets up an alarm for a future time atime.
+
+\lstinputlisting[language=rexx,label=alarmcancelmethod.rexx,caption=alarmcancelmethod.rexx]{alarmcancelmethod.rexx}
 
@@
 
--local {[}`OUTPUT'{]} = .monitor\textasciitilde new(.output)
+\lstinputlisting[language=rexx,label=monitorclass.rexx,caption=monitorclass.rexx]{monitorclass.rexx}
 
-::class monitor
-
-\hypertarget{init-1}{%
-\subsection{INIT}\label{init-1}}
+\subsection{INIT}\label{init-1}
 
@@
 
-::method init /* oDESTINATION */ expose Destination Destination =
-.queue\textasciitilde new if arg(1,`E') then
-Destination\textasciitilde push (arg(1)) return
+\lstinputlisting[language=rexx,label=monitorinitmethod.rexx,caption=monitorinitmethod.rexx]{monitorinitmethod.rexx}
 
-\hypertarget{current}{%
-\subsection{CURRENT}\label{current}}
+\subsection{CURRENT}\label{current}
 
@@
 
-::smethod current expose Destination return Destination {[}1{]}
+\lstinputlisting[language=rexx,label=monitorcurrentmethod.rexx,caption=monitorcurrentmethod.rexx]{monitorcurrentmethod.rexx}
 
-\hypertarget{destination}{%
-\subsection{DESTINATION}\label{destination}}
+\subsection{DESTINATION}\label{destination}
 
@@
 
-::method destination /* oDESTINATION */ expose Destination if arg(1,`E')
-then Destination\textasciitilde push (arg(1)) else
-Destination\textasciitilde pull return Destination {[}1{]}
+\lstinputlisting[language=rexx,label=monitordestinationmethod.rexx,caption=monitordestinationmethod.rexx]{monitordestinationmethod.rexx}
 
-\hypertarget{unknown-1}{%
-\subsection{UNKNOWN}\label{unknown-1}}
+\subsection{UNKNOWN}\label{unknown-1}
 
@@
 
-::method unknown expose Destination
+\lstinputlisting[language=rexx,label=monitorunknownmethod.rexx,caption=monitorunknownmethod.rexx]{monitorunknownmethod.rexx}
 
-Extra parens needed here in original OREXX syntax
-
-forward to destination{[}1{]} message arg(1) arguments arg(2) return
+\emph{Extra parens needed here in original OREXX syntax}
+\lstinputlisting[language=rexx,label=monitorforwardsyntax.rexx,caption=monitorforwardsyntax.rexx]{monitorforwardsyntax.rexx}

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/syntax_constructs.tex
add filenames

@@
 \item
-  a left-hand side (called identifier);
+  a left-hand side (called \emph{identifier});
 \item
-  the characters ':=``;
+  the characters \texttt{\textquotesingle{}:="};
 \item
-  a right-hand side (called bnf\_expression). The left-hand side
-  identifies syntactic constructs. The right-hand side describes valid
-  ways of writing a specific syntactic construct. The right-hand side
-  consists of operands and operators, and may be grouped.
+  a right-hand side (called \emph{bnf\_expression}).
 \end{itemize}
+
+The left-hand side identifies syntactic constructs. The right-hand side
+describes valid ways of writing a specific syntactic construct.
+
+The right-hand side consists of operands and operators, and may be
+grouped.
 
@@
 otherwise it is called a terminal. Terminals are either literal or
-symbolic. Literal terminals are enclosed in quotes and represent
-literally (apart from case) what must be present in the source being
-described. Symbolic terminals formed with lower case characters
-represent something which the configuration may, or may not, allow in
-the source program, see nnn, nnn, nnn, nnn. Symbolic terminals formed
-with uppercase characters represent events and tokens, see nnn and nnn.
-\#\#\# Operators The following lists the valid operators, their meaning,
-and their precedence; the operator listed first has the highest
-precedence; apart from precedence recognition is from left to right:
+symbolic.
+
+Literal terminals are enclosed in quotes and represent literally (apart
+from case) what must be present in the source being described.
+
+Symbolic terminals formed with lower case characters represent something
+which the configuration may, or may not, allow in the source program,
+see nnn, nnn, nnn, nnn.
+
+Symbolic terminals formed with uppercase characters represent events and
+tokens, see nnn and nnn.
+
+\subsection{Operators}\label{operators}
+
+The following lists the valid operators, their meaning, and their
+precedence; the operator listed first has the highest precedence; apart
+from precedence recognition is from left to right:
 
@@
 \item
-  the operator `\textbar{}' specifies alternatives between the preceding
-  and the following constructs.
+  the operator \texttt{\textquotesingle{}\textbar{}\textquotesingle{}}
+  specifies alternatives between the preceding and the following
+  constructs.
 \end{itemize}
@@
 
-production := identifier `:=' bnf expression
-
-bnf expression t= abuttal \textbar{} bnf expression `\textbar{}' abuttal
-
-abuttal t= {[}abuttal{]} bnf primary
-
-bnf primary := `{[}T' bnf expression '{]}' \textbar{} `(' bnf expression
-`)' \textbar{} literal \textbar{}
-
-identifier \textbar{} message identifier \textbar{} bnf primary `+'
+\lstinputlisting[language=rexx,label=ebnfgrouping.ebnf,caption=ebnfgrouping.ebnf]{ebnfgrouping.ebnf}
 
@@
 The syntax descriptions (see nnn and nnn) make use of
-message\_identifiers which are shown as Msgnn.nn or Msgnn, where nn is a
-number. These actions produce the correspondingly numbered error
+\emph{message\_identifiers} which are shown as Msgnn.nn or Msgnn, where
+nn is a number. These actions produce the correspondingly numbered error
 messages (see nnn and nnn).
@@
 \item
-  EOL occurs at the end of a line of the source. It is provided by
-  Config\_SourceChar, see nnn;
+  \emph{EOL} occurs at the end of a line of the source. It is provided
+  by \texttt{Config\_SourceChar}, see nnn;
 \item
-  EOS occurs at the end of the source program. It is provided by
-  Config\_SourceChar;
+  \emph{EOS} occurs at the end of the source program. It is provided by
+  \texttt{Config\_SourceChar};
 \item
-  RADIX occurs when the character about to be scanned is `X' or `x' or
-  `B' or `b' not followed by a general\_letter, or a digit, or'.';
+  \emph{RADIX} occurs when the character about to be scanned is
+  \texttt{\textquotesingle{}X\textquotesingle{}} or
+  \texttt{\textquotesingle{}x\textquotesingle{}} or
+  \texttt{\textquotesingle{}B\textquotesingle{}} or
+  \texttt{\textquotesingle{}b\textquotesingle{}} not followed by a
+  \emph{general\_letter}, or a \emph{digit}, or
+  \texttt{\textquotesingle{}.\textquotesingle{}};
 \item
-  CONTINUE occurs when the character about to be scanned is `,', and the
-  characters after the `,' up to EOL represent a repetition of comment
-  or blank, and the EOL is not immediately followed by an EOS;
+  \emph{CONTINUE} occurs when the character about to be scanned is
+  \texttt{\textquotesingle{},\textquotesingle{}}, and the characters
+  after the \texttt{\textquotesingle{},\textquotesingle{}} up to
+  \emph{EOL} represent a repetition of \emph{comment} or \emph{blank},
+  and the \emph{EOL} is not immediately followed by an \emph{EOS};
 \item
-  EXPONENT\_SIGN occurs when the character about to be scanned is `+' or
-  `-', and the characters to the left of the sign, currently parsed as
-  part of Const\_symbol, represent a plain\_number followed by `E' or
-  `e', and the characters to the right of the sign represent a
-  repetition of digit not followed by a general\_letter or'.'.
-\item
-  would put ASSIGN here for the leftmost `=' in a clause that is not
-  within parentheses or brackets. But Simon not
+  \emph{EXPONENT\_SIGN} occurs when the character about to be scanned is
+  \texttt{\textquotesingle{}+\textquotesingle{}} or
+  \texttt{\textquotesingle{}-\textquotesingle{}}, and the characters to
+  the left of the sign, currently parsed as part of
+  \emph{Const\_symbol}, represent a \emph{plain\_number} followed by
+  \texttt{\textquotesingle{}E\textquotesingle{}} or
+  \texttt{\textquotesingle{}e\textquotesingle{}}, and the characters to
+  the right of the sign represent a repetition of \emph{digit} not
+  followed by a \emph{general\_letter} or
+  \texttt{\textquotesingle{}.\textquotesingle{}}.
+\end{itemize}
 
-  happy with message term being an assignment? \#\#\#\# Actions and
-  tokens Mixed case identifiers with an initial capital letter cause an
-  action when they appear as operands ina production. These actions
-  perform further tests and create tokens for use by the top syntax
-  level. The following actions are defined:
+\emph{- I would put \emph{ASSIGN} here for the leftmost
+\texttt{\textquotesingle{}=\textquotesingle{}} in a clause that is not
+within parentheses or brackets. But Simon not happy with message term
+being an assignment?}
+
+\subsubsection{Actions and tokens}\label{actions-and-tokens}
+
+Mixed case identifiers with an initial capital letter cause an action
+when they appear as operands ina production. These actions perform
+further tests and create tokens for use by the top syntax level. The
+following actions are defined:
+
+\begin{itemize}
 \item
@@
 
-starter:=x3j18
-
-x3j18:=program Eos \textbar{} Msg35.1
-
-program := {[}label list{]} {[}ncl{]} {[}requires+{]}
-{[}prolog\_instruction+{]} (class definition {[}requires+{]})+ requires
-:= `REQUIRES' ( taken constant \textbar{} Mgg19.8 ) ```;'+ prolog
-instruction:= (package \textbar{} import \textbar{} options) nel package
-= `PACKAGE'( NAME \textbar{} Msgnn ) import = `IMPORT' ( NAME \textbar{}
-Msgnn ) {[}`.'{]} options := `OPTIONS' ( symbol+ \textbar{} Msgnn ) nel
-:= null \_clause+ \textbar{} Msg21.1 null clause = `;' {[}label list{]}
-label list = (LABEL `;')+ class definition = class {[}property info{]}
-{[}method definition+{]} class = `CLASS' ( taken constant \textbar{}
-Msg19.12 ) {[}class\_option+{]}
-
-{[}`INHERIT' ( taken constant \textbar{} Msg19.13 )+{]} nel visibility
-\textbar{} modifier \textbar{} `BINARY' \textbar{} `DEPRECATED'
-`EXTENDS' ( NAME \textbar{} Msgnn ) `USES' ( NAMElist \textbar{} Msgnn )
-\textbar{} `IMPLEMENTS' ( NAMElist \textbar{} Msgnn )
-
-class option
-
-numeric digits:= `DIGITS' {[}expression{]}
-
-external \textbar{} metaclass \textbar{} submix /* \textbar{} 'PUBLIC!
-*/
-
-external = `EXTERNAL' (STRING \textbar{} Msg19.14) metaclass =
-`METACLASS' ( taken constant \textbar{} Msg19.15 ) submix = `MIXINCLASS'
-( taken constant \textbar{} Msg19.16 ) \textbar{} `SUBCLASS' ( taken
-constant \textbar{} Msg19.17 ) visibility = `PUBLIC' \textbar{}
-`PRIVATE' modifier = `ABSTRACT' \textbar{} `FINAL' \textbar{}
-`INTERFACE' \textbar{} `ADAPTER'! NAMElist = NAME {[}(`,' ( NAME
-\textbar{} Msgnn ) )+{]} property \_info = numeric \textbar{} property
-assignment \textbar{} properties \textbar{} trace numeric = `NUMERIC'
-(numeric digits \textbar{} numeric form \textbar{} Msg25.15) numeric
-form = `FORM' {[}`ENGINEERING' \textbar{} `SCIENTIFIC'{]} property
-agsignment := NAME \textbar{} assignment properties := `PROPERTIES' (
-properties option+ \textbar{} Msgnn) properties option := properties
-vigibility \textbar{} properties modifier properties visibility :=
-`INHERITABLE' \textbar{} `PRIVATE' \textbar{} `PUBLIC' \textbar{}
-`INDIRECT' properties modifier := `CONSTANT' \textbar{} `STATIC'
-\textbar{} `VOLATILE' \textbar{} `TRANSIENT' trace := `TRACE' {[}`ALL'
-\textbar{} `METHODS' \textbar{} `OFF' \textbar{} `RESULTS'{]} method
-definition = (method {[}expose ncl{]}\textbar{} routine) balanced expose
-= `EXPOSE' variable list method = `METHOD' (taken constant \textbar{}
-Msg19.9) {[} `(' assigncommalist \textbar{} Msgnn ( `)' \textbar{} Msgnn
-){]} {[}method option+{]} nel assigncommalist assignment {[}(`,' (
-assignment \textbar{} Msgnn ) )+{]}
-
-method visibility \textbar{} method modifier \textbar{} `PROTECT'
-\textbar{} `RETURNS' ( term \textbar{} Msgnn ) \textbar{} `SIGNAL' (
-termcommalist \textbar{} Msgnn )
-
-method option
-
-`DEPRECATED'! `CLASS' \textbar{} `ATTRIBUTE' \textbar{} /\emph{`PRIVATE'
-\textbar{} }/ guarded guarded := `GUARDED' \textbar{} `UNGUARDED! method
-visibility := 'INHERITABLE' \textbar{} `PRIVATE' \textbar{} `PUBLIC'
-\textbar{} `SHARED!' method modifier := `ABSTRACT' \textbar{} `CONSTANT'
-\textbar{} `FINAL' \textbar{} `NATIVE' \textbar{} `STATIC' termcommalist
-:= term {[}(`,' ( term \textbar{} Msgnn ) )+{]} routine := `ROUTINE' (
-taken constant \textbar{} Msg19.11 ) {[}`PUBLIC'{]} nel balanced:=
-instruction list {[}`END' Msgl0.1{]} instruction list:= instruction+
-
-/* The second part ig about groups */
-
-instruction = group \textbar{} gingle instruction nel group = do ncl
-\textbar{} if \textbar{} loop nel \textbar{} select nel do = do
-specification nel {[}instruction+{]} {[}group\_handler{]} (`END'
-{[}NAME{]} \textbar{} Eos Msg14.1 \textbar{} Msg35.1) group option :=
-`LABEL' ( NAME \textbar{} Msgnn ) \textbar{} `PROTECT' ( term \textbar{}
-Msgnn )
-
-group handler catch \textbar{} finally catch finally
-
-`CATCH' {[} NAME `=' {]} ( NAME \textbar{} Msgnn) nel {[}instruction+{]}
-
-catch = /* FINALLY implies a semicolon. */ finally = `FINALLY' nel (
-instruction+ \textbar{} Msgnn ) if := `IF' expression {[}ncl{]} (then
-\textbar{} Msg18.1) {[}else{]} then := `THEN' nel (instruction
-\textbar{} EOS Msg14.3 \textbar{} `END' Msg10.5) else := `ELSE' nel
-(instruction \textbar{} EOS Msg14.4 \textbar{} `END' Msg10.6) loop :=
-`LOOP' {[}group\_option+{]} {[}repetitor{]} {[}conditional{]} nel
-
-instruction+ {[}group\_handler{]} loop ending
-
-loop ending `END' {[}VAR SYMBOL{]} \textbar{} EOS Msg14.n \textbar{}
-Msg35.1
-
-conditional = `WHILE' whileexpr \textbar{} `UNTIL' untilexpr untilexpr =
-expression whileexpr = expression repetitor = assignment {[}count
-option+{]} \textbar{} expression \textbar{} over `FOREVER' count option
-= loopt \textbar{} loopb \textbar{} loopf loopt = `TO' expression loopb
-= `BY' expression loopf = `FOR' expression over = VAR SYMBOL `OVER'
-expression
-
-NUMBER `OVER' Msg31.1\\
-CONST SYMBOL `OVER' (Msg31.2 \textbar{} Msg31.3)
-
-select := `SELECT' {[}group option+{]} ncl select \_ body
-{[}group\_handler{]} (`END' {[}NAME Msg10.4{]} \textbar{} EOS Msgl14.2
-\textbar{} Msg7.2)
-
-(when \textbar{} Msg7.1) {[}when+{]} {[}otherwise{]}
-
-`WHEN' expresgion {[}ncl{]} (then \textbar{} Msg18.2)
-
-`OTHERWISE' ncl {[}instruction+{]}
-
-select body when otherwise
-
-/* Third part is for single instructions. */ single instruction:=
-assignment \textbar{} message instruction \textbar{} keyword instruction
-\textbar{} command assignment := VAR SYMBOL `\#' expression NUMBER `\#'
-Msg31.1 CONST SYMBOL `\#!' (Msg31.2 \textbar{} Mgg31.3)
-
-message instruction := message term \textbar{} message term `\#'
-expression keyword instruction:= address \textbar{} arg \textbar{} call
-\textbar{} drop \textbar{} exit interpret \textbar{} iterate \textbar{}
-leave
-
-nop \textbar{} numeric \textbar{} options parse \textbar{} procedure
-\textbar{} pull \textbar{} push \textbar{} queue
-
-raise \textbar{} reply \textbar{} return \textbar{} say \textbar{}
-signal \textbar{} trace \textbar{} use `THEN' Msg8.1 `ELSE' Msg8.2
-`WHEN' Msg9.1 \textbar{} `OTHERWISE' Msg9.2 command = expression address
-= `ADDRESS' {[}(taken constant {[}expression{]}
-
-Msg19.1 \textbar{} valueexp) {[} `WITH' connection{]} {]}
-
-taken constant symbol \textbar{} STRING
-
-valueexp = `VALUE' expression connection = ad option+ ad option = error
-\textbar{} input \textbar{} output \textbar{} Msg25.5 error = `ERROR!'
-(resourceo \textbar{} Msg25.14) input = `INPUT' (resourcei \textbar{}
-Msg25.6) resourcei := resources \textbar{} `NORMAL' output = `OUTPUT'
-(resourceo \textbar{} Mgg25.7) resourceo := `APPEND' (resources
-\textbar{} Msg25.8) \textbar{} `REPLACE' (resources \textbar{} Msg25.9)
-\textbar{} resources \textbar{} `NORMAL' resources := `STREAM'
-(VAR\_SYMBOL \textbar{} Msg53.1) \textbar{} `STEM' (VAR\_SYMBOL
-\textbar{} Msg53.2) vref := `(' var symbol (`)' \textbar{} Msg46.1) var
-symbol = VAR\_SYMBOL \textbar{} Msg20.1
-
-arg := `ARG' {[}template list{]}
-
-eall := `CALL' (callon\_spec \textbar{} (taken constant \textbar{} vref
-\textbar{} Msg19.2) {[}expression list{]} ) callon spec := `ON'
-(callable condition \textbar{} Msg25.1)
-
-{[}`NAME' (symbol constant term \textbar{} Msg19.3){]} \textbar{} `OFF'
-(callable condition \textbar{} Msg25.2)
-
-symbol constant term := term
-
-callable condition:= `ANY' \textbar{} `ERROR' \textbar{} `FAILURE'
-\textbar{} `HALT' \textbar{} `NOTREADY' \textbar{} `USER' ( symbol
-constant term \textbar{} Msg19.18 )
-
-condition := callable condition \textbar{} `LOSTDIGITS' \textbar{}
-`NOMETHOD' \textbar{} `NOSTRING' \textbar{} `NOVALUE' \textbar{}
-'SYNTAX!
-
-expression list do specification do simple
-
-expr \textbar{} {[}expr{]} `,' {[}expression list{]} do simple
-\textbar{} do repetitive `DO' {[}group\_option+{]}
-
-do repetitive = do simple (dorep \textbar{} conditional \textbar{} dorep
-conditional) dorep = `FOREVER' \textbar{} repetitor drop = `DROP'
-variable list variable list = (vref \textbar{} var\_symbol)+ exit =
-`EXIT' {[}expression{]} forward = `FORWARD' {[}forward \_option+
-\textbar{} Msg25.18{]} forward option = `CONTINUE' \textbar{}
-ArrayArgOption \textbar{} MessageOption \textbar{} ClassOption
-\textbar{} ToOption ArrayArgOption:=`ARRAY' arguments \textbar{}
-`ARGUMENTS' term MessageOption :=`MESSAGE' term ClassOption =`CLASS'
-term ToOption =`TO' term guard = `GUARD' (`ON' \textbar{} Msg25.22)
-{[}(`WHEN' \textbar{} Msg25.21) expression{]} \textbar{} ( `OFF'
-\textbar{} Msg25.19) {[}(`WHEN' \textbar{} Msg25.21) expression{]}
-interpret = `INTERPRET' expression iterate = `ITERATE' {[}VAR SYMBOL
-\textbar{} Msg20.2{]} leave = `LEAVE' {[}VAR SYMBOL \textbar{}
-Msg20.2{]} nop = `NOP!' numeric = `NUMERIC' (numeric digits \textbar{}
-numeric form
-
-numeric fuzz \textbar{} Msg25.15)
-
-numeric digits `DIGITS' {[}expression{]} numeric form `FORM' {[}numeric
-form suffix{]} numeric form suffix:=(`ENGINEERING'
-\textbar{}`SCIENTIFIC'\textbar valueexp \textbar{} Msg25.11) numeric
-fuzz `FUZZ' {[}expression{]}
-
-options = `OPTIONS' expression parse = `PARSE' {[}translations{]} (parse
-type \textbar Msg25.12) {[}template list{]} translations := `CASELESS'
-{[}`UPPER' \textbar{} `LOWER'{]} \textbar{} (`UPPER' \textbar{} `LOWER')
-{[}`CASELESS'{]} parse type = parse key \textbar{} parse value
-\textbar{} parse var \textbar{} term parse key = `ARG' \textbar{} `PULL'
-\textbar{} `SOURCE' \textbar{} `LINEIN' \textbar{} `VERSION! parse value
-= 'VALUE' {[}expression{]} (`WITH' \textbar{} Mgg38.3) parse var = `VAR'
-var symbol template := NAME {[}( {[}pattern{]} NAME) +{]} pattern:=
-STRING \textbar{} {[}indicator{]} NUMBER \textbar{} {[}indicator{]} `('
-symbol `)' indicator := `+' \textbar{} `-' \textbar{} `s! procedure =
-'PROCEDURE' {[}expose \textbar{} Msg25.17{]} pull = `PULL' {[}template
-list{]} push = `PUSH' {[}expression{]} queue = `QUEUE' {[}expression{]}
-raise = `RAISE' conditions (raise option \textbar{} Msg25.24) conditions
-= `ANY' \textbar{} `ERROR' term \textbar{} `FAILURE' term \textbar{}
-`HALT'\textbar{} `LOSTDIGITS' \textbar{} `NOMETHOD' \textbar{}
-`NOSTRING' \textbar{} ``NOTREADY' \textbar{} `NOVALUE' \textbar{}
-`PROPAGATE' \textbar{} `SYNTAX' term \textbar{} `USER' ( symbol constant
-term \textbar{} Msg19.18) \textbar{} Msg25.23 raise option :=
-ExitRetOption \textbar{} Description \textbar{} ArrayOption
-
-ExitRetOption := `EXIT! {[}term{]} \textbar{} 'RETURN' {[}term{]}
-Description =`DESCRIPTION' term ArrayOption = `ADDITIONAL' term
-\textbar{} ```ARRAY' arguments reply = `REPLY' {[} expression{]} return
-= `RETURN' {[}expression{]} say = `SAY' {[}expression{]} signal =
-`SIGNAL' (signal spec \textbar{} valueexp
-
-symbol constant term \textbar{} Msg19.4)
-
-signal spec := `ON' (condition \textbar{} Msg25.3) {[}`NAME' (symbol
-constant term \textbar{} Msg19.3){]} \textbar{} `OFF' (condition
-\textbar{} Msg25.4)
-
-trace = `TRACE' {[}(taken\_constant \textbar{} Msgl19.6) \textbar{}
-valueexp{]} use = `USE' (`ARG' \textbar{} Msg25.26) {[}use list{]} use
-list = VAR\_SYMBOL \textbar{} {[}VAR SYMBOL{]} `,' {[}use list{]}
-
-/* Note: The next part describes templates. */ template list template
-\textbar{} {[}template{]} `,' {[}template list{]}
-
-template = (trigger \textbar{} target \textbar{} Msg38.1)+ target =
-VAR\_SYMBOL \textbar{} `.!' trigger = pattern \textbar{} positional
-pattern = STRING \textbar{} vrefp vrefp = `(' (VAR\_SYMBOL \textbar{}
-Msgl19.7) (`)' \textbar{} Msg46.1) positional = absolute positional
-\textbar{} relative positional absolute positional:= NUMBER `=' position
-pogition := NUMBER \textbar{} vrefp \textbar{} Msg38.2 relative
-posgitional:= (`+' \textbar{} `-``') position
-
-/* Note: The final part specifies the various forms of symbol, and
-expression. */
-
-symbol = VAR\_SYMBOL \textbar{} CONST SYMBOL \textbar{} NUMBER
-expression = expr {[}(`,' Msg37.1) \textbar{} (`)' Msg37.2 ){]} expr =
-expr alias
-
-expr alias and expression
-
-expr alias or operator and expression
-
-or operator := `\textbar{}' \textbar{} `\&\&! and expression :=
-comparison \textbar{} and expression'\&' comparison comparison :=
-concatenation \textbar{} comparison comparison operator concatenation
-comparison \_operator:= normal compare \textbar{} strict compare normal
-compare:= `=! \textbar{} ``\s! \textbar{} hep! \textbar{} tact
-\textbar{} tot \textbar{} tet \textbar{} toet len! W\textgreater!
-\textless! strict\_compare:='==' \textbar{} `\s=' \textbar{}
-`\textgreater\textgreater!' \textbar{} `\textless\textless!' \textbar{}
-`\textgreater\textgreater e! \textbar{}'\textless\textless s! \textbar{}
-```\e\textgreater t \textbar{}'\textless c! concatenation := addition
-\textbar{} concatenation (' ' \textbar{} `\textbar\textbar{}') addition
-addition := multiplication \textbar{} addition additive operator
-multiplication additive operator:= `+' \textbar{} '-! multiplication t=
-power expression
-
-multiplication multiplicative operator
-
-power expression multiplicative operator:= '*' \textbar{} `/' \textbar{}
-`//' \textbar{} `\%! power expression := prefix expression \textbar{}
-power expression'**' prefix expression prefix expression := (`+'
-\textbar{} `-' \textbar{} '\textquotesingle) prefix expression term
-\textbar{} Msg35.1 /* ``Stub'' has to be identified semantically? */
-
-term = simple term {[} `.' ( term \textbar{} Msgnn ){]} simple term :=
-symbol \textbar{} STRING \textbar{} invoke \textbar{} indexed `('
-expression ( `)' \textbar{} Msg36 ) initializer \textbar{} message term
-`\#\#! message term:= term ('\textasciitilde' \textbar{}
-`\textasciitilde\textasciitilde{}') method name {[}arguments{]} term
-`{[}'{[} expression list {]} ('{]}' \textbar{} Msg36.2)
-
-method name:=(taken constant \textbar{} Msg19.19) {[}`:' ( VAR\_SYMBOL
-\textbar{} Msg19.21 ){]} /* Method-call without arguments ig
-syntactically like symbol. \emph{/ /} Editor - not sure of my notes
-about here. */
-
-invoke := (symbol \textbar{} STRING) arguments arguments := `\#('
-{[}expression list{]} (`)' \textbar{} Meg36) expression list :=
-expregsion \textbar{} {[}expression{]} `,' {[}expression list{]} indexed
-= (symbol \textbar{} STRING) indices indices = `\#{[}' {[}expression
-list{]} ('{]}' \textbar{} Msg36.n) initializer = `{[}'expression list
-('{]}' \textbar{} Msg36.n)
+\lstinputlisting[language=rexx,label=syntaxlevel.ebnf,caption=syntaxlevel.ebnf]{syntaxlevel.ebnf}
 
@@
 
-Any VAR\_SYMBOL in a do\_ending must be matched by the same VAR\_SYMBOL
-occurring at the start of an assignment contained in the
-do\_specification of the do that contains both the do\_specification and
-the do\_ending, as described in nnn. If there is a VAR\_SYMBOL in a
-do\_ending for which there is no assignment in the corresponding
-do\_specification then message Msg10.3 is produced and no further
-activity is defined. If there is a VAR\_SYMBOL in a do\_ending which
-does not match the one occurring in the assignment then message Msg10.2
-is produced and no further activity is defined. An iterate or leave must
-be contained in the instruction\_list of some do with a
-do\_specification which is do\_repetitive, otherwise a message (Msg28.2
-or Msg28.1 respectively) is produced and no further activity is defined.
-If an iterate or leave contains a VAR\_SYMBOL there must be a matching
-VAR\_SYMBOL ina do\_specification, otherwise a message (Msg28.1,
-Msg28.2, Msg28.3 or Msg28.4 appropriately) is produced and no further
-activity is defined. The matching VAR\_SYMBOL will occur at the start of
-an assignment in the do\_specification. Tne do\_specification will be
-associated with a do by nnn. The /ferafe or leave will be a single
-instruction in an instruction\_list associated with a do by nnn. These
-two dos shall be the same, or the latter nested one or more levels
-within the former. The number of levels is called the
-nesting\_correction and influences the semantics of the iterafe or
-leave. It is zero if the two dos are the same. The nesting\_correction
-for /ferates or leaves that do not contain VAR\_SYMBOL is zero.
+Any \texttt{VAR\_SYMBOL} in a \emph{do\_ending} must be matched by the
+same \texttt{VAR\_SYMBOL} occurring at the start of an \emph{assignment}
+contained in the \emph{do\_specification} of the \emph{do} that contains
+both the \emph{do\_specification} and the \emph{do\_ending}, as
+described in nnn.
+
+If there is a \emph{VAR\_SYMBOL} in a \emph{do\_ending} for which there
+is no \emph{assignment} in the corresponding \emph{do\_specification}
+then message Msg10.3 is produced and no further activity is defined.
+
+If there is a :VAR\_SYMBOL\_ in a \emph{do\_ending} which does not match
+the one occurring in the \emph{assignment} then message Msg10.2 is
+produced and no further activity is defined.
+
+An \emph{iterate} or \emph{leave} must be contained in the
+\emph{instruction\_list} of some \emph{do} with a
+\emph{do\_specification} which is \emph{do\_repetitive}, otherwise a
+message (Msg28.2 or Msg28.1 respectively) is produced and no further
+activity is defined.
+
+If an \emph{iterate} or \emph{leave} contains a \emph{VAR\_SYMBOL} there
+must be a matching \emph{VAR\_SYMBOL} in a \emph{do\_specification},
+otherwise a message (Msg28.1, Msg28.2, Msg28.3 or Msg28.4 appropriately)
+is produced and no further activity is defined. The matching
+\texttt{VAR\_SYMBOL} will occur at the start of an \emph{assignment} in
+the \emph{do\_specification}. Tne \emph{do\_specification} will be
+associated with a \emph{do} by nnn. The \emph{iterate} or \emph{leave}
+will be a single \emph{instruction} in an \emph{instruction\_list}
+associated with a \emph{do} by nnn. These two dos shall be the same, or
+the latter nested one or more levels within the former. The number of
+levels is called the \emph{nesting\_correction} and influences the
+semantics of the \emph{iterate} or \emph{leave}. It is zero if the two
+dos are the same. The \emph{nesting\_correction} for \emph{iterates} or
+\emph{leaves} that do not contain \emph{VAR\_SYMBOL} is zero.
 
@@
 
-Instances of LABEL which occur within a grouping\_instruction and are
-not in a nc/ at the end of that grouping\_instruction are instances of
-trace-only labels.
+Instances of \emph{LABEL} which occur within a
+\emph{grouping\_instruction} and are not in a \emph{ncl} at the end of
+that \emph{grouping\_instruction} are instances of trace-only labels.
 
@@
 token. The semicolon may be explicit in the program or inferred. The
-line number of a clause is one more than the number of EOL events
+line number of a clause is one more than the number of \emph{EOL} events
 recognized before the first token of the clause was recognized.
@@
 
-The syntax specification nnn allows `IF' instructions to be nested and
-does not fully specify the association of an `ELSE' keyword with an `IF'
-keyword. An `ELSE' associates with the closest prior `IF' that it can
-associate with in conformance with the syntax.
+The syntax specification nnn allows
+\texttt{\textquotesingle{}IF\textquotesingle{}} instructions to be
+nested and does not fully specify the association of an
+\texttt{\textquotesingle{}ELSE\textquotesingle{}} keyword with an
+\texttt{\textquotesingle{}IF\textquotesingle{}} keyword. An
+\texttt{\textquotesingle{}ELSE\textquotesingle{}} associates with the
+closest prior \texttt{\textquotesingle{}IF\textquotesingle{}} that it
+can associate with in conformance with the syntax.
 
@@
 \item
-  Msg15.1 shall be preferred to Msg15.3 if the choice of Msg15.3 would
-  result in the replacement for the insertion being a blank character;
+  \emph{Msg15.1} shall be preferred to \emph{Msg15.3} if the choice of
+  \emph{Msg15.3} would result in the replacement for the insertion being
+  a blank character;
 \item
-  Msg15.2 shall be preferred to Msg15.4 if the choice of Msg15.4 would
-  result in the replacement for the insertion being a blank character;
+  \emph{Msg15.2} shall be preferred to \emph{Msg15.4} if the choice of
+  \emph{Msg15.4} would result in the replacement for the insertion being
+  a blank character;
 \item
-  Msg31.3 shall be preferred to Msg31.2 if the replacement for the
-  insertion in the message starts with a period;
+  \emph{Msg31.3} shall be preferred to \emph{Msg31.2} if the replacement
+  for the insertion in the message starts with a period;
 \item
   Preference is given to the message that appears later in the list:
-  Msg21.1, Msg27.1, Msg25.16, Msg36, Msg38.3, Msg35.1, other messages.
+  \emph{Msg21.1}, \emph{Msg27.1}, \emph{Msg25.16}, \emph{Msg36},
+  \emph{Msg38.3}, \emph{Msg35.1}, other messages.
 \end{itemize}
@@
 
-The message\_identifiers in clause 6 correlate with the tails of stem
-\#ErrorText., which is initialized in nnn to identify particular
+The \emph{message\_identifiers} in clause 6 correlate with the tails of
+stem \#ErrorText., which is initialized in nnn to identify particular
 messages. The action of producing an error message will replace any
@@
 with information on the origin of the error, to the configuration by
-writing on the default error stream. Further activity by the language
-processor is permitted, but not defined by this standard. The effect of
-an error during the writing of an error message is not defined.
+writing on the default error stream.
+
+Further activity by the language processor is permitted, but not defined
+by this standard.
+
+The effect of an error during the writing of an error message is not
+defined.
 
@@
 The error message selected by the message number is preceded by a
-prefix. The text of the prefix is \#ErrorText.0.1 except when the error
-is in source that execution of an interactive trace interpret
-instruction (see nnn) is processing, in which case the text is
-\#ErrorText.0.2. The insert called in these texts is the message number.
-The insert called is the line number of the error. The line number of
-the error is one more than the number of EOL events encountered before
-the error was detectable, except for messages Msg6.1, Msg14, Msg14.1,
-Msg14.2, Msg14.3, and Msg14.4. For Msg6.1 it is one more than the number
-of EOL events encountered before the line containing the unmatched '/*'.
-For the others, it is the line number of the clause containing the
-keyword referenced in the message text. The insert called is the value
-provided on the API\_ Start function which started processing of the
-program, see nnn.
+prefix. The text of the prefix is \emph{\#ErrorText.0.1} except when the
+error is in source that execution of an interactive trace
+\emph{interpret} instruction (see nnn) is processing, in which case the
+text is \emph{\#ErrorText.0.2}. The insert called
+\texttt{\textless{}value\textgreater{}} in these texts is the message
+number. The insert called \texttt{\textless{}linenumber\textgreater{}}
+is the line number of the error. The line number of the error is one
+more than the number of \emph{EOL} events encountered before the error
+was detectable, except for messages \emph{Msg6.1}, \emph{Msg14},
+\emph{Msg14.1}, \emph{Msg14.2}, \emph{Msg14.3}, and \emph{Msg14.4}. For
+\emph{Msg6.1} it is one more than the number of \emph{EOL} events
+encountered before the line containing the unmatched
+\texttt{\textquotesingle{}/*\textquotesingle{}}. For the others, it is
+the line number of the clause containing the keyword referenced in the
+message text.
+
+The insert called \texttt{\textless{}source\textgreater{}} is the value
+provided on the \texttt{API\_Start} function which started processing of
+the program, see nnn.
 
@@
 Within the text of error messages, an insertion consists of the
-characters `\textless{}', `\textgreater{}', and what is between those
-characters. There will be a word in the insertion that specifies the
-replacement text, with the following meaning:
+characters \texttt{\textquotesingle{}\textless{}\textquotesingle{}},
+\texttt{\textquotesingle{}\textgreater{}\textquotesingle{}}, and what is
+between those characters. There will be a word in the insertion that
+specifies the replacement text, with the following meaning:
 
@@
 \item
-  if the word is `hex-encoding' and the message is not Msg23.1 then the
-  replacement text is the value of the leftmost character which caused
-  the source to be syntactically incorrect. The value is in hexadecimal
-  notation;
+  if the word is
+  \texttt{\textquotesingle{}hex-encoding\textquotesingle{}} and the
+  message is not \emph{Msg23.1} then the replacement text is the value
+  of the leftmost character which caused the source to be syntactically
+  incorrect. The value is in hexadecimal notation;
 \item
-  if the word is `token' then the replacement text is the part of the
-  source program which was recognized as the detection token, or in the
-  case of Msg31.1 and Msg31.2, the token before the detection token. The
-  detection token is the leftmost token for which the program up to and
-  including the token could not be parsed as the left part of a program
-  without causing a message. If the detection token is a semicolon that
-  was not present in the source but was supplied during recognition then
-  the replacement is the previous token;
-\item
-  if the word is `position' then the replacement text is a number
-  identifying the detection character. The detection character is the
-  leftmost character in the hex\_string or binary\_string which did not
-  match the required syntax. The number is a count of the characters in
-  the string which preceded the detection character, including the
-  initial quote or apostrophe. In deciding the leftmost blank in a
-  quoted string of radix `X' or `B' that is erroneous not that: A blank
-  as the first character of the quoted string is an error. The leftmost
-  embedded sequence of blanks can validly follow any number of non-blank
-  characters. Otherwise a blank run that follows an odd numbered
-  sequence of non-blanks (or a number not a multiple of four in the case
-  of radix `B') is not valid. If the string is invalid for a reason not
-  described above, the leftmost blank of the rightmost sequence of
-  blanks is the invalid blank to be referenced in the message;
-\item
-  if the word is `char' then the replacement text is the detection
-  character;
-\item
-  if the word is `linenumber' then the replacement text is the line
-  number of a clause associated with the error. The wording of the
-  message text specifies which clause that is;
-\item
-  if the word is `keywords' then the replacement text is a list of the
-  keywords that the syntax would allow at the context where the error
-  occurred. If there are two keywords they shall be separated by the
-  four characters ' or `. If more, the last shall be preceded by the
-  three characters 'or' and the others shall be followed by the two
-  characters `,'. The keywords will be uppercased and in alphabetical
-  order.
+  if the word is \texttt{\textquotesingle{}token\textquotesingle{}} then
+  the replacement text is the part of the source program which was
+  recognized as the detection token, or in the case of \emph{Msg31.1}
+  and \emph{Msg31.2}, the token before the detection token.
 \end{itemize}
 
-Replacement text is truncated to \#Limit\_Messagelnsert characters if it
-would otherwise be longer than that, except for a keywords replacement.
-When an insert is both truncated and appears within quotes in the
-message, the three characters `\ldots{}' are inserted in the message
-after the trailing quote.
+The detection token is the leftmost token for which the program up to
+and including the token could not be parsed as the left part of a
+program without causing a message. If the detection token is a semicolon
+that was not present in the source but was supplied during recognition
+then the replacement is the previous token;
+
+\begin{itemize}
+\item
+  if the word is \texttt{\textquotesingle{}position\textquotesingle{}}
+  then the replacement text is a number identifying the detection
+  character. The detection character is the leftmost character in the
+  \emph{hex\_string} or \emph{binary\_string} which did not match the
+  required syntax. The number is a count of the characters in the string
+  which preceded the detection character, including the initial quote or
+  apostrophe. In deciding the leftmost blank in a quoted string of radix
+  \texttt{\textquotesingle{}X\textquotesingle{}} or
+  \texttt{\textquotesingle{}B\textquotesingle{}} that is erroneous note
+  that:
+
+  \begin{itemize}
+  \item
+    A blank as the first character of the quoted string is an error.
+  \item
+    The leftmost embedded sequence of blanks can validly follow any
+    number of non-blank characters.
+  \item
+    Otherwise a blank run that follows an odd numbered sequence of
+    non-blanks (or a number not a multiple of four in the case of radix
+    \texttt{\textquotesingle{}B\textquotesingle{}}) is not valid.
+  \item
+    If the string is invalid for a reason not described above, the
+    leftmost blank of the rightmost sequence of blanks is the invalid
+    blank to be referenced in the message;
+  \end{itemize}
+\item
+  if the word is \texttt{\textquotesingle{}char\textquotesingle{}} then
+  the replacement text is the detection character;
+\item
+  if the word is \texttt{\textquotesingle{}linenumber\textquotesingle{}}
+  then the replacement text is the line number of a clause associated
+  with the error. The wording of the message text specifies which clause
+  that is;
+\item
+  if the word is \texttt{\textquotesingle{}keywords\textquotesingle{}}
+  then the replacement text is a list of the keywords that the syntax
+  would allow at the context where the error occurred. If there are two
+  keywords they shall be separated by the four characters
+  \texttt{\textquotesingle{}\ or\ \textquotesingle{}}. If more, the last
+  shall be preceded by the three characters
+  \texttt{\textquotesingle{}or\textquotesingle{}} and the others shall
+  be followed by the two characters
+  \texttt{\textquotesingle{},\textquotesingle{}}. The keywords will be
+  uppercased and in alphabetical order.
+\end{itemize}
+
+Replacement text is truncated to \texttt{\#Limit\_Messagelnsert}
+characters if it would otherwise be longer than that, except for a
+keywords replacement. When an insert is both truncated and appears
+within quotes in the message, the three characters
+\texttt{\textquotesingle{}...\textquotesingle{}} are inserted in the
+message after the trailing quote.
 
@@
 
-If a message\_term contains a `{[}' it is regarded as an equivalent
-message\_term without a `{[}', for execution. The equivalent is
-term\textasciitilde{}`{[}{]}'(expression\_list). See nnn. If a
-message\_instruction has the construction message\_term `=' expression
-it is regarded as equivalent to a message\_term with the same components
-as the message\_term left of the `=', except that the taken\_constant
-has an `=' character appended and arguments has the expression from the
-right of the `=' as an extra first argument. See nnn.
+If a message\_term contains a
+\texttt{\textquotesingle{}{[}\textquotesingle{}} it is regarded as an
+equivalent message\_term without a
+\texttt{\textquotesingle{}{[}\textquotesingle{}}, for execution. The
+equivalent is
+\texttt{term\textasciitilde{}\textquotesingle{}{[}{]}\textquotesingle{}(expression\_list)}.
+See nnn. If a \emph{message\_instruction} has the construction
+\texttt{message\_term\ \textquotesingle{}=\textquotesingle{}\ expression}
+it is regarded as equivalent to a \emph{message\_term} with the same
+components as the \emph{message\_term} left of the
+\texttt{\textquotesingle{}=\textquotesingle{}}, except that the
+\emph{taken\_constant} has an
+\texttt{\textquotesingle{}=\textquotesingle{}} character appended and
+\emph{arguments} has the expression from the right of the
+\texttt{\textquotesingle{}=\textquotesingle{}} as an extra first
+argument. See nnn.

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/to_be_processed.tex
work in standard book production process including readme.md
upd
add filenames

@@
-\hypertarget{to-be-processed}{%
-\chapter{To be processed:}\label{to-be-processed}}
+%preprocessed texin
+\chapter{To be processed:}\label{to-be-processed}
 
-The following decisons are abstracted from minutes. We need to ensure
-they are covered in the main standard and their rationale appropriately
-reworded for this annex.
+\emph{The following decisons are abstracted from minutes. We need to
+ensure they are covered in the main standard and their rationale
+appropriately reworded for this annex.}
 
@@
 different variables without upsetting the usage of the object).
-Classess. There will be `factory' objects capable of creating multiple
+Classess. There will be factory' objects capable of creating multiple
 new objects which have common characteristics about how they can be
@@
 adding to the semantics of another class. This relation is used to form
-a tree. We prefer a singly rooted tree, rooted in the class `Object'
+a tree. We prefer a singly rooted tree, rooted in the class Object'
 which is built-in to the language. Other classes will also be built-in.
@@
 from classic Rexx to justify a new syntax construct. The new syntax is
-Receiver \textasciitilde{} MethodName(Arguments) and implies both a
-different search for the method to be invoked and a special role for the
-receiver as opposed to the other arguments of the invocation.
+\texttt{Receiver\textasciitilde{}MethodName(Arguments)} and implies both
+a different search for the method to be invoked and a special role for
+the receiver as opposed to the other arguments of the invocation.
 
-Packaging: In principle a `program builder' could be used in developing
+Packaging: In principle a program builder' could be used in developing
 Rexx programs with many classes and methods, and that builder could hide
@@
 objects without shriek names are not candidates for standardising, eg
-SYSTEM, .KERNEL.
+\texttt{.SYSTEM}, \texttt{.KERNEL}.
 
@@
 Persistent objects. It is our belief that support for very-long-running
-programs is required. It is a moot point whether the ENVIRONMENT
-directory is enough.
+programs is required. It is a moot point whether the
+\texttt{.ENVIRONMENT} directory is enough.
 
@@
 
-Critical sections. The GUARD mechanism can be used in a `critical
-section' style. Nothing will be added to the definition.
+Critical sections. The \texttt{GUARD} mechanism can be used in a
+critical section' style. Nothing will be added to the definition.
 
@@
 programs. (A warning about this was put in A8.3.3 of X3.274.) It
-produces surprises for OO programmers, eg a==b after
-a=.stem\textasciitilde new; b=.stem\textasciitilde new. The proposed
-alternative is to make the presence/absence of a dot at the end of the
-name determine whether coercion to string is done. The `classic' meaning
-of A.=B. would be restored but AA=BB, AA==BB etc. would have their OO
-meanings. The meaning of USE ARG with a dotted name would be defined to
-allow `by reference' passing of a stem. Square brackets could be used
-with both dotted and undotted names. A further proposal is to note that
-this leaves few differences between the DIRECTORY class and the
-non-dotted STEM class so that it might be a further improvement if the
-DIRECTORY class was extended to the extent that the STEM class was
-unnecessary.
+produces surprises for OO programmers, eg \texttt{a==b} after
+\texttt{a=.stem\textasciitilde{}new;\ b=.stem\textasciitilde{}new}. The
+proposed alternative is to make the presence/absence of a dot at the end
+of the name determine whether coercion to string is done. The `classic'
+meaning of \texttt{A.=B.} would be restored but \texttt{AA=BB},
+\texttt{AA==BB} etc. would have their OO meanings. The meaning of
+\texttt{USE\ ARG} with a dotted name would be defined to allow by
+reference' passing of a stem. Square brackets could be used with both
+dotted and undotted names. A further proposal is to note that this
+leaves few differences between the DIRECTORY class and the non-dotted
+STEM class so that it might be a further improvement if the DIRECTORY
+class was extended to the extent that the STEM class was unnecessary.
 
@@
 analysis may show that more of I/O could (\& should?) be made standard
-or may show that some OO! I/O language should not be standardized.
+or may show that some OOI I/O language should not be standardized.
 
@@
 suitable to be defined in a standard. Potentially, all the builtin
-classes and objects (which are reachable from .ENVIRONMENT) might be
-standardized. However, names which start with an exclamation mark denote
-unsuitable things. The committee also thought the following unsuitable:
+classes and objects (which are reachable from \texttt{.ENVIRONMENT})
+might be standardized. However, names which start with an exclamation
+mark denote unsuitable things. The committee also thought the following
+unsuitable:
 
@@
 \item
-  Anything specific to SOM. - RX\_QUEUE - Stream\_Supplier - Parts of
-  LOCAL other than direct reference to the default streams. There is a
-  naming problem with this. The names in OOI are STDIN, STDOUT and
-  STDERR. We would prefer INPUT, OUTPUT, and ERROR to be\_ consistent
+  Anything specific to SOM.
+\item
+  \texttt{RX\_QUEUE}
+\item
+  Stream\_Supplier
+\item
+  Parts of \texttt{.LOCAL} other than direct reference to the default
+  streams. There is a naming problem with this. The names in OOI are
+  \texttt{STDIN}, \texttt{STDOUT} and \texttt{STDERR}. We would prefer
+  \texttt{INPUT}, \texttt{OUTPUT}, and \texttt{ERROR} to be consistent
   with the keywords. OOI has used those names for something else. We
   will work on the proposal that we use the prefered names and the
-  MONITOR class is dropped. (Users who want the monitor function can get
-  it with a few lines of directive.)
+  \texttt{MONITOR} class is dropped. (Users who want the monitor
+  function can get it with a few lines of directive.)
 \end{itemize}
@@
 The committee feels that OOI over-specifies the index of an item in a
-LIST. In OO it is a count giving the sequence over time of the
+\texttt{LIST}. In OOI it is a count giving the sequence over time of the
 insertions in the list. The risk in using numbers is that they may be
 (wrongly) used as positions, and arithmetic done on them. It is proposed
-that the index of a list item be of class OBJECT rather than of class
-STRING.
+that the index of a list item be of class \texttt{OBJECT} rather than of
+class \texttt{STRING}.
 
-In OOI, the .ENVIRONMENT is global, not read-only, and contains builtin
-objects such as .TRUE and .FALSE. The committee regards this as too
-risky - suppose that .TRUE was accidentally or maliciously revalued as
-0! It seems sufficient to add read-only as a characteristic of
-directories. (This characteristic at the element level might be
-expensive to implement.) Reserved symbols (X8-274 clause 6.2.3.1) also
-provide a mechanism for preventing the override of builtin names. It
-won't be possible for a standard to exactly define in a
-system-independent way the scopes/lifetimes of -ENVIRONMENT and .LOCAL
-but (as with OOl) the .LOCAL will relate to ``One API\_START'' and
--ENVIRONMENT will have a wider scope. (Power on to power off of some
-system'). The proposed ``search order'' is:
+In OOI, the \texttt{.ENVIRONMENT} is global, not read-only, and contains
+builtin objects such as \texttt{.TRUE} and \texttt{.FALSE}. The
+committee regards this as too risky - suppose that \texttt{.TRUE} was
+accidentally or maliciously revalued as 0!
+
+It seems sufficient to add read-only as a characteristic of directories.
+(This characteristic at the element level might be expensive to
+implement.)
+
+Reserved symbols (X8-274 clause 6.2.3.1) also provide a mechanism for
+preventing the override of builtin names.
+
+It won't be possible for a standard to exactly define in a
+system-independent way the scopes/lifetimes of \texttt{.ENVIRONMENT} and
+\texttt{.LOCAL} but (as with OOI) the \texttt{.LOCAL} will relate to
+``One \texttt{API\_START}'' and \texttt{.ENVIRONMENT} will have a wider
+scope. (Power on to power off of some system?).
+
+The proposed ``search order'' is:
 
@@
   Things provided by the system which no user is expected to want to
-  override. Perhaps .TRUE .FALSE NIL.
+  override. Perhaps \texttt{.TRUE} \texttt{.FALSE} \texttt{NIL.}
 \item
-  The .LOCAL read/write directory, initialized with the default streams,
-  changable by the user for individual program executions. Perhaps
-  METHODS here.
+  The \texttt{.LOCAL} read/write directory, initialized with the default
+  streams, changable by the user for individual program executions.
+  Perhaps \texttt{METHODS} here.
 \item
   The read-only part of the environment, that is the builtin classes and
-  objects. Also .SYSTEM perhaps.
+  objects. Also \texttt{.SYSTEM} perhaps.
 \item
-  The read/write ENVIRONMENT directory. Changable by programmers
-  co-operating at the system level. Final placement of all builtins
-  needs discussion, but the read-only true\&false requirement will be
-  met. Note that the algorithm of method lookup does not change if ``old
-  objects see newest methods'' is desired. What changes is whether the
-  method tables are updated in place or copied-and-updated when they are
-  changed.
-\item
-  There have been sugestions to allow the REQUIRES directive appear in
-  more places. The committee agrees with this and proposes:
+  The read/write \texttt{.ENVIRONMENT} directory. Changable by
+  programmers co-operating at the system level.
 \end{enumerate}
 
-\begin{enumerate}
-\def\labelenumi{\Alph{enumi})}
-\tightlist
-\item
-  All REQUIRES directives must appear together in the file. B) These
-  directives may appear anywhere the OOI implementation currently allows
-  them to appear.
-\end{enumerate}
+Final placement of all builtins needs discussion, but the read-only
+true\&false requirement will be met. Note that the algorithm of method
+lookup does not change if ``old objects see newest methods'' is desired.
+What changes is whether the method tables are updated in place or
+copied-and-updated when they are changed.
 
@@
 \def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
+\item
+  There have been sugestions to allow the \texttt{REQUIRES} directive
+  appear in more places. The committee agrees with this and proposes:
+
+  \begin{enumerate}
+  \def\labelenumii{\Alph{enumii})}
+  \item
+    All \texttt{REQUIRES} directives must appear together in the file.
+  \item
+    These directives may appear anywhere the OOI implementation
+    currently allows them to appear.
+  \end{enumerate}
 \item
@@
 \item
-  Proposed language, eg FORWARD, METHOD, and CLASS clauses, allow for
-  many options which can appear in any order. These can be written in
-  the BNF (in the manner that TO BY FOR were handled in the first
-  standard) but it is neater to extend the BNF metalanguage.
+  Proposed language, eg \texttt{FORWARD}, \texttt{METHOD}, and
+  \texttt{CLASS} clauses, allow for many options which can appear in any
+  order. These can be written in the BNF (in the manner that \texttt{TO}
+  \texttt{BY} \texttt{FOR} were handled in the first standard) but it is
+  neater to extend the BNF metalanguage.
 \item
-  The OOI syntax used in the FORWARD instruction has examples of the
-  `argument' construct, which is either a symbol-or-string taken as a
-  constant or is an expression in parentheses. The committee will define
-  `term' to be allowed in such places. This is a change to the OOI for
-  valid programs only in the case where a MESSAGE option used a symbol
-  intending it to be `taken as a constant'. (As opposed to taken as a
-  variable with the value defaulting to its name when uninitialized.)
+  The OOI syntax used in the \texttt{FORWARD} instruction has examples
+  of the `argument' construct, which is either a symbol-or-string taken
+  as a constant or is an expression in parentheses. The committee will
+  define `term' to be allowed in such places. This is a change to the
+  OOI for valid programs only in the case where a \texttt{MESSAGE}
+  option used a symbol intending it to be `taken as a constant'. (As
+  opposed to taken as a variable with the value defaulting to its name
+  when uninitialized.)
 \item
@@
   for OO, to:
-\end{enumerate}
 
@@
   \end{itemize}
+\end{enumerate}
 
@@
 applied to local pools. (Local pools may access non-local pool items by
-reason of EXPOSE.) In principle this leads to different threads of
-execution (resulting from REPLY) being able to execute the API. (In
-practice OOI has a restriction to executing the API only on the `main'
-thread and the committee needs to know if this is due to a generally
-applicable difficulty.)
+reason of \texttt{EXPOSE}.)
+
+In principle this leads to different threads of execution (resulting
+from \texttt{REPLY}) being able to execute the API. (In practice OOI has
+a restriction to executing the API only on the `main' thread and the
+committee needs to know if this is due to a generally applicable
+difficulty.)
 
@@
 
-PEEK on queue unnecessary - same as AT{[}1{]}?
+\texttt{PEEK} on queue unnecessary - same as \texttt{AT{[}1{]}}?
 
-Also need to resolve the issues on Monitor class and on run time
-inspection.
+\emph{Also need to resolve the issues on \texttt{Monitor} class and on
+run time inspection.}

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/built-in_functions.tex
updates

--- 2024-02-16 ARB/standards/work-in-progress/document/tex/standard/built-in_functions.tex
@@
 
-/* Check arguments. Some further checks will be made in particular
-built-ins.\emph{/ /} The argument to CheckArgs is a checklist for the
-allowable arguments. */
-
-/* NUM, WHOLENUM and WHOLE have a side-effect, `normalizing' the number.
-*/
-
-/* Calls to raise syntax conditions will not return. */
-
-CheckArgs: CheckList = arg(1) /* This refers to the argument of
-CheckArgs. */
-
-/* Move the checklist information from a string to individual variables
-\emph{/ ArgType. = '\,'\,' ArgPos = 0 /} To count arguments \emph{/
-MinArgs = 0 do j = 1 to length (CheckList) ArgPos = ArgPos+1 /} Count
-the required arguments. \emph{/ if substr(CheckList,j,1) == `r' then
-MinArgs = MinArgs + 1 /} Collect type information. \emph{/ do while j
-\textless{} length(CheckList) j=eajae1 t = substr(CheckList,j,1) if t=='
-' then leave ArgType.ArgPos = ArgType.ArgPos \textbar\textbar{} t end /}
-A single space delimits parts. */ end j MaxArgs = ArgPos
-
-/* Check the number of arguments to the built-in, in this instance. */
-NumArgs = \#Bif Arg.0
-
-if NumArgs \textless{} MinArgs then call Raise 40.3, MinArgs
-
-if NumArgs \textgreater{} MaxArgs then call Raise 40.4, MaxArgs
-
-/* Check the type(s) of the arguments to the built-in. */ do ArgPos = 1
-to NumArgs if \#Bif ArgExists.ArgPos then call CheckType else if ArgPos
-\textless= MinArgs then call Raise 40.5, ArgPos end ArgPos
-
-/* No errors found by CheckArgs. */
-
-return CheckType: Value = \#Bif Arg.ArgPos Type = ArgType.ArgPos select
-when Type == `ANY' then nop /* Any string \emph{/ when Type == `NUM'
-then do /} Any number */
-
-/* This check is made with the caller's digits setting. \emph{/ if
-\Cdatatype(Value, `N') then if \#DatatypeResult==`E' then call Raise
-40.9, ArgPos, Value else call Raise 40.11, ArgPos, Value \#Bif
-Arg.ArgPos=\#DatatypeResult /} Update argument copy. */ end
-
-when Type == `WHOLE' then do /* Whole number \emph{/ /} This check is
-made with digits setting for the built-in. */ if \Edatatype(Value,`W')
-then call Raise 40.12, ArgPos, Value \#Bif\_ Arg.ArgPos=\#DatatypeResult
-end
-
-when Type == `WHOLE\textgreater=0' then do /* Non-negative whole number
-*/ if \Edatatype(Value,`W') then call Raise 40.12, ArgPos, Value if
-\#DatatypeResult \textless{} 0 then call Raise 40.13, ArgPos, Value
-\#Bif\_ Arg.ArgPos=\#DatatypeResult end
-
-when Type == `WHOLE\textgreater0O' then do /* Positive whole number */
-if \Edatatype(Value,`W') then call Raise 40.12, ArgPos, Value if
-\#DatatypeResult \textless= 0 then call Raise 40.14, ArgPos, Value
-\#Bif\_ Arg.ArgPos=\#DatatypeResult end
-
-when Type == `WHOLENUM' then do /* D2X type whole number \emph{/ /} This
-check is made with digits setting of the caller. */ if
-\Cdatatype(Value,`W') then call Raise 40.12, ArgPos, Value \#Bif\_
-Arg.ArgPos=\#DatatypeResult end
-
-when Type == `WHOLENUM\textgreater=0' then do /* D2X Non-negative whole
-number */ if \Cdatatype(Value,`W') then call Raise 40.12, ArgPos, Value
-if \#DatatypeResult \textless{} 0 then call Raise 40.13, ArgPos, Value
-\#Bif\_ Arg.ArgPos=\#DatatypeResult end
-
-when Type == `0 90' then do /* Errortext */
-
-if \Edatatype(Value,`N') then call Raise 40.11, ArgPos, Value
-
-Value=\#DatatypeResult
-
-\#Bif\_ Arg.ArgPos=Value
-
-Major=Value \% 1
-
-Minor=Value - Major
-
-if Major \textless{} 0 \textbar{} Major \textgreater{} 90 \textbar{}
-Minor \textgreater{} .9 \textbar{} pos(`E',Value)\textgreater0 then call
-Raise 40.17, Value /* ArgPos will be 1 */
-
-end
-
-when Type == `PAD' then do /* Single character, usually a pad. */
-
-if length(Value) = 1 then call Raise 40.23, ArgPos, Value end
-
-when Type == `HEX' then /* Hexadecimal string \emph{/ if
-\datatype (Value, `X') then call Raise 40.25, Value /} ArgPos will be 1
-\emph{/ when Type == `BIN' then /} Binary string \emph{/ if
-\datatype(Value,`B') then call Raise 40.24, Value /} ArgPos will be 1
-\emph{/ when Type == `SYM' then /} Symbol */
-
-if \datatype (Value, `S') then call Raise 40.26, Value /* ArgPos will be
-1 */
-
-when Type == `STREAM' then do call Config Stream Qualify Value if left
-(\#Response, 1) == `B' then call Raise 40.27, Value /* ArgPos will be 1
-*/ end
-
-when Type = `ACEFILNOR' then do /* Trace \emph{/ Val = Value /} Allow
-`?' alone */
-
-if val == `?' then do /* Allow leading `?' */ if left(Val,1) == `?' then
-Val = substr(Val,2)
-
-if pos(translate(left(Val, 1)), `ACEFILNOR') = 0 then call Raise 40.28,
-ArgPos, Type, Val
-
-end end otherwise do /* Options \emph{/ /} The checklist item is a list
-of allowed characters */ if Value == '\,' then
-
-call Raise 40.21, ArgPos \#Bif Arg.ArgPos = translate(left(Value, 1)) if
-pos(\#Bif Arg.ArgPos, Type) = 0 then call Raise 40.28, ArgPos, Type,
-Value end
-
-end /* Select */ return
-
-Cdatatype: /* This check is made with the digits setting of the caller.
-\emph{/ /} \#DatatypeResult will be set by use of datatype() */
-
-numeric digits \#Digits.\#Level
-
-numeric form value \#Form.\#Level
-
-return datatype(arg(1), arg(2))
-
-Edatatype: /* This check is made with digits setting for the particular
-built-in. \emph{/ /} \#DatatypeResult will be set by use of datatype()
-*/
-
-numeric digits \#Bif Digits.\#Bif
-
-numeric form scientific
-
-return datatype(arg(1),arg(2))
-
-10.2.2 Date calculations Time2Date: if arg(1) \textless{} 0 then call
-Raise 40.18 if arg(1) \textgreater= 315537897600000000 then call Raise
-40.18 return Time2Date2 (arg(1))
-
-Time: procedure /* This routine is essentially the code from the
-standard, put in stand-alone form. The only `tricky bit' is that there
-is no Rexx way for it to fail with the same error codes as a ``real''
-implementation would. It can however give a SYNTAX error, albeit not the
-desirable one. This causing of an error is done by returning with no
-value. Since the routine will have been called as a function, this
-produces an error. */
-
-/* Backslash is avoided as some systems don't handle that negation sign.
-\emph{/ if argQ)\textgreater3 then returm numeric digits 18 if
-arg(1,`E') then if pos(translate(left(arg(1),1)),``CEHLMNRS'')=0 then
-return /} (The standard would also allow `O' but what this code is
-running on would not.) \emph{/ if arg(3,`E') then if
-pos(translate(left(arg(3),1)),``CHLMNS'')=0 then return /} Tf the third
-argument is given then the second is mandatory. \emph{/ if arg(3,`E') \&
-arg(2,`E')=0 then returm /} Default the first argument. \emph{/ if
-arg(1,`E') then Option = translate(left(arg(1),1)) else Option = `N'' /}
-Tf there is no second argument, the current time is returned. \emph{/ if
-arg(2,`E') = 0 then if arg(1,`E') then return ``TIME'(arg(1)) else
-return `TIME'() /} One cannot convert to elapsed times. */ if
-pos(Option, `ERO') \textgreater{} 0 then returm InValue = arg(2) if
-arg(3,`E') then InOption = arg(3) else InOption = `N' HH =0 MM = 0 SS =0
-HourAdjust = 0 select
-
-when InOption == `C' then do parse var InValue HH `:'. +1 MM +2 XX if HH
-= 12 then HH =0 if XX == `pm' then HourAdjust = 12
-
-end when InOption == `H' then HH = InValue
-
-when InOption == `L'' \textbar{} InOption == `N' then parse var InValue
-HH `:' MM `:' SS when InOption == `M' then MM = InValue otherwise SS =
-InValue end if datatype(HH,`W')=0 \textbar{} datatype(MM,`W')=0
-\textbar{} datatype(SS,`N')=0 then returm HH = HH + HourAdjust /*
-Convert to microseconds \emph{/ Micro = trunc((((HH } 60) + MM) * 60 +
-SS) * 1000000) /* There is no special message for time-out-of-range; the
-bad-format message is used. \emph{/ if Micro\textless0O \textbar{} Micro
-\textgreater{} 24}3600* 1000000 then returm /* Reconvert to further
-check the original. */ if TimeFormat(Micro,InOption) == InValue then
-return TimeFormat(Micro, Option) return
-
-TimeFormat: procedure /* Convert from microseconds to given format.
-\emph{/ /} The day will be irrelevant; actually it will be the first day
-possible. */ x = Time2Date2(arg(1)) parse value x with Year Month Day
-Hour Minute Second Microsecond Base Days select when arg(2) == `C' then
-select when Hour\textgreater12 then return
-Hour-12':`right(Minute,2,'0'')'pm' when Hour=12 then return
-`12:'right(Minute,2,'0')`pm' when Hour\textgreater0 then return
-Hour':`right(Minute,2,'0')`am' when Hour=0 then return
-`12:'right(Minute,2,'0')`am'
-
-when arg(2) == `H' then return Hour when arg(2) == `L' then return
-right(Hour,?2,`0')`:'right(Minute,2,'0')`:'right(Second,2,'0'),
-\textbar\textbar{} `.'right(Microsecond,6,'0') when arg(2) == `M' then
-return 60\emph{Hour+Minute when arg(2) == `N' then return
-right(Hour,?2,`0')`:'right(Minute,2,'0')`:'right(Second,2,'0') otherwise
-/} arg(2) ==`S' \emph{/ return 3600}Hour+60* Minute+Second end
-
-Time2Date2: Procedure /* Convert a timestamp to a date. Argument is a
-timestamp (the number of microseconds relative to 0001 01 01
-00:00:00.000000) Returns a date in the form: year month day hour minute
-second microsecond base days */
-
-/* Argument is relative to the virtual date 0001 01 01 00:00:00.000000
-*/ Time = arg(1)
-
-Second = Time \% 1000000 \} Microsecond = Time // 1000000 Minute =
-Second \% 60 ; Second = Second // 60 Hour = Minute \% 60 ; Minute =
-Minute // 60 Day = Hour \% 24 ;\} Hour = Hour // 24
-
-/* At this point, the days are the days since the 0001 base date. */
-BaseDays = Day Day = Day + 1
-
-/* Compute either the fitting year, or some year not too far earlier.
-Compute the number of days left on the first of January of this year. */
-
-Year = Day \% 366 Day = Day - (Year*365 + Year\%4 - Year\%100 +
-Year\%400) Year = Year +1
-
-/* Now if the number of days left is larger than the number of days in
-the year we computed, increment the year, and decrement the number of
-days accordingly. */ do while Day \textgreater{} (365 + Leap (Year) )
-
-Day = Day - (365 + Leap(Year) )
-
-Year = Year + 1 end
-
-/* At this point, the days left pertain to this year. */ YearDays = Day
-
-/* Now step through the months, increment the number of the month, and
-decrement the number of days accordingly (taking into consideration that
-in a leap year February has 29 days), until further reducing the number
-of days and incrementing the month would lead to a negative number of
-days */ Days = `31 28 31 30 31 30 31 31 30 31 30 31' do Month = 1 to
-words (Days)
-
-ThisMonth = Word(Days, Month) + (Month = 2) * Leap (Year)
-
-if Day \textless= ThisMonth then leave
-
-Day = Day - ThisMonth end
-
-return Year Month Day Hour Minute Second Microsecond BaseDays YearDays
-
-Leap: procedure /* Return 1 if the year given as argument is a leap
-year, or 0 otherwise. */ return (arg(1)//4 = 0) \& ((arg(1)//100
-\textless\textgreater{} 0) \textbar{} (arg(1)//400 = 0))
+\lstinputlisting[language=rexx,label=argumentchecking.rexx,caption=argumentchecking.rexx]{argumentchecking.rexx}
+
+10.2.2 Date calculations
+
+\lstinputlisting[language=rexx,label=datecalculations.rexx,caption=datecalculations.rexx]{datecalculations.rexx}
 
@@
 
-ReRadix: /* Converts Arg(1) from radix Arg(2) to radix Arg(3) */
-
-procedure
-
-Subj ect=arg(1)
-
-FromRadix=arg (2)
-
-ToRadix=arg (3)
-
-/* Radix range is 2-16. Conversion is via decimal */
-
-Integer=0
-
-do j=1 to length (Subject) /* Individual digits have already been
-checked for range. \emph{/ Integer=Integer}FromRadix+pos (substr
-(Subject,j,1),`0123456789ABCDEF')-1 end
-
-rete
-
-do while Integer\textgreater0
-
-r= substr(`0123456789ABCDEF',1 + Integer // ToRadix, 1)
-\textbar\textbar{} r Integer = Integer \% ToRadix end
-
-/* When between 2 and 16, there is no zero suppression. \emph{/ if
-FromRadix = 2 \& ToRadix = 16 then r=eright(r, (length(Subject)+3) \% 4,
-`0') else if FromRadix = 16 \& ToRadix = 2 then reright(r,
-length(Subject) } 4, `0') return r
+\lstinputlisting[language=rexx,label=reradix.rexx,caption=reradix.rexx]{reradix.rexx}
 
@@
 
-Raise:
-
-/* These 40.nn messages always include the built-in name as an
-insert.\emph{/ call \#Raise `SYNTAX', arg(1), \#Bif, arg(2), arg(3),
-arg(4) /} \#Raise does not return. */
+\lstinputlisting[language=rexx,label=raise.rexx,caption=raise.rexx]{raise.rexx}
 
@@
 
-ABBREV returns `1' if the second argument is equal to the leading
-characters of the first and the length of the second argument is not
-less than the third argument.
-
-call CheckArgs `rANY rANY oWHOLE\textgreater=0'
-
-Subject \#Bif Arg.1
-
-Subj \#Bif Arg.2
-
-if \#Bif\_ArgExists.3 then Length = \#Bif Arg.3 else Length = length
-(Subj)
-
-Condl = length(Subject) \textgreater= length(Subj) Cond2 = length(Subj)
-\textgreater= Length Cond3 = substr(Subject, 1, length(Subj)) == Subj
-
-return Condl \& Cond2 \& Cond3
+\lstinputlisting[language=rexx,label=abbrev.rexx,caption=abbrev.rexx]{abbrev.rexx}
 
@@
 
-call CheckArgs `rANY rWHOLE\textgreater=0 oPAD'
-
-String = \#Bif Arg.1
-
-Length = \#Bif Arg.2
-
-if \#Bif\_ArgExists.3 then Pad = \#Bif Arg.3 else Pad = ' '!
-
-Trim = length(String) - Length
-
-if Trim \textgreater{} 0 then return substr(String, Trim \% 2 + 1,
-Length)
-
-return overlay(String, copies(Pad, Length), -Trim \% 2 + 1)
+\lstinputlisting[language=rexx,label=center.rexx,caption=center.rexx]{center.rexx}
 
@@
 
-call CheckArgs ```rANY rANY rANY'
-
-Output = '\,' Position = 1 do forever FoundPos = pos(\#Bif Arg.1, \#Bif
-Arg.2, Position) if FoundPos = 0 then leave Output = Output
-\textbar\textbar{} substr(\#Bif\_ Arg.2, Position, FoundPos - Position),
-\textbar\textbar{} \#Bif Arg.3 Position = FoundPos + length(\#Bif Arg.1)
-end return Output \textbar\textbar{} substr(\#Bif Arg.2, Position)
+\lstinputlisting[language=rexx,label=changestr.rexx,caption=changestr.rexx]{changestr.rexx}
 
@@
 
-call CheckArgs `rANY rANY oPAD'
-
-Strl = \#Bif\_Arg.1
-
-Str2 = \#Bif\_Arg.2
-
-if \#Bif\_ArgExists.3 then Pad else Pad
-
-\#Bif Arg.3
-
-/* Compare the strings from left to right one character at a time */ if
-length(Str1) \textgreater{} length(Str2) then do
-
-Length = length(Str1)
-
-Str2=left (Str2, Length, Pad)
-
-end else do
-
-Length = length(Str2)
-
-Strl=left (Str1, Length, Pad)
-
-end
-
-do i= 1 to Length if substr(Stri, i, 1) == substr(Str2, i, 1) then
-return i end
-
-return 0
+\lstinputlisting[language=rexx,label=compare.rexx,caption=compare.rexx]{compare.rexx}
 
@@
 
-call CheckArgs ```rANY rWHOLE\textgreater=0' Output = '\,' do \#Bif
-Arg.2 Output = Output \textbar\textbar{} \#Bif\_Arg.1 end
-
-return Output
+\lstinputlisting[language=rexx,label=copies.rexx,caption=copies.rexx]{copies.rexx}
 
@@
 
-call CheckArgs ```rANY rANY'
-
-Output = 0
-
-Position = pos (\#Bif Arg.1,\#Bif Arg.2)
-
-do while Position \textgreater{} 0 Output = Output + 1 Position =
-pos(\#Bif Arg.1, \#Bif Arg.2, Position + length(\#Bif Arg.1)) end
-
-return Output
+\lstinputlisting[language=rexx,label=counstr.rexx,caption=counstr.rexx]{counstr.rexx}
 
@@
 
-call CheckArgs `rANY oABLMNSUWX'
-
-/* As well as returning the type, the value for a `NUM' is set in
-\#DatatypeResult. This is a convenience when DATATYPE is used
-
-by CHECKARGS. */
-
-String = \#Bif Arg.1
-
-/* If no second argument, DATATYPE checks whether the first is a number.
-*/ if \#Bif\_ArgExists.2 then return DtypeOne()
-
-Type = \#Bif\_ Arg.2 /* Null strings are a special case. */
-
-if String == '\,' then do if Type == ``X'' then return 1 if Type ==
-``B'' then return 1 return 0 end
-
-/* Several of the options are shorthands for VERIFY \emph{/
-azl=``abcdefghijklmnopqrstuvwxyz'' AZU= ``ABCDEFGHIJKLMNOPQRSTUVWXYZ''
-DO09=``0123456789'' if Type == ``A'' then return
-verify(String,az1\textbar\textbar AzU\textbar\textbar D09)=0 if Type ==
-``B'' then do /} Check blanks in allowed places. \emph{/ if pos (left
-(String,1),\#Al1Blanks)\textgreater0 then return 0 if pos (right
-(String,1),\#Al1Blanks)\textgreater0 then return 0 BinaryDigits=0 do j =
-length(String) by -1 tol ce = substr(String,j,1) if
-pos(c,\#Al11Blanks)\textgreater0 then do /} Blanks need four
-BinaryDigits to the right of them. */ if BinaryDigits//4 = 0 then return
-0 end else do if verify(c,``01'') = 0 then return 0 BinaryDigits =
-BinaryDigits + 1
-
-end end j
-
-return 1
-
-end /* B */ if Type == ``L'' then return (verify (String,azl1)=0) if
-Type == ``M'' then return (verify (String, azl1\textbar\textbar AZU)=0)
-if Type == ``N'' then return (datatype (String) ==``NUM'') if Type ==
-``S'' then return(symbol (String) ==`BAD') if Type == ``U'' then return
-(verify (String, AZU)=0) if Type == ``W'' then do
-
-/* It may not be a number. \emph{/ if DtypeOne(String) == `CHAR' then
-return `0' /} It can be ``Whole'' even if originally in exponential
-notation, provided it can be written as non-exponential. \emph{/ if
-pos(`E',\#DatatypeResult)\textgreater0 then return `0' /} It won't be
-``Whole'' if there is a non-zero after the decimal point. */
-InFraction=`0'
-
-do j = 1 to length (String) ce = substr(String,j,1) if
-pos(c,`Ee')\textgreater0 then leave j if InFraction \&
-pos(c,`+-')\textgreater0 then leave j if c == `.' then InFraction=`1'
-else if InFraction \& a==`0' then return 0 end j /* All tests for Whole
-passed. \emph{/ \#DatatypeResult = \#DatatypeResult \% 1 return 1 end /}
-W \emph{/ /} Type will be ``x'' \emph{/ if pos (left
-(String,1),\#Al1Blanks)\textgreater0 then return 0 if pos (right
-(String,1),\#Al1Blanks)\textgreater0 then return 0 HexDigits=0 do
-j=length(String) by -1 to 1 c=substr (String,j,1) if
-pos(c,\#Al11Blanks)\textgreater0 then do /} Blanks need a pair of
-HexDigits to the right of them. \emph{/ if HexDigits//2 = 0 then return
-0 end else do if verify(c,``abcdefABCDEF''D09) = 0 then return 0
-HexDigits=HexDigits+1 end end return 1 /} end X */
-
-DtypeOne: /* See section nnn for the syntax of a number. \emph{/
-\#DatatypeResult = `S' /} If not syntactically a number \emph{/ Residue
-= strip(String) /} Blanks are allowed at both ends. */ if Residue ==
-'\,' then return ``CHAR'' Sign = '\,' if left(Residue,1) == `+'
-\textbar{} left(Residue,1) == `-' then do
-
-Sign = left(Residue, 1) Residue = strip(substr(Residue,2),`L') /* Blanks
-after sign */
-
-end if Residue == '\,' then return ``CHAR'' /* Now testing Number,
-section nnn */ if left(Residue,1) == `.' then do Residue = substr
-(Residue, 2) Before = '\,' After = DigitRun() if After == '\,' then
-return ``CHAR'' end else do Before = DigitRun() if Before == '\,' then
-return ``CHAR'' if left(Residue,1) == `.' then do Residue = substr
-(Residue, 2) After = DigitRun() end end Exponent = 0 if Residue == '\,'
-then do if left(Residue, 1) == `e' \& left(Residue, 1) == `E' then
-
-return ``CHAR'' Residue = substr (Residue, 2)
-
-if Residue == '\,' then return ``CHAR''
-
-Esign = '\,'
-
-if left(Residue, 1) == `+' \textbar{} left(Residue, 1) == `-' then do
-Esign = left(Residue, 1) Residue = substr (Residue, 2) if Residue ==
-'\,' then return ``CHAR'' end
-
-Exponent = DigitRun()
-
-if Exponent == '\,' then return ``CHAR''
-
-Exponent = Esign \textbar\textbar{} Exponent
-
-end
-
-if Residue == '\,' then return ``CHAR''
-
-102 /\emph{DATATYPE tests for exponent out of range. }/
-
-\#DatatypeResult = `E' /* If exponent out of range */ Before =
-strip(Before,`L',`0') if Before == '\,' then Before = `0'
-
-Exponent = Exponent + length(Before) -1 /* For SCIENTIFIC */
-
-/* ``Engineering notation causes powers of ten to expressed as a
-multiple of 3 - the integer part may therefore range from 1 through
-9910.'' \emph{/ g=1 if \#Form.\#Level == `E' then do /} Adjustment to
-make exponent a multiple of 3 */ g = Exponent//3 if g \textless{} 0 then
-g = g + Exponent = Exponent - end
-
-/* Check on the exponent. */ if Exponent \textgreater{} \#Limit
-ExponentDigits then return ``CHAR'' if -\#Limit ExponentDigits
-\textgreater{} Exponent then return ``CHAR''
-
-/* Format to the numeric setting of the caller of DATATYPE */ numeric
-digits \#Digits.\#Level
-
-numeric form value \#Form.\#Level
-
-\#DatatypeResult = 0 + \#Bif\_Arg.1
-
-return ``NUM''
-
-DigitRun: Outcome = '\,' do while Residue == '\,' if pos(left (Residue,
-1), `0123456789') = 0 then leave
-
-Outcome = Outcome \textbar\textbar{} left(Residue, 1) Residue =
-substr(Residue, 2) end
-
-return Outcome
+\lstinputlisting[language=rexx,label=datatype.rexx,caption=datatype.rexx]{datatype.rexx}
 
@@
 
-call CheckArgs `rANY rWHOLE\textgreater0 oWHOLE\textgreater=0' String
-\#Bif Arg.1
-
-Num \#Bif Arg.2 if \#Bif\_ArgExists.3 then Len = \#Bif\_Arg.3
-
-if Num \textgreater{} length(String) then return String
-
-Output = substr(String, 1, Num - 1) if \#Bif\_ArgExists.3 then if Num +
-Len \textless= length(String) then Output = Output \textbar\textbar{}
-substr(String, Num + Len) return Output
+\lstinputlisting[language=rexx,label=delstr.rexx,caption=delstr.rexx]{delstr.rexx}
 
@@
 
-call CheckArgs `rANY rWHOLE\textgreater0 oWHOLE\textgreater=0' String
-\#Bif Arg.1
-
-Num \#Bif Arg.2 if \#Bif\_ArgExists.3 then Len = \#Bif\_Arg.3
-
-if Num \textgreater{} words(String) then return String
-
-EndLeft = wordindex(String, Num) - 1 Output = left(String, EndLeft) if
-\#Bif\_ArgExists.3 then do BeginRight = wordindex(String, Num + Len)
-
-if BeginRight\textgreater0 then Output =
-
-end return Output
+\lstinputlisting[language=rexx,label=delword.rexx,caption=delword.rexx]{delword.rexx}
 
@@
 
-call CheckArgs
-
-New \#Bif\_ Arg.1 Target \#Bif Arg.2 if \#Bif\_ArgExists.3
-
-then else then else then else
-
-Num Num Length = \#Bif Arg.4 Length = length (New) Pad \#Bif Arg.5 Pad
-ro
-
-\#Bif Arg.3 0
-
-if \#Bif\_ArgExists.4
-
-if \#Bif\_ArgExists.5
-
-return left(Target, Num, Pad) left (New, Length, Pad), substr(Target,
-Num + 1)
-
-``
+\lstinputlisting[language=rexx,label=insert.rexx,caption=insert.rexx]{insert.rexx}
 
@@
 
-`rANY rANY oWHOLE\textgreater=0 oWHOLE\textgreater=0 oPAD'
-
-\emph{/ }/
-
-LASTPOS returns the position of the last occurrence of the first
-argument within the second. The third
-
-argument is a starting position for the search.
-
-call CheckArgs `rANY rANY oWHOLE\textgreater0'
-
-Needle = \#Bif\_Arg.1 Haystack = \#Bif Arg.2 if \#Bif\_ArgExists.3 then
-Start = \#Bif\_Arg.3 else Start = length(Haystack)
-
-NeedleLength = length (Needle) if NeedleLength = 0 then return 0 Start =
-Start - NeedleLength + 1 do i= Start by -1 while i \textgreater{} 0 if
-substr(Haystack, i, NeedleLength) end i return 0
+\lstinputlisting[language=rexx,label=lastpos.rexx,caption=lastpos.rexx]{lastpos.rexx}
 
@@
 
-length of the result and the third is the padding character. call
-CheckArgs `rANY rWHOLE\textgreater=0 oPAD'
-
-if \#Bif\_ArgExists.3 then Pad else Pad
-
-\#Bif Arg.3
-
-return substr(\#Bif Arg.1, 1, \#Bif\_ Arg.2, Pad)
-
-10.1.14 LENGTH
-
-Needle then return i
-
-The second argument specifies the
-
-Length returns a count of the number of characters in the argument.
-
-call CheckArgs `rANY'
-
-String = \#Bif Arg.1
-
-\#Response = Config Length(String)
-
-Length = \#Outcome
-
-call Config Substr \#Response, 1
-
-if \#Outcome == `E' then return Length
-
-/* Here if argument was not a character string. call Config C2B String
-call \#Raise `SYNTAX',
-
-23.1, b2x(\#Outcome)
-
-\emph{/ /} No return to here */
+length of the result and the third is the padding character.
+
+\lstinputlisting[language=rexx,label=left.rexx,caption=left.rexx]{left.rexx}
 
@@
 OVERLAY overlays the first argument onto the second. The third argument
-is the starting position of the
-
-overlay. The fourth argument is the length of the overlay and the fifth
-is the padding character.
-
-call CheckArgs `rANY rANY oWHOLE\textgreater0 oOWHOLE\textgreater=0
-oPAD'
-
-New = \#Bif\_Arg.1
-
-Target = \#Bif Arg.2
-
-if \#Bif\_ArgExists.3 then Num = \#Bif\_Arg.3 else Num = 1
-
-if \#Bif\_ArgExists.4 then Length = \#Bif\_ Arg.4 else Length = length
-(New)
-
-if \#Bif\_ArgExists.5 then Pad = \#Bif\_ Arg.5 else Pad = ' ' return
-left(Target, Num - 1, Pad), /* To left of overlay \emph{/
-\textbar\textbar{} left (New, Length, Pad), /} New string overlaid
-\emph{/ \textbar\textbar{} substr (Target, Num + Length) /} To right of
-overlay */
-
-10.1.16 POS POS returns the position of the first argument within the
-second.
-
-call CheckArgs `rANY rANY oWHOLE\textgreater0'
-
-Needle \#Bif Arg.1
-
-Haystack = \#Bif Arg.2
-
-if \#Bif\_ ArgExists.3 then Start else Start
-
-\#Bif Arg.3 1
-
-if length(Needle) = 0 then return 0
-
-do i = Start to length (Haystack) +1-length (Needle) if substr(Haystack,
-i, length(Needle)) == Needle then return i end i
-
-return 0
+is the starting position of the overlay. The fourth argument is the
+length of the overlay and the fifth is the padding character.
+
+\lstinputlisting[language=rexx,label=overlay.rexx,caption=overlay.rexx]{overlay.rexx}
 
@@
 
-call CheckArgs `rANY'
-
-String
-
-\#Bif Arg.1 Output = '\,' do i= 1 to length (String) Output =
-substr(String,i,1) \textbar\textbar{} Output end return Output
+\lstinputlisting[language=rexx,label=reverse,caption=reverse]{reverse}
 
@@
 
-length of the result and the third is the padding character. call
-CheckArgs `rANY rWHOLE\textgreater=0 oPAD'
-
-String = \#Bif Arg.1 Length = \#Bif Arg.2 if \#Bif\_ArgExists.3 then Pad
-
-\#Bif Arg.3 else Pad '
-
-Trim = length(String) - Length if Trim \textgreater= 0 then return
-substr(String,Trim + 1) return copies(Pad, -Trim) \textbar\textbar{}
-String /* Pad string on the left */
+length of the result and the third is the padding character.
+
+\lstinputlisting[language=rexx,label=right.rexx,caption=right.rexx]{right.rexx}
 
@@
 
-call CheckArgs `rANY oOWHOLE\textgreater=0 oPAD'
-
-String = \#Bif Arg.1
-
-if \#Bif\_ArgExists.2 then Num = \#Bif\_ Arg.2 else Num = 1 if \#Bif
-ArgExists.3 then Pad = \#Bif Arg.3 else Pad = ' '! Padding = copies(Pad,
-Num) Output = subword(String, 1, 1) do i = 2 to words (String) Output =
-Output \textbar\textbar{} Padding \textbar\textbar{} subword(String, i,
-1)
-
-end return Output
+\lstinputlisting[language=rexx,label=space.rexx,caption=space.rexx]{space.rexx}
 
@@
 
-call CheckArgs `rANY oLTB oPAD'
-
-String = \#Bif Arg.1 if \#Bif\_ArgExists.2 then Option = \#Bif Arg.2
-else Option = `B' if \#Bif\_ ArgExists.3 then Unwanted = \#Bif\_ Arg.3
-else Unwanted = \#A11Blanks\textless Index ``\#Al11Blanks'' \# ``\,''
-\textgreater{}
-
-if Option == `L' \textbar{} Option == `B' then do /* Strip leading
-characters */ do while String == '\,' \& pos(left(String, 1), Unwanted)
-\textgreater{} 0 String = substr(String, 2) end end
-
-if Option == `T' \textbar{} Option == `B' then do /* Strip trailing
-characters \emph{/ do while String == '\,' \& pos(right(String, 1),
-Unwanted) \textgreater{} 0 String = left(String, length(String) -1) end
-/} of while */ end return String
+\lstinputlisting[language=rexx,label=strip.rexx,caption=strip.rexx]{strip.rexx}
 
@@
 the length of the sub-string. The fourth argument is the padding
-
 character.
 
-call CheckArgs `rANY rWHOLE\textgreater0 oOWHOLE\textgreater=0 oPAD'
-
-String = \#Bif Arg.1 Num = \#Bif\_ Arg.2 if \#Bif\_ArgExists.3 then
-Length = \#Bif Arg.3 else Length = max(length (String) +1-Num, 0) if
-\#Bif\_ArgExists.4 then Pad = \#Bif\_ Arg.4 else Pad = ' `! Output =''
-do Length \#Response Config Substr(String,Num) /* Attempt to fetch
-character.*/ Character \#Outcome
-
-Num = Num + 1 call Config Substr \#Response,1 /* Was there such a
-character? */ if \#Outcome == `E' then do
-
-/* Here if argument was not a character string. */
-
-call Config C2B String call \#Raise `SYNTAX', 23.1, b2x(\#Outcome) /* No
-return to here */
-
-end if \#Outcome == `M' then Character = Pad Output=Output \textbar{}
-\textbar{} Character end
-
-return Output
+\lstinputlisting[language=rexx,label=substr.rexx,caption=substr.rexx]{substr.rexx}
 
@@
 
-call CheckArgs `rANY rWHOLE\textgreater0 oWHOLE\textgreater=0'
-
-String \#Bif Arg.1
-
-Num \#Bif Arg.2
-
-if \#Bif\_ ArgExists.3 then Length else Length
-
-\#Bif\_Arg.3 length(String) /* Avoids call \emph{/ /} to WORDS() */
-
-if Length = 0 then return '\,'
-
-/* Find position of first included word */
-
-Start = wordindex (String, Num)
-
-if Start = 0 then return '\,' /* Start is beyond end */
-
-/* Find position of first excluded word */ End = wordindex (String,
-Num+Length) if End = 0 then End = length(String)+1
-
-Output=substr (String, Start, End-Start)
-
-/* Drop trailing blanks */
-
-do while Output == '\,' if pos (right (Output,1),\#A11Blanks) = 0 then
-leave Output = left (Output, length (Output) -1) end
-
-return Output
+\lstinputlisting[language=rexx,label=subword.rexx,caption=subword.rexx]{subword.rexx}
 
@@
 
-call CheckArgs `rANY oANY oANY oPAD' String = \#Bif Arg.1 /* If neither
-input nor output tables, uppercase. */ if \#Bif ArgExists.2 \&
-\#Bif\_ArgExists.3 then do Output = '\,' do j=1 to length (String)
-\#Response = Config Upper (substr (String,j,1))
-
-Output = Output \textbar\textbar{} \#Outcome end j
-
-return Output
-
-end
-
-/* The input table defaults to all characters. \emph{/ if \#Bif
-ArgExists.3 then do \#Response = Config Xrange() Tablei = \#Outcome end
-else Tablei = \#Bif\_Arg.3 /} The output table defaults to null \emph{/
-if \#Bif\_ArgExists.2 then Tableo = \#Bif Arg.2 else Tableo = '\,' /}
-The tables are made the same length */ if \#Bif\_ArgExists.4 then Pad =
-\#Bif\_ Arg.4 else Pad = ' ' Tableo=left (Tableo, length (Tablei) , Pad)
-
-107 Output='\,' do j=1 to length (String) c=substr (String,j,1)
-k=pos(c,Tablei) if k=0 then Output=Output \textbar\textbar c else
-Output=Output \textbar{} \textbar{} substr (Tableo,k,1) end j return
-Output
+\lstinputlisting[language=rexx,label=translate,caption=translate]{translate}
 
@@
 
-call CheckArgs `rANY rANY oMN oWHOLE\textgreater0'
-
-String = \#Bif Arg.1
-
-Reference = \#Bif\_Arg.2
-
-if \#Bif\_ ArgExists.3 then Option \#Bif Arg.3 else Option 'N!
-
-if \#Bif\_ ArgExists.4 then Start = \#Bif\_Arg.4 else Start = 1
-
-Last = length(String) if Start \textgreater{} Last then return 0 if
-Reference == '\,' then if Option == `N' then return Start else return 0
-
-do i = Start to Last t = pos(substr(String, i, 1), Reference) if Option
-== `N' then do if t = 0 then return i /* Return position of NoMatch
-character. \emph{/ end else if t \textgreater{} 0 then return i /}
-Return position of Matched character. */ end i return 0
+\lstinputlisting[language=rexx,label=verify.rexx,caption=verify.rexx]{verify.rexx}
 
@@
 
-call CheckArgs `rANY rwWHOLE\textgreater0'
-
-return subword(\#Bif Arg.1, \#Bif\_Arg.2, 1)
+\lstinputlisting[language=rexx,label=word.rexx,caption=word.rexx]{word.rexx}
 
@@
 
-call CheckArgs `rANY rwWHOLE\textgreater0'
-
-String Num
-
-\#Bif Arg.1 \#Bif Arg.2
-
-/* Find starting position */
-
-Start = 1 Count = 0 do forever Start = verify(String,
-\#Al1Blanks\textless Index ``\#A11Blanks'' \# ``\,'' \textgreater, `N',
-Start) /\emph{ Find non-blank }/ if Start = 0 then return 0 /* Start is
-beyond end \emph{/ Count = Count + 1 /} Words found \emph{/ if Count =
-Num then leave Start = verify(String, \#Al1Blanks\textless Index
-``\#Al11Blanks'' \# ``\,'' \textgreater, `M', Start + 1) /}
-
-Find blank */
-
-if Start = 0 then return 0 /* Start is beyond end */ end return Start
+\lstinputlisting[language=rexx,label=wordindex.rexx,caption=wordindex.rexx]{wordindex.rexx}
 
@@
 
-call CheckArgs `rANY rwWHOLE\textgreater0'
-
-return length(subword(\#Bif Arg.1, \#Bif\_Arg.2, 1))
+\lstinputlisting[language=rexx,label=wordlength.rexx,caption=wordlength.rexx]{wordlength.rexx}
 
@@
 
-call CheckArgs `rANY rANY oWHOLE\textgreater0'
-
-Phrase = \#Bif\_Arg.1
-
-String = \#Bif Arg.2
-
-if \#Bif\_ArgExists.3 then Start = \#Bif\_Arg.3 else Start = 1
-
-Phrase = space (Phrase) PhraseWords = words (Phrase) if PhraseWords = 0
-then return 0 String = space (String) StringWords = words (String) do
-WordNumber = Start to StringWords - PhraseWords + 1 if Phrase ==
-subword(String, WordNumber, PhraseWords) then return WordNumber end
-WordNumber return 0
+\lstinputlisting[language=rexx,label=wordpos.rexx,caption=wordpos.rexx]{wordpos.rexx}
 
@@
 
-call CheckArgs `rANY'
-
-do Count = 0 by 1 if subword(\#Bif Arg.1, Count + 1) == '\,' then return
-Count end Count
+\lstinputlisting[language=rexx,label=words.rexx,caption=words.rexx]{words.rexx}
 
@@
 
-call CheckArgs `oPAD oPAD'
-
-if \#Bif\_ArgExists.1 then \#Bif\_Arg.1 mr if \#Bif\_ArgExists.2 then
-\#Bif Arg.2 \#Response = Config Xrange(\#Bif Arg.1, \#Bif Arg.2) return
-\#Outcome
+\lstinputlisting[language=rexx,label=xrange.rexx,caption=xrange.rexx]{xrange.rexx}
 
@@
 
-ABS returns the absolute value of its argument.
-
-call CheckArgs `rNUM' Number=\#Bif Arg.1
-
-if left (Number,1) = `-' then Number = substr (Number, 2) return Number
+\lstinputlisting[language=rexx,label=abs.rexx,caption=abs.rexx]{abs.rexx}
 
@@
 
-call CheckArgs, `rNUM OWHOLE\textgreater=0 OWHOLE\textgreater=0
-OWHOLE\textgreater=0 OWHOLE\textgreater=0'
-
-if \#Bif\_ArgExists.2 then Before \#Bif Arg.2 if \#Bif\_ArgExists.3 then
-After \#Bif Arg.3 if \#Bif\_ArgExists.4 then Expp \#Bif Arg.4
-
-if \#Bif\_ArgExists.5 then Expt = \#Bif Arg.5
-
-/* In the simplest case the first is the only argument. */ Number=\#Bif
-Arg.1
-
-if \#Bif\_Arg.0 \textless{} 2 then return Number
-
-/* Dissect the Number. It is in the normal Rexx format. */ parse var
-Number Mantissa `E' Exponent
-
-if Exponent == '\,' then Exponent = 0 Sign = 0 if left (Mantissa,1) ==
-`-' then do Sign = 1 Mantissa = substr(Mantissa, 2) end parse var
-Mantissa Befo `.' Afte
-
-/* Count from the left for the decimal point. */
-
-Point = length (Befo)
-
-/* Sign Mantissa and Exponent now reflect the Number. Befo Afte and
-Point reflect Mantissa. */
-
-/* The fourth and fifth arguments allow for exponential notation.
-\emph{/ /} Decide whether exponential form to be used, setting ShowExp.
-\emph{/ ShowExp = 0 if \#Bif\_ArgExists.4 \#Bif ArgExists.5 then do if
-\#Bif\_ArgExists.5 then Expt = \#Digits.\#Level /} Decide whether
-exponential form to be used. \emph{/ if (Point + Exponent)
-\textgreater{} Expt then ShowExp = 1 /} Digits before rule. \emph{/
-LeftOfPoint = 0 if length(Befo) \textgreater{} 0 then LeftOfPoint = Befo
-/} Value left of the point */
-
-/* Digits after point rule for exponentiation: */
-
-/* Count zeros to right of point. */
-
-ze=O0
-
-do while substr(Afte,z+1,1) == `0' Zeze+41tl end
-
-if LeftOfPoint = 0 \& (z - Exponent) \textgreater{} 5 then ShowExp = 1
-
-/* An extra rule for exponential form: */ if \#Bif\_ArgExists.4 then if
-Expp = 0 then ShowExp = 0
-
-/* Construct the exponential part of the result. */ if ShowExp then do
-
-Exponent = Exponent + ( Point - 1 ) Point = 1 /* As required for
-'SCIENTIFIC! */ if \#Form.\#Level == `ENGINEERING' then do while
-Exponent//3 = 0
-
-Point = Point+1l
-
-Exponent = Exponent-1
-
-end end
-
-if \ShowExp then Point = Point + Exponent end /* Expp or Expt given
-\emph{/ else do /} Even if Expp and Expt are not given, exponential
-notation will be used if the original Number+0 done by CheckArgs led to
-it. */ if Exponent = 0 then do ShowExp = 1
-
-110 111
-
-end end
-
-/* ShowExp now indicates whether to show an exponent, Exponent is its
-value. \emph{/ /} Make this a Number without a point.
-
-Integer = Befo\textbar\textbar Afte
-
-*/
-
-/* Make sure Point position isn't disjoint from Integer. \emph{/ if
-Point\textless1 then do /} Extra zeros on the left. */
-
-Integer = Point = 1 end
-
-copies(`0',1 - Point)
-
-if Point \textgreater{} length(Integer) then Integer =
-left(Integer,Point,`0') /* And maybe on the right.
-
-/* Deal with right of decimal point first since that can affect the
-
-\textbar\textbar{} Integer
-
-left. Ensure the requested number of digits there. Afters =
-length(Integer) -Point
-
-if \#Bif\_ArgExists.3 = 0 then After = /* Make Afters match the
-requested After */
-
-do while Afters \textless{} After Afters = Afters+1 Integer = Integer'0'
-end
-
-if Afters \textgreater{} After then do
-
-/* Round by adding 5 at the right place.
-
-Afters
-
-r=substr (Integer, Point + After + 1,
-
-Integer = left (Integer,
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-\end{enumerate}
-
-Point + After)
-
-if r \textgreater= `5' then Integer = Integer + 1 /* This can leave the
-result zero. If Integer = 0 then Sign = 0
-
-/* The case when rounding makes the integer longer is an awkward
-
-*/
-
-*/
-
-one. The exponent will have to be adjusted. */ if length(Integer)
-\textgreater{} Point + After then do
-
-Point = Point+l end
-
-if ShowExp = 1 then do
-
-Exponent=Exponent + (Point - 1)
-
-Point = 1 /* As required for 'SCIENTIFIC! */
-
-if form() = `ENGINEERING' then do while Exponent//3 = 0
-
-Point = Point+
-
-1
-
-Exponent = Exponent-1
-
-end end
-
-t = Point-length (Integer) if t \textgreater{} 0 then Integer =
-Integer\textbar\textbar copies(`0',t)
-
-end /* Rounded */
-
-/* Right part is final if After \textgreater{} 0 then Afte else Afte
-
-now. */ `.'\textbar{} \textbar substr (Integer, Point+1,After)
-
-/* Now deal with the integer part of the result. Integer = left
-(Integer, Point)
-
-if \#Bif ArgExists.2 =
-
-0 then Before
-
-/* Make Point match Before */ if Point \textgreater{} Before - Sign then
-call Raise
-
-do while Point\textless Before Point = Point+1l Integer = '0'Integer end
-
-40.38,
-
-*/
-
-Point + Sign /* Note default.
-
-2,
-
-/* Find the Sign position and blank leading zeroes.
-
-re ter Triggered = 0
-
-do j = 1 to length (Integer) Digit = substr(Integer,j,1) /* Triggered is
-set when sign inserted or blanking finished. if Triggered = 1 then do
-
-r= r\textbar\textbar Daigit iterate end
-
-*/
-
-/* Note default.
-
-\#Bif Arg.1
-
-*/
-
-*/
-
-*/
-
-*/
-
-\emph{/ /} If before sign insertion point then blank out zero. */
-
-if Digit = `0' then if ats cael = `0' \& j+l\textless length(Integer)
-then do re r ' ' iterate end /* j is the sign insertion point. */ if
-Digit = `0' \& j = length(Integer) then Digit = ' '
-
-if Sign = 1 then Digit = `-' r= xr\textbar\textbar Digit Triggered = 1
-end j Number = r\textbar\textbar Afte
-
-if ShowExp = 1 then do /* Format the exponent. \emph{/ Expart = '\,'
-SignExp = 0 if Exponent\textless0 then do SignExp = 1 Exponent =
--Exponent end /} Make the exponent to the requested width. */ if \#Bif\_
-ArgExists.4 = 0 then Expp = length (Exponent) if length(Exponent)
-\textgreater{} Expp then call Raise 40.38, 4, \#Bif\_ Arg.1
-Exponent=right (Exponent,Expp,`0') if Exponent = 0 then do if \#Bif\_
-ArgExists.4 then Expart = copies(' ',expp+2) end else if SignExp = 0
-then Expart else Expart Number = Number \textbar{} \textbar Expart end
-return Number
-
-`E+' Exponent `E-' Exponent
+\lstinputlisting[language=rexx,label=format.rexx,caption=format.rexx]{format.rexx}
 
@@
 
-if \#Bif\_Arg.0 \textless1 then call Raise 40.3, 1 call CheckArgs
-`rNUM'\textbar\textbar copies(' rNUM', \#Bif Arg.0 - 1)
-
-Max = \#Bif Arg.1
-
-do i = 2 to \#Bif Arg.0 by 1 Next = \#Bif Arg.i if Max \textless{} Next
-then Max = Next end i
-
-return Max
+\lstinputlisting[language=rexx,label=max.rexx,caption=max.rexx]{max.rexx}
 
@@
 
-if \#Bif\_Arg.0 \textless1 then call Raise 40.3, 1 call CheckArgs
-`rNUM'\textbar\textbar copies(' rNUM', \#Bif Arg.0 - 1)
-
-Min = \#Bif Arg.1
-
-do i = 2 to \#Bif Arg.0 by 1 Next = \#Bif Arg.i if Min \textgreater{}
-Next then Min = Next end i
-
-return Min
+\lstinputlisting[language=rexx,label=min.rexx,caption=min.rexx]{min.rexx}
 
@@
 
-call CheckArgs `rNUM'
-
-Number = \#Bif Arg.1
-
-select
-
-when Number \textless{} 0 then Output = -1 when Number = 0 then Output =
-0 when Number \textgreater{} 0 then Output = 1
-
-end return Output
+\lstinputlisting[language=rexx,label=sign.rexx,caption=sign.rexx]{sign.rexx}
 
@@
 
-call CheckArgs `rNUM oWHOLE\textgreater=0'
-
-Number = \#Bif Arg.1 if \#Bif\_ArgExists.2 then Num
-
-\#Bif Arg.2 else Num 0
-
-Integer =(10**Num * Number) \%1 if Num=0 then return Integer
-
-t=length (Integer) -Num if t\textless=0 then return
-`0.'right(Integer,Num,'0') else return insert(`.',Integer,t)
+\lstinputlisting[language=rexx,label=trunc.rexx,caption=trunc.rexx]{trunc.rexx}
 
@@
 
-call CheckArgs `oEINO'
-
-if \#Bif\_ArgExists.1 then Optionl = \#Bif\_Arg.1 else Optionl=`N'
-
-if Optionl == `N' then return \#Env\_Name.ACTIVE. \#Level
-
-Tail = Optionl' .ACTIVE. '\#Level return \#Env\_Position.Tail
-\#Env\_Type.Tail \#Env\_Resource.Tail
+\lstinputlisting[language=rexx,label=address.rexx,caption=address.rexx]{address.rexx}
 
@@
 
-ArgData = `OWHOLE\textgreater0 oENO' if \#Bif\_ ArgExists.2 then ArgData
-= `rWHOLE\textgreater0 rENO' call CheckArgs ArgData
-
-if \#Bif\_ArgExists.1 then return \#Arg.\#Level.0
-
-ArgNum=\#Bif Arg.1
-
-if \#Bif\_ArgExists.2 then return \#Arg.\#Level.ArgNum
-
-if \#Bif\_Arg.2 ==`0O' then return \#ArgExists.\#Level.ArgNum else
-return \#ArgExists.\#Level .ArgNum
+\lstinputlisting[language=rexx,label=arg.rexx,caption=arg.rexx]{arg.rexx}
 
@@
 
-call CheckArgs `oCDEIS'
-
-/* Values are null if this is not following a condition. */ if
-\#Condition.\#Level == '\,' then do \#ConditionDescription.\#Level =
-'\,' \#ConditionExtra.\#Level = '\,' \#ConditionInstruction.\#Level end
-
-Option=\#Bif Arg.1
-
-if Option==`C' then return \#Condition.\#Level
-
-if Option==`D' then return \#ConditionDescription. \#Level if
-Option==`E' then return \#ConditionExtra.\#Level
-
-if Option==`I' then return \#ConditionInstruction. \#Level /* State is
-the current state. */
-
-if \#Condition.\#Level = '\,' then return ``\,'' return
-\#Enabling.\#Condition.\#Level \#\#\# DIGITS
+\lstinputlisting[language=rexx,label=condition.rexx,caption=condition.rexx]{condition.rexx}
+
+\subsection{DIGITS}\label{digits}
 
@@
 
-return \#Digits.\#Level
+\lstinputlisting[language=rexx,label=digits,caption=digits]{digits}
 
@@
 
-call CheckArgs `r0\_90 oSN'
-
-msgcode = \#Bif Arg.1
-
-if \#Bif\_ ArgExists.2 then Option else Option
-
-return \#ErrorText .msgcode
+\lstinputlisting[language=rexx,label=errortext.rexx,caption=errortext.rexx]{errortext.rexx}
 
@@
 
-\#Bif\_Arg.2 tint
-
-call CheckArgs '\,'
-
-return \#Form.\#Level
+\lstinputlisting[language=rexx,label=form.rexx,caption=form.rexx]{form.rexx}
 
@@
 
-call CheckArgs '\,'
-
-return \#Fuzz.\#Level
+\lstinputlisting[language=rexx,label=fuzz.rexx,caption=fuzz.rexx]{fuzz.rexx}
 
@@
 
-call CheckArgs `oWHOLE\textgreater0'
-
-if \#Bif\_ArgExists.1 then return \#SourceLine.0 Num = \#Bif\_Arg.1 if
-Num \textgreater{} \#SourceLine.0O then call Raise 40.34, Num,
-\#SourceLine.0 return \#SourceLine.Num
+\lstinputlisting[language=rexx,label=sourceline.rexx,caption=sourceline.rexx]{sourceline.rexx}
 
@@
 
-call CheckArgs `oACEFILNOR' /* Also checks for `?!' \emph{/ /} With no
-argument, this a simple query. */ Output=\#Tracing.\#Level
-
-if \#Interactive.\#Level then Output = `?'\textbar\textbar Output if
-\#Bif\_ArgExists.1 then return Output
-
-Value=\#Bif Arg.1
-
-\#Interactive.\#Level=0
-
-/* A question mark sets the interactive flag. */ if left(Value,1)==`?'
-then do
-
-\#Interactive.\#Level = 1
-
-Value=substr (Value, 2)
-
-end /* Absence of a letter leaves the setting unchanged. */ if
-Value=='\,' then do
-
-Value=translate (left (Value,1)) if Value==`0O' then
-\#Interactive.\#Level=`0' \#Tracing.\#Level = Value end return Output
+\lstinputlisting[language=rexx,label=trace.rexx,caption=trace.rexx]{trace.rexx}
 
@@
 
-\subsection{B2xX}\label{b2xx}
+\subsection{B2X}\label{b2x}
 
@@
 
-call CheckArgs `rBIN'
-
-String = space (\#Bif Arg.1,0) return ReRadix(String,2,16)
+\lstinputlisting[language=rexx,label=b2x.rexx,caption=b2x.rexx]{b2x.rexx}
 
@@
 
-call CheckArgs `rANY oANY oPAD'
-
-Stringl = \#Bif\_ Arg.1 if \#Bif ArgExists.2 then String2
-
-\#Bif Arg.2 else String2 rr
-
-/* Presence of a pad implies character strings. */ if \#Bif\_ArgExists.3
-then if length(Stringl) \textgreater{} length(String2) then String2=left
-(String2,length(String1l),\#Bif\_Arg.3) else
-Stringl=left(Stringl,length(String2),\#Bif\_Arg.3)
-
-/* Change to manifest bit representation. \emph{/ \#Response=Config
-C2B(String1) String1l=\#Outcome \#Response=Config C2B(String2)
-String2=\#0utcome /} Exchange if necessary to make shorter second. */ if
-length(Stringl)\textless length(String2) then do
-
-t=Stringl
-
-Stringl=String2
-
-String2=t
-
-end
-
-/* Operate on common length of those bit strings. */ r='\,' do j=1 to
-length (String2)
-
-bl=substr (Stringl,j,1)
-
-b2=substr (String2,j,1)
-
-select when \#Bif=`BITAND' then b1=b1\&b2 when \#Bif=`BITOR' then
-bl=b1\textbar b2
-
-115 when \#Bif=`BITXOR' then b1=b1\&\&b2 end r=r\textbar\textbar bl end
-j rer \textbar\textbar{} right (Stringl, length (String1)
--length(String2) )
-
-/* Convert back to encoded characters. */ return x2c (b2x(r))
+\lstinputlisting[language=rexx,label=bitand.rexx,caption=bitand.rexx]{bitand.rexx}
 
@@
 
-call CheckArgs `rANY oWHOLE\textgreater=0' if length(\#Bif Arg.1)=0 then
-return 0
-
-if \#Bif\_ArgExists.2 then do /* Size specified \emph{/ Size =
-\#Bif\_Arg.2 if Size = 0 then return 0 /} Pad will normally be zeros
-\emph{/ t=right (\#Bif Arg.1,Size,left (xrange(),1)) /} Convert to
-manifest bit \emph{/ call Config C2B t /} And then to signed decimal.
-\emph{/ Sign = Left (\#Outcome,1) \#Outcome = substr(\#Outcome, 2)
-t=ReRadix (\#Outcome, 2,10) /} Sign indicates 2s-complement. \emph{/ if
-Sign then t=t-2\textbf{length(\#Outcome) if abs(t) \textgreater{} 10 }
-\#Digits.\#Level - 1 then call Raise 40.35, t return t end /} Size not
-specified. */ call Config C2B \#Bif\_Arg.1 t = ReRadix(\#Outcome, 2,10)
-if t \textgreater{} 10 ** \#Digits.\#Level - 1 then call Raise 40.35, t
-return t
-
-10.4.6 C2Xx C2X performs coded to hexadecimal conversion.
-
-call CheckArgs `rANY' if length(\#Bif Arg.1) = 0 then return '\,'
-
-call Config C2B \#Bif\_Arg.1 return ReRadix (\#Outcome,2,16)
+\lstinputlisting[language=rexx,label=c2d.rexx,caption=c2d.rexx]{c2d.rexx}
+
+10.4.6 C2X C2X performs coded to hexadecimal conversion.
+
+\lstinputlisting[language=rexx,label=c2x.rexx,caption=c2x.rexx]{c2x.rexx}
 
@@
 
-`rWHOLENUM\textgreater=0'! `rWHOLENUM rWHOLE\textgreater=0'
-
-if \#Bif\_ArgExists.2 then ArgData else ArgData call CheckArgs ArgData
-
-/* Convert to manifest binary \emph{/ Subject = abs(\#Bif Arg.1) r =
-ReRadix(Subject,10,2) /} Make length a multiple of 8, as required for
-Config B2C */ Length = length(r) do while Length//8 = 0 Length =
-Length+1 end r= right(r,Length,`0')
-
-/* 2s-complement for negatives. */ if \#Bif\_Arg.1\textless0 then do
-Subject = 2**length(r)-Subject r = ReRadix(Subject,10,2) end /* Convert
-to characters */ \#Response = Config B2C(r) Output = \#Outcome if
-\#Bif\_ArgExists.2 then return Output
-
-/* Adjust the length with appropriate characters. */ if
-\#Bif\_Arg.1\textgreater=0 then return right (Output,
-\#Bif\_Arg.2,left(xrange(),1)) else return right (Output, \#Bif
-Arg.2,right (xrange(),1))
+\lstinputlisting[language=rexx,label=d2c.rexx,caption=d2c.rexx]{d2c.rexx}
 
@@
 
-D2X performs decimal to hexadecimal conversion. if \#Bif\_ArgExists.2
-then ArgData = `rWHOLENUM\textgreater=0' else ArgData = `rWHOLENUM
-rWHOLE\textgreater=0'
-
-call CheckArgs ArgData
-
-/* Convert to manifest hexadecimal \emph{/ Subject = abs(\#Bif Arg.1 ) r
-= ReRadix(Subject,10,16) /} Twos-complement for negatives */ if
-\#Bif\_Arg.1\textless0 then do Subject = 16**length(r) -Subject r =
-ReRadix(Subject,10,16) end if \#Bif\_ArgExists.2 then return r /* Adjust
-the length with appropriate characters. */ if \#Bif\_
-Arg.1\textgreater=0 then return right(r,\#Bif Arg.2,`0') else return
-right(r,\#Bif Arg.2,`F')
+D2X performs decimal to hexadecimal conversion.
+
+\lstinputlisting[language=rexx,label=d2x.rexx,caption=d2x.rexx]{d2x.rexx}
 
@@
 
-call CheckArgs `rHEX'
-
-Subject = \#Bif Arg.1
-
-if Subject == '\,' then return '\,'
-
-/* Blanks were checked by CheckArgs, here they are ignored. */ Subject =
-space (Subject, 0)
-
-return ReRadix(translate (Subject) ,16,2)
+\lstinputlisting[language=rexx,label=x2b.rexx,caption=x2b.rexx]{x2b.rexx}
 
@@
 
-call CheckArgs `rHEX'
-
-Subject = \#Bif Arg.1
-
-if Subject == '\,' then return '\,'
-
-Subject = space (Subject, 0)
-
-/* Convert to manifest binary */
-
-r = ReRadix(translate (Subject) ,16,2)
-
-/* Convert to character */
-
-Length = 8*( (length (Subject) +1) \%2) \#Response = Config
-B2C(right(r,Length,`0')) return \#Outcome
+\lstinputlisting[language=rexx,label=x2c.rexx,caption=x2c.rexx]{x2c.rexx}
 
@@
 
-call CheckArgs `rHEX OWHOLE\textgreater=0'
-
-Subject = \#Bif Arg.1 if Subject == '\,' then return `0'
-
-Subject = translate (space (Subject,0))
-
-if \#Bif ArgExists.2 then
-
-Subject = right (Subject, \#Bif Arg.2,`0') if Subject =='\,' then return
-`0' /* Note the sign */ if \#Bif ArgExists.2 then SignBit
-
-else SignBit
-
-/* Convert to decimal \emph{/ r = ReRadix(Subject,16,10) /}
-Twos-complement */ if SignBit then r = 2**(4*\#Bif Arg.2) - r if
-abs(r)\textgreater10 ** \#Digits.\#Level - 1 then call Raise 40.35, t
-return r
-
-left (x2b (Subject) ,1) ror
+\lstinputlisting[language=rexx,label=x2d.rexx,caption=x2d.rexx]{x2d.rexx}
 
@@
 nnn. When the operations are successful the following characteristics
-shall be exhibited: - The CHARIN/CHAROUT functions are insensitive to
-the lengths of the arguments. The data written to a stream by CHAROUT
-can be read by a different number of CHARINs. - The CHARIN/CHAROUT
-functions are reflective, that is, the concatenation of the data read
-from a persistent stream by CHARIN (after positioning to 1, while
-CHARS(Stream)\textgreater0), will be the same as the concatenation of
-the data put by CHAROUT. - All characters can be used as CHARIN/CHAROUT
-data. - The CHARS(Stream, `N') function will return zero only when a
+shall be exhibited:
+
+\begin{itemize}
+\item
+  The CHARIN/CHAROUT functions are insensitive to the lengths of the
+  arguments. The data written to a stream by CHAROUT can be read by a
+  different number of CHARINs.
+\item
+  The CHARIN/CHAROUT functions are reflective, that is, the
+  concatenation of the data read from a persistent stream by CHARIN
+  (after positioning to 1, while CHARS(Stream)\textgreater0), will be
+  the same as the concatenation of the data put by CHAROUT.
+\item
+  All characters can be used as CHARIN/CHAROUT data.
+\item
+  The CHARS(Stream, `N') function will return zero only when a
   subsequent read (without positioning) is guaranteed to raise the
-NOTREADY condition. - The LINEIN/LINEOUT functions are sensitive to the
-length of the arguments, that is, the length of a line written by
-LINEOUT is the same as the length of the string returned by successful
-LINEIN of the line. - Some characters, call them line-banned characters,
-cannot reliably be used as data for LINEIN/LINEOUT. If these are not
-used, LINEIN/LINEOUT is reflective. If they are used, the result is not
-defined. The set of characters which are line-barred is a property of
-the configuration. - The LINES(Stream, `N') function will return zero
-only when a subsequent LINEIN (without positioning) is guaranteed to
-raise the NOTREADY condition. - When a persistent stream is repositioned
-and written to with CHAROUT, the previously written data is not lost,
-except for the data overwritten by this latest CHAROUT. - When a
-persistent stream is repositioned and written to with LINEOUT, the
-previously written data is not lost, except for the data overwritten by
-this latest LINEOUT, which may leave lines partially overwritten. \#\#\#
-CHARIN CHARIN returns a string read from the stream named by the first
+  NOTREADY condition.
+\item
+  The LINEIN/LINEOUT functions are sensitive to the length of the
+  arguments, that is, the length of a line written by LINEOUT is the
+  same as the length of the string returned by successful LINEIN of the
+  line.
+\item
+  Some characters, call them line-banned characters, cannot reliably be
+  used as data for LINEIN/LINEOUT. If these are not used, LINEIN/LINEOUT
+  is reflective. If they are used, the result is not defined. The set of
+  characters which are line-barred is a property of the configuration.
+\item
+  The LINES(Stream, `N') function will return zero only when a
+  subsequent LINEIN (without positioning) is guaranteed to raise the
+  NOTREADY condition.
+\item
+  When a persistent stream is repositioned and written to with CHAROUT,
+  the previously written data is not lost, except for the data
+  overwritten by this latest CHAROUT.
+\item
+  When a persistent stream is repositioned and written to with LINEOUT,
+  the previously written data is not lost, except for the data
+  overwritten by this latest LINEOUT, which may leave lines partially
+  overwritten.
+\end{itemize}
+
+\subsection{CHARIN}\label{charin}
+
+CHARIN returns a string read from the stream named by the first
 argument.
 
-call CheckArgs `oSTREAM oOWHOLE\textgreater0 oOWHOLE\textgreater=0'
-
-if \#Bif\_ArgExists.1 then Stream else Stream \#StreamState.Stream =
-'\,' /* Argument 2 is positioning. */ if \#Bif\_ArgExists.2 then do
-\#Response = Config Stream Position(Stream, `CHARIN', \#Bif Arg.2)
-
-\#Bif Arg.1
-
-if left(\#Response, 1) == `R' then call Raise 40.41, 2, \#Bif\_ Arg.2 if
-left(\#Response, 1) == `T' then call Raise 40.42,Stream end
-
-/* Argument 3 is how many. \emph{/ if \#Bif\_ArgExists.3 then Count else
-Count if Count = 0 then do call Config Stream Charin Stream, `NULL' /}
-``Touch'' the stream \emph{/ return '! end /} The unit may be eight bits
-(as characters) or one character. */ call Config Stream Query Stream
-
-\#Bif Arg.3 1
-
-Mode = \#Outcome
-
-do until Count = 0
-
-\#Response = Config Stream Charin(Stream, `CHARIN')
-
-if left(\#Response, 1) == `N' then do if left (\#Response, 1) == `E'
-then \#StreamState.Stream = `ERROR' /* This call will return. */ call
-\#Raise `NOTREADY', Stream, substr(\#Response, 2) leave end
-
-r = r\textbar{} \textbar\#Outcome
-
-Count = Count-1
-
-end if Mode == `B' then do call Config B2C r r = \#Outcome end return r
+\lstinputlisting[language=rexx,label=charin.rexx,caption=charin.rexx]{charin.rexx}
 
@@
 
-call CheckArgs `oSTREAM oANY oWHOLE\textgreater0'
-
-if \#Bif\_ArgExists.1 then Stream else Stream
-
-\#Bif Arg.1
-
-\#StreamState.Stream = '\,' if \#Bif ArgExists.2 \& \#Bif\_ArgExists.3
-then do /* Position to end of stream. */ \#Response = Config Stream
-Close (Stream) if left (\#Response,1) == `T' then call Raise
-40.42,Stream return 0 end
-
-if \#Bif\_ArgExists.3 then do /* Explicit positioning. */ \#Response =
-Config Stream Position(Stream, `CHAROUT', \#Bif Arg.3)
-
-if left (\#Response,1) == `T' then call Raise 40.42,Stream if
-left(\#Response, 1) == `R' then call Raise 40.41, 3, \#Bif\_ Arg.3 end
-
-if \#Bif\_ArgExists.2 \textbar{} \#Bif\_Arg.2 == '\,' then do call
-Config Stream \_Charout Stream, `NULL' /* ``Touch'' the stream */ return
-0 end
-
-String = \#Bif Arg.2 call Config Stream Query Stream Mode = \#Outcome if
-Mode == `B' then do call Config C2B String String = \#Outcome Stride = 8
-Residue = length(String)/8 end else do Stride = 1 Residue = length
-(String) end
-
-Cursor = 1 do while Residue\textgreater0 Piece = substr (String, Cursor,
-Stride) Cursor = Cursor+Stride call Config Stream Charout Stream, Piece
-if left (\#Response, 1) == `N' then do if left (\#Response, 1) == `E'
-then \#StreamState.Stream = `ERROR' call \#Raise `NOTREADY', Stream,
-substr(\#Response, 2)
-
-return Residue end Residue = Residue - 1 end return 0
+\lstinputlisting[language=rexx,label=charout.rexx,caption=charout.rexx]{charout.rexx}
 
@@
 
-call CheckArgs `oSTREAM oCN'
-
-if \#Bif\_ArgExists.1 then Stream = \#Bif\_Arg.1 else Stream = '\,'
-
-if \#Bif\_ArgExists.2 then Option = \#Bif Arg.2 else Option = `N'
-
-call Config Stream Count Stream, `CHARS', Option return \#Outcome
+\lstinputlisting[language=rexx,label=chars.rexx,caption=chars.rexx]{chars.rexx}
 
@@
 
-call CheckArgs `oSTREAM oOWHOLE\textgreater0 oOWHOLE\textgreater=0'
-
-if \#Bif\_ArgExists.1 then Stream else Stream \#StreamState.Stream =
-'\,' if \#Bif\_ArgExists.2 then do \#Response = Config Stream
-Position(Stream, `LINEIN', \#Bif Arg2) if left (\#Response, 1) `T' then
-call Raise 40.42,Stream if left (\#Response, 1) `R' then call Raise
-40.41, 2, \#Bif\_ Arg.2 end if \#Bif\_ArgExists.3 then Count \#Bif Arg.3
-else Count 1 if Count\textgreater1 then call Raise 40.39, Count if Count
-= 0 then do call Config Stream Charin Stream, `NULL' /* ``Touch'' the
-stream \emph{/ return `! end /* A configuration may recognise lines even
-in 'binary' mode. }/ call Config Stream Query Stream Mode = \#Outcome re
-ter t = \#Linein Position.Stream /* Config Stream Charin will alter
-\#Linein Position. */ do until t = \#Linein Position.Stream \#Response =
-Config Stream Charin(Stream, `LINEIN') if left(\#Response, 1) == `N'
-then do if left (\#Response, 1) == `E' then \#StreamState.Stream =
-`ERROR' call \#Raise `NOTREADY', Stream, substr(\#Response, 2) leave end
-r = r\textbar\textbar\#Outcome end if Mode == `B' then do call Config
-B2C r r= \#Outcome end return r
-
-\#Bif Arg.1
+\lstinputlisting[language=rexx,label=linein.rexx,caption=linein.rexx]{linein.rexx}
 
@@
 LINEOUT returns `1' or `0', indicating whether the second argument has
-been successfully written to the
-
-stream named by the first argument. A result of `1' means an
-unsuccessful write. call CheckArgs `oSTREAM oANY oWHOLE\textgreater0'
-
-if \#Bif\_ArgExists.1 then Stream else Stream
-
-\#Bif Arg.1
-
-\#StreamState.Stream = '\,'
-
-if \#Bif ArgExists.2 \& \#Bif\_ArgExists.3 then do
-
-/* Position to end of stream.
-
-*/
-
-\#Response = Config Stream Close (Stream) if left (\#Response,1) == `T'
-then call Raise 40.42,Stream
-
-return 0 end
-
-if \#Bif\_ArgExists.3 then do
-
-\#Response = Config Stream Position(Stream, `LINEOUT', \#Bif Arg.3)
-
-if left(\#Response, 1) == `T' then call Raise 40.42,Stream if left
-(\#Response, 1) == `R' then call Raise 40.41, 3, \#Bif\_Arg.3 end
-
-if \#Bif ArgExists.2 then do
-
-call Config Stream \_Charout Stream, '\,' /* ``Touch'' the stream */
-
-return 0 end
-
-String \#Bif Arg.2 Stride 1 call Config Stream Query Stream Mode =
-\#Outcome if Mode == `B' then do call Config C2B String String =
-\#Outcome Stride = 8 Residue = length(String)/8 end else do Stride = 1
-Residue = length(String) end Cursor = 1 do while Residue \textgreater{}
-0
-
-Piece = substr (String, Cursor, Stride)
-
-Cursor = Cursor+Stride
-
-call Config Stream Charout Stream, Piece then do
-
-then \#StreamState.Stream = `ERROR'
-
-call \#Raise `NOTREADY', Stream, substr(\#Response, 2)
-
-if left(\#Response, 1) == `N' if left (\#Response, 1) == `E' return 1
-end Residue = Residue-1 end call Config Stream Charout Stream, return 0
-\#\#\# LINES
+been successfully written to the stream named by the first argument. A
+result of `1' means an unsuccessful write.
+
+\lstinputlisting[language=rexx,label=lineout.rexx,caption=lineout.rexx]{lineout.rexx}
+
+\subsection{LINES}\label{lines}
 
@@
 
-call CheckArgs `oSTREAM oCN'
-
-if \#Bif\_ArgExists.1 then Stream else Stream if \#Bif\_ ArgExists.2
-then Option else Option
-
-Call Config Stream Count Stream, return \#Outcome
+\lstinputlisting[language=rexx,label=lines.rexx,caption=lines.rexx]{lines.rexx}
 
@@
 
-\#Bif Arg.1 ter \#Bif\_Arg.2 tint
-
-LINES', Option
-
-QUALIFY returns a name for the stream named by the argument. The two
-names are currently associated with the same resource and the result of
-QUALIFY may be more persistently associated with
-
-that resource. call CheckArgs `oSTREAM'
-
-if \#Bif\_ArgExists.1 then Stream else Stream
-
-\#Bif Arg.1 \#Response = Config Stream Qualified (Stream) return
-\#Outcome
+\lstinputlisting[language=rexx,label=qualify.rexx,caption=qualify.rexx]{qualify.rexx}
 
@@
 
-/* Third argument is only correct with `C' */
-
-if \#Bif ArgExists.2 \& translate(left(\#Bif Arg.2, 1)) == `C' then
-ArgData = `rSTREAM rCDS rANY'
-
-else ArgData = `rSTREAM oCDS'
-
-call CheckArgs ArgData
-
-Stream = \#Bif\_Arg.1
-
-if \#Bif\_ ArgExists.2 then Operation = \#Bif\_Arg.2 else Operation =
-`S' Select when Operation == `C' then do
-
-call Config Stream Command Stream, \#Bif Arg.3 return \#Outcome
-
-end when Operation == `D' then do \#Response = Config Stream State
-(Stream) return substr(\#Response, 2) end when Operation == `S' then do
-if StreamState.Stream == `ERROR' then return `ERROR' \#Response = Config
-Stream State (Stream)
-
-`N' then return `READY' `U' then return `UNKNOWN'
-
-if left (\#Response, if left (\#Response, return `NOTREADY' end
-
-end
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-  ==
-\item
-  ==
-\end{enumerate}
+\lstinputlisting[language=rexx,label=stream.rexx,caption=stream.rexx]{stream.rexx}
 
@@
 
-call CheckArgs `oBDEMNOSUW oANY oOBDENOSU oSEP oSEP' /* If the third
-argument is given then the second is mandatory. */ if \#Bif\_ArgExists.3
-\& \#Bif\_ArgExists.2 then
-
-call Raise 40.19, '\,', \#Bif Arg.3
-
-if \#Bif\_ ArgExists.1 then Option else Option
-
-\#Bif Arg.1 tint
-
-/* The date/time is `frozen' throughout a clause. \emph{/ if
-\#ClauseTime.\#Level == '\,' then do \#Response = Config Time ()
-\#ClauseTime.\#Level = \#Time \#ClauseLocal.\#Level = \#Time +
-\#Adjust\textless Index ``\#Adjust'' \# ``\,'' \textgreater{} end /}
-English spellings are used, even if messages not in English are used. */
-Months = `January February March April May June July', `August September
-October November December' WeekDays = `Monday Tuesday Wednesday Thursday
-Friday Saturday Sunday'
-
-/* If there is no second argument, the current date is returned. */ if
-\#Bif ArgExists.2 then return DateFormat (\#ClauseLocal.\#Level, Option)
-
-/* If there is a second argument it provides the date to be converted.
-*/
-
-Value = \#Bif\_Arg.2 if \#Bif\_ ArgExists.3 then InOption else InOption
-if Option == `S' then OutSeparator else OutSeparator if
-\#Bif\_ArgExists.4 then do
-
-if OutSeparator == `x' then call OutSeparator = \#Bif.Arg.4 end
-
-if InOption == `S' then InSeparator
-
-else InSeparator if \#Bif\_ArgExists.5 then do
-
-\#Bif\_Arg.3 'nt
-
-translate (Option, ``xx/x //x'',``BDEMNOUW'')
-
-Raise 40.46, Option, 4
-
-translate(InOption,``xx/ //'',``BDENOU'')
-
-if InSeparator == `x' then call Raise 40.46, InOption, 5
-
-InSeparator = \#Bif.Arg.5 end /* First try for Year Month Day */ Logic =
-`NS' select when InOption == `N' then do if InSeparator == '\,' then do
-
-if length(Value)\textless9 then return Year = right (Value, 4)
-
-MonthIs = substr (right (Value,7),1,3) Day = left (Value, length (Value)
--7)
-
-end else
-
-parse var Value Day (InSeparator) MonthIs (InSeparator) Year
-
-do Month = 1 to 12 if left(word(Months, Month), 3)
-
-== MonthIs then leave
-
-parse var Value Year (InSeparator) Month (InSeparator) Day
-
-end Month
-
-end
-
-when InOption == `S' then if InSeparator == '\,' then
-
-parse var Value Year +4 Month +2 Day
-
-else
-
-otherwise Logic = `EOU' /* or BD */
-
-end
-
-/* Next try for year without century */
-
-parse var Value Day (InSeparator) Month (InSeparator) YY
-
-parse var Value YY (InSeparator) Month (InSeparator) Day
-
-parse var Value Month (InSeparator) Day (InSeparator) YY
-
-if logic = `EOU' then Select when InOption == `E' then if InSeparator ==
-'\,' then parse var Value Day +2 Month +2 YY else when InOption == `O'
-then if InSeparator == '\,' then parse var Value YY +2 Month +2 Day else
-when InOption == `U' then if InSeparator == '\,' then parse var Value
-Month +2 Day +2 YY else otherwise Logic = `BD' end if Logic = `EOU' then
-do
-
-/* The century is assumed, on the basis of the current year. */
-
-if datatype(YY,`W')=0 then return
-
-YearNow = left(`DATE'(`S'),4)
-
-Year = YY
-
-do while Year \textless{} YearNow-50 Year = Year + 100
-
-end
-
-end /* Century assumption */
-
-if Logic \textless\textgreater{} `BD' then do /* Convert Month \& Day to
-Days of year. \emph{/ if datatype(Month,`W')=0 \textbar{}
-datatype(Day,`W')=0 \textbar{} datatype(Year,`W')=0 then return Days =
-word(`0 31 59 90 120 151 181 212 243 273 304 334',Month), +
-(Month\textgreater2)}Leap(Year) + Day-1
-
-end else
-
-if datatype(Value,`W')=0 then
-
-return
-
-if InOption == `D' then do
-
-Year = left(`DATE' (`S'),4)
-
-Days = Value - 1 /* `D' includes current day */ end
-
-/* Convert to BaseDays */ if InOption \textless\textgreater{} `B' then
-
-BaseDays = (Year-1)*365 + (Year-1)\%4 - (Year-1)\%100 + (Year-1)\%400 +
-Days else
-
-Basedays = Value
-
-/* Convert to microseconds from 0001 \emph{/ Micro = BaseDays } 86400 *
-1000000
-
-/* Reconvert to check the original. (eg for Month = 99) */
-
-if DateFormat (Micro,InOption,InSeparator) == Value then call Raise
-40.19, Value, InOption
-
-return DateFormat (Micro, Option, OutSeparator)
-
-DateFormat: /* Convert from microseconds to given format. */ parse value
-Time2Date(arg(1)) with, Year Month Day Hour Minute Second Microsecond
-Base Days
-
-select
-
-when arg(2) == `B' then
-
-return Base when arg(2) == `D' then
-
-return Days when arg(2) == `E' then
-
-return right (Day,2,`0') (arg(3)) right (Month,2,`0') (arg(3)) right
-(Year,2,`0') when arg(2) == `M' then
-
-return word (Months ,Month) when arg(2) == `N' then
-
-return (Day) (arg(3)) left (word(Months,Month),3) (arg(3))right
-(Year,4,`0') when arg(2) == `O' then
-
-return right (Year,2,`0') (arg(3))right (Month,2,`0') (arg(3))right
-(Day,2,`0') when arg(2) == `S' then
-
-return right (Year,4,`0') (arg(3))right (Month,2,`0') (arg(3))right
-(Day,2,`0') when arg(2) == `U' then
-
-return right (Month,2,`0') (arg(3)) right (Day,2,`0') (arg(3)) right
-(Year,2,`0') otherwise /* arg(2) == `W' */
-
-return word (Weekdays, 1+Base//7)
-
-end
+\lstinputlisting[language=rexx,label=date.rexx,caption=date.rexx]{date.rexx}
 
@@
 
-call CheckArgs '\,'
-
-\#Response = Config Queued() return \#Outcome
+\lstinputlisting[language=rexx,label=queued.rexx,caption=queued.rexx]{queued.rexx}
 
@@
 
-call CheckArgs `oWHOLE\textgreater=0 oWHOLE\textgreater=0
-oWHOLE\textgreater=0'
-
-if \#Bif\_Arg.0 = 1 then do Minimum = 0
-
-Maximum = \#Bif Arg.1 if Maximum\textgreater100000 then call Raise
-40.31, Maximum
-
-end else do if \#Bif\_ArgExists.1 then Minimum = \#Bif Arg.1 else
-Minimum = 0 if \#Bif\_ArgExists.2 then Maximum = \#Bif\_Arg.2 else
-Maximum = 999
-
-end
-
-if Maximum-Minimum\textgreater100000 then call Raise 40.32, Minimum,
-Maximum
-
-if Maximum-Minimum\textless0 then call Raise 40.33, Minimum, Maximum
-
-if \#Bif\_ArgExists.3 then call Config Random Seed \#Bif\_Arg.3 call
-Config Random Next Minimum, Maximum return \#Outcome
+\lstinputlisting[language=rexx,label=random.rexx,caption=random.rexx]{random.rexx}
 
@@
 
-call CheckArgs `oCEHLMNORS oANY oCHLMNS' /* If the third argument is
-given then the second is mandatory. */ if \#Bif\_ArgExists.3 \&
-\#Bif\_ArgExists.2 then
-
-call Raise 40.19, '\,', \#Bif Arg.3
-
-if \#Bif\_ ArgExists.1 then Option else Option
-
-\#Bif Arg.1 tint
-
-/* The date/time is `frozen' throughout a clause. */
-
-if \#ClauseTime.\#Level == '\,' then do \#Response = Config Time ()
-\#ClauseTime.\#Level = \#Time \#ClauseLocal.\#Level = \#Time +
-\#Adjust\textless Index ``\#Adjust'' \# ``\,'' \textgreater{} end
-
-/* If there is no second argument, the current time is returned. */ if
-\#Bif ArgExists.2 then return TimeFormat (\#ClauseLocal.\#Level, Option)
-
-/* If there is a second argument it provides the time to be converted.
-*/ if pos(Option, `ERO') \textgreater{} 0 then call Raise 40.29, Option
-InValue = \#Bif Arg.2
-
-if \#Bif\_ArgExists.3 then InOption = \#Bif\_Arg.3 else InOption =
-
-HH = 0
-
-MM = 0
-
-ss = 0
-
-HourAdjust = 0
-
-select
-
-when InOption == `C' then do parse var InValue HH `:' . +1 MM +2 XX
-
-if HH = 12 then
-
-125 HH = 0
-
-if XX == `pm' then HourAdjust = 12 end when InOption == `H' then HH =
-InValue when InOption == `L' \textbar{} InOption == `N' then parse var
-InValue HH `:' MM `:' SS when InOption == `M' then MM = InValue
-otherwise SS = InValue
-
-end
-
-if datatype(HH,`W')=0 \textbar{} datatype(MM,`W')=0 \textbar{}
-datatype(SS,`N')=0 then call Raise 40.19, InValue, InOption
-
-HH = HH + HourAdjust
-
-/* Convert to microseconds */
-
-Micro = trunc((((HH * 60) + MM) * 60 + SS) * 1000000)
-
-/* There is no special message for time-out-of-range; the bad-format
-
-message is used. */
-
-if Micro\textless0 \textbar{} Micro \textgreater{} 24\emph{3600}1000000
-then call Raise 40.19, InValue, InOption
-
-/* Reconvert to check the original. (eg for hour = 99) */
-
-if TimeFormat (Micro,InOption) == InValue then call Raise 40.19,
-InValue, InOption
-
-return TimeFormat (Micro, Option)
-
-end /* Conversion */
-
-TimeFormat: procedure /* Convert from microseconds to given format.
-\emph{/ /} The day will be irrelevant; actually it will be the first day
-possible. \emph{/ x = Time2Date2(arg(1)) parse value x with Year Month
-Day Hour Minute Second Microsecond Base Days select when arg(2) == `C'
-then select when Hour\textgreater12 then return Hour-12':`right
-(Minute,2,'0')`pm' when Hour=12 then return `12:'right(Minute,2,'0')`pm'
-when Hour\textgreater0 then return Hour':`right (Minute,2,'0')`am' when
-Hour=0 then return `12:'right(Minute,2,'0')`am' end when arg(2) == `H'
-then return Hour when arg(2) == `L' then return right
-(Hour,2,`0')`:'right (Minute,2,'0')`:'right(Second,2,'0'),
-\textbar\textbar{} `.'right (Microsecond,6,'0') when arg(2) == `M' then
-return 60}Hour+Minute when arg(2) == `N' then return right
-(Hour,2,`0')`:'right (Minute,2,'0')`:'right(Second,2,'0') otherwise /*
-arg(2) == `S' \emph{/ return 3600}Hour+60*Minute+Second end
-
-Time2Date: /* These are checks on the range of the date. */ if arg(1)
-\textless{} 0 then call Raise 40.19, InValue, InOption if arg(1)
-\textgreater= 315537897600000000 then call Raise 40.19, InValue,
-InOption return Time2Date2 (arg(1))
+\lstinputlisting[language=rexx,label=time.rexx,caption=time.rexx]{time.rexx}
 
 10.6.1 VALUE VALUE returns the value of the symbol named by the first
-argument, and optionally assigns it a new
-
-value.
-
-`rANY oOANY oANY'
-
-if \#Bif ArgExists.3 then ArgData `rSYM oANY oOANY'
-
-else ArgData call CheckArgs ArgData
-
-Subject = \#Bif Arg.1
-
-if \#Bif\_ArgExists.3 then do /* An external pool, or the reserved pool.
-\emph{/ /} The reserved pool uses a null string as its pool identifier.
-*/ Pool = \#Bif\_Arg.3
-
-if Pool == '\,' then do Subject = `.' \textbar\textbar{} translate
-(Subject) /* The dot on the name is implied. \emph{/ Value =
-.environment {[}Subject{]} /} Was the translate redundant? */
-
-if \#Bif ArgExists.2 then .environment {[}Subject{]} = \#Bif\_ Arg.2
-return Value end
-
-/* Fetch the original value */
-
-\#Response = Config Get (Pool, Subject)
-
-\#Indicator = left (\#Response,1)
-
-if \#Indicator == `F' then call Raise 40.36, Subject if \#Indicator ==
-`P' then
-
-call Raise 40.37, Pool Value = \#Outcome if \#Bif\_ArgExists.2 then do
-/* Set the new value. \emph{/ \#Response = Config Set(Pool,Subject,\#Bif
-Arg.2) if \#Indicator == `P' then call Raise 40.37, Pool if \#Indicator
-== `F' then call Raise 40.36, Subject end /} Return the original value.
-\emph{/ return Value end /} Not external \emph{/ Subject =
-translate(Subject) /} See nnn \emph{/ Pp = pos(Subject, `.') if p = 0
-\textbar{} p = length(Subject) then do /} Not compound \emph{/
-\#Response = Var Value(\#Pool, Subject, `0') /} The caller, in the code
-of the standard, may need to test whether the Subject was dropped.
-\emph{/ \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
-ArgExists.2 then \#Response = Var Set(\#Pool, Subject, `0', \#Bif\_
-Arg.2) return Value end /} Compound \emph{/ Expanded = left(Subject,p-1)
-/} The stem \emph{/ do forever Start = p+l1 Pp = pos(Subject,`.',Start)
-if p = 0 then p = length(Subject) Item = substr(Subject,Start,p-Start)
-/} Tail component symbol \emph{/ if Item=='\,' then if
-pos(left(Item,1),`0123456789') = 0 then do \#Response = Var
-Value(\#Pool, Item, `0') Item = \#Outcome end /} Add tail component. */
-Expanded = Expanded'.'Item end \#Response = Var Value(\#Pool, Expanded,
-`1') \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
-ArgExists.2 then \#Response = Var Set(\#Pool, Expanded, `1', \#Bif\_
-Arg.2)
-
-return Value
+argument, and optionally assigns it a new value.
+
+\lstinputlisting[language=rexx,label=value.rexx,caption=value.rexx]{value.rexx}
 
@@
 
-call CheckArgs '\,' \#Response = Config Queued() return \#Outcome
+\lstinputlisting[language=rexx,label=queued.rexx,caption=queued.rexx]{queued.rexx}
 
@@
 
-call CheckArgs `oWHOLE\textgreater=0 oWHOLE\textgreater=0
-oWHOLE\textgreater=0'
-
-if \#Bif\_Arg.0 = 1 then do Minimum = 0 Maximum = \#Bif\_Arg.1 if
-Maximum\textgreater100000 then call Raise 40.31, Maximum
-
-end else do if \#Bif\_ArgExists.1 then Minimum = \#Bif Arg.1 else
-Minimum = 0 if \#Bif\_ArgExists.2 then Maximum = \#Bif\_Arg.2 else
-Maximum = 999
-
-end
-
-if Maximum-Minimum\textgreater100000 then call Raise 40.32, Minimum,
-Maximum
-
-if Maximum-Minimum\textless0 then call Raise 40.33, Minimum, Maximum
-
-if \#Bif\_ArgExists.3 then call Config Random Seed \#Bif\_Arg.3 call
-Config Random Next Minimum, Maximum return \#Outcome
+\lstinputlisting[language=rexx,label=random.rexx,caption=random.rexx]{random.rexx}
 
@@
 
-call CheckArgs `oCEHLMNORS oANY oCHLMNS' /* If the third argument is
-given then the second is mandatory. */ if \#Bif\_ArgExists.3 \&
-\#Bif\_ArgExists.2 then
-
-call Raise 40.19, '\,', \#Bif Arg.3
-
-if \#Bif\_ ArgExists.1 then Option else Option
-
-\#Bif Arg.1 tint
-
-/* The date/time is `frozen' throughout a clause. */
-
-if \#ClauseTime.\#Level == '\,' then do \#Response = Config Time ()
-\#ClauseTime.\#Level = \#Time \#ClauseLocal.\#Level = \#Time +
-\#Adjust\textless Index ``\#Adjust'' \# ``\,'' \textgreater{} end
-
-/* If there is no second argument, the current time is returned. */ if
-\#Bif ArgExists.2 then return TimeFormat (\#ClauseLocal.\#Level, Option)
-
-/* If there is a second argument it provides the time to be
-
-converted. */ if pos(Option, `ERO') \textgreater{} 0 then
-
-128 call Raise 40.29, Option InValue = \#Bif Arg.2
-
-if \#Bif\_ArgExists.3 then InOption = \#Bif\_Arg.3 else InOption = `N'
-HH = 0 MM = 0 ss = 0 HourAdjust = 0 select when InOption == `C' then do
-parse var InValue HH `:' . +1 MM +2 XX if XX == `pm' then HourAdjust =
-12 end when InOption == `H' then HH = InValue when InOption == `L'
-\textbar{} InOption == `N' then parse var InValue HH `:' MM `:' SS when
-InOption == `M' then MM = InValue otherwise SS = InValue end
-
-if \datatype(HH,`W') \textbar{} \datatype(MM,`W') \textbar{}
-\datatype(SS,`N') then call Raise 40.19, InValue, InOption
-
-HH = HH + HourAdjust
-
-/* Convert to microseconds */
-
-Micro = trunc((((HH * 60) + MM) * 60 + SS) * 1000000)
-
-/* Reconvert to check the original. (eg for hour = 99) */
-
-if TimeFormat (Micro,InOption) == InValue then call Raise 40.19,
-InValue, InOption
-
-return TimeFormat (Micro, Option)
-
-end /* Conversion */
-
-TimeFormat: /* Convert from microseconds to given format. */ parse value
-Time2Date(arg(1)) with, Year Month Day Hour Minute Second Microsecond
-Base Days select when arg(2) == `C' then if Hour\textgreater12 then
-return Hour-12':`right (Minute,2,'0')`pm'
-
-else return Hour':`right (Minute,2,'0')`am' when arg(2) == `E'
-\textbar{} arg(2) == `R' then do
-
-/* Special case first time */
-
-if \#StartTime.\#Level == '\,' then do \#StartTime.\#Level
-\#ClauseTime.\#Level return `0O'
-
-end Output = \#ClauseTime.\#Level-\#StartTime. \#Level if arg(2) == `R'
-then
-
-\#StartTime.\#Level = \#ClauseTime.\#Level return Output * 1E-6 end /* E
-or R */ when arg(2) == `H' then return Hour when arg(2) == `L' then
-return right (Hour,2,`0')`:'right (Minute,2,'0')`:'right(Second,2,'0'),
-\textbar\textbar{} `.'right (Microsecond,6,'0')
-
-when arg(2) == `M' then return 60\emph{Hour+Minute when arg(2) == `N'
-then return right (Hour,2,`0')`:'right
-(Minute,2,'0')`:'right(Second,2,'0') when arg(2) == `O' then return
-trunc(\#ClauseLocal.\#Level - \#ClauseTime.\#Level) otherwise /} arg(2)
-== `S' \emph{/ return 3600}Hour+60*Minute+Second end 10.6.5 VALUE
+\lstinputlisting[language=rexx,label=time.rexx,caption=time.rexx]{time.rexx}
+
+10.6.5 VALUE
 
@@
 
-`rANY oOANY oANY' `rSYM OANY oANY'
-
-if \#Bif ArgExists.3 then ArgData else ArgData call CheckArgs ArgData
-
-129 130
-
-Subject = \#Bif Arg.1
-
-if \#Bif\_ArgExists.3 then do /* An external pool, or the reserved pool.
-\emph{/ /} The reserved pool uses a null string as its pool identifier.
-*/ Pool = \#Bif\_Arg.3
-
-if Pool == '\,' then do Subject = `.' \textbar\textbar{} translate
-(Subject) /* The dot on the name is implied. \emph{/ Value =
-.environment {[}Subject{]} /} Was the translate redundant? */
-
-if \#Bif ArgExists.2 then .environment {[}Subject{]} = \#Bif\_ Arg.2
-return Value end
-
-/* Fetch the original value */
-
-\#Response = Config Get (Pool, Subject)
-
-\#Indicator = left (\#Response,1)
-
-if \#Indicator == `F' then call Raise 40.36, Subject if \#Indicator ==
-`P' then
-
-call Raise 40.37, Pool Value = \#Outcome if \#Bif\_ArgExists.2 then do
-/* Set the new value. \emph{/ \#Response = Config Set(Pool,Subject,\#Bif
-Arg.2) if \#Indicator == `P' then call Raise 40.37, Pool if \#Indicator
-== `F' then call Raise 40.36, Subject end /} Return the original value.
-\emph{/ return Value end /} Not external \emph{/ Subject =
-translate(Subject) /} See nnn \emph{/ Pp = pos(Subject, `.') if p = 0
-\textbar{} p = length(Subject) then do /} Not compound \emph{/
-\#Response = Var Value(\#Pool, Subject, `0') /} The caller, in the code
-of the standard, may need to test whether the Subject was dropped.
-\emph{/ \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
-ArgExists.2 then \#Response = Var Set(\#Pool, Subject, `0', \#Bif\_
-Arg.2) return Value end /} Compound \emph{/ Expanded = left(Subject,p-1)
-/} The stem \emph{/ do forever Start = p+l1 Pp = pos(Subject,`.',Start)
-if p = 0 then p = length(Subject) Item = substr(Subject,Start,p-Start)
-/} Tail component symbol \emph{/ if Item=='\,' then if
-pos(left(Item,1),`0123456789') = 0 then do \#Response = Var
-Value(\#Pool, Item, `0') Item = \#Outcome end /} Add tail component. */
-Expanded = Expanded'.'Item end \#Response = Var Value(\#Pool, Expanded,
-`1') \#Indicator = left(\#Response, 1) Value = \#Outcome if \#Bif\_
-ArgExists.2 then \#Response = Var Set(\#Pool, Expanded, `1', \#Bif\_
-Arg.2)
-
-return Value
+\lstinputlisting[language=rexx,label=value.rexx,caption=value.rexx]{value.rexx}

### 2024-04-12 ARB/standards/work-in-progress/document/tex/standard/directives.tex
updates

@@
 Return \texttt{\textquotesingle{}1\textquotesingle{}} if the
-\emph{production} identitied by \emph{Identifier} contained a
+\emph{production} identified by \emph{Identifier} contained a
 \emph{bnf\_primary} identified by \texttt{BNF\_primary}, otherwise
@@
 \lstinputlisting[language=rexx,label=configotherblankcharacters.rexx,caption=configotherblankcharacters.rexx]{configotherblankcharacters.rexx}
-our model are only distinquished by the values within their pool so we
-can construct the builtin classes incomplete and then complete them with
-directives.}
+model are only distinquished by the values within their pool so we can
+construct the builtin classes incomplete and then complete them with
+directives.\_
 
-\emph{Can we initialize the methods of .nil by directives?}
+\_Can we initialize the methods of .nil by directives?\_
 

### 2024-04-26 ARB/standards/work-in-progress/document/tex/boilerplate/bookmeta.tex
tex skeleton
wip
paragraph numbering of both documents now equal

@@
 Netherlands.\\[1cm]
-The RexxLA Symposium Series is registered under ISSN 1534-8954\\
-The \edition{} edition is registered under ISBN \isbn \\[1cm]
+
+This \edition{} edition is registered under ISBN \isbn \\[1cm]
 \psset{unit=1in}
-\begin{pspicture}(3.5,1in)
-  \psbarcode{1534-8954}{includetext guardwhitespace}{issn}
-\end{pspicture}
 \begin{pspicture}(3.5,1in)
@@
 \begin{description}
-\item[2023-03-14] First printing
+\item[2024-05-31] First printing
 \end{description}
